### 2015-12-03 writeup/local_pca_paper.tex
initial skeleton

\documentclass{article}
\usepackage{fullpage}

\begin{document}

Put the text in here!

\end{document}

### 2016-01-28 writeup/local_pca_paper.tex
example
latex
outline of intro

@@
 \documentclass{article}
+\usepackage{amsmath,amssymb}
 \usepackage{fullpage}
+\usepackage[authoryear]{natbib}
+\usepackage{graphicx}
+\usepackage[hidelinks]{hyperref}
+%\usepackage{latexml} % for \iflatexml
+\usepackage{color}
+\usepackage[dvipsnames]{xcolor}
+
+\bibliographystyle{plainnat}
+
+\title{Population structure along the genome}
+\author{
+Han Li
+\\and\\
+Peter Ralph
+}
 
@@
 
-Put the text in here!
+Other phrases that might go into a better title?
+\begin{itemize}
+    \item How selection affects relatedness
+    \item Inhomogeneous kinship
+    \item There's more than one kinship matrix
+    \item Local PCA reveals inversions and the effects of selection
+    \item Unsupervised learning of genomic patterns of relatedness
+\end{itemize}
+
+%%%%%%%%%%%%
+\section{Introduction}
+
+Catchy start: the kinship matrix goes back to almost Mendel and is essential in GWAS;
+however, it is well-known that actual relatednesses have a lot of noise about the expected value,
+and depend on where on the genome you look;
+this is why scans for selective sweeps work.
+
+Review of kinship matrix: 
+it's either an expected kinship, given the pedigree;
+or an estimated genome-wide average.
+Wright's path coefficients \citep{wright1943isolation}.
+Why it helps with confounding for GWAS.
+Graham \& Vince's paper, maybe.
+Like IBD, is only well-defined in a known pedigree, up to the founders.
+
+Kinship matrices differ for sex chromosomes and the like.
+
+Review of selection causing differential patterns along the genome.
+Locally everything is treelike (gene trees);
+kinship matrix is an average of these (write equation for this).
+Selective sweeps cause local recent ancestry/short trees.
+Balancing selection causes deep trees.
+Background selection, shallow ones.
+Refs: hitchhiking \citep{maynardsmith1974hitchhiking},
+selection on LD \citep{mcvean2007structure},
+
+Review of methods looking along the genome:
+argweaver, HMM between species, ???
+
+What is "population structure"?
+Asks which "populations" are closely related, more diverged, how much diversity do they harbor.
+Often geographic.
+Vital in exploratory data analysis.
+It is a summary of kinship: 
+lack of migration between pops causes a deficit in connections through the pedigree,
+and so affects kinship.
+
+Review of methods for visualizing pop structure:
+PCA, etcetera.
+Include maps of homozygosity.
+
+Other semi-related stuff:
+estimation of covariance matrices;
+local pca(?);
+
+\subsection{Applications}
+
+humans:
+effects of selection \citep{sabeti2006positive}
+dataset(s)
+
+drosophila:
+effects of selection \citep{sella2009pervasive},
+inversions,
+dataset(s)
+
+medicago:
+
+%%%%%%%%%%%%
+\section{Methods}
+
+%%%%%%%%%%%%
+\section{Results}
+
+%%%%%%%%%%%%
+\section{Discussion}
+
+%%%%%%%%%%%%%
+\bibliography{references}
 

### 2016-01-29 writeup/local_pca_paper.tex
added more refs

@@
 Like IBD, is only well-defined in a known pedigree, up to the founders.
+Kinship and confounding reviewed in \citet{astle2009population}.
 
@@
 Background selection, shallow ones.
+Extreme examples of free gene flow in some places between species: e.g. Heliconius.
+Introgression may be nonuniform, e.g. neanderthal, others.
 Refs: hitchhiking \citep{maynardsmith1974hitchhiking},
-selection on LD \citep{mcvean2007structure},
+\citet{kim2011hitchhiking} study hitchhiking in a spatially subdivided population,
+\citet{mcvean2007structure} looks at the effect of selection on LD.
+\citet{barton2000genetic} reviews hitchhiking.
+\citet{bierne2010distinctive} discusses how hitchhiking effect decreases with geographic distance.
+\citet{charlesworth2003review} reviews patterns of diversity, relating spatial structure to effects of selection.
 
@@
 and so affects kinship.
+Wright defined $F_{ST}$ in \citep{wright1949genetical}, and says ``It has probably occurred to the reader that the coefficientof inbreeding may mean very different
+things in different cases.''
 
 Review of methods for visualizing pop structure:
-PCA, etcetera.
-Include maps of homozygosity.
-
+PCA, structure \citep{falush2003inference}, EEMS, \citep{petkova2014visualizing}, \citep{yang2012modelbased},
+Maps of heterozygosity \citep{ramachandran2005support}.
+Genealogical interpretation of PCA by \citet{mcvean2009genealogical}.
 Other semi-related stuff:
@@
 
+We apply this to three species.
+
 humans:
-effects of selection \citep{sabeti2006positive}
+effects of selection are seen \citep{sabeti2006positive}.
+\citet{duforetfrebourg2015detecting} showed that variants known to underly locally adaptive traits correlated most strongly with PCs.
+\citet{tian2008analysis} noted that removing the segregating inversion on chromosome 8 changed the PCA for a set of 1400 Europeans
 dataset(s)

### 2016-02-10 local_pca_proposal.tex
Create local_pca_proposal.tex

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{setspace}
\title{Using Local PCA to Investigate Population Structure Along the Genome}
\author{Han Li, Peter Ralph}
%\date{}							% Activate to display a given date or no date

\begin{document}
\maketitle
%\section{}
%\subsection{}
\doublespacing
\section{Introduction}

About 37 years ago, Cavalli-Sforza et al. first applied Principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \cite{key1}. Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data. PCA results are derived from the covariance matrix of genotype matrix, which describes population structure, demographic history, and pedigree tree and so on. The results of PCA can be directly related to the underlying genealogical history of the samples, such as coalescence time (time to most recent common ancestor) and migration rate between populations \cite{key1,key2,key3}. Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and plots of first two Principal components (PCs) can mimic the samples' geographic origin to some extent. The closeness of two samples in PC plots represents the similarity of their genetic information. Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \cite{key4,key6}.

\noindent  Most time, PCA used in genome-wide association study (GWAS) is for stratification correction \cite{key7}. However, for different part of genome, they have different genetic features. First, each site of DNA may have different gene tree. For a DNA region, if individuals have SNPs closer in gene tree, they tends to be more close in PC projections for that region. Different DNA segments may have different gene tree for their SNPs and would result in different population structure for those segments. Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. Third, if chromosome inversion exists, the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints, thus results in different genome structure and population structure. Investigating the genomic variation along the genome can help us to have a better understanding of the relation between genome structure and population structure. Therefore, we cut the chromosomes into windows with hundreds/thousands of SNPs and apply PCA locally on each window. In this project, we used SNP data for human, \textit{Medicago truncatula}, and whole genome sequencing data for \textit{Drosophila}.

\noindent Based on the principal components, we can estimate the similarity of population structure contained in each genome window. To visualize the closeness for each window with regard to the population structure, we constructed matrix based on the first two PCs for each window and got the pairwise Euclidean distance between those matrixes, then use multidimensional scaling to reduces the pairwise distance matrix to lower dimension while preserve the distance information between windows as well as possible \cite{key5}. To interpret the results of MDS, we combine known genome feature information for each species, such as the distribution of inversions, heterochromatin chunk and gene density along the genome. For different species, we got different explanation for the variation of population structure along the genome.

\section{Method}
\subsection{Recode the DNA sequence to a matrix consisting of 0,1,2 (and NA).}
For human SNP data from POPRES \cite{key8}, we use the allele that has highest frequency in the samples as the reference allele for each position. If an allele is same with the reference allele, we recode it as 0; if an allele is different from the reference allele, we recode it as 1; for positions that have missing data, we recode them as NA. Since human genome is diploid, we add the value for the two alleles and then one chromosome will eventually be recode to a sequence consisting of 0,1,2 (and NA). There are 3965 samples in total, (346 African-Americas; 73 Asians; 3187 Europeans; 359 Indian Asians); then the genome data for human is recoded to a matrix that has 3965 columns, each column for an individual's genotype. We process the data separately for the 22 autosomes in human. For \textit{Drosophila}, we first process the sequencing data (from DPGP and John pool's lab) to SNP data by eliminating the positions that have all the same alleles. Due to high density of missing data for some parts in the genome, we then delete the samples with more than 8\% NAs and positions with more than 20\%. The cutoff points 8\% and 20\% are determined from the corresponding distributions of NAs in samples and at positions. After we got the SNP data for \textit{Drosophila}, we recode it \textit{Drosophila}to matrices with 0,1,2 (and NA) for each chromosome arms (Chr2L, Chr2R, Chr3L, Chr3R, ChrX) similar to the process for human genome. Since the \textit{Drosophila} samples here are all homogenous, the matrices are indeed consisting of 0,1 (and NA). For \textit{Medicago truncatula}, we use the SNP data from Medicago Hapmap and recode it to matrices with 0,1,2 (and NA) for each of the 8 chromosomes.

\subsection{Cut each genome into windows}
We cut each recoded matrix into sub matrixes by that have the same columns but fewer rows than the original matrix. Then apply Principal Component Analysis (PCA) on each window. Here's a brief summary about how is PCA carried out on genomic data \cite{key3}. Starting with the recoded genotype matrix Z, where Z is a L*N matrix ( L is SNP number ; N is sample size), then zero-center the matrix Z to X to make the data rows have equal variance. Get the covariance matrix of X (denote as matrix C) and compute the eigenvectors and eigenvalues of the covariance matrix C. The \textit{i}th principal component is the \textit{i}th eigenvector of C.

\begin{equation}
X_{si}=Z_{si}-\frac{1}{n}\sum_{j=1}^{n}Z_{sj} 
\end{equation}
\begin{equation}
C=\frac{1}{n-1}XX^{T}
\end{equation}

\subsubsection{Genomic PCA on windows}
PCA on each window is getting the eigenvectors of covariance matrix of the cut recoded matrix. PCA plots (generally using PC2 against PC1) can show population structure, which describes how different samples are related. We want to investigate the variation of population structure along the genome. So we apply PCA on each window to check the population structure along the genome.
\subsubsection{Choose window length}
The window length should neither be too long or too short. The longer the windows, the more accurate is the estimate of population structure in that window. However, for better resolution, we need to find a length that is reasonably short. If we use the first principal component as a measure of population structure, then to choose a proper length for a
window, we need to find a balance between variance of the first principal components inside a window and that between windows. The variance between windows is estimated as mean variance of the first principal component for each window. The variance inside a window is estimated using the jackknife block when cutting the window into 10 equal size smaller windows \cite{key9}. Table 1 shows the comparison of variance within a window and that between windows for chromosome arms in \textit{Drosophila}. Finally, we choose 100 SNPs, 1000 SNPs and 10000 SNPs as window length for human, \textit{Drosophila}, and \textit{Medicago} separately.

Table 1.

\subsubsection{Similarity of population structure between windows}
We use the first two principal components (PCs) and the corresponding eigenvalues from PCA, and construct a new matrix with the two PCs for each window to stand for the population structure information. For example, the constructed matrixes for ith and jth window are as following. ($\lambda _{1i}$ and $\lambda _{2i}$ are the eigenvalues for the first two PCs for \textit{i}th window; $\lambda _{1j}$ and $\lambda _{2j}$ are the eigenvalues for the first two PCs for \textit{j}th window.)

\begin{equation}
M_{i}=\frac{\lambda _{1i}PC1_{i}PC1_{i}^{T}+\lambda _{2i}PC2_{i}PC2_{i}^{T}}{\lambda _{1i}+\lambda _{2i}} \qquad
M_{j}=\frac{\lambda _{1j}PC1_{j}PC1_{j}^{T}+\lambda _{2j}PC2_{j}PC2_{j}^{T}}{\lambda _{1j}+\lambda _{2j}}
\end{equation}

The Euclidean distance between the constructed matrices for windows stands for the
similarity of population structure for windows. Due to the property of eigenvectors,we
could use the following method to calculate the pairwise distance greatly saving time and space.

\begin{equation}
V_{1}=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} \qquad
V_{2}=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i}
\end{equation}

\begin{multline}
Distance_{ij} \\ =\sqrt{\left \{\left ( V_{1}\cdot V_{1}^{} \right )^{2}+\left ( V_{2}\cdot V_{2}^{} \right )^{2}+\left ( U_{1}\cdot U_{1}^{} \right )^{2} +\left ( U_{2}\cdot U_{2}^{} \right )^{2} -2\left [ \left ( V_{1}\cdot U_{1}^{} \right )^{2}+\left ( V_{1}\cdot U_{2}^{} \right )^{2}+\left ( V_{2}\cdot U_{1}^{} \right )^{2} +\left ( V_{2}\cdot U_{2}^{} \right )^{2}\right ]\right \}}
\end{multline}

Using this procedure, we get the pairwise distance matrix that says how similar population structure is in each pair of genomic windows.
\subsubsection{Visualize the pairwise distance matrix}
To do this, I use Multidimensional scaling (MDS), which is a visualization method commonly applied to distance matrixes. It can reduce the dimensionality of a distance matrix while preserve the distance information between objects as well as possible. The aimed dimension \lq\lq M\rq\rq can be set based on your need. We used M=1 and M=2 in our study. This allowed us to visualize the information in the distance matrix in one or two dimensional relation between windows' population structure.

\section{Results}
\textbf{In all these 3 species, PCA plots vary along the genome.}

\noindent Since PCA plots can show population structure, this shows that the population structure varies along the genome. Each PC plot comes from the covariance matrix of a different section of the genome, which may result from different tree for each position, chromosome inversions, linked selection and so on. We investigate the pattern in each species here separately, combining MDS method and genome features.
Here are the MDS results for each species.
\subsection{\textit{Drosophila}}
We checked the results for chromosome arms Chr2L, Chr2R, Chr3L, Chr3R and ChrX separately. When setting the dimension parameter M=2 in MDS method, that is, plotting the first 2 coordinates reduced from the distance matrix, each plot looks like triangle.
(Fig1.a) Since the relative position for each window in the plot shows the relative similarity between windows, it tells us there are 3 extreme types of population structure shown in the 3 peaks of the \lq\lq triangle\rq\rq, and other windows are between them, which means other window's population structure might be a mixture of those extremes. We then want to investigate more information at the extremes.

Fig1.

\noindent We pick a window for each extreme, and take out 5\% windows that have the smallest distance to it in the original pairwise distance matrix, then combine those windows for each extreme and apply PCA on them. (Fig.2) We can see the obvious difference between their PCA plots. This difference stands for variation of population structure along the genome in \textit{Drosophila} Chr2L.

Fig2.

\noindent There's a known inversion in Chr2L in \textit{Drosophila}, In(2L)t, with breakpoints at 2225744bp and 13154180bp. \cite{key11} We recolored the PCA plots in Fig.2 a,b,c by the direction of the inversion for each sample using data from John Pool's lab. (Fig.3) The results show that the clustering in PCA plots are mainly due to inversions for the red and green extremes in Fig.2a,b (especially for green extreme). What's more, we found that the two breakpoints of In(2L)t are located in the two clusters of the green points in Fig.1c. Similar results are found in other chromosome arms that have known inversions (Chr2R,
Chr3L, Chr3R) in Drosophila. (See supplementary for more details) These facts show that the variation of population structure along the Drosophila genome is mainly due to inversions.

Fig3.

\subsection{Human}
We ran our method separately on all 22 autosomes. For chromosomes 3, 8, 15 and 17 have known inversions and the corresponding breakpoints \cite{key12}. We found that the outlying windows in one-dimensional MDS plots coincide with the position of inversions for those 4 chromosomes.(Fig.4) As in Drosophila, the biggest source of variability in population structure along human chromosomes is inversion. Other chromosomes haven't had experimentally validated inversions, but there are many predicted inversions, and PCA might provide a way to validate those predicted inversion. \cite{key13} (See supplementary for other chromosomes' MDS results for human)

Fig4.

\subsection{\textit{Medicago truncatula}}
We checked the MDS results for all 8 chromosomes. We found the position of the peak for each MDS plot has a coincidence with the position of heterochromatic regions. This means the population structure in the windows located in heterochromatin tends to have higher similarity, since those windows are closer in MDS plots. (Fig.5) Biologically,
heterochromatic regions have lower gene density and may be less subject to selection \cite{key15,key16}. Then we checked the MDS results against gene density along the genome for each chromosome using gene models in Mt4.0 JBrowse. Numerically, the first MDS coordinate value is negatively correlated to the gene count for each window, which shows the correlation between MDS result and gene density. (Fig.6) This fact shows that in \textit{Medicago}, the variation of population structure is correlated with heterochromatin and gene density, and is perhaps due to linked selection.

Fig5.

Fig6.

\section{Future work}
1. For human and Drosophila, we want to eliminate the regions under known inversions and check the variation of population structure for the remaining part by removing those sections. We try to check whether they will give similar results as in \textit{Medicago truncatula}, that is whether the variation is closely related to heterochromatin or gene density.

\noindent 2. Uneven sampling has a strong influence on PCA projections \cite{key3}. Our human data, POPRES, is unevenly sampled including 346 African-Americas, 73 Asians, 359 Indian Asians and 3187 Europeans. First, we'll try sub-sampling Europeans to balance the population size for the 4 population and repeat the process on the resampled data. Second, we'll try to apply the whole process on only European samples to see the genetic variation inside European samples. Third, we want to try different scheme of adding a weighting matrix to the covariance matrix of genotype data, thus to the reduce the influence of uneven sampling.

\noindent 3. Since regions that have low recombination rate tend to have similar PCs, we'll try cutting the chromosomes into windows with same distance in genetic map instead of same SNP numbers.

\noindent 4. Euclidean distance between the contracted matrix based on PCs is one measure of the similarity for window's population structure. We want to try other methods of distance between windows, for example, we used the distance for PCs to reduce noise, however the distance between covariance matrixes of genotype matrix might also be informative.

\noindent 5. Although the first two coordinates contains the main part of information, we'd like to see the information contained in higher PCs (e.g. the third PC, the forth PC), and higher dimension of MDS.

\begin{thebibliography}{99}
\bibitem{key1} Menozzi, P., Piazza, A. \& Cavalli-Sforza, L. \lq\lq Synthetic maps of human gene frequencies in Europeans.'' Science 201, 786--792 (1978).
\bibitem{key2} Novembre, John, and Matthew Stephens. \lq\lq Interpreting principal component analyses of spatial population genetic variation." Nature genetics 40.5 (2008): 646-649.
\bibitem{key3} McVean, Gil. \lq\lq A genealogical interpretation of principal components analysis." PLoS Genet 5.10 (2009): e1000686.
\bibitem{key4} Novembre, John, et al. \lq\lq Genes mirror geography within Europe." Nature 456.7218 (2008): 98-101.
\bibitem{key5} Borg, Ingwer, and Patrick JF Groenen. Modern multidimensional scaling: Theory and applications. Springer Science \& Business Media, 2005.
\bibitem{key6} Patterson, N., A. L. Price, and D. Reich. \lq\lq Population structure and eigenanalysis." PLoS Genet 2.12 (2006): e190.
\bibitem{key7} Price, Alkes L., et al. \lq\lq Principal components analysis corrects for stratification in genome- wide association studies." Nature genetics 38.8 (2006): 904-909.
\bibitem{key8} Nelson, Matthew R., et al. \lq\lq The Population Reference Sample, POPRES: a resource for population, disease, and pharmacological genetics research." The American Journal of Human Genetics 83.3 (2008): 347-358.
\bibitem{key9} Efron, Bradley, and B. Efron. The jackknife, the bootstrap and other resampling plans. Vol. 38. Philadelphia: Society for industrial and applied mathematics, 1982.
\bibitem{key10} Engelhardt, Barbara E., and Matthew Stephens. \lq\lq Analysis of population structure: a unifying framework and novel methods based on sparse factor analysis." PLoS Genet 6.9 (2010): e1001117.
\bibitem{key11} Corbett-Detig, Russell B., and Daniel L. Hartl. \lq\lq Population genomics of inversion polymorphisms in Drosophila melanogaster." (2012): e1003056.
\bibitem{key12} Antonacci, Francesca, et al. \lq\lq Characterization of six human disease-associated inversion polymorphisms." Human molecular genetics 18.14 (2009): 2555-2566.
\bibitem{key13} Ma, Jianzhong, and Christopher I. Amos. \lq\lq Investigation of inversion polymorphisms in the human genome using principal components analysis." (2012): e40224.
\bibitem{key14} Kulikova, Olga, et al. \lq\lq Satellite repeats in the functional centromere and pericentromeric heterochromatin of Medicago truncatula." Chromosoma 113.6 (2004): 276-283.
\bibitem{key15} Kulikova, Olga, et al. \lq\lq Integration of the FISH pachytene and genetic maps of Medicago truncatula." The Plant Journal 27.1 (2001): 49-58.
\bibitem{key16} Paape, Timothy, et al. \lq\lq Selection, genome-wide fitness effects and evolutionary rates in the model legume Medicago truncatula." Molecular ecology 22.13 (2013): 3525-3538.

\end{thebibliography}

\end{document}

### 2016-02-10 local_pca_proposal.tex
switched bibliography to natbib

@@
 \usepackage{setspace}
+\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
+\usepackage[hidelinks]{hyperref}
+
 \title{Using Local PCA to Investigate Population Structure Along the Genome}
@@
 %\date{}							% Activate to display a given date or no date
+
+\usepackage{color}
+\newcommand{\plr}[1]{{\em \color{blue} #1}}
 
@@
 
-About 37 years ago, Cavalli-Sforza et al.
-first applied Principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \cite{key1}.
+About 37 years ago, \citet{menozzi1978synthetic} first applied Principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}.
 Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.
 PCA results are derived from the covariance matrix of genotype matrix, which describes population structure, demographic history, and pedigree tree and so on.
-The results of PCA can be directly related to the underlying genealogical history of the samples, such as coalescence time (time to most recent common ancestor) and migration rate between populations \cite{key1,key2,key3}.
+The results of PCA can be directly related to the underlying genealogical history of the samples, such as coalescence time (time to most recent common ancestor) and migration rate between populations \citep{menozzi1978synthetic,novembre2008interpreting,mcvean2009genealogical}.
 Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and plots of first two Principal components (PCs) can mimic the samples' geographic origin to some extent.
 The closeness of two samples in PC plots represents the similarity of their genetic information.
-Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \cite{key4,key6}.
+Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008europe,key6}.
 
@@
 Then apply Principal Component Analysis (PCA) on each window.
-Here's a brief summary about how is PCA carried out on genomic data \cite{key3}.
+Here's a brief summary about how is PCA carried out on genomic data \cite{mcvean2009genealogical}.
 Starting with the recoded genotype matrix Z, where Z is a L*N matrix ( L is SNP number ; N is sample size), then zero-center the matrix Z to X to make the data rows have equal variance.
@@
 
-\begin{multline}
-Distance_{ij} \\ =\sqrt{\left \{\left ( V_{1}\cdot V_{1}^{} \right )^{2}+\left ( V_{2}\cdot V_{2}^{} \right )^{2}+\left ( U_{1}\cdot U_{1}^{} \right )^{2} +\left ( U_{2}\cdot U_{2}^{} \right )^{2} -2\left [ \left ( V_{1}\cdot U_{1}^{} \right )^{2}+\left ( V_{1}\cdot U_{2}^{} \right )^{2}+\left ( V_{2}\cdot U_{1}^{} \right )^{2} +\left ( V_{2}\cdot U_{2}^{} \right )^{2}\right ]\right \}}
-\end{multline}
+\plr{changed "Distance" to $D$}
+\begin{align}
+    \begin{split}
+    D_{ij} &= 
+        \left \{
+                \left ( 
+                    V_{1}\cdot V_{1}^{} 
+                \right )^{2}
+                + \left ( V_{2}\cdot V_{2}^{} \right )^{2}
+                + \left ( U_{1}\cdot U_{1}^{} \right )^{2} 
+                +\left ( U_{2}\cdot U_{2}^{} \right )^{2} 
+        \right. \\ 
+        & \left. \qquad {}  
+            -2\left [ 
+                \left ( V_{1}\cdot U_{1}^{} \right )^{2}
+                + \left ( V_{1}\cdot U_{2}^{} \right )^{2}
+                + \left ( V_{2}\cdot U_{1}^{} \right )^{2} 
+                + \left ( V_{2}\cdot U_{2}^{} \right )^{2}
+            \right ]
+        \right \}^{1/2}
+    \end{split}
+\end{align}
 
-Using this procedure, we get the pairwise distance matrix that says how similar population structure is in each pair of genomic windows.
+Using this procedure, 
+we get the pairwise distance matrix that says how similar population structure is 
+in each pair of genomic windows.
 \subsubsection{Visualize the pairwise distance matrix}
-To do this, I use Multidimensional scaling (MDS), which is a visualization method commonly applied to distance matrixes.
+To do this, I use Multidimensional scaling (MDS), 
+which is a visualization method commonly applied to distance matrixes. 
 It can reduce the dimensionality of a distance matrix while preserve the distance information between objects as well as possible.
@@
 
-Fig1.
+\begin{figure}
+    \begin{center}
+       % \includegraphics[width=0.7\textwidth]{figs/example_cline}
+    \end{center}
+    \caption{
+        The caption goes here.
+        \label{fig:pde_clines}
+    }
+\end{figure}
 
@@
 \noindent 2.
-Uneven sampling has a strong influence on PCA projections \cite{key3}.
+Uneven sampling has a strong influence on PCA projections \cite{mcvean2009genealogical}.
 Our human data, POPRES, is unevenly sampled including 346 African-Americas, 73 Asians, 359 Indian Asians and 3187 Europeans.
@@
 
-\begin{thebibliography}{99}
-\bibitem{key1} Menozzi, P., Piazza, A. \& Cavalli-Sforza, L. \lq\lq Synthetic maps of human gene frequencies in Europeans.'' Science 201, 786--792 (1978).
-\bibitem{key2} Novembre, John, and Matthew Stephens. \lq\lq Interpreting principal component analyses of spatial population genetic variation." Nature genetics 40.5 (2008): 646-649.
-\bibitem{key3} McVean, Gil. \lq\lq A genealogical interpretation of principal components analysis." PLoS Genet 5.10 (2009): e1000686.
-\bibitem{key4} Novembre, John, et al. \lq\lq Genes mirror geography within Europe." Nature 456.7218 (2008): 98-101.
-\bibitem{key5} Borg, Ingwer, and Patrick JF Groenen. Modern multidimensional scaling: Theory and applications. Springer Science \& Business Media, 2005.
-\bibitem{key6} Patterson, N., A. L. Price, and D. Reich. \lq\lq Population structure and eigenanalysis." PLoS Genet 2.12 (2006): e190.
-\bibitem{key7} Price, Alkes L., et al. \lq\lq Principal components analysis corrects for stratification in genome- wide association studies." Nature genetics 38.8 (2006): 904-909.
-\bibitem{key8} Nelson, Matthew R., et al.
+\bibliographystyle{plainnat}
+\bibliography{references}  % if bibtex file is called references.bib
+
+% \begin{thebibliography}{99}
+% \bibitem{menozzi1978synthetic} Menozzi, P., Piazza, A. \& Cavalli-Sforza, L. \lq\lq Synthetic maps of human gene frequencies in Europeans.'' Science 201, 786--792 (1978).
+% \bibitem{novembre2008interpreting} Novembre, John, and Matthew Stephens.
+\lq\lq Interpreting principal component analyses of spatial population genetic variation." Nature genetics 40.5 (2008): 646-649.
+% \bibitem{mcvean2009genealogical} McVean, Gil. \lq\lq A genealogical interpretation of principal components analysis." PLoS Genet 5.10 (2009): e1000686.
+% \bibitem{novembre2008europe} Novembre, John, et al. \lq\lq Genes mirror geography within Europe." Nature 456.7218 (2008): 98-101.
+% \bibitem{key5} Borg, Ingwer, and Patrick JF Groenen. Modern multidimensional scaling: Theory and applications. Springer Science \& Business Media, 2005.
+% \bibitem{key6} Patterson, N., A. L. Price, and D. Reich. \lq\lq Population structure and eigenanalysis." PLoS Genet 2.12 (2006): e190.
+% \bibitem{key7} Price, Alkes L., et al. \lq\lq Principal components analysis corrects for stratification in genome- wide association studies." Nature genetics 38.8 (2006): 904-909.
+% \bibitem{key8} Nelson, Matthew R., et al.
 \lq\lq The Population Reference Sample, POPRES: a resource for population, disease, and pharmacological genetics research." The American Journal of Human Genetics 83.3 (2008): 347-358.
-\bibitem{key9} Efron, Bradley, and B. Efron. The jackknife, the bootstrap and other resampling plans. Vol. 38. Philadelphia: Society for industrial and applied mathematics, 1982.
-\bibitem{key10} Engelhardt, Barbara E., and Matthew Stephens.
+% \bibitem{key9} Efron, Bradley, and B. Efron. The jackknife, the bootstrap and other resampling plans. Vol. 38. Philadelphia: Society for industrial and applied mathematics, 1982.
+% \bibitem{key10} Engelhardt, Barbara E., and Matthew Stephens.
 \lq\lq Analysis of population structure: a unifying framework and novel methods based on sparse factor analysis." PLoS Genet 6.9 (2010): e1001117.
-\bibitem{key11} Corbett-Detig, Russell B., and Daniel L. Hartl. \lq\lq Population genomics of inversion polymorphisms in Drosophila melanogaster." (2012): e1003056.
-\bibitem{key12} Antonacci, Francesca, et al. \lq\lq Characterization of six human disease-associated inversion polymorphisms." Human molecular genetics 18.14 (2009): 2555-2566.
-\bibitem{key13} Ma, Jianzhong, and Christopher I. Amos. \lq\lq Investigation of inversion polymorphisms in the human genome using principal components analysis." (2012): e40224.
-\bibitem{key14} Kulikova, Olga, et al. \lq\lq Satellite repeats in the functional centromere and pericentromeric heterochromatin of Medicago truncatula." Chromosoma 113.6 (2004): 276-283.
-\bibitem{key15} Kulikova, Olga, et al. \lq\lq Integration of the FISH pachytene and genetic maps of Medicago truncatula." The Plant Journal 27.1 (2001): 49-58.
-\bibitem{key16} Paape, Timothy, et al. \lq\lq Selection, genome-wide fitness effects and evolutionary rates in the model legume Medicago truncatula." Molecular ecology 22.13 (2013): 3525-3538.
-
-\end{thebibliography}
+% \bibitem{key11} Corbett-Detig, Russell B., and Daniel L. Hartl. \lq\lq Population genomics of inversion polymorphisms in Drosophila melanogaster." (2012): e1003056.
+% \bibitem{key12} Antonacci, Francesca, et al. \lq\lq Characterization of six human disease-associated inversion polymorphisms." Human molecular genetics 18.14 (2009): 2555-2566.
+% \bibitem{key13} Ma, Jianzhong, and Christopher I. Amos. \lq\lq Investigation of inversion polymorphisms in the human genome using principal components analysis." (2012): e40224.
+% \bibitem{key14} Kulikova, Olga, et al. \lq\lq Satellite repeats in the functional centromere and pericentromeric heterochromatin of Medicago truncatula." Chromosoma 113.6 (2004): 276-283.
+% \bibitem{key15} Kulikova, Olga, et al. \lq\lq Integration of the FISH pachytene and genetic maps of Medicago truncatula." The Plant Journal 27.1 (2001): 49-58.
+% \bibitem{key16} Paape, Timothy, et al. \lq\lq Selection, genome-wide fitness effects and evolutionary rates in the model legume Medicago truncatula." Molecular ecology 22.13 (2013): 3525-3538.
+% \end{thebibliography}
 

### 2016-02-22 writeup/local_pca_paper.tex
moved writeup to writeup/

@@
-\documentclass{article}
-\usepackage{amsmath,amssymb}
-\usepackage{fullpage}
-\usepackage[authoryear]{natbib}
-\usepackage{graphicx}
+\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
+\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
+\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
+%\geometry{landscape}                		% Activate for for rotated page geometry
+%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
+\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
+								% TeX will automatically convert eps --> pdf in pdflatex		
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{setspace}
+\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
 \usepackage[hidelinks]{hyperref}
-%\usepackage{latexml} % for \iflatexml
+
+\title{Using Local PCA to Investigate Population Structure Along the Genome}
+\author{Han Li, Peter Ralph}
+%\date{}							% Activate to display a given date or no date
+
 \usepackage{color}
-\usepackage[dvipsnames]{xcolor}
-
-\bibliographystyle{plainnat}
-
-\title{Population structure along the genome}
-\author{
-Han Li
-\\and\\
-Peter Ralph
-}
+\newcommand{\plr}[1]{{\em \color{blue} #1}}
 
 \begin{document}
+\maketitle
+%\section{}
+%\subsection{}
+\doublespacing
+\section{Introduction}
 
-Other phrases that might go into a better title?
-\begin{itemize}
-    \item How selection affects relatedness
-    \item Inhomogeneous kinship
-    \item There's more than one kinship matrix
-    \item Local PCA reveals inversions and the effects of selection
-    \item Unsupervised learning of genomic patterns of relatedness
-\end{itemize}
+About 37 years ago, \citet{menozzi1978synthetic} first applied Principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}.
+Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.
+PCA results are derived from the covariance matrix of genotype matrix, which describes population structure, demographic history, and pedigree tree and so on.
+The results of PCA can be directly related to the underlying genealogical history of the samples, such as coalescence time (time to most recent common ancestor) and migration rate between populations \citep{menozzi1978synthetic,novembre2008interpreting,mcvean2009genealogical}.
+Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and plots of first two Principal components (PCs) can mimic the samples' geographic origin to some extent.
+The closeness of two samples in PC plots represents the similarity of their genetic information.
+Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008europe,key6}.
 
-%%%%%%%%%%%%
-\section{Introduction}
+\noindent  Most time, PCA used in genome-wide association study (GWAS) is for stratification correction \cite{key7}.
+However, for different part of genome, they have different genetic features.
+First, each site of DNA may have different gene tree.
+For a DNA region, if individuals have SNPs closer in gene tree, they tends to be more close in PC projections for that region.
+Different DNA segments may have different gene tree for their SNPs and would result in different population structure for those segments.
+Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region.
+Third, if chromosome inversion exists, the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints, thus results in different genome structure and population structure.
+Investigating the genomic variation along the genome can help us to have a better understanding of the relation between genome structure and population structure.
+Therefore, we cut the chromosomes into windows with hundreds/thousands of SNPs and apply PCA locally on each window.
+In this project, we used SNP data for human, \textit{Medicago truncatula}, and whole genome sequencing data for \textit{Drosophila}.
+
+\noindent Based on the principal components, we can estimate the similarity of population structure contained in each genome window.
+To visualize the closeness for each window with regard to the population structure, we constructed matrix based on the first two PCs for each window and got the pairwise Euclidean distance between those matrixes, then use multidimensional scaling to reduces the pairwise distance matrix to lower dimension while preserve the distance information between windows as well as possible \cite{key5}.
+To interpret the results of MDS, we combine known genome feature information for each species, such as the distribution of inversions, heterochromatin chunk and gene density along the genome.
+For different species, we got different explanation for the variation of population structure along the genome.
+
+\section{Other Introduction}
 
@@
 
-\subsection{Applications}
+\section{Method}
+\subsection{Recode the DNA sequence to a matrix consisting of 0,1,2 (and NA).}
+For human SNP data from POPRES \cite{key8}, we use the allele that has highest frequency in the samples as the reference allele for each position.
+If an allele is same with the reference allele, we recode it as 0; if an allele is different from the reference allele, we recode it as 1; for positions that have missing data, we recode them as NA.
+Since human genome is diploid, we add the value for the two alleles and then one chromosome will eventually be recode to a sequence consisting of 0,1,2 (and NA).
+There are 3965 samples in total, (346 African-Americas; 73 Asians; 3187 Europeans; 359 Indian Asians); then the genome data for human is recoded to a matrix that has 3965 columns, each column for an individual's genotype.
+We process the data separately for the 22 autosomes in human.
+For \textit{Drosophila}, we first process the sequencing data (from DPGP and John pool's lab) to SNP data by eliminating the positions that have all the same alleles.
+Due to high density of missing data for some parts in the genome, we then delete the samples with more than 8\% NAs and positions with more than 20\%.
+The cutoff points 8\% and 20\% are determined from the corresponding distributions of NAs in samples and at positions.
+After we got the SNP data for \textit{Drosophila}, we recode it \textit{Drosophila}to matrices with 0,1,2 (and NA) for each chromosome arms (Chr2L, Chr2R, Chr3L, Chr3R, ChrX) similar to the process for human genome.
+Since the \textit{Drosophila} samples here are all homogenous, the matrices are indeed consisting of 0,1 (and NA).
+For \textit{Medicago truncatula}, we use the SNP data from Medicago Hapmap and recode it to matrices with 0,1,2 (and NA) for each of the 8 chromosomes.
 
-We apply this to three species.
+\subsection{Cut each genome into windows}
+We cut each recoded matrix into sub matrixes by that have the same columns but fewer rows than the original matrix.
+Then apply Principal Component Analysis (PCA) on each window.
+Here's a brief summary about how is PCA carried out on genomic data \cite{mcvean2009genealogical}.
+Starting with the recoded genotype matrix Z, where Z is a L*N matrix ( L is SNP number ; N is sample size), then zero-center the matrix Z to X to make the data rows have equal variance.
+Get the covariance matrix of X (denote as matrix C) and compute the eigenvectors and eigenvalues of the covariance matrix C.
+The \textit{i}th principal component is the \textit{i}th eigenvector of C.
 
-humans:
-effects of selection are seen \citep{sabeti2006positive}.
-\citet{duforetfrebourg2015detecting} showed that variants known to underly locally adaptive traits correlated most strongly with PCs.
-\citet{tian2008analysis} noted that removing the segregating inversion on chromosome 8 changed the PCA for a set of 1400 Europeans
-dataset(s)
+\begin{equation}
+X_{si}=Z_{si}-\frac{1}{n}\sum_{j=1}^{n}Z_{sj} 
+\end{equation}
+\begin{equation}
+C=\frac{1}{n-1}XX^{T}
+\end{equation}
 
-drosophila:
-effects of selection \citep{sella2009pervasive},
-inversions,
-dataset(s)
+\subsubsection{Genomic PCA on windows}
+PCA on each window is getting the eigenvectors of covariance matrix of the cut recoded matrix.
+PCA plots (generally using PC2 against PC1) can show population structure, which describes how different samples are related.
+We want to investigate the variation of population structure along the genome.
+So we apply PCA on each window to check the population structure along the genome.
+\subsubsection{Choose window length}
+The window length should neither be too long or too short.
+The longer the windows, the more accurate is the estimate of population structure in that window.
+However, for better resolution, we need to find a length that is reasonably short.
+If we use the first principal component as a measure of population structure, then to choose a proper length for a
+window, we need to find a balance between variance of the first principal components inside a window and that between windows.
+The variance between windows is estimated as mean variance of the first principal component for each window.
+The variance inside a window is estimated using the jackknife block when cutting the window into 10 equal size smaller windows \cite{key9}.
+Table 1 shows the comparison of variance within a window and that between windows for chromosome arms in \textit{Drosophila}.
+Finally, we choose 100 SNPs, 1000 SNPs and 10000 SNPs as window length for human, \textit{Drosophila}, and \textit{Medicago} separately.
 
-medicago:
+Table 1.
 
-%%%%%%%%%%%%
-\section{Methods}
+\subsubsection{Similarity of population structure between windows}
+We use the first two principal components (PCs) and the corresponding eigenvalues from PCA, and construct a new matrix with the two PCs for each window to stand for the population structure information.
+For example, the constructed matrixes for ith and jth window are as following.
+($\lambda _{1i}$ and $\lambda _{2i}$ are the eigenvalues for the first two PCs for \textit{i}th window; $\lambda _{1j}$ and $\lambda _{2j}$ are the eigenvalues for the first two PCs for \textit{j}th window.)
 
-%%%%%%%%%%%%
+\begin{equation}
+M_{i}=\frac{\lambda _{1i}PC1_{i}PC1_{i}^{T}+\lambda _{2i}PC2_{i}PC2_{i}^{T}}{\lambda _{1i}+\lambda _{2i}} \qquad
+M_{j}=\frac{\lambda _{1j}PC1_{j}PC1_{j}^{T}+\lambda _{2j}PC2_{j}PC2_{j}^{T}}{\lambda _{1j}+\lambda _{2j}}
+\end{equation}
+
+The Euclidean distance between the constructed matrices for windows stands for the
+similarity of population structure for windows. Due to the property of eigenvectors,we
+could use the following method to calculate the pairwise distance greatly saving time and space.
+
+\begin{equation}
+V_{1}=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} \qquad
+V_{2}=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i}
+\end{equation}
+
+\plr{changed "Distance" to $D$}
+\begin{align}
+    \begin{split}
+    D_{ij} &= 
+        \left \{
+                \left ( 
+                    V_{1}\cdot V_{1}^{} 
+                \right )^{2}
+                + \left ( V_{2}\cdot V_{2}^{} \right )^{2}
+                + \left ( U_{1}\cdot U_{1}^{} \right )^{2} 
+                +\left ( U_{2}\cdot U_{2}^{} \right )^{2} 
+        \right. \\ 
+        & \left. \qquad {}  
+            -2\left [ 
+                \left ( V_{1}\cdot U_{1}^{} \right )^{2}
+                + \left ( V_{1}\cdot U_{2}^{} \right )^{2}
+                + \left ( V_{2}\cdot U_{1}^{} \right )^{2} 
+                + \left ( V_{2}\cdot U_{2}^{} \right )^{2}
+            \right ]
+        \right \}^{1/2}
+    \end{split}
+\end{align}
+
+Using this procedure, 
+we get the pairwise distance matrix that says how similar population structure is 
+in each pair of genomic windows.
+\subsubsection{Visualize the pairwise distance matrix}
+To do this, I use Multidimensional scaling (MDS), 
+which is a visualization method commonly applied to distance matrixes. 
+It can reduce the dimensionality of a distance matrix while preserve the distance information between objects as well as possible.
+The aimed dimension \lq\lq M\rq\rq can be set based on your need.
+We used M=1 and M=2 in our study.
+This allowed us to visualize the information in the distance matrix in one or two dimensional relation between windows' population structure.
+
 \section{Results}
+\textbf{In all these 3 species, PCA plots vary along the genome.}
 
-%%%%%%%%%%%%
-\section{Discussion}
+\noindent Since PCA plots can show population structure, this shows that the population structure varies along the genome.
+Each PC plot comes from the covariance matrix of a different section of the genome, which may result from different tree for each position, chromosome inversions, linked selection and so on.
+We investigate the pattern in each species here separately, combining MDS method and genome features.
+Here are the MDS results for each species.
+\subsection{\textit{Drosophila}}
+We checked the results for chromosome arms Chr2L, Chr2R, Chr3L, Chr3R and ChrX separately.
+When setting the dimension parameter M=2 in MDS method, that is, plotting the first 2 coordinates reduced from the distance matrix, each plot looks like triangle.
+(Fig1.a) Since the relative position for each window in the plot shows the relative similarity between windows, it tells us there are 3 extreme types of population structure shown in the 3 peaks of the \lq\lq triangle\rq\rq, and other windows are between them, which means other window's population structure might be a mixture of those extremes.
+We then want to investigate more information at the extremes.
 
-%%%%%%%%%%%%%
-\bibliography{references}
+\begin{figure}
+    \begin{center}
+       % \includegraphics[width=0.7\textwidth]{figs/example_cline}
+    \end{center}
+    \caption{
+        The caption goes here.
+        \label{fig:pde_clines}
+    }
+\end{figure}
+
+\noindent We pick a window for each extreme, and take out 5\% windows that have the smallest distance to it in the original pairwise distance matrix, then combine those windows for each extreme and apply PCA on them.
+(Fig.2) We can see the obvious difference between their PCA plots.
+This difference stands for variation of population structure along the genome in \textit{Drosophila} Chr2L.
+
+Fig2.
+
+\noindent There's a known inversion in Chr2L in \textit{Drosophila}, In(2L)t, with breakpoints at 2225744bp and 13154180bp.
+\cite{key11} We recolored the PCA plots in Fig.2 a,b,c by the direction of the inversion for each sample using data from John Pool's lab.
+(Fig.3) The results show that the clustering in PCA plots are mainly due to inversions for the red and green extremes in Fig.2a,b (especially for green extreme).
+What's more, we found that the two breakpoints of In(2L)t are located in the two clusters of the green points in Fig.1c.
+Similar results are found in other chromosome arms that have known inversions (Chr2R,
+Chr3L, Chr3R) in Drosophila. (See supplementary for more details) These facts show that the variation of population structure along the Drosophila genome is mainly due to inversions.
+
+Fig3.
+
+\subsection{Human}
+We ran our method separately on all 22 autosomes.
+For chromosomes 3, 8, 15 and 17 have known inversions and the corresponding breakpoints \cite{key12}.
+We found that the outlying windows in one-dimensional MDS plots coincide with the position of inversions for those 4 chromosomes.(Fig.4) As in Drosophila, the biggest source of variability in population structure along human chromosomes is inversion.
+Other chromosomes haven't had experimentally validated inversions, but there are many predicted inversions, and PCA might provide a way to validate those predicted inversion.
+\cite{key13} (See supplementary for other chromosomes' MDS results for human)
+
+Fig4.
+
+\subsection{\textit{Medicago truncatula}}
+We checked the MDS results for all 8 chromosomes.
+We found the position of the peak for each MDS plot has a coincidence with the position of heterochromatic regions.
+This means the population structure in the windows located in heterochromatin tends to have higher similarity, since those windows are closer in MDS plots.
+(Fig.5) Biologically,
+heterochromatic regions have lower gene density and may be less subject to selection \cite{key15,key16}.
+Then we checked the MDS results against gene density along the genome for each chromosome using gene models in Mt4.0 JBrowse.
+Numerically, the first MDS coordinate value is negatively correlated to the gene count for each window, which shows the correlation between MDS result and gene density.
+(Fig.6) This fact shows that in \textit{Medicago}, the variation of population structure is correlated with heterochromatin and gene density, and is perhaps due to linked selection.
+
+Fig5.
+
+Fig6.
+
+\section{Future work}
+1.
+For human and Drosophila, we want to eliminate the regions under known inversions and check the variation of population structure for the remaining part by removing those sections.
+We try to check whether they will give similar results as in \textit{Medicago truncatula}, that is whether the variation is closely related to heterochromatin or gene density.
+
+\noindent 2.
+Uneven sampling has a strong influence on PCA projections \cite{mcvean2009genealogical}.
+Our human data, POPRES, is unevenly sampled including 346 African-Americas, 73 Asians, 359 Indian Asians and 3187 Europeans.
+First, we'll try sub-sampling Europeans to balance the population size for the 4 population and repeat the process on the resampled data.
+Second, we'll try to apply the whole process on only European samples to see the genetic variation inside European samples.
+Third, we want to try different scheme of adding a weighting matrix to the covariance matrix of genotype data, thus to the reduce the influence of uneven sampling.
+
+\noindent 3. Since regions that have low recombination rate tend to have similar PCs, we'll try cutting the chromosomes into windows with same distance in genetic map instead of same SNP numbers.
+
+\noindent 4.
+Euclidean distance between the contracted matrix based on PCs is one measure of the similarity for window's population structure.
+We want to try other methods of distance between windows, for example, we used the distance for PCs to reduce noise, however the distance between covariance matrixes of genotype matrix might also be informative.
+
+\noindent 5.
+Although the first two coordinates contains the main part of information, we'd like to see the information contained in higher PCs (e.g.
+the third PC, the forth PC), and higher dimension of MDS.
+
+\bibliographystyle{plainnat}
+\bibliography{references}  % if bibtex file is called references.bib
+
+% \begin{thebibliography}{99}
+% \bibitem{menozzi1978synthetic} Menozzi, P., Piazza, A. \& Cavalli-Sforza, L. \lq\lq Synthetic maps of human gene frequencies in Europeans.'' Science 201, 786--792 (1978).
+% \bibitem{novembre2008interpreting} Novembre, John, and Matthew Stephens.
+\lq\lq Interpreting principal component analyses of spatial population genetic variation." Nature genetics 40.5 (2008): 646-649.
+% \bibitem{mcvean2009genealogical} McVean, Gil. \lq\lq A genealogical interpretation of principal components analysis." PLoS Genet 5.10 (2009): e1000686.
+% \bibitem{novembre2008europe} Novembre, John, et al. \lq\lq Genes mirror geography within Europe." Nature 456.7218 (2008): 98-101.
+% \bibitem{key5} Borg, Ingwer, and Patrick JF Groenen. Modern multidimensional scaling: Theory and applications. Springer Science \& Business Media, 2005.
+% \bibitem{key6} Patterson, N., A. L. Price, and D. Reich. \lq\lq Population structure and eigenanalysis." PLoS Genet 2.12 (2006): e190.
+% \bibitem{key7} Price, Alkes L., et al. \lq\lq Principal components analysis corrects for stratification in genome- wide association studies." Nature genetics 38.8 (2006): 904-909.
+% \bibitem{key8} Nelson, Matthew R., et al.
+\lq\lq The Population Reference Sample, POPRES: a resource for population, disease, and pharmacological genetics research." The American Journal of Human Genetics 83.3 (2008): 347-358.
+% \bibitem{key9} Efron, Bradley, and B. Efron. The jackknife, the bootstrap and other resampling plans. Vol. 38. Philadelphia: Society for industrial and applied mathematics, 1982.
+% \bibitem{key10} Engelhardt, Barbara E., and Matthew Stephens.
+\lq\lq Analysis of population structure: a unifying framework and novel methods based on sparse factor analysis." PLoS Genet 6.9 (2010): e1001117.
+% \bibitem{key11} Corbett-Detig, Russell B., and Daniel L. Hartl. \lq\lq Population genomics of inversion polymorphisms in Drosophila melanogaster." (2012): e1003056.
+% \bibitem{key12} Antonacci, Francesca, et al. \lq\lq Characterization of six human disease-associated inversion polymorphisms." Human molecular genetics 18.14 (2009): 2555-2566.
+% \bibitem{key13} Ma, Jianzhong, and Christopher I. Amos. \lq\lq Investigation of inversion polymorphisms in the human genome using principal components analysis." (2012): e40224.
+% \bibitem{key14} Kulikova, Olga, et al. \lq\lq Satellite repeats in the functional centromere and pericentromeric heterochromatin of Medicago truncatula." Chromosoma 113.6 (2004): 276-283.
+% \bibitem{key15} Kulikova, Olga, et al. \lq\lq Integration of the FISH pachytene and genetic maps of Medicago truncatula." The Plant Journal 27.1 (2001): 49-58.
+% \bibitem{key16} Paape, Timothy, et al. \lq\lq Selection, genome-wide fitness effects and evolutionary rates in the model legume Medicago truncatula." Molecular ecology 22.13 (2013): 3525-3538.
+% \end{thebibliography}
 

### 2016-02-22 writeup/local_pca_paper.tex
some notes

@@
 \doublespacing
+
+\plr{Note: your intro focuses on PCA more than mine, which is trying to be about population structure generally.}
+
 \section{Introduction}
@@
 The closeness of two samples in PC plots represents the similarity of their genetic information.
-Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008europe,key6}.
+Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008genes,key6}.
 
-\noindent  Most time, PCA used in genome-wide association study (GWAS) is for stratification correction \cite{key7}.
+Most time, PCA used in genome-wide association study (GWAS) is for stratification correction \cite{key7}.
 However, for different part of genome, they have different genetic features.
@@
 Investigating the genomic variation along the genome can help us to have a better understanding of the relation between genome structure and population structure.
+
+\plr{(this is starting to be methods and results)}
 Therefore, we cut the chromosomes into windows with hundreds/thousands of SNPs and apply PCA locally on each window.
@@
 
-\noindent Based on the principal components, we can estimate the similarity of population structure contained in each genome window.
+Based on the principal components, we can estimate the similarity of population structure contained in each genome window.
 To visualize the closeness for each window with regard to the population structure, we constructed matrix based on the first two PCs for each window and got the pairwise Euclidean distance between those matrixes, then use multidimensional scaling to reduces the pairwise distance matrix to lower dimension while preserve the distance information between windows as well as possible \cite{key5}.
@@
 
-\section{Method}
+\section{Methods}
+
+\plr{How about: first describe the method; then afterwards, present the three datasets.
+
 \subsection{Recode the DNA sequence to a matrix consisting of 0,1,2 (and NA).}
+
+\plr{Separate the different datasets into different paragraphs.}
+
 For human SNP data from POPRES \cite{key8}, we use the allele that has highest frequency in the samples as the reference allele for each position.
@@
 
-\noindent Since PCA plots can show population structure, this shows that the population structure varies along the genome.
+Since PCA plots can show population structure, this shows that the population structure varies along the genome.
 Each PC plot comes from the covariance matrix of a different section of the genome, which may result from different tree for each position, chromosome inversions, linked selection and so on.
@@
 
-\noindent We pick a window for each extreme, and take out 5\% windows that have the smallest distance to it in the original pairwise distance matrix, then combine those windows for each extreme and apply PCA on them.
+We pick a window for each extreme, and take out 5\% windows that have the smallest distance to it in the original pairwise distance matrix, then combine those windows for each extreme and apply PCA on them.
 (Fig.2) We can see the obvious difference between their PCA plots.
@@
 
-\noindent There's a known inversion in Chr2L in \textit{Drosophila}, In(2L)t, with breakpoints at 2225744bp and 13154180bp.
+There's a known inversion in Chr2L in \textit{Drosophila}, In(2L)t, with breakpoints at 2225744bp and 13154180bp.
 \cite{key11} We recolored the PCA plots in Fig.2 a,b,c by the direction of the inversion for each sample using data from John Pool's lab.

### 2016-02-22 writeup/local_pca_paper_updated.tex
Create local_pca_paper_updated.tex

--- 2016-02-10 local_pca_proposal.tex
@@
 \usepackage{setspace}
-\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
+\usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
 \usepackage[hidelinks]{hyperref}
@@
 The closeness of two samples in PC plots represents the similarity of their genetic information.
-Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008europe,key6}.
+Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008genes,patterson2006population}.
 
-\noindent  Most time, PCA used in genome-wide association study (GWAS) is for stratification correction \cite{key7}.
+\noindent  Most time, PCA used in genome-wide association study (GWAS) is for stratification correction \citep{price2006principal}.
 However, for different part of genome, they have different genetic features.
@@
 \noindent Based on the principal components, we can estimate the similarity of population structure contained in each genome window.
-To visualize the closeness for each window with regard to the population structure, we constructed matrix based on the first two PCs for each window and got the pairwise Euclidean distance between those matrixes, then use multidimensional scaling to reduces the pairwise distance matrix to lower dimension while preserve the distance information between windows as well as possible \cite{key5}.
+To visualize the closeness for each window with regard to the population structure, we constructed matrix based on the first two PCs for each window and got the pairwise Euclidean distance between those matrixes, then use multidimensional scaling to reduces the pairwise distance matrix to lower dimension while preserve the distance information between windows as well as possible \citep{borg2005modern}.
 To interpret the results of MDS, we combine known genome feature information for each species, such as the distribution of inversions, heterochromatin chunk and gene density along the genome.
@@
 \subsection{Recode the DNA sequence to a matrix consisting of 0,1,2 (and NA).}
-For human SNP data from POPRES \cite{key8}, we use the allele that has highest frequency in the samples as the reference allele for each position.
+For human SNP data from POPRES \citet{nelson2008population}, we use the allele that has highest frequency in the samples as the reference allele for each position.
 If an allele is same with the reference allele, we recode it as 0; if an allele is different from the reference allele, we recode it as 1; for positions that have missing data, we recode them as NA.
@@
 Then apply Principal Component Analysis (PCA) on each window.
-Here's a brief summary about how is PCA carried out on genomic data \cite{mcvean2009genealogical}.
+Here's a brief summary about how is PCA carried out on genomic data \citep{mcvean2009genealogical}.
 Starting with the recoded genotype matrix Z, where Z is a L*N matrix ( L is SNP number ; N is sample size), then zero-center the matrix Z to X to make the data rows have equal variance.
@@
 The variance between windows is estimated as mean variance of the first principal component for each window.
-The variance inside a window is estimated using the jackknife block when cutting the window into 10 equal size smaller windows \cite{key9}.
+The variance inside a window is estimated using the jackknife block when cutting the window into 10 equal size smaller windows \cite{efron1982jackknife}.
 Table 1 shows the comparison of variance within a window and that between windows for chromosome arms in \textit{Drosophila}.
@@
 
-Table 1.
+\noindent Table 1. Comparison of variance within a window and that between windows for chromosome arms in \textit{Drosophila}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rllrrrrr}
+  \hline
+ & chr\_name & win\_length(in SNPs) & 100 & 500 & 10\verb|^|3 & 10\verb|^|4 & 10\verb|^|5 \\ 
+  \hline
+1 & Chr2L & SE\verb|^|2(within) & 2.05e-03 & 1.64e-03 & 1.18e-03 & 1.68e-04 & 4.02e-05 \\ 
+  2 & Chr2L & Var(between) & 2.76e-03 & 2.69e-03 & 2.23e-03 & 6.74e-04 & 3.12e-04 \\ 
+  3 & Chr2R & SE\verb|^|2(within) & 2.18e-03 & 1.92e-03 & 1.63e-03 & 5.76e-04 & 1.35e-04 \\ 
+  4 & Chr2R & Var(between) & 2.78e-03 & 2.70e-03 & 2.65e-03 & 2.31e-03 & 1.82e-03 \\ 
+  5 & Chr3L & SE\verb|^|2(within) & 2.08e-03 & 2.00e-03 & 1.64e-03 & 7.32e-04 & 2.45e-04 \\ 
+  6 & Chr3L & Var(between) & 2.60e-03 & 2.52e-03 & 2.40e-03 & 1.68e-03 & 1.89e-03 \\ 
+  7 & Chr3R & SE\verb|^|2(within) & 1.95e-03 & 1.76e-03 & 1.44e-03 & 5.87e-04 & 2.03e-04 \\ 
+  8 & Chr3R & Var(between) & 2.58e-03 & 2.51e-03 & 2.44e-03 & 1.96e-03 & 1.40e-03 \\ 
+  9 & ChrX & SE\verb|^|2(within) & 2.48e-03 & 2.04e-03 & 1.54e-03 & 1.62e-03 & 1.68e-04 \\ 
+  10 & ChrX & Var(between) & 2.61e-03 & 2.43e-03 & 2.30e-03 & 3.24e-04 & 1.14e-03 \\ 
+   \hline
+\end{tabular}
+\end{table}
 
@@
 
-\plr{changed "Distance" to $D$}
 \begin{align}
@@
     \begin{center}
-       % \includegraphics[width=0.7\textwidth]{figs/example_cline}
+       \includegraphics[width=0.9\textwidth]{Fig1_Together_MDS_plot_Chr2L}
     \end{center}
     \caption{
-        The caption goes here.
+         a) Two dimensional MDS plot for chromosome arm Chr2L in \textit{Drosophila}.
+Each point in the plot stands for a window.
+The distance between windows shows the similarity of population structure for windows.
+b) The color corresponding to that in a) and thus shows a relative position for each window in a) along the genome.
+c) MDS 1D plot for Chr2L in \textit{Drosophila}, using the first coordinate in a) against the window id along the genome.
+The green, red, blue points are corresponding to the windows in the green, red, blue circled extremes in a).
         \label{fig:pde_clines}
@@
 
-Fig2.
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{Fig2_all_pca_plots_for_Chr2L_3peaks}
+    \end{center}
+    \caption{
+         PCA plots for each peak in Fig.1a and for whole Chr2L in \textit{Drosophila}. The color of the square corresponds to the circles of the 3 extremes in Fig.1a. (Also labeled same as 1,2,3)
+        \label{fig:pde_clines}
+    }
+\end{figure}
 
 \noindent There's a known inversion in Chr2L in \textit{Drosophila}, In(2L)t, with breakpoints at 2225744bp and 13154180bp.
-\cite{key11} We recolored the PCA plots in Fig.2 a,b,c by the direction of the inversion for each sample using data from John Pool's lab.
+\citep{corbett2012population} We recolored the PCA plots in Fig.2 a,b,c by the direction of the inversion for each sample using data from John Pool's lab.
 (Fig.3) The results show that the clustering in PCA plots are mainly due to inversions for the red and green extremes in Fig.2a,b (especially for green extreme).
@@
 
-Fig3.
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{Fig3_all_pca_plots_for_Chr2L_3peaks_color_by_In(2L)t}
+    \end{center}
+    \caption{
+         PCA plots colored by direction of In(2L)t direction for each peak in Chr2L MDS 2D plot. Red dots for samples with inversion, green dots for samples without inversion.
+        \label{fig:pde_clines}
+    }
+\end{figure}
 
@@
 We ran our method separately on all 22 autosomes.
-For chromosomes 3, 8, 15 and 17 have known inversions and the corresponding breakpoints \cite{key12}.
+For chromosomes 3, 8, 15 and 17 have known inversions and the corresponding breakpoints \citep{antonacci2009characterization}.
 We found that the outlying windows in one-dimensional MDS plots coincide with the position of inversions for those 4 chromosomes.(Fig.4) As in Drosophila, the biggest source of variability in population structure along human chromosomes is inversion.
 Other chromosomes haven't had experimentally validated inversions, but there are many predicted inversions, and PCA might provide a way to validate those predicted inversion.
-\cite{key13} (See supplementary for other chromosomes' MDS results for human)
+\citep{ma2012investigation} (See supplementary for other chromosomes' MDS results for human)
 
-Fig4.
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.7\textwidth]{fig4}
+    \end{center}
+    \caption{
+         MDS results for human genome (chr3, chr8, chr15, chr17, chr6, chr12).
+The x-axis is the position on the genome in window number, and y-axis is the first coordinate of MDS.
+Black dots stand for windows, red lines show positions of inversions.
+        \label{fig:pde_clines}
+    }
+\end{figure}
 
@@
 (Fig.5) Biologically,
-heterochromatic regions have lower gene density and may be less subject to selection \cite{key15,key16}.
+heterochromatic regions have lower gene density and may be less subject to selection \citep{kulikova2001integration,paape2013selection}.
 Then we checked the MDS results against gene density along the genome for each chromosome using gene models in Mt4.0 JBrowse.
@@
 
-Fig5.
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.7\textwidth]{fig5}
+    \end{center}
+    \caption{
+         MDS results for the \textit{Medicago} genome (chr1-8), using first coordinate of MDS against the middle position of each window along each chromosome. 
+         The bars under MDS plots are diagrams for each chromosome showing the relative position of heterochromatin (the deep purple part on each bar) \citep{kulikova2004satellite, kulikova2001integration}.
+        \label{fig:pde_clines}
+    }
+\end{figure}
 
-Fig6.
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{Fig6_MDS_1D_and_gene_count_and_lm_chr3_based_all_chr_win104}
+    \end{center}
+    \caption{
+        The first coordinate of MDS plotting against gene models gene count for chromosome 3 in \textit{Medicago}.
+(See supplementary for other chromosomes' MDS results against gene count for Medicago)
+        \label{fig:pde_clines}
+    }
+\end{figure}
 
@@
 
-\noindent 2.
-Uneven sampling has a strong influence on PCA projections \cite{mcvean2009genealogical}.
+\noindent 2. Uneven sampling has a strong influence on PCA projections \citep{mcvean2009genealogical}. 
 Our human data, POPRES, is unevenly sampled including 346 African-Americas, 73 Asians, 359 Indian Asians and 3187 Europeans. 
@@
 
-\noindent 4.
-Euclidean distance between the contracted matrix based on PCs is one measure of the similarity for window's population structure.
+\noindent 4. Euclidean distance between the contracted matrix based on PCs is one measure of the similarity for window's population structure.
 We want to try other methods of distance between windows, for example, we used the distance for PCs to reduce noise, however the distance between covariance matrixes of genotype matrix might also be informative.
@@
 \bibliographystyle{plainnat}
-\bibliography{references}  % if bibtex file is called references.bib
-
-% \begin{thebibliography}{99}
-% \bibitem{menozzi1978synthetic} Menozzi, P., Piazza, A. \& Cavalli-Sforza, L. \lq\lq Synthetic maps of human gene frequencies in Europeans.'' Science 201, 786--792 (1978).
-% \bibitem{novembre2008interpreting} Novembre, John, and Matthew Stephens.
-\lq\lq Interpreting principal component analyses of spatial population genetic variation." Nature genetics 40.5 (2008): 646-649.
-% \bibitem{mcvean2009genealogical} McVean, Gil. \lq\lq A genealogical interpretation of principal components analysis." PLoS Genet 5.10 (2009): e1000686.
-% \bibitem{novembre2008europe} Novembre, John, et al. \lq\lq Genes mirror geography within Europe." Nature 456.7218 (2008): 98-101.
-% \bibitem{key5} Borg, Ingwer, and Patrick JF Groenen. Modern multidimensional scaling: Theory and applications. Springer Science \& Business Media, 2005.
-% \bibitem{key6} Patterson, N., A. L. Price, and D. Reich. \lq\lq Population structure and eigenanalysis." PLoS Genet 2.12 (2006): e190.
-% \bibitem{key7} Price, Alkes L., et al. \lq\lq Principal components analysis corrects for stratification in genome- wide association studies." Nature genetics 38.8 (2006): 904-909.
-% \bibitem{key8} Nelson, Matthew R., et al.
-\lq\lq The Population Reference Sample, POPRES: a resource for population, disease, and pharmacological genetics research." The American Journal of Human Genetics 83.3 (2008): 347-358.
-% \bibitem{key9} Efron, Bradley, and B. Efron. The jackknife, the bootstrap and other resampling plans. Vol. 38. Philadelphia: Society for industrial and applied mathematics, 1982.
-% \bibitem{key10} Engelhardt, Barbara E., and Matthew Stephens.
-\lq\lq Analysis of population structure: a unifying framework and novel methods based on sparse factor analysis." PLoS Genet 6.9 (2010): e1001117.
-% \bibitem{key11} Corbett-Detig, Russell B., and Daniel L. Hartl. \lq\lq Population genomics of inversion polymorphisms in Drosophila melanogaster." (2012): e1003056.
-% \bibitem{key12} Antonacci, Francesca, et al. \lq\lq Characterization of six human disease-associated inversion polymorphisms." Human molecular genetics 18.14 (2009): 2555-2566.
-% \bibitem{key13} Ma, Jianzhong, and Christopher I. Amos. \lq\lq Investigation of inversion polymorphisms in the human genome using principal components analysis." (2012): e40224.
-% \bibitem{key14} Kulikova, Olga, et al. \lq\lq Satellite repeats in the functional centromere and pericentromeric heterochromatin of Medicago truncatula." Chromosoma 113.6 (2004): 276-283.
-% \bibitem{key15} Kulikova, Olga, et al. \lq\lq Integration of the FISH pachytene and genetic maps of Medicago truncatula." The Plant Journal 27.1 (2001): 49-58.
-% \bibitem{key16} Paape, Timothy, et al. \lq\lq Selection, genome-wide fitness effects and evolutionary rates in the model legume Medicago truncatula." Molecular ecology 22.13 (2013): 3525-3538.
-% \end{thebibliography}
-
+\bibliography{references}  
 \end{document}

### 2016-02-22 writeup/local_pca_paper.tex
a few comments

@@
 The closeness of two samples in PC plots represents the similarity of their genetic information.
-Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008genes,key6}.
+Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008genes,patterson2006population}.
 
-Most time, PCA used in genome-wide association study (GWAS) is for stratification correction \cite{key7}.
+Most time, PCA used in genome-wide association study (GWAS) is for stratification correction \citep{price2006principal}.
 However, for different part of genome, they have different genetic features.
@@
 
-\plr{(this is starting to be methods and results)}
+\plr{(This is starting to be methods/results: less detail?)}
 Therefore, we cut the chromosomes into windows with hundreds/thousands of SNPs and apply PCA locally on each window.
@@
 Based on the principal components, we can estimate the similarity of population structure contained in each genome window.
-To visualize the closeness for each window with regard to the population structure, we constructed matrix based on the first two PCs for each window and got the pairwise Euclidean distance between those matrixes, then use multidimensional scaling to reduces the pairwise distance matrix to lower dimension while preserve the distance information between windows as well as possible \cite{key5}.
+To visualize the closeness for each window with regard to the population structure, we constructed matrix based on the first two PCs for each window and got the pairwise Euclidean distance between those matrixes, then use multidimensional scaling to reduces the pairwise distance matrix to lower dimension while preserve the distance information between windows as well as possible \citep{borg2005modern}.
 To interpret the results of MDS, we combine known genome feature information for each species, such as the distribution of inversions, heterochromatin chunk and gene density along the genome.
@@
 
-\plr{How about: first describe the method; then afterwards, present the three datasets.
+\plr{How about: first describe the method; then afterwards, present the three datasets.}
 
@@
 
-\plr{Separate the different datasets into different paragraphs.}
+\plr{Separate paragraphs for each dataset. Say how many sites, what kind of data, etcetera.}
 
-For human SNP data from POPRES \cite{key8}, we use the allele that has highest frequency in the samples as the reference allele for each position.
+For human SNP data from POPRES \citet{nelson2008population}, we use the allele that has highest frequency in the samples as the reference allele for each position.
 If an allele is same with the reference allele, we recode it as 0; if an allele is different from the reference allele, we recode it as 1; for positions that have missing data, we recode them as NA.
@@
 \subsection{Cut each genome into windows}
+
+\plr{this describes PCA, not cutting into windows? and, the next section?}
 We cut each recoded matrix into sub matrixes by that have the same columns but fewer rows than the original matrix.
 Then apply Principal Component Analysis (PCA) on each window.
-Here's a brief summary about how is PCA carried out on genomic data \cite{mcvean2009genealogical}.
+Here's a brief summary about how is PCA carried out on genomic data \citep{mcvean2009genealogical}.
 Starting with the recoded genotype matrix Z, where Z is a L*N matrix ( L is SNP number ; N is sample size), then zero-center the matrix Z to X to make the data rows have equal variance.
@@
 The \textit{i}th principal component is the \textit{i}th eigenvector of C.
-
+\plr{(put equations into the text in the right place)}
 \begin{equation}
@@
 The variance between windows is estimated as mean variance of the first principal component for each window.
-The variance inside a window is estimated using the jackknife block when cutting the window into 10 equal size smaller windows \cite{key9}.
-Table 1 shows the comparison of variance within a window and that between windows for chromosome arms in \textit{Drosophila}.
+The variance inside a window is estimated using the jackknife block when cutting the window into 10 equal size smaller windows \cite{efron1982jackknife}.
+Table \ref{tab:window_sizes} shows the comparison of variance within a window and that between windows for chromosome arms in \textit{Drosophila}.
 Finally, we choose 100 SNPs, 1000 SNPs and 10000 SNPs as window length for human, \textit{Drosophila}, and \textit{Medicago} separately.
 
-Table 1.
+\begin{table}[ht]
+\centering
+\begin{tabular}{rllrrrrr}
+  \hline
+ & chr\_name & win\_length(in SNPs) & 100 & 500 & 10\verb|^|3 & 10\verb|^|4 & 10\verb|^|5 \\ 
+  \hline
+1 & Chr2L & SE\verb|^|2(within) & 2.05e-03 & 1.64e-03 & 1.18e-03 & 1.68e-04 & 4.02e-05 \\ 
+  2 & Chr2L & Var(between) & 2.76e-03 & 2.69e-03 & 2.23e-03 & 6.74e-04 & 3.12e-04 \\ 
+  3 & Chr2R & SE\verb|^|2(within) & 2.18e-03 & 1.92e-03 & 1.63e-03 & 5.76e-04 & 1.35e-04 \\ 
+  4 & Chr2R & Var(between) & 2.78e-03 & 2.70e-03 & 2.65e-03 & 2.31e-03 & 1.82e-03 \\ 
+  5 & Chr3L & SE\verb|^|2(within) & 2.08e-03 & 2.00e-03 & 1.64e-03 & 7.32e-04 & 2.45e-04 \\ 
+  6 & Chr3L & Var(between) & 2.60e-03 & 2.52e-03 & 2.40e-03 & 1.68e-03 & 1.89e-03 \\ 
+  7 & Chr3R & SE\verb|^|2(within) & 1.95e-03 & 1.76e-03 & 1.44e-03 & 5.87e-04 & 2.03e-04 \\ 
+  8 & Chr3R & Var(between) & 2.58e-03 & 2.51e-03 & 2.44e-03 & 1.96e-03 & 1.40e-03 \\ 
+  9 & ChrX & SE\verb|^|2(within) & 2.48e-03 & 2.04e-03 & 1.54e-03 & 1.62e-03 & 1.68e-04 \\ 
+  10 & ChrX & Var(between) & 2.61e-03 & 2.43e-03 & 2.30e-03 & 3.24e-04 & 1.14e-03 \\ 
+   \hline
+\end{tabular}
+\caption{
+Comparison of variance within a window and that between windows for chromosome arms in \textit{Drosophila}.
+} \label{tab:window_sizes}
+\end{table}
 
@@
 ($\lambda _{1i}$ and $\lambda _{2i}$ are the eigenvalues for the first two PCs for \textit{i}th window; $\lambda _{1j}$ and $\lambda _{2j}$ are the eigenvalues for the first two PCs for \textit{j}th window.)
+\plr{
+Say what $M_i$ and $M_j$ are in the text.  
+Also, how about we write $V$ instead of $PC$?  I like to use only one letter for variables, so it's clear it's just one thing, not the product of $P$ and $C$.
+And, isn't it $\sqrt{\lambda_{1j}^2+\lambda_{2j}^2}$ on the bottom? Check in the R code (pc{\textunderscore}dist.R).
+}
 
-\begin{equation}
-M_{i}=\frac{\lambda _{1i}PC1_{i}PC1_{i}^{T}+\lambda _{2i}PC2_{i}PC2_{i}^{T}}{\lambda _{1i}+\lambda _{2i}} \qquad
-M_{j}=\frac{\lambda _{1j}PC1_{j}PC1_{j}^{T}+\lambda _{2j}PC2_{j}PC2_{j}^{T}}{\lambda _{1j}+\lambda _{2j}}
-\end{equation}
+\begin{align}
+    M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} \\
+    M_{j} &= \frac{\lambda_{1j}PC1_{j}PC1_{j}^{T}+\lambda_{2j}PC2_{j}PC2_{j}^{T}}{\lambda_{1j}+\lambda_{2j}}
+\end{align}
 
@@
 could use the following method to calculate the pairwise distance greatly saving time and space.
+\plr{Define $D$.}
 
@@
 in each pair of genomic windows.
+
 \subsubsection{Visualize the pairwise distance matrix}
@@
 It can reduce the dimensionality of a distance matrix while preserve the distance information between objects as well as possible.
-The aimed dimension \lq\lq M\rq\rq can be set based on your need.
+The aimed dimension `` M'' can be set based on your need.
 We used M=1 and M=2 in our study.
@@
 Here are the MDS results for each species.
+
 \subsection{\textit{Drosophila}}
@@
 When setting the dimension parameter M=2 in MDS method, that is, plotting the first 2 coordinates reduced from the distance matrix, each plot looks like triangle.
-(Fig1.a) Since the relative position for each window in the plot shows the relative similarity between windows, it tells us there are 3 extreme types of population structure shown in the 3 peaks of the \lq\lq triangle\rq\rq, and other windows are between them, which means other window's population structure might be a mixture of those extremes.
+(Figure \ref{fig:mds_chr2L}a) Since the relative position for each window in the plot shows the relative similarity between windows, it tells us there are 3 extreme types of population structure shown in the 3 peaks of the ``triangle'', and other windows are between them, which means other window's population structure might be a mixture of those extremes.
 We then want to investigate more information at the extremes.
@@
     \begin{center}
-       % \includegraphics[width=0.7\textwidth]{figs/example_cline}
+       \includegraphics[width=0.9\textwidth]{Fig1_Together_MDS_plot_Chr2L}
     \end{center}
     \caption{
-        The caption goes here.
-        \label{fig:pde_clines}
+         a) Two dimensional MDS plot for chromosome arm Chr2L in \textit{Drosophila}.
+Each point in the plot stands for a window.
+The distance between windows shows the similarity of population structure for windows.
+b) The color corresponding to that in a) and thus shows a relative position for each window in a) along the genome.
+c) MDS 1D plot for Chr2L in \textit{Drosophila}, using the first coordinate in a) against the window id along the genome.
+The green, red, blue points are corresponding to the windows in the green, red, blue circled extremes in a).
+         \plr{(I think it's a bit confusing having both MDS1D and MDS2D: can we instead of MDS1D on the right, just plot the first coordinate of MDS2D?
+And, somehow make the colors agree between the two figures?
+(not sure how to do this)}
+         \label{fig:mds_chr2L}
     }
@@
 
-Fig2.
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{Fig2_all_pca_plots_for_Chr2L_3peaks}
+    \end{center}
+    \caption{
+        PCA plots for each peak in Figure \ref{fig:mds_chr2L}a and for whole Chr2L in \textit{Drosophila}.
+The color of the square corresponds to the circles of the 3 extremes in Fig.1a.
+(Also labeled same as 1,2,3)
+         \label{fig:pca_chr2L}
+    }
+\end{figure}
+
+\plr{Use figure labels, like ``Figure {\textbackslash}ref\{fig:mds{\textunderscore}chr2L\}'' instead of ``Figure 1''.}
 
 There's a known inversion in Chr2L in \textit{Drosophila}, In(2L)t, with breakpoints at 2225744bp and 13154180bp.
-\cite{key11} We recolored the PCA plots in Fig.2 a,b,c by the direction of the inversion for each sample using data from John Pool's lab.
-(Fig.3) The results show that the clustering in PCA plots are mainly due to inversions for the red and green extremes in Fig.2a,b (especially for green extreme).
+\citep{corbett2012population} We recolored the PCA plots in Fig.2 a,b,c by the direction of the inversion for each sample using data from John Pool's lab.
+(Fig.3)
+\plr{(citation?)}
+The results show that the clustering in PCA plots are mainly due to inversions for the red and green extremes in Fig.2a,b (especially for green extreme).
 What's more, we found that the two breakpoints of In(2L)t are located in the two clusters of the green points in Fig.1c.
@@
 
-Fig3.
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{Fig3_all_pca_plots_for_Chr2L_3peaks_color_by_In_2L_t}
+    \end{center}
+    \caption{
+         PCA plots colored by direction of In(2L)t direction for each peak in Chr2L MDS 2D plot. Red dots for samples with inversion, green dots for samples without inversion.
+        \label{fig:pde_clines}
+    }
+\end{figure}
 
@@
 We ran our method separately on all 22 autosomes.
-For chromosomes 3, 8, 15 and 17 have known inversions and the corresponding breakpoints \cite{key12}.
+For chromosomes 3, 8, 15 and 17 have known inversions and the corresponding breakpoints \citep{antonacci2009characterization}.
 We found that the outlying windows in one-dimensional MDS plots coincide with the position of inversions for those 4 chromosomes.(Fig.4) As in Drosophila, the biggest source of variability in population structure along human chromosomes is inversion.
 Other chromosomes haven't had experimentally validated inversions, but there are many predicted inversions, and PCA might provide a way to validate those predicted inversion.
-\cite{key13} (See supplementary for other chromosomes' MDS results for human)
+\citep{ma2012investigation} (See supplementary for other chromosomes' MDS results for human)
 
-Fig4.
+\plr{Readers will wonder if the "triangle" looks the same in humans and Medicago as in Drosophila. Should we add them, or just put these in the supplement?}
+
+\plr{Do we know what happens when we run all 22 autosomes together?}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.7\textwidth]{fig4}
+    \end{center}
+    \caption{
+         MDS results for human genome (chr3, chr8, chr15, chr17, chr6, chr12).
+The x-axis is the position on the genome in window number, and y-axis is the first coordinate of MDS.
+Black dots stand for windows, red lines show positions of inversions.
+        \label{fig:pde_clines}
+    }
+\end{figure}
 
@@
 We checked the MDS results for all 8 chromosomes. 
+\plr{The results looked different than the other two species, with much less pronounced peaks...}
 We found the position of the peak for each MDS plot has a coincidence with the position of heterochromatic regions.
@@
 
-Fig5.
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.7\textwidth]{fig5}
+    \end{center}
+    \caption{
+         MDS results for the \textit{Medicago} genome (chr1-8), using first coordinate of MDS against the middle position of each window along each chromosome. 
+         The bars under MDS plots are diagrams for each chromosome showing the relative position of heterochromatin (the deep purple part on each bar) \citep{kulikova2004satellite, kulikova2001integration}.
+        \label{fig:pde_clines}
+    }
+\end{figure}
 
-Fig6.
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{Fig6_MDS_1D_and_gene_count_and_lm_chr3_based_all_chr_win104}
+    \end{center}
+    \caption{
+        The first coordinate of MDS plotting against gene models gene count for chromosome 3 in \textit{Medicago}.
+(See supplementary for other chromosomes' MDS results against gene count for Medicago)
+        \label{fig:pde_clines}
+    }
+\end{figure}
 
-\section{Future work}
+%%%%%%% %%%%%%%%%
+\section{Discussion}
+
+We've shown that local summaries of tree shape, i.e., kinship, vary significantly on a large scale along the genome,
+and that exploratory visualization of this variation can be useful.
+
+There are many ways to visualize population structure; perhaps others would show different things.
+
+Come back to some points in the Introduction.
+
+\subsection{Future work}
 1.

### 2016-03-17 writeup/local_pca_paper.tex
Update local_pca_paper.tex

@@
 
+The kinship matrix goes back to almost Mendel.
+It contains the kinship coefficient for pairwise individuals, which defines the genetic relatedness between individuals and could be estimated from genome-wide covariances of genotype markers.
+Populations are often structured in some way while there are systematic genetic variation.
+Linear Mixed models make use of the kinship matrix and can correct for population structure in GWAS.
+\citep{astle2009population,sul2013mixed}
 About 37 years ago, \citet{menozzi1978synthetic} first applied Principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}.
 Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.
-PCA results are derived from the covariance matrix of genotype matrix, which describes population structure, demographic history, and pedigree tree and so on.
+PCA results are derived from the covariance matrix of genotype matrix, which could be used to estimate kinship matrix and describes information about population structure, demographic history, pedigree tree and so on.
 The results of PCA can be directly related to the underlying genealogical history of the samples, such as coalescence time (time to most recent common ancestor) and migration rate between populations \citep{menozzi1978synthetic,novembre2008interpreting,mcvean2009genealogical}.
-Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and plots of first two Principal components (PCs) can mimic the samples' geographic origin to some extent.
+Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography.
+Plots of first two Principal components (PCs) can mimic the samples' geographic origin to some extent.
 The closeness of two samples in PC plots represents the similarity of their genetic information.
 Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008genes,patterson2006population}.
-
+It is worth mentioning that this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns.
+\citep{astle2009population}
+It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, and depend on where on the genome you look; this is why scans for selective sweeps work.
 Most time, PCA used in genome-wide association study (GWAS) is for stratification correction \citep{price2006principal}. 
 However, for different part of genome, they have different genetic features. 
-First, each site of DNA may have different gene tree.
+First, Locally everything is treelike and each site of DNA may have different gene tree. 
+The covariance matrix of genotype matrix averages those gene trees. 
 For a DNA region, if individuals have SNPs closer in gene tree, they tends to be more close in PC projections for that region. 
@@
 
-For human SNP data from POPRES \citet{nelson2008population}, we use the allele that has highest frequency in the samples as the reference allele for each position.
+For human, we use SNP chip data from POPRES \citet{nelson2008population}.
+There are 3965 samples in total, (346 African-Americas; 73 Asians; 3187 Europeans; 359 Indian Asians), and the 22 autosomes together have 447267 SNPs in this dataset.
+We use the allele that has highest frequency in the samples as the reference allele for each position. 
 If an allele is same with the reference allele, we recode it as 0; if an allele is different from the reference allele, we recode it as 1; for positions that have missing data, we recode them as NA. 
 Since human genome is diploid, we add the value for the two alleles and then one chromosome will eventually be recode to a sequence consisting of 0,1,2 (and NA).
-There are 3965 samples in total, (346 African-Americas; 73 Asians; 3187 Europeans; 359 Indian Asians); then the genome data for human is recoded to a matrix that has 3965 columns, each column for an individual's genotype.
+Then the genome data for human is recoded to a matrix that has 3965 columns, each column for an individual's genotype. 
 We process the data separately for the 22 autosomes in human.
-For \textit{Drosophila}, we first process the sequencing data (from DPGP and John pool's lab) to SNP data by eliminating the positions that have all the same alleles.
+
+For \textit{Drosophila}, we use the sequencing data from DPGP and John pool's lab, which together has 380 samples from 16 countries.
+We first process the sequencing data (from DPGP and John pool's lab) to SNP data by eliminating the positions that have all the same alleles. 
+Each chromosome arms we investigated (Chr2L, Chr2R, Chr3L, Chr3R, ChrX) has 2-3 million SNPs.
 Due to high density of missing data for some parts in the genome, we then delete the samples with more than 8\% NAs and positions with more than 20\%. 
@@
 Since the \textit{Drosophila} samples here are all homogenous, the matrices are indeed consisting of 0,1 (and NA).
-For \textit{Medicago truncatula}, we use the SNP data from Medicago Hapmap and recode it to matrices with 0,1,2 (and NA) for each of the 8 chromosomes.
+
+For \textit{Medicago truncatula}, we use the SNP data from Medicago Hapmap. It has 263 samples from 24 countries. 
+We recode it to matrices with 0,1,2 (and NA) for each of the 8 chromosomes. Each chromosome has 3-5 million SNPs
 
@@
 For example, the constructed matrixes for ith and jth window are as following. 
-($\lambda _{1i}$ and $\lambda _{2i}$ are the eigenvalues for the first two PCs for \textit{i}th window; $\lambda _{1j}$ and $\lambda _{2j}$ are the eigenvalues for the first two PCs for \textit{j}th window.)
+($\lambda _{1i}$ and $\lambda _{2i}$ are the eigenvalues for the first two PCs for \textit{i}th window; $\lambda _{1j}$ and $\lambda _{2j}$ are the eigenvalues for the first two PCs for \textit{j}th window.) $M_i$ and $M_j$ are the constructed new matrix for \textit{i}th window and \textit{j}th window.
 \plr{
@@
 
-The Euclidean distance between the constructed matrices for windows stands for the
-similarity of population structure for windows. Due to the property of eigenvectors,we
-could use the following method to calculate the pairwise distance greatly saving time and space.
+The Euclidean distance $D_{ij}$ between the constructed matrices $M_i$ and $M_j$ stands for the similarity of population structure for the \textit{i}th window and \textit{j}th window. 
+Due to the property of eigenvectors,we could use the following method to calculate the pairwise distance greatly saving time and space.
+
 \plr{Define $D$.}

### 2016-03-30 writeup/local_pca_paper.tex
Update local_pca_paper.tex

@@
 \section{Introduction}
+The kinship matrix contains the kinship coefficient for pairwise individuals. 
+Kinship coefficient defines the genetic relatedness between individuals.
+It is the probability that two alleles randomly selected from two individuals are inherited from the most recent common ancester.
+It could be estimated from given pedigree or from genome-wide covariances of genotype markers. 
+It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, and depend on where on the genome you look; this is why scans for selective sweeps work.
+Populations are often structured in some way while there are systematic genetic variation between populations.
 
-The kinship matrix goes back to almost Mendel.
-It contains the kinship coefficient for pairwise individuals, which defines the genetic relatedness between individuals and could be estimated from genome-wide covariances of genotype markers.
-Populations are often structured in some way while there are systematic genetic variation.
-Linear Mixed models make use of the kinship matrix and can correct for population structure in GWAS.
-\citep{astle2009population,sul2013mixed}
-About 37 years ago, \citet{menozzi1978synthetic} first applied Principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}.
+About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
 Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.  
-PCA results are derived from the covariance matrix of genotype matrix, which could be used to estimate kinship matrix and describes information about population structure, demographic history, pedigree tree and so on.
-The results of PCA can be directly related to the underlying genealogical history of the samples, such as coalescence time (time to most recent common ancestor) and migration rate between populations \citep{menozzi1978synthetic,novembre2008interpreting,mcvean2009genealogical}.
+PCA results are derived from the covariance matrix of genotype matrix. 
+The results of PCA can be directly related to the underlying genealogical history of the samples, 
+such as coalescence time (time to most recent common ancestor) and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}. 
 Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
-Plots of first two Principal components (PCs) can mimic the samples' geographic origin to some extent.
-The closeness of two samples in PC plots represents the similarity of their genetic information.
-Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008genes,patterson2006population}.
-It is worth mentioning that this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns.
-\citep{astle2009population}
-It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, and depend on where on the genome you look; this is why scans for selective sweeps work.
-Most time, PCA used in genome-wide association study (GWAS) is for stratification correction \citep{price2006principal}. 
-However, for different part of genome, they have different genetic features. 
-First, Locally everything is treelike and each site of DNA may have different gene tree. 
+Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
+Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}.
+However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}
+
+PCA is often used in genome-wide association studies (GWAS) for stratification correction \citep{price2006principal}. 
+However, different parts of genome have different genetic features. 
+First, each site of DNA may have different gene tree. 
 The covariance matrix of genotype matrix averages those gene trees. 
-For a DNA region, if individuals have SNPs closer in gene tree, they tends to be more close in PC projections for that region. 
-Different DNA segments may have different gene tree for their SNPs and would result in different population structure for those segments. 
+For a genomic region, if individuals have alleles closer in gene tree, they tends to be more close in PC projections for that region. 
+Different DNA segments may have different gene tree and therefore different population structure for those segments. 
 Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
-Third, if chromosome inversion exists, the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints, thus results in different genome structure and population structure.
-Investigating the genomic variation along the genome can help us to have a better understanding of the relation between genome structure and population structure.
+Selective sweeps cause local recent ancestry or short trees.
+Balancing selection causes deep trees.
+Background selection causes shallow ones.
+Third, if a chromosome inversion is polymorphic in the sample, the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints.
+Recombination suppression across inversions thus results in different genome structure and population structure.
+Other effects, like noise, introgression might also influence population structure.
 
-\plr{(This is starting to be methods/results: less detail?)}
-Therefore, we cut the chromosomes into windows with hundreds/thousands of SNPs and apply PCA locally on each window.
+Investigating the genomic variation along the genome can help us to have a better understanding of the relation between genome structure and population structure, and could possibly lead to more powerful methods for GWAS.
+
+To investigate this, we cut each chromosome into windows (with hundreds to thousands of SNPs in each), applied PCA to each window, and visualized how population structure, as summarized by PCA, varies along windows.
+
 In this project, we used SNP data for human, \textit{Medicago truncatula}, and whole genome sequencing data for \textit{Drosophila}.
+Based on the principal components, we can estimate the similarity of population structure contained in each genome window. 
+To quantify similarity of population structure between windows, we constructed for each window an approximate, scaled covariance matrix based on the first two PCs measured the pairwise Euclidean distance between those matrices.
+We use multidimensional scaling to visualize the relationships between windows, which reduces the pairwise distance matrix to lower dimension while preserving the distance information between windows as well as possible \citep{borg2005modern}.
+To interpret the results of MDS, we combine known genome feature information for each species, such as the distribution of inversions, and heterochromatin and gene density along the genome.
+Each species showed distinct patterns, reflecting differences in their biology.
 
-Based on the principal components, we can estimate the similarity of population structure contained in each genome window.
-To visualize the closeness for each window with regard to the population structure, we constructed matrix based on the first two PCs for each window and got the pairwise Euclidean distance between those matrixes, then use multidimensional scaling to reduces the pairwise distance matrix to lower dimension while preserve the distance information between windows as well as possible \citep{borg2005modern}.
-To interpret the results of MDS, we combine known genome feature information for each species, such as the distribution of inversions, heterochromatin chunk and gene density along the genome.
-For different species, we got different explanation for the variation of population structure along the genome.
+Other methods for visualizing population structure are like STRUCTURE, \citep{pritchard2000inference,falush2003inference, falush2007inference,hubisz2009inferring} model-based approach, \citep{yang2012modelbased} maps of heterozygosity \citep{ramachandran2005support}.
+The term ``local PCA" is also used in some other region for different meanings.
+\citep{manjon2013diffusion,kambhatla1997dimension,weingessel2000local}
 

### 2016-03-30 writeup/local_pca_paper.tex
peter comments to intro

@@
 \usepackage{color}
+\usepackage{marginnote}
 \newcommand{\plr}[1]{{\em \color{blue} #1}}
+\newcommand{\plrm}[1]{\marginnote{\em \color{blue} #1}}
 
@@
 
-\plr{Note: your intro focuses on PCA more than mine, which is trying to be about population structure generally.}
-
 \section{Introduction}
+\plr{Should cite something, probably Astle \& Balding in this paragraph.}
+\plr{(Try to make these next three sentences into one sentence.)}
 The kinship matrix contains the kinship coefficient for pairwise individuals. 
@@
 It is the probability that two alleles randomly selected from two individuals are inherited from the most recent common ancester.
+\plr{(This definition is not quite right: two alleles are \emph{always} inherited from their most recent common ancestor.)}
 It could be estimated from given pedigree or from genome-wide covariances of genotype markers. 
-It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, and depend on where on the genome you look; this is why scans for selective sweeps work.
+\plr{(Estimating the "kinship matrix" from pedigrees, or from covariances, are two different things.)}
+It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, 
+\plr{(explain what the expected value is)}
+and depend on where on the genome you look; 
+this is why scans for selective sweeps work.
+\plr{(do you think this last comment about scans for selective sweeps needs explanation?)}
 Populations are often structured in some way while there are systematic genetic variation between populations. 
+\plr{(Instead, how about: 
+Structure in the kinship matrix, such as a correlation between geographic proximity and relatedness,
+is often referred to as ``population structure''.)}
 
-About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
+About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \plr{(instead, ``summarizing population structure''?)} \citep{menozzi1978synthetic}.
 Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.  
 PCA results are derived from the covariance matrix of genotype matrix.
+\plr{(We can call this the ``genetic covariance matrix''.)}
 The results of PCA can be directly related to the underlying genealogical history of the samples, 
-such as coalescence time (time to most recent common ancestor) and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}. 
+such as \plr{mean} coalescence time (time to most recent common ancestor) and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}. 
 Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
@@
 Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}.
-However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}
+However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns, for example, social or religious groups. \citep{astle2009population}
 
 PCA is often used in genome-wide association studies (GWAS) for stratification correction \citep{price2006principal}. 
-However, different parts of genome have different genetic features. 
+However, different parts of \plr{the} genome have different genetic features. 
 First, each site of DNA may have different gene tree. 
@@
 For a genomic region, if individuals have alleles closer in gene tree, they tends to be more close in PC projections for that region. 
+\plr{(I don't think it makes sense to say that alleles are close in a gene tree.)}
 Different DNA segments may have different gene tree and therefore different population structure for those segments. 
 Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
+\plr{(Add citations for the following.)}
 Selective sweeps cause local recent ancestry or short trees.
@@
 Recombination suppression across inversions thus results in different genome structure and population structure.
+\plr{(I don't think noise affects population structure, since I'd define population structure to be something like the mean, and that noise just obscures population structure.)}
+\plr{(Should mention (and cite) why the effects of introgression might be different along the genome.)}
 Other effects, like noise, introgression might also influence population structure.
@@
 
+\plr{(Replace ``this'' with what you mean by ``this'' in the next sentence.)}
 To investigate this, we cut each chromosome into windows (with hundreds to thousands of SNPs in each), applied PCA to each window, and visualized how population structure, as summarized by PCA, varies along windows.
@@
 Based on the principal components, we can estimate the similarity of population structure contained in each genome window. 
-To quantify similarity of population structure between windows, we constructed for each window an approximate, scaled covariance matrix based on the first two PCs measured the pairwise Euclidean distance between those matrices.
-We use multidimensional scaling to visualize the relationships between windows, which reduces the pairwise distance matrix to lower dimension while preserving the distance information between windows as well as possible \citep{borg2005modern}.
+To quantify similarity of population structure between windows, we constructed for each window an approximate, scaled covariance matrix based on the first two PCs \plr{and} measured the pairwise Euclidean distance between those matrices.
+We use multidimensional scaling \plr{(MDS)} to visualize the relationships between windows, which reduces the pairwise distance matrix to lower dimension while preserving the distance information between windows as well as possible \citep{borg2005modern}.
 To interpret the results of MDS, we combine known genome feature information for each species, such as the distribution of inversions, and heterochromatin and gene density along the genome.
@@
 
+\plr{(These two sentences need more detail.
+How does what we're doing differ from them?
+Especially, say, chromopainter, argweaver, and \url{http://www.ncbi.nlm.nih.gov/pubmed/26945783}?
+What does local PCA mean in other places, or at lesat, what other fields is it used in?)}
 Other methods for visualizing population structure are like STRUCTURE, \citep{pritchard2000inference,falush2003inference, falush2007inference,hubisz2009inferring} model-based approach, \citep{yang2012modelbased} maps of heterozygosity \citep{ramachandran2005support}.

### 2016-04-12 1kG/workflow.Rmd
thousand genomes init

---
title: "Local PCA on 1000 genomes"
date: "`r date()`"
---

```{r setup, include=FALSE}
library(devtools)
load_all("../package")
# Chromosome map lengths, from:
.chrlens <- c( 262.00830, 244.03647, 209.97332, 197.48005, 189.58412, 174.70526, 172.35610, 156.52810, 146.28861, 160.40336, 143.88007, 156.40816, 119.34065, 104.83476, 110.93523, 118.68929, 119.61966, 105.74726, 92.57134, 83.62057, 54.79314, 55.59191 )
.chrstarts <- c(0, cumsum(.chrlens) )
# and in bp
.chrbps <- as.numeric( c( 247135059, 242663303, 199318155, 191167888, 180625439, 170747902, 158798338, 146264218, 140185941, 135284541, 134449982, 132276874, 114092980, 106356482, 100210760, 88684276, 78605474, 76115554, 63785051, 62376958, 46902240, 49522492 ) )
.chrbpstarts <- c(0,cumsum(.chrbps))
```

# The data

We have VCFs from *phase 3* of the 1000 genomes, in separate VCF files by chromosome,
from 2504 individuals.
Of these, 24 are relatives; we'll remove these.
```{r vcfs}
vcfs <- list.files("phase3",".*vcf.gz$", full.names=TRUE)
full.sample.info <- read.csv("20130606_sample_info.csv",header=TRUE,stringsAsFactors=FALSE)
vcf.header <- scan( pipe( paste0( "zcat ", vcfs[1], " | head -n 1000 | grep '^#[^#]'")), what='char' )
vcf.indivs <- vcf.header[-(1:9)]
all( vcf.indivs %in% full.sample.info$Sample )
sample.info <- subset(full.sample.info, Sample %in% vcf.indivs)
rellies <- unlist(subset(sample.info, ! ( Relationship %in% c("", "unrels", "unrel", "not father")) )[,7:14])
rellies <- unlist( strsplit( rellies[nchar(rellies)>0], "[ ,]" ) )
included.rellies <- intersect( rellies, vcf.indivs )
sample.info$usethese <- ! ( sample.info$Sample %in% included.rellies )
table(subset(sample.info,usethese)$Population)
```

Now, we would like to find the eigendecomposition
of each genomic window.
We'll start with chromosome 22.
```{r chr22, cache=TRUE}
vcf.chr22 <- grep("chr22",vcfs,value=TRUE)
# first do a scan to find the scale we're dealing with
coords <- scan( pipe( paste0( "zcat ", vcf.chr22, " | grep -v '^#' | cut -f 2 " ) ) )
length(coords)
summary( diff(coords) )
# percent of spacings less than 100bp
mean(diff(coords)<100)
length(coords)
```
On chromsome 22, which is 1.7% of the genome, there are 1.1 million SNPs,
median spacing between SNPs is 20bp, and 95% of spacings are below 100bp.

### 2016-04-12 writeup/local_pca_paper.tex
Update local_pca_paper.tex

--- 2016-03-30 writeup/local_pca_paper.tex
@@
 
-\title{Using Local PCA to Investigate Population Structure Along the Genome}
+\title{Local PCA Shows How Population Structure Differs Along the Genome}
 \author{Han Li, Peter Ralph}
@@
 \section{Introduction}
+
 The kinship matrix contains the kinship coefficient for pairwise individuals. 
@@
 
-For \textit{Drosophila}, we use the sequencing data from DPGP and John pool's lab, which together has 380 samples from 16 countries.
+For \textit{Drosophila}, we use the sequencing data from DPGP and John pool's lab \citet{lack2015drosophila}, which together has 380 samples from 16 countries.
 We first process the sequencing data (from DPGP and John pool's lab) to SNP data by eliminating the positions that have all the same alleles. 
@@
 
-\subsection{Cut each genome into windows}
+\subsection{PCA in genomic windows}
 
@@
 Here's a brief summary about how is PCA carried out on genomic data \citep{mcvean2009genealogical}.
-Starting with the recoded genotype matrix Z, where Z is a L*N matrix ( L is SNP number ; N is sample size), then zero-center the matrix Z to X to make the data rows have equal variance.
-Get the covariance matrix of X (denote as matrix C) and compute the eigenvectors and eigenvalues of the covariance matrix C.
-The \textit{i}th principal component is the \textit{i}th eigenvector of C.
+Starting with the recoded genotype matrix Z, where Z is a L*N matrix ( L is SNP number ; N is sample size), then zero-center the matrix Z to X to make the data rows have equal mean.
+($X_{si}=Z_{si}-\frac{1}{n}\sum_{j=1}^{n}Z_{sj}$) Get the covariance matrix of X (denote as matrix C) and compute the eigenvectors and eigenvalues of the covariance matrix C.
+($C=\frac{1}{n-1}XX^{T}$ ) The \textit{i}th principal component is the \textit{i}th eigenvector of C.
 \plr{(put equations into the text in the right place)}
-\begin{equation}
-X_{si}=Z_{si}-\frac{1}{n}\sum_{j=1}^{n}Z_{sj} 
-\end{equation}
-\begin{equation}
-C=\frac{1}{n-1}XX^{T}
-\end{equation}
 
-\subsubsection{Genomic PCA on windows}
-PCA on each window is getting the eigenvectors of covariance matrix of the cut recoded matrix.
-PCA plots (generally using PC2 against PC1) can show population structure, which describes how different samples are related.
+PCA on each window is getting the eigenvectors of covariance matrix of the cut recoded matrix. PCA plots generally use PC2 against PC1.
+PC1 and PC2 refer to the first and second principal component respectively. 
+PCA plots can show population structure, which describes how different samples are related. 
 We want to investigate the variation of population structure along the genome. 
@@
 V_{1}=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} \qquad
-V_{2}=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i}
+V_{2}=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} \qquad
+U_{1}=\sqrt{\frac{\lambda _{1j}}{\lambda _{1j}+\lambda _{2j}}}PC1_{j} \qquad
+U_{2}=\sqrt{\frac{\lambda _{2j}}{\lambda _{1j}+\lambda _{2j}}}PC2_{j}
 \end{equation}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig1_Together_MDS_plot_Chr2L}
+       \includegraphics[width=0.9\textwidth]{Fig1_Together_MDS_plot_Chr2L_final_abline}
     \end{center}
     \caption{
-         a) Two dimensional MDS plot for chromosome arm Chr2L in \textit{Drosophila}.
-Each point in the plot stands for a window.
-The distance between windows shows the similarity of population structure for windows.
-b) The color corresponding to that in a) and thus shows a relative position for each window in a) along the genome.
-c) MDS 1D plot for Chr2L in \textit{Drosophila}, using the first coordinate in a) against the window id along the genome.
-The green, red, blue points are corresponding to the windows in the green, red, blue circled extremes in a).
+         Variation in structure for windows on chromosome arm Chr2L. 
+         Each represents a window: (a) shows the MDS visualization of relationships between windows. 
+         (b) and (c) show the position (midpoint) of each window against the first and second MDS coordinates, respectively. 
+         Colors are consistent for (a) (b) and (c).
+         Vertical black lines show the breakpoints of known polymorphic inversions In(2L)t.
          \plr{(I think it's a bit confusing having both MDS1D and MDS2D: can we instead of MDS1D on the right, just plot the first coordinate of MDS2D?
@@
 
-We pick a window for each extreme, and take out 5\% windows that have the smallest distance to it in the original pairwise distance matrix, then combine those windows for each extreme and apply PCA on them.
-(Fig.2) We can see the obvious difference between their PCA plots.
+We pick a window for each extreme, and take out 5\% windows that are closest to it in the MDS 2D plot , then combine those windows for each extreme and apply PCA on them.
+(Figure \ref{fig:pca_chr2L}) We can see the obvious difference between their PCA plots.
 This difference stands for variation of population structure along the genome in \textit{Drosophila} Chr2L.
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig2_all_pca_plots_for_Chr2L_3peaks}
+       \includegraphics[width=0.9\textwidth]{Fig2_all_pca_plots_for_Chr2L_3peaks_byMDS}
     \end{center}
     \caption{
-        PCA plots for each peak in Figure \ref{fig:mds_chr2L}a and for whole Chr2L in \textit{Drosophila}.
-The color of the square corresponds to the circles of the 3 extremes in Fig.1a.
-(Also labeled same as 1,2,3)
+        PCA plots for each peak in Figure \ref{fig:mds_chr2L}a. 
+        Each point represents a sample. 
+        The plots are plotting the second principal components against the first principal components. 
+        (a) shows the windows colored green in Figure 1; 
+        (b) shows orange windows; and (c) shows purple windows.
          \label{fig:pca_chr2L}
@@
 There's a known inversion in Chr2L in \textit{Drosophila}, In(2L)t, with breakpoints at 2225744bp and 13154180bp.
-\citep{corbett2012population} We recolored the PCA plots in Fig.2 a,b,c by the direction of the inversion for each sample using data from John Pool's lab.
-(Fig.3)
+\citep{corbett2012population} We recolored the PCA plots in Figure \ref{fig:pca_chr2L} a,b,c by the direction of the inversion for each sample using data from John Pool's lab \citep{lack2015drosophila}.
+Figure \ref{fig:color_inver}
 \plr{(citation?)}
-The results show that the clustering in PCA plots are mainly due to inversions for the red and green extremes in Fig.2a,b (especially for green extreme).
-What's more, we found that the two breakpoints of In(2L)t are located in the two clusters of the green points in Fig.1c.
-Similar results are found in other chromosome arms that have known inversions (Chr2R,
-Chr3L, Chr3R) in Drosophila. (See supplementary for more details) These facts show that the variation of population structure along the Drosophila genome is mainly due to inversions.
+The results show that PCA plots, and thus population structure, varies significantly along the genome, with most variation closely related to inversions. 
+Extremes of structure occurred around inversion breakpoints or in the centers of inversions (green and orange parts in Chr2L).
+The corresponding PCA plots show that locally, population structure is mostly determined by which orientation of the inversion each sample has.
+Similar results are found in other chromosome arms that have known polymorphic inversions (Chr2R, Chr3L, Chr3R) in \textit{Drosophila}. (See supplementary for more details)
+These facts show that the variation of population structure along the Drosophila genome is mainly due to inversions.
 
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig3_all_pca_plots_for_Chr2L_3peaks_color_by_In_2L_t}
+       \includegraphics[width=0.9\textwidth]{Fig3_all_pca_plots_for_Chr2L_3peaks_color_by_In2Lt_byMDS}
     \end{center}
     \caption{
-         PCA plots colored by direction of In(2L)t direction for each peak in Chr2L MDS 2D plot. Red dots for samples with inversion, green dots for samples without inversion.
-        \label{fig:pde_clines}
+         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversion, In(2L)t.
+        \label{fig:color_inver}
     }
@@
 We ran our method separately on all 22 autosomes. 
-For chromosomes 3, 8, 15 and 17 have known inversions and the corresponding breakpoints \citep{antonacci2009characterization}.
-We found that the outlying windows in one-dimensional MDS plots coincide with the position of inversions for those 4 chromosomes.(Fig.4) As in Drosophila, the biggest source of variability in population structure along human chromosomes is inversion.
-Other chromosomes haven't had experimentally validated inversions, but there are many predicted inversions, and PCA might provide a way to validate those predicted inversion.
-\citep{ma2012investigation} (See supplementary for other chromosomes' MDS results for human)
+We found that the windows that are outliers in the first MDS coordinate (Figure \ref{fig:mds_human}b) coincide with the position of known polymorphic inversions on 8p23. 
+Similar results are found in other chromosomes that have known polymorphic inversions (Chromosome 15, Chromosome 17) in Human. (See supplementary for more details)
+Similar to Drosophila, the biggest source of variability in population structure along each human chromosome seems to be polymorphic inversions.
+Other chromosomes showed similar results around predicted inversions: PCA might provide an additional way to identify inversions \citep{ma2012investigation}.
+When we run the process for all 22 autosomes together, the outlying signal of chromosome 8 is still impressing. (See supplementary for all 22 autosomes' MDS results for human)
 
@@
     \begin{center}
-       \includegraphics[width=0.7\textwidth]{fig4}
+       \includegraphics[width=0.7\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_final_abline}
     \end{center}
     \caption{
-         MDS results for human genome (chr3, chr8, chr15, chr17, chr6, chr12).
-The x-axis is the position on the genome in window number, and y-axis is the first coordinate of MDS.
-Black dots stand for windows, red lines show positions of inversions.
-        \label{fig:pde_clines}
+         Variation in structure between windows on human chromosome 8. Each point in the plot represents a window. 
+         (a) shows the MDS visualization of relationships between windows; 
+         (b) and third (c) show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
+         The vertical red lines show the breakpoints of the known inversion on 8p23. \citep{antonacci2009characterization}
+        \label{fig:mds_human}
     }
@@
 This means the population structure in the windows located in heterochromatin tends to have higher similarity, since those windows are closer in MDS plots.
-(Fig.5) Biologically,
+(Figure \ref{fig:mds_medicago}) Biologically,
 heterochromatic regions have lower gene density and may be less subject to selection \citep{kulikova2001integration,paape2013selection}.
@@
 Numerically, the first MDS coordinate value is negatively correlated to the gene count for each window, which shows the correlation between MDS result and gene density.
-(Fig.6) This fact shows that in \textit{Medicago}, the variation of population structure is correlated with heterochromatin and gene density, and is perhaps due to linked selection.
+(Figure \ref{fig:mds_gene_count})
+This fact shows that in \textit{Medicago truncatula}, variation in population structure, as measured by the first MDS coordinate, is negatively correlated to gene density. 
+Unlike in the human and Drosophila genomes, in \textit{Medicago truncatula}, major variation in population structure is likely due to linked selection.
 
@@
          The bars under MDS plots are diagrams for each chromosome showing the relative position of heterochromatin (the deep purple part on each bar) \citep{kulikova2004satellite, kulikova2001integration}.
-        \label{fig:pde_clines}
+        \label{fig:mds_medicago}
     }
@@
     \caption{
-        The first coordinate of MDS plotting against gene models gene count for chromosome 3 in \textit{Medicago}.
-(See supplementary for other chromosomes' MDS results against gene count for Medicago)
-        \label{fig:pde_clines}
+        First MDS coordinate against gene density for chromosome 3. 
+        (A) First MDS coordinate against the position (midpoint) of each window. 
+        (B) Gene count for each window against the position of each window; 
+        (C) The first MDS coordinate is significantly correlated with gene count (r=XXX, p=XXX). 
+        (See supplementary for other chromosomes' MDS results against gene count for \textit{Medicago})
+        \label{fig:mds_gene_count}
     }

### 2016-04-12 writeup/local_pca_paper.tex
removed some comments

@@
 \begin{equation}
-V_{1}=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} \qquad
-V_{2}=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} \qquad
-U_{1}=\sqrt{\frac{\lambda _{1j}}{\lambda _{1j}+\lambda _{2j}}}PC1_{j} \qquad
-U_{2}=\sqrt{\frac{\lambda _{2j}}{\lambda _{1j}+\lambda _{2j}}}PC2_{j}
+    \begin{aligned}
+        V_{1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} \qquad
+        V_{2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} \\
+        U_{1}&=\sqrt{\frac{\lambda _{1j}}{\lambda _{1j}+\lambda _{2j}}}PC1_{j} \qquad
+        U_{2}&=\sqrt{\frac{\lambda _{2j}}{\lambda _{1j}+\lambda _{2j}}}PC2_{j}
+    \end{aligned}
 \end{equation}
@@
          Vertical black lines show the breakpoints of known polymorphic inversions In(2L)t.
-         \plr{(I think it's a bit confusing having both MDS1D and MDS2D: can we instead of MDS1D on the right, just plot the first coordinate of MDS2D?
-And, somehow make the colors agree between the two figures?
-(not sure how to do this)}
          \label{fig:mds_chr2L}
@@
 \end{figure}
-
-\plr{Use figure labels, like ``Figure {\textbackslash}ref\{fig:mds{\textunderscore}chr2L\}'' instead of ``Figure 1''.}
 

### 2016-04-13 writeup/local_pca_paper.tex
added local pca paragraph

@@
 
-Investigating the genomic variation along the genome can help us to have a better understanding of the relation between genome structure and population structure, and could possibly lead to more powerful methods for GWAS.
+Investigating the genomic variation along the genome can help us to have a
+better understanding of the relation between genome structure and population
+structure, and could possibly lead to more powerful methods for GWAS.
 
-To investigate this, we cut each chromosome into windows (with hundreds to thousands of SNPs in each), applied PCA to each window, and visualized how population structure, as summarized by PCA, varies along windows.
+To investigate this, we cut each chromosome into windows (with hundreds to
+thousands of SNPs in each), applied PCA to each window, and visualized how
+population structure, as summarized by PCA, varies along windows.
 
-In this project, we used SNP data for human, \textit{Medicago truncatula}, and whole genome sequencing data for \textit{Drosophila}.
-Based on the principal components, we can estimate the similarity of population structure contained in each genome window. 
-To quantify similarity of population structure between windows, we constructed for each window an approximate, scaled covariance matrix based on the first two PCs measured the pairwise Euclidean distance between those matrices.
-We use multidimensional scaling to visualize the relationships between windows, which reduces the pairwise distance matrix to lower dimension while preserving the distance information between windows as well as possible \citep{borg2005modern}.
-To interpret the results of MDS, we combine known genome feature information for each species, such as the distribution of inversions, and heterochromatin and gene density along the genome.
-Each species showed distinct patterns, reflecting differences in their biology.
+In this project, we used SNP data for human, \textit{Medicago truncatula}, and
+whole genome sequencing data for \textit{Drosophila}.  Based on the principal
+components, we can estimate the similarity of population structure contained in
+each genome window.  To quantify similarity of population structure between
+windows, we constructed for each window an approximate, scaled covariance
+matrix based on the first two PCs measured the pairwise Euclidean distance
+between those matrices.  We use multidimensional scaling to visualize the
+relationships between windows, which reduces the pairwise distance matrix to
+lower dimension while preserving the distance information between windows as
+well as possible \citep{borg2005modern}.  To interpret the results of MDS, we
+combine known genome feature information for each species, such as the
+distribution of inversions, and heterochromatin and gene density along the
+genome. Each species showed distinct patterns, reflecting differences in their
+biology.
 
-Other methods for visualizing population structure are like STRUCTURE, \citep{pritchard2000inference,falush2003inference, falush2007inference,hubisz2009inferring} model-based approach, \citep{yang2012modelbased} maps of heterozygosity \citep{ramachandran2005support}.
-The term ``local PCA" is also used in some other region for different meanings.
-\citep{manjon2013diffusion,kambhatla1997dimension,weingessel2000local}
+Other methods for visualizing population structure are like STRUCTURE,
+\citep{pritchard2000inference,falush2003inference,
+falush2007inference,hubisz2009inferring} 
+model-based approach,
+\citep{yang2012modelbased} 
+maps of heterozygosity
+\citep{ramachandran2005support}.
+
+A number of methods for dimensionality reduction also use a strategy of ``local PCA''
+\citep[e.g.][]{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear},
+performing PCA not on the entire dataset but instead on subsets,
+providing ``local'' pictures which are then stitched back together to give a global picture.
+Our method shares this common thread, but differs somewhat in the context and goals,
+and future methods may benefit from this substantial literature \citep[reviewed in][]{vandermaaten2009dimensionality}.
 

### 2016-04-13 writeup/local_pca_paper.tex
corrected local pca bit

@@
 providing ``local'' pictures which are then stitched back together to give a global picture.
-Our method shares this common thread, but differs somewhat in the context and goals,
-and future methods may benefit from this substantial literature \citep[reviewed in][]{vandermaaten2009dimensionality}.
+This differs from our use of the term in that we restrict to subsets of \emph{variables} instead of subsets of observations.
+Future methods for visualization of genomic data
+may benefit from other advances in this substantial literature \citep[reviewed in][]{vandermaaten2009dimensionality}.
 

### 2016-04-13 writeup/local_pca_paper.tex
one more bit

@@
 
+\plr{Improve this paragraph:}
 Other methods for visualizing population structure are like STRUCTURE,
@@
 \citep[e.g.][]{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear},
-performing PCA not on the entire dataset but instead on subsets,
+performing PCA not on the entire dataset but instead on subsets of observations,
 providing ``local'' pictures which are then stitched back together to give a global picture.
-This differs from our use of the term in that we restrict to subsets of \emph{variables} instead of subsets of observations.
+At first sight,
+this differs from our use of the term in that we restrict to subsets of \emph{variables} instead of subsets of observations.
+However, if we flip perspectives and think of each genetic variant as an observation,
+our method shares common threads, although the ultimate goals and methods for visualization are different.
 Future methods for visualization of genomic data

### 2016-05-04 medicago/local_pca_on_medicago.Rmd
small example windows in Rmd file

# Local PCA on *Medicago truncatula* Hapmap

# Data

Filtered SNP calls from Mt4.0 were downloaded from [medicagohapmap.org](http://medicagohapmap.org/) on 5/4/2016:
```
mkdir -p data; cd data
wget http://www.medicagohapmap.org/downloads/Mt40/Mt4.0_HapMap_README.pdf
for CHROM in 1 2 3 4 5 6 7 8 u;
do
    echo "Getting $CHROM"
    wget http://www.medicagohapmap.org/downloads/Mt40/snps_by_chr/chr${CHROM}-filtered-set-2014Apr15.bcf
    wget http://www.medicagohapmap.org/downloads/Mt40/snps_by_chr/chr${CHROM}-filtered-set-2014Apr15.csi
done
```
For some reason these .csi index files don't work; so we need to remake them:
```
for CHROM in 2 3 4 5 6 7 8 u;
do
    bcftools index -f chr${CHROM}-filtered-set-2014Apr15.bcf
done
```

## Samples

The file of sample info was obtained from the HTML table at [http://www.medicagohapmap.org/hapmap/germplasm](http://www.medicagohapmap.org/hapmap/germplasm).
To check this contains information about all samples,
first get the list of samples in the .bcf files:
```
bcftools query -l chr1-filtered-set-2014Apr15.bcf > samples_in_bcf.txt
```
then compare to the table of information:
```{r sample_info}
samples <- read.table("sample_info.tsv", sep="\t", header=TRUE)
bcf.samples <- scan("samples_in_bcf.txt", what='char')
setdiff( bcf.samples , samples$ID )
```
For some reason, the bcf files refer to `HM020-I` while in the the sample info file there is `HM020`.
We'll assume they are the same (but keep an eye on that one!).
Here's where they're all from:
```{r show_sample_info}
bcf.samples <- gsub("-I","",bcf.samples)
samples <- droplevels(subset(samples, ID %in% bcf.samples))
sort( table(samples$Country.of.Origin) )
```

## Recoding

Recall that VCF files have the following columns:

> CHROM  POS     ID      REF     ALT     QUAL    FILTER  INFO    FORMAT

We use [bcftools](https://github.com/samtools/bcftools) to extract information, for instance:
```
# list of sites
bcftools query -f '%CHROM\t%POS\n' -r chr1:1-400 chr1-filtered-set-2014Apr15.bcf
# genotypes
bcftools query -f '%CHROM\t%POS[\t%GT]\n' -r chr1:1-400 chr1-filtered-set-2014Apr15.bcf
# genotypes of first four individuals, numerically
bcftools query -f '%CHROM\t%POS[\t%GT]\n' -r chr1:1-400 chr1-filtered-set-2014Apr15.bcf -s 'HM001,HM002,HM003,HM004' | sed -e 's_0/0_0_g' -e 's_\(0/1\)\|\(1/0\)_1_g' -e 's_1/1_2_g'
```

To get windows out of the VCF file and recode them, we'll just pipe a call to `bcftools`.
Here's a dummy version that will pull out windows of 100 SNPs for only the first four individuals.
```{r win_fn}
get.indivs <- "'HM001,HM002,HM003,HM004'"
bcf.file <- "chr1-filtered-set-2014Apr15.bcf"
# first read in the info about sites
bcf.con <- pipe(paste("bcftools query -f '%CHROM\\t%POS\\n' -r chr1:1-100000",bcf.file),open="r")
bcf.sites <- read.table(bcf.con, sep='\t')
colnames(bcf.sites) <- c("chrom","pos")
close(bcf.con)
# now use bcftools to pull out sites we want
#' This gives a string of the form "chr:start-end,chr:start-end" as required by bcftools
get_regions <- function (sites) {
    if (nrow(sites)==0) { return( NULL ) }
    pos <- as.numeric( sites[,2] )
    do.call( paste, c( tapply( 1:nrow(sites), sites[,1], function (kk) {
                            paste0( sites[kk[1],1], ":", min(pos[kk]), "-", max(pos[kk]) )
                        } ), list(sep=",") ) )
}
#' This will pull out bits of the bcf file corresponding to a regions string and convert it to numeric
get_sites <- function (regions) {
    gt.text <- data.table::fread( paste( c("bcftools", "query", "-f", "'[ %GT]\\n'", "-r", regions, "-s", get.indivs, bcf.file), collapse=" " ), header=FALSE, sep=' ', data.table=FALSE )
    gt <- c(0,1,1,2)[match( unlist(gt.text), c("0/0","0/1","1/0","1/1") )]
    dim(gt) <- dim(gt.text)
    return(gt)
}
win.fn <- function (n,win.snps=100) {
    sites <- bcf.sites[seq(1+(n-1)*win.snps,length.out=win.snps),]
    regions <- get_regions(sites)
    get_sites(regions)
}
win.fn(20,10)  # get the 20th window of 10 SNPs
```

Here's an alternate version that pulls out windows based on basepairs.
Note that this version assumes we have only one chromosome, starting at zero.
```{r win_fn_bp}
win.fn.bp <- function (n,win.bp=200) {
    win.start <- 1+(n-1)*win.bp
    win.end <- win.start + win.bp-1
    regions <- get_regions( cbind( chrom="chr1", pos=c(win.start,win.end) ) )
    get_sites(regions)
}
win.fn.bp(20,50)
```

## Computing eigenvectors of covariance matrices

Now we are set up to use `lostruct::eigen_windows()` to compute and write out the eigenstructure for windows along the genome.

Here's how to do that for the first 4 windows of length 100 SNPs, with the simplified version: just four individuals.
```{r simple_snp_pcas}
eigen_windows( 
        data=bcf.file,
        win.fn=function(data,n,...){ win.fn(n,...) },
        do.windows=1:4,
        k=2,
        win.snps=100
    )
```
Here's the same thing on windows of length 1000bps:
```{r simple_bp_pcas}
eigen_windows( 
        data=bcf.file,
        win.fn=function(data,n,...){ win.fn.bp(n,...) },
        do.windows=1:4,
        k=2,
        win.bp=1000
    )
```

### 2016-05-04 package/R/eigen_windows.R

@@ -6,26 +6,34 @@
 #' and returns the given number of eigenvector/eigenvalue pairs.
 #' Omits the last (short) window.
 #'
-#' @param data A numeric matrix.
-#' @param win Number of rows to take as one window.
+#' @param data The object containing the genotype data, e.g., a numeric matrix.
+#' @param win.fn Function that takes \code{data} and a positive integer \code{n} and returns the numeric matrix that correspond to the \code{n}th window.
+#' @param do.windows Vector of integers (or, arguments to \code{win.fn}) that correspond to the windows used.
+#' @param win If \code{win.fn} is not provided, each contiguous block of \code{win} rows of the matrix \code{data} will be one window.
 #' @param k Number of eigenvalue/eigenvector pairs to return.
 #' @param mc.cores If this is greater than 1, parallel::mclapply will be used.
+#' @param ... Other parameters to be passed to \code{win.fn}.
 #' @return A named list, with components
 #'   $values : A numeric matrix of eigenvalues, one column for each window and k rows.
 #'   $vectors : A numeric matrix, one column for each windows, having the eigenvectors in order (so it has k * ncol(matrix) rows).
 #' @export
-eigen_windows <- function ( data, win, k, mc.cores=1 ) {
+eigen_windows <- function ( 
+            data, 
+            win.fn=function(data,n,...){ data[seq( ((n-1)*win+1), (n*win) ),] }, 
+            do.windows=1:floor(nrow(data)/win),
+            win, k, mc.cores=1, ... ) {
+    if (missing(win) && missing(win.fn)) { stop("Must supply either win or win.fn.") }
     this.lapply <- if (mc.cores>1) { function (...) parallel::mclapply(...,mc.cores=mc.cores) } else { lapply }
-    data <- sweep( data, 1, rowMeans(data,na.rm=TRUE) )
     .local <- function(x) {
-        chunk <- data[((x-1)*win + 1):(x*win), ]
+        chunk <- win.fn(data,n=x,...)
+        chunk <- sweep( chunk, 1, rowMeans(chunk,na.rm=TRUE) )
         cov <- cov(chunk,use="pairwise")
         if(any(is.na(cov))) {return(rep(NA,2*(nrow(cov)+1)))}
         PCA <- eigen(cov)
         # returns in order (values, vectors)
         return( c( PCA$values[1:k], PCA$vectors[,1:k] ) )
     }
[…]

### 2016-05-04 package/R/read_data.R

@@ -9,7 +9,7 @@
 #'
 #' @param x Character matrix of alleles.
 #' @param ploidy Number of columns per individual.
-#' @param triallelic Include triallelic sites? Defaults to TRUE.
+#' @param triallelic Include triallelic sites? Defaults to TRUE; otherwise sets these to NA.
 #' @return An integer matrix with one row per site and one column per individual.
 #' @export
 recode_numeric <- function (x, ploidy=2, triallelic=TRUE) {
@@ -56,7 +56,7 @@ recode_numeric <- function (x, ploidy=2, triallelic=TRUE) {
 #'
 #' @param file Input .tped filename.
 #' @param chrom Vector of chromosome name(s) to include (defaults to all).
-#' @param triallelic Include triallelic sites? Defaults to TRUE.
+#' @param triallelic Include triallelic sites? Defaults to TRUE; otherwise sets these to NA.
 #' @param phased Are the data phased? (i.e., should return one column per individual or two?)
 #' @return An integer matrix with one row per site and one (or two, if phased) column per individual.
 #' @export

### 2016-05-05 medicago/local_pca_on_medicago.Rmd
rewrote eigen_windows to take functions

@@
 # Local PCA on *Medicago truncatula* Hapmap
+
+```{r setup}
+devtools::load_all("../package")
+```
 
@@
 ```{r sample_info}
-samples <- read.table("sample_info.tsv", sep="\t", header=TRUE)
-bcf.samples <- scan("samples_in_bcf.txt", what='char')
+samples <- read.table("data/sample_info.tsv", sep="\t", header=TRUE)
+bcf.samples <- scan("data/samples_in_bcf.txt", what='char')
 setdiff( bcf.samples , samples$ID )
@@
 ```{r win_fn}
-get.indivs <- "'HM001,HM002,HM003,HM004'"
-bcf.file <- "chr1-filtered-set-2014Apr15.bcf"
+get.indivs <- c("HM001","HM002","HM003","HM004")
+bcf.file <- "data/chr1-filtered-set-2014Apr15.bcf"
 # first read in the info about sites
-bcf.con <- pipe(paste("bcftools query -f '%CHROM\\t%POS\\n' -r chr1:1-100000",bcf.file),open="r")
-bcf.sites <- read.table(bcf.con, sep='\t')
-colnames(bcf.sites) <- c("chrom","pos")
+bcf.con <- pipe(paste("bcftools query -f '%POS\\n' -r chr1:1-100000",bcf.file),open="r")
+bcf.sites <- list( chr1=scan(bcf.con) )
 close(bcf.con)
 # now use bcftools to pull out sites we want
-#' This gives a string of the form "chr:start-end,chr:start-end" as required by bcftools
-get_regions <- function (sites) {
-    if (nrow(sites)==0) { return( NULL ) }
-    pos <- as.numeric( sites[,2] )
-    do.call( paste, c( tapply( 1:nrow(sites), sites[,1], function (kk) {
-                            paste0( sites[kk[1],1], ":", min(pos[kk]), "-", max(pos[kk]) )
-                        } ), list(sep=",") ) )
-}
-#' This will pull out bits of the bcf file corresponding to a regions string and convert it to numeric
-get_sites <- function (regions) {
-    gt.text <- data.table::fread( paste( c("bcftools", "query", "-f", "'[ %GT]\\n'", "-r", regions, "-s", get.indivs, bcf.file), collapse=" " ), header=FALSE, sep=' ', data.table=FALSE )
-    gt <- c(0,1,1,2)[match( unlist(gt.text), c("0/0","0/1","1/0","1/1") )]
-    dim(gt) <- dim(gt.text)
-    return(gt)
-}
 win.fn <- function (n,win.snps=100) {
-    sites <- bcf.sites[seq(1+(n-1)*win.snps,length.out=win.snps),]
-    regions <- get_regions(sites)
-    get_sites(regions)
+    sites <- list( chr1=bcf.sites[[1]][seq(1+(n-1)*win.snps,length.out=win.snps)] )
+    regions <- region_string(sites)
+    query_genotypes( bcf.file, regions, get.indivs )
 }
@@
     win.end <- win.start + win.bp-1
-    regions <- get_regions( cbind( chrom="chr1", pos=c(win.start,win.end) ) )
-    get_sites(regions)
+    regions <- paste0( "chr1", ":", win.start, ",", win.end )
+    query_genotypes( bcf.file, regions, get.indivs )
 }
@@
 ```
+
+## Doing this for everything
+
+Now let's see how it goes to do this for an entire chromosome.
+On chromosome 1 there are about 5 million SNPs, so we'll do it in windows of 5,000.
+First, windows by SNPs.
+```{r lpca_chr1}
+sites <- vcf_positions(bcf.file)
+win.fn.snp <- vcf_windower(bcf.file, size=5e3, type="snp", sites=sites, samples=get.indivs) 
+```

### 2016-05-05 package/R/eigen_windows.R

@@ -6,11 +6,10 @@
 #' and returns the given number of eigenvector/eigenvalue pairs.
 #' Omits the last (short) window.
 #'
-#' @param data The object containing the genotype data, e.g., a numeric matrix.
-#' @param win.fn Function that takes \code{data} and a positive integer \code{n} and returns the numeric matrix that correspond to the \code{n}th window.
-#' @param do.windows Vector of integers (or, arguments to \code{win.fn}) that correspond to the windows used.
-#' @param win If \code{win.fn} is not provided, each contiguous block of \code{win} rows of the matrix \code{data} will be one window.
+#' @param data Either a numeric matrix or a window extractor function, that takes a positive integer \code{n} and returns the numeric matrix that correspond to the \code{n}th window.
 #' @param k Number of eigenvalue/eigenvector pairs to return.
+#' @param do.windows Vector of integers that correspond to the windows used.
+#' @param win If \code{data} is a matrix, each contiguous block of \code{win} rows of the matrix \code{data} will be one window.
 #' @param mc.cores If this is greater than 1, parallel::mclapply will be used.
 #' @param ... Other parameters to be passed to \code{win.fn}.
 #' @return A named list, with components
@@ -18,14 +17,41 @@
 #'   $vectors : A numeric matrix, one column for each windows, having the eigenvectors in order (so it has k * ncol(matrix) rows).
 #' @export
 eigen_windows <- function ( 
-            data, 
-            win.fn=function(data,n,...){ data[seq( ((n-1)*win+1), (n*win) ),] }, 
-            do.windows=1:floor(nrow(data)/win),
-            win, k, mc.cores=1, ... ) {
-    if (missing(win) && missing(win.fn)) { stop("Must supply either win or win.fn.") }
+            data, k,
+            do.windows=NULL,
+            win, mc.cores=1, ... ) {
     this.lapply <- if (mc.cores>1) { function (...) parallel::mclapply(...,mc.cores=mc.cores) } else { lapply }
+    eigen.mat <- if (inherits(data,"function")) {
+                eigen_windows_winfn( win.fn=data, do.windows=do.windows, k=k, mc.cores=mc.cores, ... )
+            } else {
[…]

### 2016-05-05 medicago/local_pca_on_medicago.Rmd
.

@@
 win.fn.snp <- vcf_windower(bcf.file, size=5e3, type="snp", sites=sites, samples=get.indivs) 
+snp.pca <- eigen_windows(win.fn.snp,k=2)
 ```

### 2016-05-05 medicago/local_pca_on_medicago.Rmd
functions for querying vcf and finding corners

--- 2016-05-05 medicago/local_pca_on_medicago.Rmd
@@
 devtools::load_all("../package")
+library(Matrix)
+library(colorspace)
 ```
@@
 ```{r sample_info}
-samples <- read.table("data/sample_info.tsv", sep="\t", header=TRUE)
+samples <- read.table("data/sample_info.tsv", sep="\t", header=TRUE,stringsAsFactors=FALSE)
 bcf.samples <- scan("data/samples_in_bcf.txt", what='char')
@@
 ```{r show_sample_info}
-bcf.samples <- gsub("-I","",bcf.samples)
+wierd.names <- grep("-I",bcf.samples,value=TRUE)
+samples$ID[ match(gsub("-I","",wierd.names),wierd.names) ] <- wierd.names
 samples <- droplevels(subset(samples, ID %in% bcf.samples))
@@
     sites <- list( chr1=bcf.sites[[1]][seq(1+(n-1)*win.snps,length.out=win.snps)] )
-    regions <- region_string(sites)
+    regions <- enclosing_region(sites)
     query_genotypes( bcf.file, regions, get.indivs )
@@
     win.end <- win.start + win.bp-1
-    regions <- paste0( "chr1", ":", win.start, ",", win.end )
+    regions <- data.frame( chrom="chr1", start=win.start, end=win.end )
     query_genotypes( bcf.file, regions, get.indivs )
@@
 eigen_windows( 
-        data=bcf.file,
-        win.fn=function(data,n,...){ win.fn(n,...) },
+        data=win.fn,
         do.windows=1:4,
@@
 eigen_windows( 
-        data=bcf.file,
-        win.fn=function(data,n,...){ win.fn.bp(n,...) },
+        data=win.fn.bp,
         do.windows=1:4,
@@
 First, windows by SNPs.
-```{r lpca_chr1}
+```{r lpca_chr1_snp, cache=TRUE}
 sites <- vcf_positions(bcf.file)
-win.fn.snp <- vcf_windower(bcf.file, size=5e3, type="snp", sites=sites, samples=get.indivs) 
+do.indivs <- sample( samples$ID, 50 )
+win.fn.snp <- vcf_windower(bcf.file, size=5e3, type="snp", sites=sites, samples=do.indivs) 
+system.time( snp.pca <- eigen_windows(win.fn.snp,k=2) )
 ```
+It takes about 2 minutes to do chromosome 1, for ten individuals, on phoebe.
+Next, we compute the distance between windows with `pc_dist()` and do MDS visualization on the result.
+This is just with ten samples, so we don't expect a real result, but let's check it all works:
+```{r mds_chr1_snp, cache=TRUE, depends="lpca_chr1_snp"}
+system.time( chr1.pcdist <- pc_dist( snp.pca ) )
+chr1.mds <- cmdscale( chr1.pcdist, eig=TRUE, k=4 )
+colnames(chr1.mds$points) <- paste("MDS coordinate", 1:ncol(chr1.mds$points))
+```
+
+## Visualization
+
+Here's the results:
+```{r show_results_snp}
+# the distance matrix itself
+image( Matrix( chr1.pcdist ) )
+# the MDS results
+win.regions <- region(win.fn.snp)()
+win.mids <- (win.regions$start+win.regions$end)/2
+chrom.cols <- rainbow_hcl(128, c=90, end=.9*360)[as.numeric(cut(win.mids,128))]
+pairs( chr1.mds$points, pch=20, col=adjustcolor(chrom.cols,0.75) )
+```
+
+Now let's extract the extreme windows in the MDS plot:
+```{r get_corners}
+mincirc <- enclosing_circle( chr1.mds$points[,1:2] )
+corners <- corners( chr1.mds$points[,1:2], prop=.05 )
+corner.cols <- c("red","blue","purple")
+ccols <- rep("black",nrow(chr1.mds$points))
+for (k in 1:ncol(corners)) {
+    ccols[ corners[,k] ] <- corner.cols[k]
+}
+plot( chr1.mds$points[,1:2], pch=20, col=adjustcolor(ccols,0.75), 
+        xlab="MDS coordinate 1", ylab="MDS coordinate 2", 
+        xlim=mincirc$ctr[1]+c(-1,1)*mincirc$rad,
+        ylim=mincirc$ctr[2]+c(-1,1)*mincirc$rad )
+plot_circle( mincirc, col='red' )
+points( mincirc$three, col='red', cex=2 )
+```
+
+... and look at where these lie along the chromosomes
+```{r show_corners}
+for (k in 1:ncol(chr1.mds$points)) {
+    plot( win.mids/1e6, chr1.mds$points[,k], pch=20, 
+            xlab="Position (Mb)", ylab=paste("MDS coordinate",k),
+            col=adjustcolor(ccols,0.75) )
+}
+```

### 2016-05-05 package/R/corners.R

@@ -1,0 +1,19 @@
+#' Find Points in a 2D Cloud Closest to the "Corners"
+#'
+#' Finds the three "extreme points" that lie on the minimal enclosing circle,
+#' and returns a list of length three containing the portion of points 
+#' that lie closest to those extreme points.
+#'
+#' @param xy A two-column numeric matrix of coordinates.
+#' @param prop The proportion of points to return for each corner.
+#' @return A three column integer matrix giving the indices of the corresponding points.
+#' @export
+corners <- function (xy, prop) {
+    mincirc <- enclosing_circle( xy )
+    cidx <- mincirc$index
+    dists <- sqrt( outer( xy[,1], xy[cidx,1], "-" )^2 + outer( xy[,2], xy[cidx,2], "-" )^2 )
+    return( apply( dists, 2, function (u) {
+                which( u <= quantile(u,probs=prop) )
+            } ) )
+}
+

### 2016-05-05 package/R/eigen_windows.R

@@ -35,7 +35,7 @@ eigen_windows <- function (
 
 #' Sets up eigen_windows_winfn to work on a matrix.
 eigen_windows_matrix <- function (
-            data, k, win
+            data, k, win,
             do.windows=NULL,
             mc.cores ) {
     win.fn <- function(n,...){ data[seq( ((n-1)*win+1), (n*win) ),] }

### 2016-05-05 package/R/enclosing_circle.R

@@ -1,0 +1,153 @@
+#' Minimum Enclosing Circle
+#'
+#' Finds the minimum enclosing circle to a set of points,
+#' "Skyum algorithm based on the convex hull", as implemented at 
+#'    http://www.uni-kiel.de/psychologie/rexrepos/posts/diagBounding.html#minimum-enclosing-circle
+#'
+#' @param xy The 2D coordinates.
+#' @param circ A circle as returned by \code{enclosing_circle}.
+#' @param n The number of angles to draw the circle at.
+#' @param plot.it Actually do plotting or just return points?
+#' @return A named list with entries \code{ctr} (the coordinates of the center), \code{rad} (the radius),
+#' \code{three} (the three points in xy that define the circle),
+#' and \code{index} (the indices of those three points in xy).
+enclosing_circle <- function (xy) { getMinCircle(xy) }
+
+#' @documentIn enclosing_circle Plots a circle.
+plot_circle <- function (circ,n=200, plot.it=TRUE, ...) {
+    angles <- seq(0, 2*pi, length.out=n)
+    circ.lines <- cbind(circ$ctr[1] + circ$rad*cos(angles),
+                    circ$ctr[2] + circ$rad*sin(angles))
+    if (plot.it) lines(circ.lines, ... )
+    return(invisible(circ.lines))
+}
+
+#' Circle defined by three points
+getCircleFrom3 <- function(xy) {
+    stopifnot(is.matrix(xy), is.numeric(xy), nrow(xy) == 3, ncol(xy) == 2)
+
+    aa <- xy[1,  ]
+    bb <- xy[2,  ]
+    cc <- xy[3,  ]
+    y  <- xy[ , 2]
+
+    xDeltaA <- bb[1] - aa[1]
+    yDeltaA <- bb[2] - aa[2]
+    xDeltaB <- cc[1] - bb[1]
+    yDeltaB <- cc[2] - bb[2]
+    xDeltaC <- cc[1] - aa[1]
+    yDeltaC <- cc[2] - aa[2]
+
+    ## check if the points are collinear: qr(xy)$rank == 1, or:
+    ## determinant of difference matrix = 0, no need to use det()
+    dMat <- rbind(c(xDeltaA, yDeltaA), c(xDeltaB, yDeltaB))
+    if(isTRUE(all.equal(dMat[1,1]*dMat[2,2] - dMat[1,2]*dMat[2,1], 0, check.attributes=FALSE))) {
+        ## define the circle as the one that's centered between the points
[…]

### 2016-05-05 package/R/pc_dist.R

@@ -8,15 +8,14 @@
 #' \code{values} is replaced by \code{values/sqrt(sum(values^2))}.
 #'
 #' @param x List as output by eigen_windows().
-#' @param normalize Normalize the matrices to have the same norm?
+#' @param normalize Normalize the matrices to have the same norm?  Can be "L1" (default), "L2", or FALSE.
 #' @param mc.cores If this is greater than 1, parallel::mclapply will be used.
 #' @return A symmetric, numeric matrix with number of columns equal to the number of columns in eigen.win$values.
 #' @export
-pc_dist <- function( x, normalize=TRUE, mc.cores=1 ) {
+pc_dist <- function( x, normalize="L1", mc.cores=1 ) {
     this.lapply <- if (mc.cores>1) { function (...) parallel::mclapply(...,mc.cores=mc.cores) } else { lapply }
-    if (normalize) {
-        x$values <- sweep( x$values, 2, sqrt(colSums(x$values^2)), "/" )
-    }
+    if (normalize=="L1") { x$values <- sweep( x$values, 2, abs(colSums(x$values)), "/" ) }
+    if (normalize=="L2") { x$values <- sweep( x$values, 2, sqrt(colSums(x$values^2)), "/" ) }
     n <- ncol(x$values)
     k <- nrow(x$values)
     # should be symmetric, oh well.

### 2016-05-05 package/R/query_vcf.R

@@ -1,0 +1,159 @@
+#' Find Minimal Enclosing Regions
+#'
+#' Given a named list of positions, whose names correspond to chromosomes,
+#' return the data frame of regions with names chrom,start,end that encloses them.
+#'
+#' @param sites Named list of positions, as output by \code{vcf_positions()}.
+#' @return A data frame with variables chrom, start, and end.
+#' @export
+enclosing_region <- function (sites) {
+    if (length(sites)==0 || max(sapply(sites,length))==0) { return( NULL ) }
+    return( data.frame(
+            chrom=names(sites),
+            start=sapply(sites,min,na.rm=TRUE),
+            end=sapply(sites,max,na.rm=TRUE)
+        ) )
+}
+
+#' Make Regions String from Data Frame of Regions
+#'
+#' Given a data frame with variables \code{chrom}, \code{start}, and \code{end},
+#' make a regions string that can be passed to bcftools.
+#'
+#' @param regions A data frame with \code{chrom}, \code{start}, and \code{end}.
+#' @return A character string of the form "chr1:start1-end1,chr2:start2,end2".
+#' @export
+region_string <- function (regions) {
+    if (nrow(regions)==0) { return( NULL ) }
+    paste( paste0( regions$chrom, ":", as.numeric(regions$start), "-", as.numeric(regions$end) ), collapse="," )
+}
+
+
+#' Query a VCF File For Numeric Genotypes
+#'
+#' Uses bcftools to query vcf or bcf files for genotypes in a collection of regions,
+#' returning the genotypes in numeric format.
+#' Assumes the data are diploid.
+#'
+#' @param file The name of the indexed vcf file.
+#' @param regions A data frame containing chrom, start, and end of the regions to be extracted.
+#' @param samples A character vector of sample names to be extracted.
+#' @param verbose Whether to output with \code{cat()} the bcftools calls.
+#' @return An integer matrix, giving the number of alternate alleles for each sample.
+#' @export
+query_genotypes <- function (file,regions,samples, verbose=FALSE) {
+	bcf.args <- c("bcftools", "query", "-f", "'[ %GT]\\n'", "-r", region_string(regions))
[…]

### 2016-05-06 medicago/medicago_data_setup.Rmd
Rscript for medicago

--- 2016-05-05 medicago/local_pca_on_medicago.Rmd
@@
 wierd.names <- grep("-I",bcf.samples,value=TRUE)
-samples$ID[ match(gsub("-I","",wierd.names),wierd.names) ] <- wierd.names
+samples$ID[ match(gsub("-I","",wierd.names),samples$ID) ] <- wierd.names
 samples <- droplevels(subset(samples, ID %in% bcf.samples))
@@
 system.time( chr1.pcdist <- pc_dist( snp.pca ) )
-chr1.mds <- cmdscale( chr1.pcdist, eig=TRUE, k=4 )
-colnames(chr1.mds$points) <- paste("MDS coordinate", 1:ncol(chr1.mds$points))
+# there may be windows with missing data
+na.inds <- is.na( snp.pca$values[1,] )
+chr1.mds <- cmdscale( chr1.pcdist[!na.inds,!na.inds], eig=TRUE, k=4 )
+mds.coords <- chr1.mds$points[ ifelse( na.inds, NA, cumsum(!na.inds) ), ]
+colnames(mds.coords) <- paste("MDS coordinate", 1:ncol(mds.coords))
 ```
@@
 chrom.cols <- rainbow_hcl(128, c=90, end=.9*360)[as.numeric(cut(win.mids,128))]
-pairs( chr1.mds$points, pch=20, col=adjustcolor(chrom.cols,0.75) )
+pairs( mds.coords, pch=20, col=adjustcolor(chrom.cols,0.75) )
 ```
@@
 ```{r get_corners}
-mincirc <- enclosing_circle( chr1.mds$points[,1:2] )
-corners <- corners( chr1.mds$points[,1:2], prop=.05 )
+mincirc <- enclosing_circle( mds.coords[,1:2] )
+mds.corners <- corners( mds.coords[,1:2], prop=.05 )
 corner.cols <- c("red","blue","purple")
-ccols <- rep("black",nrow(chr1.mds$points))
-for (k in 1:ncol(corners)) {
-    ccols[ corners[,k] ] <- corner.cols[k]
+ccols <- rep("black",nrow(mds.coords))
+for (k in 1:ncol(mds.corners)) {
+    ccols[ mds.corners[,k] ] <- corner.cols[k]
 }
-plot( chr1.mds$points[,1:2], pch=20, col=adjustcolor(ccols,0.75), 
+plot( mds.coords[,1:2], pch=20, col=adjustcolor(ccols,0.75), 
         xlab="MDS coordinate 1", ylab="MDS coordinate 2", 
@@
 points( mincirc$three, col='red', cex=2 )
+points( mds.coords[mincirc$index,], col='red', cex=1.5 )
 ```
 
-... and look at where these lie along the chromosomes
+Here is where these lie along the chromosomes:
 ```{r show_corners}
-for (k in 1:ncol(chr1.mds$points)) {
-    plot( win.mids/1e6, chr1.mds$points[,k], pch=20, 
+for (k in 1:ncol(mds.coords)) {
+    plot( win.mids/1e6, mds.coords[,k], pch=20, 
             xlab="Position (Mb)", ylab=paste("MDS coordinate",k),
@@
 ```
+
+And, here are the PCA plots corresponding to these corners:
+```{r pc_corners, cache=TRUE, depends="mds_chr1_snp"}
+corner.winfn <- function (n) { win.fn.snp( mds.corners[,n] ) }
+corner.pcs <- eigen_windows( data=corner.winfn, 
+                    k=2, do.windows=1:ncol(mds.corners) )
+```
+
+```{r plot_pc_corners}
+countries <- samples$Country.of.Origin[match(do.indivs,samples$ID)]
+country.names <- unique(countries)
+country.cols <- rainbow_hcl( length(country.names), c=90 )
+country.pch <- rep(1:6,length.out=length(country.names))
+for (k in 1:ncol(corner.pcs$vectors)) {
+    xy <- matrix( corner.pcs$vectors[,k], nrow=nrow(corner.pcs$vectors)/nrow(corner.pcs$values) )
+    plot(xy, xlab="PC 1", ylab="PC 2", main=paste("corner",k), 
+            col=country.cols[match(countries,country.names)], 
+            pch=country.pch[match(countries,country.names)]  )
+}
+legend('topright', col=country.cols, pch=country.pch, legend=country.names)
+```

### 2016-05-06 medicago/run_on_medicago.R

@@ -1,0 +1,61 @@
+#!/usr/bin/Rscript --vanilla
+library(optparse)
+
+invocation <- commandArgs()
+
+usage <- "\
+Does a local PCA analysis on the Medicago truncatula hapmap data (see medicago_data_setup.html for details).
+"
+
+option_list <- list(
+    # input/output
+        make_option( c("-t","--type"),   type="character",             help="Window by SNP or by bp?"),
+        make_option( c("-s","--size"),   type="integer",               help="Size of the window, in units of type."),
+        make_option( c("-k","--npca"),   type="integer",   default=2L, help="Number of principal components to compute for each window. [default: %default]"),
+        make_option( c("-m","--nmds"),   type="integer",   default=2L, help="Number of principal coordinates (MDS variables) to compute. [default: %default]"),
+        make_option( c("-o","--outdir"), type="character",             help="Directory to save results to.  [default: lostruct/results_%type_%size_%jobid/]"),
+        make_option( c("-j","--jobid"),  type="character", default=formatC(1e6*runif(1),width=6,format="d",flag="0"),   help="Unique job id. [default random]")
+    )
+opt <- parse_args(OptionParser(option_list=option_list,description=usage))
+if (is.null(opt$outdir)) { opt$outdir <- file.path("lostruct", sprintf( "results_%s_%d_%s", opt$type, opt$size, opt$jobid ) ) }
+print(opt) # this will go in the logfile
+Sys.time()
+
+dir.create( opt$outdir, showWarnings=FALSE, recursive=TRUE )
+
+# setup
+devtools::load_all("../package")
+
+# VCF files
+chroms <- paste0("chr",1:8)
+bcf.files <- file.path( "data", paste0(chroms,"-filtered-set-2014Apr15.bcf") )
+names(bcf.files) <- chroms
+
+all.pcas <- numeric(0)
+
+# local PCA, by chromosome
+for (bcf.file in bcf.files) {
+    pca.file <- file.path( opt$outdir, sprintf(gsub(".bcf",".pca.csv",bcf.file)) )
+    if (file.exists(pca.file)) { stop(paste("File",pca.file,"already exists! Not overwriting.")) }
+    cat("Finding PCs for", bcf.file, "and writing out to", pca.file, "\n")
[…]

### 2016-05-06 package/R/eigen_windows.R

@@ -11,15 +11,17 @@
 #' @param do.windows Vector of integers that correspond to the windows used.
 #' @param win If \code{data} is a matrix, each contiguous block of \code{win} rows of the matrix \code{data} will be one window.
 #' @param mc.cores If this is greater than 1, parallel::mclapply will be used.
+#' @param return.mat Return a single matrix instead of values and vectors split out?
 #' @param ... Other parameters to be passed to \code{win.fn}.
 #' @return A named list, with components
 #'   $values : A numeric matrix of eigenvalues, one column for each window and k rows.
-#'   $vectors : A numeric matrix, one column for each windows, having the eigenvectors in order (so it has k * ncol(matrix) rows).
+#'   $vectors : A numeric matrix, one column for each window, having the eigenvectors in order (so it has k * ncol(matrix) rows).
 #' @export
-eigen_windows <- function ( 
+eigen_windows <- function (
             data, k,
             do.windows=NULL,
-            win, mc.cores=1, ... ) {
+            win, mc.cores=1, 
+            return.mat=FALSE, ... ) {
     this.lapply <- if (mc.cores>1) { function (...) parallel::mclapply(...,mc.cores=mc.cores) } else { lapply }
     eigen.mat <- if (inherits(data,"function")) {
                 eigen_windows_winfn( win.fn=data, do.windows=do.windows, k=k, mc.cores=mc.cores, ... )
@@ -27,10 +29,15 @@ eigen_windows <- function (
                 if (missing(win)) { stop("Must supply either a function or a matrix and a value for win.") }
                 eigen_windows_matrix( data=data, win=win, do.windows=do.windows, k=k, mc.cores=mc.cores )
             }
-    return( list(
-            values = eigen.mat[1:k,],
-            vectors = eigen.mat[-(1:k),]
-        ) )
+    return( if (return.mat) { 
+               eigen.mat 
+           } else {
+               list(
+                    values = eigen.mat[1:k,],
+                    vectors = eigen.mat[-(1:k),]
+                ) 
+           }
+       )
 }
 
 #' Sets up eigen_windows_winfn to work on a matrix.

### 2016-05-06 medicago/medicago_data_setup.Rmd
added manpages

@@
 sort( table(samples$Country.of.Origin) )
+```
+
+## Statistics by chromosome
+
+We should know what the density of SNPs looks like, for choosing window sizes.
+Chromosome lengths are from the VCF files.
+```{r snp_density, cache=TRUE}
+chrom.lens <- c( chr1=52991155, chr2=45729672, chr3=55515152, chr4=56582383, chr5=43630510, chr6=35275713, chr7=49172423, chr8=45569985, chl_Mt=124033 )
+chrom.stats <- sapply( paste0("chr",1:8), function (chrom) {
+            bcf.file <- file.path("data",sprintf("%s-filtered-set-2014Apr15.bcf",chrom))
+            sites <- vcf_positions(bcf.file)
+            spacings <- diff(sites[[chrom]])
+            out <- c( nsnps = length(sites[[chrom]]),
+                    nbp = chrom.lens[chrom],
+                    spacing.mean = mean(spacings),
+                    spacing = quantile(spacings,.05),
+                    spacing = quantile(spacings,.25),
+                    spacing = quantile(spacings,.50),
+                    spacing = quantile(spacings,.75),
+                    spacing = quantile(spacings,.95),
+                    spacing.max = max(spacings,.95) )
+            return(out)
+        } )
+floor(chrom.stats)
 ```

### 2016-05-06 medicago/run_on_medicago.R

@@ -4,7 +4,17 @@ library(optparse)
 invocation <- commandArgs()
 
 usage <- "\
-Does a local PCA analysis on the Medicago truncatula hapmap data (see medicago_data_setup.html for details).
+\
+Does a local PCA analysis on the Medicago truncatula hapmap data (see medicago_data_setup.html for details). \
+To run an analysis on windows of 1000 SNPs, for instance: \
+    Rscript run_on_medicago.R -t snp -s 1000 \
+while to run it on windows of 1000 bps: \
+    Rscript run_on_medicago.R -t bp -s 1000 \
+\
+Do
+    Rscript run_on_medicago.R -h\
+for options.\
+\
 "
 
 option_list <- list(
@@ -13,11 +23,13 @@ option_list <- list(
         make_option( c("-s","--size"),   type="integer",               help="Size of the window, in units of type."),
         make_option( c("-k","--npca"),   type="integer",   default=2L, help="Number of principal components to compute for each window. [default: %default]"),
         make_option( c("-m","--nmds"),   type="integer",   default=2L, help="Number of principal coordinates (MDS variables) to compute. [default: %default]"),
-        make_option( c("-o","--outdir"), type="character",             help="Directory to save results to.  [default: lostruct/results_%type_%size_%jobid/]"),
+        make_option( c("-o","--outdir"), type="character",             help="Directory to save results to.  [default: lostruct/results_type_%type_size_%size_jobid_%jobid/]"),
         make_option( c("-j","--jobid"),  type="character", default=formatC(1e6*runif(1),width=6,format="d",flag="0"),   help="Unique job id. [default random]")
     )
 opt <- parse_args(OptionParser(option_list=option_list,description=usage))
-if (is.null(opt$outdir)) { opt$outdir <- file.path("lostruct", sprintf( "results_%s_%d_%s", opt$type, opt$size, opt$jobid ) ) }
+if (is.null(opt$outdir)) { opt$outdir <- file.path("lostruct", sprintf( "results_type_%s_size_%d_jobid_%s", opt$type, opt$size, opt$jobid ) ) }
+if (is.null(opt$type) || is.null(opt$size)) { stop(usage) }
+
 print(opt) # this will go in the logfile
 Sys.time()
 
@@ -35,7 +47,7 @@ all.pcas <- numeric(0)
 
 # local PCA, by chromosome
[…]

### 2016-05-07 medicago/medicago_data_setup.Rmd
.

@@
 library(colorspace)
+options(width=100)
 ```
@@
 ```
+
+So, thats a SNP every 10bp.
+Since we have `r length(bcf.samples)` samples,
+each local PC will take `r 1+length(bcf.samples)` doubles to record;
+a double is 8 bytes;
+so in, say 1Gb we can fit one local PC for `r floor(1e9 / (8*(1+length(bcf.samples))))` windows.
+With a total of `r sum(chrom.stats['nsnps',])` SNPs,
+that'd be `r floor(sum(chrom.stats['nsnps',])/floor(1e9 / (8*(1+length(bcf.samples)))))` SNPs per window.
+A bigger concern is the distance matrix we compute between windows
+(we don't have to save this to disk, though).
+With this many windows, that would be `r floor(1e9 / (8*(1+length(bcf.samples))))^2` floats,
+or `r floor(1e9 / (8*(1+length(bcf.samples))))^2 / 8e9` Gb.
+That's still possible, but barely.
 

### 2016-05-07 medicago/summarize_run.Rmd
.

---
title: "Local PCA results"
date: "`r date()`"
---

```{r setup, include=FALSE}
library(lostruct)
library(colorspace)
library(jsonlite)
opt <- fromJSON("config.json")

    pca.file <- file.path( opt$outdir, sprintf(gsub(".bcf",".pca.csv",basename(bcf.file))) )
    if (file.exists(pca.file)) { stop(paste("File",pca.file,"already exists! Not overwriting.")) }
    cat("Finding PCs for", bcf.file, "and writing out to", pca.file, "\n")
    win.fn <- vcf_windower(bcf.file, size=opt$size, type=tolower(opt$type) )

win.regions <- region(win.fn.snp)()
win.mids <- (win.regions$start+win.regions$end)/2
chrom.cols <- rainbow_hcl(128, c=90, end=.9*360)[as.numeric(cut(win.mids,128))]
pairs( mds.coords, pch=20, col=adjustcolor(chrom.cols,0.75) )

```

### 2016-05-07 medicago/run_on_medicago.R

@@ -30,8 +30,9 @@ opt <- parse_args(OptionParser(option_list=option_list,description=usage))
 if (is.null(opt$outdir)) { opt$outdir <- file.path("lostruct", sprintf( "results_type_%s_size_%d_jobid_%s", opt$type, opt$size, opt$jobid ) ) }
 if (is.null(opt$type) || is.null(opt$size)) { stop(usage) }
 
+opt$start.time <- Sys.time()
 print(opt) # this will go in the logfile
-Sys.time()
+cat( jsonlite::toJSON( opt, pretty=TRUE ), file=file.path( opt$outdir, "config.json" ) )
 
 dir.create( opt$outdir, showWarnings=FALSE, recursive=TRUE )
 

### 2016-05-11 medicago/medicago_data_setup.Rmd
run_on_medicago.R runs, saves regions and labeled MDS coords

@@
 options(width=100)
+fig.dim <- 4
+knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
+```
+
 ```
@@
     regions <- enclosing_region(sites)
-    query_genotypes( bcf.file, regions, get.indivs )
+    vcf_query( bcf.file, regions, get.indivs )
 }
@@
     regions <- data.frame( chrom="chr1", start=win.start, end=win.end )
-    query_genotypes( bcf.file, regions, get.indivs )
+    vcf_query( bcf.file, regions, get.indivs )
 }
@@
 ```{r get_corners}
-mincirc <- enclosing_circle( mds.coords[,1:2] )
+mincirc <- lostruct:::enclosing_circle( mds.coords[,1:2] )
 mds.corners <- corners( mds.coords[,1:2], prop=.05 )

### 2016-05-11 medicago/summarize_run.Rmd
run_on_medicago.R runs, saves regions and labeled MDS coords

@@
 library(jsonlite)
+fig.dim <- 4
+knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
+```
+
+```{r data_setup, include=FALSE}
 opt <- fromJSON("config.json")
 
-    pca.file <- file.path( opt$outdir, sprintf(gsub(".bcf",".pca.csv",basename(bcf.file))) )
-    if (file.exists(pca.file)) { stop(paste("File",pca.file,"already exists! Not overwriting.")) }
-    cat("Finding PCs for", bcf.file, "and writing out to", pca.file, "\n")
-    win.fn <- vcf_windower(bcf.file, size=opt$size, type=tolower(opt$type) )
+# data files, precomputed
+pca.files <- list.files(".",".*.pca.csv")
+mds.file <- "mds_coords.csv"
+regions.file <- gsub("[.]pca[.]csv$",".regions.csv",pca.files)
 
-win.regions <- region(win.fn.snp)()
-win.mids <- (win.regions$start+win.regions$end)/2
-chrom.cols <- rainbow_hcl(128, c=90, end=.9*360)[as.numeric(cut(win.mids,128))]
-pairs( mds.coords, pch=20, col=adjustcolor(chrom.cols,0.75) )
+# read in mds
+mds.coords <- read.csv(mds.file,header=TRUE)
+mds.cols <- (1:ncol(mds.coords))[-(1:2)]
 
+# position information
+regions <- do.call( rbind, lapply( regions.file, read.csv, header=TRUE ) )
+regions$mid <- (regions$start+regions$end)/2
+chrom.cols <- rainbow_hcl(128, c=90, end=.9*360)[as.numeric(cut(regions$mid,128))]
 ```
+
+This run had these parameters:
+
+- *window type* : `r opt$type`
+- *window size* : `r opt$size`
+- *number of pcs* : `r opt$npc`
+- *number of MDS coordinates* : `r opt$nmds`
+
+Here are the computed MDS coordinates, colored by chromosome:
+```{r mds_pairplot}
+pairs( mds.coords[,mds.cols], pch=20, col=adjustcolor(chrom.cols,0.75) )
+```
+
+Here are the extreme windows in the MDS plot:
+```{r get_corners}
+mds.corners <- corners( mds.coords[,mds.cols[1:2]], prop=.05 )
+corner.cols <- c("red","blue","purple")
+ccols <- rep("black",nrow(mds.coords))
+for (k in 1:ncol(mds.corners)) {
+    ccols[ mds.corners[,k] ] <- corner.cols[k]
+}
+
+plot( mds.coords[,mds.cols[1:2]], pch=20, col=adjustcolor(ccols,0.75), 
+        xlab="MDS coordinate 1", ylab="MDS coordinate 2" )
+
+for (k in mds.cols) {
+    plot( regions$mid/1e6, mds.coords[,k], pch=20, 
+            xlab="Position (Mb)", ylab=paste("MDS coordinate",k),
+            col=adjustcolor(ccols,0.75) )
+}
+```

### 2016-05-11 medicago/run_on_medicago.R

@@ -21,7 +21,7 @@ option_list <- list(
     # input/output
         make_option( c("-t","--type"),   type="character",             help="Window by SNP or by bp?"),
         make_option( c("-s","--size"),   type="integer",               help="Size of the window, in units of type."),
-        make_option( c("-k","--npca"),   type="integer",   default=2L, help="Number of principal components to compute for each window. [default: %default]"),
+        make_option( c("-k","--npc"),   type="integer",   default=2L, help="Number of principal components to compute for each window. [default: %default]"),
         make_option( c("-m","--nmds"),   type="integer",   default=2L, help="Number of principal coordinates (MDS variables) to compute. [default: %default]"),
         make_option( c("-o","--outdir"), type="character",             help="Directory to save results to.  [default: lostruct/results_type_%type_size_%size_jobid_%jobid/]"),
         make_option( c("-j","--jobid"),  type="character", default=formatC(1e6*runif(1),width=6,format="d",flag="0"),   help="Unique job id. [default random]")
@@ -31,10 +31,11 @@ if (is.null(opt$outdir)) { opt$outdir <- file.path("lostruct", sprintf( "results
 if (is.null(opt$type) || is.null(opt$size)) { stop(usage) }
 
 opt$start.time <- Sys.time()
+opt$run.dir <- normalizePath(".")
 print(opt) # this will go in the logfile
-cat( jsonlite::toJSON( opt, pretty=TRUE ), file=file.path( opt$outdir, "config.json" ) )
 
 dir.create( opt$outdir, showWarnings=FALSE, recursive=TRUE )
+cat( jsonlite::toJSON( opt, pretty=TRUE ), file=file.path( opt$outdir, "config.json" ) )
 
 # setup
 devtools::load_all("../package")
@@ -44,31 +45,48 @@ chroms <- paste0("chr",1:8)
 bcf.files <- file.path( "data", paste0(chroms,"-filtered-set-2014Apr15.bcf") )
 names(bcf.files) <- chroms
 
-all.pcas <- numeric(0)
+all.pcas <- numeric(0)       # will be a numeric matrix of eigen values/vectors
+all.lengths <- numeric(0)    # will be a numeric vector of numbers of windows per chromosome
[…]

### 2016-05-11 package/R/corners.R

@@ -12,6 +12,7 @@
 #' @return A three column integer matrix giving the indices of the corresponding points.
 #' @export
 corners <- function (xy, prop, k=3) {
+    xy <- as.matrix(xy)
     mincirc <- enclosing_circle( xy )
     cidx <- mincirc$index
     dpt <- function (uv) { sqrt( (xy[,1]-uv[1])^2 + (xy[,2]-uv[2])^2 ) }

### 2016-05-11 package/R/eigen_windows.R

@@ -11,17 +11,15 @@
 #' @param do.windows Vector of integers that correspond to the windows used.
 #' @param win If \code{data} is a matrix, each contiguous block of \code{win} rows of the matrix \code{data} will be one window.
 #' @param mc.cores If this is greater than 1, parallel::mclapply will be used.
-#' @param return.mat Return a single matrix instead of values and vectors split out?
 #' @param ... Other parameters to be passed to \code{win.fn}.
-#' @return A named list, with components
-#'   $values : A numeric matrix of eigenvalues, one column for each window and k rows.
-#'   $vectors : A numeric matrix, one column for each window, having the eigenvectors in order (so it has k * ncol(matrix) rows).
+#' @return A numeric matrix with one row for each window; the first k columns giving eigenvalues and each subsequent group of ncol(data) columns giving the corresponding eigenvector.
+#' The result has an attribute, "npc", which is equal to the number of eigenvalues computed (k).
 #' @export
 eigen_windows <- function (
             data, k,
             do.windows=NULL,
             win, mc.cores=1, 
-            return.mat=FALSE, ... ) {
+            ... ) {
     this.lapply <- if (mc.cores>1) { function (...) parallel::mclapply(...,mc.cores=mc.cores) } else { lapply }
     eigen.mat <- if (inherits(data,"function")) {
                 eigen_windows_winfn( win.fn=data, do.windows=do.windows, k=k, mc.cores=mc.cores, ... )
@@ -29,15 +27,16 @@ eigen_windows <- function (
                 if (missing(win)) { stop("Must supply either a function or a matrix and a value for win.") }
                 eigen_windows_matrix( data=data, win=win, do.windows=do.windows, k=k, mc.cores=mc.cores )
             }
-    return( if (return.mat) { 
-               eigen.mat 
-           } else {
-               list(
-                    values = eigen.mat[1:k,],
-                    vectors = eigen.mat[-(1:k),]
-                ) 
-           }
-       )
+    return( eigen.mat )
+}
+
+#' Sets up a winfun-like accessor to a matrix.
+winfun_matrix <- function (data,win) {
[…]

### 2016-05-11 package/R/enclosing_circle.R

@@ -19,7 +19,7 @@ enclosing_circle <- function (xy) {
     return(out)
 }
 
-#' @documentIn enclosing_circle Plots a circle.
+#' @describeIn enclosing_circle Plots a circle.
 plot_circle <- function (circ,n=200, plot.it=TRUE, ...) {
     angles <- seq(0, 2*pi, length.out=n)
     circ.lines <- cbind(circ$ctr[1] + circ$rad*cos(angles),

### 2016-05-11 package/R/pc_dist.R

@@ -1,28 +1,30 @@
 #' Find Matrix of Distances Between Matrices from Eigenvector/values
 #'
-#' Given a list of sets of eigenvalues/vectors for a set of n symmetric matrices as output by eigen_windows(),
+#' Given a matrix of sets of eigenvalues/vectors for a set of n symmetric matrices as output by eigen_windows(),
 #' return the (n x n) matrix whose [i,j]th element is the Frobenius norm of the difference between the i-th and the j-th matrices,
 #' as approximated by the given eigenvalues/vectors. The approximation is in pseudocode
 #'      M = values[1] * outer(vectors[1],vectors[1]) + values[2] * outer(vectors[2],vectors[2]) + ...
 #' If \code{normalize} is TRUE, then the the matrices are normalized to have norm 1, so that in the definition of M,
 #' \code{values} is replaced by \code{values/sqrt(sum(values^2))}.
 #'
-#' @param x List as output by eigen_windows().
+#' @param x Matrix as output by eigen_windows().
+#' @param npc Number of pcas computed.
 #' @param normalize Normalize the matrices to have the same norm?  Can be "L1" (default), "L2", or FALSE.
 #' @param mc.cores If this is greater than 1, parallel::mclapply will be used.
 #' @return A symmetric, numeric matrix with number of columns equal to the number of columns in eigen.win$values.
 #' @export
-pc_dist <- function( x, normalize="L1", mc.cores=1 ) {
+pc_dist <- function( x, npc, normalize="L1", mc.cores=1 ) {
     this.lapply <- if (mc.cores>1) { function (...) parallel::mclapply(...,mc.cores=mc.cores) } else { lapply }
-    if (normalize=="L1") { x$values <- sweep( x$values, 2, abs(colSums(x$values)), "/" ) }
-    if (normalize=="L2") { x$values <- sweep( x$values, 2, sqrt(colSums(x$values^2)), "/" ) }
-    n <- ncol(x$values)
-    k <- nrow(x$values)
-    # should be symmetric, oh well.
-    emat <- function (u) { matrix(u,ncol=k) }
+    values <- x[,1:npc,drop=FALSE]
+    vectors <- x[,-(1:npc)]
+    if (normalize=="L1") { values <- sweep( values, 1, abs(rowSums(values)), "/" ) }
[…]

### 2016-05-11 package/R/query_vcf.R

@@ -33,21 +33,24 @@ region_string <- function (regions) {
 #'
 #' Uses bcftools to query vcf or bcf files for genotypes in a collection of regions,
 #' returning the genotypes in numeric format.
-#' Assumes the data are diploid.
+#' Assumes the data are diploid and that every site is diallelic.
 #'
 #' @param file The name of the indexed vcf file.
 #' @param regions A data frame containing chrom, start, and end of the regions to be extracted.
 #' @param samples A character vector of sample names to be extracted.
 #' @param verbose Whether to output with \code{cat()} the bcftools calls.
+#' @param recode Whether to recode genotypes as numeric (assuming diallelic sites only).
 #' @return An integer matrix, giving the number of alternate alleles for each sample.
 #' @export
-query_genotypes <- function (file,regions,samples, verbose=FALSE) {
-	bcf.args <- c("bcftools", "query", "-f", "'[ %GT]\\n'", "-r", region_string(regions))
+vcf_query <- function (file, regions, samples, verbose=FALSE, recode=TRUE) {
+	bcf.args <- c("bcftools", "query", "-f", "'[ %GT]\\n'")
+	if (!missing(regions) && length(regions)>0) { bcf.args <- c( bcf.args, "-r", region_string(regions) ) }
 	if (!missing(samples) && length(samples)>0) { bcf.args <- c( bcf.args, "-s", paste(samples,collapse=',') ) }
 	bcf.args <- c( bcf.args, file )
     bcf.call <- paste(bcf.args, collapse=" ")
     if (verbose) cat(bcf.call, "\n")
     gt.text <- data.table::fread( bcf.call, header=FALSE, sep=' ', data.table=FALSE )
+    if (!recode) { return(gt.text) }
     gt <- c(0L,1L,1L,2L)[match( unlist(gt.text), c("0/0","0/1","1/0","1/1") )]
     dim(gt) <- dim(gt.text)
     return(gt)
@@ -70,6 +73,18 @@ vcf_positions <- function (file) {
 	return( tapply( bcf.sites$pos, bcf.sites$chrom, identity) )
 }
 
+#' Find Sample IDs from a VCF File
+#'
+#' Returns the character vector of samples (by calling \code{bcftools query -l}).
+#'
+#' @param file The name of the indexed vcf file.
+#' @return A character vector of sample IDs.
+#' @export
+vcf_samples <- function (file) {
+    return( system2( "bcftools", c("query","-l",file), stdout=TRUE ) )
+}
+
+
[…]

### 2016-05-11 package/R/read_data.R

@@ -10,14 +10,14 @@
 #' @param x Character matrix of alleles.
 #' @param ploidy Number of columns per individual.
 #' @param triallelic Include triallelic sites? Defaults to TRUE; otherwise sets these to NA.
+#' @param alleles A character vector of valid alleles.
 #' @return An integer matrix with one row per site and one column per individual.
 #' @export
-recode_numeric <- function (x, ploidy=2, triallelic=TRUE) {
-    bases <- c("A","C","G","T")
-    geno <- match(as.matrix(x),bases)
+recode_numeric <- function (x, ploidy=2, triallelic=TRUE, alleles=c("A","C","G","T")) {
+    geno <- match(as.matrix(x),alleles)
     dim(geno) <- c( nrow(x), ncol(x) )
     # totals[j,2] gives the total number of C's at the j-th site
-    totals <- sapply( seq_along(bases), function (k) { rowSums( geno==k, na.rm=TRUE ) } )
+    totals <- sapply( seq_along(alleles), function (k) { rowSums( geno==k, na.rm=TRUE ) } )
     if (!triallelic) {
         # remove triallelic sites
         which.triallelic <- ( rowSums(totals>0) > 2 )
@@ -25,7 +25,7 @@ recode_numeric <- function (x, ploidy=2, triallelic=TRUE) {
     }
     maxcounts <- do.call( pmax, lapply(1:ncol(totals),function(k)totals[,k]) )
     major <- rep(NA,nrow(geno))
-    for (k in seq_along(bases)) { major[ totals[,k] == maxcounts ] <- k }
+    for (k in seq_along(alleles)) { major[ totals[,k] == maxcounts ] <- k }
 
     # code the genotype matrix
     coded <- matrix( NA, nrow=nrow(geno), ncol(geno) )
@@ -76,8 +76,6 @@ read_tped <- function (file, chrom="[^ \t]*", triallelic=TRUE, phased=FALSE) {
 #' where "major" is the most common allele at that site, and all others are lumped as "minor".
 #' Assumes that anything not in "ACGT" is NA (case-sensitive).
 #'
-#' The work is done by \code{vcfR::extract.gt( )}; see there for details.
-#'
 #' @param file Input .vcf file, possibly gzip'ped.
[…]

### 2016-05-11 package/tests/testthat/test_pc_dist.R

@@ -11,12 +11,8 @@ y.eig <- eigen(y)
 
 elist <- function (k) {
     # transform to the format pc_dist wants
-    list(
-         values = cbind(x.eig$values[1:k],y.eig$values[1:k]),
-         vectors = cbind(
-                    as.vector( x.eig$vectors[,1:k] ),
-                    as.vector( y.eig$vectors[,1:k] )
-                 ) )
+    rbind( c(x.eig$values[1:k], as.vector( x.eig$vectors[,1:k] ) ),
+           c(y.eig$values[1:k], as.vector( y.eig$vectors[,1:k] ) ) )
 }
 
 f <- function (k,eig) {
@@ -36,9 +32,9 @@ for (k in kk) {
     # non-normalized
     ff <- sum( (f(k,x.eig) - f(k,y.eig))^2 )
     fm <- matrix(c(0,ff,ff,0),nrow=2)
-    expect_equal( fm, pc_dist(elist(k),normalize=FALSE) )
+    expect_equal( fm, pc_dist(elist(k),npc=k,normalize=FALSE) )
     # normalized
     gg <- sum( (g(k,x.eig) - g(k,y.eig))^2 )
     gm <- matrix(c(0,gg,gg,0),nrow=2)
-    expect_equal( gm, pc_dist(elist(k),normalize=TRUE) )
+    expect_equal( gm, pc_dist(elist(k),npc=k,normalize=TRUE) )
 }

### 2016-05-11 popres/popres_example.R

@@ -8,7 +8,7 @@ chr22 <- read_tped("POPRES_Genotypes_QC2_v2_TXT.tped.gz", chrom=22)
 eigenstuff <- eigen_windows(chr22, win=100, k=2)
 
 # as in POPRES_distance.R
-windist <- pc_dist( eigenstuff )
+windist <- pc_dist( eigenstuff, npc=2 )
 
 # as in POPRES_MDS.R
 fit2d <- cmdscale( windist, eig=TRUE, k=2 )

### 2016-05-12 medicago/summarize_run.Rmd
added corner plots to summarize_run

--- 2016-05-07 medicago/summarize_run.Rmd
@@
 library(jsonlite)
+library(RColorBrewer)
+fig.dim <- 4
+knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
+```
+
+```{r data_setup, include=FALSE}
 opt <- fromJSON("config.json")
 
-    pca.file <- file.path( opt$outdir, sprintf(gsub(".bcf",".pca.csv",basename(bcf.file))) )
-    if (file.exists(pca.file)) { stop(paste("File",pca.file,"already exists! Not overwriting.")) }
-    cat("Finding PCs for", bcf.file, "and writing out to", pca.file, "\n")
-    win.fn <- vcf_windower(bcf.file, size=opt$size, type=tolower(opt$type) )
+# data files, precomputed
+pca.files <- list.files(".",".*.pca.csv")
+mds.file <- "mds_coords.csv"
+regions.files <- list.files(".",".*.regions.csv")
 
-win.regions <- region(win.fn.snp)()
-win.mids <- (win.regions$start+win.regions$end)/2
-chrom.cols <- rainbow_hcl(128, c=90, end=.9*360)[as.numeric(cut(win.mids,128))]
-pairs( mds.coords, pch=20, col=adjustcolor(chrom.cols,0.75) )
+# read in mds
+mds.coords <- read.csv(mds.file,header=TRUE)
+mds.cols <- (1:ncol(mds.coords))[-(1:2)]
 
+# position information
+regions <- do.call( rbind, lapply( regions.files, read.csv, header=TRUE ) )
+# figure out where to plot things at
+chroms <- unique(regions$chrom)
+chrom.starts <- tapply( regions$start, regions$chrom, min, na.rm=TRUE )
+chrom.ends <- tapply( regions$end, regions$chrom, max, na.rm=TRUE )
+chrom.spacing <- floor(.05*mean(chrom.ends))
+chrom.offsets <- c(0,cumsum(chrom.spacing+chrom.ends))
+names(chrom.offsets) <- c(chroms,"end")
+chrom.dividers <- c(0,chrom.offsets[-1])-chrom.spacing/2
+chrom.mids <- chrom.dividers[-1] - diff(chrom.dividers)/2
+
+# this is where to plot windows at when plotting with all chromosomes
+regions$pos <- chrom.offsets[regions$chrom]+(regions$start+regions$end)/2
+
+chrom.cols <- rainbow_hcl(length(chroms), c=90, end=.9*360)[as.numeric(regions$chrom)]
+
+# set up to plot all chromosomes together
+chrom.plot <- function (y,ylab='',main='',...) {
+    plot(0, type='n', xlim=range(chrom.offsets/1e6), ylim=range(y,finite=TRUE), 
+         xlab='', xaxt='n', ylab=ylab, main=main)
+    for (k in 1:floor(length(chroms)/2)) {
+        rect( xleft=chrom.dividers[2*k-1]/1e6, xright=chrom.dividers[2*k]/1e6, 
+             ybottom=par("usr")[3], ytop=par("usr")[4], 
+             border=NA, col=adjustcolor("grey",0.25) )
+    }
+    abline( v=chrom.dividers/1e6, lty=3, col=adjustcolor("grey",0.5), lwd=2 )
+    axis( 1, at=chrom.mids/1e6, labels=chroms, las=0, tick=FALSE )
+    points( regions$pos/1e6, y, ...)
+}
 ```
+
+This run had these parameters:
+
+- *window type* : `r opt$type`
+- *window size* : `r opt$size`
+- *number of pcs* : `r opt$npc`
+- *number of MDS coordinates* : `r opt$nmds`
+
+Here are the number of windows per chromsome,
+and the computed MDS coordinates, colored by chromosome:
+```{r mds_pairplot}
+table(regions$chrom)
+pairs( mds.coords[,mds.cols], pch=20, col=adjustcolor(chrom.cols,0.75) )
+```
+
+Here are the extreme windows in the MDS plot:
+```{r get_corners}
+mds.corners <- corners( mds.coords[,mds.cols[1:2]], prop=.05 )
+# set up colors and pchs for corners
+corner.cols <- brewer.pal(3,"Dark2")
+corner.pch <- c(15,17,19)
+ccols <- rep("black",nrow(mds.coords))
+cpch <- rep(20,nrow(mds.coords))
+for (k in 1:ncol(mds.corners)) {
+    ccols[ mds.corners[,k] ] <- corner.cols[k]
+    cpch[ mds.corners[,k] ] <- corner.pch[k]
+}
+# plot corners and MDS along the chromosome
+layout(matrix(c(1,1,2,3),nrow=2),widths=c(1,2))
+plot( mds.coords[,mds.cols[1:2]], pch=cpch, 
+     col=adjustcolor(ccols,0.75), 
+     xlab="MDS coordinate 1", ylab="MDS coordinate 2" )
+for (k in mds.cols) {
+    chrom.plot( mds.coords[,k], pch=20, 
+            xlab="Position (Mb)", main=paste("MDS coordinate",match(k,mds.cols)),
+            ylab=colnames(mds.coords)[k],
+            col=adjustcolor(ccols,0.75) )
+}
+```

### 2016-05-12 writeup/local_pca_paper.tex
paper update

@@
 \plr{How about: first describe the method; then afterwards, present the three datasets.}
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{the-method-diagram}
+    \end{center}
+    \caption{
+         The diagram of our method.
+         \label{fig:diagram}
+    }
+\end{figure}
 
@@
 
-\plr{Separate paragraphs for each dataset. Say how many sites, what kind of data, etcetera.}
+For human, we use SNP chip data from POPRES \citep{nelson2008population}.
+There are 3,965 samples in total, (346 African-Americas; 73 Asians; 3,187 Europeans; 359 Indian Asians), and the 22 autosomes together have 447,267 SNPs in this dataset.
+We use the allele that has highest frequency in the samples as the reference allele for each position. 
+The first step is to recode genetic data as a numeric matrix; 
+which we take to have one row per variant and one column per sample. (omitting monomorphic variants)
+In our applications, for each variant we picked a reference allele, 
+and recoded genotypes as the number of non-reference alleles carried by the individual;
+obtaining 0,1, or 2 for autosomal variants in diploids, or NA for missing data.
+A normalization step (see below) ensures the result does not depend on the choice of reference allele.
 
-For human, we use SNP chip data from POPRES \citet{nelson2008population}.
-There are 3965 samples in total, (346 African-Americas; 73 Asians; 3187 Europeans; 359 Indian Asians), and the 22 autosomes together have 447267 SNPs in this dataset.
-We use the allele that has highest frequency in the samples as the reference allele for each position. 
-If an allele is same with the reference allele, we recode it as 0; if an allele is different from the reference allele, we recode it as 1; for positions that have missing data, we recode them as NA. 
-Since human genome is diploid, we add the value for the two alleles and then one chromosome will eventually be recode to a sequence consisting of 0,1,2 (and NA).
-Then the genome data for human is recoded to a matrix that has 3965 columns, each column for an individual's genotype. 
-We process the data separately for the 22 autosomes in human.
+For \textit{Drosophila melanogaster}, we use the sequencing data from Drosophila Population Genomics Project (DPGP) and John Pool's lab \citep{lack2015drosophila}, 
+which together has 380 samples from 16 countries across Africa and Europe.
+Each chromosome arm we investigated (Chr2L, Chr2R, Chr3L, Chr3R) has 2-3 million SNPs.
+Due to high density of missing data for some parts in the genome, we then delete the samples with more than 8\% missing genotypes and then deleting positions with more than 20\% missing data. 
+The cutoff points 8\% and 20\% are determined from the corresponding distributions of NAs in samples and at positions. 
+Since the \textit{Drosophila} samples are from inbred lines with little heterozygosity, we treat the samples as haploid when recoding.
 
-For \textit{Drosophila}, we use the sequencing data from DPGP and John pool's lab \citet{lack2015drosophila}, which together has 380 samples from 16 countries.
-We first process the sequencing data (from DPGP and John pool's lab) to SNP data by eliminating the positions that have all the same alleles. 
-Each chromosome arms we investigated (Chr2L, Chr2R, Chr3L, Chr3R, ChrX) has 2-3 million SNPs.
-Due to high density of missing data for some parts in the genome, we then delete the samples with more than 8\% NAs and positions with more than 20\%. 
-The cutoff points 8\% and 20\% are determined from the corresponding distributions of NAs in samples and at positions. 
-After we got the SNP data for \textit{Drosophila}, we recode it \textit{Drosophila}to matrices with 0,1,2 (and NA) for each chromosome arms (Chr2L, Chr2R, Chr3L, Chr3R, ChrX) similar to the process for human genome.
-Since the \textit{Drosophila} samples here are all homogenous, the matrices are indeed consisting of 0,1 (and NA).
+For \textit{Medicago truncatula}, we use the SNP data from whole genome sequencing from \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved}. 
+It has 263 samples from 24 countries. 
+Each of the 8 chromosomes has 3-5 million SNPs.
 
-For \textit{Medicago truncatula}, we use the SNP data from Medicago Hapmap. It has 263 samples from 24 countries. 
-We recode it to matrices with 0,1,2 (and NA) for each of the 8 chromosomes. Each chromosome has 3-5 million SNPs
+\begin{table}[ht]
+\centering
+\begin{tabular}{p{0.8in}p{1in}p{1in}p{1in}p{1.2in}}
+  \hline
+ & species number of SNPs per window & mean length in bp per window & mean number of windows per chromosome & mean ratio of variance explained by top 2 PCs \\ 
+  \hline
+\textit{Drosophila melanogaster} & 1000 & 9019 & 2674 & 0.53 \\ 
+  Human & 100 & 636494 & 203 & 0.55 \\ 
+  \textit{Medicago truncatula} & 10000 & 102580 & 467 & 0.50 \\ 
+   \hline
+\end{tabular}
+\caption {some statistic for each species' genome}
+\end{table}
 
@@
 
-\plr{this describes PCA, not cutting into windows? and, the next section?}
-We cut each recoded matrix into sub matrixes by that have the same columns but fewer rows than the original matrix.
-Then apply Principal Component Analysis (PCA) on each window.
-Here's a brief summary about how is PCA carried out on genomic data \citep{mcvean2009genealogical}.
-Starting with the recoded genotype matrix Z, where Z is a L*N matrix ( L is SNP number ; N is sample size), then zero-center the matrix Z to X to make the data rows have equal mean.
-($X_{si}=Z_{si}-\frac{1}{n}\sum_{j=1}^{n}Z_{sj}$) Get the covariance matrix of X (denote as matrix C) and compute the eigenvectors and eigenvalues of the covariance matrix C.
-($C=\frac{1}{n-1}XX^{T}$ ) The \textit{i}th principal component is the \textit{i}th eigenvector of C.
-\plr{(put equations into the text in the right place)}
+After recoding, we divided the recoded matrix into contigous matrixes that have the same columns but fewer rows than the original matrix, 
+and applied Principal Component Analysis (PCA) separately on each window, 
+following the procedure in \citet{mcvean2009genealogical} as follows. Starting with the recoded genotype matrix $Z$, 
+where $Z$ is a $L\times N$ matrix ( $L$ is SNP number ; $N$ is sample size), 
+first compute the mean-centered matrix $X$, $X_{si}=Z_{si}-\overline{Z_{s}}$, 
+where $\overline{Z_{s}}$ is the mean of non-missing entries, $\overline{Z_{s}}=\frac{1}{n}\sum_{j=1}^{n}Z_{sj}$.
+The mean-centering step makes the result not depend on the choice of reference allele.
+Then find the covariance matrix of $X$, denoted $C=\frac{1}{n-1}XX^{T}$. 
+We compute the covariance matrix using the R function cov() , with use=``pairwise",
+which computes the covariance between each pair of individuals using all complete pairs of SNPs on those individuals.
+The principal components are eigenvector of C, ordered by magnitude of the eigenvalues.
 
-PCA on each window is getting the eigenvectors of covariance matrix of the cut recoded matrix. PCA plots generally use PC2 against PC1.
-PC1 and PC2 refer to the first and second principal component respectively. 
-PCA plots can show population structure, which describes how different samples are related. 
-We want to investigate the variation of population structure along the genome. 
-So we apply PCA on each window to check the population structure along the genome.
-\subsubsection{Choose window length}
+The top few principal components generally display population structure; we usually use the first two (referred to as PC1 and PC2).
+
+\subsubsection{Choosing window length}
 The window length should neither be too long or too short. 
-The longer the windows, the more accurate is the estimate of population structure in that window.
-However, for better resolution, we need to find a length that is reasonably short.
-If we use the first principal component as a measure of population structure, then to choose a proper length for a
-window, we need to find a balance between variance of the first principal components inside a window and that between windows.
-The variance between windows is estimated as mean variance of the first principal component for each window.
-The variance inside a window is estimated using the jackknife block when cutting the window into 10 equal size smaller windows \cite{efron1982jackknife}.
+In general, longer windows have more SNPs, 
+the noise (the mean standard error within a window) will be smaller
+and so population structure in that window will be more accurately estimated.
+However, to better resolve features along the genome, we need reasonably short windows 
+to have stronger signal (the mean standard variation between windows).
+If we use the first principal component as a measure of population structure, 
+then to choose a proper length for a window, 
+we need to find a balance between variance of the first principal components inside a window and that between windows. 
+The variance between windows is estimated by the mean variance of the first principal component for each window. 
+$Var_{between}= \frac{1}{N}\sum_{j=1}^{N}\left ( \frac{1}{K}\sum_{i=1}^{K}\left ( PC1_{i,j} -\overline{PC1_{,j}} \right )^{2} \right )$ ($K$ is the number of windows on the genome)
+The variance inside a window is estimated using the block jackknife cutting the window into 10 equal size smaller windows \citep{efron1982jackknife}. 
+$Var_{inside}= \frac{1}{N}\sum_{j=1}^{N}\left ( \frac{9}{10}\sum_{i=1}^{10}\left ( PC1_{i,j} -\overline{PC1_{,j}} \right )^{2} \right )$
 Table \ref{tab:window_sizes} shows the comparison of variance within a window and that between windows for chromosome arms in \textit{Drosophila}.
-Finally, we choose 100 SNPs, 1000 SNPs and 10000 SNPs as window length for human, \textit{Drosophila}, and \textit{Medicago} separately.
+Finally, we choose 100 SNPs, 1000 SNPs and 10000 SNPs as window length for human, \textit{Drosophila}, and \textit{Medicago} respectively.
 
@@
   \hline
- & chr\_name & win\_length(in SNPs) & 100 & 500 & 10\verb|^|3 & 10\verb|^|4 & 10\verb|^|5 \\ 
+ &   & window length(in SNPs) & 100 & 500 & 10\verb|^|3 & 10\verb|^|4 & 10\verb|^|5 \\ 
   \hline
-1 & Chr2L & SE\verb|^|2(within) & 2.05e-03 & 1.64e-03 & 1.18e-03 & 1.68e-04 & 4.02e-05 \\ 
-  2 & Chr2L & Var(between) & 2.76e-03 & 2.69e-03 & 2.23e-03 & 6.74e-04 & 3.12e-04 \\ 
-  3 & Chr2R & SE\verb|^|2(within) & 2.18e-03 & 1.92e-03 & 1.63e-03 & 5.76e-04 & 1.35e-04 \\ 
-  4 & Chr2R & Var(between) & 2.78e-03 & 2.70e-03 & 2.65e-03 & 2.31e-03 & 1.82e-03 \\ 
-  5 & Chr3L & SE\verb|^|2(within) & 2.08e-03 & 2.00e-03 & 1.64e-03 & 7.32e-04 & 2.45e-04 \\ 
-  6 & Chr3L & Var(between) & 2.60e-03 & 2.52e-03 & 2.40e-03 & 1.68e-03 & 1.89e-03 \\ 
-  7 & Chr3R & SE\verb|^|2(within) & 1.95e-03 & 1.76e-03 & 1.44e-03 & 5.87e-04 & 2.03e-04 \\ 
-  8 & Chr3R & Var(between) & 2.58e-03 & 2.51e-03 & 2.44e-03 & 1.96e-03 & 1.40e-03 \\ 
-  9 & ChrX & SE\verb|^|2(within) & 2.48e-03 & 2.04e-03 & 1.54e-03 & 1.62e-03 & 1.68e-04 \\ 
-  10 & ChrX & Var(between) & 2.61e-03 & 2.43e-03 & 2.30e-03 & 3.24e-04 & 1.14e-03 \\ 
+ & Chr2L & SE\verb|^|2(within) & 2.05e-03 & 1.64e-03 & 1.18e-03 & 1.68e-04 & 4.02e-05 \\ 
+  & 	 & Var(between) & 2.76e-03 & 2.69e-03 & 2.23e-03 & 6.74e-04 & 3.12e-04 \\ 
+  & Chr2R & SE\verb|^|2(within) & 2.18e-03 & 1.92e-03 & 1.63e-03 & 5.76e-04 & 1.35e-04 \\ 
+  & 	 & Var(between) & 2.78e-03 & 2.70e-03 & 2.65e-03 & 2.31e-03 & 1.82e-03 \\ 
+  & Chr3L & SE\verb|^|2(within) & 2.08e-03 & 2.00e-03 & 1.64e-03 & 7.32e-04 & 2.45e-04 \\ 
+  & 	 & Var(between) & 2.60e-03 & 2.52e-03 & 2.40e-03 & 1.68e-03 & 1.89e-03 \\ 
+  & Chr3R & SE\verb|^|2(within) & 1.95e-03 & 1.76e-03 & 1.44e-03 & 5.87e-04 & 2.03e-04 \\ 
+  & 	 & Var(between) & 2.58e-03 & 2.51e-03 & 2.44e-03 & 1.96e-03 & 1.40e-03 \\ 
+  & ChrX & SE\verb|^|2(within) & 2.48e-03 & 2.04e-03 & 1.54e-03 & 1.62e-03 & 1.68e-04 \\ 
+  & 	 & Var(between) & 2.61e-03 & 2.43e-03 & 2.30e-03 & 3.24e-04 & 1.14e-03 \\ 
    \hline
@@
 \subsubsection{Similarity of population structure between windows}
-We use the first two principal components (PCs) and the corresponding eigenvalues from PCA, and construct a new matrix with the two PCs for each window to stand for the population structure information.
+We compared population structure in different genomic windows 
+using the first two principal components (PCs) and the corresponding eigenvalues from PCA.
+We do this, rather than using the entire covariance matrix for computational efficiency 
+and because the top PCs summarize important population structure, 
+using only these should reduce the effect of noise.
 For example, the constructed matrixes for ith and jth window are as following. 
-($\lambda _{1i}$ and $\lambda _{2i}$ are the eigenvalues for the first two PCs for \textit{i}th window; $\lambda _{1j}$ and $\lambda _{2j}$ are the eigenvalues for the first two PCs for \textit{j}th window.) $M_i$ and $M_j$ are the constructed new matrix for \textit{i}th window and \textit{j}th window.
+($\lambda _{1i}$ and $\lambda _{2i}$ are the eigenvalues for the first two PCs for \textit{i}th window; 
+$M_i$ is the constructed new matrix for \textit{i}th window and \textit{j}th window.
 \plr{
@@
 \begin{align}
-    M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} \\
-    M_{j} &= \frac{\lambda_{1j}PC1_{j}PC1_{j}^{T}+\lambda_{2j}PC2_{j}PC2_{j}^{T}}{\lambda_{1j}+\lambda_{2j}}
+    M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
 \end{align}
 
-The Euclidean distance $D_{ij}$ between the constructed matrices $M_i$ and $M_j$ stands for the similarity of population structure for the \textit{i}th window and \textit{j}th window. 
-Due to the property of eigenvectors,we could use the following method to calculate the pairwise distance greatly saving time and space.
+The Euclidean distance $D_{ij}$ between the matrices $M_i$ and $M_j$ stands for the similarity of population structure for the \textit{i}th window and \textit{j}th window. 
+Due to the orthogonality of eigenvectors,we could use the following method to calculate the pairwise distance greatly saving time and space.
 
@@
     \begin{aligned}
-        V_{1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} \qquad
-        V_{2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} \\
-        U_{1}&=\sqrt{\frac{\lambda _{1j}}{\lambda _{1j}+\lambda _{2j}}}PC1_{j} \qquad
-        U_{2}&=\sqrt{\frac{\lambda _{2j}}{\lambda _{1j}+\lambda _{2j}}}PC2_{j}
+        V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} \qquad
+        V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} \\
+        V_{j1}&=\sqrt{\frac{\lambda _{1j}}{\lambda _{1j}+\lambda _{2j}}}PC1_{j} \qquad
+        V_{j2}&=\sqrt{\frac{\lambda _{2j}}{\lambda _{1j}+\lambda _{2j}}}PC2_{j}
     \end{aligned}
 \end{equation}
+
+\begin{align}
+    M_{i} &= V_{i1}V_{i1}^{T}+V_{i2}V_{i2}^{T}
+\end{align}
 
@@
                 \left ( 
-                    V_{1}\cdot V_{1}^{} 
+                    V_{i1}\cdot V_{i1}^{} 
                 \right )^{2}
-                + \left ( V_{2}\cdot V_{2}^{} \right )^{2}
-                + \left ( U_{1}\cdot U_{1}^{} \right )^{2} 
-                +\left ( U_{2}\cdot U_{2}^{} \right )^{2} 
+                + \left ( V_{i2}\cdot V_{i2}^{} \right )^{2}
+                + \left ( V_{j1}\cdot V_{j1}^{} \right )^{2} 
+                +\left ( V_{j2}\cdot V_{j2}^{} \right )^{2} 
         \right. \\ 
@@
             -2\left [ 
-                \left ( V_{1}\cdot U_{1}^{} \right )^{2}
-                + \left ( V_{1}\cdot U_{2}^{} \right )^{2}
-                + \left ( V_{2}\cdot U_{1}^{} \right )^{2} 
-                + \left ( V_{2}\cdot U_{2}^{} \right )^{2}
+                \left ( V_{i1}\cdot V_{j1}^{} \right )^{2}
+                + \left ( V_{i1}\cdot V_{j2}^{} \right )^{2}
+                + \left ( V_{i2}\cdot V_{j1}^{} \right )^{2} 
+                + \left ( V_{i2}\cdot V_{j2}^{} \right )^{2}
             \right ]
@@
 \subsubsection{Visualize the pairwise distance matrix}
-To do this, I use Multidimensional scaling (MDS), 
-which is a visualization method commonly applied to distance matrixes. 
+We use Multidimensional scaling (MDS) method to visualize the distance matrices. 
 It can reduce the dimensionality of a distance matrix while preserve the distance information between objects as well as possible. 
-The aimed dimension `` M'' can be set based on your need.
-We used M=1 and M=2 in our study.
-This allowed us to visualize the information in the distance matrix in one or two dimensional relation between windows' population structure.
+The result is a set of coordinates for each sample with the property that 
+the first M coordinates give the arrangement in M-dimensional space that best recapitulates the original distance matrix.
+We use M=2 to produce one or two dimensional visualization of relationships between windows' population structure.
 
 \section{Results}
-\textbf{In all these 3 species, PCA plots vary along the genome.}
+In all these 3 species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
+This implies that variation is truly due to population structure, 
+since fluctuations due to demographic noise are not expected to show long distance correlations. 
+Below, we display the results and investigate likely underlying causes.
 
-Since PCA plots can show population structure, this shows that the population structure varies along the genome.
-Each PC plot comes from the covariance matrix of a different section of the genome, which may result from different tree for each position, chromosome inversions, linked selection and so on.
-We investigate the pattern in each species here separately, combining MDS method and genome features.
-Here are the MDS results for each species.
-
-\subsection{\textit{Drosophila}}
-We checked the results for chromosome arms Chr2L, Chr2R, Chr3L, Chr3R and ChrX separately.
-When setting the dimension parameter M=2 in MDS method, that is, plotting the first 2 coordinates reduced from the distance matrix, each plot looks like triangle.
-(Figure \ref{fig:mds_chr2L}a) Since the relative position for each window in the plot shows the relative similarity between windows, it tells us there are 3 extreme types of population structure shown in the 3 peaks of the ``triangle'', and other windows are between them, which means other window's population structure might be a mixture of those extremes.
-We then want to investigate more information at the extremes.
+\subsection{\textit{Drosophila melanogaster}}
+We ran the above method on chromosome arms 2L, 2R, 3L, 3R and X separately. 
+For each, the two-dimensional MDS visualization resembles a triangle. (eg. Figure \ref{fig:mds_chr2L}a) 
+Since the relative position of each window in this plot shows the similarity between windows, this suggests that there are 3 extreme types of population structure shown in the 3 peaks of the ``triangle'', and that other window's population structure might be a mixture of those extremes.
 
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig1_Together_MDS_plot_Chr2L_final_abline}
+       \includegraphics[width=0.9\textwidth]{Fig1_temp}
     \end{center}
     \caption{
-         Variation in structure for windows on chromosome arm Chr2L. 
-         Each represents a window: (a) shows the MDS visualization of relationships between windows. 
-         (b) and (c) show the position (midpoint) of each window against the first and second MDS coordinates, respectively. 
-         Colors are consistent for (a) (b) and (c).
-         Vertical black lines show the breakpoints of known polymorphic inversions In(2L)t.
+         Variation in structure for windows on chromosome arms 2L, 2R, and 3R. 
+         Each represents a window: the first column shows the MDS visualization of relationships between windows. 
+         The second and third columns show the position (midpoint) of each window against the first and second MDS coordinates, respectively. 
+         Rows show results from different chromosome arms, from top to bottom: Chr2L, Chr2R, Chr3L. 
+         Colors are consistent for plots in each row. 
+         Vertical black lines show the breakpoints of known polymorphic inversions (from top to bottom: In(2L)t, In(2R)NS, In(3L)OK).
          \label{fig:mds_chr2L}
@@
 
-We pick a window for each extreme, and take out 5\% windows that are closest to it in the MDS 2D plot , then combine those windows for each extreme and apply PCA on them.
-(Figure \ref{fig:pca_chr2L}) We can see the obvious difference between their PCA plots.
-This difference stands for variation of population structure along the genome in \textit{Drosophila} Chr2L.
+To investigate these extremes, we pick a window for each extreme, 
+and take out the 5\% of windows that are closest to it in the MDS coordinates, 
+then combine those windows for each extreme and apply PCA on the corresponding sections of genome. 
+We can see in Figure \ref{fig:pca_chr2L} the obvious difference between their PCA plots.
 
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig2_all_pca_plots_for_Chr2L_3peaks_byMDS}
+       \includegraphics[width=0.9\textwidth]{Fig2_temp}
     \end{center}
     \caption{      
-        PCA plots for each peak in Figure \ref{fig:mds_chr2L}a. 
+        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_chr2L}. 
         Each point represents a sample. 
         The plots are plotting the second principal components against the first principal components. 
-        (a) shows the windows colored green in Figure 1; 
-        (b) shows orange windows; and (c) shows purple windows.
+        The first column shows the windows colored green in Figure 1; the second column shows orange windows; and third shows purple windows.
          \label{fig:pca_chr2L}
@@
 
-There's a known inversion in Chr2L in \textit{Drosophila}, In(2L)t, with breakpoints at 2225744bp and 13154180bp.
-\citep{corbett2012population} We recolored the PCA plots in Figure \ref{fig:pca_chr2L} a,b,c by the direction of the inversion for each sample using data from John Pool's lab \citep{lack2015drosophila}.
+There's a large inversion on Chr2L in that is polymorphic in these samples, In(2L)t \citep{corbett2012population}.
+We recolored the PCA plots in Figure \ref{fig:pca_chr2L} by the orientation of the inversion for each sample.
 Figure \ref{fig:color_inver}
-\plr{(citation?)}
-The results show that PCA plots, and thus population structure, varies significantly along the genome, with most variation closely related to inversions. 
-Extremes of structure occurred around inversion breakpoints or in the centers of inversions (green and orange parts in Chr2L).
+Two regions of similar, extreme population structure (green in Figure \ref{fig:mds_chr2L}) are found around inversion breakpoints, 
+and the other two extremes occur in the center of the inversion and between the inversion and centromere.
 The corresponding PCA plots show that locally, population structure is mostly determined by which orientation of the inversion each sample has.
-Similar results are found in other chromosome arms that have known polymorphic inversions (Chr2R, Chr3L, Chr3R) in \textit{Drosophila}. (See supplementary for more details)
-These facts show that the variation of population structure along the Drosophila genome is mainly due to inversions.
+Similar results are found in other chromosome arms that have known polymorphic inversions (Chr2R, Chr3L, Chr3R) in \textit{Drosophila}.
 
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig3_all_pca_plots_for_Chr2L_3peaks_color_by_In2Lt_byMDS}
+       \includegraphics[width=0.9\textwidth]{Fig3_temp}
     \end{center}
     \caption{
-         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversion, In(2L)t.
+         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversion, In(2L)t, In(2R)NS, In(3L)OK. (Data from\citet{lack2015drosophila}.
         \label{fig:color_inver}
@@
 \subsection{Human}
-We ran our method separately on all 22 autosomes. 
-We found that the windows that are outliers in the first MDS coordinate (Figure \ref{fig:mds_human}b) coincide with the position of known polymorphic inversions on 8p23. 
-Similar results are found in other chromosomes that have known polymorphic inversions (Chromosome 15, Chromosome 17) in Human. (See supplementary for more details)
-Similar to Drosophila, the biggest source of variability in population structure along each human chromosome seems to be polymorphic inversions.
+We ran our method separately on all 22 human autosomes. 
+For instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}b) coincide with the position of known polymorphic inversions on 8p23. 
+Similar results are found in other chromosomes that have known inversions (eg. Chromosome 15, Chromosome 17).
+We found that the primary axis of variation in population structure differentiated only one or a few windows on each chromosome, 
+unlike the countinous variation in population structure we see in Drosophila.
 Other chromosomes showed similar results around predicted inversions: PCA might provide an additional way to identify inversions \citep{ma2012investigation}.
-When we run the process for all 22 autosomes together, the outlying signal of chromosome 8 is still impressing. (See supplementary for all 22 autosomes' MDS results for human)
 
-\plr{Readers will wonder if the "triangle" looks the same in humans and Medicago as in Drosophila. Should we add them, or just put these in the supplement?}
-
-\plr{Do we know what happens when we run all 22 autosomes together?}
+When we run the method on all 22 autosomes together, the outlying signal of chromosome 8 is still visible. (See supplementary for all 22 autosomes)
 
@@
     \begin{center}
-       \includegraphics[width=0.7\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_final_abline}
+       \includegraphics[width=0.9\textwidth]{Fig4_temp}
     \end{center}
     \caption{
-         Variation in structure between windows on human chromosome 8. Each point in the plot represents a window. 
-         (a) shows the MDS visualization of relationships between windows; 
-         (b) and third (c) show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
-         The vertical red lines show the breakpoints of the known inversion on 8p23. \citep{antonacci2009characterization}
+         Variation in structure between windows on human chromosomes 8, 15, and 17. 
+         Each point in the plot represents a window.
+         The first column shows the MDS visualization of relationships between windows;
+         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
+         Rows, from top to bottom show chromosomes 8, 15, and 17. 
+         The vertical red lines show the breakpoints of the known inversions \citep{antonacci2009characterization}.
         \label{fig:mds_human}
@@
 \subsection{\textit{Medicago truncatula}}
-We checked the MDS results for all 8 chromosomes. 
+We ran our method on all 8 chromosomes of \textit{Medicago truncatula} seperately. 
+We also ran all chromosomes together.
 \plr{The results looked different than the other two species, with much less pronounced peaks...}
-We found the position of the peak for each MDS plot has a coincidence with the position of heterochromatic regions.
-This means the population structure in the windows located in heterochromatin tends to have higher similarity, since those windows are closer in MDS plots.
-(Figure \ref{fig:mds_medicago}) Biologically,
-heterochromatic regions have lower gene density and may be less subject to selection \citep{kulikova2001integration,paape2013selection}.
-Then we checked the MDS results against gene density along the genome for each chromosome using gene models in Mt4.0 JBrowse.
-Numerically, the first MDS coordinate value is negatively correlated to the gene count for each window, which shows the correlation between MDS result and gene density.
-(Figure \ref{fig:mds_gene_count})
-This fact shows that in \textit{Medicago truncatula}, variation in population structure, as measured by the first MDS coordinate, is negatively correlated to gene density. 
+We found the position of the peak for each MDS plot has a coincidence with the position of heterochromatic regions. (Figure \ref{fig:mds_medicago})
+This means that population structure in the windows located in heterochromatin tends to have higher similarity, since those windows are closer in MDS plots.
+Biologically, heterochromatic regions have lower gene density and may be less subject to selection \citep{kulikova2001integration,paape2013selection}.
+To investigate this, we computed gene density near each window using gene models in Mt4.0 JBrowse \citep{tang2014improved}. 
+Indeed, the first MDS coordinate value is negatively correlated to the gene count for each window (Figure \ref{fig:mds_gene_count}).
+Besides, we found the PCA plots are quite consistent for the peaks that more clustering away from the heterochromatic regions, 
+while the PCA plots are barely consistent for the peaks that more clustering around heterochromatic regions for the 8 chromosomes.
 Unlike in the human and Drosophila genomes, in \textit{Medicago truncatula}, major variation in population structure is likely due to linked selection.
@@
     \begin{center}
+       \includegraphics[width=0.9\textwidth]{medicago_mds}
+    \end{center}
+    \caption{
+     MDS visualization for Medicago chromosome 3. Each point in the plot stands for a window (length 1000 SNPs).
+       \label{fig:mds12_medicago}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{medicago_peaks_pca}
+    \end{center}
+    \caption{
+        PCA plots for the three sets of genomic windows colored separately in Figure \ref{fig:mds12_medicago}. 
+        From left to right, the windows colored green, orange, and purple. Each point corresponds to a sample, colored by origin.
+        \label{fig:pca_medicago}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
        \includegraphics[width=0.9\textwidth]{Fig6_MDS_1D_and_gene_count_and_lm_chr3_based_all_chr_win104}
@@
         (B) Gene count for each window against the position of each window; 
-        (C) The first MDS coordinate is significantly correlated with gene count (r=XXX, p=XXX). 
+        (C) The first MDS coordinate is significantly correlated with gene count (r=0.20, p=$2.2\times 10^{-16}$). 
         (See supplementary for other chromosomes' MDS results against gene count for \textit{Medicago})
@@
 
+We aim to investigate the population structure variation along the genome, so we look into it for each genomic window locally. 
+The choice of window length depends on the genomic data's features, such as data size and noise.
+
 We've shown that local summaries of tree shape, i.e., kinship, vary significantly on a large scale along the genome,
@@
 There are many ways to visualize population structure; perhaps others would show different things.
-
-Come back to some points in the Introduction.
+For example, SMARTPCA removes effects of linked blocks, while our method don't.
 

### 2016-05-13 medicago/medicago_data_setup.Rmd
added in total variance output to eigen_windows and split out cov_pca

@@
 knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
-```
-
 ```
@@
 win.fn <- function (n,win.snps=100) {
-    sites <- list( chr1=bcf.sites[[1]][seq(1+(n-1)*win.snps,length.out=win.snps)] )
-    regions <- enclosing_region(sites)
+    regions <- data.frame( chrom="chr1", start=1+(n-1)*win.snps, end=n*win.snps )
     vcf_query( bcf.file, regions, get.indivs )
@@
 # there may be windows with missing data
-na.inds <- is.na( snp.pca$values[1,] )
+na.inds <- is.na( snp.pca[1,] )
 chr1.mds <- cmdscale( chr1.pcdist[!na.inds,!na.inds], eig=TRUE, k=4 )
@@
 country.pch <- rep(1:6,length.out=length(country.names))
-for (k in 1:ncol(corner.pcs$vectors)) {
-    xy <- matrix( corner.pcs$vectors[,k], nrow=nrow(corner.pcs$vectors)/nrow(corner.pcs$values) )
+for (k in 1:nrow(corner.pcs)) {
+    xy <- matrix( corner.pcs[k,-(1:3)], ncol=2 )
     plot(xy, xlab="PC 1", ylab="PC 2", main=paste("corner",k), 

### 2016-05-13 medicago/summarize_run.Rmd
added in total variance output to eigen_windows and split out cov_pca

@@
 }
+```
+```{r plot_corners}
 # plot corners and MDS along the chromosome

### 2016-05-13 package/R/cov_pca.R

@@ -1,0 +1,21 @@
+#' Find the Top k Eigenvalues and Eigenvectors of the Covariance Matrix
+#'
+#' This takes a numeric matrix, subtracts row means, finds the covariance matrix of the columns (using \code{use="pairwise"}),
+#' and returns a vector containing: the sum of squared entries in the covariance matrix; the top k eigenvalues, and the top k eigenvectors.
+#'
+#' @param x A numeric matrix, possibly with missing values.
+#' @param k Number of eigenvalue/eigenvector pairs to return.
+#' @return A numeric vector: 
+#' the first entry gives the total sum of squared values of the covariance matrix 
+#' (i.e., the sum of the eigenvectors, for computing the proportion of variance explained);
+#' the next k entires give eigenvalues and each subsequent group of ncol(data) columns give the corresponding eigenvector.
+#' @export
+cov_pca <- function (x,k) {
+    x <- sweep( x, 1, rowMeans(x,na.rm=TRUE) )
+    covmat <- cov(x,use="pairwise")
+    if(any(is.na(covmat))) {return(rep(NA,2*(nrow(covmat)+1)))}
+    PCA <- eigen(covmat)
+    # returns in order (values, vectors)
+    return( c( PCA$values[1:k], PCA$vectors[,1:k] ) )
+}
+

### 2016-05-13 package/R/eigen_windows.R

@@ -55,16 +55,11 @@ eigen_windows_winfn <- function (
     this.lapply <- if (mc.cores>1) { function (...) parallel::mclapply(...,mc.cores=mc.cores) } else { lapply }
     if (is.null(do.windows)) { do.windows <- seq_len(attr(win.fn,"max.n")) }
     .local <- function(x) {
-        chunk <- win.fn(n=x,...)
-        chunk <- sweep( chunk, 1, rowMeans(chunk,na.rm=TRUE) )
-        cov <- cov(chunk,use="pairwise")
-        if(any(is.na(cov))) {return(rep(NA,2*(nrow(cov)+1)))}
-        PCA <- eigen(cov)
-        # returns in order (values, vectors)
-        return( c( PCA$values[1:k], PCA$vectors[,1:k] ) )
+        # returns in order (total, values, vectors)
+        cov_pca( win.fn(n=x,...), k=k )
     }
     eigen.mat <- do.call( rbind, this.lapply( do.windows, .local ) )
-    colnames(eigen.mat) <- c( paste0("lam_",1:k), as.vector( paste0("PC_", t(outer( 1:k, samples(win.fn), paste, sep="_" )) ) ) )
+    colnames(eigen.mat) <- c( "total", paste0("lam_",1:k), as.vector( paste0("PC_", t(outer( 1:k, samples(win.fn), paste, sep="_" )) ) ) )
     attr(eigen.mat,"npc") <- k
     return( eigen.mat )
 }

### 2016-05-13 package/R/pc_dist.R

@@ -4,8 +4,8 @@
 #' return the (n x n) matrix whose [i,j]th element is the Frobenius norm of the difference between the i-th and the j-th matrices,
 #' as approximated by the given eigenvalues/vectors. The approximation is in pseudocode
 #'      M = values[1] * outer(vectors[1],vectors[1]) + values[2] * outer(vectors[2],vectors[2]) + ...
-#' If \code{normalize} is TRUE, then the the matrices are normalized to have norm 1, so that in the definition of M,
-#' \code{values} is replaced by \code{values/sqrt(sum(values^2))}.
+#' If \code{normalize} is set, then the the matrices are normalized to have norm 1, so that in the definition of M,
+#' \code{values} is replaced by \code{values/sqrt(sum(values))}.
 #'
 #' @param x Matrix as output by eigen_windows().
 #' @param npc Number of pcas computed.
@@ -15,8 +15,8 @@
 #' @export
 pc_dist <- function( x, npc, normalize="L1", mc.cores=1 ) {
     this.lapply <- if (mc.cores>1) { function (...) parallel::mclapply(...,mc.cores=mc.cores) } else { lapply }
-    values <- x[,1:npc,drop=FALSE]
-    vectors <- x[,-(1:npc)]
+    values <- x[,1+(1:npc),drop=FALSE]
+    vectors <- x[,-(1:(1+npc))]
     if (normalize=="L1") { values <- sweep( values, 1, abs(rowSums(values)), "/" ) }
     if (normalize=="L2") { values <- sweep( values, 1, sqrt(rowSums(values^2)), "/" ) }
     n <- nrow(values)

### 2016-05-13 package/R/query_vcf.R

@@ -1,19 +1,3 @@
-#' Find Minimal Enclosing Regions
-#'
-#' Given a named list of positions, whose names correspond to chromosomes,
-#' return the data frame of regions with names chrom,start,end that encloses them.
-#'
-#' @param sites Named list of positions, as output by \code{vcf_positions()}.
-#' @return A data frame with variables chrom, start, and end.
-#' @export
-enclosing_region <- function (sites) {
-    if (length(sites)==0 || max(sapply(sites,length))==0) { return( NULL ) }
-    return( data.frame(
-            chrom=names(sites),
-            start=sapply(sites,min,na.rm=TRUE),
-            end=sapply(sites,max,na.rm=TRUE)
-        ) )
-}
 
 #' Make Regions String from Data Frame of Regions
 #'

### 2016-05-13 medicago/summarize_run.Rmd
multi-file vcf querying

--- 2016-05-12 medicago/summarize_run.Rmd
@@
 opt <- fromJSON("config.json")
+
+# original data files
+chroms <- paste0("chr",1:8)
+bcf.files <- file.path( "..", "..", "data", paste0(chroms,"-filtered-set-2014Apr15.bcf") )
+names(bcf.files) <- chroms
 
@@
 }
+```
+```{r plot_corners, fig.width=3*fig.dim}
 # plot corners and MDS along the chromosome
@@
 ```
+
+Now, we'll look at PCA plots from the extracted corners.
+```{r corner_pca, cache=TRUE}
+corner.regions <- lapply( 1:ncol(mds.corners), function (k) {
+                regions[ mds.corners[,k], ]
+    } )
+corner.pca <- lapply( corner.regions, function (reg) {
+            geno <- do.call( rbind, tapply( 1:nrow(reg), reg$chrom, function (k) {
+                        chrom <- reg$chrom[k[1]]
+                        vcf_query( bcf.files[chrom], reg[k,] )
+                }, simplify=FALSE ) )
+            cov_pca(geno,k=4)
+    } )
+```

### 2016-05-13 package/R/cov_pca.R

@@ -16,6 +16,6 @@ cov_pca <- function (x,k) {
     if(any(is.na(covmat))) {return(rep(NA,2*(nrow(covmat)+1)))}
     PCA <- eigen(covmat)
     # returns in order (values, vectors)
-    return( c( PCA$values[1:k], PCA$vectors[,1:k] ) )
+    return( c( sum(covmat^2), PCA$values[1:k], PCA$vectors[,1:k] ) )
 }
 

### 2016-05-13 package/R/query_vcf.R

@@ -18,6 +18,7 @@ region_string <- function (regions) {
 #' Uses bcftools to query vcf or bcf files for genotypes in a collection of regions,
 #' returning the genotypes in numeric format.
 #' Assumes the data are diploid and that every site is diallelic.
+#' Data are returned in the order encountered in the VCF file even if the regions are not in this order.
 #'
 #' @param file The name of the indexed vcf file.
 #' @param regions A data frame containing chrom, start, and end of the regions to be extracted.
@@ -38,6 +39,26 @@ vcf_query <- function (file, regions, samples, verbose=FALSE, recode=TRUE) {
     gt <- c(0L,1L,1L,2L)[match( unlist(gt.text), c("0/0","0/1","1/0","1/1") )]
     dim(gt) <- dim(gt.text)
     return(gt)
+}
+
+#' Query Multiple VCF Files
+#'
+#' Applies \code{vcf_query} to multiple VCF files: must specify which chromosomes are in which file with \code{chrom.list}.
+#'
+#' @param chrom.list A list of character vectors of chromosome names, the \code{n}th saying which chromosomes are available in the \code{n}th vcf file.
+#' @param file The name(s) of the indexed vcf file(s).
+#' @param regions A data frame containing chrom, start, and end of the regions to be extracted.
+#' @param ... Other arguments passed to \code{vcf_query}.
+#' @return The results of the separate \code{vcf_query} calls, \code{rbind}'ed together
+#' (this will *not* necessarily be in the same order as \code{regions}!).
+#' @export
+vcf_multi_query <- function (chrom.list, file=names(chrom.list), regions, ... ) {
+    chroms <- unlist(chrom.list)
+    files <- file[ rep(seq_along(chrom.list),sapply(chrom.list,length)) ]
+    do.call( rbind, lapply( seq_along(file), function (kf) {
+                this.regions <- regions[ files[match(regions$chrom,chroms)]==file[k], ]
+                vcf_query( file[k], this.regions, ... )
+        } ) )
 }
 
 #' Read Position From a VCF File
@@ -155,11 +176,28 @@ vcf_windower_snp <- function (file, sites, size, samples=vcf_samples(file)) {
 	return(win.fn)
 }
 
-#' @describeIn vcf_windower Returns the \code{region} function of a window extractor function.
+#' Window Extractor Functions
+#'
+#' A window extractor function ("winfun") takes an integer vector and returns an integer matrix, 
+#' giving the number of alternate alleles for each sample for the corresponding windows.
[…]

### 2016-05-13 medicago/summarize_run.Rmd
corner pcas added to summarize_run

@@
 Now, we'll look at PCA plots from the extracted corners.
+(warning: takes awhile and a lot of RAM)
 ```{r corner_pca, cache=TRUE}
+corner.npc <- 4
 corner.regions <- lapply( 1:ncol(mds.corners), function (k) {
@@
 corner.pca <- lapply( corner.regions, function (reg) {
-            geno <- do.call( rbind, tapply( 1:nrow(reg), reg$chrom, function (k) {
-                        chrom <- reg$chrom[k[1]]
-                        vcf_query( bcf.files[chrom], reg[k,] )
-                }, simplify=FALSE ) )
-            cov_pca(geno,k=4)
+            geno <- vcf_multi_query( chrom.list=chroms, file=bcf.files, regions=reg )
+            cov_pca(geno,k=corner.npc)
     } )
 ```
+
+First, here are the first `r corner.npc` PCs for each corner, together:
+```{r plot_corner_pca, fig.width=3*fig.dim}
+for (k in seq_along(mds.corners)) {
+    vectors <- matrix( corner.pca[[k]][-(1:(1+corner.npc))], ncol=corner.npc )
+    colnames(vectors) <- paste("PC", 1:corner.npc)
+    pairs(vectors)
+}
+```

### 2016-05-16 writeup/local_pca_paper.tex
added abstract and discussion

@@
 
-\plr{Note: your intro focuses on PCA more than mine, which is trying to be about population structure generally.}
+\section*{Abstract}
+
+Dimension reduction techniques,
+such as principal components,
+are often used to discover and display large-scale structure in genomic datasets
+found in the patterns of kinship
+between the genotyped individuals,
+and to control for the confounding effects of population structure in genome-wide association studies.
+The genome-wide mean kinship this uses
+is an average of the relationships across all locus-specific genealogical trees.
+However, many biological factors,
+including linked selection,
+can systematically skew patterns of kinship over intermediate genomic scales.
+We show how to use principal components analysis (PCA) to describe this meso-scale variation in kinship,
+and apply the method to genomic data from three species.
+In each species we find substantial variation on the scale of megabases to tens of megabases.
+In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
+In a dataset of African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
+is explained by known chromosomal inversions thought to be under recent selection.
+In a range-wide dataset of \textit{Medicago truncatula},
+common axes of variation in population structure are shared between chromosomes,
+correlates with local gene density,
+and may be caused by background selection or local adaptation.
+
+\plr{Add percent of variance explained by first two MDS coordinates, after subtracting the genome-wide mean covariance matrix.}
+
+\plr{Add whole-genome PCA plots for comparison (unless they look just like one of the three corners; in which case say so.}
 
@@
 
-We aim to investigate the population structure variation along the genome, so we look into it for each genomic window locally. 
-The choice of window length depends on the genomic data's features, such as data size and noise.
+The phrase ``population structure'' refers to reduced gene flow between subpopulations, often because of geographical isolation.
+This leads to systematic patterns in genome-wide mean kinship,
+and so visualizations of kinship are said to depict population structure,
+rather than depicting the \emph{effects} of population structure (which would be more accurate).
+However, it is widely recognized that because of selection the effects of gene flow are not equal everywhere on the genome \citep{introgression_review},
+and patterns of polymorphism and divergence vary significantly depending on factors including local gene density \citep{linked_selection,sattath}.
+This implies that the population structure of a species depends on which part of the genome is being examined.
 
-We've shown that local summaries of tree shape, i.e., kinship, vary significantly on a large scale along the genome,
-and that exploratory visualization of this variation can be useful.
+With growing appreciation of the heterogeneous effects of selection across the genome,
+especially the importance of adaptive introgression, hybrid speciation \citep{tomato,mosquito,introgression_review},
+local adaptation \citep{local_adapt_review},
+and inversion polymorphisms \citep{mosquito,kirkpatrick_inversions},
+local PCA may prove to be a useful exploratory tool to discover important genomic features.
+It is unclear whether the technique will be useful on reduced representation genotyping datasets 
+due to marker density and issues with missing data --
+our investigations with one such dataset were inconclusive --
+but even low coverage, whole-genome sequence is very promising.
 
-There are many ways to visualize population structure; perhaps others would show different things.
-For example, SMARTPCA removes effects of linked blocks, while our method don't.
+\paragraph{Confounding in GWAS}
+So-called cryptic relatedness between samples
+has been one of the major sources of confounding in genome-wide association studies (GWAS)
+and so methods must account for it by modeling population structure or kinship \citep{gwas_confounding_review,mixed_models}.
+Since population structure is not constant along the genome,
+this could in principle lead to an inflation of false positives parts of the genome
+with stronger population structure than the genome-wide average.
+Fortunately, in our human dataset this does not seem likely to have a strong effect:
+most variation is due to small, independent regions, possibly primarily inversions,
+and so may not have a major effect on GWAS.
+In the other species we examined, particularly \textit{Drosophila melanogaster},
+treating population structure as a single quantity could be severely misleading.
+
+\paragraph{Parameter choices}
+There are several choices in the method that may affect the results.
+As with whole-genome PCA,
+the choice of samples is important,
+as variation not strongly represented in the sample will not be discovered.
+The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
+but downweighting may be a better option that does not discard data.
+Next, the choice of window size may be important,
+although in our applications results were not sensitive to this,
+indicating that the limit of resolution was smaller than the scale on which patterns of kinship varies along the genome.
+Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in figure \ref{fig:diagram}),
+along with the method used to discover common structure,
+will affect results.
+We used MDS, applied to either each chromosome separately or to the entire genome;
+for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosome,
+but genome-wide, their signal is obscured by the numerous other signals of comparable strength.
+
+Besides window length, there is also the question of how to choose windows.
+In these applications we have used nonoverlapping windows with equal numbers of polymorphic sites.
+Alternatively, windows could be chosen to have equal length in genetic distance,
+so that each would have roughly the same amount of phylogenetic information.
+However, given the insensitivity of our results to window length,
+this seems unlikely to give different results.
+\plr{If we check: We did not have this choice to have a substantial effect on results.}
+
+More generally, there are many possible methods to discover common structure in different parts of the genome.
+The methods we chose discovered strong biological signal of different types in three datasets;
+but it is possible that other methods for measuring dissimilarity between windows' covariance matrices
+or for summarizing the matrix of pairwise distances between windows
+would lead to different insights.
+Minor points we have not explored include how to decide how many PCs to use in approximating structure of each window
+(equations XX),
+how many MDS coordinates to use when describing the distance matrix between windows,
+or how to choose interesting regions of the genome when the MDS plot is not triangular.
+These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
+we encourage the user to experiment.
+
+\paragraph{Chromosomal inversions}
+A major driver of variation in population structure in two datasets we examined are inversions.
+This may be common,
+but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
+PCA has been proposed as a method for discovering inversions \citep{ma_identify_inversions};
+However, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
+or simply regions of reduced recombination.
+However, in many applications, inversions are a nuisance.
+For instance, SMARTPCA \citep{smartpca} reduces their effect on PCA plots
+by regressing out the effect of linked SNPs on each other.
+It would be interesting to see if smoehow removing the effects of inversions in the \textit{Drosophila melanogaster} or human datasets
+would produce a pattern similar to that seen in \textit{Medicago truncatula}.
+
+\plr{Check if I've missed anything, then remove these?}
 

### 2016-05-16 writeup/local_pca_paper.tex
update

--- 2016-05-12 writeup/local_pca_paper.tex
@@
         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} \\
-        V_{j1}&=\sqrt{\frac{\lambda _{1j}}{\lambda _{1j}+\lambda _{2j}}}PC1_{j} \qquad
-        V_{j2}&=\sqrt{\frac{\lambda _{2j}}{\lambda _{1j}+\lambda _{2j}}}PC2_{j}
     \end{aligned}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig1_temp}
+       \includegraphics[width=0.9\textwidth]{Fig1}
     \end{center}
     \caption{
-         Variation in structure for windows on chromosome arms 2L, 2R, and 3R. 
+         Variation in structure for windows on chromosome arms 2L, 2R, 3L, 3R and X. 
          Each represents a window: the first column shows the MDS visualization of relationships between windows. 
          The second and third columns show the position (midpoint) of each window against the first and second MDS coordinates, respectively. 
-         Rows show results from different chromosome arms, from top to bottom: Chr2L, Chr2R, Chr3L. 
+         Rows show results from different chromosome arms, from top to bottom: Chr2L, Chr2R, Chr3L, Chr3R, ChrX. 
          Colors are consistent for plots in each row. 
-         Vertical black lines show the breakpoints of known polymorphic inversions (from top to bottom: In(2L)t, In(2R)NS, In(3L)OK).
+         Vertical black lines show the breakpoints of known polymorphic inversions for Chr2L, Chr2R, Chr3L, Chr3R (from top to down: In(2L)t, In(2R)NS, In(3L)OK, In(3R)K).
          \label{fig:mds_chr2L}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig2_temp}
+       \includegraphics[width=0.9\textwidth]{Fig2}
     \end{center}
@@
 Similar results are found in other chromosome arms that have known polymorphic inversions (Chr2R, Chr3L, Chr3R) in \textit{Drosophila}. 
+Other known polymorphic inversions are like In(3L)P on Chr3L, In(3R)Mo and  In(3R)P on Chr3R. 
+In(3L)P and In(3R)Mo are just a few in our samples, thus we don't see the significant influence of population structure of them. 
+The situation of Chr3R is a little complicated. The coexisting inversions make the MDS visualiazation of it more difficult to pick the extremes.
+For the extremes we picked in Figure \ref{fig:mds_chr2L}, In(3R)K could best explain the PCA clustering result, 
+while picking other windows as extremes might show that In(3R)P explain the PCA clustering result.
 
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig3_temp}
+       \includegraphics[width=0.9\textwidth]{Fig3}
     \end{center}
     \caption{
-         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversion, In(2L)t, In(2R)NS, In(3L)OK. (Data from\citet{lack2015drosophila}.
+         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversion, In(2L)t, In(2R)NS, In(3L)OK and In(3R)K. (Data from \citet{lack2015drosophila}.
         \label{fig:color_inver}
@@
 \subsection{\textit{Medicago truncatula}}
-We ran our method on all 8 chromosomes of \textit{Medicago truncatula} seperately. 
-We also ran all chromosomes together.
+We ran our method on all 8 chromosomes of \textit{Medicago truncatula} seperately 
+and found the correlation between each MDS plot and gene density along the genome in all 8 chromosomes. 
+This consistency implies that the factor driving the population structure for each chromosome might be the same, for example, background selection.
+So we ran all chromosomes together, that is, calculating the pairwise distance for all the windows along 8 chromosomes,
+and then get the MDS coordinates for each chromosome by locating the whole MDS coordinates.
 \plr{The results looked different than the other two species, with much less pronounced peaks...}  
-We found the position of the peak for each MDS plot has a coincidence with the position of heterochromatic regions. (Figure \ref{fig:mds_medicago})
+We computed gene density near each window using gene models in Mt4.0 JBrowse \citep{tang2014improved}. 
+The first MDS coordinate value is negatively correlated to the gene count for each window (Figure \ref{fig:mds_gene_count}).
+We found the position of the peak for each MDS plot has a coincidence with the position of heterochromatic regions. 
 This means that population structure in the windows located in heterochromatin tends to have higher similarity, since those windows are closer in MDS plots.
 Biologically, heterochromatic regions have lower gene density and may be less subject to selection \citep{kulikova2001integration,paape2013selection}. 
-To investigate this, we computed gene density near each window using gene models in Mt4.0 JBrowse \citep{tang2014improved}. 
-Indeed, the first MDS coordinate value is negatively correlated to the gene count for each window (Figure \ref{fig:mds_gene_count}).
 Besides, we found the PCA plots are quite consistent for the peaks that more clustering away from the heterochromatic regions, 
@@
     \caption{
-         MDS results for the \textit{Medicago} genome (chr1-8), using first coordinate of MDS against the middle position of each window along each chromosome. 
-         The bars under MDS plots are diagrams for each chromosome showing the relative position of heterochromatin (the deep purple part on each bar) \citep{kulikova2004satellite, kulikova2001integration}.
+         MDS results and gene density for the \textit{Medicago} genome (chromosome 1-8). 
+         For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
+         The black plot below is gene count for each window against the position of each window.
          \label{fig:mds_medicago}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig6_MDS_1D_and_gene_count_and_lm_chr3_based_all_chr_win104}
+       \includegraphics[width=0.9\textwidth]{Fig6_MDS_1D_against_gene_count_all_chr_win104_with_lm}
     \end{center}
     \caption{
-        First MDS coordinate against gene density for chromosome 3. 
-        (A) First MDS coordinate against the position (midpoint) of each window. 
-        (B) Gene count for each window against the position of each window; 
-        (C) The first MDS coordinate is significantly correlated with gene count (r=0.20, p=$2.2\times 10^{-16}$). 
-        (See supplementary for other chromosomes' MDS results against gene count for \textit{Medicago})
+        First MDS coordinate against gene density for all 8 chromosomes. 
+        The first MDS coordinate is significantly correlated with gene count (r=0.146, p=$2.2\times 10^{-16}$). 
+        (See supplementary for each single chromosome's MDS result against gene count for \textit{Medicago})
         \label{fig:mds_gene_count}

### 2016-05-25 medicago/summarize_run.Rmd
added abstract and discussion
.

@@
 knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
+```
+
+Render this, for instance, like:
+```
+templater::render_template("summarize_run.Rmd",output="lostruct/results_type_snp_size_10000_jobid_324902/run_summary.html",change.rootdir=TRUE)
 ```
@@
 
-# set up to plot all chromosomes together
+#' Set up to plot all chromosomes together
+#' will plot the vector of values 'y' along all the chromosomes.
 chrom.plot <- function (y,ylab='',main='',...) {

### 2016-05-25 writeup/local_pca_paper.tex
update
peter rewrote introduction

@@
 common axes of variation in population structure are shared between chromosomes,
-correlates with local gene density,
+correlate with local gene density,
 and may be caused by background selection or local adaptation.
+The method is a useful addition to the exploratory toolbox
+of population genomics.
 
@@
 
-The kinship matrix contains the kinship coefficient for pairwise individuals. 
-Kinship coefficient defines the genetic relatedness between individuals.
-It is the probability that two alleles randomly selected from two individuals are inherited from the most recent common ancester.
-It could be estimated from given pedigree or from genome-wide covariances of genotype markers. 
-It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, and depend on where on the genome you look; this is why scans for selective sweeps work.
-Populations are often structured in some way while there are systematic genetic variation between populations.
+% The kinship matrix contains the kinship coefficient for pairwise individuals. 
+% Kinship coefficient defines the genetic relatedness between individuals.
+% It is the probability that two alleles randomly selected from two individuals are inherited from the most recent common ancester.
+% It could be estimated from given pedigree or from genome-wide covariances of genotype markers. 
+% It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, and depend on where on the genome you look; this is why scans for selective sweeps work.
+% Populations are often structured in some way while there are systematic genetic variation between populations.
 
-About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
-Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.  
-PCA results are derived from the covariance matrix of genotype matrix. 
+The phrase ``population structure'' refers to reduced gene flow between subpopulations, 
+often because of geographical isolation.
+However, it is widely recognized that because of selection the effects of gene flow are not equal everywhere on the genome,
+and patterns of polymorphism and divergence can vary significantly depending on factors including local gene density.
+This implies that, paradoxically, 
+the population structure of a species depends on which part of the genome is being examined.
+
+Population structure leads to systematic patterns in genome-wide mean kinship,
+and so it is said that visualizations of kinship depict population structure,
+rather than the \emph{effects} of population structure (which would be more accurate).
+The \emph{kinship coefficient} for a pair of individuals
+gives the expected proportion of their genome
+that the two have inherited identically by descent;
+for a single individual it is the inbreeding coefficient.
+However, as \citet{wright1949genetical} wrote,
+``It has probably occurred to the reader that the coefficient of inbreeding
+may mean very different things in different cases.''
+The kinship coefficient originally referred to the expected probability
+of coinheritance within a given pedigree;
+while modern applications to ``unrelated'' individuals
+use a genetic covariance matrix to estimate the \emph{realized} proportion of the genome
+coinherited from sufficiently recent ancestors.
+
+Realized kinship summarizes the shapes
+of the genealogical that relate the samples
+at each location along the genome.
+Since these trees vary along the genome, so does realized kinship;
+but averaging over sufficiently many trees we hope to get a stable estimate,
+independent of the genomic region chosen.
+This hope is not entirely justified: for instance,
+kinship on sex chromosomes is expected to differ from the autosomes;
+and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
+\citep{maynardsmith1974hitchhiking,barton2000genetic,neher2013genetic,charlesworth2012effects},
+and therefore patterns of kinship.
+Indeed,
+chromosome-scale variation in diversity and divergence has been observed in many species
+\citep[e.g.][]{sattath2011pervasive};
+species phylogenies can differ along the genome 
+due to incomplete lineage sorting \citep[e.g.][]{pease2013accurate},
+adaptive introgression and/or local adaptation \citep[e.g.][]{ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
+and theoretical expectations predict that geographic patterns of relatedness should depend on selection
+\citet{charlesworth2003review}.
+Nonetheless, it is not generally known to what extent patterns of kinship vary along the genome,
+nor what the major axes of variation are.
+
+Patterns in genome-wide kinship are often summarized
+by applying principal components analysis (PCA) \citep{patterson2006population} 
+to the genetic covariance matrix,
+as pioneered by \citet{menozzi1978synthetic}.
+% About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}.
+% Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.  
+% PCA results are derived from the covariance matrix of genotype matrix. 
 The results of PCA can be directly related to the underlying genealogical history of the samples, 
-such as coalescence time (time to most recent common ancestor) and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}. 
-Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
-Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
-Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}.
-However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}
+such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}. 
+These patterns are often geographical, producing ``maps'' of population structure
+that reflect the samples' geographic origin distorted by rates of gene flow
+\citep{novembre2008genes},
+although other patterns emerge if recent migration or nongeographic kinship patterns are important
+\citep{astle2009population}.
+% Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
+% Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
+% Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}.
+% However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}
 
-PCA is often used in genome-wide association studies (GWAS) for stratification correction \citep{price2006principal}. 
-However, different parts of genome have different genetic features. 
-First, each site of DNA may have different gene tree. 
-The covariance matrix of genotype matrix averages those gene trees. 
-For a genomic region, if individuals have alleles closer in gene tree, they tends to be more close in PC projections for that region. 
-Different DNA segments may have different gene tree and therefore different population structure for those segments. 
-Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
-Selective sweeps cause local recent ancestry or short trees.
-Balancing selection causes deep trees.
-Background selection causes shallow ones.
-Third, if a chromosome inversion is polymorphic in the sample, the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints.
-Recombination suppression across inversions thus results in different genome structure and population structure.
-Other effects, like noise, introgression might also influence population structure.
+As reviewed by \citet{astle2009population}, 
+modeling ``background'' kinship between samples
+is essential to successful genome-wide association studies,
+and PCA has often been used for stratification correction \citep{price2006principal},
+and so investigating variation in kinship along the genome can help us to have a
+better understanding of the relation between genome structure 
+and population structure, possibly leading to more powerful methods for association studies.
 
-Investigating the genomic variation along the genome can help us to have a
-better understanding of the relation between genome structure and population
-structure, and could possibly lead to more powerful methods for GWAS.
-
-To investigate this, we cut each chromosome into windows (with hundreds to
-thousands of SNPs in each), applied PCA to each window, and visualized how
-population structure, as summarized by PCA, varies along windows.
-
-In this project, we used SNP data for human, \textit{Medicago truncatula}, and
-whole genome sequencing data for \textit{Drosophila}.  Based on the principal
-components, we can estimate the similarity of population structure contained in
-each genome window.  To quantify similarity of population structure between
+To investigate how population structure varies along the genome in several datasets, 
+we cut each genome into windows (with hundreds to thousands of SNPs in each), 
+applied PCA to each window, 
+and visualized the major ways that population structure, as summarized by PCA, varies among windows.
+% In this project, we used SNP data for human, \textit{Medicago truncatula}, and
+whole genome sequencing data for \textit{Drosophila}.  
+% Based on the principal components, we can estimate the similarity of population structure contained in each genome window.  
+To quantify similarity of population structure between
 windows, we constructed for each window an approximate, scaled covariance
-matrix based on the first two PCs measured the pairwise Euclidean distance
-between those matrices.  We use multidimensional scaling to visualize the
+matrix based on the first few principal components, and measured the pairwise Euclidean distance
+between those matrices.  
+We then use multidimensional scaling to visualize the
 relationships between windows, which reduces the pairwise distance matrix to
 lower dimension while preserving the distance information between windows as
-well as possible \citep{borg2005modern}.  To interpret the results of MDS, we
-combine known genome feature information for each species, such as the
-distribution of inversions, and heterochromatin and gene density along the
-genome. Each species showed distinct patterns, reflecting differences in their
-biology.
+well as possible \citep{borg2005modern}.
 
-\plr{Improve this paragraph:}
-Other methods for visualizing population structure are like STRUCTURE,
-\citep{pritchard2000inference,falush2003inference,
-falush2007inference,hubisz2009inferring} 
-model-based approach,
-\citep{yang2012modelbased} 
-maps of heterozygosity
-\citep{ramachandran2005support}.
+% To interpret the results of MDS, we combine known genome feature information for each species, such as the
+% distribution of inversions, and heterochromatin and gene density along the genome. 
+Each species showed distinct patterns, reflecting differences in their biology;
+before presenting results,
+it may help to have a prior idea of what factors are expected to affect population structure,
+and how.
+PCA summarizes patterns in kinship found in the genetic covariance matrix,
+which is an average across locus-specific genealogies.
+% However, different parts of genome have different genetic features. 
+% First, each site of DNA may have different gene tree. 
+% The covariance matrix of genotype matrix averages those gene trees. 
+If individuals are closer in the genealogies of a given genomic region, 
+they tends to be more close in the PCA maps for that region. 
+% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
+% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
+% Selective sweeps cause local recent ancestry or short trees.
+% Background selection causes shallow ones.
+Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
+characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
+A genomic region with many targets for selection experiences
+background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
+which would tend to generally shorten genealogical trees in the region,
+similar to a reduction in effective population size \citep{hudson1995deleterious}.
+% Balancing selection causes deep trees.
+On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
+and population structure locally describes which individuals have which alleles
+rather than geographical proximity.
+Finally, 
+since recombination is suppressed between opposite orientations of a chromosomal inversion
+near its breakpoints,
+genealogies in these regions separate samples carrying the two orientations of the inversion.
+In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
+(indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
+as a way to identify inversions).
+% if a chromosomal inversion is polymorphic in the sample, 
+% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints.
+Recombination suppression across inversions thus results in different genome structure and population structure.
+% Other effects, like noise, introgression might also influence population structure.
+We will look for strong variation in population structure shared across large regions of the genome.
+Many of the effects listed above (such as single selective sweeps or inversions) 
+are not expected to have similar effects on population structure in different regions of the genome
+because of randomness in which samples end up in which group;
+but if these coincide with a region of reduced recombination (such as an inversion),
+these could drive major patterns of variation.
+The more subtle effects of genome-wide linked selection could be shared across large regions
+due to large-scale variation in gene density
+(as by background selection or local adaptation at many genes).
 
-A number of methods for dimensionality reduction also use a strategy of ``local PCA''
+We chose PCA to summarize population structure,
+but other methods, such as STRUCTURE \citep{falush2003inference},
+SPA \citep{yang2012modelbased},
+SpaceMix \citep{bradburd2015spatial},
+or other matrix factorization methods \citep{engelhardt2010analysis}
+would highlight different sources of variation in the data.
+% There are other methods for visualizing population structure are like STRUCTURE,
+% \citep{pritchard2000inference,falush2003inference, falush2007inference,hubisz2009inferring} 
+% model-based approach,
+% maps of heterozygosity
+% \citep{ramachandran2005support}.
+
+Finally, a number of general methods for dimensionality reduction also use a strategy of ``local PCA''
 \citep[e.g.][]{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear},
@@
 Future methods for visualization of genomic data
-may benefit from other advances in this substantial literature \citep[reviewed in][]{vandermaaten2009dimensionality}.
-
-\section{Other Introduction}
-
-Catchy start: the kinship matrix goes back to almost Mendel and is essential in GWAS;
-however, it is well-known that actual relatednesses have a lot of noise about the expected value,
-and depend on where on the genome you look;
-this is why scans for selective sweeps work.
-
-Review of kinship matrix: 
-it's either an expected kinship, given the pedigree;
-or an estimated genome-wide average.
-Wright's path coefficients \citep{wright1943isolation}.
-Why it helps with confounding for GWAS.
-Graham \& Vince's paper, maybe.
-Like IBD, is only well-defined in a known pedigree, up to the founders.
-Kinship and confounding reviewed in \citet{astle2009population}.
-
-Kinship matrices differ for sex chromosomes and the like.
-
-Review of selection causing differential patterns along the genome.
-Locally everything is treelike (gene trees);
-kinship matrix is an average of these (write equation for this).
-Selective sweeps cause local recent ancestry/short trees.
-Balancing selection causes deep trees.
-Background selection, shallow ones.
-Extreme examples of free gene flow in some places between species: e.g. Heliconius.
-Introgression may be nonuniform, e.g. neanderthal, others.
-Refs: hitchhiking \citep{maynardsmith1974hitchhiking},
-\citet{kim2011hitchhiking} study hitchhiking in a spatially subdivided population,
-\citet{mcvean2007structure} looks at the effect of selection on LD.
-\citet{barton2000genetic} reviews hitchhiking.
-\citet{bierne2010distinctive} discusses how hitchhiking effect decreases with geographic distance.
-\citet{charlesworth2003review} reviews patterns of diversity, relating spatial structure to effects of selection.
-
-Review of methods looking along the genome:
-argweaver, HMM between species, ???
-
-What is "population structure"?
-Asks which "populations" are closely related, more diverged, how much diversity do they harbor.
-Often geographic.
-Vital in exploratory data analysis.
-It is a summary of kinship: 
-lack of migration between pops causes a deficit in connections through the pedigree,
-and so affects kinship.
-Wright defined $F_{ST}$ in \citep{wright1949genetical}, and says ``It has probably occurred to the reader that the coefficientof inbreeding may mean very different
-things in different cases.''
-
-Review of methods for visualizing pop structure:
-PCA, structure \citep{falush2003inference}, EEMS, \citep{petkova2014visualizing}, \citep{yang2012modelbased},
-Maps of heterozygosity \citep{ramachandran2005support}.
-Genealogical interpretation of PCA by \citet{mcvean2009genealogical}.
-Other semi-related stuff:
-estimation of covariance matrices;
-local pca(?);
+may benefit from other advances in this substantial literature 
+\citep[reviewed in][]{vandermaaten2009dimensionality}.
 
@@
 
-The phrase ``population structure'' refers to reduced gene flow between subpopulations, often because of geographical isolation.
-This leads to systematic patterns in genome-wide mean kinship,
-and so visualizations of kinship are said to depict population structure,
-rather than depicting the \emph{effects} of population structure (which would be more accurate).
-However, it is widely recognized that because of selection the effects of gene flow are not equal everywhere on the genome \citep{introgression_review},
-and patterns of polymorphism and divergence vary significantly depending on factors including local gene density \citep{linked_selection,sattath}.
-This implies that the population structure of a species depends on which part of the genome is being examined.
+(sentence about how we found variation in pop structure)
 

### 2016-05-25 writeup/local_pca_paper.tex
added more refs

@@
 chromosome-scale variation in diversity and divergence has been observed in many species
-\citep[e.g.][]{sattath2011pervasive};
+\citep[e.g.][]{langley2012genomic};
 species phylogenies can differ along the genome 
@@
 which would tend to generally shorten genealogical trees in the region,
-similar to a reduction in effective population size \citep{hudson1995deleterious}.
+similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
 % Balancing selection causes deep trees.
@@
 
-(sentence about how we found variation in pop structure)
+Our investigations have found a remarkable amount of variation in population structure across the genomes
+of three diverse species,
+revealing distinct biological processes driving this variation in each species.
+More investigation, particularly on more species and datasets will help uncover which patterns are generalizable.
 
 With growing appreciation of the heterogeneous effects of selection across the genome,
-especially the importance of adaptive introgression, hybrid speciation \citep{tomato,mosquito,introgression_review},
-local adaptation \citep{local_adapt_review},
-and inversion polymorphisms \citep{mosquito,kirkpatrick_inversions},
+especially the importance of adaptive introgression, hybrid speciation \citep{pool2015natural,brandvain2014speciation,hufford2013genomic,fitzpatrick2010rapid,staubach2012genome},
+local adaptation \citep{lenormand2002limits,wang2014isolation},
+and inversion polymorphisms \citep{kirkpatrick2015chromosome,kirkpatrick2010chromosome},
 local PCA may prove to be a useful exploratory tool to discover important genomic features.

### 2016-05-26 writeup/local_pca_paper.tex
switched out eigen for RSpectra::eigs

@@
 \bibliography{references}
+
+\appendix
+
+\subsection{Weighted PCA}
+
+PCA can be thought of as finding a good low-dimensional matrix factorization \citep{engelhardt2010analysis}
+that well-approximates the original data in the least-squares sense:
+if $C$ is the $N \times N$ matrix of genotypes, 
+then to find the top $k$ principal components, 
+we find an orthogonal $N \times k$ matrix $U$,
+and a $k \times k$ diagonal matrix $\Lambda$ with diagonal entries $\Lambda_{ii}=\lambda_i$ to minimize
+\begin{align} \label{eqn:objective}
+    \| C - U \Lambda U^T \|^2 = \sum_{ij} \left( G_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 .
+\end{align}
+The columns of $U$, known as the principal components, are the eigenvectors of $C$,
+the eigenvalues of $C$ are $\lambda$, and the proportion of variance explained by the $m^\text{th}$ component is
+\begin{align*}
+    \frac{ \lambda_m }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m U_{im} U_{jm} )^2 }{ \sum_{ij} C_{ij}^2 } .
+\end{align*}
+
+Thinking about the problem as a least-squares approximation problem
+makes it clear why unbalanced sample sizes can result in undesireable outcomes.
+If we want to describe variation \emph{between} populations,
+but 80\% of the samples are from a single population,
+then unless populations are highly differentiated, 
+a better approximation to $C$ may be obtained by using the columns of $U$ to describe variation \emph{within} the overrepresented population
+rather than between the populations.
+A common workaround is to remove samples,
+but a more elegant solution can be found by reweighting the objective function in \eqref{eqn:objective}
+Let $w_{i}$ be a weight associated with sample $i$,
+$W$ the diagonal matrix with $w$ along the diagonal,
+and instead seek to minimize
+\begin{align} \label{eqn:weighted_objective}
+    \| W^{1/2} (C - U \Lambda U^T) W^{1/2} \|^2 = \sum_{ij} \sqrt{W_i W_j} \left( G_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 ,
+\end{align}
+and now for convenince we require $U$ to be orthogonal in $\ell_2(w)$, i.e., that $U^T W U =I$.
+We then would choose $w$ to give roughly equal weight to each \emph{population},
+instead of each individual.
+We have used with good results the weightings
+$w_i = 1/\max(10,n_i)$,
+where $n_i$ is, if there are discrete populations,
+the number of samples in the same population as sample $i$;
+or, for continuously sampled individuals,
+the number of samples within a certain distance of sample $i$.
+
+To solve \eqref{eqn:weighted_objective},
+let $\lambda$ and $V$ denote the eigenvalues and eigenvectors of $W^{1/2} C W^{1/2}$,
+so that $V \Lambda V^T$ is the closest in least squares to $W^{1/2} C W^{1/2}$;
+so if we define $U = W^{-1/2} V$ then 
+then $U^T W U = V^T V = I$,
+and 
+\begin{align*}
+    W^{-1/2} V \Lambda V^T W^{-1/2} 
+    =
+    U \Lambda U^T
+\end{align*}
+is the low-dimensional approximation to $C$.
+The proportion of variance explained is calculated exactly as before,
+but the variance is computed in $\ell_2(w)$:
+\begin{align*}
+    \frac{ \lambda_m }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m w_i w_j U_{im} U_{jm} )^2 }{ \sum_{ij} w_i w_j C_{ij}^2 } .
+\end{align*}
+
+Note finally that since we only need find the top $k$ eigenvectors;
+in our R implementation we use the Spectra library \citep{qiu2016rspectra}.
+
 \end{document}

### 2016-05-26 package/R/cov_pca.R

@@ -14,8 +14,10 @@ cov_pca <- function (x,k) {
     x <- sweep( x, 1, rowMeans(x,na.rm=TRUE) )
     covmat <- cov(x,use="pairwise")
     if(any(is.na(covmat))) {return(rep(NA,2*(nrow(covmat)+1)))}
-    PCA <- eigen(covmat)
-    # returns in order (values, vectors)
-    return( c( sum(covmat^2), PCA$values[1:k], PCA$vectors[,1:k] ) )
+    # PCA <- eigen(covmat)
+    # return( c( sum(covmat^2), PCA$values[1:k], PCA$vectors[,1:k] ) )
+    PCA <- Rspectra::eigs_sym(covmat,k=k)
+    # returns in order (total sumsq, values, vectors)
+    return( c( sum(covmat^2), PCA$values, PCA$vectors ) )
 }
 

### 2016-05-26 package/tests/testthat/test_covariance.R

@@ -1,0 +1,31 @@
+# verify what cov(use="pairwise") does
+
+x <- structure(
+               c(2L, NA, 8L, 8L, 
+                 NA, NA, 8L, 6L, 
+                 NA, 1L, 9L, 2L), 
+           .Dim = c(4L, 3L))
+
+cov.x <- structure(
+                   c(12,  0,  0, 
+                      0,  2,  7, 
+                      0,  7,  19), 
+           .Dim = c(3L, 3L))
+
+expect_equal( cov(x,use='pairwise'), cov.x )
+
+cov_fn <- function (x,y=x) { 
+    ut <- (!is.na(x)) & (!is.na(y))
+    u <- x[ut]; 
+    v <- y[ut]; 
+    sum( (u-mean(u))*(v-mean(v)) )/(sum(ut)-1)
+}
+
+my.cov <- matrix(0,nrow=nrow(cov.x),ncol=ncol(cov.x))
+for (i in 1:ncol(x)) {
+    for (j in 1:ncol(x)) {
+        my.cov[i,j] <- cov_fn( x[,i], x[,j] )
+    }
+}
+
+expect_equal( my.cov, cov.x )

### 2016-05-26 package/tests/testthat/test_vcf_windower.R

@@ -1,0 +1,15 @@
+context("reading windows from VCF files")
+
+
+# has 7 individuals at 71 loci across 4334 bases
+bcf.file <- "test.vcf"
+
+# read in data
+win.fn <- vcf_windower(bcf.file, size=7, type='snp' )
+
+# get regions
+these.regions <- region(win.fn)()
+
+# do local PCA
+pca.stuff <- eigen_windows( win.fn, k=4 )
+

### 2016-05-26 writeup/local_pca_paper.tex
added writeup of weighted pca to appendix

@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig1}
+       \includegraphics[width=1\textwidth]{Fig1_Together_MDS_plot}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig2}
+       \includegraphics[width=0.9\textwidth]{Fig2_all_pca_plots}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=0.7\textwidth]{fig5}
+       \includegraphics[width=0.7\textwidth]{Medicago_MDS_1D_and_gene_count}
     \end{center}
@@
 \begin{align*}
-    \frac{ \lambda_m }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m U_{im} U_{jm} )^2 }{ \sum_{ij} C_{ij}^2 } .
+    \frac{ \lambda_m^2 }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m U_{im} U_{jm} )^2 }{ \sum_{ij} C_{ij}^2 } .
 \end{align*}
@@
 The proportion of variance explained is calculated exactly as before,
-but the variance is computed in $\ell_2(w)$:
+but the variance is computed in $\ell_2(\sqrt{w})$:
 \begin{align*}
-    \frac{ \lambda_m }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m w_i w_j U_{im} U_{jm} )^2 }{ \sum_{ij} w_i w_j C_{ij}^2 } .
+    \frac{ \lambda_m^2 }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m \sqrt{w_i w_j} U_{im} U_{jm} )^2 }{ \sum_{ij} \sqrt{w_i w_j} C_{ij}^2 } .
 \end{align*}

### 2016-05-27 writeup/local_pca_paper.tex
two more refs

@@
 but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
-PCA has been proposed as a method for discovering inversions \citep{ma_identify_inversions};
+PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
 However, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
@@
 However, in many applications, inversions are a nuisance.
-For instance, SMARTPCA \citep{smartpca} reduces their effect on PCA plots
+For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
 by regressing out the effect of linked SNPs on each other.

### 2016-05-28 medicago/summarize_run.Rmd
weighting; stepwise covariance computation

@@
 opt <- fromJSON("config.json")
+if (is.null(opt$weights)) { opt$weights <- 1 }
 
@@
 #' will plot the vector of values 'y' along all the chromosomes.
-chrom.plot <- function (y,ylab='',main='',...) {
+chrom.plot <- function (y,ylab='',main='',chrom.labels=TRUE,...) {
     plot(0, type='n', xlim=range(chrom.offsets/1e6), ylim=range(y,finite=TRUE), 
@@
     abline( v=chrom.dividers/1e6, lty=3, col=adjustcolor("grey",0.5), lwd=2 )
-    axis( 1, at=chrom.mids/1e6, labels=chroms, las=0, tick=FALSE )
+    if (chrom.labels) axis( 1, at=chrom.mids/1e6, labels=chroms, las=0, tick=FALSE )
     points( regions$pos/1e6, y, ...)
@@
 and the computed MDS coordinates, colored by chromosome:
-```{r mds_pairplot}
+```{r mds_pairplot, fig.height=2*fig.dim}
 table(regions$chrom)
@@
 ```
-```{r plot_corners, fig.width=3*fig.dim}
+```{r plot_corners, fig.width=3*fig.dim, fig.height=fig.dim*ceiling(length(mds.cols)/2)}
 # plot corners and MDS along the chromosome
-layout(matrix(c(1,1,2,3),nrow=2),widths=c(1,2))
+layout(matrix(c(rep(1,length(mds.cols)),1+seq_along(mds.cols)),ncol=2),
+        widths=c(1,2), heights=c(rep(1,length(mds.cols)-1),1.25))
 plot( mds.coords[,mds.cols[1:2]], pch=cpch, 
-     col=adjustcolor(ccols,0.75), 
+     col=adjustcolor(ccols,0.75),  asp=1,
      xlab="MDS coordinate 1", ylab="MDS coordinate 2" )
+opar <- par(mar=c(par("mar"),1.1)[c(5,2,5,4)])
 for (k in mds.cols) {
+    lastone <- (k==mds.cols[length(mds.cols)])
+    if (lastone) {
+        par(mar=c(opar$mar,0.2)[c(1,2,5,4)])
+    }
     chrom.plot( mds.coords[,k], pch=20, 
-            xlab="Position (Mb)", main=paste("MDS coordinate",match(k,mds.cols)),
+            xlab=if (lastone) { "Position (Mb)"} else { "" }, # main=paste("MDS coordinate",match(k,mds.cols)),
+            chrom.labels=lastone,
             ylab=colnames(mds.coords)[k],
@@
 }
+par(opar)
 ```
@@
             geno <- vcf_multi_query( chrom.list=chroms, file=bcf.files, regions=reg )
-            cov_pca(geno,k=corner.npc)
+            cov_pca(geno,k=corner.npc,w=opt$weights)
     } )

### 2016-05-28 medicago/data/make-weights.R

@@ -1,0 +1,11 @@
+samples <- read.table("sample_info.tsv",sep="\t",header=TRUE)
+
+samples$nsamples <- table(samples$Country.of.Origin)[samples$Country.of.Origin]
+samples$weight <- 1/pmax(10,samples$nsamples)
+
+bcf.samples <- vcf_samples("chr1-filtered-set-2014Apr15.bcf")
+bcf.samples[bcf.samples=="HM020-I"] <- "HM020"
+
+weights <- data.frame( ID=bcf.samples, weight=samples$weight[match(bcf.samples,samples$ID)] )
+
+write.table(weights, sep="\t", file="inverse-samplesize-weights.tsv", row.names=FALSE)

### 2016-05-28 medicago/run_on_medicago.R

@@ -22,7 +22,7 @@ option_list <- list(
         make_option( c("-t","--type"),   type="character",             help="Window by SNP or by bp?"),
         make_option( c("-s","--size"),   type="integer",               help="Size of the window, in units of type."),
         make_option( c("-k","--npc"),   type="integer",   default=2L,  help="Number of principal components to compute for each window. [default: %default]"),
-        make_option( c("-w","--weights"), type="character",            help="File name containing weights to apply to PCA."),
+        make_option( c("-w","--weightfile"), type="character",            help="File name containing weights to apply to PCA."),
         make_option( c("-m","--nmds"),   type="integer",   default=2L, help="Number of principal coordinates (MDS variables) to compute. [default: %default]"),
         make_option( c("-o","--outdir"), type="character",             help="Directory to save results to.  [default: lostruct/results_type_%type_size_%size_jobid_%jobid/]"),
         make_option( c("-j","--jobid"),  type="character", default=formatC(1e6*runif(1),width=6,format="d",flag="0"),   help="Unique job id. [default random]")
@@ -32,7 +32,7 @@ if (is.null(opt$outdir)) { opt$outdir <- file.path("lostruct",
                                    sprintf( "results_type_%s_size_%d_weights_%s_jobid_%s", 
                                            opt$type, 
                                            opt$size, 
-                                           if (is.null(opt$weights)) { "none" } else { gsub("[.].*","",basename(opt$weights)) }, 
+                                           if (is.null(opt$weightfile)) { "none" } else { gsub("[.].*","",basename(opt$weightfile)) }, 
                                            opt$jobid ) ) }
 if (is.null(opt$type) || is.null(opt$size)) { stop(usage) }
 
@@ -40,18 +40,18 @@ opt$start.time <- Sys.time()
 opt$run.dir <- normalizePath(".")
 print(opt) # this will go in the logfile
 
+# weights
+if (is.null(opt$weightfile)) {
+    opt$weights <- 1
+} else {
+    opt$weights <- read.table(opt$weightfile,sep='\t',header=TRUE)$weight
+}
+
 dir.create( opt$outdir, showWarnings=FALSE, recursive=TRUE )
 cat( jsonlite::toJSON( opt, pretty=TRUE ), file=file.path( opt$outdir, "config.json" ) )
 
 # setup
[…]

### 2016-05-28 package/R/cov_pca.R

@@ -31,3 +31,33 @@ cov_pca <- function (x,k,w=1) {
     return( c( sum(covmat^2), PCA$values, PCA$vectors ) )
 }
 
+#' Construct the Covariance Matrix from the Data in Pieces
+#'
+#' This will compute the covariance matrix for a dataset
+#' that is only available in disjoint pieces (because it is too big to keep in memory at once, for instance).
+#'
+#' Suppose the entire data matrix is D, and n is the logical vector indicating when both D[,i] and D[,j] are nonmissing:
+#'    cov(D)[i,j] = sum( D[nij,i]*D[nij,j] )/(sum(nij)-1) - sum(D[nij,i])*sum(D[nij,i])/( sum(nij)*(sum(nij)-1) )
+#' @param f A function such that f(k) returns the k-th chunk of data.
+#' @param n The number of chunks of data, or a vector of indices.
+#' @export
+running_cov <- function (f,n) {
+    if (is.numeric(n) && length(n)==1) { n <- seq_len(n) }
+    x <- f(1)
+    colm <- colMeans(x,na.rm=TRUE)  # estimate of the column means
+    x <- sweep(x,2,colm,"-")
+    nn <- crossprod(!is.na(x))      # matrix of number of shared nonmissings
+    mu <- colMeans(x,na.rm=TRUE)
+    x[is.na(x)] <- 0
+    xp <- crossprod(x)/(nn-1)       # matrix of mean product of shared nonmissings
+    for (k in n) {
+        x <- sweep( f(k), 2, colm, "-" )
+        new.nn <- crossprod(!is.na(x))
+        nn <- nn + new.nn
+        mu <- ( diag(new.nn)/diag(nn) ) * mu + (1/diag(nn)) * colSums(x,na.rm=TRUE)
+        x[is.na(x)] <- 0
+        xp <- ( new.nn/(nn-1) ) * xp + (1/(nn-1)) * crossprod(x)/(nn-1)
+    }
+    return( xp - (nn/(nn-1))*outer(mu,mu,"*") )
+}
+

### 2016-05-28 package/R/query_vcf.R

@@ -188,7 +188,6 @@ vcf_windower_snp <- function (file, sites, size, samples=vcf_samples(file)) {
 #' @param max.n Index of the largest window.
 #' @param samples Character vector of sample IDs corresponding to columns of extracted data.
 #' @param region A function taking an integer vector returning the chromosome, start, and end of the corresponding windows.
-#' @name winfuns
 as.winfun <- function (f,max.n,samples,region) {
     attr(f,"max.n") <- max.n
     attr(f,"samples") <- samples
@@ -205,7 +204,10 @@ region <- function (f) { function (n=seq_len(attr(f,"max.n"))) { attr(f,"region"
 samples <- function (f) { attr(f,"samples") }
 
 
-#' Gives the number of samples of matrices returned by a window extractor function (access with ncol( )).
+#' Number of Samples Returned by a Window Extractor Function
+#' 
+#' Gives the number of samples of matrices returned by a window extractor function (the number of rows is NA).
+#'
 #' @export 
 #' @method dim winfun
 dim.winfun <- function (f) { c(NA,length(attr(f,"samples"))) }

### 2016-05-28 package/tests/testthat/test_running_cov.R

@@ -1,0 +1,13 @@
+context("running covariance")
+
+xx <- matrix( 1:27, ncol=3 )
+
+f <- function (k) {
+    xx[ (k-1)*3+(1:3), ]
+}
+
+n <- 1:3
+
+expect_equal( xx, do.call(rbind, lapply(n,f) ) )
+
+

### 2016-05-28 medicago/summarize_run.Rmd
different sort of querying function

@@
 corner.pca <- lapply( corner.regions, function (reg) {
-            geno <- vcf_multi_query( chrom.list=chroms, file=bcf.files, regions=reg )
+            geno <- multi_vcf_query( chrom.list=chroms, file=bcf.files, regions=reg )
             cov_pca(geno,k=corner.npc,w=opt$weights)

### 2016-05-28 package/R/cov_pca.R

@@ -17,10 +17,9 @@
 #' and the sum-of-squares (equal to the sum of the eigenvalues squared) is
 #'       sum_{ij}  ( w[i] w[j] )^(1/2) C[i,j]^2 .
 #' @export
-cov_pca <- function (x,k,w=1) {
-    x <- sweep( x, 1, rowMeans(x,na.rm=TRUE) )
+cov_pca <- function (x,k,w=1,
+                     covmat=cov(sweep(x,1,rowMeans(x,na.rm=TRUE),"-"),use='pairwise') ) {
     sqrt.w <- rep_len(sqrt(w),ncol(x))
-    covmat <- cov(x,use="pairwise")
     covmat <- sweep( sweep( covmat, 1, sqrt.w, "*" ), 2, sqrt.w, "*" )
     if(any(is.na(covmat))) {return(rep(NA,2*(nrow(covmat)+1)))}
     # PCA <- eigen(covmat)

### 2016-05-28 package/R/query_vcf.R

@@ -52,7 +52,7 @@ vcf_query <- function (file, regions, samples, verbose=FALSE, recode=TRUE) {
 #' @return The results of the separate \code{vcf_query} calls, \code{rbind}'ed together
 #' (this will *not* necessarily be in the same order as \code{regions}!).
 #' @export
-vcf_multi_query <- function (chrom.list, file=names(chrom.list), regions, ... ) {
+multi_vcf_query <- function (chrom.list, file=names(chrom.list), regions, ... ) {
     chroms <- unlist(chrom.list)
     files <- file[ rep(seq_along(chrom.list),sapply(chrom.list,length)) ]
     do.call( rbind, lapply( seq_along(file), function (kf) {
@@ -60,6 +60,30 @@ vcf_multi_query <- function (chrom.list, file=names(chrom.list), regions, ... ) 
                 vcf_query( file[k], this.regions, ... )
         } ) )
 }
+
+#' Function to Query Multiple VCF Files
+#'
+#' Returns a function that when called with an integer \code{n} will return the \code{n}th region.
+#'
+#' @param chrom.list A list of character vectors of chromosome names, the \code{n}th saying which chromosomes are available in the \code{n}th vcf file.
+#' @param file The name(s) of the indexed vcf file(s).
+#' @param regions A data frame containing chrom, start, and end of the regions to be extracted.
+#' @param ... Other arguments passed to \code{vcf_query}.
+#' @return A function, say, \code{f}, so that \code{f(n)} is the result of \code{n}th \code{vcf_query} call.
+#' @export
+multi_vcf_query_fn <- function (chrom.list, file=names(chrom.list), regions, ... ) {
+    chroms <- unlist(chrom.list)
+    files <- file[ rep(seq_along(chrom.list),sapply(chrom.list,length)) ]
+    qfun <- function (n) {
+        this.region <- regions[n,]
+        vcf_query( files[match(this.region$chrom,chroms)], this.region, ... )
+    }
+    attr(qfun,"max.n") <- nrow(regions)
+    attr(qfun,"region") <- regions
+    attr(qfun,"samples") <- samples(files[1])
+    return(qfun)
+}
+
 
 #' Read Position From a VCF File
 #'

### 2016-05-28 medicago/summarize_run.Rmd
added corner plots to summarize_run.Rmd

@@
 names(bcf.files) <- chroms
+
+sample.ids <- vcf_samples(bcf.files[1])
+samp.file <- file.path( "..", "..", "data", "sample_info.tsv" )
+samps <- read.table(samp.file,sep="\t",header=TRUE)
+samps <- droplevels( samps[match(gsub("-I$","",sample.ids),samps$ID),] )
 
@@
 Now, we'll look at PCA plots from the extracted corners.
-(warning: takes awhile and a lot of RAM)
+(this is done without a ton of memory by accumulating the covariance matrix in `running_cov`):
 ```{r corner_pca, cache=TRUE}
@@
 corner.regions <- lapply( 1:ncol(mds.corners), function (k) {
-            regions[ mds.corners[,k], ]
+            regions[ mds.corners[,k],]
     } )
-corner.pca <- lapply( corner.regions, function (reg) {
-            geno <- multi_vcf_query( chrom.list=chroms, file=bcf.files, regions=reg )
-            cov_pca(geno,k=corner.npc,w=opt$weights)
+corner.covmats <- lapply( 1:ncol(mds.corners), function (k) {
+            reg <- regions[ mds.corners[,k], ]
+            qfun <- multi_vcf_query_fn( chrom.list=chroms, file=bcf.files, regions=reg )
+            running_cov(qfun,1:nrow(reg))
+        } )
+corner.pca <- lapply( corner.covmats, function (covmat) {
+            cov_pca(covmat=covmat,k=corner.npc,w=opt$weights)
     } )
@@
 
-First, here are the first `r corner.npc` PCs for each corner, together:
+Here is the color scheme:
+```{r pop_colors, fig.height=2*fig.dim}
+pop.names <- levels(samps$Country.of.Origin)
+pop.cols <- rainbow_hcl(nlevels(samps$Country.of.Origin))
+pop.pch <- seq_len(nlevels(samps$Country.of.Origin))
+plot( rep(1,length(pop.names)), seq_along(pop.names), pch=pop.pch, col=pop.cols, xlim=c(0,length(pop.names)),
+       xlab='', ylab='', xaxt='n', yaxt='n' )
+text( rep(1,length(pop.names)), seq_along(pop.names), labels=pop.names, pos=4 )
+```
+
+Here are all pairwise plots of the first `r corner.npc` PCs for each of the three corners:
 ```{r plot_corner_pca, fig.width=3*fig.dim}
-for (k in seq_along(mds.corners)) {
-    vectors <- matrix( corner.pca[[k]][-(1:(1+corner.npc))], ncol=corner.npc )
-    colnames(vectors) <- paste("PC", 1:corner.npc)
-    pairs(vectors)
+layout(t(1:3))
+for (i in 1:(corner.npc-1)) {
+    for (j in (i+1):corner.npc) {
+        for (k in 1:ncol(mds.corners)) {
+            vectors <- matrix( corner.pca[[k]][-(1:(1+corner.npc))], ncol=corner.npc )[,c(i,j)]
+            colnames(vectors) <- paste("PC", c(i,j))
+            plot(vectors, pch=pop.pch[samps$Country.of.Origin], 
+                    col=pop.cols[samps$Country.of.Origin] )
+            if (i==1 && j==2) {
+                mtext(paste("corner",k),side=3,cex=2)
+            }
+        }
+    }
 }

### 2016-05-30 writeup/local_pca_paper.tex
fixed math in apx

@@
 \begin{align} \label{eqn:weighted_objective}
-    \| W^{1/2} (C - U \Lambda U^T) W^{1/2} \|^2 = \sum_{ij} \sqrt{W_i W_j} \left( G_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 ,
+    \| W^{1/2} (C - U \Lambda U^T) W^{1/2} \|^2 = \sum_{ij} W_i W_j \left( G_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 ,
 \end{align}
@@
 so that $V \Lambda V^T$ is the closest in least squares to $W^{1/2} C W^{1/2}$;
-so if we define $U = W^{-1/2} V$ then 
+so if we define $U = W^{-1/2} V$
 then $U^T W U = V^T V = I$,
@@
 is the low-dimensional approximation to $C$.
-The proportion of variance explained is calculated exactly as before,
-but the variance is computed in $\ell_2(\sqrt{w})$:
+The proportion of variance explained is calculated from eigenvalues as before,
+but has the interpretation
 \begin{align*}
-    \frac{ \lambda_m^2 }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m \sqrt{w_i w_j} U_{im} U_{jm} )^2 }{ \sum_{ij} \sqrt{w_i w_j} C_{ij}^2 } .
+    \frac{ \lambda_m^2 }{ \sum_\ell \lambda_\ell^2 } 
+    = 
+    \frac{ \sum_{ij} w_i w_j ( \lambda_m U_{im} U_{jm} )^2 }{ \sum_{ij} w_i w_j C_{ij}^2 } .
 \end{align*}

### 2016-05-30 medicago/summarize_run.Rmd
.

@@
 ```{r get_corners}
-mds.corners <- corners( mds.coords[,mds.cols[1:2]], prop=.05 )
+mds.corners <- corners( mds.coords[,mds.cols[1:2],drop=FALSE], prop=.05 )
+corner.coords <- t( apply( mds.corners, 2, function (k) { colMeans( mds.coords[k,mds.cols,drop=FALSE], na.rm=TRUE ) } ) )
 # set up colors and pchs for corners
@@
      xlab="MDS coordinate 1", ylab="MDS coordinate 2" )
+text( corner.coords, labels=1:3 )
 opar <- par(mar=c(par("mar"),1.1)[c(5,2,5,4)])
@@
             plot(vectors, pch=pop.pch[samps$Country.of.Origin], 
-                    col=pop.cols[samps$Country.of.Origin] )
+                    col=pop.cols[samps$Country.of.Origin], cex.axis=1.5 )
             if (i==1 && j==2) {
-                mtext(paste("corner",k),side=3,cex=2)
+                mtext(paste("corner",k),side=3,cex=2,col=corner.cols[k])
             }

### 2016-05-30 medicago/run_on_medicago.R

@@ -89,7 +89,7 @@ rm(pca.stuff)
 
 # distance matrix
 cat("Done finding PCs, computing distances.\n")
-system.time( pc.distmat <- pc_dist( all.pcas, npc=opt$npc ) )
+system.time( pc.distmat <- pc_dist( all.pcas, npc=opt$npc, w=opt$weights ) )
 
 # MDS on the resulting distance matrix
 mds.file <- file.path( opt$outdir, "mds_coords.csv" )

### 2016-06-04 writeup/local_pca_paper.tex
update
various edits

@@
 \newcommand{\plr}[1]{{\em \color{blue} #1}}
+
+\newcommand{\pcone}{PC1}
+\newcommand{\pctwo}{PC2}
+\newcommand{\given}{\,\vert\,}
+\newcommand{\st}{\,\colon\,}
+\renewcommand{\and}{\,\&\,}
+\newcommand{\E}{\mathbb{E}}
+\renewcommand{\P}{\mathbb{P}}
+\DeclareMathOperator{\sgn}{sgn}
+\DeclareMathOperator{\var}{Var}
+\DeclareMathOperator{\cov}{Cov}
 
@@
 
-\plr{How about: first describe the method; then afterwards, present the three datasets.}
+The general steps in our method could be carried out in many ways.
+Referring to Figure \ref{fig:diagram}, these are
+(1) divide the genome into windows,
+(2) summarize the patterns of relatedness (population structure) in each window,
+(3) measure dissimilarity in population structure between each pair of windows,
+(4) visualize the resulting dissimilarity matrix,
+and (5) combine similar windows to more accurately estimate local population structure.
+Details of how we carried these out are given below.
+
 \begin{figure}
@@
 
-\subsection{Recode the DNA sequence to a matrix consisting of 0,1,2 (and NA).}
+\subsection{Datasets}
+% Recode the DNA sequence to a matrix consisting of 0,1,2 (and NA).
 
-For human, we use SNP chip data from POPRES \citep{nelson2008population}.
-There are 3,965 samples in total, (346 African-Americas; 73 Asians; 3,187 Europeans; 359 Indian Asians), and the 22 autosomes together have 447,267 SNPs in this dataset.
-We use the allele that has highest frequency in the samples as the reference allele for each position. 
-The first step is to recode genetic data as a numeric matrix; 
-which we take to have one row per variant and one column per sample. (omitting monomorphic variants)
-In our applications, for each variant we picked a reference allele, 
-and recoded genotypes as the number of non-reference alleles carried by the individual;
-obtaining 0,1, or 2 for autosomal variants in diploids, or NA for missing data.
+We used three publically available datasets (summarized in Table \ref{tab:data_stats});
+as usual for genetic data we converted the data to a numeric matrix
+(with one row per polymorphic variant and one column per sample)
+by replacing each genotype with the number of nonreference alleles
+(or NA for missing data).
 A normalization step (see below) ensures the result does not depend on the choice of reference allele.
 
-For \textit{Drosophila melanogaster}, we use the sequencing data from Drosophila Population Genomics Project (DPGP) and John Pool's lab \citep{lack2015drosophila}, 
+\paragraph{Human:}
+We used genomic data from the entire POPRES dataset \citep{nelson2008population},
+which has array-derived genotype information for 447,267 SNPs across the autosomes
+of 3,965 samples in total: 346 African-Americas, 73 Asians, 3,187 Europeans and 359 Indian Asians.
+(We excluded the sex chromosomes and the mitochondria.)
+We use the allele that has highest frequency in the samples as the reference allele for each position.
+
+\paragraph{\textit{Drosophila melanogaster}:}
+We also used whole-genome sequencing data from Drosophila Population Genomics Project (DPGP) and John Pool's lab \citep{lack2015drosophila}, 
 which together has 380 samples from 16 countries across Africa and Europe.
-Each chromosome arm we investigated (Chr2L, Chr2R, Chr3L, Chr3R) has 2-3 million SNPs.
-Due to high density of missing data for some parts in the genome, we then delete the samples with more than 8\% missing genotypes and then deleting positions with more than 20\% missing data. 
-The cutoff points 8\% and 20\% are determined from the corresponding distributions of NAs in samples and at positions. 
-Since the \textit{Drosophila} samples are from inbred lines with little heterozygosity, we treat the samples as haploid when recoding.
+\plr{Check: are all of our samples described in \citet{lack2015drosophila}? If yes, just say DPGP; if no, add appropriate citation (ask John what this should be). Also, add website?}
+Since the \textit{Drosophila} samples are from inbred lines, we treat the samples as haploid when recoding;
+regions with residual heterozygosity were marked as missing in the original dataset.
+Each chromosome arm we investigated (Chr2L, Chr2R, Chr3L, and Chr3R) has 2--3 million SNPs.
+Due to high density of missing data for some parts in the genome, 
+we first deleted any samples with more than 8\% missing genotypes, 
+and then deleted positions with more than 20\% missing data. 
+(We chose these cutoffs as the tails of the relevant empirical distributions.)
 
-For \textit{Medicago truncatula}, we use the SNP data from whole genome sequencing from \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved}. 
-It has 263 samples from 24 countries. 
-Each of the 8 chromosomes has 3-5 million SNPs.
+\paragraph{\textit{Medicago truncatula}:}
+Finally, we used whole-genome sequencing data from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
+which has 263 samples from 24 countries,
+primarily distributed around the Mediterranean basin.
+Each of the 8 chromosomes has 3-5 million SNPs;
+we did not use the mitochondria or chloroplasts.
 
@@
 \centering
-\begin{tabular}{p{0.8in}p{1in}p{1in}p{1in}p{1.2in}}
+\begin{tabular}{p{0.8in}p{.8in}p{1in}p{1.2in}p{1.4in}}
   \hline
- & species number of SNPs per window & mean length in bp per window & mean number of windows per chromosome & mean ratio of variance explained by top 2 PCs \\ 
+    species & \# SNPs per window & mean window length (bp) & mean \# windows per chromosome & mean \% variance explained by top 2 PCs \\ 
   \hline
-\textit{Drosophila melanogaster} & 1000 & 9019 & 2674 & 0.53 \\ 
-  Human & 100 & 636494 & 203 & 0.55 \\ 
-  \textit{Medicago truncatula} & 10000 & 102580 & 467 & 0.50 \\ 
+  Human & 100 & 636,494 & 203 & 0.55 \\ 
+  \textit{Drosophila melanogaster} & 1,000 & 9,019 & 2,674 & 0.53 \\ 
+  \textit{Medicago truncatula} & 10,000 & 102,580 & 467 & 0.50 \\ 
    \hline
 \end{tabular}
-\caption {some statistic for each species' genome}
+\caption{
+    Descriptive statistics for each dataset used,
+    with the chosen window sizes.
+    See text for how window sizes were chosen;
+    in each window we performed PCA,
+    computed the percent varince explained by the top two principal components;
+    the mean across windows is given in the last column.
+    \label{tab:data_stats}
+}
 \end{table}
@@
 
-After recoding, we divided the recoded matrix into contigous matrixes that have the same columns but fewer rows than the original matrix, 
-and applied Principal Component Analysis (PCA) separately on each window, 
-following the procedure in \citet{mcvean2009genealogical} as follows. Starting with the recoded genotype matrix $Z$, 
-where $Z$ is a $L\times N$ matrix ( $L$ is SNP number ; $N$ is sample size), 
-first compute the mean-centered matrix $X$, $X_{si}=Z_{si}-\overline{Z_{s}}$, 
-where $\overline{Z_{s}}$ is the mean of non-missing entries, $\overline{Z_{s}}=\frac{1}{n}\sum_{j=1}^{n}Z_{sj}$.
-The mean-centering step makes the result not depend on the choice of reference allele.
-Then find the covariance matrix of $X$, denoted $C=\frac{1}{n-1}XX^{T}$. 
-We compute the covariance matrix using the R function cov() , with use=``pairwise",
-which computes the covariance between each pair of individuals using all complete pairs of SNPs on those individuals.
-The principal components are eigenvector of C, ordered by magnitude of the eigenvalues.
+After recoding, we divided the genome into contiguous segments (``windows''; see below),
+% recoded matrix into contigous matrices that have the same columns but fewer rows than the original matrix, 
+and applied Principal Component Analysis (PCA) as described in \citet{mcvean2009genealogical}
+separately to the submatrices that corresponded to each window.
+Specifically, we did PCA as follows:
+denote by $Z$ the $L\times N$ recoded genotype matrix ($L$ is the number of SNPs and $N$ is the sample size), 
+and by $\overline{Z_{s}}$ the mean of non-missing entries for allele $s$, 
+so that $\overline{Z_{s}}=\frac{1}{n_s}\sum_j Z_{sj}$, 
+where the sum is over the $n_s$ nonmissing genotypes.
+We first compute the mean-centered matrix $X$, as $X_{si}=Z_{si}-\overline{Z_{s}}$,
+and preserving missingness.
+(This mean-centering makes the result not depend on the choice of reference allele,
+exactly if there is no missing data, and approximately otherwise.)
+Next, we find the covariance matrix of $X$, denoted $C$,
+as $C_{ij} = \frac{1}{m_{ij}-1} \sum_s X_{si} X_{sj} - \frac{1}{m_{ij}(m_{ij}-1)} (\sum_s X_{si})(\sum_s X_{sj})$,
+where all sums are over the $m_{ij}$ sites where both sample $i$ and sample $j$ have nonmissing genotypes.
+\plr{Is that correct? This is not $X^T X/(m-1)$, which is what was here before, 
+     but is what cov(), use='pairwise' does: see \url{package/tests/testthat/test_covariance.R}.}
+% We compute the covariance matrix using the R function cov() , with use=``pairwise",
+% which computes the covariance between each pair of individuals using all complete pairs of SNPs on those individuals.
+The principal components are the eigenvectors of $C$, 
+normalized to have Euclidean length equal to one,
+and ordered by magnitude of the eigenvalues.
 
-The top few principal components generally display population structure; we usually use the first two (referred to as PC1 and PC2).
+The top few principal components generally display population structure; 
+we usually use the first two (referred to as $\pcone$ and $\pctwo$).
+The above procedure can be performed on any subset of the data;
+for future reference, denote by $\pcone_j$ and $\pctwo_j$
+the result after applying to all SNPs in the $j^\text{th}$ window.
 
-\subsubsection{Choosing window length}
-The window length should neither be too long or too short. 
-In general, longer windows have more SNPs, 
-the noise (the mean standard error within a window) will be smaller
-and so population structure in that window will be more accurately estimated.
-However, to better resolve features along the genome, we need reasonably short windows 
-to have stronger signal (the mean standard variation between windows).
-If we use the first principal component as a measure of population structure, 
-then to choose a proper length for a window, 
-we need to find a balance between variance of the first principal components inside a window and that between windows. 
-The variance between windows is estimated by the mean variance of the first principal component for each window. 
-$Var_{between}= \frac{1}{N}\sum_{j=1}^{N}\left ( \frac{1}{K}\sum_{i=1}^{K}\left ( PC1_{i,j} -\overline{PC1_{,j}} \right )^{2} \right )$ ($K$ is the number of windows on the genome)
-The variance inside a window is estimated using the block jackknife cutting the window into 10 equal size smaller windows \citep{efron1982jackknife}. 
-$Var_{inside}= \frac{1}{N}\sum_{j=1}^{N}\left ( \frac{9}{10}\sum_{i=1}^{10}\left ( PC1_{i,j} -\overline{PC1_{,j}} \right )^{2} \right )$
-Table \ref{tab:window_sizes} shows the comparison of variance within a window and that between windows for chromosome arms in \textit{Drosophila}.
+\plr{I'm talking about flipping signs up here, even though in the code we do it when computing variances, since it makes sense conceptually here.}
+Since eigenvectors are still only defined up to sign,
+when comparing between windows we choose the sign to best match each other:
+after choosing $\pcone_1$, for instance, and $u$ is the first eigenvector obtained from the covariance matrix
+for window $j$,
+then $\pcone_j = \pm u$,
+where the sign is chosen according to which of 
+$\sum_i (\pcone_{i1} - u_i)$ or
+$\sum_i (\pcone_{i1} + u_i)$ 
+is smaller.
+
+\subsection{Choosing window length}
+
+The choice of window length entails a balance between several factors.
+In very short windows, genealogies of the samples will only be represented by a few trees,
+so variation between windows represents demographic noise rather than meaningful variation in population structure.
+Longer windows generally have more distinct trees (and SNPs), 
+allowing for less noisy estimation of local population structure.
+However, to better resolve meaningful signal, i.e., differences in population structure along the genome, 
+we would like reasonably short windows.
+Window length choice therefore entails a signal versus noise tradeoff
+in the estimates of population structure.
+% If we use the first principal component as a measure of population structure, 
+% then to choose the best window length,
+% we need to find a balance between the standard error of the first principal components for each window 
+% and the standard deviation between windows. 
+Since we summarize population structure using relative positions in the principal component maps,
+we quantify ``noise'' as the standard error of a sample's position on PC1 in a particular window,
+averaged across windows and samples,
+and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
+averaged over samples.
+(Recall that the signs for PCs are chosen to match each other.)
+Then, the mean variance across windows is
+\begin{align*}
+    \sigma_\text{signal}^2
+    = 
+    \frac{1}{N} \sum_{j=1}^{N}
+        \frac{1}{L}\sum_{i=1}^{L}\left ( \pcone_{ij} -\overline\pcone_{j} \right )^{2} ,
+\end{align*}
+where $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
+We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife}:
+we divide the $j^\text{th}$ block into 10 equal-sized pieces,
+and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
+then
+$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$,
+and we measure noise by averaging over samples and windows:
+\begin{align*}
+    \sigma^2_\text{noise}
+    &=
+    \frac{1}{N} \sum_{j=1}^{N} \frac{1}{L}\sum_{i=1}^{L} \sigma^2_{ij} .
+\end{align*}
+
+These values are shown calculated separately for each chromosome arm in the \textit{Drosophila} dataset in Table \ref{tab:window_sizes}.
 Finally, we choose 100 SNPs, 1000 SNPs and 10000 SNPs as window length for human, \textit{Drosophila}, and \textit{Medicago} respectively.
@@
 \centering
-\begin{tabular}{rllrrrrr}
+    \begin{tabular}{rlp{1.1in}rrrrr}
   \hline
- &   & window length(in SNPs) & 100 & 500 & 10\verb|^|3 & 10\verb|^|4 & 10\verb|^|5 \\ 
+ & chrom.\ arm  & window length (in SNPs) & 100 & 500 & 1,000 & 10,000 & 100,000 \\ 
   \hline
- & Chr2L & SE\verb|^|2(within) & 2.05e-03 & 1.64e-03 & 1.18e-03 & 1.68e-04 & 4.02e-05 \\ 
-  & 	 & Var(between) & 2.76e-03 & 2.69e-03 & 2.23e-03 & 6.74e-04 & 3.12e-04 \\ 
-  & Chr2R & SE\verb|^|2(within) & 2.18e-03 & 1.92e-03 & 1.63e-03 & 5.76e-04 & 1.35e-04 \\ 
-  & 	 & Var(between) & 2.78e-03 & 2.70e-03 & 2.65e-03 & 2.31e-03 & 1.82e-03 \\ 
-  & Chr3L & SE\verb|^|2(within) & 2.08e-03 & 2.00e-03 & 1.64e-03 & 7.32e-04 & 2.45e-04 \\ 
-  & 	 & Var(between) & 2.60e-03 & 2.52e-03 & 2.40e-03 & 1.68e-03 & 1.89e-03 \\ 
-  & Chr3R & SE\verb|^|2(within) & 1.95e-03 & 1.76e-03 & 1.44e-03 & 5.87e-04 & 2.03e-04 \\ 
-  & 	 & Var(between) & 2.58e-03 & 2.51e-03 & 2.44e-03 & 1.96e-03 & 1.40e-03 \\ 
-  & ChrX & SE\verb|^|2(within) & 2.48e-03 & 2.04e-03 & 1.54e-03 & 1.62e-03 & 1.68e-04 \\ 
-  & 	 & Var(between) & 2.61e-03 & 2.43e-03 & 2.30e-03 & 3.24e-04 & 1.14e-03 \\ 
+    & 2L & $\sigma_\text{noise}$  &    4.53 & 4.05 & 3.44 & 1.30 & 0.63 \\ 
+    & 	 & $\sigma_\text{signal}$ &    5.25 & 5.19 & 4.72 & 2.60 & 1.77 \\ 
+    & 2R & $\sigma_\text{noise}$  &    4.67 & 4.38 & 4.04 & 2.40 & 1.16 \\ 
+    & 	 & $\sigma_\text{signal}$ &    5.27 & 5.20 & 5.15 & 4.81 & 4.27 \\ 
+    & 3L & $\sigma_\text{noise}$  &    4.56 & 4.47 & 4.05 & 2.71 & 1.57 \\ 
+    & 	 & $\sigma_\text{signal}$ &    5.10 & 5.02 & 4.90 & 4.10 & 4.35 \\ 
+    & 3R & $\sigma_\text{noise}$  &    4.42 & 4.20 & 3.79 & 2.42 & 1.42 \\ 
+    & 	 & $\sigma_\text{signal}$ &    5.08 & 5.01 & 4.94 & 4.43 & 3.74 \\ 
+    & X  & $\sigma_\text{noise}$  &    4.98 & 4.52 & 3.92 & 4.02 & 1.30 \\ 
+    & 	 & $\sigma_\text{signal}$ &    5.11 & 4.93 & 4.80 & 1.80 & 3.38 \\
    \hline
@@
 \caption{
-Comparison of variance within a window and that between windows for chromosome arms in \textit{Drosophila}.
+    Measures of signal and noise,
+    computed separately for each chromosome arm in the \textit{Drosophila} dataset,
+    at different window sizes.
+    All values are multiplied by $100$
+    (so typical variation is of order of 50\% of the actual values).
+    Starting at windows of 1,000 SNPs, the signal (variation of PC1 between windows)
+    starts to be substantially larger than the noise (standard error of PC1 for each window).
+    \plr{make 'window length' in bigger bar above the values}
 } \label{tab:window_sizes}
@@
 
-\subsubsection{Similarity of population structure between windows}
+\subsection{Similarity of population structure between windows}
 We compared population structure in different genomic windows 

### 2016-06-04 package/tests/testthat/test_covariance.R

@@ -21,11 +21,20 @@ cov_fn <- function (x,y=x) {
     sum( (u-mean(u))*(v-mean(v)) )/(sum(ut)-1)
 }
 
-my.cov <- matrix(0,nrow=nrow(cov.x),ncol=ncol(cov.x))
+cov_fn_2 <- function (x,y=x) { 
+    ut <- (!is.na(x)) & (!is.na(y))
+    u <- x[ut]; 
+    v <- y[ut]; 
+    sum( u*v )/(sum(ut)-1) - sum(u)*sum(v)/(sum(ut)*(sum(ut)-1))
+}
+
+my.cov.2 <- my.cov <- matrix(0,nrow=nrow(cov.x),ncol=ncol(cov.x))
 for (i in 1:ncol(x)) {
     for (j in 1:ncol(x)) {
         my.cov[i,j] <- cov_fn( x[,i], x[,j] )
+        my.cov.2[i,j] <- cov_fn_2( x[,i], x[,j] )
     }
 }
 
 expect_equal( my.cov, cov.x )
+expect_equal( my.cov.2, cov.x )

### 2016-06-06 writeup/local_pca_paper.tex
peter edited up through methods

@@
 \newcommand{\pctwo}{PC2}
+\newcommand{\pc}[1]{PC#1}
 \newcommand{\given}{\,\vert\,}
@@
 \DeclareMathOperator{\cov}{Cov}
+\DeclareMathOperator{\tr}{tr}
 
@@
 The method is a useful addition to the exploratory toolbox
-of population genomics.
+of population genomics,
+and is implemented as an R package.
 
@@
     \caption{
-         The diagram of our method.
+         An illustration of the method; see text for details.
+         % (1) the genome is divided into windows;
+         % (2) population structure is summarized in each with PCA, and 
+         % (3) dissimilarities for each are computed;
+         % (4) the patterns of dissimilarity are visualized with MDS,
+         % and (5) genomic regions clustering together are aggregated and visualized, again with PCA.
          \label{fig:diagram}
@@
 \centering
-\begin{tabular}{p{0.8in}p{.8in}p{1in}p{1.2in}p{1.4in}}
+    \begin{tabular}{p{0.8in}rrrr}
   \hline
-    species & \# SNPs per window & mean window length (bp) & mean \# windows per chromosome & mean \% variance explained by top 2 PCs \\ 
+    species 
+    & \parbox[t]{.8in}{\# SNPs per \\ window} 
+    & \parbox[t]{1in}{mean window\\ length (bp)}
+    & \parbox[t]{1.2in}{mean \# windows \\ per chromosome} 
+    & \parbox[t]{1.4in}{mean \% variance ex-\\plained by top 2 PCs} \\ 
   \hline
@@
 
-\plr{I'm talking about flipping signs up here, even though in the code we do it when computing variances, since it makes sense conceptually here.}
+\plr{I'm talking about flipping signs up here, even though in the code we do it when computing variances. Do you think it makes sense here?}
 Since eigenvectors are still only defined up to sign,
@@
 and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
-then
+then the estimate of the squared standard error is
 $\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$,
@@
 
-These values are shown calculated separately for each chromosome arm in the \textit{Drosophila} dataset in Table \ref{tab:window_sizes}.
-Finally, we choose 100 SNPs, 1000 SNPs and 10000 SNPs as window length for human, \textit{Drosophila}, and \textit{Medicago} respectively.
+We calculated $\sigma^2_\text{signal}$ and $\sigma^2_\text{noise}$
+for a range of window sizes,
+and for our main results
+chose the smallest window for which $\sigma^2_\text{signal}$ was consistently larger than $\sigma^2_\text{noise}$;
+the values for various window sizes across \textit{Drosophila} chromosomes are shown in Table \ref{tab:window_sizes},
+and the choices for all taxa are in Table \ref{tab:data_stats}.
+In practice, even substantially different window sizes result in the same genome-wide patterns.
+We also chose windows to each consist of the same number of neighboring SNPs.
+We compared to the results when windows are each of the same physical length (in bp) but of the same average length,
+and results were nearly identical.
+
+% # Variances within and between windows for Drosophila, multiplied by 1000:
+% w <- structure(list(chrom = structure(c(1L, 1L, 2L, 2L, 3L, 3L, 4L, 
+% 4L, 5L, 5L), .Label = c("2L", "2R", "3L", "3R", "X"), class = "factor"), 
+%     type = structure(c(1L, 2L, 1L, 2L, 1L, 2L, 1L, 2L, 1L, 2L
+%     ), .Label = c("noise", "signal"), class = "factor"), `100` = c(2.05209, 
+%     2.75625, 2.18089, 2.77729, 2.07936, 2.601, 1.95364, 2.58064, 
+%     2.48004, 2.61121), `500` = c(1.64025, 2.69361, 1.91844, 2.704, 
+%     1.99809, 2.52004, 1.764, 2.51001, 2.04304, 2.43049), `1000` = c(1.18336, 
+%     2.22784, 1.63216, 2.65225, 1.64025, 2.401, 1.43641, 2.44036, 
+%     1.53664, 2.304), `10000` = c(0.169, 0.676, 0.576, 2.31361, 
+%     0.73441, 1.681, 0.58564, 1.96249, 1.61604, 0.324), `1e+05` = c(0.03969, 
+%     0.31329, 0.13456, 1.82329, 0.24649, 1.89225, 0.20164, 1.39876, 
+%     0.169, 1.14244)), .Names = c("chrom", "type", "100", "500", 
+% "1000", "10000", "1e+05"), row.names = c(NA, -10L), class = "data.frame")
+% print(xtable(w),include.rownames=FALSE)
 
@@
 \centering
-    \begin{tabular}{rlp{1.1in}rrrrr}
+    \begin{tabular}{cccrrrrr}
   \hline
- & chrom.\ arm  & window length (in SNPs) & 100 & 500 & 1,000 & 10,000 & 100,000 \\ 
+        & & & \multicolumn{5}{c}{window length (SNPs)} \\
+ & chrom.\ arm  & & 100 & 500 & 1,000 & 10,000 & 100,000 \\ 
   \hline
-    & 2L & $\sigma_\text{noise}$  &    4.53 & 4.05 & 3.44 & 1.30 & 0.63 \\ 
-    & 	 & $\sigma_\text{signal}$ &    5.25 & 5.19 & 4.72 & 2.60 & 1.77 \\ 
-    & 2R & $\sigma_\text{noise}$  &    4.67 & 4.38 & 4.04 & 2.40 & 1.16 \\ 
-    & 	 & $\sigma_\text{signal}$ &    5.27 & 5.20 & 5.15 & 4.81 & 4.27 \\ 
-    & 3L & $\sigma_\text{noise}$  &    4.56 & 4.47 & 4.05 & 2.71 & 1.57 \\ 
-    & 	 & $\sigma_\text{signal}$ &    5.10 & 5.02 & 4.90 & 4.10 & 4.35 \\ 
-    & 3R & $\sigma_\text{noise}$  &    4.42 & 4.20 & 3.79 & 2.42 & 1.42 \\ 
-    & 	 & $\sigma_\text{signal}$ &    5.08 & 5.01 & 4.94 & 4.43 & 3.74 \\ 
-    & X  & $\sigma_\text{noise}$  &    4.98 & 4.52 & 3.92 & 4.02 & 1.30 \\ 
-    & 	 & $\sigma_\text{signal}$ &    5.11 & 4.93 & 4.80 & 1.80 & 3.38 \\
+    & 2L & $\sigma^2_\text{noise}$  & 2.05  &  1.64  &  1.18  &  0.17  &  0.04 \\
+    & 	 & $\sigma^2_\text{signal}$ & 2.76  &  2.69  &  2.23  &  0.68  &  0.31 \\
+    & 2R & $\sigma^2_\text{noise}$  & 2.18  &  1.92  &  1.63  &  0.58  &  0.13 \\
+    & 	 & $\sigma^2_\text{signal}$ & 2.78  &  2.70  &  2.65  &  2.31  &  1.82 \\
+    & 3L & $\sigma^2_\text{noise}$  & 2.08  &  2.00  &  1.64  &  0.73  &  0.25 \\
+    & 	 & $\sigma^2_\text{signal}$ & 2.60  &  2.52  &  2.40  &  1.68  &  1.89 \\
+    & 3R & $\sigma^2_\text{noise}$  & 1.95  &  1.76  &  1.44  &  0.59  &  0.20 \\
+    & 	 & $\sigma^2_\text{signal}$ & 2.58  &  2.51  &  2.44  &  1.96  &  1.40 \\
+    & X  & $\sigma^2_\text{noise}$  & 2.48  &  2.04  &  1.54  &  1.62  &  0.17 \\
+    & 	 & $\sigma^2_\text{signal}$ & 2.61  &  2.43  &  2.30  &  0.32  &  1.14 \\
    \hline
@@
     at different window sizes.
-    All values are multiplied by $100$
+    All values are multiplied by $1,000$
     (so typical variation is of order of 50\% of the actual values).
@@
     starts to be substantially larger than the noise (standard error of PC1 for each window).
-    \plr{make 'window length' in bigger bar above the values}
 } \label{tab:window_sizes}
@@
 \subsection{Similarity of population structure between windows}
-We compared population structure in different genomic windows 
-using the first two principal components (PCs) and the corresponding eigenvalues from PCA.
-We do this, rather than using the entire covariance matrix for computational efficiency 
+
+We think of local population structure as being summarized by \emph{relative} position of the sampmles
+in the space defined by the top principal components.
+Since rotations of this space imply identical population structure,
+we do not compare population structure of different genomic windows using the PCs directly, 
+but instead compare the low-dimensional approximations of the local covariance matrices
+obtained using the top $k$ PCs.
+(For results shown here, we use $k=2$;
+results using larger numbers of PCs were nearly identical.)
+We do this, rather than using the entire covariance matrix, 
+both for computational efficiency 
 and because the top PCs summarize important population structure, 
-using only these should reduce the effect of noise.
-For example, the constructed matrixes for ith and jth window are as following. 
-($\lambda _{1i}$ and $\lambda _{2i}$ are the eigenvalues for the first two PCs for \textit{i}th window; 
-$M_i$ is the constructed new matrix for \textit{i}th window and \textit{j}th window.
-\plr{
-Say what $M_i$ and $M_j$ are in the text.  
-Also, how about we write $V$ instead of $PC$?  I like to use only one letter for variables, so it's clear it's just one thing, not the product of $P$ and $C$.
-And, isn't it $\sqrt{\lambda_{1j}^2+\lambda_{2j}^2}$ on the bottom? Check in the R code (pc{\textunderscore}dist.R).
-}
+so using only these should reduce the effect of noise.
+To remove the effect of variation in mutation rate between windows,
+we also rescale each covariance matrix so that the underlying data matrix has trace norm equal to one.
+To do this, define the $N \times k$ matrix $V(i)$ so that the $\ell^\text{th}$ column of $V(i)$
+is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
+multiplied by $\sqrt{ \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} }$.
+% \begin{equation}
+%     \begin{aligned}
+%         V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} ,
+%         \qquad
+%         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} 
+%     \end{aligned}
+% \end{equation}
+Then, the rescaled, $k$-dimensional covariance matrix for the $i^\text{th}$ window is
+\begin{align}
+    M(i) &= \sum_{\ell=1}^k V_{i\ell} V_{i\ell}^T .
+\end{align}
+% \begin{align}
+%     M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
+% \end{align}
 
+We then use
+Euclidean distance $D_{ij}$ between the matrices $M(i)$ and $M(j)$ 
+to measure the similarity of population structure for the $i^\text{th}$ window and $j^\text{th}$ window. 
+Since 
+$\sum_{ij} (A_{ij}-B_{ij})^2 = \sum_{ij} (A^2_{ij} + B^2_{ij}) - 2 \tr(A^T B)$,
+and $M(i) = V(i) V(i)^T$,
+then due to the orthogonality of eigenvectors and the cyclic invariance of trace,
+$D_{ij}$ can be computed efficiently as
 \begin{align}
-    M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
+    D_{ij} 
+    = 
+    \frac{ \sum_{\ell=1}^k \lambda_{\ell i}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell i})^2 }
+    + \frac{ \sum_{\ell=1}^k \lambda_{\ell j}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell j})^2 }
+    - 2 \sum_{\ell, m=1}^k (V(i)^T V(j))^2_{\ell m} .
 \end{align}
+% from the code for pc_dist():
+% #' In the unweighted case, suppose that (u) and (v) are sets of vectors, and that 
+% #'    A = a_1 * u_1 u_1^T + ... + a_j * u_j u_j^T = U diag(a) U^T
+% #'    B = b_1 * v_1 v_1^T + ... + b_j * v_k v_k^T = V diag(b) V^T,
+% #' where U is the matrix whose columns are (u), and likewise for V.
+% #' Then 
+% #'    (A-B)^T (A-B) = A^T A + B^T B - A^T B - B^T A
+% #' and so
+% #'    ||A-B||^2 = ||A||^2 + ||B||^2 - 2 tr( A^T B )
+% #' By the cyclic invariance of trace, if X = U^T V then
+% #'    tr( A^T B ) = tr( U diag(a) U^T V diag(b) V^T ) = tr( diag(a) X diag(b) X^T ) .
+% \begin{align}
+%     \begin{split}
+%     D_{ij} &= 
+%         \left \{
+%                 \left ( 
+%                     V_{i1}\cdot V_{i1}^{} 
+%                 \right )^{2}
+%                 + \left ( V_{i2}\cdot V_{i2}^{} \right )^{2}
+%                 + \left ( V_{j1}\cdot V_{j1}^{} \right )^{2} 
+%                 +\left ( V_{j2}\cdot V_{j2}^{} \right )^{2} 
+%         \right. \\ 
+%         & \left. \qquad {}  
+%             -2\left [ 
+%                 \left ( V_{i1}\cdot V_{j1}^{} \right )^{2}
+%                 + \left ( V_{i1}\cdot V_{j2}^{} \right )^{2}
+%                 + \left ( V_{i2}\cdot V_{j1}^{} \right )^{2} 
+%                 + \left ( V_{i2}\cdot V_{j2}^{} \right )^{2}
+%             \right ]
+%         \right \}^{1/2}
+%     \end{split}
+% \end{align}
 
-The Euclidean distance $D_{ij}$ between the matrices $M_i$ and $M_j$ stands for the similarity of population structure for the \textit{i}th window and \textit{j}th window. 
-Due to the orthogonality of eigenvectors,we could use the following method to calculate the pairwise distance greatly saving time and space.
+\subsection{Visualization of results}
 
-\plr{Define $D$.}
+We use multidimensional scaling (MDS) to visualize relationships between windows
+as summarized by the dissimilarity matrix $D$,
+that produces a set of $m$ coordinates for each window that
+give the arrangement in $m$-dimensional space that best recapitulates the original distance matrix.
+For results here, we use $m=2$ to produce one- or two-dimensional visualizations of relationships between windows' population structure.
 
-\begin{equation}
-    \begin{aligned}
-        V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} \qquad
-        V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} \\
-    \end{aligned}
-\end{equation}
+We then locate variation in population structure along the genome
+by choosing collections of windows that are nearby in MDS cooridinates,
+and map their positions along the genome.
+Their common population structure is identified by extracting the corresponding genomic regions
+and performing PCA on all, aggregated regions.
 
-\begin{align}
-    M_{i} &= V_{i1}V_{i1}^{T}+V_{i2}V_{i2}^{T}
-\end{align}
+\subsection{Software}
 
-\begin{align}
-    \begin{split}
-    D_{ij} &= 
-        \left \{
-                \left ( 
-                    V_{i1}\cdot V_{i1}^{} 
-                \right )^{2}
-                + \left ( V_{i2}\cdot V_{i2}^{} \right )^{2}
-                + \left ( V_{j1}\cdot V_{j1}^{} \right )^{2} 
-                +\left ( V_{j2}\cdot V_{j2}^{} \right )^{2} 
-        \right. \\ 
-        & \left. \qquad {}  
-            -2\left [ 
-                \left ( V_{i1}\cdot V_{j1}^{} \right )^{2}
-                + \left ( V_{i1}\cdot V_{j2}^{} \right )^{2}
-                + \left ( V_{i2}\cdot V_{j1}^{} \right )^{2} 
-                + \left ( V_{i2}\cdot V_{j2}^{} \right )^{2}
-            \right ]
-        \right \}^{1/2}
-    \end{split}
-\end{align}
+The methods described here
+are implemented in an open-source R package
+available at \url{https://github.com/petrelharp/lostruct},
+as well as scripts to perform all analyses from VCF files
+at various parameter settings.
 
-Using this procedure, 
-we get the pairwise distance matrix that says how similar population structure is 
-in each pair of genomic windows.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Results}
 
-\subsubsection{Visualize the pairwise distance matrix}
-We use Multidimensional scaling (MDS) method to visualize the distance matrices. 
-It can reduce the dimensionality of a distance matrix while preserve the distance information between objects as well as possible. 
-The result is a set of coordinates for each sample with the property that 
-the first M coordinates give the arrangement in M-dimensional space that best recapitulates the original distance matrix.
-We use M=2 to produce one or two dimensional visualization of relationships between windows' population structure.
-
-\section{Results}
 In all these 3 species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.

### 2016-06-06 writeup/local_pca_paper.tex

@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{the-method-diagram}
+       \includegraphics{the-method-diagram}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=1\textwidth]{Fig1_Together_MDS_plot}
+       \includegraphics{Fig1_allchr_Together_MDS_plot_compact}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig2_all_pca_plots}
+       \includegraphics{Fig2_pca_plots_allchr_3peaks_byMDS}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig3}
+       \includegraphics{Fig3_pca_plots_color_by_inv_allchr}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig4_temp}
+       \includegraphics{Fig4_temp}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=0.7\textwidth]{Medicago_MDS_1D_and_gene_count}
+       \includegraphics{Medicago_MDS_1D_and_gene_count}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{medicago_mds}
+       \includegraphics{medicago_mds}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{medicago_peaks_pca}
+       \includegraphics{medicago_peaks_pca}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig6_MDS_1D_against_gene_count_all_chr_win104_with_lm}
+       \includegraphics{Fig6_MDS_1D_against_gene_count_all_chr_win104_with_lm}
     \end{center}

### 2016-06-08 writeup/local_pca_paper.tex
update

@@
     \begin{center}
-       \includegraphics{Fig4_temp}
+       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
     \end{center}
@@
     \begin{center}
-       \includegraphics{Medicago_MDS_1D_and_gene_count}
+       \includegraphics[width=0.9\textwidth]{Fig5_MDS_and_gene_count_allchr}
     \end{center}
@@
     \begin{center}
-       \includegraphics{medicago_mds}
+       \includegraphics[width=0.9\textwidth]{Fig6_Together_MDS_plot_chr3_final}
     \end{center}
@@
     \begin{center}
-       \includegraphics{medicago_peaks_pca}
+       \includegraphics[width=0.9\textwidth]{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
     \end{center}

### 2016-06-14 writeup/local_pca_paper.tex
update

@@
 Dimension reduction techniques,
-such as principal components,
+such as principal component analysis (PCA),
 are often used to discover and display large-scale structure in genomic datasets
@@
 and to control for the confounding effects of population structure in genome-wide association studies.
-The genome-wide mean kinship this uses
+The genome-wide mean kinship that principal component uses
 is an average of the relationships across all locus-specific genealogical trees.
@@
 and apply the method to genomic data from three species.
-In each species we find substantial variation on the scale of megabases to tens of megabases.
+In each species we find that population structure varies on the scale of megabases to tens of megabases.
 In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
-In a dataset of African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
+In a dataset of primarily African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
 is explained by known chromosomal inversions thought to be under recent selection.
@@
 Realized kinship summarizes the shapes
-of the genealogical that relate the samples
+of the genealogical trees that relate the samples
 at each location along the genome.
@@
 and theoretical expectations predict that geographic patterns of relatedness should depend on selection
-\citet{charlesworth2003review}.
+\citep{charlesworth2003review}.
 Nonetheless, it is not generally known to what extent patterns of kinship vary along the genome,
@@
 \citep{novembre2008genes},
-although other patterns emerge if recent migration or nongeographic kinship patterns are important
+although other patterns emerge if recent migration or nongeographic kinship patterns are having a strong role in relatedness determining.
 \citep{astle2009population}.
@@
 before presenting results,
-it may help to have a prior idea of what factors are expected to affect population structure,
+it may be helpful to have a prior idea of what factors are expected to affect population structure,
 and how.
@@
 If individuals are closer in the genealogies of a given genomic region, 
-they tends to be more close in the PCA maps for that region. 
+they tend to be closer in the PCA maps for that region. 
 % Different DNA segments may have different gene tree and therefore different population structure for those segments. 
@@
 background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
-which would tend to generally shorten genealogical trees in the region,
+which would tend to shorten genealogical trees in the region,
 similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
@@
 \paragraph{\textit{Drosophila melanogaster}:}
-We also used whole-genome sequencing data from Drosophila Population Genomics Project (DPGP) and John Pool's lab \citep{lack2015drosophila}, 
+We also used whole-genome sequencing data from Drosophila Population Genomics Project (DPGP) \citep{lack2015drosophila}, 
 which together has 380 samples from 16 countries across Africa and Europe.
@@
 where the sign is chosen according to which of 
-$\sum_i (\pcone_{i1} - u_i)$ or
-$\sum_i (\pcone_{i1} + u_i)$ 
+$\sum_i (\pcone_{i1} - u_i)^{2}$ or
+$\sum_i (\pcone_{i1} + u_i)^{2}$ 
 is smaller.
@@
     \begin{center}
-       \includegraphics{Fig2_pca_plots_allchr_3peaks_byMDS}
+       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
     \end{center}
@@
     \begin{center}
-       \includegraphics{Fig6_MDS_1D_against_gene_count_all_chr_win104_with_lm}
+       \includegraphics[width=0.9\textwidth]{MDS_1D_against_gene_count_all_chr_win104_with_lm}
     \end{center}

### 2016-06-17 writeup/local_pca_paper.tex
update

@@
 
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=1\textwidth]{MDS_1D_win100_all_chr_col_by_chr_with_legend}
+    \end{center}
+    \caption{
+        MDS 1D plot along all 22 autosomes in Human.
+        \label{fig:mds1_along_allchr_human}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr}
+    \end{center}
+    \caption{
+        MDS plots for all 8 chromosomes in \textit{Medicago truncatula}.
+        \label{fig:mds_medicago_allchr}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.8\textwidth]{FigS_pca_plots_for_Medicago_allchr_3peaks_byMDS}
+    \end{center}
+    \caption{
+        PCA plots for all 8 chromosomes' 3 peaks in \textit{Medicago truncatula}.
+        \label{fig:pca_peaks_medicago_allchr}
+    }
+\end{figure}
+
 \end{document}

### 2016-06-22 writeup/local_pca_paper.tex
update

@@
     \begin{center}
-       \includegraphics[width=0.8\textwidth]{FigS_pca_plots_for_Medicago_allchr_3peaks_byMDS}
+       \includegraphics[width=0.82\textwidth]{FigS_pca_plots_for_Medicago_allchr_3peaks_byMDS}
     \end{center}
@@
 
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=1\textwidth]{FigS_pca_plots_allchr_drosophila}
+    \end{center}
+    \caption{
+        PCA plots for chromosome arms 2L, 2R, 3L, 3R and X in \textit{Drosophila}.
+        \label{fig:pca_drosophila_allchr}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human}
+    \end{center}
+    \caption{
+        PCA plots for all 22 autosomes in human.
+        \label{fig:pca_human_allchr}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=1\textwidth]{FigS_pca_plots_medicago_allchr}
+    \end{center}
+    \caption{
+        PCA plots for all 8 chromosomes in \textit{Medicago truncatula}.
+        \label{fig:pca_medicago_allchr}
+    }
+\end{figure}
+
 \end{document}

### 2016-06-23 writeup/local_pca_paper.tex
edited partway through results

@@
 
-In all these 3 species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
-This implies that variation is truly due to population structure, 
+In these three species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
+This implies that variation is due to meaningful variation in population structure, 
 since fluctuations due to demographic noise are not expected to show long distance correlations. 
-Below, we display the results and investigate likely underlying causes.
+Below, we discuss the results and likely underlying causes.
 
+%%%%%%%%%%%
 \subsection{\textit{Drosophila melanogaster}}
-We ran the above method on chromosome arms 2L, 2R, 3L, 3R and X separately. 
-For each, the two-dimensional MDS visualization resembles a triangle. (eg. Figure \ref{fig:mds_chr2L}a) 
-Since the relative position of each window in this plot shows the similarity between windows, this suggests that there are 3 extreme types of population structure shown in the 3 peaks of the ``triangle'', and that other window's population structure might be a mixture of those extremes.
+
+We ran the method on chromosome arms 2L, 2R, 3L, 3R and X separately. 
+For each, the two-dimensional MDS visualization resembles a triangle,
+as seen in Figure \ref{fig:mds_chr2L}.
+Since the relative position of each window in this plot shows the similarity between windows, 
+this suggests that there are at least three extreme types of population structure 
+typified by windows found in the three corners of the triangle,
+and that other windows' population structure may be a mixture of those extremes.
 
@@
     \caption{
-         Variation in structure for windows on chromosome arms 2L, 2R, 3L, 3R and X. 
-         Each represents a window: the first column shows the MDS visualization of relationships between windows. 
-         The second and third columns show the position (midpoint) of each window against the first and second MDS coordinates, respectively. 
-         Rows show results from different chromosome arms, from top to bottom: Chr2L, Chr2R, Chr3L, Chr3R, ChrX. 
+        Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms.
+        In all plots, each point represents one window along the genome.
+         The first column shows the MDS visualization of relationships between windows,
+         and the second and third columns show the position (midpoint) of each window along the genome
+         against the first and second MDS coordinates, respectively. 
+         Rows show results from different chromosome arms, from top to bottom these are 2L, 2R, 3L, 3R, and X. 
          Colors are consistent for plots in each row. 
-         Vertical black lines show the breakpoints of known polymorphic inversions for Chr2L, Chr2R, Chr3L, Chr3R (from top to down: In(2L)t, In(2R)NS, In(3L)OK, In(3R)K).
+         Vertical black lines show the breakpoints of known polymorphic inversions
+         (from top to bottom: In(2L)t, In(2R)NS, In(3L)OK, In(3R)K).
          \label{fig:mds_chr2L}
@@
 
-To investigate these extremes, we pick a window for each extreme, 
-and take out the 5\% of windows that are closest to it in the MDS coordinates, 
-then combine those windows for each extreme and apply PCA on the corresponding sections of genome. 
-We can see in Figure \ref{fig:pca_chr2L} the obvious difference between their PCA plots.
+To investigate these extremes, for each chromosome arm we visually selected three ``extreme'' windows in the MDS plot,
+and take out the 5\% of windows that are closest to it in the MDS coordinates;
+highlighted these windows' positions along the genome (colors in Figure \ref{fig:mds_chr2L}),
+and produced PCA plots for each collection of windows
+(shown in Figure \ref{fig:pca_by_pop}).
 
@@
     \caption{      
-        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_chr2L}. 
-        Each point represents a sample. 
-        The plots are plotting the second principal components against the first principal components. 
-        The first column shows the windows colored green in Figure 1; the second column shows orange windows; and third shows purple windows.
-         \label{fig:pca_chr2L}
+        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_chr2L},
+        on each chromosome arm of \textit{Drosophila melanogaster}.
+        In all plots, each point represents a sample. 
+        The first column shows the combined PCA plot for windows 
+        whose points are colored green in Figure \ref{fig:mds_chr2L}; 
+        the second is for orange windows; 
+        and third is for purple windows.
+         \label{fig:pca_by_pop}
     }
@@
 
-There's a large inversion on Chr2L in that is polymorphic in these samples, In(2L)t \citep{corbett2012population}.
-We recolored the PCA plots in Figure \ref{fig:pca_chr2L} by the orientation of the inversion for each sample.
-Figure \ref{fig:color_inver}
-Two regions of similar, extreme population structure (green in Figure \ref{fig:mds_chr2L}) are found around inversion breakpoints, 
-and the other two extremes occur in the center of the inversion and between the inversion and centromere.
-The corresponding PCA plots show that locally, population structure is mostly determined by which orientation of the inversion each sample has.
-Similar results are found in other chromosome arms that have known polymorphic inversions (Chr2R, Chr3L, Chr3R) in \textit{Drosophila}. 
-Other known polymorphic inversions are like In(3L)P on Chr3L, In(3R)Mo and  In(3R)P on Chr3R. 
-In(3L)P and In(3R)Mo are just a few in our samples, thus we don't see the significant influence of population structure of them. 
-The situation of Chr3R is a little complicated. The coexisting inversions make the MDS visualiazation of it more difficult to pick the extremes.
-For the extremes we picked in Figure \ref{fig:mds_chr2L}, In(3R)K could best explain the PCA clustering result, 
-while picking other windows as extremes might show that In(3R)P explain the PCA clustering result.
+The variation in population structure turns out to be almost entirely explained by
+several large inversions that are polymorphic in these samples, identified in \citet{corbett2012population}. 
+Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
+\plr{Add this table.}
+We recolored the PCA plots in Figure \ref{fig:pca_by_pop} by the orientation of the inversion for each sample,
+obtaining Figure \ref{fig:pca_by_inversion}.
+Taking chromosome arm 2L as an example,
+the two regions of similar, extreme population structure
+shown in green in the first row of Figure \ref{fig:mds_chr2L}
+lie directly around the breakpoints of the iversion In(2L)t,
+and the PCA plots in the first rows of figure \ref{fig:pca_by_inversion}
+shows that population structure here is mostly determined by inversion orientation.
+The regions shown in purple lie near the centromere,
+and have population structure reflective of two axes of variation,
+seen in figure \ref{fig:pca_by_pop},
+which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
+(see \citet{pool2015natural} for more details).
+\plr{check that reference}
+The regions shown in orange mostly lie inside the inversion,
+and show population structure that is a mixture between the other two,
+as expected due to recombination within the (long) inversion \citep{kirkpatrick2015chromosome}.
+\plr{check that reference too}
+Similar results are found in other chromosome arms,
+\plr{add chromosome X}
+except that the situation of Chr3R is a little more complicated,
+due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
+However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
+although we have not highlighted portions of the MDS plot that pick these out.
+Two further inversions, In(3L)P and In(3R)Mo, are at very low frequency in these samples, 
+and do not significantly influence population structure.
 
@@
     \caption{
-         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversion, In(2L)t, In(2R)NS, In(3L)OK and In(3R)K. (Data from \citet{lack2015drosophila}.
+         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
+         In(2L)t, In(2R)NS, In(3L)OK and In(3R)K respectively. 
+         (Data from \citet{lack2015drosophila}.
         \label{fig:color_inver}
@@
 
+%%%%%%%%%%%
 \subsection{Human}
-We ran our method separately on all 22 human autosomes. 
-For instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}b) coincide with the position of known polymorphic inversions on 8p23. 
-Similar results are found in other chromosomes that have known inversions (eg. Chromosome 15, Chromosome 17).
-We found that the primary axis of variation in population structure differentiated only one or a few windows on each chromosome, 
-unlike the countinous variation in population structure we see in Drosophila.
-Other chromosomes showed similar results around predicted inversions: PCA might provide an additional way to identify inversions \citep{ma2012investigation}.
 
-When we run the method on all 22 autosomes together, the outlying signal of chromosome 8 is still visible. (See supplementary for all 22 autosomes)
+As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes, 
+with results shown in Supplementary Figure XXX.
+\plr{Add plots for remaining autosomes to supplement.}
+On each, variation in population structure was dominated by a small number of windows
+having similar population structure to each other that differed dramatically from the rest of the chromosome.
+Such regions seem to often coincide with inversions:
+for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}b) 
+coincide with the position of known polymorphic inversions on 8p23. 
+Similar results are found in other chromosomes that have known inversions,
+shown in Figure \ref{fig:mds_human}.
+PCA plots of these windows show a characteristic trimodal shape,
+presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
+(although we do not have data on orientation status).
+PCA has been previously proposed as a method to computationally identify inversions \citep{ma2012investigation},
+but understanding how well the method can distinguish inversions from regions of low recombination
+is beyond the scope of this paper.
+
+We also ran the method on all 22 autosomes together, 
+and found that, remarkably, 
+the inversion on chromosome 8 is still the most striking outlying signal.
+Further investigation would likely benefit from a denser set of SNPs,
+which would allow a finer genomic resolution.
 
@@
          Variation in structure between windows on human chromosomes 8, 15, and 17. 
-         Each point in the plot represents a window.
+         Each point in each plot represents a window.
          The first column shows the MDS visualization of relationships between windows;
@@
          Rows, from top to bottom show chromosomes 8, 15, and 17. 
-         The vertical red lines show the breakpoints of the known inversions \citep{antonacci2009characterization}.
+         The vertical red lines show the breakpoints of known inversions from \citep{antonacci2009characterization},
+         whose values are listed in \plr{add these in a table}.
         \label{fig:mds_human}
@@
 
-\plr{Check if I've missed anything, then remove these?}
+%%%%%%%%
+\subsection*{Acknowledgements}
 
-\subsection{Future work}
-1.
-For human and Drosophila, we want to eliminate the regions under known inversions and check the variation of population structure for the remaining part by removing those sections.
-We try to check whether they will give similar results as in \textit{Medicago truncatula}, that is whether the variation is closely related to heterochromatin or gene density.
-
-\noindent 2. Uneven sampling has a strong influence on PCA projections \citep{mcvean2009genealogical}. 
-Our human data, POPRES, is unevenly sampled including 346 African-Americas, 73 Asians, 359 Indian Asians and 3187 Europeans. 
-First, we'll try sub-sampling Europeans to balance the population size for the 4 population and repeat the process on the resampled data. 
-Second, we'll try to apply the whole process on only European samples to see the genetic variation inside European samples. 
-Third, we want to try different scheme of adding a weighting matrix to the covariance matrix of genotype data, thus to the reduce the influence of uneven sampling.
-
-\noindent 3. Since regions that have low recombination rate tend to have similar PCs, we'll try cutting the chromosomes into windows with same distance in genetic map instead of same SNP numbers.
-
-\noindent 4. Euclidean distance between the contracted matrix based on PCs is one measure of the similarity for window's population structure.
-We want to try other methods of distance between windows, for example, we used the distance for PCs to reduce noise, however the distance between covariance matrixes of genotype matrix might also be informative.
-
-\noindent 5.
-Although the first two coordinates contains the main part of information, we'd like to see the information contained in higher PCs (e.g.
-the third PC, the forth PC), and higher dimension of MDS.
+We are indebted to John Pool, Russ Corbett-Detig, Matilde Cordeiro, and Peter Chang 
+for assistence with obtaining data and interpreting results
+(especially inversion status of \textit{D.~melanogaster} samples).
+Thanks also go to Yaniv Brandvain and Barbara Engelhardt for helpful comments 
+and for encouraging the project.
 

### 2016-06-23 writeup/local_pca_paper.tex
edited through results

@@
 
-\plr{Add whole-genome PCA plots for comparison (unless they look just like one of the three corners; in which case say so.}
+\plr{Add references to supplemental whole-genome PCA plots in the text.}
 
@@
 the inversion on chromosome 8 is still the most striking outlying signal.
-Further investigation would likely benefit from a denser set of SNPs,
-which would allow a finer genomic resolution.
+Further investigation with a denser set of SNPs,
+allowing a finer genomic resolution,
+may yield other patterns.
 
@@
 
+%%%%%%%%%%%
 \subsection{\textit{Medicago truncatula}}
-We ran our method on all 8 chromosomes of \textit{Medicago truncatula} seperately 
-and found the correlation between each MDS plot and gene density along the genome in all 8 chromosomes. 
-This consistency implies that the factor driving the population structure for each chromosome might be the same, for example, background selection.
-So we ran all chromosomes together, that is, calculating the pairwise distance for all the windows along 8 chromosomes,
-and then get the MDS coordinates for each chromosome by locating the whole MDS coordinates.
-\plr{The results looked different than the other two species, with much less pronounced peaks...}  
-We computed gene density near each window using gene models in Mt4.0 JBrowse \citep{tang2014improved}. 
-The first MDS coordinate value is negatively correlated to the gene count for each window (Figure \ref{fig:mds_gene_count}).
-We found the position of the peak for each MDS plot has a coincidence with the position of heterochromatic regions. 
-This means that population structure in the windows located in heterochromatin tends to have higher similarity, since those windows are closer in MDS plots.
-Biologically, heterochromatic regions have lower gene density and may be less subject to selection \citep{kulikova2001integration,paape2013selection}. 
-Besides, we found the PCA plots are quite consistent for the peaks that more clustering away from the heterochromatic regions, 
-while the PCA plots are barely consistent for the peaks that more clustering around heterochromatic regions for the 8 chromosomes.
-Unlike in the human and Drosophila genomes, in \textit{Medicago truncatula}, major variation in population structure is likely due to linked selection.
+
+Unlike the other two species,
+the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
+showed similar patterns of gradual change in population structure across each chromosome,
+with no indications of chromosome-specific patterns.
+This consistency suggests that the factor driving the population structure for each chromosome is the same,
+as might be caused by varying strengths of linked selection.
+To verify that variation in population structure is shared across chromosomes,
+we applied the method to all chromosomes together.
+Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago}
+(other chromosomes are similar).
+
+Across chromosomes, the high values of MDS coordinate 1 coincide with the position of the heterochromatic regions surrounding the centromere,
+which often have lower gene density and may therefore be less subject to linked selection \citep{kulikova2001integration,paape2013selection}. 
+To verify that this is a possible explanation,
+we computed gene density near each window using gene models in Mt4.0 JBrowse \citep{tang2014improved}. 
+\plr{Say exactly what was computed.}
+These values are shown juxtaposed with
+the first MDS coordinate of each window is shown in Figure \ref{fig:mds_medicago},
+and are significantly correlated, as shown in Figure \ref{fig:mds_gene_count}.
+Other genomic features covary with gene density, however,
+so we cannot rule out alternative hypotheses.
+
+We also found nearly identical results when choosing shorter windows of 1,000 SNPs;
+or choosing windows of equal length in base pairs rather than SNPs.
 
@@
     \caption{
-         MDS results and gene density for the \textit{Medicago} genome (chromosome 1-8). 
+         MDS visualization and gene density for each window in the \textit{Medicago} genome,
+         for chromosomes 1--8 (numbered above each pair of figures).
          For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
@@
     \caption{
-     MDS visualization for Medicago chromosome 3. Each point in the plot stands for a window (length 1000 SNPs).
+     MDS visualization for Medicago chromosome 3. Each point in the plot represents a window.
        \label{fig:mds12_medicago}
@@
     \caption{
-        PCA plots for the three sets of genomic windows colored separately in Figure \ref{fig:mds12_medicago}. 
-        From left to right, the windows colored green, orange, and purple. Each point corresponds to a sample, colored by origin.
+        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
+        Each point corresponds to a sample, colored by country of origin.
         \label{fig:pca_medicago}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{MDS_1D_against_gene_count_all_chr_win104_with_lm}
+       \includegraphics{MDS_1D_against_gene_count_all_chr_win104_with_lm}
     \end{center}
@@
         First MDS coordinate against gene density for all 8 chromosomes. 
-        The first MDS coordinate is significantly correlated with gene count (r=0.146, p=$2.2\times 10^{-16}$). 
+        The first MDS coordinate is significantly correlated with gene count ($r=0.146$, $p=2.2\times 10^{-16}$). 
         (See supplementary for each single chromosome's MDS result against gene count for \textit{Medicago})
@@
 this seems unlikely to give different results.
-\plr{If we check: We did not have this choice to have a substantial effect on results.}
 
@@
 and for encouraging the project.
+\plr{Others?}
 
@@
 
-\subsection{Weighted PCA}
+\section{Weighted PCA}
 

### 2016-06-25 writeup/local_pca_paper.tex
edits in results

@@
 several large inversions that are polymorphic in these samples, identified in \citet{corbett2012population}. 
-Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
-\plr{Add this table.}
+% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
 We recolored the PCA plots in Figure \ref{fig:pca_by_pop} by the orientation of the inversion for each sample,
@@
          (Data from \citet{lack2015drosophila}.
-        \label{fig:color_inver}
+        \label{fig:pca_by_inversion}
     }
@@
 having similar population structure to each other that differed dramatically from the rest of the chromosome.
-Such regions seem to often coincide with inversions:
+These may be primarily inversions:
 for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}b) 
-coincide with the position of known polymorphic inversions on 8p23. 
-Similar results are found in other chromosomes that have known inversions,
+coincide with the position of a particularly large, polymorphic inversion on 8p23. 
+More generally, we see signals coincident with 
+three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
 shown in Figure \ref{fig:mds_human}.
@@
 presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
-(although we do not have data on orientation status).
-PCA has been previously proposed as a method to computationally identify inversions \citep{ma2012investigation},
-but understanding how well the method can distinguish inversions from regions of low recombination
-is beyond the scope of this paper.
+(although we do not have data on orientation status),
+a signal that has been proposed as a method to identify inversions \citep{ma2012investigation}.
+However, distinguishing this hypothesis from others,
+such as regions of low recombination rate,
+would require additional data.
 
@@
          Rows, from top to bottom show chromosomes 8, 15, and 17. 
-         The vertical red lines show the breakpoints of known inversions from \citep{antonacci2009characterization},
-         whose values are listed in \plr{add these in a table}.
+         The vertical red lines show the breakpoints of known inversions from \citep{antonacci2009characterization}.
         \label{fig:mds_human}
@@
 we applied the method to all chromosomes together.
-Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago}
-(other chromosomes are similar).
+Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
+and other chromosomes are similar:
+the two types of population structure with higher gene density (colored green and purple in Figure \ref{fig:mds12_medicago}
+are nearly identical across all chromosomes,
+but population structure associated with low gene density (colored orange)
+shows more chromsome-specific patterns (see Supplemental Figure \ref{fig:pca_drosophila_allchr}).
 
@@
      MDS visualization for Medicago chromosome 3. Each point in the plot represents a window.
+     Plots for remaining chromosomes are shown in Supplemental Figure \ref{fig:mds_medicago_allchr}.
        \label{fig:mds12_medicago}
@@
         Each point corresponds to a sample, colored by country of origin.
+        Plots for remaining chromosomes are shown in Supplemental Figure \ref{fig:pca_peaks_medicago_allchr}.
         \label{fig:pca_medicago}
@@
         PCA plots for all 8 chromosomes' 3 peaks in \textit{Medicago truncatula}.
+        (A) green, (B) orange, and (C) purple.
         \label{fig:pca_peaks_medicago_allchr}

### 2016-06-25 writeup/medicago-maps.R

@@ -1,0 +1,37 @@
+library(maps)
+data(world.cities)
+samples <- read.table("../medicago/data/sample_info.tsv",sep="\t",header=TRUE)
+samples$country <- samples$Country.of.Origin
+levels(samples$country)[ levels(samples$country) ==  "Greece, Crete" ] <- "Greece_Crete"
+levels(samples$country)[ levels(samples$country) ==  "France, Corsica" ] <- "France_Corsica"
+levels(samples$country) <- gsub(",.*","",levels(samples$country))
+levels(samples$country)[ levels(samples$country)=='' ] <- "blank"
+
+countries <- setdiff( sort( levels( samples$country ) ), c("Bulgaria", "China", "Russia", "Uzbekistan") )
+country.cols <- rainbow(length(countries))
+
+other.cities <- c(
+                  "France_Corsica" = "Ajaccio",
+                  "Greece_Crete" = "Malia",
+                  "U.S." = "Los Angeles"
+                  )
+cities <- subset( do.call( rbind, tapply( 1:nrow(world.cities), world.cities$country.etc,
+                                 function (kk) { 
+                                     thisone <- which.max( world.cities$pop[kk] )
+                                     world.cities[kk[thisone],]
+                                 } ) ), tolower(country.etc) %in% tolower(countries) )
+for (k in seq_along(other.cities)) {
+    thisone <- subset( world.cities, name == other.cities[k] )
+    thisone$country.etc <- names(other.cities[k])
+    cities <- rbind( cities, thisone[nrow(thisone),] )
+}
+cities <- cities[ match(countries,cities$country.etc), ]
+
+lims <- structure(c(29.025, 62.085, -12.099, 41.649), .Dim = c(2L, 2L), .Dimnames = list(NULL, c("lat", "long")))
+
+map("world", xlim=lims[,'long'], ylim=lims[,'lat'] )
+points( cities[,"long"], cities[,"lat"], col=country.cols, pch=seq_along(countries), cex=3, lwd=2 )
+text( cities[,"long"], cities[,"lat"], labels=cities$country.etc, col=country.cols, 
+    pos=3, adj=1 )
+
+

### 2016-06-28 writeup/local_pca_paper.tex
update

@@
     \begin{center}
-       \includegraphics{Fig1_allchr_Together_MDS_plot_compact}
+       \includegraphics{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
     \end{center}
@@
 shown in green in the first row of Figure \ref{fig:mds_chr2L}
-lie directly around the breakpoints of the iversion In(2L)t,
+lie directly around the breakpoints of the inversion In(2L)t,
 and the PCA plots in the first rows of figure \ref{fig:pca_by_inversion}
@@
     \begin{center}
-       \includegraphics{Fig3_pca_plots_color_by_inv_allchr}
+       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
     \end{center}

### 2016-06-30 writeup/local_pca_paper.tex
notes

@@
 
+\plr{Peter to add bit about weighted PCA}
+
 The general steps in our method could be carried out in many ways.
@@
 or choosing windows of equal length in base pairs rather than SNPs.
+\plr{Peter add sentence about weighted PCA.}
 
@@
 but downweighting may be a better option that does not discard data.
+\plr{and we have methods to do this}
 Next, the choice of window size may be important,
@@
 this seems unlikely to give different results.
+\plr{we tried this with bp}
 
@@
 (equations XX),
+\plr{(say we have explored)}
 how many MDS coordinates to use when describing the distance matrix between windows,

### 2016-07-07 writeup/local_pca_paper.tex
update

@@
 which together has 380 samples from 16 countries across Africa and Europe.
-\plr{Check: are all of our samples described in \citet{lack2015drosophila}? If yes, just say DPGP; if no, add appropriate citation (ask John what this should be). Also, add website?}
 Since the \textit{Drosophila} samples are from inbred lines, we treat the samples as haploid when recoding;
@@
 where all sums are over the $m_{ij}$ sites where both sample $i$ and sample $j$ have nonmissing genotypes.
-\plr{Is that correct? This is not $X^T X/(m-1)$, which is what was here before, 
-     but is what cov(), use='pairwise' does: see \url{package/tests/testthat/test_covariance.R}.}
 % We compute the covariance matrix using the R function cov() , with use=``pairwise",
@@
 
-\plr{I'm talking about flipping signs up here, even though in the code we do it when computing variances. Do you think it makes sense here?}
 Since eigenvectors are still only defined up to sign,
@@
 Then, the rescaled, $k$-dimensional covariance matrix for the $i^\text{th}$ window is
-\begin{align}
+\begin{align} \label{eqn:est_cov}
     M(i) &= \sum_{\ell=1}^k V_{i\ell} V_{i\ell}^T .
@@
 For each, the two-dimensional MDS visualization resembles a triangle,
-as seen in Figure \ref{fig:mds_chr2L}.
+as seen in Figure \ref{fig:mds_allchr}.
 Since the relative position of each window in this plot shows the similarity between windows, 
@@
     \begin{center}
-       \includegraphics{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
+       \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
     \end{center}
@@
          The first column shows the MDS visualization of relationships between windows, 
-         and the second and third columns show the position (midpoint) of each window along the genome
-         against the first and second MDS coordinates, respectively. 
-         Rows show results from different chromosome arms, from top to bottom these are 2L, 2R, 3L, 3R, and X. 
+         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
+         rows correspond to chromosome arms.
          Colors are consistent for plots in each row. 
-         Vertical black lines show the breakpoints of known polymorphic inversions
-         (from top to bottom: In(2L)t, In(2R)NS, In(3L)OK, In(3R)K).
-         \label{fig:mds_chr2L}
+         Vertical lines show the breakpoints of known polymorphic inversions.   
+         Solid black lines are for the inversions we used to obtaining Figure \ref{fig:pca_by_inversion},
+         while dotted grey lines are for other known inversions.     
+         \label{fig:mds_allchr}
     }
@@
 and take out the 5\% of windows that are closest to it in the MDS coordinates;
-highlighted these windows' positions along the genome (colors in Figure \ref{fig:mds_chr2L}),
-and produced PCA plots for each collection of windows
-(shown in Figure \ref{fig:pca_by_pop}).
+these three groups of windows are shown in color in Figure \ref{fig:mds_allchr} and PCA plots for each collection are shown in Figure \ref{fig:pca_by_pop}.
 
@@
     \caption{      
-        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_chr2L},
+        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_allchr},
         on each chromosome arm of \textit{Drosophila melanogaster}.
@@
         The first column shows the combined PCA plot for windows 
-        whose points are colored green in Figure \ref{fig:mds_chr2L}; 
+        whose points are colored green in Figure \ref{fig:mds_allchr}; 
         the second is for orange windows; 
@@
 the two regions of similar, extreme population structure
-shown in green in the first row of Figure \ref{fig:mds_chr2L}
+shown in green in the first row of Figure \ref{fig:mds_allchr}
 lie directly around the breakpoints of the inversion In(2L)t,
-and the PCA plots in the first rows of figure \ref{fig:pca_by_inversion}
+and the PCA plots in the first rows of Figure \ref{fig:pca_by_inversion}
 shows that population structure here is mostly determined by inversion orientation.
-The regions shown in purple lie near the centromere,
+The regions shown in purple on chromosome 2L lie near the centromere,
 and have population structure reflective of two axes of variation,
-seen in figure \ref{fig:pca_by_pop},
+seen in Figure \ref{fig:pca_by_pop},
 which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
@@
 \plr{check that reference}
-The regions shown in orange mostly lie inside the inversion,
+The regions shown in orange on chromosome 2L mostly lie inside the inversion,
 and show population structure that is a mixture between the other two,
@@
 Similar results are found in other chromosome arms,
-\plr{add chromosome X}
 except that the situation of Chr3R is a little more complicated,
@@
 although we have not highlighted portions of the MDS plot that pick these out.
-Two further inversions, In(3L)P and In(3R)Mo, are at very low frequency in these samples, 
+Two more inversions, In(3L)P and In(3R)Mo, are at very low frequency in these samples, 
 and do not significantly influence population structure.
@@
          As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
-         In(2L)t, In(2R)NS, In(3L)OK and In(3R)K respectively. 
+         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
          (Data from \citet{lack2015drosophila}.
@@
 As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes, 
-with results shown in Supplementary Figure XXX.
-\plr{Add plots for remaining autosomes to supplement.}
+with results shown in Supplementary Figure \ref{fig:mds12_chr1_8_human}, \ref{fig:mds12_chr9_16_human}, \ref{fig:mds12_chr17_22_human}.
 On each, variation in population structure was dominated by a small number of windows
@@
 These may be primarily inversions:
-for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}b) 
+for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}) 
 coincide with the position of a particularly large, polymorphic inversion on 8p23. 
@@
 shown in Figure \ref{fig:mds_human}.
-PCA plots of these windows show a characteristic trimodal shape,
+PCA plots of these windows show a characteristic trimodal shape (Figure \ref{fig:pca_chr8_outliers}),
 presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
@@
 and found that, remarkably, 
-the inversion on chromosome 8 is still the most striking outlying signal.
+the inversion on chromosome 8 is still the most striking outlying signal (Figure \ref{fig:mds1_along_allchr_human}). 
 Further investigation with a denser set of SNPs,
@@
 To verify that this is a possible explanation,
-we computed gene density near each window using gene models in Mt4.0 JBrowse \citep{tang2014improved}. 
-\plr{Say exactly what was computed.}
+we computed gene density for each window using gene models in Mt4.0 JBrowse \citep{tang2014improved}. 
+we downloaded the gene model track from Mt4.0 JBrowse at jcvi.org,
+extracted the start and end positions of each predicted mRNA transcript,
+counted how many of these fell in each window, 
+and use the mean number of the starts and ends in a window to represent the gene count for that window.
+The endpoints of the windows were defined to be halfway between the last SNP of each window and the first SNP in the next window.
 These values are shown juxtaposed with
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig5_MDS_and_gene_count_allchr}
+       \includegraphics[width=0.9\textwidth]{Fig5_MDS_and_gene_count_allchr_update}
     \end{center}
@@
     \begin{center}
-       \includegraphics{MDS_1D_against_gene_count_all_chr_win104_with_lm}
+       \includegraphics{MDS_1D_against_gene_count_all_chr_win104_with_lm_update}
     \end{center}
@@
         First MDS coordinate against gene density for all 8 chromosomes. 
-        The first MDS coordinate is significantly correlated with gene count ($r=0.146$, $p=2.2\times 10^{-16}$). 
+        The first MDS coordinate is significantly correlated with gene count ($r=0.149$, $p=2.2\times 10^{-16}$). 
         (See supplementary for each single chromosome's MDS result against gene count for \textit{Medicago})
@@
 revealing distinct biological processes driving this variation in each species.
-More investigation, particularly on more species and datasets will help uncover which patterns are generalizable.
+More investigation, particularly on more species and datasets, will help to uncover which patterns are generalizable.
 
@@
 Since population structure is not constant along the genome,
-this could in principle lead to an inflation of false positives parts of the genome
+this could in principle lead to an inflation of false positives in parts of the genome
 with stronger population structure than the genome-wide average.
@@
 although in our applications results were not sensitive to this,
-indicating that the limit of resolution was smaller than the scale on which patterns of kinship varies along the genome.
-Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in figure \ref{fig:diagram}),
+indicating that we can see variation on a sufficiently fine scale.
+Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in Figure \ref{fig:diagram}),
 along with the method used to discover common structure,
@@
 Minor points we have not explored include how to decide how many PCs to use in approximating structure of each window
-(equations XX),
+(equations \ref{eqn:est_cov}),
 \plr{(say we have explored)}
@@
 by regressing out the effect of linked SNPs on each other.
-It would be interesting to see if smoehow removing the effects of inversions in the \textit{Drosophila melanogaster} or human datasets
+It would be interesting to see if somehow removing the effects of inversions in the \textit{Drosophila melanogaster} or human datasets
 would produce a pattern similar to that seen in \textit{Medicago truncatula}.
@@
     \begin{center}
+       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr1_8_all_valid_and_pre}
+    \end{center}
+    \caption{
+         MDS plots for human chromosome 1-8.
+         The first column shows the MDS visualization of relationships between windows, 
+         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
+         rows correspond to chromosomes.        
+         Colorful vertical lines show the breakpoints of known valid inversions, 
+         while grey vertical lines show the breakpoints of predicted inversions.
+        \label{fig:mds12_chr1_8_human}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr9_16_all_valid_and_pre}
+    \end{center}
+    \caption{
+        MDS plots for human chromosome 9-16.
+        \label{fig:mds12_chr9_16_human}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr17_22_all_valid_and_pre}
+    \end{center}
+    \caption{
+        MDS plots for human chromosome 17-22.
+        \label{fig:mds12_chr17_22_human}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{FigS_PCA_chr8_outliers_combine}
+    \end{center}
+    \caption{
+        PCA plots comparison for outlier windows and non-outlier windows
+        \label{fig:pca_chr8_outliers}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
        \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr}

### 2016-07-09 writeup/local_pca_paper.tex
fixed up refs; cut down intro drastically

@@
 
-Dimension reduction techniques,
-such as principal component analysis (PCA),
-are often used to discover and display large-scale structure in genomic datasets
-found in the patterns of kinship
-between the genotyped individuals,
-and to control for the confounding effects of population structure in genome-wide association studies.
-The genome-wide mean kinship that principal component uses
-is an average of the relationships across all locus-specific genealogical trees.
-However, many biological factors,
-including linked selection,
-can systematically skew patterns of kinship over intermediate genomic scales.
-We show how to use principal components analysis (PCA) to describe this meso-scale variation in kinship,
-and apply the method to genomic data from three species.
-In each species we find that population structure varies on the scale of megabases to tens of megabases.
+Dimension reduction techniques such as principal component analysis (PCA)
+are often used to discover and visualize the patterns 
+of mean kinship between individuals in large genomic datasets,
+known as population structure.
+% Modeling such kinship is key to controling the confounding effects of population structure in genome-wide association studies.
+Genome-wide mean kinship
+is an average of the relationships across all locus-specific genealogical trees,
+which can be strongly affected on intermediate genomic scales by many biological factors,
+such as linked selection.
+We show how to use local principal components analysis to describe this meso-scale variation,
+and apply the method to genomic data from three species,
+finding in each that population structure varies on the scale of megabases to tens of megabases.
 In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
@@
 and may be caused by background selection or local adaptation.
-The method is a useful addition to the exploratory toolbox
-of population genomics,
-and is implemented as an R package.
+The method is implemented as an R package, called lostruct.
 
@@
 
-The phrase ``population structure'' refers to reduced gene flow between subpopulations, 
-often because of geographical isolation.
+The phrase ``population structure'',
+used to describe patterns in genome-wide mean kinship,
+calls to mind to reduced gene flow between subpopulations. %, often because of geographical isolation.
 However, it is widely recognized that because of selection the effects of gene flow are not equal everywhere on the genome,
 and patterns of polymorphism and divergence can vary significantly depending on factors including local gene density.
-This implies that, paradoxically, 
+This implies that, counterintuitively, 
 the population structure of a species depends on which part of the genome is being examined.
 
-Population structure leads to systematic patterns in genome-wide mean kinship,
-and so it is said that visualizations of kinship depict population structure,
-rather than the \emph{effects} of population structure (which would be more accurate).
-The \emph{kinship coefficient} for a pair of individuals
-gives the expected proportion of their genome
-that the two have inherited identically by descent;
-for a single individual it is the inbreeding coefficient.
-However, as \citet{wright1949genetical} wrote,
-``It has probably occurred to the reader that the coefficient of inbreeding
-may mean very different things in different cases.''
-The kinship coefficient originally referred to the expected probability
-of coinheritance within a given pedigree;
-while modern applications to ``unrelated'' individuals
-use a genetic covariance matrix to estimate the \emph{realized} proportion of the genome
-coinherited from sufficiently recent ancestors.
+% Population structure leads to systematic patterns in genome-wide mean kinship,
+% and so it is said that visualizations of kinship depict population structure,
+% rather than the \emph{effects} of population structure (which would be more accurate).
+% The \emph{kinship coefficient} for a pair of individuals
+% gives the expected proportion of their genome
+% that the two have inherited identically by descent;
+% for a single individual it is the inbreeding coefficient.
+% However, as \citet{wright1949genetical} wrote,
+% ``It has probably occurred to the reader that the coefficient of inbreeding
+% may mean very different things in different cases.''
+% The kinship coefficient originally referred to the expected probability
+% of coinheritance within a given pedigree;
+% while modern applications to ``unrelated'' individuals
+% use a genetic covariance matrix to estimate the \emph{realized} proportion of the genome
+% coinherited from sufficiently recent ancestors.
 
 Realized kinship summarizes the shapes
-of the genealogical trees that relate the samples
+of the genealogical trees % that relate the samples
 at each location along the genome.
@@
 and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
-\citep{maynardsmith1974hitchhiking,barton2000genetic,neher2013genetic,charlesworth2012effects},
-and therefore patterns of kinship.
+\citep{maynardsmith1974hitchhiking,barton2000genetic,neher2013genetic,charlesworth2012effects}.
 Indeed,
-chromosome-scale variation in diversity and divergence has been observed in many species
-\citep[e.g.][]{langley2012genomic};
+many species show chromosome-scale variation in diversity and divergence
+\citep[sometimes dramatic:][]{langley2012genomic};
 species phylogenies can differ along the genome 
-due to incomplete lineage sorting \citep[e.g.][]{pease2013accurate},
-adaptive introgression and/or local adaptation \citep[e.g.][]{ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
+due to incomplete lineage sorting \citep[e.g.,][]{pease2013accurate},
+adaptive introgression and/or local adaptation \citep[e.g.,][]{ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
 and theoretical expectations predict that geographic patterns of relatedness should depend on selection
@@
 Nonetheless, it is not generally known to what extent patterns of kinship vary along the genome,
-nor what the major axes of variation are.
+nor for what reasons.
 
 Patterns in genome-wide kinship are often summarized
-by applying principal components analysis (PCA) \citep{patterson2006population} 
+by applying principal components analysis \citep[PCA,][]{patterson2006population} 
 to the genetic covariance matrix,
@@
 % Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.  
-% PCA results are derived from the covariance matrix of genotype matrix. 
-The results of PCA can be directly related to the underlying genealogical history of the samples, 
-such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}. 
-These patterns are often geographical, producing ``maps'' of population structure
+The results of PCA can be related to the genealogical history of the samples, 
+such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}, 
+and sometimes produce ``maps'' of population structure
 that reflect the samples' geographic origin distorted by rates of gene flow
-\citep{novembre2008genes},
-although other patterns emerge if recent migration or nongeographic kinship patterns are having a strong role in relatedness determining.
-\citep{astle2009population}.
+\citep{novembre2008genes}.
+% although other patterns can be caused by recent migration or nongeographic kinship patterns.  % \citep{astle2009population}.
 % Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
@@
 
-As reviewed by \citet{astle2009population}, 
+Furthermore,
 modeling ``background'' kinship between samples
-is essential to successful genome-wide association studies,
-and PCA has often been used for stratification correction \citep{price2006principal},
-and so investigating variation in kinship along the genome can help us to have a
-better understanding of the relation between genome structure 
-and population structure, possibly leading to more powerful methods for association studies.
-
-To investigate how population structure varies along the genome in several datasets, 
-we cut each genome into windows (with hundreds to thousands of SNPs in each), 
-applied PCA to each window, 
-and visualized the major ways that population structure, as summarized by PCA, varies among windows.
-% In this project, we used SNP data for human, \textit{Medicago truncatula}, and
-whole genome sequencing data for \textit{Drosophila}.  
-% Based on the principal components, we can estimate the similarity of population structure contained in each genome window.  
-To quantify similarity of population structure between
-windows, we constructed for each window an approximate, scaled covariance
-matrix based on the first few principal components, and measured the pairwise Euclidean distance
-between those matrices.  
-We then use multidimensional scaling to visualize the
-relationships between windows, which reduces the pairwise distance matrix to
-lower dimension while preserving the distance information between windows as
-well as possible \citep{borg2005modern}.
-
-% To interpret the results of MDS, we combine known genome feature information for each species, such as the
-% distribution of inversions, and heterochromatin and gene density along the genome. 
-Each species showed distinct patterns, reflecting differences in their biology;
-before presenting results,
-it may be helpful to have a prior idea of what factors are expected to affect population structure,
-and how.
-PCA summarizes patterns in kinship found in the genetic covariance matrix,
-which is an average across locus-specific genealogies.
-% However, different parts of genome have different genetic features. 
-% First, each site of DNA may have different gene tree. 
-% The covariance matrix of genotype matrix averages those gene trees. 
-If individuals are closer in the genealogies of a given genomic region, 
-they tend to be closer in the PCA maps for that region. 
-% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
-% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
-% Selective sweeps cause local recent ancestry or short trees.
-% Background selection causes shallow ones.
-Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
-characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
-A genomic region with many targets for selection experiences
-background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
-which would tend to shorten genealogical trees in the region,
-similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
-% Balancing selection causes deep trees.
-On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
-and population structure locally describes which individuals have which alleles
-rather than geographical proximity.
-Finally, 
-since recombination is suppressed between opposite orientations of a chromosomal inversion
-near its breakpoints,
-genealogies in these regions separate samples carrying the two orientations of the inversion.
-In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
-(indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
-as a way to identify inversions).
-% if a chromosomal inversion is polymorphic in the sample, 
-% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints.
-Recombination suppression across inversions thus results in different genome structure and population structure.
-% Other effects, like noise, introgression might also influence population structure.
-We will look for strong variation in population structure shared across large regions of the genome.
-Many of the effects listed above (such as single selective sweeps or inversions) 
-are not expected to have similar effects on population structure in different regions of the genome
-because of randomness in which samples end up in which group;
-but if these coincide with a region of reduced recombination (such as an inversion),
-these could drive major patterns of variation.
-The more subtle effects of genome-wide linked selection could be shared across large regions
-due to large-scale variation in gene density
-(as by background selection or local adaptation at many genes).
-
-We chose PCA to summarize population structure,
-but other methods, such as STRUCTURE \citep{falush2003inference},
-SPA \citep{yang2012modelbased},
-SpaceMix \citep{bradburd2015spatial},
-or other matrix factorization methods \citep{engelhardt2010analysis}
-would highlight different sources of variation in the data.
-% There are other methods for visualizing population structure are like STRUCTURE,
-% \citep{pritchard2000inference,falush2003inference, falush2007inference,hubisz2009inferring} 
-% model-based approach,
-% maps of heterozygosity
-% \citep{ramachandran2005support}.
-
-Finally, a number of general methods for dimensionality reduction also use a strategy of ``local PCA''
-\citep[e.g.][]{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear},
+is essential to genome-wide association studies \citep[GWAS,][]{price2006principal,astle2009population},
+% and PCA has often been used in this context \citep{price2006principal}.
+and so understanding variation in kinship along the genome 
+% can help us to have a better understanding of the relation between genome structure and population structure, 
+could lead to more generally powerful methods,
+and is essential for doing GWAS in species with substantial variation in population structure,
+
+%% summary of the method: this happens in the method
+% To investigate how population structure varies along the genome in several datasets, 
+% we cut each genome into windows (with hundreds to thousands of SNPs in each), 
+% applied PCA to each window, 
+% and visualized the major ways that population structure, as summarized by PCA, varies among windows.
+% % In this project, we used SNP data for human, \textit{Medicago truncatula}, and
+% % whole genome sequencing data for \textit{Drosophila}.  
+% % Based on the principal components, we can estimate the similarity of population structure contained in each genome window.  
+% To quantify similarity of population structure between
+% windows, we constructed for each window an approximate, scaled covariance
+% matrix based on the first few principal components, and measured the pairwise Euclidean distance
+% between those matrices.  
+% We then use multidimensional scaling to visualize the
+% relationships between windows, which reduces the pairwise distance matrix to
+% lower dimension while preserving the distance information between windows as
+% well as possible \citep{borg2005modern}.
+
+In this work we describe variation in population structure using local PCA,
+where ``local'' refers to proximity along the genome.
+A number of general methods for dimensionality reduction also use a strategy of ``local PCA''
+\citep[e.g.,][]{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear},
 performing PCA not on the entire dataset but instead on subsets of observations,
-providing ``local'' pictures which are then stitched back together to give a global picture.
+providing local pictures which are then stitched back together to give a global picture.
 At first sight,
-this differs from our use of the term in that we restrict to subsets of \emph{variables} instead of subsets of observations.
+this differs from our method in that we restrict to subsets of \emph{variables} instead of subsets of observations.
 However, if we flip perspectives and think of each genetic variant as an observation,
@@
 \section{Methods}
-
-\plr{Peter to add bit about weighted PCA}
 
@@
 \paragraph{\textit{Drosophila melanogaster}:}
-We also used whole-genome sequencing data from Drosophila Population Genomics Project (DPGP) \citep{lack2015drosophila}, 
-which together has 380 samples from 16 countries across Africa and Europe.
+We also used whole-genome sequencing data 
+from the Drosophila Genome Nexus \citep[\url{http://www.johnpool.net/genomes.html}][]{lack2015drosophila},
+consisting of the Drosophila Population Genomics Project phases 1--3 \citep[DPGP][]{langley2012genomic,pool2012population},
+and additional African genomes \citep[AGES][]{lack2015drosophila},
+After removing 20 genomes with more than 8\% missing data,
+we were left with 380 samples from 16 countries across Africa and Europe.
 Since the \textit{Drosophila} samples are from inbred lines, we treat the samples as haploid when recoding;
-regions with residual heterozygosity were marked as missing in the original dataset.
-Each chromosome arm we investigated (Chr2L, Chr2R, Chr3L, and Chr3R) has 2--3 million SNPs.
-Due to high density of missing data for some parts in the genome, 
-we first deleted any samples with more than 8\% missing genotypes, 
-and then deleted positions with more than 20\% missing data. 
-(We chose these cutoffs as the tails of the relevant empirical distributions.)
+regions with residual heterozygosity were marked as missing in the original dataset;
+we also removed positions with more than 20\% missing data. 
+% (We chose these cutoffs as the tails of the relevant empirical distributions.)
+Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs.
+
+% lack2015drosophila: describes the DGN, including the AGES samples.  reports diversity and mentions inversions but mostly a 'data' paper
+% pool2012population: describes 129 subsaharan samples (DPGP3)
 
@@
 is smaller.
+
+Several of the datasets we use have unbalanced representations of diverged populations,
+which can have a strong impact on the results of PCA.
+(The principal axes may describe variation \emph{within} an overrepresented group
+rather than more significant variation between groups.)
+Therefore, to check the effect of sampling on our results,
+we used a variant of PCA that gives roughly equal weight to each group of samples,
+rather than to each sample.
+The rationale and implementation of this method are described in Appendix \ref{apx:weighted_pca}.
 
@@
 where $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
-We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife}:
+We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}:
 we divide the $j^\text{th}$ block into 10 equal-sized pieces,
@@
 
+\plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years, probably why it doesn't show up. 
+Also cite \citet{corbettdetig2012sequencebased} for a computational method for detecting inversions.}
+% corbettdetig2012sequencebased suggests these inversions are recent-ish (but before out-of-Africa)
+
 \begin{figure}
@@
 which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
-(see \citet{pool2015natural} for more details).
-\plr{check that reference}
+(see \citet{lack2015drosophila} for more about admixture in this sample).
 The regions shown in orange on chromosome 2L mostly lie inside the inversion,
 and show population structure that is a mixture between the other two,
-as expected due to recombination within the (long) inversion \citep{kirkpatrick2015chromosome}.
-\plr{check that reference too}
+as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
 Similar results are found in other chromosome arms,
@@
          In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
-         (Data from \citet{lack2015drosophila}.
+         \citep[Data from][.]{lack2015drosophila}
         \label{fig:pca_by_inversion}
@@
 or choosing windows of equal length in base pairs rather than SNPs.
-\plr{Peter add sentence about weighted PCA.}
+Similarly, the results were not substantially changed
+when using weighted PCA to downweight the large group of Tunisian samples.
 
@@
 has been one of the major sources of confounding in genome-wide association studies (GWAS)
-and so methods must account for it by modeling population structure or kinship \citep{gwas_confounding_review,mixed_models}.
+and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
 Since population structure is not constant along the genome,
@@
 
+\paragraph{The effect of selection}
+\plr{intro sentence}
+PCA summarizes patterns in kinship found in the genetic covariance matrix,
+which is an average across locus-specific genealogies.
+% However, different parts of genome have different genetic features. 
+% First, each site of DNA may have different gene tree. 
+% The covariance matrix of genotype matrix averages those gene trees. 
+If individuals are closer in the genealogies of a given genomic region, 
+they tend to be closer in the PCA maps for that region. 
+% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
+% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
+% Selective sweeps cause local recent ancestry or short trees.
+% Background selection causes shallow ones.
+Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
+characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
+A genomic region with many targets for selection experiences
+background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
+which would tend to shorten genealogical trees in the region,
+similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
+% Balancing selection causes deep trees.
+On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
+and population structure locally describes which individuals have which alleles
+rather than geographical proximity.
+Finally, 
+since recombination is suppressed between opposite orientations of a chromosomal inversion
+near its breakpoints,
+genealogies in these regions separate samples carrying the two orientations of the inversion.
+In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
+(indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
+as a way to identify inversions).
+% if a chromosomal inversion is polymorphic in the sample, 
+% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints.
+Recombination suppression across inversions thus results in different genome structure and population structure.
+% Other effects, like noise, introgression might also influence population structure.
+We will look for strong variation in population structure shared across large regions of the genome.
+Many of the effects listed above (such as single selective sweeps or inversions) 
+are not expected to have similar effects on population structure in different regions of the genome
+because of randomness in which samples end up in which group;
+but if these coincide with a region of reduced recombination (such as an inversion),
+these could drive major patterns of variation.
+The more subtle effects of genome-wide linked selection could be shared across large regions
+due to large-scale variation in gene density
+(as by background selection or local adaptation at many genes).
+
 \paragraph{Parameter choices}
@@
 The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
-but downweighting may be a better option that does not discard data.
-\plr{and we have methods to do this}
+but downweighting may be a better option that does not discard data
+(and here we present a method to do this).
 Next, the choice of window size may be important,
@@
 so that each would have roughly the same amount of phylogenetic information.
-However, given the insensitivity of our results to window length,
-this seems unlikely to give different results.
-\plr{we tried this with bp}
+However, we found little change in results when using different window sizes
+or when measuring windows in physical distance (in bp),
+so this seems unlikely to give different results.
 
@@
 would lead to different insights.
-Minor points we have not explored include how to decide how many PCs to use in approximating structure of each window
-(equations \ref{eqn:est_cov}),
-\plr{(say we have explored)}
-how many MDS coordinates to use when describing the distance matrix between windows,
-or how to choose interesting regions of the genome when the MDS plot is not triangular.
+Our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
+and how many MDS coordinates to use when describing the distance matrix between windows,
+but in our exploration, changing these has not produced dramatically different results.
 These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
 we encourage the user to experiment.
+
+Finally,
+in this work we use PCA to summarize population structure,
+but other methods, such as STRUCTURE \citep{falush2003inference},
+SPA \citep{yang2012modelbased},
+SpaceMix \citep{bradburd2015spatial},
+or other matrix factorization methods \citep{engelhardt2010analysis}
+may highlight different sources of variation in the data.
 
@@
 \section{Weighted PCA}
-
-PCA can be thought of as finding a good low-dimensional matrix factorization \citep{engelhardt2010analysis}
+\label{apx:weighted_pca}
+
+Principal components analysis can be thought of as finding a good low-dimensional matrix factorization \citep{engelhardt2010analysis}
 that well-approximates the original data in the least-squares sense:

### 2016-07-09 writeup/local_pca_paper.tex
done editing

@@
 The general steps in our method could be carried out in many ways.
-Referring to Figure \ref{fig:diagram}, these are
+As depicted in Figure \ref{fig:diagram}, these are
 (1) divide the genome into windows,
@@
 We used three publically available datasets (summarized in Table \ref{tab:data_stats});
-as usual for genetic data we converted the data to a numeric matrix
+we converted each to a numeric matrix
 (with one row per polymorphic variant and one column per sample)
@@
 
-\paragraph{Human:}
-We used genomic data from the entire POPRES dataset \citep{nelson2008population},
-which has array-derived genotype information for 447,267 SNPs across the autosomes
-of 3,965 samples in total: 346 African-Americas, 73 Asians, 3,187 Europeans and 359 Indian Asians.
-(We excluded the sex chromosomes and the mitochondria.)
-We use the allele that has highest frequency in the samples as the reference allele for each position.
-
 \paragraph{\textit{Drosophila melanogaster}:}
-We also used whole-genome sequencing data 
-from the Drosophila Genome Nexus \citep[\url{http://www.johnpool.net/genomes.html}][]{lack2015drosophila},
-consisting of the Drosophila Population Genomics Project phases 1--3 \citep[DPGP][]{langley2012genomic,pool2012population},
-and additional African genomes \citep[AGES][]{lack2015drosophila},
+We used whole-genome sequencing data 
+from the Drosophila Genome Nexus \citep[\url{http://www.johnpool.net/genomes.html},][]{lack2015drosophila},
+consisting of the Drosophila Population Genomics Project phases 1--3 \citep{langley2012genomic,pool2012population},
+and additional African genomes \citep{lack2015drosophila}.
 After removing 20 genomes with more than 8\% missing data,
 we were left with 380 samples from 16 countries across Africa and Europe.
-Since the \textit{Drosophila} samples are from inbred lines, we treat the samples as haploid when recoding;
+Since the \textit{Drosophila} samples are from inbred lines or haploid embryos, 
+we treat the samples as haploid when recoding;
 regions with residual heterozygosity were marked as missing in the original dataset;
@@
 % (We chose these cutoffs as the tails of the relevant empirical distributions.)
-Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs.
+Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs;
+PCA plots for each arm are shown in Figure \ref{fig:pca_drosophila_allchr}.
+
+\paragraph{Human:}
+We also used genomic data from the entire POPRES dataset \citep{nelson2008population},
+which has array-derived genotype information for 447,267 SNPs across the 22 autosomes
+of 3,965 samples in total: 346 African-Americas, 73 Asians, 3,187 Europeans and 359 Indian Asians.
+Since these data derive from genotyping arrays, the SNP density is much lower than the other datasets,
+which are each derived from whole genome sequencing.
+We excluded the sex chromosomes and the mitochondria.
+PCA plots for each chromosome, separately, are shown in Figure \ref{fig:pca_human_allchr}.
+% We use the allele that has highest frequency in the samples as the reference allele for each position.
 
@@
 primarily distributed around the Mediterranean basin.
-Each of the 8 chromosomes has 3-5 million SNPs;
+Each of the 8 chromosomes has 3--5 million SNPs;
+PCA plots for these are shown in Figure \ref{fig:pca_medicago_allchr}.
 we did not use the mitochondria or chloroplasts.
@@
   \hline
+  \textit{Drosophila melanogaster} & 1,000 & 9,019 & 2,674 & 0.53 \\ 
   Human & 100 & 636,494 & 203 & 0.55 \\ 
-  \textit{Drosophila melanogaster} & 1,000 & 9,019 & 2,674 & 0.53 \\ 
   \textit{Medicago truncatula} & 10,000 & 102,580 & 467 & 0.50 \\ 
@@
 \caption{
-    Descriptive statistics for each dataset used,
-    with the chosen window sizes.
-    See text for how window sizes were chosen;
-    in each window we performed PCA,
-    computed the percent varince explained by the top two principal components;
-    the mean across windows is given in the last column.
+    Descriptive statistics for each dataset used.
+    % with the chosen window sizes.
+    % See text for how window sizes were chosen;
+    Columns 2--4 give statistics describing the window sizes used for most analyses for each dataset
+    (windows had a fixed number of SNPs).
+    The final column gives the percent variance explained by the top two principal components,
+    averaged across independent PCA of each window.
     \label{tab:data_stats}
@@
 separately to the submatrices that corresponded to each window.
-Specifically, we did PCA as follows:
+% Specifically, we did PCA as follows:
+Precisely,
 denote by $Z$ the $L\times N$ recoded genotype matrix ($L$ is the number of SNPs and $N$ is the sample size), 
@@
 Since eigenvectors are still only defined up to sign,
-when comparing between windows we choose the sign to best match each other:
-after choosing $\pcone_1$, for instance, and $u$ is the first eigenvector obtained from the covariance matrix
+when comparing between windows we choose signs to best match each other:
+after choosing $\pcone_1$, for instance, 
+if $u$ is the first eigenvector obtained from the covariance matrix
 for window $j$,
-then $\pcone_j = \pm u$,
+then we next choose $\pcone_j = \pm u$,
 where the sign is chosen according to which of 
-$\sum_i (\pcone_{i1} - u_i)^{2}$ or
-$\sum_i (\pcone_{i1} + u_i)^{2}$ 
+$\| \pcone_{1} - u \|$ or
+$\| \pcone_{1} + u \|$ 
 is smaller.
@@
 
-The choice of window length entails a balance between several factors.
+The choice of window length entails a balance between signal and noise.
 In very short windows, genealogies of the samples will only be represented by a few trees,
@@
 we would like reasonably short windows.
-Window length choice therefore entails a signal versus noise tradeoff
-in the estimates of population structure.
+% Window length choice therefore entails a signal versus noise tradeoff
+% in the estimates of population structure.
 % If we use the first principal component as a measure of population structure, 
@@
 % and the standard deviation between windows.
+
 Since we summarize population structure using relative positions in the principal component maps,
@@
 \end{align*}
-where $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
+where $\pcone_{ij}$ is the position of the $i^\text{th}$ individual on $\pcone$ in window $j$,
+and $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
 We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}:
-we divide the $j^\text{th}$ block into 10 equal-sized pieces,
+we divide the $j^\text{th}$ window into 10 equal-sized pieces,
 and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
 then the estimate of the squared standard error is
-$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$,
-and we measure noise by averaging over samples and windows:
+$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$.
+Averaging over samples and windows,
 \begin{align*}
@@
 
-We calculated $\sigma^2_\text{signal}$ and $\sigma^2_\text{noise}$
-for a range of window sizes,
-and for our main results
-chose the smallest window for which $\sigma^2_\text{signal}$ was consistently larger than $\sigma^2_\text{noise}$;
-the values for various window sizes across \textit{Drosophila} chromosomes are shown in Table \ref{tab:window_sizes},
-and the choices for all taxa are in Table \ref{tab:data_stats}.
-In practice, even substantially different window sizes result in the same genome-wide patterns.
-We also chose windows to each consist of the same number of neighboring SNPs.
-We compared to the results when windows are each of the same physical length (in bp) but of the same average length,
-and results were nearly identical.
+For the main analysis, we defined windows to each consist of the same number of neighboring SNPs,
+and calculated $\sigma^2_\text{signal}$ and $\sigma^2_\text{noise}$
+for a range of window sizes (i.e., numbers of SNPs).
+For our main results we
+chose the smallest window for which $\sigma^2_\text{signal}$ was consistently larger than $\sigma^2_\text{noise}$ (but checked other sizes);
+the values for various window sizes across \textit{Drosophila} chromosomes are shown in Table \ref{tab:window_sizes}.
+% and the choices for all taxa are in Table \ref{tab:data_stats}.
+% In practice, even substantially different window sizes result in the same genome-wide patterns.
+In the cases we examined, we found nearly identical results after varying window size,
+and choosing windows to be of the same physical length (in bp) rather than in numbers of SNPs.
 
@@
 
-We think of local population structure as being summarized by \emph{relative} position of the sampmles
+We think of local population structure as being summarized by \emph{relative} position of the samples
 in the space defined by the top principal components.
@@
 we also rescale each covariance matrix so that the underlying data matrix has trace norm equal to one.
-To do this, define the $N \times k$ matrix $V(i)$ so that the $\ell^\text{th}$ column of $V(i)$
+
+To do this, define the $N \times k$ matrix $V(i)$ so that $V(i)_{\cdot \ell}$, 
+the $\ell^\text{th}$ column of $V(i)$,
 is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
-multiplied by $\sqrt{ \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} }$.
+multiplied by $\sqrt{ \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} }$,
+where $\lambda_{\ell i}$ is the $\ell^\text{th}$ eigenvalue of the genetic covariance matrix.
 % \begin{equation}
@@
 % \end{equation}
-Then, the rescaled, $k$-dimensional covariance matrix for the $i^\text{th}$ window is
+Then, the rescaled, rank $k$ approximate covariance matrix for the $i^\text{th}$ window is
 \begin{align} \label{eqn:est_cov}
-    M(i) &= \sum_{\ell=1}^k V_{i\ell} V_{i\ell}^T .
+    M(i) &= \sum_{\ell=1}^k V(i)_{\cdot \ell} V(i)_{\cdot \ell}^T .
 \end{align}
@@
 We use multidimensional scaling (MDS) to visualize relationships between windows
-as summarized by the dissimilarity matrix $D$,
-that produces a set of $m$ coordinates for each window that
+as summarized by the dissimilarity matrix $D$.
+MDS produces a set of $m$ coordinates for each window that
 give the arrangement in $m$-dimensional space that best recapitulates the original distance matrix.
@@
 We then locate variation in population structure along the genome
-by choosing collections of windows that are nearby in MDS cooridinates,
+by choosing collections of windows that are nearby in MDS coordinates,
 and map their positions along the genome.
-Their common population structure is identified by extracting the corresponding genomic regions
-and performing PCA on all, aggregated regions.
+Their common population structure is visualized by extracting the corresponding genomic regions
+and performing PCA on all, aggregated, regions.
 
@@
          Vertical lines show the breakpoints of known polymorphic inversions.   
-         Solid black lines are for the inversions we used to obtaining Figure \ref{fig:pca_by_inversion},
+         Solid black lines are for the inversions we used in Figure \ref{fig:pca_by_inversion},
          while dotted grey lines are for other known inversions.     
@@
 The variation in population structure turns out to be almost entirely explained by
-several large inversions that are polymorphic in these samples, identified in \citet{corbett2012population}. 
+several large inversions that are polymorphic in these samples, 
+disussed in \citet{corbett2012population} and \citet{langley2012genomic}.
 % Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
-We recolored the PCA plots in Figure \ref{fig:pca_by_pop} by the orientation of the inversion for each sample,
-obtaining Figure \ref{fig:pca_by_inversion}.
+To depict this, Figure \ref{fig:pca_by_inversion} shows
+the PCA plots in Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
 Taking chromosome arm 2L as an example,
@@
 Similar results are found in other chromosome arms,
-except that the situation of Chr3R is a little more complicated,
-due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
-However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
-although we have not highlighted portions of the MDS plot that pick these out.
-Two more inversions, In(3L)P and In(3R)Mo, are at very low frequency in these samples, 
-and do not significantly influence population structure.
+albeit complicated by the coexistence of more than one polymorphic inversion;
+however, each breakpoint visibly affects patterns in the MDS coordinates.
+% except that the situation of Chr3R is a little more complicated,
+% due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
+% However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
+% although we have not highlighted portions of the MDS plot that pick these out.
+% The inversions In(3L)P and In(3R)Mo, are at low frequency in these samples, 
+% and visibly affect the MDS plots although less than the more common inversion on each arm.
+% The very young inversion In(X)be shows little signal 
+% compared to the older inversion In(X)A \citep{corbett2012population}.
 
@@
          In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
-         \citep[Data from][.]{lack2015drosophila}
+         \citep[data from][]{lack2015drosophila}
         \label{fig:pca_by_inversion}
@@
 
-As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes, 
-with results shown in Supplementary Figure \ref{fig:mds12_chr1_8_human}, \ref{fig:mds12_chr9_16_human}, \ref{fig:mds12_chr17_22_human}.
+As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes.
 On each, variation in population structure was dominated by a small number of windows
 having similar population structure to each other that differed dramatically from the rest of the chromosome.
-These may be primarily inversions:
-for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}) 
-coincide with the position of a particularly large, polymorphic inversion on 8p23. 
-More generally, we see signals coincident with 
+These may be primarily inversions: outlying windows coincide with
 three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
-shown in Figure \ref{fig:mds_human}.
-PCA plots of these windows show a characteristic trimodal shape (Figure \ref{fig:pca_chr8_outliers}),
+notably a particularly large, polymorphic inversion on 8p23 (Figure \ref{fig:mds_human}). 
+% for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}) 
+Similar plots for all chromosomes are shown
+in Supplementary Figures \ref{fig:mds12_chr1_8_human}, \ref{fig:mds12_chr9_16_human}, \ref{fig:mds12_chr17_22_human}.
+PCA plots of many outlying windows show a characteristic trimodal shape 
+(shown for chromosome 8 in Figure \ref{fig:pca_chr8_outliers}),
 presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
-(although we do not have data on orientation status),
-a signal that has been proposed as a method to identify inversions \citep{ma2012investigation}.
-However, distinguishing this hypothesis from others,
+(although we do not have data on orientation status).
+This trimodal shape has been proposed as a method to identify inversions \citep{ma2012investigation},
+but distinguishing this hypothesis from others,
 such as regions of low recombination rate,
@@
 
-We also ran the method on all 22 autosomes together, 
+We also ran the method grouping all 22 autosomes together, 
 and found that, remarkably, 
@@
          Rows, from top to bottom show chromosomes 8, 15, and 17. 
-         The vertical red lines show the breakpoints of known inversions from \citep{antonacci2009characterization}.
+         The vertical red lines show the breakpoints of known inversions from \citet{antonacci2009characterization}.
         \label{fig:mds_human}
@@
 and other chromosomes are similar:
-the two types of population structure with higher gene density (colored green and purple in Figure \ref{fig:mds12_medicago}
-are nearly identical across all chromosomes,
-but population structure associated with low gene density (colored orange)
-shows more chromsome-specific patterns (see Supplemental Figure \ref{fig:pca_drosophila_allchr}).
-
-Across chromosomes, the high values of MDS coordinate 1 coincide with the position of the heterochromatic regions surrounding the centromere,
+% the two types of population structure with higher gene density 
+% (colored green and purple in Figure \ref{fig:mds12_medicago})
+% are nearly identical across all chromosomes,
+% but population structure associated with low gene density (colored orange)
+% shows more chromsome-specific patterns (see Supplemental Figure \ref{fig:pca_drosophila_allchr}).
+across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
 which often have lower gene density and may therefore be less subject to linked selection \citep{kulikova2001integration,paape2013selection}. 
 To verify that this is a possible explanation,
-we computed gene density for each window using gene models in Mt4.0 JBrowse \citep{tang2014improved}. 
-we downloaded the gene model track from Mt4.0 JBrowse at jcvi.org,
-extracted the start and end positions of each predicted mRNA transcript,
-counted how many of these fell in each window, 
-and use the mean number of the starts and ends in a window to represent the gene count for that window.
-The endpoints of the windows were defined to be halfway between the last SNP of each window and the first SNP in the next window.
-These values are shown juxtaposed with
+we computed gene densities in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
+which are shown juxtaposed with
 the first MDS coordinate of each window is shown in Figure \ref{fig:mds_medicago},
 and are significantly correlated, as shown in Figure \ref{fig:mds_gene_count}.
+(Values shown are the number of start and end positions of each predicted mRNA transcript,
+divided by two, assigned to the nearest window.)
+% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
 Other genomic features covary with gene density, however,
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig5_MDS_and_gene_count_allchr_update}
-    \end{center}
-    \caption{
-         MDS visualization and gene density for each window in the \textit{Medicago} genome,
+       \includegraphics{Fig6_Together_MDS_plot_chr3_final}
+    \end{center}
+    \caption{
+       MDS visualization of population structure on \textit{M.~truncatula} chromosome 3. 
+       Each point in the plot represents a window.
+       Plots for remaining chromosomes are shown in Supplemental Figure \ref{fig:mds_medicago_allchr}.
+       \label{fig:mds12_medicago}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
+    \end{center}
+    \caption{
+        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
+        Each point corresponds to a sample, colored by country of origin.
+        Plots for remaining chromosomes are shown in Supplemental Figure \ref{fig:pca_peaks_medicago_allchr}.
+        \label{fig:pca_medicago}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics{Fig5_MDS_and_gene_count_allchr_update}
+    \end{center}
+    \caption{
+         MDS coordinate and gene density for each window in the \textit{Medicago} genome,
          for chromosomes 1--8 (numbered above each pair of figures).
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig6_Together_MDS_plot_chr3_final}
-    \end{center}
-    \caption{
-     MDS visualization for Medicago chromosome 3. Each point in the plot represents a window.
-     Plots for remaining chromosomes are shown in Supplemental Figure \ref{fig:mds_medicago_allchr}.
-       \label{fig:mds12_medicago}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
-    \end{center}
-    \caption{
-        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
-        Each point corresponds to a sample, colored by country of origin.
-        Plots for remaining chromosomes are shown in Supplemental Figure \ref{fig:pca_peaks_medicago_allchr}.
-        \label{fig:pca_medicago}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
        \includegraphics{MDS_1D_against_gene_count_all_chr_win104_with_lm_update}
@@
     \caption{
-        First MDS coordinate against gene density for all 8 chromosomes. 
+        First MDS coordinate against gene density for all 8 chromosomes of \textit{M.~truncatula}.
         The first MDS coordinate is significantly correlated with gene count ($r=0.149$, $p=2.2\times 10^{-16}$). 
-        (See supplementary for each single chromosome's MDS result against gene count for \textit{Medicago})
         \label{fig:mds_gene_count}
@@
 local PCA may prove to be a useful exploratory tool to discover important genomic features.
-It is unclear whether the technique will be useful on reduced representation genotyping datasets 
-due to marker density and issues with missing data --
-our investigations with one such dataset were inconclusive --
-but even low coverage, whole-genome sequence is very promising.
+% It is unclear whether the technique will be useful on reduced representation genotyping datasets 
+% due to marker density and issues with missing data --
+% our investigations with one such dataset were inconclusive --
+% but even low coverage, whole-genome sequence is very promising.
 
@@
 \paragraph{The effect of selection}
-\plr{intro sentence}
+How can we expect selection to affect population structure?
 PCA summarizes patterns in kinship found in the genetic covariance matrix,
-which is an average across locus-specific genealogies.
-% However, different parts of genome have different genetic features. 
-% First, each site of DNA may have different gene tree. 
-% The covariance matrix of genotype matrix averages those gene trees. 
-If individuals are closer in the genealogies of a given genomic region, 
+which is an average across locus-specific genealogies:
+if individuals are closer in the genealogies of a given genomic region, 
 they tend to be closer in the PCA maps for that region. 
@@
 genealogies in these regions separate samples carrying the two orientations of the inversion.
-In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
-(indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
-as a way to identify inversions).
+% In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
+% (indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
+% as a way to identify inversions).
 % if a chromosomal inversion is polymorphic in the sample, 
@@
 % Other effects, like noise, introgression might also influence population structure.
-We will look for strong variation in population structure shared across large regions of the genome.
+
 Many of the effects listed above (such as single selective sweeps or inversions) 
@@
 because of randomness in which samples end up in which group;
-but if these coincide with a region of reduced recombination (such as an inversion),
+but if these coincide with a region of reduced recombination (or an inversion),
 these could drive major patterns of variation.
@@
 Alternatively, windows could be chosen to have equal length in genetic distance,
-so that each would have roughly the same amount of phylogenetic information.
+so that each would have roughly the same number of independent trees.
 However, we found little change in results when using different window sizes
-or when measuring windows in physical distance (in bp),
-so this seems unlikely to give different results.
-
-More generally, there are many possible methods to discover common structure in different parts of the genome.
-The methods we chose discovered strong biological signal of different types in three datasets;
-but it is possible that other methods for measuring dissimilarity between windows' covariance matrices
-or for summarizing the matrix of pairwise distances between windows
-would lead to different insights.
-Our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
-and how many MDS coordinates to use when describing the distance matrix between windows,
-but in our exploration, changing these has not produced dramatically different results.
-These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
-we encourage the user to experiment.
-
-Finally,
+or when measuring windows in physical distance (in bp).
+
+More generally, there are many possible methods to discover common structure in different parts of the genome:
 in this work we use PCA to summarize population structure,
@@
 may highlight different sources of variation in the data.
+It is also possible that other methods for measuring dissimilarity between windows' covariance matrices
+or for summarizing the matrix of pairwise distances between windows
+would lead to different insights.
+
+Finally, our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
+and how many MDS coordinates to use when describing the distance matrix between windows,
+but in our exploration, changing these has not produced dramatically different results.
+These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
+we encourage the user to experiment.
 
@@
 PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
-However, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
-or simply regions of reduced recombination.
+however, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
+or simply regions of reduced recombination
+without additional lines of evidence.
 However, in many applications, inversions are a nuisance.
@@
 \appendix
+\setcounter{table}{0}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
 
@@
 that well-approximates the original data in the least-squares sense:
-if $C$ is the $N \times N$ matrix of genotypes, 
+if $C$ is the $N \times N$ genetic covariance matrix,
 then to find the top $k$ principal components, 
@@
 \begin{align} \label{eqn:objective}
-    \| C - U \Lambda U^T \|^2 = \sum_{ij} \left( G_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 .
+    \| C - U \Lambda U^T \|^2 = \sum_{ij} \left( C_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 .
 \end{align}
 The columns of $U$, known as the principal components, are the eigenvectors of $C$,
-the eigenvalues of $C$ are $\lambda$, and the proportion of variance explained by the $m^\text{th}$ component is
+the entries of $\lambda$ are the eigenvalues of $C$, 
+and the proportion of variance explained by the $m^\text{th}$ component is
 \begin{align*}
@@
 A common workaround is to remove samples,
-but a more elegant solution can be found by reweighting the objective function in \eqref{eqn:objective}
+but a more elegant solution can be found by reweighting the objective function in \eqref{eqn:objective}.
 Let $w_{i}$ be a weight associated with sample $i$,
@@
 \begin{align} \label{eqn:weighted_objective}
-    \| W^{1/2} (C - U \Lambda U^T) W^{1/2} \|^2 = \sum_{ij} W_i W_j \left( G_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 ,
+    \| W^{1/2} (C - U \Lambda U^T) W^{1/2} \|^2 = \sum_{ij} w_i w_j \left( G_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 ,
 \end{align}
@@
 To solve \eqref{eqn:weighted_objective},
-let $\lambda$ and $V$ denote the eigenvalues and eigenvectors of $W^{1/2} C W^{1/2}$,
-so that $V \Lambda V^T$ is the closest in least squares to $W^{1/2} C W^{1/2}$;
+let $\lambda$ and $V$ denote the top $k$ eigenvalues and eigenvectors of $W^{1/2} C W^{1/2}$,
+so that $V \Lambda V^T$ is the rank $k$ matrix closest in least squares to $W^{1/2} C W^{1/2}$;
 so if we define $U = W^{-1/2} V$
@@
 \end{align*}
-
-Note finally that since we only need find the top $k$ eigenvectors;
-in our R implementation we use the Spectra library \citep{qiu2016rspectra}.
+In our R implementation we use the Spectra library \citep{qiu2016rspectra}
+to find only the top $k$ eigenvectors.
 
@@
     \caption{
-        MDS 1D plot along all 22 autosomes in Human. 
+        MDS visualization of variation in population structure amongst windows across \emph{all} human autosomes simultaneously.
+        The small group of windows with positive MDS values lie around the inversion at 8p23.
         \label{fig:mds1_along_allchr_human}
@@
         PCA plots comparison for outlier windows and non-outlier windows
+        \plr{better caption: which windows?}
         \label{fig:pca_chr8_outliers}
@@
     \caption{
-        MDS plots for all 8 chromosomes in \textit{Medicago truncatula}.
+        MDS visualizations of population structure for all 8 chromosomes of the \textit{Medicago truncatula} data.
         \label{fig:mds_medicago_allchr}
@@
     \caption{
-        PCA plots for all 8 chromosomes' 3 peaks in \textit{Medicago truncatula}.
+        PCA plots for regions colored in Figure \ref{fig:mds_medicago_allchr}
+        on all 8 chromosomes of \textit{Medicago truncatula}:
         (A) green, (B) orange, and (C) purple.
@@
     \caption{
-        PCA plots for chromosome arms 2L, 2R, 3L, 3R and X in \textit{Drosophila}.
+        PCA plots for chromosome arms 2L, 2R, 3L, 3R and X of the \textit{Drosophila melanogaster} dataset.
         \label{fig:pca_drosophila_allchr}
@@
     \caption{
-        PCA plots for all 22 autosomes in human.
+        PCA plots for all 22 huan autosomes from the POPRES data.
         \label{fig:pca_human_allchr}
@@
     \caption{
-        PCA plots for all 8 chromosomes in \textit{Medicago truncatula}.
+        PCA plots for all 8 chromosomes in the \textit{Medicago truncatula} dataset.
         \label{fig:pca_medicago_allchr}

### 2016-07-12 writeup/local_pca_paper.tex
updated readme
update

@@
 and for encouraging the project.
-\plr{Others?}
 
@@
     \begin{center}
-       \includegraphics[width=1\textwidth]{MDS_1D_win100_all_chr_col_by_chr_with_legend}
+       \includegraphics[width=1\textwidth]{FigS_MDS1D_plot_all_chr_human}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=0.9\textwidth]{FigS_PCA_chr8_outliers_combine}
-    \end{center}
-    \caption{
-        PCA plots comparison for outlier windows and non-outlier windows
-        \plr{better caption: which windows?}
+       \includegraphics[width=0.9\textwidth]{FigS_PCA_chr8_outliers_combine_update}
+    \end{center}
+    \caption{
+        PCA plots comparison for outlier windows and non-outlier windows.
         \label{fig:pca_chr8_outliers}

### 2016-07-14 writeup/submission/cover-letter.tex
cover letter

To the editor(s) --

We are writing to submit our paper,
``Local PCA shows how population structure differs along the genome'',
for consideration in JOURNAL XXX.
The paper brings together several topics that are currently of substantial interest 
to a wide audience:
First, principal components analysis is one of the most widely used descriptive methods 
applied to genomic data;
however, the results depend on the portion of the genome examined, 
and we provide a systematic method to examine this variation.
Second, the three species we study show substantially different patterns
reflective of substantially different biology,
most notably, the role and scale of chromosomal inversions.
Finally,
although we don't develop these ideas,
our method points the way towards move powerful methods for genome-wide association studies.
Furthermore,
the methods are provided as an R package,
to allow others to easily use the method.

In summary, the paper aims to both
provide conceptual unification to a range of observations
(variation along the genome, linked selection, population structure),
and provide a new statistical method,
which leads to biologically interesting observations.

Sincerely,

 Peter Ralph and Han Li

### 2016-08-02 writeup/local_pca_paper.tex
update

@@
     \begin{center}
-       \includegraphics{Fig5_MDS_and_gene_count_allchr_update}
+       \includegraphics{Fig5_MDS_and_gene_count_allchr_update_correction}
     \end{center}
@@
         MDS visualization of variation in population structure amongst windows across \emph{all} human autosomes simultaneously.
-        The small group of windows with positive MDS values lie around the inversion at 8p23.
+        The small group of windows with positive outlying MDS values lie around the inversion at 8p23.
         \label{fig:mds1_along_allchr_human}

### 2016-08-10 writeup/local_pca_paper.tex
added new figs; modified method figure

@@
 of mean kinship between individuals in large genomic datasets,
-known as population structure.
+\plr{known as population structure.}
 % Modeling such kinship is key to controling the confounding effects of population structure in genome-wide association studies.
-Genome-wide mean kinship
-is an average of the relationships across all locus-specific genealogical trees,
+Mean kinship
+is an average of the relationships across locus-specific genealogical trees,
 which can be strongly affected on intermediate genomic scales by many biological factors,
@@
 The method is implemented as an R package, called lostruct.
-
-\plr{Add percent of variance explained by first two MDS coordinates, after subtracting the genome-wide mean covariance matrix.}
-
-\plr{Add references to supplemental whole-genome PCA plots in the text.}
+\plr{conclusion sentence.}
+\plr{methods or results?}
+
+\plr{check refs to supp figs}
 
@@
 but averaging over sufficiently many trees we hope to get a stable estimate,
-independent of the genomic region chosen.
-This hope is not entirely justified: for instance,
+that doesn't depend much on the genetic markers chosen.
+This is not guarenteed:
+for instance,
 kinship on sex chromosomes is expected to differ from the autosomes;
@@
 
+\plr{short paragraph describing what the approach does/brings to the table}
+
 \section{Methods}
 
-The general steps in our method could be carried out in many ways.
-As depicted in Figure \ref{fig:diagram}, these are
+As depicted in Figure \ref{fig:diagram}, the general steps to the method are:
 (1) divide the genome into windows,
@@
 \end{figure}
-
-\subsection{Datasets}
-% Recode the DNA sequence to a matrix consisting of 0,1,2 (and NA).
-
-We used three publically available datasets (summarized in Table \ref{tab:data_stats});
-we converted each to a numeric matrix
-(with one row per polymorphic variant and one column per sample)
-by replacing each genotype with the number of nonreference alleles
-(or NA for missing data).
-A normalization step (see below) ensures the result does not depend on the choice of reference allele.
-
-\paragraph{\textit{Drosophila melanogaster}:}
-We used whole-genome sequencing data 
-from the Drosophila Genome Nexus \citep[\url{http://www.johnpool.net/genomes.html},][]{lack2015drosophila},
-consisting of the Drosophila Population Genomics Project phases 1--3 \citep{langley2012genomic,pool2012population},
-and additional African genomes \citep{lack2015drosophila}.
-After removing 20 genomes with more than 8\% missing data,
-we were left with 380 samples from 16 countries across Africa and Europe.
-Since the \textit{Drosophila} samples are from inbred lines or haploid embryos, 
-we treat the samples as haploid when recoding;
-regions with residual heterozygosity were marked as missing in the original dataset;
-we also removed positions with more than 20\% missing data. 
-% (We chose these cutoffs as the tails of the relevant empirical distributions.)
-Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs;
-PCA plots for each arm are shown in Figure \ref{fig:pca_drosophila_allchr}.
-
-\paragraph{Human:}
-We also used genomic data from the entire POPRES dataset \citep{nelson2008population},
-which has array-derived genotype information for 447,267 SNPs across the 22 autosomes
-of 3,965 samples in total: 346 African-Americas, 73 Asians, 3,187 Europeans and 359 Indian Asians.
-Since these data derive from genotyping arrays, the SNP density is much lower than the other datasets,
-which are each derived from whole genome sequencing.
-We excluded the sex chromosomes and the mitochondria.
-PCA plots for each chromosome, separately, are shown in Figure \ref{fig:pca_human_allchr}.
-% We use the allele that has highest frequency in the samples as the reference allele for each position.
-
-% lack2015drosophila: describes the DGN, including the AGES samples.  reports diversity and mentions inversions but mostly a 'data' paper
-% pool2012population: describes 129 subsaharan samples (DPGP3)
-
-\paragraph{\textit{Medicago truncatula}:}
-Finally, we used whole-genome sequencing data from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
-which has 263 samples from 24 countries,
-primarily distributed around the Mediterranean basin.
-Each of the 8 chromosomes has 3--5 million SNPs;
-PCA plots for these are shown in Figure \ref{fig:pca_medicago_allchr}.
-we did not use the mitochondria or chloroplasts.
-
-\begin{table}[ht]
-\centering
-    \begin{tabular}{p{0.8in}rrrr}
-  \hline
-    species 
-    & \parbox[t]{.8in}{\# SNPs per \\ window} 
-    & \parbox[t]{1in}{mean window\\ length (bp)}
-    & \parbox[t]{1.2in}{mean \# windows \\ per chromosome} 
-    & \parbox[t]{1.4in}{mean \% variance ex-\\plained by top 2 PCs} \\ 
-  \hline
-  \textit{Drosophila melanogaster} & 1,000 & 9,019 & 2,674 & 0.53 \\ 
-  Human & 100 & 636,494 & 203 & 0.55 \\ 
-  \textit{Medicago truncatula} & 10,000 & 102,580 & 467 & 0.50 \\ 
-   \hline
-\end{tabular}
-\caption{
-    Descriptive statistics for each dataset used.
-    % with the chosen window sizes.
-    % See text for how window sizes were chosen;
-    Columns 2--4 give statistics describing the window sizes used for most analyses for each dataset
-    (windows had a fixed number of SNPs).
-    The final column gives the percent variance explained by the top two principal components,
-    averaged across independent PCA of each window.
-    \label{tab:data_stats}
-}
-\end{table}
 
@@
 
+\subsection{Datasets}
+% Recode the DNA sequence to a matrix consisting of 0,1,2 (and NA).
+
+We used three publically available datasets (summarized in Table \ref{tab:data_stats});
+we converted each to a numeric matrix
+(with one row per polymorphic variant and one column per sample)
+by replacing each genotype with the number of nonreference alleles
+(or NA for missing data).
+A normalization step (see below) ensures the result does not depend on the choice of reference allele.
+
+\paragraph{\textit{Drosophila melanogaster}:}
+We used whole-genome sequencing data 
+from the Drosophila Genome Nexus \citep[\url{http://www.johnpool.net/genomes.html},][]{lack2015drosophila},
+consisting of the Drosophila Population Genomics Project phases 1--3 \citep{langley2012genomic,pool2012population},
+and additional African genomes \citep{lack2015drosophila}.
+After removing 20 genomes with more than 8\% missing data,
+we were left with 380 samples from 16 countries across Africa and Europe.
+Since the \textit{Drosophila} samples are from inbred lines or haploid embryos, 
+we treat the samples as haploid when recoding;
+regions with residual heterozygosity were marked as missing in the original dataset;
+we also removed positions with more than 20\% missing data. 
+% (We chose these cutoffs as the tails of the relevant empirical distributions.)
+Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs;
+PCA plots for each arm are shown in Figure \ref{fig:pca_drosophila_allchr}.
+
+\paragraph{Human:}
+We also used genomic data from the entire POPRES dataset \citep{nelson2008population},
+which has array-derived genotype information for 447,267 SNPs across the 22 autosomes
+of 3,965 samples in total: 346 African-Americas, 73 Asians, 3,187 Europeans and 359 Indian Asians.
+Since these data derive from genotyping arrays, the SNP density is much lower than the other datasets,
+which are each derived from whole genome sequencing.
+We excluded the sex chromosomes and the mitochondria.
+PCA plots for each chromosome, separately, are shown in Figure \ref{fig:pca_human_allchr}.
+% We use the allele that has highest frequency in the samples as the reference allele for each position.
+
+% lack2015drosophila: describes the DGN, including the AGES samples.  reports diversity and mentions inversions but mostly a 'data' paper
+% pool2012population: describes 129 subsaharan samples (DPGP3)
+
+\paragraph{\textit{Medicago truncatula}:}
+Finally, we used whole-genome sequencing data from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
+which has 263 samples from 24 countries,
+primarily distributed around the Mediterranean basin.
+Each of the 8 chromosomes has 3--5 million SNPs;
+PCA plots for these are shown in Figure \ref{fig:pca_medicago_allchr}.
+we did not use the mitochondria or chloroplasts.
+
+\begin{table}[ht]
+\centering
+    \begin{tabular}{p{0.8in}rrrr}
+  \hline
+    species 
+    & \parbox[t]{.8in}{\# SNPs per \\ window} 
+    & \parbox[t]{1in}{mean window\\ length (bp)}
+    & \parbox[t]{1.2in}{mean \# windows \\ per chromosome} 
+    & \parbox[t]{1.4in}{mean \% variance ex-\\plained by top 2 PCs} \\ 
+  \hline
+  \textit{Drosophila melanogaster} & 1,000 & 9,019 & 2,674 & 0.53 \\ 
+  Human & 100 & 636,494 & 203 & 0.55 \\ 
+  \textit{Medicago truncatula} & 10,000 & 102,580 & 467 & 0.50 \\ 
+   \hline
+\end{tabular}
+\caption{
+    Descriptive statistics for each dataset used.
+    % with the chosen window sizes.
+    % See text for how window sizes were chosen;
+    Columns 2--4 give statistics describing the window sizes used for most analyses for each dataset
+    (windows had a fixed number of SNPs).
+    The final column gives the percent variance explained by the top two principal components,
+    averaged across independent PCA of each window.
+    \label{tab:data_stats}
+}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-08-10 writeup/local_pca_paper.tex
added grahams comments

@@
 
-After recoding, we divided the genome into contiguous segments (``windows''; see below),
+We first recoded sampled genotypes as numeric matrices in the usual manner,
+by recording the number of nonreference alleles seen at each locus for each sample.
+We then divided the genome into contiguous segments 
+(``windows'')
 % recoded matrix into contigous matrices that have the same columns but fewer rows than the original matrix, 
@@
 separately to the submatrices that corresponded to each window.
+The choice of window length entails a tradeoff between signal and noise,
+since shorter windows allow better resolution along the genome but allow less precise estimates of kinship.
+A method for choosing a window length is given in Appendix \ref{apx:window_length}.
 % Specifically, we did PCA as follows:
@@
 The top few principal components generally display population structure; 
-we usually use the first two (referred to as $\pcone$ and $\pctwo$).
+we follow common practice and usually only use the first two (referred to as $\pcone$ and $\pctwo$).
 The above procedure can be performed on any subset of the data;
@@
 
-\subsection{Choosing window length}
-
-The choice of window length entails a balance between signal and noise.
-In very short windows, genealogies of the samples will only be represented by a few trees,
-so variation between windows represents demographic noise rather than meaningful variation in population structure.
-Longer windows generally have more distinct trees (and SNPs), 
-allowing for less noisy estimation of local population structure.
-However, to better resolve meaningful signal, i.e., differences in population structure along the genome,
-we would like reasonably short windows.
-% Window length choice therefore entails a signal versus noise tradeoff
-% in the estimates of population structure.
-% If we use the first principal component as a measure of population structure, 
-% then to choose the best window length,
-% we need to find a balance between the standard error of the first principal components for each window 
-% and the standard deviation between windows.
-
-Since we summarize population structure using relative positions in the principal component maps,
-we quantify ``noise'' as the standard error of a sample's position on PC1 in a particular window,
-averaged across windows and samples,
-and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
-averaged over samples.
-(Recall that the signs for PCs are chosen to match each other.)
-Then, the mean variance across windows is
-\begin{align*}
-    \sigma_\text{signal}^2
-    = 
-    \frac{1}{N} \sum_{j=1}^{N}
-        \frac{1}{L}\sum_{i=1}^{L}\left ( \pcone_{ij} -\overline\pcone_{j} \right )^{2} ,
-\end{align*}
-where $\pcone_{ij}$ is the position of the $i^\text{th}$ individual on $\pcone$ in window $j$,
-and $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
-We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}:
-we divide the $j^\text{th}$ window into 10 equal-sized pieces,
-and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
-then the estimate of the squared standard error is
-$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$.
-Averaging over samples and windows,
-\begin{align*}
-    \sigma^2_\text{noise}
-    &=
-    \frac{1}{N} \sum_{j=1}^{N} \frac{1}{L}\sum_{i=1}^{L} \sigma^2_{ij} .
-\end{align*}
-
-For the main analysis, we defined windows to each consist of the same number of neighboring SNPs,
-and calculated $\sigma^2_\text{signal}$ and $\sigma^2_\text{noise}$
-for a range of window sizes (i.e., numbers of SNPs).
-For our main results we
-chose the smallest window for which $\sigma^2_\text{signal}$ was consistently larger than $\sigma^2_\text{noise}$ (but checked other sizes);
-the values for various window sizes across \textit{Drosophila} chromosomes are shown in Table \ref{tab:window_sizes}.
-% and the choices for all taxa are in Table \ref{tab:data_stats}.
-% In practice, even substantially different window sizes result in the same genome-wide patterns.
-In the cases we examined, we found nearly identical results after varying window size,
-and choosing windows to be of the same physical length (in bp) rather than in numbers of SNPs.
-
-% # Variances within and between windows for Drosophila, multiplied by 1000:
-% w <- structure(list(chrom = structure(c(1L, 1L, 2L, 2L, 3L, 3L, 4L, 
-% 4L, 5L, 5L), .Label = c("2L", "2R", "3L", "3R", "X"), class = "factor"), 
-%     type = structure(c(1L, 2L, 1L, 2L, 1L, 2L, 1L, 2L, 1L, 2L
-%     ), .Label = c("noise", "signal"), class = "factor"), `100` = c(2.05209, 
-%     2.75625, 2.18089, 2.77729, 2.07936, 2.601, 1.95364, 2.58064, 
-%     2.48004, 2.61121), `500` = c(1.64025, 2.69361, 1.91844, 2.704, 
-%     1.99809, 2.52004, 1.764, 2.51001, 2.04304, 2.43049), `1000` = c(1.18336, 
-%     2.22784, 1.63216, 2.65225, 1.64025, 2.401, 1.43641, 2.44036, 
-%     1.53664, 2.304), `10000` = c(0.169, 0.676, 0.576, 2.31361, 
-%     0.73441, 1.681, 0.58564, 1.96249, 1.61604, 0.324), `1e+05` = c(0.03969, 
-%     0.31329, 0.13456, 1.82329, 0.24649, 1.89225, 0.20164, 1.39876, 
-%     0.169, 1.14244)), .Names = c("chrom", "type", "100", "500", 
-% "1000", "10000", "1e+05"), row.names = c(NA, -10L), class = "data.frame")
-% print(xtable(w),include.rownames=FALSE)
-
-\begin{table}[ht]
-\centering
-    \begin{tabular}{cccrrrrr}
-  \hline
-        & & & \multicolumn{5}{c}{window length (SNPs)} \\
- & chrom.\ arm  & & 100 & 500 & 1,000 & 10,000 & 100,000 \\ 
-  \hline
-    & 2L & $\sigma^2_\text{noise}$  & 2.05  &  1.64  &  1.18  &  0.17  &  0.04 \\
-    & 	 & $\sigma^2_\text{signal}$ & 2.76  &  2.69  &  2.23  &  0.68  &  0.31 \\
-    & 2R & $\sigma^2_\text{noise}$  & 2.18  &  1.92  &  1.63  &  0.58  &  0.13 \\
-    & 	 & $\sigma^2_\text{signal}$ & 2.78  &  2.70  &  2.65  &  2.31  &  1.82 \\
-    & 3L & $\sigma^2_\text{noise}$  & 2.08  &  2.00  &  1.64  &  0.73  &  0.25 \\
-    & 	 & $\sigma^2_\text{signal}$ & 2.60  &  2.52  &  2.40  &  1.68  &  1.89 \\
-    & 3R & $\sigma^2_\text{noise}$  & 1.95  &  1.76  &  1.44  &  0.59  &  0.20 \\
-    & 	 & $\sigma^2_\text{signal}$ & 2.58  &  2.51  &  2.44  &  1.96  &  1.40 \\
-    & X  & $\sigma^2_\text{noise}$  & 2.48  &  2.04  &  1.54  &  1.62  &  0.17 \\
-    & 	 & $\sigma^2_\text{signal}$ & 2.61  &  2.43  &  2.30  &  0.32  &  1.14 \\
-   \hline
-\end{tabular}
-\caption{
-    Measures of signal and noise,
-    computed separately for each chromosome arm in the \textit{Drosophila} dataset,
-    at different window sizes.
-    All values are multiplied by $1,000$
-    (so typical variation is of order of 50\% of the actual values).
-    Starting at windows of 1,000 SNPs, the signal (variation of PC1 between windows)
-    starts to be substantially larger than the noise (standard error of PC1 for each window).
-} \label{tab:window_sizes}
-\end{table}
-
 \subsection{Similarity of population structure between windows}
@@
 in the space defined by the top principal components.
-Since rotations of this space imply identical population structure,
-we do not compare population structure of different genomic windows using the PCs directly, 
-but instead compare the low-dimensional approximations of the local covariance matrices
-obtained using the top $k$ PCs.
+However, 
+we do not compare population structure of different genomic by directly comparing the PCs,
+since rotations or reflections of these imply identical population structure.
+Instead, we compare the low-dimensional approximations of the local covariance matrices
+obtained using the top $k$ PCs,
+which is invariant under reflections and rotations and yet contains all other information about the PCs.
 (For results shown here, we use $k=2$;
 results using larger numbers of PCs were nearly identical.)
-We do this, rather than using the entire covariance matrix, 
-both for computational efficiency 
-and because the top PCs summarize important population structure, 
-so using only these should reduce the effect of noise.
-To remove the effect of variation in mutation rate between windows,
-we also rescale each covariance matrix so that the underlying data matrix has trace norm equal to one.
+Furthermore, to remove the effect of artifacts such as mutation rate variation,
+we also rescale each approximate covariance matrix 
+so that the underlying data matrix has trace norm equal to one.
 
@@
 
-We then use
-Euclidean distance $D_{ij}$ between the matrices $M(i)$ and $M(j)$ 
-to measure the similarity of population structure for the $i^\text{th}$ window and $j^\text{th}$ window. 
+To measure the similarity of population structure for the $i^\text{th}$ window and $j^\text{th}$ window,
+we then use
+Euclidean distance $D_{ij}$ between the matrices $M(i)$ and $M(j)$:
+$D_{ij}^2 = \sum_{k\ell} ( M(i)_{k,\ell} - M(j)_{k,\ell} )^2$.
+
+Although we were motivated by comparing PC plots up to rotation and reflection,
+the method suggests instead directly comparing entire local covariance matrices. 
+However, with thousands of samples and tens of thousands of windows,
+computing the distance matrix would take months of CPU time,
+while $D$ can be computed in minutes using the following method.
 Since 
@@
 This implies that variation is due to meaningful variation in population structure, 
-since fluctuations due to demographic noise are not expected to show long distance correlations. 
+since noise due to randomness in choice of local genealogical trees
+are not expected to show long distance correlations. 
 Below, we discuss the results and likely underlying causes.
@@
 For each, the two-dimensional MDS visualization resembles a triangle,
+sometimes with additional points,
 as seen in Figure \ref{fig:mds_allchr}.
@@
 
-To investigate these extremes, for each chromosome arm we visually selected three ``extreme'' windows in the MDS plot,
-and take out the 5\% of windows that are closest to it in the MDS coordinates;
-these three groups of windows are shown in color in Figure \ref{fig:mds_allchr} and PCA plots for each collection are shown in Figure \ref{fig:pca_by_pop}.
+To investigate each extreme on a particular chromosome arm, 
+we selected an ``extreme'' window in the MDS plot
+and the 5\% of windows that are closest to it in the MDS coordinates,
+then highlighted these windows' positions along the genome,
+and created PCA plots for the windows, combined.
+Representative plots are shown for three groups of windows on each chromosome arm
+in Figure \ref{fig:mds_allchr} (groups are shown in color)
+in Figure \ref{fig:pca_by_pop} (PCA plots).
 
@@
        MDS visualization of population structure on \textit{M.~truncatula} chromosome 3. 
-       Each point in the plot represents a window.
+       Each point in the plot represents a window;
+       the structure revealed by the MDS plot is strongly clustered along the chromosome,
+       with windows in the upper-right corner of the MDS plot (colored red) clustered around the centromere,
+       windows in the upper-left corner (purple) furthest from the centromere,
+       and the remaining corner (green) intermediate.
        Plots for remaining chromosomes are shown in Supplemental Figure \ref{fig:mds_medicago_allchr}.
@@
 
-Our investigations have found a remarkable amount of variation in population structure across the genomes
+Our investigations have found substantial variation in population structure across the genomes
 of three diverse species,
@@
 More investigation, particularly on more species and datasets, will help to uncover which patterns are generalizable.
-
 With growing appreciation of the heterogeneous effects of selection across the genome,
@@
 
-\paragraph{Confounding in GWAS}
-So-called cryptic relatedness between samples
-has been one of the major sources of confounding in genome-wide association studies (GWAS)
-and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
-Since population structure is not constant along the genome,
-this could in principle lead to an inflation of false positives in parts of the genome
-with stronger population structure than the genome-wide average.
-Fortunately, in our human dataset this does not seem likely to have a strong effect:
-most variation is due to small, independent regions, possibly primarily inversions,
-and so may not have a major effect on GWAS.
-In the other species we examined, particularly \textit{Drosophila melanogaster},
-treating population structure as a single quantity could be severely misleading.
+We now discuss possible implications of this variation in population structure,
+the impact of various parameter choices in implementing the method,
+and possible additional applications.
+
+\paragraph{Chromosomal inversions}
+A major driver of variation in population structure in two datasets we examined are inversions.
+This may be common,
+but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
+PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
+however, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
+or simply regions of reduced recombination
+without additional lines of evidence.
+However, in many applications, inversions are a nuisance.
+For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
+by regressing out the effect of linked SNPs on each other.
+Removing samples with the less common orientation of each inversion reduced,
+but did not eliminate, the signal of inversions
+seen in the \textit{Drosophila melanogaster} dataset,
+demonstrating that the genomic effects of transiently polymorphic inversions
+may outlast the inversions themselves.
 
 \paragraph{The effect of selection}
-How can we expect selection to affect population structure?
-PCA summarizes patterns in kinship found in the genetic covariance matrix,
-which is an average across locus-specific genealogies:
-if individuals are closer in the genealogies of a given genomic region, 
-they tend to be closer in the PCA maps for that region. 
+It seems clear that the variation in population structure we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
+is likely due to selection.
+Furthermore, 
+the selection must be affecting many targets across the genome,
+since we see similar effects across long distances (even distinct chromosomes).
+For this reason, the most likely candidate may be selection against linked deleterious mutations, known as ``background selection'' \citep{charlesworth1993effect,charlesworth2013background}.
+Background selection can be thought of as reducing the number of potential contributors to the gene pool in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious};
+for this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently,
+Regions of the genome harboring many targets of local adaptation may show similar patterns,
+since migrant alleles in these regions will be selected against,
+and so locally gene trees will more closely reflect spatial proximity.
+
+A third possibility is that variation in population structure is due to recent admixture between previously separated populations,
+the effects of which were not uniform across the genome due to selection.
+For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
+is due to selection against Neanderthal genes, leading to greater introgression in regions of lower gene density
+\citep{harris2016genetic,juric2016strength}.
+African \textit{Drosophila melanogaster} are known to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
+if selection regularly favors genes from one origin,
+this could lead to substantial variation in population structure correlated with local gene density.
+
+There has been substantial debate over the relative impacts of different forms of selection.
+These have been difficult to disentangle in part because most concrete theoretical predictions
+come in the form of predicted forms of the allele frequency spectrum which are only strictly valid in randomly mating (i.e., unstructured) populations,
+and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
+It is possible that statistics making stronger use of spatial information
+will have more luck.
+
+% How do we expect selection to affect population structure?
+% PCA summarizes patterns in kinship found in the genetic covariance matrix,
+% which is an average across locus-specific genealogies:
+% if individuals are closer in the genealogies of a given genomic region, 
+% they tend to be closer in the PCA maps for that region. 
 % Different DNA segments may have different gene tree and therefore different population structure for those segments. 
@@
 % Background selection causes shallow ones.
-Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
-characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
-A genomic region with many targets for selection experiences
-background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
-which would tend to shorten genealogical trees in the region,
-similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
-% Balancing selection causes deep trees.
-On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
-and population structure locally describes which individuals have which alleles
-rather than geographical proximity.
-Finally, 
-since recombination is suppressed between opposite orientations of a chromosomal inversion
-near its breakpoints,
-genealogies in these regions separate samples carrying the two orientations of the inversion.
+% Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
+% characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
+% A genomic region with many targets for selection experiences
+% background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
+% which would tend to shorten genealogical trees in the region,
+% similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
+% % Balancing selection causes deep trees.
+% On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
+% and population structure locally describes which individuals have which alleles
+% rather than geographical proximity.
+% Finally, 
+% since recombination is suppressed between opposite orientations of a chromosomal inversion
+% near its breakpoints,
+% genealogies in these regions separate samples carrying the two orientations of the inversion.
 % In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
@@
 % Other effects, like noise, introgression might also influence population structure.
-
-Many of the effects listed above (such as single selective sweeps or inversions) 
-are not expected to have similar effects on population structure in different regions of the genome
-because of randomness in which samples end up in which group;
-but if these coincide with a region of reduced recombination (or an inversion),
-these could drive major patterns of variation.
-The more subtle effects of genome-wide linked selection could be shared across large regions
-due to large-scale variation in gene density
-(as by background selection or local adaptation at many genes).
+% 
+% Many of the effects listed above (such as single selective sweeps or inversions) 
+% are not expected to have similar effects on population structure in different regions of the genome
+% because of randomness in which samples end up in which group;
+% but if these coincide with a region of reduced recombination (or an inversion),
+% these could drive major patterns of variation.
+% The more subtle effects of genome-wide linked selection could be shared across large regions
+% due to large-scale variation in gene density
+% (as by background selection or local adaptation at many genes).
 
@@
 
-\paragraph{Chromosomal inversions}
-A major driver of variation in population structure in two datasets we examined are inversions.
-This may be common,
-but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
-PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
-however, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
-or simply regions of reduced recombination
-without additional lines of evidence.
-However, in many applications, inversions are a nuisance.
-For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
-by regressing out the effect of linked SNPs on each other.
-It would be interesting to see if somehow removing the effects of inversions in the \textit{Drosophila melanogaster} or human datasets
-would produce a pattern similar to that seen in \textit{Medicago truncatula}.
+\paragraph{Applications}
+So-called cryptic relatedness between samples
+has been one of the major sources of confounding in genome-wide association studies (GWAS)
+and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
+Since population structure is not constant along the genome,
+this could in principle lead to an inflation of false positives in parts of the genome
+with stronger population structure than the genome-wide average.
+A method such as ours might be used to provide a more sensitive correction.
+Fortunately, in our human dataset this does not seem likely to have a strong effect:
+most variation is due to small, independent regions, possibly primarily inversions,
+and so may not have a major effect on GWAS.
+In the other species we examined, particularly \textit{Drosophila melanogaster},
+treating population structure as a single quantity would entail a substantial loss of power,
+and could potentially be misleading.
 
@@
 (especially inversion status of \textit{D.~melanogaster} samples).
-Thanks also go to Yaniv Brandvain and Barbara Engelhardt for helpful comments 
+Thanks also go to Yaniv Brandvain, Barbara Engelhardt, Charles Langley, Graham Coop, and Jeremy Berg for helpful comments
 and for encouraging the project.
@@
 \renewcommand{\thefigure}{S\arabic{figure}}
+
+\section{Choosing window length}
+\label{apx:window_length}
+
+The choice of window length entails a balance between signal and noise.
+In very short windows, genealogies of the samples will only be represented by a few trees,
+so variation between windows represents demographic noise rather than meaningful variation in population structure.
+Longer windows generally have more distinct trees (and SNPs), 
+allowing for less noisy estimation of local population structure.
+However, to better resolve meaningful signal, i.e., differences in population structure along the genome,
+we would like reasonably short windows.
+% Window length choice therefore entails a signal versus noise tradeoff
+% in the estimates of population structure.
+% If we use the first principal component as a measure of population structure, 
+% then to choose the best window length,
+% we need to find a balance between the standard error of the first principal components for each window 
+% and the standard deviation between windows.
+
+Since we summarize population structure using relative positions in the principal component maps,
+we quantify ``noise'' as the standard error of a sample's position on PC1 in a particular window,
+averaged across windows and samples,
+and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
+averaged over samples.
+(Recall that the signs for PCs are chosen to match each other.)
+Then, the mean variance across windows is
+\begin{align*}
+    \sigma_\text{signal}^2
+    = 
+    \frac{1}{N} \sum_{j=1}^{N}
+        \frac{1}{L}\sum_{i=1}^{L}\left ( \pcone_{ij} -\overline\pcone_{j} \right )^{2} ,
+\end{align*}
+where $\pcone_{ij}$ is the position of the $i^\text{th}$ individual on $\pcone$ in window $j$,
+and $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
+We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}:
+we divide the $j^\text{th}$ window into 10 equal-sized pieces,
+and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
+then the estimate of the squared standard error is
+$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$.
+Averaging over samples and windows,
+\begin{align*}
+    \sigma^2_\text{noise}
+    &=
+    \frac{1}{N} \sum_{j=1}^{N} \frac{1}{L}\sum_{i=1}^{L} \sigma^2_{ij} .
+\end{align*}
+
+For the main analysis, we defined windows to each consist of the same number of neighboring SNPs,
+and calculated $\sigma^2_\text{signal}$ and $\sigma^2_\text{noise}$
+for a range of window sizes (i.e., numbers of SNPs).
+For our main results we
+chose the smallest window for which $\sigma^2_\text{signal}$ was consistently larger than $\sigma^2_\text{noise}$ (but checked other sizes);
+the values for various window sizes across \textit{Drosophila} chromosomes are shown in Table \ref{tab:window_sizes}.
+% and the choices for all taxa are in Table \ref{tab:data_stats}.
+% In practice, even substantially different window sizes result in the same genome-wide patterns.
+In the cases we examined, we found nearly identical results after varying window size,
+and choosing windows to be of the same physical length (in bp) rather than in numbers of SNPs.
+
+% # Variances within and between windows for Drosophila, multiplied by 1000:
+% w <- structure(list(chrom = structure(c(1L, 1L, 2L, 2L, 3L, 3L, 4L, 
+% 4L, 5L, 5L), .Label = c("2L", "2R", "3L", "3R", "X"), class = "factor"), 
+%     type = structure(c(1L, 2L, 1L, 2L, 1L, 2L, 1L, 2L, 1L, 2L
+%     ), .Label = c("noise", "signal"), class = "factor"), `100` = c(2.05209, 
+%     2.75625, 2.18089, 2.77729, 2.07936, 2.601, 1.95364, 2.58064, 
+%     2.48004, 2.61121), `500` = c(1.64025, 2.69361, 1.91844, 2.704, 
+%     1.99809, 2.52004, 1.764, 2.51001, 2.04304, 2.43049), `1000` = c(1.18336, 
+%     2.22784, 1.63216, 2.65225, 1.64025, 2.401, 1.43641, 2.44036, 
+%     1.53664, 2.304), `10000` = c(0.169, 0.676, 0.576, 2.31361, 
+%     0.73441, 1.681, 0.58564, 1.96249, 1.61604, 0.324), `1e+05` = c(0.03969, 
+%     0.31329, 0.13456, 1.82329, 0.24649, 1.89225, 0.20164, 1.39876, 
+%     0.169, 1.14244)), .Names = c("chrom", "type", "100", "500", 
+% "1000", "10000", "1e+05"), row.names = c(NA, -10L), class = "data.frame")
+% print(xtable(w),include.rownames=FALSE)
+
+\begin{table}[ht]
+\centering
+    \begin{tabular}{cccrrrrr}
+  \hline
+        & & & \multicolumn{5}{c}{window length (SNPs)} \\
+ & chrom.\ arm  & & 100 & 500 & 1,000 & 10,000 & 100,000 \\ 
+  \hline
+    & 2L & $\sigma^2_\text{noise}$  & 2.05  &  1.64  &  1.18  &  0.17  &  0.04 \\
+    & 	 & $\sigma^2_\text{signal}$ & 2.76  &  2.69  &  2.23  &  0.68  &  0.31 \\
+    & 2R & $\sigma^2_\text{noise}$  & 2.18  &  1.92  &  1.63  &  0.58  &  0.13 \\
+    & 	 & $\sigma^2_\text{signal}$ & 2.78  &  2.70  &  2.65  &  2.31  &  1.82 \\
+    & 3L & $\sigma^2_\text{noise}$  & 2.08  &  2.00  &  1.64  &  0.73  &  0.25 \\
+    & 	 & $\sigma^2_\text{signal}$ & 2.60  &  2.52  &  2.40  &  1.68  &  1.89 \\
+    & 3R & $\sigma^2_\text{noise}$  & 1.95  &  1.76  &  1.44  &  0.59  &  0.20 \\
+    & 	 & $\sigma^2_\text{signal}$ & 2.58  &  2.51  &  2.44  &  1.96  &  1.40 \\
+    & X  & $\sigma^2_\text{noise}$  & 2.48  &  2.04  &  1.54  &  1.62  &  0.17 \\
+    & 	 & $\sigma^2_\text{signal}$ & 2.61  &  2.43  &  2.30  &  0.32  &  1.14 \\
+   \hline
+\end{tabular}
+\caption{
+    Measures of signal and noise,
+    computed separately for each chromosome arm in the \textit{Drosophila} dataset,
+    at different window sizes.
+    All values are multiplied by $1,000$
+    (so typical variation is of order of 50\% of the actual values).
+    Starting at windows of 1,000 SNPs, the signal (variation of PC1 between windows)
+    starts to be substantially larger than the noise (standard error of PC1 for each window).
+} \label{tab:window_sizes}
+\end{table}
 

### 2016-08-10 writeup/local_pca_paper.tex
.

@@
 However, it is widely recognized that because of selection the effects of gene flow are not equal everywhere on the genome,
-and patterns of polymorphism and divergence can vary significantly depending on factors including local gene density.
+and patterns of polymorphism within species and divergence between species can vary significantly depending on factors including local gene density.
 This implies that, counterintuitively, 
@@
 and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
-\citep{maynardsmith1974hitchhiking,barton2000genetic,neher2013genetic,charlesworth2012effects}.
+\citep{kim2002hitchhiking,charlesworth1993effect,barton2000genetic}.
 Indeed,
 many species show chromosome-scale variation in diversity and divergence
-\citep[sometimes dramatic:][]{langley2012genomic};
+\citep[e.g.,][]{langley2012genomic};
 species phylogenies can differ along the genome 
@@
 \citep{charlesworth2003review}.
-Nonetheless, it is not generally known to what extent patterns of kinship vary along the genome,
-nor for what reasons.
 
@@
 
+Population genetics studies often describe variation in particular statistics along the genome,
+often with the aim of detecting particular signals.
+In this paper, we use a visualization technique to obtain an unbiased view
+of how patterns of kinship vary along the genome,
+in genome-wide, spatially distributed datasets for three species,
+to start to discover common patterns and differences.
+
+%%%
 \section{Methods}

### 2016-08-10 writeup/local_pca_paper.tex
edited after chucks comments

@@
 However, if we flip perspectives and think of each genetic variant as an observation,
-our method shares common threads, although the ultimate goals and methods for visualization are different.
-Future methods for visualization of genomic data
-may benefit from other advances in this substantial literature 
-\citep[reviewed in][]{vandermaaten2009dimensionality}.
+our method shares common threads, although 
+our method does not subsequently use adjacency along the genome,
+as we aim to identify similar regions that may be distant.
+% Future methods for visualization of genomic data
+% may benefit from other advances in this substantial literature 
+% \citep[reviewed in][]{vandermaaten2009dimensionality}.
 
@@
 across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
-which often have lower gene density and may therefore be less subject to linked selection \citep{kulikova2001integration,paape2013selection}. 
+which often have lower gene density and may therefore be less subject to linked selection.
+% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
 To verify that this is a possible explanation,
@@
 will have more luck.
+\plr{expand}
 

### 2016-08-11 writeup/local_pca_paper.tex
added another drosophila fig without inversions

@@
 \citep{charlesworth2003review}.
+
+\plr{mention Fst along the genome}
 
@@
 and that other windows' population structure may be a mixture of those extremes. 
+\plr{make clear corners are arbitrary}
 
@@
 \end{figure}
+
+To attempt to see how population structure varies without the effect of polymorphic inversions,
+we then performed the same analyses after removing, for each chromosome arm,
+the samples having the less frequent orientation of any inversion on that chromosome arm.
+The result showed substantially less variation in MDS axes,
+although some variation associated with inversion breakpoints was still visible,
+as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions}.
+For instance, the majority of the variation along 3L in Figure \ref{fig:mds_allchr}
+is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
+there is also a relatively small dip on the right end of the arm (near the centromere).
+In contrast, Supplemental Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
+remaining structure is dominated by the dip near the centromere.
+The remaining variation seems to vary systematically along the chromosome,
+following roughly similar patterns to that seen in recombination rate and diversity in \citet{langley2012genomic} and \citet{mackay2012drosophila}.
+
+\plr{gene density from RefSeq Genes from UCSC, showing genes from NCBI RNA references sequence collections. we use txStart and txEnd (transcription start/end).}
 
@@
 
+\begin{figure}
+    \begin{center}
+        \includegraphics{MDS_allchr_Together_plot_samples_noinv}
+    \end{center}
+    \caption{
+        Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms: without inversions.
+         As in Figure \label{fig:mds_allchr}, but after omitting for each chromosome arm individuals carrying the less frequent orientation
+         of any inversions on that chromosome arm.
+        In all plots, each point represents one window along the genome.
+         The first column shows the MDS visualization of relationships between windows, 
+         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
+         rows correspond to chromosome arms.
+         Colors are consistent for plots in each row. 
+         Vertical lines show the breakpoints of known polymorphic inversions.   
+         \label{sfig:mds_allchr_noinversions}
+    }
+\end{figure}
+
 \end{document}

### 2016-08-12 writeup/local_pca_paper.tex
.

@@
 %\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
+\usepackage{graphicx}				% Use pdf, png, jpg, or epsÂ§ with pdflatex; use eps in DVI mode
 								% TeX will automatically convert eps --> pdf in pdflatex		
@@
 PCA plots for these are shown in Figure \ref{fig:pca_medicago_allchr}.
-we did not use the mitochondria or chloroplasts.
+We did not use the mitochondria or chloroplasts.
 
@@
 
-To attempt to see how population structure varies without the effect of polymorphic inversions,
+To see how population structure varies without the effect of polymorphic inversions,
 we then performed the same analyses after removing, for each chromosome arm,
@@
 remaining structure is dominated by the dip near the centromere.
-The remaining variation seems to vary systematically along the chromosome,
+The remaining variation varies systematically along the chromosome,
 following roughly similar patterns to that seen in recombination rate and diversity in \citet{langley2012genomic} and \citet{mackay2012drosophila}.
+Indeed, correlations between the recombination rate \plr{XXX} near each window and the position on the first MDS coordinate are highly significant
+(Figure \ref{fig:drosophila_recomb_rate}).
+This is consistent with the hypothesis that variation in population structure
+is due to selection, since the strength of linked selection increases with local gene density, measured in units of recombination distance.
+However, the number of genes -- measured as the number of transcription start and end sites within each window (obtained from the RefGene table, UCSC genome browser) --
+was more weakly correlated with MDS coordinate (Supplemental Figures \ref{sfig:drosophila_gene_density_plots} and \ref{sfig:drosophila_gene_density_correlations}).
+
+\begin{figure}
+    \begin{center}
+       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
+    \end{center}
+    \caption{
+         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
+         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
+         \citep[data from][]{lack2015drosophila}
+        \label{fig:pca_by_inversion}
+    }
+\end{figure}
 
@@
 
+\begin{figure}
+    \begin{center}
+        \includegraphics{MDS_and_gene_count_allchr_drosophila_win104_try}
+    \end{center}
+    \caption{
+        Gene number and population structure for \textit{Drosophila melanogaster}:
+        The first MDS coordinate along the genome (black, above) is as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions};
+        below is the number of genes' transcription start and end sites within each window, divided by two.
+        Transcription start and end sites were obtained from the RefGene table from the UCSC browser, assembly dm3, on August 11, 2016.
+        The extremely dense cluster of histone genes on chromosome arm 2L is excluded.
+        \label{sfig:drosophila_gene_density_plots}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{MDS_against_gene_count_allchr_drosophila_win104}
+    \end{center}
+    \caption{
+        Gene number and population structure for \textit{Drosophila melanogaster}:
+        this shows the first MDS coordinate and gene number, as in Supplementary Figure \ref{sfig:drosophila_gene_density_plots},
+        against each other.
+        The correlation coefficient and $p$-values for each linear regression are as follows:
+        \plr{FILL THIS IN}.
+% >summary of the linear regression of MDS against gene count for 2L
+% 
+% Call:
+% lm(formula = coordinate1_2L[-k] ~ gene_count1[-k])
+% 
+% Residuals:
+%      Min       1Q   Median       3Q      Max 
+% -0.40257 -0.15326  0.04554  0.15037  0.32149 
+% 
+% Coefficients:
+%                   Estimate Std. Error t value Pr(>|t|)
+% (Intercept)      0.0087783  0.0182530   0.481    0.631
+% gene_count1[-k] -0.0003592  0.0006379  -0.563    0.574
+% 
+% Residual standard error: 0.1851 on 273 degrees of freedom
+% Multiple R-squared:  0.00116,	Adjusted R-squared:  -0.002498 
+% F-statistic: 0.3172 on 1 and 273 DF,  p-value: 0.5738
+% 
+% 
+% 
+% > summary of the linear regression of MDS against gene count for 2R
+% 
+% Call:
+% lm(formula = coordinate1_2R ~ gene_count2)
+% 
+% Residuals:
+%      Min       1Q   Median       3Q      Max 
+% -0.30622 -0.08482  0.02146  0.08438  0.30339 
+% 
+% Coefficients:
+%               Estimate Std. Error t value Pr(>|t|)    
+% (Intercept)  0.0846724  0.0163485   5.179 5.08e-07 ***
+% gene_count2 -0.0028749  0.0004704  -6.112 4.52e-09 ***
+% ---
+% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+% 
+% Residual standard error: 0.1285 on 217 degrees of freedom
+% Multiple R-squared:  0.1469,	Adjusted R-squared:  0.1429 
+% F-statistic: 37.35 on 1 and 217 DF,  p-value: 4.516e-09
+% 
+% 
+% 
+% 
+% > summary of the linear regression of MDS against gene count for 3L
+% 
+% Call:
+% lm(formula = coordinate1_3L ~ gene_count3)
+% 
+% Residuals:
+%      Min       1Q   Median       3Q      Max 
+% -0.24754 -0.09927 -0.04406  0.09895  0.34383 
+% 
+% Coefficients:
+%               Estimate Std. Error t value Pr(>|t|)    
+% (Intercept) -0.0455675  0.0135826  -3.355  0.00091 ***
+% gene_count3  0.0018695  0.0004338   4.309 2.31e-05 ***
+% ---
+% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+% 
+% Residual standard error: 0.1393 on 265 degrees of freedom
+% Multiple R-squared:  0.06548,	Adjusted R-squared:  0.06196 
+% F-statistic: 18.57 on 1 and 265 DF,  p-value: 2.312e-05
+% 
+% 
+% 
+% 
+% 
+% 
+% > summary of the linear regression of MDS against gene count for 3R
+% 
+% Call:
+% lm(formula = coordinate1_3R ~ gene_count4)
+% 
+% Residuals:
+%     Min      1Q  Median      3Q     Max 
+% -0.2883 -0.1316 -0.0424  0.0892  0.4433 
+% 
+% Coefficients:
+%               Estimate Std. Error t value Pr(>|t|)    
+% (Intercept) -0.0606512  0.0158644  -3.823 0.000161 ***
+% gene_count4  0.0022703  0.0004608   4.927  1.4e-06 ***
+% ---
+% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+% 
+% Residual standard error: 0.1713 on 291 degrees of freedom
+% Multiple R-squared:  0.077,	Adjusted R-squared:  0.07383 
+% F-statistic: 24.28 on 1 and 291 DF,  p-value: 1.403e-06
+% 
+% 
+% 
+% 
+% 
+% > summary of the linear regression of MDS against gene count for X
+% 
+% Call:
+% lm(formula = coordinate1_X ~ gene_count5)
+% 
+% Residuals:
+%      Min       1Q   Median       3Q      Max 
+% -0.46790 -0.09270  0.01922  0.11102  0.35063 
+% 
+% Coefficients:
+%               Estimate Std. Error t value Pr(>|t|)    
+% (Intercept)  0.0411519  0.0139012   2.960 0.003331 ** 
+% gene_count5 -0.0020778  0.0005487  -3.787 0.000186 ***
+% ---
+% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+% 
+% Residual standard error: 0.1471 on 286 degrees of freedom
+% Multiple R-squared:  0.04774,	Adjusted R-squared:  0.04441 
+% F-statistic: 14.34 on 1 and 286 DF,  p-value: 0.0001862
+
+        \label{sfig:drosophila_gene_density_correlations}
+    }
+\end{figure}
+
 \end{document}

### 2016-08-12 writeup/local_pca_paper.tex
moved the cart before the horse

@@
 \usepackage[hidelinks]{hyperref}
-
-\title{Local PCA Shows How Population Structure Differs Along the Genome}
+\usepackage[utf8]{inputenc}
+
+\title{Local PCA Shows How the Effect of Population Structure Differs Along the Genome}
 \author{Han Li, Peter Ralph}
@@
 %\subsection{}
-\doublespacing
+% \doublespacing
 
@@
 
-Dimension reduction techniques such as principal component analysis (PCA)
-are often used to discover and visualize the patterns 
-of mean kinship between individuals in large genomic datasets,
-\plr{known as population structure.}
-% Modeling such kinship is key to controling the confounding effects of population structure in genome-wide association studies.
-Mean kinship
+Population structure leads to systematic patterns in measures of mean relatedness
+between individuals in large genomic datasets,
+which are often discovered and visualized using dimension reduction techniques such as principal component analysis (PCA).
+Mean relatedness
 is an average of the relationships across locus-specific genealogical trees,
@@
 and apply the method to genomic data from three species,
-finding in each that population structure varies on the scale of megabases to tens of megabases.
+finding in each that the effect of population structure varies on the scale of megabases to tens of megabases.
 In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
-In a dataset of primarily African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
-is explained by known chromosomal inversions thought to be under recent selection.
 In a range-wide dataset of \textit{Medicago truncatula},
-common axes of variation in population structure are shared between chromosomes,
+common axes of variation are shared between chromosomes,
 correlate with local gene density,
 and may be caused by background selection or local adaptation.
-The method is implemented as an R package, called lostruct.
-\plr{conclusion sentence.}
-\plr{methods or results?}
-
-\plr{check refs to supp figs}
+In a dataset of primarily African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
+is explained by known chromosomal inversions thought to be under recent selection,
+and after removing samples carrying inversions, remaining variation is correlated with recombination rate and gene density,
+again suggesting a role for linked selection.
+% The method is implemented as an R package, called lostruct.
+The visualization method provides a flexible new way to discover biological drivers of genetic variation,
+and its application to data highlights the strong effects that linked selection and chromosomal inversions
+can have on observed patterns of genetic variation.
 
 \section{Introduction}
+
+% NOTE on historical uses of 'population structure'
+% Population Structure in Toads, Albert P. Blair, The American Naturalist, Vol. 77, No. 773 (Nov. - Dec., 1943), pp. 563-568
+% "In the last decade it has become increasingly apparent that the development of diversity within species populations is intimately concerned with such factors as size of breeding populations, periodic fluctuation of population size, sex ratio, activity range and differential survival of progeny.
+The sum total of these factors, population structure, determines to a great extent the relative importance of the roles which mutation, natural selection and random fluctuation of gene frequency play in the genetic configuration of the species."
+% Wright 1949, The Genetical Structure of Populations, says that:
+% "I propose to discuss certain genetic aspects of population structure. The term is used to include such matters as numbers, composition by age and sex, and state of subdivision."
+
+\citet{wright1949genetical} defined \emph{population structure} to encompass ``such matters as numbers, composition by age and sex, and state of subdivision'',
+where ``subdivision'' refers to restricted migration between subpopulations.
+The phrase is also commonly used to refer to the genetic patterns that result from this process,
+as for instance reduced mean relatedness between individuals from distinct populations.
+However, it is not necessarily clear what aspects of demography should be included in the concept.
+For instance, 
+\citet{blair1943population} defines \emph{population structure} to be the sum total of
+``such factors as size of breeding populations, periodic fluctuation of population size, sex ratio, activity range and \emph{differential survival of progeny}'' (emphasis added).
+The definition is similar to Wright's, but differs in including the effects of natural selection.
+On closer examination, incorporating differential survival or fecundity makes the concept less clear:
+should a randomly mating population consisting of two types that are partially reproductively isolated from each other
+be said to show population structure or not?
+Whatever the definition,
+it is clear that due to natural selection, the effects of population structure --
+the \emph{realized} patterns of genetic relatedness --
+differ depending on which portion of the genome is being considered.
+For instance, strongly locally adapted alleles of a gene will be selected against in migrants to different habitats,
+increasing genetic differentiation between populations near to this gene.
+Similarly, newly adaptive alleles spread first in local populations.
+These observations motivate many methods to search for genetic loci under selection,
+as for example in \citet{huertasnchez2013genetic} and \citet{martin2016natural}.
+
+% The phrase ``population structure'',
+% used to describe patterns in genome-wide mean kinship,
+% calls to mind to reduced gene flow between subpopulations. %, often because of geographical isolation.
+% However, it is widely recognized that because of selection the effects of gene flow are not equal everywhere on the genome,
+% and patterns of polymorphism within species and divergence between species can vary significantly depending on factors including local gene density.
+% This implies that, counterintuitively, 
+% the population structure of a species depends on which part of the genome is being examined.
 
@@
 % Populations are often structured in some way while there are systematic genetic variation between populations.
-
-The phrase ``population structure'',
-used to describe patterns in genome-wide mean kinship,
-calls to mind to reduced gene flow between subpopulations. %, often because of geographical isolation.
-However, it is widely recognized that because of selection the effects of gene flow are not equal everywhere on the genome,
-and patterns of polymorphism within species and divergence between species can vary significantly depending on factors including local gene density.
-This implies that, counterintuitively, 
-the population structure of a species depends on which part of the genome is being examined.
 
@@
 
-Realized kinship summarizes the shapes
-of the genealogical trees % that relate the samples
+These realized patterns of genetic relatedness
+summarize the shapes of the genealogical trees % that relate the samples
 at each location along the genome.
-Since these trees vary along the genome, so does realized kinship;
-but averaging over sufficiently many trees we hope to get a stable estimate,
+Since these trees vary along the genome, so does relatedness,
+but averaging over sufficiently many trees we hope to get a stable estimate
 that doesn't depend much on the genetic markers chosen.
-This is not guarenteed:
+This is not guaranteed:
 for instance,
-kinship on sex chromosomes is expected to differ from the autosomes;
+relatedness on sex chromosomes is expected to differ from the autosomes;
 and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
@@
 species phylogenies can differ along the genome 
-due to incomplete lineage sorting \citep[e.g.,][]{pease2013accurate},
-adaptive introgression and/or local adaptation \citep[e.g.,][]{ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
+due to incomplete lineage sorting,
+adaptive introgression and/or local adaptation 
+\citep[e.g.,][]{pease2013accurate,ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
 and theoretical expectations predict that geographic patterns of relatedness should depend on selection
@@
 
-\plr{mention Fst along the genome}
-
-Patterns in genome-wide kinship are often summarized
+Patterns in genome-wide relatedness are often summarized
 by applying principal components analysis \citep[PCA,][]{patterson2006population} 
@@
 
-Furthermore,
-modeling ``background'' kinship between samples
+Modeling such ``background'' kinship between samples
 is essential to genome-wide association studies \citep[GWAS,][]{price2006principal,astle2009population},
@@
 could lead to more generally powerful methods,
-and is essential for doing GWAS in species with substantial variation in population structure,
+and may be essential for doing GWAS in species with substantial heterogeneity in realized patterns of mean relatedness along the genome.
 
@@
 
-In this work we describe variation in population structure using local PCA,
+A note on nomenclature: In this work we describe variation in population structure using local PCA,
 where ``local'' refers to proximity along the genome.
@@
 
-\plr{short paragraph describing what the approach does/brings to the table}
-
 Population genetics studies often describe variation in particular statistics along the genome,
@@
 
-%%%
-\section{Methods}
+As reviewed above,
+% In searching for loci of interest, 
+it is common to describe variation in genetic differentiation between pairs of populations along the genome (often using $F_{ST}$),
+and also to use methods such as PCA to visualize large-scale patterns in genome-wide relatedness.
+In this paper we aim to combine these two, by describing in an unbiased, descriptive way
+how patterns of mean relatedness vary systematically along the genome, with a focus on large, geographically distributed populations.
+Our aim is not to identify outlier loci,
+but rather to describe larger-scale variation shared by many parts of the genome.
+By doing this with three taxa with diverse population histories,
+we can compare relative contributions of different sorts of variation across taxa;
+since the method is an visualization method, we allow ourselves to be suprised by unexpected signals in the data.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Results}
 
@@
 and (5) combine similar windows to more accurately estimate local population structure.
-Details of how we carried these out are given below.
+Details of how we carried these out are given in the Methods.
 
@@
     \caption{
-         An illustration of the method; see text for details.
+         An illustration of the method; see Methods for details.
          % (1) the genome is divided into windows;
@@
 
+We applied the method to genomic datasets with good geographic sampling
+of African populations of the fruit fly \textit{Drosophila melanogaster},
+a worldwide dataset of humans,
+and a range-wide dataset of the partially selfing weedy annual plant \textit{Medicago truncatula}.
+In these three species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
+This implies that variation is due to meaningful variation in population structure, 
+since noise due to randomness in choice of local genealogical trees
+are not expected to show long distance correlations. 
+Below, we discuss the results and likely underlying causes.
+
+%%%%%%%%%%%
+\subsection{\textit{Drosophila melanogaster}}
+
+We ran the method on chromosome arms 2L, 2R, 3L, 3R and X separately. 
+For each, the two-dimensional MDS visualization resembles a triangle,
+sometimes with additional points,
+as seen in Figure \ref{fig:mds_allchr}.
+Since the relative position of each window in this plot shows the similarity between windows, 
+this suggests that there are at least three extreme types of population structure 
+typified by windows found in the three corners of the triangle,
+and that other windows' population structure may be a mixture of those extremes. 
+\plr{make clear corners are arbitrary}
+
+\plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years, probably why it doesn't show up. 
+Also cite \citet{corbettdetig2012sequencebased} for a computational method for detecting inversions.}
+% corbettdetig2012sequencebased suggests these inversions are recent-ish (but before out-of-Africa)
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
+    \end{center}
+    \caption{
+        Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms.
+        In all plots, each point represents one window along the genome.
+         The first column shows the MDS visualization of relationships between windows, 
+         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
+         rows correspond to chromosome arms.
+         Colors are consistent for plots in each row. 
+         Vertical lines show the breakpoints of known polymorphic inversions.   
+         Solid black lines are for the inversions we used in Figure \ref{fig:pca_by_inversion},
+         while dotted grey lines are for other known inversions.     
+         \label{fig:mds_allchr}
+    }
+\end{figure}
+
+To investigate each extreme on a particular chromosome arm, 
+we selected an ``extreme'' window in the MDS plot
+and the 5\% of windows that are closest to it in the MDS coordinates,
+then highlighted these windows' positions along the genome,
+and created PCA plots for the windows, combined.
+Representative plots are shown for three groups of windows on each chromosome arm
+in Figure \ref{fig:mds_allchr} (groups are shown in color)
+in Figure \ref{fig:pca_by_pop} (PCA plots).
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
+    \end{center}
+    \caption{      
+        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_allchr},
+        on each chromosome arm of \textit{Drosophila melanogaster}.
+        In all plots, each point represents a sample. 
+        The first column shows the combined PCA plot for windows 
+        whose points are colored green in Figure \ref{fig:mds_allchr}; 
+        the second is for orange windows; 
+        and third is for purple windows.
+         \label{fig:pca_by_pop}
+    }
+\end{figure}
+
+The variation in population structure turns out to be almost entirely explained by
+several large inversions that are polymorphic in these samples, 
+disussed in \citet{corbett2012population} and \citet{langley2012genomic}.
+% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
+To depict this, Figure \ref{fig:pca_by_inversion} shows
+the PCA plots in Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
+Taking chromosome arm 2L as an example,
+the two regions of similar, extreme population structure
+shown in green in the first row of Figure \ref{fig:mds_allchr}
+lie directly around the breakpoints of the inversion In(2L)t,
+and the PCA plots in the first rows of Figure \ref{fig:pca_by_inversion}
+shows that population structure here is mostly determined by inversion orientation.
+The regions shown in purple on chromosome 2L lie near the centromere,
+and have population structure reflective of two axes of variation,
+seen in Figure \ref{fig:pca_by_pop},
+which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
+(see \citet{lack2015drosophila} for more about admixture in this sample).
+The regions shown in orange on chromosome 2L mostly lie inside the inversion,
+and show population structure that is a mixture between the other two,
+as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
+Similar results are found in other chromosome arms,
+albeit complicated by the coexistence of more than one polymorphic inversion;
+however, each breakpoint visibly affects patterns in the MDS coordinates.
+% except that the situation of Chr3R is a little more complicated,
+% due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
+% However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
+% although we have not highlighted portions of the MDS plot that pick these out.
+% The inversions In(3L)P and In(3R)Mo, are at low frequency in these samples, 
+% and visibly affect the MDS plots although less than the more common inversion on each arm.
+% The very young inversion In(X)be shows little signal 
+% compared to the older inversion In(X)A \citep{corbett2012population}.
+
+\begin{figure}
+    \begin{center}
+       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
+    \end{center}
+    \caption{
+         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
+         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
+         \citep[data from][]{lack2015drosophila}
+        \label{fig:pca_by_inversion}
+    }
+\end{figure}
+
+To see how population structure varies without the effect of polymorphic inversions,
+we then performed the same analyses after removing, for each chromosome arm,
+the samples having the less frequent orientation of any inversion on that chromosome arm.
+The result showed substantially less variation in MDS axes,
+although some variation associated with inversion breakpoints was still visible,
+as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions}.
+For instance, the majority of the variation along 3L in Figure \ref{fig:mds_allchr}
+is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
+there is also a relatively small dip on the right end of the arm (near the centromere).
+In contrast, Supplemental Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
+remaining structure is dominated by the dip near the centromere.
+The remaining variation varies systematically along the chromosome,
+following roughly similar patterns to that seen in recombination rate and diversity in \citet{langley2012genomic} and \citet{mackay2012drosophila}.
+Indeed, correlations between the recombination rate \plr{XXX} near each window and the position on the first MDS coordinate are highly significant
+(Figure \ref{fig:drosophila_recomb_rate}).
+This is consistent with the hypothesis that variation in population structure
+is due to selection, since the strength of linked selection increases with local gene density, measured in units of recombination distance.
+However, the number of genes -- measured as the number of transcription start and end sites within each window --
+was more weakly correlated with MDS coordinate (Supplemental Figures \ref{sfig:drosophila_gene_density_plots} and \ref{sfig:drosophila_gene_density_correlations}).
+
+\begin{figure}
+    \begin{center}
+        % \includegraphics{}
+    \end{center}
+    \caption{
+        Population structure without inversions is correlated to recombination rate:
+        \plr{XXX}
+        \label{fig:drosophila_recomb_rate}
+    }
+\end{figure}
+
+%%%%%%%%%%%
+\subsection{Human}
+
+As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes.
+On each, variation in population structure was dominated by a small number of windows
+having similar population structure to each other that differed dramatically from the rest of the chromosome.
+These may be primarily inversions: outlying windows coincide with
+three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
+notably a particularly large, polymorphic inversion on 8p23 (Figure \ref{fig:mds_human}). 
+% for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}) 
+Similar plots for all chromosomes are shown
+in Supplementary Figures \ref{sfig:mds12_chr1_8_human}, \ref{sfig:mds12_chr9_16_human}, \ref{sfig:mds12_chr17_22_human}.
+PCA plots of many outlying windows show a characteristic trimodal shape 
+(shown for chromosome 8 in Figure \ref{sfig:pca_chr8_outliers}),
+presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
+(although we do not have data on orientation status).
+This trimodal shape has been proposed as a method to identify inversions \citep{ma2012investigation},
+but distinguishing this hypothesis from others,
+such as regions of low recombination rate,
+would require additional data.
+
+We also ran the method grouping all 22 autosomes together, 
+and found that, remarkably, 
+the inversion on chromosome 8 is still the most striking outlying signal (Figure \ref{sfig:mds1_along_allchr_human}). 
+Further investigation with a denser set of SNPs,
+allowing a finer genomic resolution,
+may yield other patterns.
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
+    \end{center}
+    \caption{
+         Variation in structure between windows on human chromosomes 8, 15, and 17. 
+         Each point in each plot represents a window.
+         The first column shows the MDS visualization of relationships between windows;
+         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
+         Rows, from top to bottom show chromosomes 8, 15, and 17. 
+         The vertical red lines show the breakpoints of known inversions from \citet{antonacci2009characterization}.
+        \label{fig:mds_human}
+    }
+\end{figure}
+
+%%%%%%%%%%%
+\subsection{\textit{Medicago truncatula}}
+
+Unlike the other two species,
+the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
+showed similar patterns of gradual change in population structure across each chromosome,
+with no indications of chromosome-specific patterns.
+This consistency suggests that the factor driving the population structure for each chromosome is the same,
+as might be caused by varying strengths of linked selection.
+To verify that variation in population structure is shared across chromosomes,
+we applied the method to all chromosomes together.
+Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
+and other chromosomes are similar:
+% the two types of population structure with higher gene density 
+% (colored green and purple in Figure \ref{fig:mds12_medicago})
+% are nearly identical across all chromosomes,
+% but population structure associated with low gene density (colored orange)
+% shows more chromsome-specific patterns (see Supplemental Figure \ref{sfig:pca_drosophila_allchr}).
+across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
+which often have lower gene density and may therefore be less subject to linked selection.
+% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
+To verify that this is a possible explanation,
+we computed gene densities in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
+which are shown juxtaposed with
+the first MDS coordinate of each window in Figure \ref{fig:mds_medicago},
+and are significantly correlated, as shown in Figure \ref{fig:mds_gene_count}.
+(Values shown are the number of start and end positions of each predicted mRNA transcript,
+divided by two, assigned to the nearest window.)
+% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
+However, other genomic features, such as distance to centromere show roughly the same patterns,
+so we cannot rule out alternative hypotheses.
+In particular, the recombination rates estimated by \citet{paape2012finescale} appear visually to be similar,
+but were not readily available for comparison.
+
+We also found nearly identical results when choosing shorter windows of 1,000 SNPs;
+or choosing windows of equal length in base pairs rather than SNPs.
+Similarly, the results were not substantially changed
+when using weighted PCA to downweight the large group of Tunisian samples.
+
+\begin{figure}
+    \begin{center}
+       \includegraphics{Fig6_Together_MDS_plot_chr3_final}
+    \end{center}
+    \caption{
+       MDS visualization of population structure on \textit{M.~truncatula} chromosome 3. 
+       Each point in the plot represents a window;
+       the structure revealed by the MDS plot is strongly clustered along the chromosome,
+       with windows in the upper-right corner of the MDS plot (colored red) clustered around the centromere,
+       windows in the upper-left corner (purple) furthest from the centromere,
+       and the remaining corner (green) intermediate.
+       Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:mds_medicago_allchr}.
+       \label{fig:mds12_medicago}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
+    \end{center}
+    \caption{
+        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
+        Each point corresponds to a sample, colored by country of origin.
+        Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:pca_peaks_medicago_allchr}.
+        \label{fig:pca_medicago}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics{Fig5_MDS_and_gene_count_allchr_update_correction}
+    \end{center}
+    \caption{
+         MDS coordinate and gene density for each window in the \textit{Medicago} genome,
+         for chromosomes 1--8 (numbered above each pair of figures).
+         For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
+         The black plot below is gene count for each window against the position of each window.
+         \label{fig:mds_medicago}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics{MDS_1D_against_gene_count_all_chr_win104_with_lm_update}
+    \end{center}
+    \caption{
+        First MDS coordinate against gene density for all 8 chromosomes of \textit{M.~truncatula}.
+        The first MDS coordinate is significantly correlated with gene count ($r=0.149$, $p=2.2\times 10^{-16}$). 
+        \label{fig:mds_gene_count}
+    }
+\end{figure}
+
+%%%%%%% %%%%%%%%%
+\section{Discussion}
+
+Our investigations have found substantial variation in population structure across the genomes
+of three diverse species,
+revealing distinct biological processes driving this variation in each species.
+More investigation, particularly on more species and datasets, will help to uncover which patterns are generalizable.
+With growing appreciation of the heterogeneous effects of selection across the genome,
+especially the importance of adaptive introgression, hybrid speciation \citep{pool2015natural,brandvain2014speciation,hufford2013genomic,fitzpatrick2010rapid,staubach2012genome},
+local adaptation \citep{lenormand2002limits,wang2014isolation},
+and inversion polymorphisms \citep{kirkpatrick2015chromosome,kirkpatrick2010chromosome},
+local PCA may prove to be a useful exploratory tool to discover important genomic features.
+% It is unclear whether the technique will be useful on reduced representation genotyping datasets 
+% due to marker density and issues with missing data --
+% our investigations with one such dataset were inconclusive --
+% but even low coverage, whole-genome sequence is very promising.
+
+We now discuss possible implications of this variation in population structure,
+the impact of various parameter choices in implementing the method,
+and possible additional applications.
+
+\paragraph{Chromosomal inversions}
+A major driver of variation in population structure in two datasets we examined are inversions.
+This may be common,
+but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
+PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
+however, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
+or simply regions of reduced recombination
+without additional lines of evidence.
+However, in many applications, inversions are a nuisance.
+For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
+by regressing out the effect of linked SNPs on each other.
+Removing samples with the less common orientation of each inversion reduced,
+but did not eliminate, the signal of inversions
+seen in the \textit{Drosophila melanogaster} dataset,
+demonstrating that the genomic effects of transiently polymorphic inversions
+may outlast the inversions themselves.
+
+\paragraph{The effect of selection}
+It seems clear that the variation in population structure we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
+is likely due to selection.
+Furthermore, 
+the selection must be affecting many targets across the genome,
+since we see similar effects across long distances (even distinct chromosomes).
+For this reason, the most likely candidate may be selection against linked deleterious mutations, known as ``background selection'' \citep{charlesworth1993effect,charlesworth2013background}.
+Background selection can be thought of as reducing the number of potential contributors to the gene pool in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious};
+for this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently,
+Regions of the genome harboring many targets of local adaptation may show similar patterns,
+since migrant alleles in these regions will be selected against,
+and so locally gene trees will more closely reflect spatial proximity.
+
+A third possibility is that variation in population structure is due to recent admixture between previously separated populations,
+the effects of which were not uniform across the genome due to selection.
+For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
+is due to selection against Neanderthal genes, leading to greater introgression in regions of lower gene density
+\citep{harris2016genetic,juric2016strength}.
+African \textit{Drosophila melanogaster} are known to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
+if selection regularly favors genes from one origin,
+this could lead to substantial variation in population structure correlated with local gene density.
+
+There has been substantial debate over the relative impacts of different forms of selection.
+These have been difficult to disentangle in part because most concrete theoretical predictions
+come in the form of predicted forms of the allele frequency spectrum which are only strictly valid in randomly mating (i.e., unstructured) populations,
+and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
+It is possible that statistics making stronger use of spatial information
+will have more luck.
+\plr{expand}
+
+% How do we expect selection to affect population structure?
+% PCA summarizes patterns in kinship found in the genetic covariance matrix,
+% which is an average across locus-specific genealogies:
+% if individuals are closer in the genealogies of a given genomic region, 
+% they tend to be closer in the PCA maps for that region. 
+% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
+% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
+% Selective sweeps cause local recent ancestry or short trees.
+% Background selection causes shallow ones.
+% Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
+% characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
+% A genomic region with many targets for selection experiences
+% background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
+% which would tend to shorten genealogical trees in the region,
+% similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
+% % Balancing selection causes deep trees.
+% On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
+% and population structure locally describes which individuals have which alleles
+% rather than geographical proximity.
+% Finally, 
+% since recombination is suppressed between opposite orientations of a chromosomal inversion
+% near its breakpoints,
+% genealogies in these regions separate samples carrying the two orientations of the inversion.
+% In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
+% (indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
+% as a way to identify inversions).
+% if a chromosomal inversion is polymorphic in the sample, 
+% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints.
+Recombination suppression across inversions thus results in different genome structure and population structure.
+% Other effects, like noise, introgression might also influence population structure.
+% 
+% Many of the effects listed above (such as single selective sweeps or inversions) 
+% are not expected to have similar effects on population structure in different regions of the genome
+% because of randomness in which samples end up in which group;
+% but if these coincide with a region of reduced recombination (or an inversion),
+% these could drive major patterns of variation.
+% The more subtle effects of genome-wide linked selection could be shared across large regions
+% due to large-scale variation in gene density
+% (as by background selection or local adaptation at many genes).
+
+\paragraph{Parameter choices}
+There are several choices in the method that may affect the results.
+As with whole-genome PCA,
+the choice of samples is important,
+as variation not strongly represented in the sample will not be discovered.
+The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
+but downweighting may be a better option that does not discard data
+(and here we present a method to do this).
+Next, the choice of window size may be important,
+although in our applications results were not sensitive to this,
+indicating that we can see variation on a sufficiently fine scale.
+Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in Figure \ref{fig:diagram}),
+along with the method used to discover common structure,
+will affect results.
+We used MDS, applied to either each chromosome separately or to the entire genome;
+for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosome,
+but genome-wide, their signal is obscured by the numerous other signals of comparable strength.
+
+Besides window length, there is also the question of how to choose windows.
+In these applications we have used nonoverlapping windows with equal numbers of polymorphic sites.
+Alternatively, windows could be chosen to have equal length in genetic distance,
+so that each would have roughly the same number of independent trees.
+However, we found little change in results when using different window sizes
+or when measuring windows in physical distance (in bp).
+
+More generally, there are many possible methods to discover common structure in different parts of the genome:
+in this work we use PCA to summarize population structure,
+but other methods, such as STRUCTURE \citep{falush2003inference},
+SPA \citep{yang2012modelbased},
+SpaceMix \citep{bradburd2015spatial},
+or other matrix factorization methods \citep{engelhardt2010analysis}
+may highlight different sources of variation in the data.
+It is also possible that other methods for measuring dissimilarity between windows' covariance matrices
+or for summarizing the matrix of pairwise distances between windows
+would lead to different insights.
+
+Finally, our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
+and how many MDS coordinates to use when describing the distance matrix between windows,
+but in our exploration, changing these has not produced dramatically different results.
+These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
+we encourage the user to experiment.
+
+\paragraph{Applications}
+So-called cryptic relatedness between samples
+has been one of the major sources of confounding in genome-wide association studies (GWAS)
+and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
+Since population structure is not constant along the genome,
+this could in principle lead to an inflation of false positives in parts of the genome
+with stronger population structure than the genome-wide average.
+A method such as ours might be used to provide a more sensitive correction.
+Fortunately, in our human dataset this does not seem likely to have a strong effect:
+most variation is due to small, independent regions, possibly primarily inversions,
+and so may not have a major effect on GWAS.
+In the other species we examined, particularly \textit{Drosophila melanogaster},
+treating population structure as a single quantity would entail a substantial loss of power,
+and could potentially be misleading.
+
+%%%%%%%%
+\subsection*{Acknowledgements}
+
+We are indebted to John Pool, Russ Corbett-Detig, Matilde Cordeiro, and Peter Chang 
+for assistance with obtaining data and interpreting results
+(especially inversion status of \textit{D.~melanogaster} samples).
+Thanks also go to Yaniv Brandvain, Barbara Engelhardt, Charles Langley, Graham Coop, and Jeremy Berg for helpful comments
+and for encouraging the project.
+
+%%%%%%%%%%%%%%%%%
+\section{Methods}
+
 \subsection{PCA in genomic windows}
@@
 (``windows'')
-% recoded matrix into contigous matrices that have the same columns but fewer rows than the original matrix, 
+% recoded matrix into contiguous matrices that have the same columns but fewer rows than the original matrix, 
 and applied Principal Component Analysis (PCA) as described in \citet{mcvean2009genealogical}
@@
 Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs;
-PCA plots for each arm are shown in Figure \ref{fig:pca_drosophila_allchr}.
+PCA plots for each arm are shown in Figure \ref{sfig:pca_drosophila_allchr}.
 
@@
 We excluded the sex chromosomes and the mitochondria.
-PCA plots for each chromosome, separately, are shown in Figure \ref{fig:pca_human_allchr}.
+PCA plots for each chromosome, separately, are shown in Figure \ref{sfig:pca_human_allchr}.
 % We use the allele that has highest frequency in the samples as the reference allele for each position.
@@
 Each of the 8 chromosomes has 3--5 million SNPs;
-PCA plots for these are shown in Figure \ref{fig:pca_medicago_allchr}.
+PCA plots for these are shown in Figure \ref{sfig:pca_medicago_allchr}.
 We did not use the mitochondria or chloroplasts.
@@
 \end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
-
-In these three species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
-This implies that variation is due to meaningful variation in population structure, 
-since noise due to randomness in choice of local genealogical trees
-are not expected to show long distance correlations. 
-Below, we discuss the results and likely underlying causes.
-
-%%%%%%%%%%%
-\subsection{\textit{Drosophila melanogaster}}
-
-We ran the method on chromosome arms 2L, 2R, 3L, 3R and X separately. 
-For each, the two-dimensional MDS visualization resembles a triangle,
-sometimes with additional points,
-as seen in Figure \ref{fig:mds_allchr}.
-Since the relative position of each window in this plot shows the similarity between windows, 
-this suggests that there are at least three extreme types of population structure 
-typified by windows found in the three corners of the triangle,
-and that other windows' population structure may be a mixture of those extremes. 
-\plr{make clear corners are arbitrary}
-
-\plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years, probably why it doesn't show up. 
-Also cite \citet{corbettdetig2012sequencebased} for a computational method for detecting inversions.}
-% corbettdetig2012sequencebased suggests these inversions are recent-ish (but before out-of-Africa)
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
-    \end{center}
-    \caption{
-        Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms.
-        In all plots, each point represents one window along the genome.
-         The first column shows the MDS visualization of relationships between windows, 
-         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
-         rows correspond to chromosome arms.
-         Colors are consistent for plots in each row. 
-         Vertical lines show the breakpoints of known polymorphic inversions.   
-         Solid black lines are for the inversions we used in Figure \ref{fig:pca_by_inversion},
-         while dotted grey lines are for other known inversions.     
-         \label{fig:mds_allchr}
-    }
-\end{figure}
-
-To investigate each extreme on a particular chromosome arm, 
-we selected an ``extreme'' window in the MDS plot
-and the 5\% of windows that are closest to it in the MDS coordinates,
-then highlighted these windows' positions along the genome,
-and created PCA plots for the windows, combined.
-Representative plots are shown for three groups of windows on each chromosome arm
-in Figure \ref{fig:mds_allchr} (groups are shown in color)
-in Figure \ref{fig:pca_by_pop} (PCA plots).
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
-    \end{center}
-    \caption{      
-        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_allchr},
-        on each chromosome arm of \textit{Drosophila melanogaster}.
-        In all plots, each point represents a sample. 
-        The first column shows the combined PCA plot for windows 
-        whose points are colored green in Figure \ref{fig:mds_allchr}; 
-        the second is for orange windows; 
-        and third is for purple windows.
-         \label{fig:pca_by_pop}
-    }
-\end{figure}
-
-The variation in population structure turns out to be almost entirely explained by
-several large inversions that are polymorphic in these samples, 
-disussed in \citet{corbett2012population} and \citet{langley2012genomic}.
-% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
-To depict this, Figure \ref{fig:pca_by_inversion} shows
-the PCA plots in Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
-Taking chromosome arm 2L as an example,
-the two regions of similar, extreme population structure
-shown in green in the first row of Figure \ref{fig:mds_allchr}
-lie directly around the breakpoints of the inversion In(2L)t,
-and the PCA plots in the first rows of Figure \ref{fig:pca_by_inversion}
-shows that population structure here is mostly determined by inversion orientation.
-The regions shown in purple on chromosome 2L lie near the centromere,
-and have population structure reflective of two axes of variation,
-seen in Figure \ref{fig:pca_by_pop},
-which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
-(see \citet{lack2015drosophila} for more about admixture in this sample).
-The regions shown in orange on chromosome 2L mostly lie inside the inversion,
-and show population structure that is a mixture between the other two,
-as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
-Similar results are found in other chromosome arms,
-albeit complicated by the coexistence of more than one polymorphic inversion;
-however, each breakpoint visibly affects patterns in the MDS coordinates.
-% except that the situation of Chr3R is a little more complicated,
-% due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
-% However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
-% although we have not highlighted portions of the MDS plot that pick these out.
-% The inversions In(3L)P and In(3R)Mo, are at low frequency in these samples, 
-% and visibly affect the MDS plots although less than the more common inversion on each arm.
-% The very young inversion In(X)be shows little signal 
-% compared to the older inversion In(X)A \citep{corbett2012population}.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
-    \end{center}
-    \caption{
-         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
-         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
-         \citep[data from][]{lack2015drosophila}
-        \label{fig:pca_by_inversion}
-    }
-\end{figure}
-
-To see how population structure varies without the effect of polymorphic inversions,
-we then performed the same analyses after removing, for each chromosome arm,
-the samples having the less frequent orientation of any inversion on that chromosome arm.
-The result showed substantially less variation in MDS axes,
-although some variation associated with inversion breakpoints was still visible,
-as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions}.
-For instance, the majority of the variation along 3L in Figure \ref{fig:mds_allchr}
-is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
-there is also a relatively small dip on the right end of the arm (near the centromere).
-In contrast, Supplemental Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
-remaining structure is dominated by the dip near the centromere.
-The remaining variation varies systematically along the chromosome,
-following roughly similar patterns to that seen in recombination rate and diversity in \citet{langley2012genomic} and \citet{mackay2012drosophila}.
-Indeed, correlations between the recombination rate \plr{XXX} near each window and the position on the first MDS coordinate are highly significant
-(Figure \ref{fig:drosophila_recomb_rate}).
-This is consistent with the hypothesis that variation in population structure
-is due to selection, since the strength of linked selection increases with local gene density, measured in units of recombination distance.
-However, the number of genes -- measured as the number of transcription start and end sites within each window (obtained from the RefGene table, UCSC genome browser) --
-was more weakly correlated with MDS coordinate (Supplemental Figures \ref{sfig:drosophila_gene_density_plots} and \ref{sfig:drosophila_gene_density_correlations}).
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
-    \end{center}
-    \caption{
-         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
-         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
-         \citep[data from][]{lack2015drosophila}
-        \label{fig:pca_by_inversion}
-    }
-\end{figure}
-
-\plr{gene density from RefSeq Genes from UCSC, showing genes from NCBI RNA references sequence collections. we use txStart and txEnd (transcription start/end).}
-
-%%%%%%%%%%%
-\subsection{Human}
-
-As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes.
-On each, variation in population structure was dominated by a small number of windows
-having similar population structure to each other that differed dramatically from the rest of the chromosome.
-These may be primarily inversions: outlying windows coincide with
-three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
-notably a particularly large, polymorphic inversion on 8p23 (Figure \ref{fig:mds_human}). 
-% for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}) 
-Similar plots for all chromosomes are shown
-in Supplementary Figures \ref{fig:mds12_chr1_8_human}, \ref{fig:mds12_chr9_16_human}, \ref{fig:mds12_chr17_22_human}.
-PCA plots of many outlying windows show a characteristic trimodal shape 
-(shown for chromosome 8 in Figure \ref{fig:pca_chr8_outliers}),
-presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
-(although we do not have data on orientation status).
-This trimodal shape has been proposed as a method to identify inversions \citep{ma2012investigation},
-but distinguishing this hypothesis from others,
-such as regions of low recombination rate,
-would require additional data.
-
-We also ran the method grouping all 22 autosomes together, 
-and found that, remarkably, 
-the inversion on chromosome 8 is still the most striking outlying signal (Figure \ref{fig:mds1_along_allchr_human}). 
-Further investigation with a denser set of SNPs,
-allowing a finer genomic resolution,
-may yield other patterns.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
-    \end{center}
-    \caption{
-         Variation in structure between windows on human chromosomes 8, 15, and 17. 
-         Each point in each plot represents a window.
-         The first column shows the MDS visualization of relationships between windows;
-         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
-         Rows, from top to bottom show chromosomes 8, 15, and 17. 
-         The vertical red lines show the breakpoints of known inversions from \citet{antonacci2009characterization}.
-        \label{fig:mds_human}
-    }
-\end{figure}
-
-%%%%%%%%%%%
-\subsection{\textit{Medicago truncatula}}
-
-Unlike the other two species,
-the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
-showed similar patterns of gradual change in population structure across each chromosome,
-with no indications of chromosome-specific patterns.
-This consistency suggests that the factor driving the population structure for each chromosome is the same,
-as might be caused by varying strengths of linked selection.
-To verify that variation in population structure is shared across chromosomes,
-we applied the method to all chromosomes together.
-Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
-and other chromosomes are similar:
-% the two types of population structure with higher gene density 
-% (colored green and purple in Figure \ref{fig:mds12_medicago})
-% are nearly identical across all chromosomes,
-% but population structure associated with low gene density (colored orange)
-% shows more chromsome-specific patterns (see Supplemental Figure \ref{fig:pca_drosophila_allchr}).
-across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
-which often have lower gene density and may therefore be less subject to linked selection.
-% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
-To verify that this is a possible explanation,
-we computed gene densities in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
-which are shown juxtaposed with
-the first MDS coordinate of each window is shown in Figure \ref{fig:mds_medicago},
-and are significantly correlated, as shown in Figure \ref{fig:mds_gene_count}.
-(Values shown are the number of start and end positions of each predicted mRNA transcript,
-divided by two, assigned to the nearest window.)
-% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
-Other genomic features covary with gene density, however,
-so we cannot rule out alternative hypotheses.
-
-We also found nearly identical results when choosing shorter windows of 1,000 SNPs;
-or choosing windows of equal length in base pairs rather than SNPs.
-Similarly, the results were not substantially changed
-when using weighted PCA to downweight the large group of Tunisian samples.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig6_Together_MDS_plot_chr3_final}
-    \end{center}
-    \caption{
-       MDS visualization of population structure on \textit{M.~truncatula} chromosome 3. 
-       Each point in the plot represents a window;
-       the structure revealed by the MDS plot is strongly clustered along the chromosome,
-       with windows in the upper-right corner of the MDS plot (colored red) clustered around the centromere,
-       windows in the upper-left corner (purple) furthest from the centromere,
-       and the remaining corner (green) intermediate.
-       Plots for remaining chromosomes are shown in Supplemental Figure \ref{fig:mds_medicago_allchr}.
-       \label{fig:mds12_medicago}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
-    \end{center}
-    \caption{
-        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
-        Each point corresponds to a sample, colored by country of origin.
-        Plots for remaining chromosomes are shown in Supplemental Figure \ref{fig:pca_peaks_medicago_allchr}.
-        \label{fig:pca_medicago}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig5_MDS_and_gene_count_allchr_update_correction}
-    \end{center}
-    \caption{
-         MDS coordinate and gene density for each window in the \textit{Medicago} genome,
-         for chromosomes 1--8 (numbered above each pair of figures).
-         For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
-         The black plot below is gene count for each window against the position of each window.
-         \label{fig:mds_medicago}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{MDS_1D_against_gene_count_all_chr_win104_with_lm_update}
-    \end{center}
-    \caption{
-        First MDS coordinate against gene density for all 8 chromosomes of \textit{M.~truncatula}.
-        The first MDS coordinate is significantly correlated with gene count ($r=0.149$, $p=2.2\times 10^{-16}$). 
-        \label{fig:mds_gene_count}
-    }
-\end{figure}
-
-%%%%%%% %%%%%%%%%
-\section{Discussion}
-
-Our investigations have found substantial variation in population structure across the genomes
-of three diverse species,
-revealing distinct biological processes driving this variation in each species.
-More investigation, particularly on more species and datasets, will help to uncover which patterns are generalizable.
-With growing appreciation of the heterogeneous effects of selection across the genome,
-especially the importance of adaptive introgression, hybrid speciation \citep{pool2015natural,brandvain2014speciation,hufford2013genomic,fitzpatrick2010rapid,staubach2012genome},
-local adaptation \citep{lenormand2002limits,wang2014isolation},
-and inversion polymorphisms \citep{kirkpatrick2015chromosome,kirkpatrick2010chromosome},
-local PCA may prove to be a useful exploratory tool to discover important genomic features.
-% It is unclear whether the technique will be useful on reduced representation genotyping datasets 
-% due to marker density and issues with missing data --
-% our investigations with one such dataset were inconclusive --
-% but even low coverage, whole-genome sequence is very promising.
-
-We now discuss possible implications of this variation in population structure,
-the impact of various parameter choices in implementing the method,
-and possible additional applications.
-
-\paragraph{Chromosomal inversions}
-A major driver of variation in population structure in two datasets we examined are inversions.
-This may be common,
-but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
-PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
-however, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
-or simply regions of reduced recombination
-without additional lines of evidence.
-However, in many applications, inversions are a nuisance.
-For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
-by regressing out the effect of linked SNPs on each other.
-Removing samples with the less common orientation of each inversion reduced,
-but did not eliminate, the signal of inversions
-seen in the \textit{Drosophila melanogaster} dataset,
-demonstrating that the genomic effects of transiently polymorphic inversions
-may outlast the inversions themselves.
-
-\paragraph{The effect of selection}
-It seems clear that the variation in population structure we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
-is likely due to selection.
-Furthermore, 
-the selection must be affecting many targets across the genome,
-since we see similar effects across long distances (even distinct chromosomes).
-For this reason, the most likely candidate may be selection against linked deleterious mutations, known as ``background selection'' \citep{charlesworth1993effect,charlesworth2013background}.
-Background selection can be thought of as reducing the number of potential contributors to the gene pool in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious};
-for this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently,
-Regions of the genome harboring many targets of local adaptation may show similar patterns,
-since migrant alleles in these regions will be selected against,
-and so locally gene trees will more closely reflect spatial proximity.
-
-A third possibility is that variation in population structure is due to recent admixture between previously separated populations,
-the effects of which were not uniform across the genome due to selection.
-For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
-is due to selection against Neanderthal genes, leading to greater introgression in regions of lower gene density
-\citep{harris2016genetic,juric2016strength}.
-African \textit{Drosophila melanogaster} are known to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
-if selection regularly favors genes from one origin,
-this could lead to substantial variation in population structure correlated with local gene density.
-
-There has been substantial debate over the relative impacts of different forms of selection.
-These have been difficult to disentangle in part because most concrete theoretical predictions
-come in the form of predicted forms of the allele frequency spectrum which are only strictly valid in randomly mating (i.e., unstructured) populations,
-and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
-It is possible that statistics making stronger use of spatial information
-will have more luck.
-\plr{expand}
-
-% How do we expect selection to affect population structure?
-% PCA summarizes patterns in kinship found in the genetic covariance matrix,
-% which is an average across locus-specific genealogies:
-% if individuals are closer in the genealogies of a given genomic region, 
-% they tend to be closer in the PCA maps for that region. 
-% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
-% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
-% Selective sweeps cause local recent ancestry or short trees.
-% Background selection causes shallow ones.
-% Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
-% characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
-% A genomic region with many targets for selection experiences
-% background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
-% which would tend to shorten genealogical trees in the region,
-% similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
-% % Balancing selection causes deep trees.
-% On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
-% and population structure locally describes which individuals have which alleles
-% rather than geographical proximity.
-% Finally, 
-% since recombination is suppressed between opposite orientations of a chromosomal inversion
-% near its breakpoints,
-% genealogies in these regions separate samples carrying the two orientations of the inversion.
-% In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
-% (indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
-% as a way to identify inversions).
-% if a chromosomal inversion is polymorphic in the sample, 
-% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints.
-Recombination suppression across inversions thus results in different genome structure and population structure.
-% Other effects, like noise, introgression might also influence population structure.
-% 
-% Many of the effects listed above (such as single selective sweeps or inversions) 
-% are not expected to have similar effects on population structure in different regions of the genome
-% because of randomness in which samples end up in which group;
-% but if these coincide with a region of reduced recombination (or an inversion),
-% these could drive major patterns of variation.
-% The more subtle effects of genome-wide linked selection could be shared across large regions
-% due to large-scale variation in gene density
-% (as by background selection or local adaptation at many genes).
-
-\paragraph{Parameter choices}
-There are several choices in the method that may affect the results.
-As with whole-genome PCA,
-the choice of samples is important,
-as variation not strongly represented in the sample will not be discovered.
-The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
-but downweighting may be a better option that does not discard data
-(and here we present a method to do this).
-Next, the choice of window size may be important,
-although in our applications results were not sensitive to this,
-indicating that we can see variation on a sufficiently fine scale.
-Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in Figure \ref{fig:diagram}),
-along with the method used to discover common structure,
-will affect results.
-We used MDS, applied to either each chromosome separately or to the entire genome;
-for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosome,
-but genome-wide, their signal is obscured by the numerous other signals of comparable strength.
-
-Besides window length, there is also the question of how to choose windows.
-In these applications we have used nonoverlapping windows with equal numbers of polymorphic sites.
-Alternatively, windows could be chosen to have equal length in genetic distance,
-so that each would have roughly the same number of independent trees.
-However, we found little change in results when using different window sizes
-or when measuring windows in physical distance (in bp).
-
-More generally, there are many possible methods to discover common structure in different parts of the genome:
-in this work we use PCA to summarize population structure,
-but other methods, such as STRUCTURE \citep{falush2003inference},
-SPA \citep{yang2012modelbased},
-SpaceMix \citep{bradburd2015spatial},
-or other matrix factorization methods \citep{engelhardt2010analysis}
-may highlight different sources of variation in the data.
-It is also possible that other methods for measuring dissimilarity between windows' covariance matrices
-or for summarizing the matrix of pairwise distances between windows
-would lead to different insights.
-
-Finally, our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
-and how many MDS coordinates to use when describing the distance matrix between windows,
-but in our exploration, changing these has not produced dramatically different results.
-These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
-we encourage the user to experiment.
-
-\paragraph{Applications}
-So-called cryptic relatedness between samples
-has been one of the major sources of confounding in genome-wide association studies (GWAS)
-and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
-Since population structure is not constant along the genome,
-this could in principle lead to an inflation of false positives in parts of the genome
-with stronger population structure than the genome-wide average.
-A method such as ours might be used to provide a more sensitive correction.
-Fortunately, in our human dataset this does not seem likely to have a strong effect:
-most variation is due to small, independent regions, possibly primarily inversions,
-and so may not have a major effect on GWAS.
-In the other species we examined, particularly \textit{Drosophila melanogaster},
-treating population structure as a single quantity would entail a substantial loss of power,
-and could potentially be misleading.
-
-%%%%%%%%
-\subsection*{Acknowledgements}
-
-We are indebted to John Pool, Russ Corbett-Detig, Matilde Cordeiro, and Peter Chang 
-for assistence with obtaining data and interpreting results
-(especially inversion status of \textit{D.~melanogaster} samples).
-Thanks also go to Yaniv Brandvain, Barbara Engelhardt, Charles Langley, Graham Coop, and Jeremy Berg for helpful comments
-and for encouraging the project.
 
@@
     \begin{center}
-       \includegraphics[width=1\textwidth]{FigS_MDS1D_plot_all_chr_human}
-    \end{center}
-    \caption{
-        MDS visualization of variation in population structure amongst windows across \emph{all} human autosomes simultaneously.
-        The small group of windows with positive outlying MDS values lie around the inversion at 8p23.
-        \label{fig:mds1_along_allchr_human}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr1_8_all_valid_and_pre}
-    \end{center}
-    \caption{
-         MDS plots for human chromosome 1-8.
-         The first column shows the MDS visualization of relationships between windows, 
-         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
-         rows correspond to chromosomes.        
-         Colorful vertical lines show the breakpoints of known valid inversions, 
-         while grey vertical lines show the breakpoints of predicted inversions.
-        \label{fig:mds12_chr1_8_human}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr9_16_all_valid_and_pre}
-    \end{center}
-    \caption{
-        MDS plots for human chromosome 9-16.
-        \label{fig:mds12_chr9_16_human}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr17_22_all_valid_and_pre}
-    \end{center}
-    \caption{
-        MDS plots for human chromosome 17-22.
-        \label{fig:mds12_chr17_22_human}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{FigS_PCA_chr8_outliers_combine_update}
-    \end{center}
-    \caption{
-        PCA plots comparison for outlier windows and non-outlier windows.
-        \label{fig:pca_chr8_outliers}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr}
-    \end{center}
-    \caption{
-        MDS visualizations of population structure for all 8 chromosomes of the \textit{Medicago truncatula} data.
-        \label{fig:mds_medicago_allchr}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.82\textwidth]{FigS_pca_plots_for_Medicago_allchr_3peaks_byMDS}
-    \end{center}
-    \caption{
-        PCA plots for regions colored in Figure \ref{fig:mds_medicago_allchr}
-        on all 8 chromosomes of \textit{Medicago truncatula}:
-        (A) green, (B) orange, and (C) purple.
-        \label{fig:pca_peaks_medicago_allchr}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
        \includegraphics[width=1\textwidth]{FigS_pca_plots_allchr_drosophila}
@@
         PCA plots for chromosome arms 2L, 2R, 3L, 3R and X of the \textit{Drosophila melanogaster} dataset.
-        \label{fig:pca_drosophila_allchr}
+        \label{sfig:pca_drosophila_allchr}
     }
@@
         PCA plots for all 22 huan autosomes from the POPRES data.
-        \label{fig:pca_human_allchr}
+        \label{sfig:pca_human_allchr}
     }
@@
         PCA plots for all 8 chromosomes in the \textit{Medicago truncatula} dataset.
-        \label{fig:pca_medicago_allchr}
+        \label{sfig:pca_medicago_allchr}
     }
@@
         Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms: without inversions.
-         As in Figure \label{fig:mds_allchr}, but after omitting for each chromosome arm individuals carrying the less frequent orientation
+         As in Figure \ref{fig:mds_allchr}, but after omitting for each chromosome arm individuals carrying the less frequent orientation
          of any inversions on that chromosome arm.
@@
 % F-statistic: 14.34 on 1 and 286 DF,  p-value: 0.0001862
-
         \label{sfig:drosophila_gene_density_correlations}
@@
 
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr1_8_all_valid_and_pre}
+    \end{center}
+    \caption{
+         MDS plots for human chromosome 1-8.
+         The first column shows the MDS visualization of relationships between windows, 
+         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
+         rows correspond to chromosomes.        
+         Colorful vertical lines show the breakpoints of known valid inversions, 
+         while grey vertical lines show the breakpoints of predicted inversions.
+        \label{sfig:mds12_chr1_8_human}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr9_16_all_valid_and_pre}
+    \end{center}
+    \caption{
+        MDS plots for human chromosome 9-16.
+        \label{sfig:mds12_chr9_16_human}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr17_22_all_valid_and_pre}
+    \end{center}
+    \caption{
+        MDS plots for human chromosome 17-22.
+        \label{sfig:mds12_chr17_22_human}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{FigS_PCA_chr8_outliers_combine_update}
+    \end{center}
+    \caption{
+        PCA plots comparison for outlier windows and non-outlier windows.
+        \label{sfig:pca_chr8_outliers}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=1\textwidth]{FigS_MDS1D_plot_all_chr_human}
+    \end{center}
+    \caption{
+        MDS visualization of variation in population structure amongst windows across \emph{all} human autosomes simultaneously.
+        The small group of windows with positive outlying MDS values lie around the inversion at 8p23.
+        \label{sfig:mds1_along_allchr_human}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr}
+    \end{center}
+    \caption{
+        MDS visualizations of population structure for all 8 chromosomes of the \textit{Medicago truncatula} data.
+        \label{sfig:mds_medicago_allchr}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_pca_plots_for_Medicago_allchr_3peaks_byMDS}
+    \end{center}
+    \caption{
+        PCA plots for regions colored in Figure \ref{sfig:mds_medicago_allchr}
+        on all 8 chromosomes of \textit{Medicago truncatula}:
+        (A) green, (B) orange, and (C) purple.
+        \label{sfig:pca_peaks_medicago_allchr}
+    }
+\end{figure}
+
 \end{document}

### 2016-08-12 writeup/local_pca_paper.tex
pop struct -> patterns of relatedness or effects of pop struct

@@
 
-A note on nomenclature: In this work we describe variation in population structure using local PCA,
+A note on nomenclature: In this work we describe variation in patterns of relatedness using local PCA,
 where ``local'' refers to proximity along the genome.
@@
 
-Population genetics studies often describe variation in particular statistics along the genome,
-often with the aim of detecting particular signals.
-In this paper, we use a visualization technique to obtain an unbiased view
-of how patterns of kinship vary along the genome,
-in genome-wide, spatially distributed datasets for three species,
-to start to discover common patterns and differences.
-
 As reviewed above,
@@
 how patterns of mean relatedness vary systematically along the genome, with a focus on large, geographically distributed populations.
+Geographic population structure sets the stage by establishing similar patterns of relatedness across much of the genome;
+we then aim to describe how this structure is affected by selection and other factors.
 Our aim is not to identify outlier loci,
@@
 (1) divide the genome into windows,
-(2) summarize the patterns of relatedness (population structure) in each window,
-(3) measure dissimilarity in population structure between each pair of windows,
+(2) summarize the patterns of relatedness in each window,
+(3) measure dissimilarity in relatedness between each pair of windows,
 (4) visualize the resulting dissimilarity matrix,
-and (5) combine similar windows to more accurately estimate local population structure.
+and (5) combine similar windows to more accurately visualize local effects of population structure.
 Details of how we carried these out are given in the Methods.
@@
 In these three species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
-This implies that variation is due to meaningful variation in population structure, 
+This implies that variation is due to meaningful variation in the effects of population structure, 
 since noise due to randomness in choice of local genealogical trees
@@
 Since the relative position of each window in this plot shows the similarity between windows, 
-this suggests that there are at least three extreme types of population structure 
+this suggests that there are at least three extreme manifestations of population structure 
 typified by windows found in the three corners of the triangle,
-and that other windows' population structure may be a mixture of those extremes. 
+and that other windows' patterns of relatedness may be a mixture of those extremes. 
 \plr{make clear corners are arbitrary}
@@
 
-The variation in population structure turns out to be almost entirely explained by
+The variation in patterns of relatedness turns out to be almost entirely explained by
 several large inversions that are polymorphic in these samples, 
@@
 Taking chromosome arm 2L as an example,
-the two regions of similar, extreme population structure
+the two regions of similar, extreme patterns of relatedness
 shown in green in the first row of Figure \ref{fig:mds_allchr}
@@
 and the PCA plots in the first rows of Figure \ref{fig:pca_by_inversion}
-shows that population structure here is mostly determined by inversion orientation.
+shows that patterns of relatedness here are mostly determined by inversion orientation.
 The regions shown in purple on chromosome 2L lie near the centromere,
-and have population structure reflective of two axes of variation,
+and have patterns of relatedness reflective of two axes of variation,
 seen in Figure \ref{fig:pca_by_pop},
@@
 The regions shown in orange on chromosome 2L mostly lie inside the inversion,
-and show population structure that is a mixture between the other two,
+and show patterns of relatedness that are a mixture between the other two,
 as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
@@
 
-To see how population structure varies without the effect of polymorphic inversions,
+To see how patterns of relatedness vary without the effect of polymorphic inversions,
 we then performed the same analyses after removing, for each chromosome arm,
@@
 (Figure \ref{fig:drosophila_recomb_rate}).
-This is consistent with the hypothesis that variation in population structure
+This is consistent with the hypothesis that variation in the effects of population structure
 is due to selection, since the strength of linked selection increases with local gene density, measured in units of recombination distance.
@@
     \caption{
-        Population structure without inversions is correlated to recombination rate:
+        The effects of population structure without inversions is correlated to recombination rate:
         \plr{XXX}
@@
 As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes.
-On each, variation in population structure was dominated by a small number of windows
-having similar population structure to each other that differed dramatically from the rest of the chromosome.
+On each, variation in patterns of relatedness was dominated by a small number of windows
+having similar patterns of relatedness to each other that differed dramatically from the rest of the chromosome.
 These may be primarily inversions: outlying windows coincide with
@@
 the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
-showed similar patterns of gradual change in population structure across each chromosome,
+showed similar patterns of gradual change in patterns of relatedness across each chromosome,
 with no indications of chromosome-specific patterns.
-This consistency suggests that the factor driving the population structure for each chromosome is the same,
+This consistency suggests that the factor affecting the population structure for each chromosome is the same,
 as might be caused by varying strengths of linked selection.
-To verify that variation in population structure is shared across chromosomes,
+To verify that variation in the effects of population structure is shared across chromosomes,
 we applied the method to all chromosomes together.
@@
     \caption{
-       MDS visualization of population structure on \textit{M.~truncatula} chromosome 3. 
+       MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3. 
        Each point in the plot represents a window;
@@
 
-Our investigations have found substantial variation in population structure across the genomes
+Our investigations have found substantial variation in the patterns of relatedness formed by population structure across the genomes
 of three diverse species,
@@
 
-We now discuss possible implications of this variation in population structure,
+We now discuss possible implications of this variation in the effects of population structure,
 the impact of various parameter choices in implementing the method,
@@
 \paragraph{Chromosomal inversions}
-A major driver of variation in population structure in two datasets we examined are inversions.
+A major driver of variation in patterns of relatedness in two datasets we examined are inversions.
 This may be common,
@@
 \paragraph{The effect of selection}
-It seems clear that the variation in population structure we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
+It seems clear that the variation in patterns of relatedness we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
 is likely due to selection.
@@
 
-A third possibility is that variation in population structure is due to recent admixture between previously separated populations,
+A third possibility is that variation in patterns of relatedness is due to recent admixture between previously separated populations,
 the effects of which were not uniform across the genome due to selection.
@@
 if selection regularly favors genes from one origin,
-this could lead to substantial variation in population structure correlated with local gene density.
+this could lead to substantial variation in patterns of relatedness correlated with local gene density.
 
@@
 More generally, there are many possible methods to discover common structure in different parts of the genome:
-in this work we use PCA to summarize population structure,
+in this work we use PCA to visualize the effects of population structure,
 but other methods, such as STRUCTURE \citep{falush2003inference},
@@
 and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
-Since population structure is not constant along the genome,
+Since the effects of population structure is not constant along the genome,
 this could in principle lead to an inflation of false positives in parts of the genome
@@
 
-The top few principal components generally display population structure; 
+The top few principal components are generally good summaries of population structure; 
 we follow common practice and usually only use the first two (referred to as $\pcone$ and $\pctwo$).
@@
 
-\subsection{Similarity of population structure between windows}
-
-We think of local population structure as being summarized by \emph{relative} position of the samples
+\subsection{Similarity of patterns of relatedness between windows}
+
+We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
 in the space defined by the top principal components.
 However, 
-we do not compare population structure of different genomic by directly comparing the PCs,
-since rotations or reflections of these imply identical population structure.
+we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
+since rotations or reflections of these imply identical patterns of relatedness.
 Instead, we compare the low-dimensional approximations of the local covariance matrices
@@
 
-To measure the similarity of population structure for the $i^\text{th}$ window and $j^\text{th}$ window,
+To measure the similarity of patterns of relatedness for the $i^\text{th}$ window and $j^\text{th}$ window,
 we then use
@@
 give the arrangement in $m$-dimensional space that best recapitulates the original distance matrix.
-For results here, we use $m=2$ to produce one- or two-dimensional visualizations of relationships between windows' population structure.
-
-We then locate variation in population structure along the genome
+For results here, we use $m=2$ to produce one- or two-dimensional visualizations of relationships between windows' patterns of relatedness.
+
+We then locate variation in patterns of relatedness along the genome
 by choosing collections of windows that are nearby in MDS coordinates,
 and map their positions along the genome.
-Their common population structure is visualized by extracting the corresponding genomic regions
+A visualization of the effects of population structure across the entire colletion is formed by extracting the corresponding genomic regions
 and performing PCA on all, aggregated, regions.
@@
 In very short windows, genealogies of the samples will only be represented by a few trees,
-so variation between windows represents demographic noise rather than meaningful variation in population structure.
+so variation between windows represents demographic noise rather than meaningful variation in patterns of relatedness.
 Longer windows generally have more distinct trees (and SNPs), 
-allowing for less noisy estimation of local population structure.
-However, to better resolve meaningful signal, i.e., differences in population structure along the genome,
+allowing for less noisy estimation of local patterns of relatedness.
+However, to better resolve meaningful signal, i.e., differences in patterns of relatedness along the genome,
 we would like reasonably short windows.
@@
 
-Since we summarize population structure using relative positions in the principal component maps,
+Since we summarize patterns of relatedness using relative positions in the principal component maps,
 we quantify ``noise'' as the standard error of a sample's position on PC1 in a particular window,
@@
     \caption{
-        Gene number and population structure for \textit{Drosophila melanogaster}:
+        Gene number and the effects of population structure for \textit{Drosophila melanogaster}:
         The first MDS coordinate along the genome (black, above) is as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions};
@@
     \caption{
-        Gene number and population structure for \textit{Drosophila melanogaster}:
+        Gene number and the effects of population structure for \textit{Drosophila melanogaster}:
         this shows the first MDS coordinate and gene number, as in Supplementary Figure \ref{sfig:drosophila_gene_density_plots},
@@
     \caption{
-        MDS visualization of variation in population structure amongst windows across \emph{all} human autosomes simultaneously.
+        MDS visualization of variation in the effects of population structure amongst windows across \emph{all} human autosomes simultaneously.
         The small group of windows with positive outlying MDS values lie around the inversion at 8p23.
@@
     \caption{
-        MDS visualizations of population structure for all 8 chromosomes of the \textit{Medicago truncatula} data.
+        MDS visualizations of the effects of population structure for all 8 chromosomes of the \textit{Medicago truncatula} data.
         \label{sfig:mds_medicago_allchr}

### 2016-08-13 writeup/local_pca_paper.tex
plots for drosophila recomb and mds

@@
 (3) measure dissimilarity in relatedness between each pair of windows,
-(4) visualize the resulting dissimilarity matrix,
-and (5) combine similar windows to more accurately visualize local effects of population structure.
-Details of how we carried these out are given in the Methods.
+(4) visualize the resulting dissimilarity matrix using multidimensional scaling (MDS),
+and (5) combine similar windows to more accurately visualize local effects of population structure using PCA.
+Details of how we carried out each step are given in the Methods,
+and code to run the analyses is provided as an R package at \url{https://github.com/petrelharp/local_pca}.
 
@@
 
-We applied the method to genomic datasets with good geographic sampling
-of African populations of the fruit fly \textit{Drosophila melanogaster},
+We applied the method to genomic datasets with good geographic sampling:
+African populations of the fruit fly \textit{Drosophila melanogaster},
 a worldwide dataset of humans,
 and a range-wide dataset of the partially selfing weedy annual plant \textit{Medicago truncatula}.
-In these three species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
-This implies that variation is due to meaningful variation in the effects of population structure, 
+In all three species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
+This implies that variation is due to meaningful heterogeneity in a biological process,
 since noise due to randomness in choice of local genealogical trees

### 2016-08-13 dpgp/gene_density/get-recomb-rates.R

@@ -1,5 +1,5 @@
-
 # paste into http://petrov.stanford.edu/cgi-bin/recombination-rates_updateR5.pl?results=0&release=5.36&query_type=batch&Submit=Submit
+# The resulting recomb rates are in cM/Mbp.
 chroms <- c("2L","2R","3L","3R","X")
 lens <- structure(c(22963456L, 21142841L, 24536634L, 27894163L, 22418422L), .Dim = 5L, .Dimnames = list(c("2L", "2R", "3L", "3R", "X")))
 

### 2016-08-13 dpgp/gene_density/plot_gene_density.R

@@ -21,12 +21,18 @@
 # cdsEndStat 	cmpl	enum('none', 'unk', 'incmpl', 'cmpl') 	values 	enum('none','unk','incmpl','cmpl')
 # exonFrames 	0,	longblob 	  	Exon frame {0,1,2}, or -1 if no frame for exon
 
+lens <- structure(c(22963456L, 21142841L, 24536634L, 27894163L, 22418422L), .Dim = 5L, .Dimnames = list(c("chr2L", "chr2R", "chr3L", "chr3R", "chrX")))
+
 genes <- read.table("refGene.txt.gz", header=FALSE, stringsAsFactors=FALSE)
 names(genes) <- c("bin", "name", "chrom", "strand", "txStart", "txEnd", "cdsStart", 
             "cdsEnd", "exonCount", "exonStarts", "exonEnds", "score", "name2", 
             "cdsStartStat", "cdsEndStat", "exonFrames")
 genes <- subset(genes, chrom %in% c("chr2L","chr2R","chr3L","chr3R","chrX"))
 genes$chrom <- factor(genes$chrom)
+
+window.length <- 1e5
+gene.breaks <- seq(1,max(genes$txEnd)+window.length,by=window.length)
+gene.mids <- gene.breaks[-1]-diff(gene.breaks)/2
 
 # recomb rates
 recomb <- read.table("recomb_rates.tsv",header=TRUE, sep='\t', stringsAsFactors=FALSE)
@@ -35,22 +41,132 @@ recomb$start <- as.integer( sapply( strsplit( recomb$Genomic.locus, "[:.]" ), "[
 recomb$end <- as.integer( sapply( strsplit( recomb$Genomic.locus, "[:.]" ), "[", 4 ) )
 recomb$mid <- (recomb$start+recomb$end)/2
 
-breaks <- seq(1,max(genes$txEnd)+1e5,by=5e5)
-mids <- breaks[-1]-diff(breaks)/2
+# MDS scores: windows of 1,000 SNPs
+mds.1e3 <- do.call( rbind, lapply( levels(genes$chrom), function (this.chrom) {
+        read.table(file.path("..","results",sprintf("mds_%s_noinversion_window_1000snp.tsv",this.chrom)),sep='\t',header=TRUE)
+    } ) )
+mds.1e3$chrom <- gsub("Chr","chr",mds.1e3$chrom)
[…]

### 2016-08-13 writeup/local_pca_paper.tex
added recomb rates to drosophila

@@
 %\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}				% Use pdf, png, jpg, or epsÂ§ with pdflatex; use eps in DVI mode
+\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
 								% TeX will automatically convert eps --> pdf in pdflatex		
@@
     \begin{center}
-        \includegraphics{MDS_and_gene_count_allchr_drosophila_win104_try}
-    \end{center}
-    \caption{
-        Gene number and the effects of population structure for \textit{Drosophila melanogaster}:
-        The first MDS coordinate along the genome (black, above) is as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions};
-        below is the number of genes' transcription start and end sites within each window, divided by two.
+        \includegraphics{drosophila_recomb_mds}
+    \end{center}
+    \caption{
+        Recombination rate, gene number, and the effects of population structure for \textit{Drosophila melanogaster}:
+        The first MDS coordinate along the genome (black, above) is as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions},
+        but inverted for chromosome 3 (as each chromosome arm was run separately, relative orientation has no meaning).
+        The second plot (in blue) shows
+        local average recombination rates in cM/Mbp, obtained as midpoint estimates for 100Kbp windows 
+        downloaded from the Drosophila recombination rate calculator \citep{fistonlavier2010drosophila} release 5,
+        derived from the estimates of \citet{comeron2012many}.
+        The third plot (in black) shows the number of genes' transcription start and end sites within each window, divided by two.
         Transcription start and end sites were obtained from the RefGene table from the UCSC browser, assembly dm3, on August 11, 2016.
@@
     \begin{center}
-        \includegraphics{MDS_against_gene_count_allchr_drosophila_win104}
-    \end{center}
-    \caption{
-        Gene number and the effects of population structure for \textit{Drosophila melanogaster}:
-        this shows the first MDS coordinate and gene number, as in Supplementary Figure \ref{sfig:drosophila_gene_density_plots},
+        \includegraphics{drosophila_recomb_mds_correlation}
+    \end{center}
+    \caption{
+        Recombination rate, and the effects of population structure for \textit{Drosophila melanogaster}:
+        this shows the first MDS coordinate and recombination rate (in cM/Mbp), as in Supplementary Figure \ref{sfig:drosophila_gene_density_plots},
         against each other.
+        Since the windows underlying estimates of Supplementary Figure \ref{sfig:drosophila_gene_density_plots} do not coincide,
+        to obtain correlations we divided the genome into 100Kbp bins, 
+        and for each variable (recombination rate and MDS coordinate 1)
+        averaged the values of each overlapping bin with weight proportional to the proportion of overlap.
         The correlation coefficient and $p$-values for each linear regression are as follows:
-        \plr{FILL THIS IN}.
-% >summary of the linear regression of MDS against gene count for 2L
-% 
-% Call:
-% lm(formula = coordinate1_2L[-k] ~ gene_count1[-k])
-% 
-% Residuals:
-%      Min       1Q   Median       3Q      Max 
-% -0.40257 -0.15326  0.04554  0.15037  0.32149 
-% 
-% Coefficients:
-%                   Estimate Std. Error t value Pr(>|t|)
-% (Intercept)      0.0087783  0.0182530   0.481    0.631
-% gene_count1[-k] -0.0003592  0.0006379  -0.563    0.574
-% 
-% Residual standard error: 0.1851 on 273 degrees of freedom
-% Multiple R-squared:  0.00116,	Adjusted R-squared:  -0.002498 
-% F-statistic: 0.3172 on 1 and 273 DF,  p-value: 0.5738
-% 
-% 
-% 
-% > summary of the linear regression of MDS against gene count for 2R
-% 
-% Call:
-% lm(formula = coordinate1_2R ~ gene_count2)
-% 
-% Residuals:
-%      Min       1Q   Median       3Q      Max 
-% -0.30622 -0.08482  0.02146  0.08438  0.30339 
-% 
-% Coefficients:
-%               Estimate Std. Error t value Pr(>|t|)    
-% (Intercept)  0.0846724  0.0163485   5.179 5.08e-07 ***
-% gene_count2 -0.0028749  0.0004704  -6.112 4.52e-09 ***
-% ---
-% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-% 
-% Residual standard error: 0.1285 on 217 degrees of freedom
-% Multiple R-squared:  0.1469,	Adjusted R-squared:  0.1429 
-% F-statistic: 37.35 on 1 and 217 DF,  p-value: 4.516e-09
-% 
-% 
-% 
-% 
-% > summary of the linear regression of MDS against gene count for 3L
-% 
-% Call:
-% lm(formula = coordinate1_3L ~ gene_count3)
-% 
-% Residuals:
-%      Min       1Q   Median       3Q      Max 
-% -0.24754 -0.09927 -0.04406  0.09895  0.34383 
-% 
-% Coefficients:
-%               Estimate Std. Error t value Pr(>|t|)    
-% (Intercept) -0.0455675  0.0135826  -3.355  0.00091 ***
-% gene_count3  0.0018695  0.0004338   4.309 2.31e-05 ***
-% ---
-% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-% 
-% Residual standard error: 0.1393 on 265 degrees of freedom
-% Multiple R-squared:  0.06548,	Adjusted R-squared:  0.06196 
-% F-statistic: 18.57 on 1 and 265 DF,  p-value: 2.312e-05
-% 
-% 
-% 
-% 
-% 
-% 
-% > summary of the linear regression of MDS against gene count for 3R
-% 
-% Call:
-% lm(formula = coordinate1_3R ~ gene_count4)
-% 
-% Residuals:
-%     Min      1Q  Median      3Q     Max 
-% -0.2883 -0.1316 -0.0424  0.0892  0.4433 
-% 
-% Coefficients:
-%               Estimate Std. Error t value Pr(>|t|)    
-% (Intercept) -0.0606512  0.0158644  -3.823 0.000161 ***
-% gene_count4  0.0022703  0.0004608   4.927  1.4e-06 ***
-% ---
-% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-% 
-% Residual standard error: 0.1713 on 291 degrees of freedom
-% Multiple R-squared:  0.077,	Adjusted R-squared:  0.07383 
-% F-statistic: 24.28 on 1 and 291 DF,  p-value: 1.403e-06
-% 
-% 
-% 
-% 
-% 
-% > summary of the linear regression of MDS against gene count for X
-% 
-% Call:
-% lm(formula = coordinate1_X ~ gene_count5)
-% 
-% Residuals:
-%      Min       1Q   Median       3Q      Max 
-% -0.46790 -0.09270  0.01922  0.11102  0.35063 
-% 
-% Coefficients:
-%               Estimate Std. Error t value Pr(>|t|)    
-% (Intercept)  0.0411519  0.0139012   2.960 0.003331 ** 
-% gene_count5 -0.0020778  0.0005487  -3.787 0.000186 ***
-% ---
-% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-% 
-% Residual standard error: 0.1471 on 286 degrees of freedom
-% Multiple R-squared:  0.04774,	Adjusted R-squared:  0.04441 
-% F-statistic: 14.34 on 1 and 286 DF,  p-value: 0.0001862
+        2L: correlation $=0.52$, $r^2=0.27$;
+        2R: correlation $=0.43$, $r^2=0.18$;
+        3L: correlation $=0.47$, $r^2=0.21$;
+        3R: correlation $=0.46$, $r^2=0.21$;
+        X:  correlation $=0.50$, $r^2=0.24$.
         \label{sfig:drosophila_gene_density_correlations}

### 2016-08-13 dpgp/gene_density/plot_gene_density.R

@@ -80,7 +80,7 @@ for (this.chrom in levels(genes$chrom)) {
     with( subset(recomb,chrom==this.chrom), {
             plot( mid/1e6, (ylims[2]/15)*Comeron.Midpoint.rate, col='blue', pch=20,
                  xlim=xlims,cex=0.5,
-               ylab=if(this.chrom=="chr2L"){'recomb rate (cM/Mbp)'}else{""},
+               ylab=if(this.chrom=="chr2L"){'recombination rate'}else{""},
                yaxt=if(this.chrom=="chr2L"){'s'}else{"n"},
                xaxt='n', xlab='' );
         })
@@ -145,12 +145,18 @@ tapply( 1:nrow(stats), stats$chrom, function (kk) {
 # (Intercept) 3.194880  0.1540236 20.742800 2.619987e-53
 # MDS1        8.012976  0.9607161  8.340629 8.701074e-15
 
-tapply( 1:nrow(stats), stats$chrom, function (kk) {
-           ( summary( lm( recomb ~ MDS1, data=stats[kk,] ) ) )$r.squared
-        } )
+do.call( rbind, 
+    tapply( 1:nrow(stats), stats$chrom, function (kk) {
+               c( cor( stats$recomb[kk], stats$MDS1[kk], use='pairwise' ), 
+               ( summary( lm( recomb ~ MDS1, data=stats[kk,] ) ) )$r.squared )
+            } )
+    )
 
-# chr2L     chr2R     chr3L     chr3R      chrX 
-# 0.2682810 0.1815880 0.2081380 0.2081019 0.2436077 
+# chr2L 0.5179585 0.2682810
+# chr2R 0.4261314 0.1815880
+# chr3L 0.4562214 0.2081380
+# chr3R 0.4561819 0.2081019
+# chrX  0.4935663 0.2436077
 
 pdf( file="../../writeup/drosophila_recomb_mds_correlation.pdf", width=6, height=2, pointsize=10 )
 layout( t(1:5), widths=c(1.3,1,1,1,1.1) )
@@ -158,7 +164,7 @@ par( mar=c(5,4,2,0.25)+.1 )
 for (this.chrom in levels(genes$chrom)) {
     with( subset( stats, chrom==this.chrom ), {
              plot( recomb, MDS1, pch=20, main=this.chrom,
-                 xlab='recomb. rate (cM/Mbp)',
+                 xlab='recomb. rate',
[…]

### 2016-08-13 writeup/local_pca_paper.tex
finished editing

@@
 We applied the method to genomic datasets with good geographic sampling:
-African populations of the fruit fly \textit{Drosophila melanogaster},
+380 African \textit{Drosophila melanogaster} from the Drosophila Genome Nexus \citep{lack2015drosophila},
 a worldwide dataset of humans,
-and a range-wide dataset of the partially selfing weedy annual plant \textit{Medicago truncatula}.
+3.965 humans from several locations worldwide from the POPRES dataset \citep{nelson2008population},
+and 263 \textit{Medicago truncatula} from 24 countries around the Mediterranean basin 
+a range-wide dataset of the partially selfing weedy annual plant 
+from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
+
 In all three species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
@@
 since noise due to randomness in choice of local genealogical trees
-are not expected to show long distance correlations. 
+is not expected to show long distance correlations. 
 Below, we discuss the results and likely underlying causes.
@@
 
-We ran the method on chromosome arms 2L, 2R, 3L, 3R and X separately. 
-For each, the two-dimensional MDS visualization resembles a triangle,
-sometimes with additional points,
-as seen in Figure \ref{fig:mds_allchr}.
+We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
+The first column of Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
+of the matrix of dissimilarities between genomic windows:
+in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
+For each chromosome arm, the MDS visualization roughly resembles a triangle,
+sometimes with additional points.
 Since the relative position of each window in this plot shows the similarity between windows, 
 this suggests that there are at least three extreme manifestations of population structure 
-typified by windows found in the three corners of the triangle,
+typified by windows found in the ``corners'' of the figure,
 and that other windows' patterns of relatedness may be a mixture of those extremes. 
-\plr{make clear corners are arbitrary}
-
-\plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years, probably why it doesn't show up. 
-Also cite \citet{corbettdetig2012sequencebased} for a computational method for detecting inversions.}
-% corbettdetig2012sequencebased suggests these inversions are recent-ish (but before out-of-Africa)
+The next two columns of Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
+plotted against the window's position along the genome,
+to show how the plot of the first column is laid out along the genome.
 
@@
 
-To investigate each extreme on a particular chromosome arm, 
-we selected an ``extreme'' window in the MDS plot
+To help visualize how clustered windows with similar patterns of relatedness are along each chromosome arm, 
+we selected three ``extreme'' windows in the MDS plot
 and the 5\% of windows that are closest to it in the MDS coordinates,
@@
 Representative plots are shown for three groups of windows on each chromosome arm
-in Figure \ref{fig:mds_allchr} (groups are shown in color)
-in Figure \ref{fig:pca_by_pop} (PCA plots).
+in Figure \ref{fig:mds_allchr} (groups are shown in color),
+and in Figure \ref{fig:pca_by_pop} (PCA plots).
+The latter plots are quite different,
+showing that genomic windows in different regions of the MDS plot indeed show quite different patterns of relatedness:
 
@@
 
-The variation in patterns of relatedness turns out to be almost entirely explained by
+The most striking variation in patterns of relatedness turns out to be explained by
 several large inversions that are polymorphic in these samples, 
-disussed in \citet{corbett2012population} and \citet{langley2012genomic}.
+discussed in \citet{corbett2012population} and \citet{langley2012genomic}.
 % Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
@@
 albeit complicated by the coexistence of more than one polymorphic inversion;
-however, each breakpoint visibly affects patterns in the MDS coordinates.
+however, each breakpoint visibly affects patterns in the MDS coordinates
+(see vertical lines in Figure \ref{fig:mds_allchr}).
+% \plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years.
+% Also cite \citet{corbettdetig2012sequencebased} for a computational method for detecting inversions.}
+% corbettdetig2012sequencebased suggests these inversions are recent-ish (but before out-of-Africa)
 % except that the situation of Chr3R is a little more complicated,
@@
     \caption{
-         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
+         As in Figure 2, except that samples are colored by orientation of the polymorphic inversions 
          In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
          \citep[data from][]{lack2015drosophila}
+         In each ``INV'' denotes an inverted genotype, ``ST'' denotes the standard orientation,
+         and ``N'' denotes unknown.
         \label{fig:pca_by_inversion}
@@
 
-To see how patterns of relatedness vary without the effect of polymorphic inversions,
-we then performed the same analyses after removing, for each chromosome arm,
-the samples having the less frequent orientation of any inversion on that chromosome arm.
-The result showed substantially less variation in MDS axes,
-although some variation associated with inversion breakpoints was still visible,
-as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions}.
+To see how patterns of relatedness vary in the absence of polymorphic inversions,
+we performed the same analyses after removing, for each chromosome arm,
+any samples carrying inversions.
+In the result, shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions},
+the striking peaks associated with inversion breakpoints are gone,
+and previously smaller-scale variation now dominates the MDS visualization.
 For instance, the majority of the variation along 3L in Figure \ref{fig:mds_allchr}
@@
 remaining structure is dominated by the dip near the centromere.
-The remaining variation varies systematically along the chromosome,
-following roughly similar patterns to that seen in recombination rate and diversity in \citet{langley2012genomic} and \citet{mackay2012drosophila}.
-Indeed, correlations between the recombination rate \plr{XXX} near each window and the position on the first MDS coordinate are highly significant
-(Figure \ref{fig:drosophila_recomb_rate}).
-This is consistent with the hypothesis that variation in the effects of population structure
-is due to selection, since the strength of linked selection increases with local gene density, measured in units of recombination distance.
-However, the number of genes -- measured as the number of transcription start and end sites within each window --
-was more weakly correlated with MDS coordinate (Supplemental Figures \ref{sfig:drosophila_gene_density_plots} and \ref{sfig:drosophila_gene_density_correlations}).
-
-\begin{figure}
-    \begin{center}
-        % \includegraphics{}
-    \end{center}
-    \caption{
-        The effects of population structure without inversions is correlated to recombination rate:
-        \plr{XXX}
+Without inversions, variation in patterns of relatedness shown in the MDS plots
+follows similar patterns to that previously seen in \textit{D.~melanogaster} recombination rate and diversity \citep{langley2012genomic,mackay2012drosophila}.
+Indeed, correlations between the recombination rate in each window and the position on the first MDS coordinate are highly significant
+(Spearman's $\rho=0.54$, $p<2 \times 10^{-16}$; Figures \ref{fig:drosophila_recomb_rate} and \ref{sfig:drosophila_gene_density_correlations}).
+This is consistent with the hypothesis that variation
+is due to selection, since the strength of linked selection increases with local gene density, 
+measured in units of recombination distance.
+The number of genes -- measured as the number of transcription start and end sites within each window --
+was not significantly correlated with MDS coordinate ($p=0.22$).
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{drosophila_recomb_mds}
+    \end{center}
+    \caption{
+        The effects of population structure without inversions is correlated to recombination rate in \textit{Drosophila melanogaster}.
+        The first MDS coordinate along the genome (black, above) is as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions},
+        but inverted for chromosome 3 (as each chromosome arm was run separately, relative orientation has no meaning).
+        The second plot (in blue) shows
+        local average recombination rates in cM/Mbp, obtained as midpoint estimates for 100Kbp windows 
+        from the Drosophila recombination rate calculator \citep{fistonlavier2010drosophila} release 5,
+        using rates from \citet{comeron2012many}.
+        The third plot (in black) shows the number of genes' transcription start and end sites within each window, divided by two.
+        Transcription start and end sites were obtained from the RefGene table from the UCSC browser. %, assembly dm3, on August 11, 2016.
+        The extremely dense cluster of histone genes on chromosome arm 2L is excluded.
         \label{fig:drosophila_recomb_rate}
@@
 
-As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes.
+As for the \textit{Drosophila} data, we applied our method separately on all 22 human autosomes.
 On each, variation in patterns of relatedness was dominated by a small number of windows
@@
 Similar plots for all chromosomes are shown
-in Supplementary Figures \ref{sfig:mds12_chr1_8_human}, \ref{sfig:mds12_chr9_16_human}, \ref{sfig:mds12_chr17_22_human}.
+in Supplementary Figures \ref{sfig:mds12_chr1_8_human}, \ref{sfig:mds12_chr9_16_human}, and \ref{sfig:mds12_chr17_22_human}.
 PCA plots of many outlying windows show a characteristic trimodal shape 
@@
 
-We also ran the method grouping all 22 autosomes together, 
+We also applied the method on all 22 autosomes together, 
 and found that, remarkably, 
@@
 the first MDS coordinate of each window in Figure \ref{fig:mds_medicago},
-and are significantly correlated, as shown in Figure \ref{fig:mds_gene_count}.
+and are significantly correlated, as shown in Supplemental Figure \ref{sfig:mds_gene_count}.
 (Values shown are the number of start and end positions of each predicted mRNA transcript,
@@
 
-\begin{figure}
-    \begin{center}
-       \includegraphics{MDS_1D_against_gene_count_all_chr_win104_with_lm_update}
-    \end{center}
-    \caption{
-        First MDS coordinate against gene density for all 8 chromosomes of \textit{M.~truncatula}.
-        The first MDS coordinate is significantly correlated with gene count ($r=0.149$, $p=2.2\times 10^{-16}$). 
-        \label{fig:mds_gene_count}
-    }
-\end{figure}
-
 %%%%%%% %%%%%%%%%
@@
 revealing distinct biological processes driving this variation in each species.
-More investigation, particularly on more species and datasets, will help to uncover which patterns are generalizable.
+More investigation, particularly on more species and datasets, will help to uncover what aspects of species history can explain these differences.
 With growing appreciation of the heterogeneous effects of selection across the genome,
-especially the importance of adaptive introgression, hybrid speciation \citep{pool2015natural,brandvain2014speciation,hufford2013genomic,fitzpatrick2010rapid,staubach2012genome},
+especially the importance of adaptive introgression and hybrid speciation \citep{pool2015natural,brandvain2014speciation,hufford2013genomic,fitzpatrick2010rapid,staubach2012genome},
 local adaptation \citep{lenormand2002limits,wang2014isolation},
@@
 without additional lines of evidence.
+Inversions show up in our method because across the inverted region,
+most gene trees share a common split that dates back to the origin of the inversion.
 However, in many applications, inversions are a nuisance.
@@
 \paragraph{The effect of selection}
-It seems clear that the variation in patterns of relatedness we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
-is likely due to selection.
+It seems that
+the variation in patterns of relatedness we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
+must be explained somehow by linked selection.
 Furthermore, 
@@
 for this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently,
-Regions of the genome harboring many targets of local adaptation may show similar patterns,
+Therefore, regions of the genome harboring many targets of local adaptation may show similar patterns,
 since migrant alleles in these regions will be selected against,
@@
 
-A third possibility is that variation in patterns of relatedness is due to recent admixture between previously separated populations,
+A related possibility is that variation in patterns of relatedness is due to recent admixture between previously separated populations,
 the effects of which were not uniform across the genome due to selection.
@@
 There has been substantial debate over the relative impacts of different forms of selection.
-These have been difficult to disentangle in part because most concrete theoretical predictions
-come in the form of predicted forms of the allele frequency spectrum which are only strictly valid in randomly mating (i.e., unstructured) populations,
+These have been difficult to disentangle in part because for the most part
+theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
 and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
-It is possible that statistics making stronger use of spatial information
-will have more luck.
-\plr{expand}
+It may be possible to design more powerful statistics that make stronger use of spatial information.
 
@@
 \paragraph{Parameter choices}
-There are several choices in the method that may affect the results.
+There are several choices in the method that may in principle affect the results.
 As with whole-genome PCA,
@@
 
-More generally, there are many possible methods to discover common structure in different parts of the genome:
-in this work we use PCA to visualize the effects of population structure,
-but other methods, such as STRUCTURE \citep{falush2003inference},
-SPA \citep{yang2012modelbased},
-SpaceMix \citep{bradburd2015spatial},
-or other matrix factorization methods \citep{engelhardt2010analysis}
-may highlight different sources of variation in the data.
-It is also possible that other methods for measuring dissimilarity between windows' covariance matrices
-or for summarizing the matrix of pairwise distances between windows
-would lead to different insights.
+% More generally, there are many possible methods to discover common structure in different parts of the genome:
+% in this work we use PCA to visualize the effects of population structure,
+% but other methods, such as STRUCTURE \citep{falush2003inference},
+% SPA \citep{yang2012modelbased},
+% SpaceMix \citep{bradburd2015spatial},
+% or other matrix factorization methods \citep{engelhardt2010analysis}
+% may highlight different sources of variation in the data.
+% It is also possible that other methods for measuring dissimilarity between windows' covariance matrices
+% or for summarizing the matrix of pairwise distances between windows
+% would lead to different insights.
 
@@
 The choice of window length entails a tradeoff between signal and noise,
-since shorter windows allow better resolution along the genome but allow less precise estimates of kinship.
-A method for choosing a window length is given in Appendix \ref{apx:window_length}.
+since shorter windows allow better resolution along the genome but provide less precise estimates of relatedness.
+A method for choosing a window length to balance these considerations is given in Appendix \ref{apx:window_length}.
 % Specifically, we did PCA as follows:
 Precisely,
-denote by $Z$ the $L\times N$ recoded genotype matrix ($L$ is the number of SNPs and $N$ is the sample size), 
+denote by $Z$ the $L\times N$ recoded genotype matrix for a given window ($L$ is the number of SNPs and $N$ is the sample size), 
 and by $\overline{Z_{s}}$ the mean of non-missing entries for allele $s$, 
@@
 rather than more significant variation between groups.)
-Therefore, to check the effect of sampling on our results,
+Therefore, to control the effect of sampling on our results,
 we used a variant of PCA that gives roughly equal weight to each group of samples,
@@
 is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
-multiplied by $\sqrt{ \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} }$,
+multiplied by $( \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} )^{1/2}$,
 where $\lambda_{\ell i}$ is the $\ell^\text{th}$ eigenvalue of the genetic covariance matrix.
@@
 
-Although we were motivated by comparing PC plots up to rotation and reflection,
-the method suggests instead directly comparing entire local covariance matrices. 
+The goal of comparing PC plots up to rotation and reflection 
+turned out to be equivalent to comparing rank-$k$ approximations to local covariance matrices.
+This suggests instead directly comparing entire local covariance matrices. 
 However, with thousands of samples and tens of thousands of windows,
 computing the distance matrix would take months of CPU time,
-while $D$ can be computed in minutes using the following method.
-Since 
+while as defined above, $D$ can be computed in minutes using the following method.
+Since for square matrices $A$ and $B$,
 $\sum_{ij} (A_{ij}-B_{ij})^2 = \sum_{ij} (A^2_{ij} + B^2_{ij}) - 2 \tr(A^T B)$,
-and $M(i) = V(i) V(i)^T$,
 then due to the orthogonality of eigenvectors and the cyclic invariance of trace,
@@
 and map their positions along the genome.
-A visualization of the effects of population structure across the entire colletion is formed by extracting the corresponding genomic regions
+A visualization of the effects of population structure across the entire collection is formed by extracting the corresponding genomic regions
 and performing PCA on all, aggregated, regions.
-
-\subsection{Software}
-
-The methods described here
-are implemented in an open-source R package
-available at \url{https://github.com/petrelharp/lostruct},
-as well as scripts to perform all analyses from VCF files
-at various parameter settings.
 
@@
 
-We used three publically available datasets (summarized in Table \ref{tab:data_stats});
-we converted each to a numeric matrix
+The three publicly available datasets we used are summarized in Table \ref{tab:data_stats}.
+We converted each to a numeric matrix
 (with one row per polymorphic variant and one column per sample)
@@
 (or NA for missing data).
-A normalization step (see below) ensures the result does not depend on the choice of reference allele.
+% A normalization step (see below) ensures the result does not depend on the choice of reference allele.
 
@@
 which has array-derived genotype information for 447,267 SNPs across the 22 autosomes
-of 3,965 samples in total: 346 African-Americas, 73 Asians, 3,187 Europeans and 359 Indian Asians.
+of 3,965 samples in total: 346 African-Americans, 73 Asians, 3,187 Europeans and 359 Indian Asians.
 Since these data derive from genotyping arrays, the SNP density is much lower than the other datasets,
@@
 \end{table}
+
+\subsection{Data accessibility}
+
+The methods described here
+are implemented in an open-source R package
+available at \url{https://github.com/petrelharp/local_pca},
+as well as scripts to perform all analyses from VCF files
+at various parameter settings.
 
@@
     \begin{center}
-        \includegraphics{drosophila_recomb_mds}
-    \end{center}
-    \caption{
-        Recombination rate, gene number, and the effects of population structure for \textit{Drosophila melanogaster}:
-        The first MDS coordinate along the genome (black, above) is as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions},
-        but inverted for chromosome 3 (as each chromosome arm was run separately, relative orientation has no meaning).
-        The second plot (in blue) shows
-        local average recombination rates in cM/Mbp, obtained as midpoint estimates for 100Kbp windows 
-        downloaded from the Drosophila recombination rate calculator \citep{fistonlavier2010drosophila} release 5,
-        derived from the estimates of \citet{comeron2012many}.
-        The third plot (in black) shows the number of genes' transcription start and end sites within each window, divided by two.
-        Transcription start and end sites were obtained from the RefGene table from the UCSC browser, assembly dm3, on August 11, 2016.
-        The extremely dense cluster of histone genes on chromosome arm 2L is excluded.
-        \label{sfig:drosophila_gene_density_plots}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
         \includegraphics{drosophila_recomb_mds_correlation}
@@
         Recombination rate, and the effects of population structure for \textit{Drosophila melanogaster}:
-        this shows the first MDS coordinate and recombination rate (in cM/Mbp), as in Supplementary Figure \ref{sfig:drosophila_gene_density_plots},
+        this shows the first MDS coordinate and recombination rate (in cM/Mbp), as in Supplementary Figure \ref{fig:drosophila_recomb_rate},
         against each other.
-        Since the windows underlying estimates of Supplementary Figure \ref{sfig:drosophila_gene_density_plots} do not coincide,
+        Since the windows underlying estimates of Supplementary Figure \ref{fig:drosophila_recomb_rate} do not coincide,
         to obtain correlations we divided the genome into 100Kbp bins, 
@@
     \begin{center}
+       \includegraphics{MDS_1D_against_gene_count_all_chr_win104_with_lm_update}
+    \end{center}
+    \caption{
+        First MDS coordinate against gene density for all 8 chromosomes of \textit{M.~truncatula}.
+        The first MDS coordinate is significantly correlated with gene count ($r=0.149$, $p=2.2\times 10^{-16}$). 
+        \label{sfig:mds_gene_count}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
        \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr}

### 2016-08-13 dpgp/gene_density/plot_gene_density.R

@@ -158,6 +158,8 @@ do.call( rbind,
 # chr3R 0.4561819 0.2081019
 # chrX  0.4935663 0.2436077
 
+cor( stats$recomb, stats$MDS1, use='pairwise' )
+
 pdf( file="../../writeup/drosophila_recomb_mds_correlation.pdf", width=6, height=2, pointsize=10 )
 layout( t(1:5), widths=c(1.3,1,1,1,1.1) )
 par( mar=c(5,4,2,0.25)+.1 )

### 2016-08-16 writeup/submission/cover-letter.tex
letter

@@
-To the editor(s) --
+\documentclass[stdletter,letterpaper,addrfromright,orderfromdateto,dateleft,11pt,noaddrto,sigleft]{newlfm}
+\usepackage[hidelinks]{hyperref}
+
+\topmarginskip{-0.35in}
+\bottommarginskip{-1.5in}
+\leftmarginsize{1in}
+\rightmarginsize{1.25in}
+\sigskipbefore{0.4in}
+\sigskipafter{0in}
+\noLines
+\nolines
+\noHeadline
+\noheadline
+\signature{Peter Ralph and Han Li}
+
+\namefrom{}
+\addrfrom{Peter Ralph \\ Fenton Hall \\ University of Oregon \\ Eugene, OR 97403-1222 USA \\ \textsc{phone:} (541) 346-5530 \\ \textsc{fax:} (541) 346-0987 \\ \href{mailto:plr@uoregon.edu}{plr@uoregon.edu}}
+\phonefrom{(541) 346-5530}
+\faxfrom{(541) 346-0987}
+\emailfrom{plr@uoregon.edu}
+
+% Department of Mathematics
+% Fenton Hall
+% University of Oregon
+% Eugene, OR 97403-1222 USA
+% Phone: 1-541-346-4705
+% FAX 1-541-346-0987
+
+\greetto{To the editor(s) --}
+\closeline{Sincerely,}
+
+\begin{document}
+
+\begin{newlfm}
+
+% A cover letter MUST include:
+% (1) name, address, telephone number, fax number, and e-mail address of author responsible for correspondence regarding the manuscript;
+% (2) paragraph highlighting the main points of the work;
+% (3) statement that the manuscript has been seen and approved by all listed authors;
+% (4) a list of potential referees (referees' names, institution, and e-mail), and, if desired, a reasonable list of individuals with potential conflict, as we work very hard to avoid sending manuscripts to competitors;
+% (5) status of any statements of personal communication or other permissions needed (any data presented as unpublished results from individuals other than the authors require permission for use);
+% (6) statement regarding databank submission of data; and
+% (7) assurance that all gene/protein names and symbols used in the paper adhere to approved nomenclature guidelines for specific species.
 
 We are writing to submit our paper,
-``Local PCA shows how population structure differs along the genome'',
-for consideration in JOURNAL XXX.
-The paper brings together several topics that are currently of substantial interest 
+``Local PCA Shows How the Effect of Population Structure Differs Along the Genome'',
+for consideration in \textit{JOURNAL XXX}.
+In this paper, we set out to describe how background patterns of relatedness 
+(colloquially, ``population structure'')
+vary along the genome at a scale larger than correlations induced by linkage.
+It was not certain that we would find anything of interest:
+in the field we tend to think of neutral, background population structure as a single thing,
+against which outlying signals of selection can be identified.
+What we found was not only substantial variation at chromosomal scales,
+but different underlying causes in the three species we consider.
+The focus of our paper is on the method we developed to identify and visualize such variation in patterns of relatedness.
+We believe our method represents a fundamental advance because it lets the data speak for itself,
+providing an unbiased view of what are the most important variation in patterns of relatedness,
+rather than looking for specific signals.
+The method is therefore complementary to more common methods that identify small regions of the genome
+that are outliers for particular summaries of relatedness (e.g., $F_{ST}$-based scans for selection).
+We think the paper will be of interest to readers of \textit{JOURNAL XXX}
+because the method fills an important gap in the genomicist's toolkit,
+that is tailored to describing variation in modern, spatially sampled datasets,
+and because it sheds light on the roles of linked selection and chromosomal inversions in 
+establishing modern genetic diversity.
+% Our results are consistent with what is known about linked selection and chromosomal inversions
+% (a possible suprise being no evidence for segregating inversions in \textit{Medicago truncatula});
+% the strength of the biological results lie in the unified framework that allows 
+% flexible discovery and comparison between the three species.
+
+Potential reviewers for this paper could include:
+John Novembre (U Chicago, \href{mailto:jnovembre@uchicago.edu}{jnovembre@uchicago.edu}),
+Joe Pickrell (NY Genome Center, \href{mailto:jkpickrell@nygenome.org}{jkpickrell@nygenome.org}),
+David Lowry (MSU, \href{mailto:dlowry@msu.edu}{dlowry@msu.edu}),
+or Andy Clark (Cornell, \href{mailto:ac347@cornell.edu}{ac347@cornell.edu}).
+
+The paper has been seen and approved by both authors.  
+All data we use is publically available, and so no permissions or databank submissions are needed;
+however, all source code and substantial derived statistics are available at 
+\href{https://github.com/petrelharp/local_pca}{our github page}.
+We adhere to all relevant nomenclature guidelines.
+
+\end{newlfm}
+\end{document}
+
+Our paper brings together several topics that are currently of substantial interest 
 to a wide audience: 
-First, principal components analysis is one of the most widely used descriptive methods 
-applied to genomic data;
-however, the results depend on the portion of the genome examined, 
-and we provide a systematic method to examine this variation.
-Second, the three species we study show substantially different patterns
-reflective of substantially different biology,
-most notably, the role and scale of chromosomal inversions.
-Finally,
-although we don't develop these ideas,
-our method points the way towards move powerful methods for genome-wide association studies.
-Furthermore,
-the methods are provided as an R package,
-to allow others to easily use the method.
-
-In summary, the paper aims to both
-provide conceptual unification to a range of observations
-(variation along the genome, linked selection, population structure),
-and provide a new statistical method,
-which leads to biologically interesting observations.
-
-Sincerely,
-
- Peter Ralph and Han Li
+Methods that search for functional variants in genomic data fundamentally search for regions of the genome
+showing unusual patterns of relatedness:
+divergence between populations in the case of scans for selection,
+or association with a phenotype in the case of GWAS.
+In this paper,
+we provide a method to visualize larger-scale variations in patterns of relatedness,
+show that this variation can be substantial, and is likely the result of both selection and structural variation.
+The method represents a fundamental advance because it lets the data speak for itself,
+providing an unbiased view of what are the most important variation in patterns of relatedness,
+rather than looking for specific signals.
+The method is also more suitable to modern, spatially sampled datasets that may not be easily separable into a small number of populations.
+On the other hand, the method will be familiar to readers as it combines two often-used approaches: 
+visualization with PCA and plots along the genome.
+The biological conclusions are also of substantial interest,
+as we find that the three species we study differ fundamentally in terms of the influence of linked selection
+and the scale and importance of chromosomal inversions,
+in particular finding strong evidence to suggest that \textit{Medicago truncatula} harbors no inversions segregating at any appreciable frequency.

### 2016-08-17 writeup/local_pca_paper.tex
han's edits

@@
 is an average of the relationships across locus-specific genealogical trees,
-which can be strongly affected on intermediate genomic scales by many biological factors,
-such as linked selection.
-We show how to use local principal components analysis to describe this meso-scale variation,
+which can be strongly affected on intermediate genomic scales by linked selection and other factors,
+We show how to use local principal components analysis to describe this 
+meso-scale heterogeneity in patterns of relatedness,
 and apply the method to genomic data from three species,
-finding in each that the effect of population structure varies on the scale of megabases to tens of megabases.
-In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
+finding in each that the effect of population structure 
+can vary substantially across only a few megabases.
+In a global human dataset, localized heterogeneity is likely explained by polymorphic chromosomal inversions.
 In a range-wide dataset of \textit{Medicago truncatula},
-common axes of variation are shared between chromosomes,
+factors that produce heterogeneity are shared between chromosomes,
 correlate with local gene density,
 and may be caused by background selection or local adaptation.
-In a dataset of primarily African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
+In a dataset of primarily African \textit{Drosophila melanogaster}, large-scale heterogeneity across each chromosome arm
 is explained by known chromosomal inversions thought to be under recent selection,
-and after removing samples carrying inversions, remaining variation is correlated with recombination rate and gene density,
+and after removing samples carrying inversions, remaining heterogeneity is correlated with recombination rate and gene density,
 again suggesting a role for linked selection.
@@
 As reviewed above,
-% In searching for loci of interest, 
-it is common to describe variation in genetic differentiation between pairs of populations along the genome (often using $F_{ST}$),
+it is common to describe variation along the genome of simple statistics such as $F_{ST}$,
 and also to use methods such as PCA to visualize large-scale patterns in genome-wide relatedness.
-In this paper we aim to combine these two, by describing in an unbiased, descriptive way
-how patterns of mean relatedness vary systematically along the genome, with a focus on large, geographically distributed populations.
+In this paper we 
+apply PCA locally along the genome,
+thus describing in an unbiased, descriptive way
+how patterns of mean relatedness vary systematically along the genome, 
+in a way particularly suited to large samples from geographically distributed populations.
 Geographic population structure sets the stage by establishing similar patterns of relatedness across much of the genome;
@@
 we can compare relative contributions of different sorts of variation across taxa;
-since the method is an visualization method, we allow ourselves to be suprised by unexpected signals in the data.
+since the method is an visualization method, 
+we allow ourselves to be suprised by unexpected signals in the data.
 
@@
 a worldwide dataset of humans,
-3.965 humans from several locations worldwide from the POPRES dataset \citep{nelson2008population},
+3,965 humans from several locations worldwide from the POPRES dataset \citep{nelson2008population},
 and 263 \textit{Medicago truncatula} from 24 countries around the Mediterranean basin 
@@
     \caption{
-        Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms.
+        Variation in patterns of relatedness for windows across \textit{Drosophila melanogaster} chromosome arms.
         In all plots, each point represents one window along the genome.
          The first column shows the MDS visualization of relationships between windows, 
-         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
+         and the second and third columns show 
+        the two MDS coordinates
+         against the midpoint of each window;
          rows correspond to chromosome arms.
@@
 we performed the same analyses after removing, for each chromosome arm,
-any samples carrying inversions.
+any samples carrying inversions on that arm.
 In the result, shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions},
@@
         The effects of population structure without inversions is correlated to recombination rate in \textit{Drosophila melanogaster}.
-        The first MDS coordinate along the genome (black, above) is as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions},
-        but inverted for chromosome 3 (as each chromosome arm was run separately, relative orientation has no meaning).
+        The first plot (in red) shows the first MDS coordinate along the genome 
+        for windows of 10,000 SNPs,
+        obtained after removing samples with inversions.
+        (A plot analogous to Figure \ref{fig:mds_allchr} is shown
+        in Supplemental Figure \ref{sfig:mds_allchr_noinversions}.)
         The second plot (in blue) shows
@@
         using rates from \citet{comeron2012many}.
-        The third plot (in black) shows the number of genes' transcription start and end sites within each window, divided by two.
+        The third plot (in black) shows the number of genes' transcription start and end sites within each 100Kbp window, divided by two.
         Transcription start and end sites were obtained from the RefGene table from the UCSC browser. %, assembly dm3, on August 11, 2016.
-        The extremely dense cluster of histone genes on chromosome arm 2L is excluded.
+        The histone gene cluster on chromosome arm 2L is excluded.
         \label{fig:drosophila_recomb_rate}
@@
 
-As for the \textit{Drosophila} data, we applied our method separately on all 22 human autosomes.
+As we did for the \textit{Drosophila} data, we applied our method separately to all 22 human autosomes.
 On each, variation in patterns of relatedness was dominated by a small number of windows
@@
 For this reason, the most likely candidate may be selection against linked deleterious mutations, known as ``background selection'' \citep{charlesworth1993effect,charlesworth2013background}.
-Background selection can be thought of as reducing the number of potential contributors to the gene pool in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious};
+Informally, background selection reduces the number of potential contributors to the gene pool 
+in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious};
 for this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently,
@@
 % recoded matrix into contiguous matrices that have the same columns but fewer rows than the original matrix, 
-and applied Principal Component Analysis (PCA) as described in \citet{mcvean2009genealogical}
+and applied principal component analysis (PCA) as described in \citet{mcvean2009genealogical}
 separately to the submatrices that corresponded to each window.
@@
 
-Since eigenvectors are still only defined up to sign,
+Since the definition of eigenvectors does not specify their sign,
 when comparing between windows we choose signs to best match each other:
@@
 rather than more significant variation between groups.)
-Therefore, to control the effect of sampling on our results,
-we used a variant of PCA that gives roughly equal weight to each group of samples,
+Therefore, to check that sampling patterns do not affect our results,
+we compared to a variant of PCA that gives roughly equal weight to each group of samples,
 rather than to each sample.
@@
     \caption{
-        Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms: without inversions.
+        Variation in structure for windows of 1,000 SNPs 
+        across \textit{Drosophila melanogaster} chromosome arms: without inversions.
          As in Figure \ref{fig:mds_allchr}, but after omitting for each chromosome arm individuals carrying the less frequent orientation
          of any inversions on that chromosome arm.
+         The values differ from those in \ref{fig:drosophila_recomb_rate}
+         in the window size used and that some MDS values were inverted
+         (but relative orientation is meaningless as chromosome arms were run separately,
+         unlike for \textit{Medicago}).
         In all plots, each point represents one window along the genome.

### 2016-09-07 writeup/local_pca_paper.tex
poster

@@
 in Figure \ref{fig:mds_allchr} (groups are shown in color),
-and in Figure \ref{fig:pca_by_pop} (PCA plots).
+and in Supplemental Figure \ref{fig:pca_by_pop} (PCA plots).
 The latter plots are quite different,
-showing that genomic windows in different regions of the MDS plot indeed show quite different patterns of relatedness:
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
-    \end{center}
-    \caption{      
-        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_allchr},
-        on each chromosome arm of \textit{Drosophila melanogaster}.
-        In all plots, each point represents a sample. 
-        The first column shows the combined PCA plot for windows 
-        whose points are colored green in Figure \ref{fig:mds_allchr}; 
-        the second is for orange windows; 
-        and third is for purple windows.
-         \label{fig:pca_by_pop}
-    }
-\end{figure}
+showing that genomic windows in different regions of the MDS plot indeed show quite different patterns of relatedness.
+
+% Fig fig:pca_by_pop WAS HERe
 
@@
 To verify that this is a possible explanation,
-we computed gene densities in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
+we counted the number of genes found in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
 which are shown juxtaposed with
@@
 In particular, the recombination rates estimated by \citet{paape2012finescale} appear visually to be similar,
-but were not readily available for comparison.
+but were not available in a form that can be compared to our MDS values.
 
@@
        \includegraphics{Fig6_Together_MDS_plot_chr3_final}
-    \end{center}
-    \caption{
-       MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3. 
+       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
+    \end{center}
+    \caption{
+       MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3,
+       with corresponding PCA plots.
        Each point in the plot represents a window;
@@
        Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:mds_medicago_allchr}.
-       \label{fig:mds12_medicago}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
-    \end{center}
-    \caption{
+       \textbf{(below)}
         PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
@@
         Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:pca_peaks_medicago_allchr}.
+       \label{fig:mds12_medicago}
        \label{fig:pca_medicago}
@@
 and could potentially be misleading.
-
-%%%%%%%%
-\subsection*{Acknowledgements}
-
-We are indebted to John Pool, Russ Corbett-Detig, Matilde Cordeiro, and Peter Chang 
-for assistance with obtaining data and interpreting results
-(especially inversion status of \textit{D.~melanogaster} samples).
-Thanks also go to Yaniv Brandvain, Barbara Engelhardt, Charles Langley, Graham Coop, and Jeremy Berg for helpful comments
-and for encouraging the project.
 
@@
 
-\subsection{Data accessibility}
+\section{Data access}
 
@@
 at various parameter settings.
+
+Datasets are available as follows:
+human (POPRES) at dbGaP with accession number phs000145.v4.p2,
+\textit{Medicago} at the Medicago Hapmap \url{http://www.medicagohapmap.org/},
+and \textit{Drosophila} at the Drosophila Genome Nexus, \url{http://www.johnpool.net/genomes.html}.
+
+%%%%%%%%
+\section*{Acknowledgements}
+
+We are indebted to John Pool, Russ Corbett-Detig, Matilde Cordeiro, and Peter Chang 
+for assistance with obtaining data and interpreting results
+(especially inversion status of \textit{D.~melanogaster} samples).
+Thanks also go to Yaniv Brandvain, Barbara Engelhardt, Charles Langley, Graham Coop, and Jeremy Berg for helpful comments
+and for encouraging the project.
+
+%%%%%%%%
+\section*{Disclosure declaration}
+
+The authors declare no conflicts of interest.
 
@@
     \begin{center}
+       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
+    \end{center}
+    \caption{      
+        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_allchr},
+        on each chromosome arm of \textit{Drosophila melanogaster}.
+        In all plots, each point represents a sample. 
+        The first column shows the combined PCA plot for windows 
+        whose points are colored green in Figure \ref{fig:mds_allchr}; 
+        the second is for orange windows; 
+        and third is for purple windows.
+         \label{fig:pca_by_pop}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
        \includegraphics[width=1\textwidth]{FigS_pca_plots_allchr_drosophila}

### 2016-09-07 writeup/poster/oxford-2016-local-pca-poster.tex
poster

\documentclass[fontscale=0.38,a0paper]{baposter}

\tracingstats=2

\usepackage{times}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{relsize}
\usepackage{multirow}
\usepackage{bm}
\usepackage{caption}
\usepackage[authoryear,round]{natbib}
\usepackage[hidelinks]{hyperref}

\usepackage{tabularx}
\usepackage{booktabs}

\usepackage{colortbl}

\usepackage{multicol}

%\usepackage{pgfbaselayers}
%\pgfdeclarelayer{background}
%\pgfdeclarelayer{foreground}
%\pgfsetlayers{background,main,foreground}

% plr macros: these are for boxes in tables
\newcommand\tLL[1]{\multicolumn{1}{|c}{#1}}
\newcommand\tRR[1]{\multicolumn{1}{c|}{#1}}
\newcommand\tLR[1]{\multicolumn{1}{|c|}{#1}}
% nucleotides:
\newcommand{\nA}{\mbox{A}}  
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}

%\usepackage{tikz}
%\usetikzlibrary{arrows,backgrounds,patterns,matrix,shapes,fit,calc,shadows,plotmarks,decorations.pathmorphing,positioning,trees}

\renewcommand{\familydefault}{\sfdefault}
\usepackage{helvet}
\usepackage{bookman}
\usepackage{palatino}

% My stuff
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
% end my stuff

\selectcolormodel{cmyk}

% \graphicspath{{images/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Some symbols used in the text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\newcommand{\sub}[2]{\ensuremath{#1_{\textrm{#2}}}}
%\newcommand{\gen}[2]{\ensuremath{\textrm{#1}_{\textrm{#2}}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Multicol Settings
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\setlength{\columnsep}{0.7em}
\setlength{\columnsep}{8mm}
\setlength{\columnseprule}{0mm} % no vertical line

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Save space in lists. Use this after the opening of the list
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\compresslist}{
\setlength{\itemsep}{1pt}
\setlength{\parskip}{0pt}
\setlength{\parsep}{0pt}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Other Macros
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \renewcommand{\labelenumi}{\alph{enumi})}
% \renewcommand{\theenumi}{\alph{enumi})}
\newcommand{\postercaption}[1]{\begin{minipage}{\linewidth}\center\smaller
  {#1}\end{minipage}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Begin of Document
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Here starts the poster
%%%---------------------------------------------------------------------------
%%% Format it to your taste with the options
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Define some colors
\definecolor{silver}{cmyk}{0,0,0,0.3}
\definecolor{black}{cmyk}{0,0,0.0,1.0}
\definecolor{white}{cmyk}{0,0,0.0,0.0}
\definecolor{darkSilver}{cmyk}{0,0,0,0.1}

\definecolor{darkYellow}{cmyk}{0,0,1.0,0.5}
\definecolor{yellow}{cmyk}{0,0,0.9,0.0}
\definecolor{reddishyellow}{cmyk}{0,0.22,1.0,0.0}
\definecolor{lightyellow}{cmyk}{0,0,0.3,0.0}
\definecolor{lighteryellow}{cmyk}{0,0,0.1,0.0}
\definecolor{lightestyellow}{cmyk}{0,0,0.05,0.0}

\definecolor{DavisBlue}{cmyk}{1,0.56,0,0.34}
%    C = 100  M = 56  Y = 0 K = 34

\definecolor{HohenheimBlue}{cmyk}{1,0.5,0,0.45}
\definecolor{HohenheimLightLightBlue}{RGB}{243,250,255}
\definecolor{HohenheimLightBlue}{RGB}{215,221,235}
\definecolor{HohenheimDarkBlue}{RGB}{52,104,152}
\definecolor{darkCyan}{cmyk}{1,1,0,0.5}
\definecolor{cyan}{cmyk}{1,0,0,0.0}
\definecolor{lightcyan}{cmyk}{0.3,0,0,0.0}
\definecolor{lightercyan}{cmyk}{0.1,0,0,0.0}
\definecolor{lightestcyan}{cmyk}{0.05,0,0,0.0}

\typeout{Poster Starts}
%Define custom background
\background{
  \begin{tikzpicture}[remember picture,overlay]%
    \draw (current page.north west)+(-2em,2em) node[anchor=north west] {\includegraphics[height=\textheight]{silhouettes_background}};
  \end{tikzpicture}%
}

\newlength{\leftimgwidth}
\begin{poster}%
  % Poster Options
  {
  % Columns and Column spacing
  columns=2,
  colspacing=1.5em,
  % Background
  % background=user,
  background=shade-tb,
  % Poster header options
  eyecatcher=yes,
  posterheaderheight=0.15\textheight,
  % Format of textbox header
  headerborder=open,
  % headershape=small-rounded,
  headershape=roundedright,
  headershade=shade-tb,
  % headershade=plain,
  headerfont=\Large\textsf, %Sans Serif
  % Format of textbox
  boxborder=small-rounded,
  % boxborder=roundedleft,    
  boxshade=plain,
  linewidth=0.5pt,
  % Color style
  % bgColorOne=lightestcyan,
  % bgColorTwo=lightcyan,
  bgColorOne=white,
  bgColorTwo=white,
  % borderColor=cyan,
  borderColor=black,
  headerColorOne=DavisBlue,
  % headerColorOne=cyan,
  headerColorTwo=black,
  % headerFontColor=black,
  headerFontColor=white,
  % boxColorOne=lightercyan,
  boxColorOne=HohenheimLightLightBlue,
  boxColorTwo=white,
  % Show grid to help with alignment
  grid=no
  % grid=yes
  }
  % Eye Catcher
  {
  \makebox[15em][r]{%
    \begin{minipage}{15em}
       \hfill
       \includegraphics[width=15em]{USC-seal} \\
    \end{minipage}
    }
  }
  % Title
  {\sf %Sans Serif
  %\bf% %bold  
  \vspace{0.5em}
     \textbf{\textcolor{DavisBlue}{Local PCA finds heterogeneity in population structure along the genome}}\vspace{0.5em}}
  % Authors
  {\sf %Sans Serif
    Han Li$^{\dagger}$ and Peter Ralph$^{\dagger\ddagger}$ \\  \vspace{-1.0mm}
    {\small \textit{$\dagger$ Molecular \& Computational Biology, University of Southern California }}\\  
    {\small \textit{$\ddagger$ Mathematics and Biology, University of Oregon} }\\
    {\small   \url{http://biorxiv.org/content/early/2016/08/21/070615} }\\
    {\small   \url{https://github.com/petrelharp/local_pca} }\\
  }
  % Project logo
  {
    \makebox[15em][r]{%
      \begin{minipage}{15em}
        \hfill
          \includegraphics[width=15em]{UOSignature-STK-BLK} \\
      \end{minipage}
    }
  }

  % Width of left inset image
%  \setlength{\leftimgwidth}{0.78em+0.0em}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Now define the boxes that make up the poster
%%%---------------------------------------------------------------------------
%%% Each box has a name and can be placed absolutely or relatively.
%%% The only inconvenience is that you can only specify a relative position 
%%% towards an already declared box. So if you have a box attached to the 
%%% bottom, one to the top and a third one which should be in between, you 
%%% have to specify the top and bottom boxes before you specify the middle 
%%% box.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{What is local population structure?}{name=intro,column=0,row=0,span=1}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\emph{Population structure}: ``such matters as numbers, composition by age and sex, and state of subdivision'' \citep{wright1949genetical};
but also commonly, the genetic patterns that result from this process.
What aspects of demography should be included?  Should natural selection?
Incorporating differential survival or fecundity makes the concept less clear:
does a randomly mating population consisting of two types that are partially reproductively isolated from each other
have ``population structure''?

\emph{Whatever the definition},
selection causes the \emph{effects} of population structure --
\emph{realized} patterns of genetic relatedness --
to differ across the genome.
Locally adapted alleles are selected against in migrants,
increasing genetic differentiation between populations around this locus.
Or, newly adaptive alleles spread first in local populations.

We aim to describe 
how patterns of mean relatedness vary along the genome.
This works because:
geography establishes similar patterns of relatedness across much of the genome,
which are distorted locally by selection and other factors;
we look for commonalities in these distortions (\emph{not} outlier loci).

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Our method}{name=method,column=0,span=1,below=intro}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \begin{center}
    \includegraphics[width=\textwidth]{../the-method-diagram}
    \end{center}

    \begin{enumerate}
            \compresslist
        \item Code genome as $\{0,1,2\}$ and divide into windows (by \# SNPs, bp, or cM).
        \item Do PCA on each window: $\lambda^{(i)}_\ell$, $V^{(i)}_\ell$ the $\ell^\text{th}$ eigenvalue/vector of the covariance matrix for the $i^\text{th}$ window.
        \item Compute distances between windows:
            \[  
            D_{ij} = \| 
            \sum_{\ell=1}^k \frac{\lambda^{(i)}_\ell}{\sum_{m=1}^k\lambda^{(i)}_m} V^{(i)}_\ell (V^{(i)}_\ell)^T
            -
            \sum_{\ell=1}^k \frac{\lambda^{(j)}_\ell}{\sum_{m=1}^k\lambda^{(j)}_m} V^{(j)}_\ell (V^{(j)}_\ell)^T
            \|
            \]
        \item Visualize distances/similarities with MDS.
        \item Highlight similar regions of the genome.
        \item See how local PCA differs between regions by doing PCA after combining similar windows.
    \end{enumerate}

    \textbf{Notes:}
    \begin{enumerate}
            \compresslist
        \item Window size chosen by optimizing signal:noise \\
            (``signal:'' variance in PC scores across a chromosome; \\
            ``noise:'' average jackknife-estimated standard error of PC scores by window);\\
            in applications, hundreds to thousands per chromosome.
        \item Also used weighted PCA to remove the effects of oversampled groups without discarding data.
        \item Results insensitive to rough choice of window size or unit of measurement.
        \item Linear algebra reduced computation time of $D$ from weeks to seconds.
    \end{enumerate}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Results: \textit{Drosophila melanogaster}}{name=drosophila,column=1,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    380 Pan-African samples from Drosophila Genome Nexus \citep{lack2015drosophila}, \url{http://www.johnpool.net/genomes.html}.

    \centering
    \includegraphics[width=\textwidth]{drosophila-MDS-extract}
    \postercaption{\textbf{Fig 1:}MDS, along the genome, for two chromosome arms.  Vertical lines corrspond to inversions.}

    \vspace{1em}

    \includegraphics[width=\textwidth]{drosophila_just_recomb_mds}
    \postercaption{\textbf{Fig 2:}
        \emph{(top)} MDS, along the genome, \emph{after} removing inverted samples;\\
        \emph{(bottom)} recombination rate from \citep{fistonlavier2010drosophila}.
    }

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Results: \textit{Homo sapiens}}{name=humans,column=1,below=drosophila}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    The POPRES dataset \citep{nelson2008population}:
    346 African-Americans, 73 Asians, 3,187 Europeans and 359 Indian Asians.
    (SNP array data; 500K markers)

    \centering
    \includegraphics[width=\textwidth]{popres_chr8}
    \postercaption{\textbf{Fig 3:}
    MDS on chromosome 8. Vertical line is a large segregating inversion.}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Results: \textit{Medicago truncatula}}{name=medicago,column=1,below=humans}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    263 samples from 24 circum-Mediterranean countries,
    from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved}.

    \centering

    \includegraphics[width=0.8\textwidth]{../Fig6_Together_MDS_plot_chr3_final}
    \postercaption{\textbf{Fig 4:}
    MDS along chromosome 3, showing continuous variation.}
    \includegraphics[width=0.8\textwidth]{medicago_PCA_chr3}
    \postercaption{\textbf{Fig 5:}
    PCA plots corresponding to windows in the three corners of the MDS triangle.}
    \includegraphics[width=0.7\textwidth]{medicago_mds_gene_count_extract}
    \postercaption{\textbf{Fig 6:}
    MDS score and local gene density along chromosomes 1--4.}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{}{name=references,column=0,above=bottom}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \textbf{Conclusion:}
    Variation mostly caused by segregating inversions
    and linked selection;
    how much varies by species.
    \vspace{1em}

    \textbf{R package:} available at \url{http://github.com/petrelharp/local_pca}\\
  \scriptsize
  \renewcommand{\section}[2]{\vskip 0.0em}
  \bibliographystyle{abbrvnat}
  \setlength{\bibsep}{0.0pt}
  \bibliography{../references}
  }
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{poster}%
\end{document}

### 2016-09-07 writeup/submission/cover-letter.tex
poster

@@
 vary along the genome at a scale larger than correlations induced by linkage.
-It was not certain that we would find anything of interest:
+It was not certain that we would find significant large-scale heterogeneity in patterns of relatedness;
 in the field we tend to think of neutral, background population structure as a single thing,
-against which outlying signals of selection can be identified.
+uniform across the genome except at a few isolated loci under selection.
 What we found was not only substantial variation at chromosomal scales,
@@
 Joe Pickrell (NY Genome Center, \href{mailto:jkpickrell@nygenome.org}{jkpickrell@nygenome.org}),
-David Lowry (MSU, \href{mailto:dlowry@msu.edu}{dlowry@msu.edu}),
+Alkes Price (Harvard, \href{mailto:aprice@hsph.harvard.edu}{aprice@hsph.harvard.edu}),
 or Andy Clark (Cornell, \href{mailto:ac347@cornell.edu}{ac347@cornell.edu}).

### 2016-09-25 writeup/local_pca_paper.tex
stuff for submission

@@
 \usepackage{setspace}
-\usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
 \usepackage[hidelinks]{hyperref}
@@
 
-\title{Local PCA Shows How the Effect of Population Structure Differs Along the Genome}
-\author{Han Li, Peter Ralph}
-%\date{}							% Activate to display a given date or no date
+% % stuff for submission
+% \usepackage[nomarkers,figuresonly]{endfloat}
+% \usepackage{setspace}
+% \doublespacing
+% \usepackage{lineno}
+% \linenumbers
+% \bibliographystyle{plos2015}
+% \newcommand\citet{\cite}
+% \newcommand\citep{\cite}
+
+% % for the nice version
+\usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
+\bibliographystyle{plainnat}
 
@@
 \DeclareMathOperator{\tr}{tr}
+
+\title{Local PCA Shows How the Effect of Population Structure Differs Along the Genome}
+\author{Han Li, Peter Ralph}
+%\date{}							% Activate to display a given date or no date
 
@@
 many species show chromosome-scale variation in diversity and divergence
-\citep[e.g.,][]{langley2012genomic};
+(e.g., \citep{langley2012genomic});
 species phylogenies can differ along the genome 
@@
 adaptive introgression and/or local adaptation 
-\citep[e.g.,][]{pease2013accurate,ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
+(e.g., \citep{pease2013accurate,ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting});
 and theoretical expectations predict that geographic patterns of relatedness should depend on selection
@@
 Patterns in genome-wide relatedness are often summarized
-by applying principal components analysis \citep[PCA,][]{patterson2006population} 
+by applying principal components analysis (PCA, \citep{patterson2006population}) 
 to the genetic covariance matrix,
@@
 Modeling such ``background'' kinship between samples
-is essential to genome-wide association studies \citep[GWAS,][]{price2006principal,astle2009population},
+is essential to genome-wide association studies (GWAS, \citep{price2006principal,astle2009population}),
 % and PCA has often been used in this context \citep{price2006principal}.
@@
 A number of general methods for dimensionality reduction also use a strategy of ``local PCA''
-\citep[e.g.,][]{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear},
+(e.g., \citep{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear}),
 performing PCA not on the entire dataset but instead on subsets of observations,
@@
 % may benefit from other advances in this substantial literature 
-% \citep[reviewed in][]{vandermaaten2009dimensionality}.
+% (reviewed in \citep{vandermaaten2009dimensionality}).
 
@@
          In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
-         \citep[data from][]{lack2015drosophila}
+         (data from \citep{lack2015drosophila})
          In each ``INV'' denotes an inverted genotype, ``ST'' denotes the standard orientation,
@@
 We used whole-genome sequencing data 
-from the Drosophila Genome Nexus \citep[\url{http://www.johnpool.net/genomes.html},][]{lack2015drosophila},
+from the Drosophila Genome Nexus (\url{http://www.johnpool.net/genomes.html}, \citep{lack2015drosophila}),
 consisting of the Drosophila Population Genomics Project phases 1--3 \citep{langley2012genomic,pool2012population},
@@
 
-\bibliographystyle{plainnat}
 \bibliography{references}

### 2016-09-25 writeup/local_pca_paper.tex
grr submission rules

@@
 
+\newif\ifsubmission
+% \submissiontrue
+\submissionfalse
+\ifsubmission
     % % stuff for submission
     % \usepackage[nomarkers,figuresonly]{endfloat}
-% \usepackage{setspace}
-% \doublespacing
-% \usepackage{lineno}
-% \linenumbers
-% \bibliographystyle{plos2015}
-% \newcommand\citet{\cite}
-% \newcommand\citep{\cite}
-
-% % for the nice version
+    \usepackage{setspace}
+    \doublespacing
+    \usepackage{lineno}
+    \linenumbers
+    \bibliographystyle{plos2015}
+    \newcommand\citet{\cite}
+    \newcommand\citep{\cite}
+    \renewcommand\includegraphics{\relax}
+    \renewcommand{\includegraphics}[2][]{\relax}
+\else
+    % for the nice version
     \usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
     \bibliographystyle{plainnat}
+\fi
+% grrr
+\newcommand{\citetx}[2][]{\ifsubmission#1 \cite{#2}\else\citet{#2}\fi}
 
@@
 
-\citet{wright1949genetical} defined \emph{population structure} to encompass ``such matters as numbers, composition by age and sex, and state of subdivision'',
+\citetx[Wright]{wright1949genetical} defined \emph{population structure} to encompass ``such matters as numbers, composition by age and sex, and state of subdivision'',
 where ``subdivision'' refers to restricted migration between subpopulations.
@@
 For instance, 
-\citet{blair1943population} defines \emph{population structure} to be the sum total of
+\citetx[Blair]{blair1943population} defines \emph{population structure} to be the sum total of
 ``such factors as size of breeding populations, periodic fluctuation of population size, sex ratio, activity range and \emph{differential survival of progeny}'' (emphasis added).
@@
 to the genetic covariance matrix,
-as pioneered by \citet{menozzi1978synthetic}.
+as pioneered by \citetx[Menozzi et al]{menozzi1978synthetic}.
 % About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}.
@@
 showing that genomic windows in different regions of the MDS plot indeed show quite different patterns of relatedness.
-
-% Fig fig:pca_by_pop WAS HERe
 
@@
 
+\section{Supplementary Figures}
+
 \begin{figure}

### 2016-09-29 writeup/local_pca_paper_with_titlepage.tex
title page

--- 2016-09-25 writeup/local_pca_paper.tex
@@
 
-\title{Local PCA Shows How the Effect of Population Structure Differs Along the Genome}
-\author{Han Li, Peter Ralph}
-%\date{}							% Activate to display a given date or no date
-
 \begin{document}
-\maketitle
-%\section{}
-%\subsection{}
-% \doublespacing
+\vspace*{0.35in}
+
+% Title must be 250 characters or less.
+% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
+\begin{flushleft}
+{\Large
+\textbf\newline{Local PCA Shows How the Effect of Population Structure Differs Along the Genome}
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Han Li\textsuperscript{1},
+Peter Ralph\textsuperscript{2,*}
+\\
+\bigskip
+\bf{1} Molecular and Computational Biology Program, Departments of Biological Sciences, University of Southern California, Los Angeles, CA , USA
+\\
+\bf{2} Department of Mathematics, University of Oregon, Eugene, OR, USA
+\\
+
+\bigskip
+
+* plr@uoregon.edu
+
+\end{flushleft}
 

### 2016-09-29 writeup/submission/cover-letter.tex
title page

@@
 ``Local PCA Shows How the Effect of Population Structure Differs Along the Genome'',
-for consideration in \textit{JOURNAL XXX}.
+for consideration in \textit{PLOS Genetics}.
 In this paper, we set out to describe how background patterns of relatedness 
@@
 that are outliers for particular summaries of relatedness (e.g., $F_{ST}$-based scans for selection).
-We think the paper will be of interest to readers of \textit{JOURNAL XXX}
+We think the paper will be of interest to readers of \textit{PLOS Genetics}
 because the method fills an important gap in the genomicist's toolkit,
@@
 or Andy Clark (Cornell, \href{mailto:ac347@cornell.edu}{ac347@cornell.edu}).
+
+If the reviewers would like to read a version of the paper that is single-spaced and has figures included, 
+they can download it from \url{http://biorxiv.org/content/early/2016/08/21/070615}.
 

### 2016-09-30 writeup/local_pca_paper.tex
added hans edits to title page and moved figure captions to end

@@
     % % stuff for submission
-    % \usepackage[nomarkers,figuresonly]{endfloat}
+    \usepackage[nofiglist,nomarkers,figuresonly]{endfloat}
     \usepackage{setspace}
@@
     \renewcommand{\includegraphics}[2][]{\relax}
+    \renewcommand{\figurename}{Fig.}
+    \newcommand{\Figure}{Fig }
+    \newcommand{\Figures}{Fig }
 \else
@@
     \bibliographystyle{plainnat}
+    \newcommand{\Figure}{{Figure }}
+    \newcommand{\Figures}{{Figures }}
 \fi
@@
 
-\title{Local PCA Shows How the Effect of Population Structure Differs Along the Genome}
-\author{Han Li, Peter Ralph}
-%\date{}							% Activate to display a given date or no date
-
 \begin{document}
-\maketitle
-%\section{}
-%\subsection{}
-% \doublespacing
+\vspace*{0.35in}
+
+% Title must be 250 characters or less.
+% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
+\begin{flushleft}
+{\Large
+\textbf\newline{Local PCA Shows How the Effect of Population Structure Differs Along the Genome}
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Han Li\textsuperscript{1},
+Peter Ralph\textsuperscript{1,2,3,*}
+\\
+\bigskip
+\bf{1} Department of Molecular and Computational Biology, University of Southern California, Los Angeles, CA , USA
+\\
+\bf{2} Institute of Ecology and Evolution, University of Oregon, Eugene, OR, USA
+\\
+\bf{3} Department of Mathematics, University of Oregon, Eugene, OR, USA
+\\
+
+\bigskip
+
+* plr@uoregon.edu
+
+\end{flushleft}
 
@@
 
-As depicted in Figure \ref{fig:diagram}, the general steps to the method are:
+As depicted in \Figure \ref{fig:diagram}, the general steps to the method are:
 (1) divide the genome into windows,
@@
 We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
-The first column of Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
+The first column of \Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
 of the matrix of dissimilarities between genomic windows:
@@
 and that other windows' patterns of relatedness may be a mixture of those extremes. 
-The next two columns of Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
+The next two columns of \Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
 plotted against the window's position along the genome,
@@
          Vertical lines show the breakpoints of known polymorphic inversions.   
-         Solid black lines are for the inversions we used in Figure \ref{fig:pca_by_inversion},
+         Solid black lines are for the inversions we used in \Figure \ref{fig:pca_by_inversion},
          while dotted grey lines are for other known inversions.     
@@
 Representative plots are shown for three groups of windows on each chromosome arm
-in Figure \ref{fig:mds_allchr} (groups are shown in color),
-and in Supplemental Figure \ref{fig:pca_by_pop} (PCA plots).
+in \Figure \ref{fig:mds_allchr} (groups are shown in color),
+and in Supplemental \Figure \ref{fig:pca_by_pop} (PCA plots).
 The latter plots are quite different,
@@
 % Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
-To depict this, Figure \ref{fig:pca_by_inversion} shows
-the PCA plots in Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
+To depict this, \Figure \ref{fig:pca_by_inversion} shows
+the PCA plots in \Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
 Taking chromosome arm 2L as an example,
 the two regions of similar, extreme patterns of relatedness
-shown in green in the first row of Figure \ref{fig:mds_allchr}
+shown in green in the first row of \Figure \ref{fig:mds_allchr}
 lie directly around the breakpoints of the inversion In(2L)t,
-and the PCA plots in the first rows of Figure \ref{fig:pca_by_inversion}
+and the PCA plots in the first rows of \Figure \ref{fig:pca_by_inversion}
 shows that patterns of relatedness here are mostly determined by inversion orientation.
@@
 and have patterns of relatedness reflective of two axes of variation,
-seen in Figure \ref{fig:pca_by_pop},
+seen in \Figures \ref{fig:pca_by_pop} and \ref{fig:pca_by_inversion},
 which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
@@
 however, each breakpoint visibly affects patterns in the MDS coordinates
-(see vertical lines in Figure \ref{fig:mds_allchr}).
+(see vertical lines in \Figure \ref{fig:mds_allchr}).
 % \plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years.
@@
     \caption{
-         As in Figure 2, except that samples are colored by orientation of the polymorphic inversions 
-         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
-         (data from \citep{lack2015drosophila})
+        PCA plots for the three sets of genomic windows colored in \Figure \ref{fig:mds_allchr},
+        on each chromosome arm of \textit{Drosophila melanogaster}.
+        In all plots, each point represents a sample. 
+        The first column shows the combined PCA plot for windows 
+        whose points are colored green in \Figure \ref{fig:mds_allchr}; 
+        the second is for orange windows; 
+        and third is for purple windows.
+         In each, samples are colored by orientation of the polymorphic inversions 
+         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively
+         (data from \citep{lack2015drosophila}).
          In each ``INV'' denotes an inverted genotype, ``ST'' denotes the standard orientation,
@@
 any samples carrying inversions on that arm.
-In the result, shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions},
+In the result, shown in Supplemental \Figure \ref{sfig:mds_allchr_noinversions},
 the striking peaks associated with inversion breakpoints are gone,
 and previously smaller-scale variation now dominates the MDS visualization.
-For instance, the majority of the variation along 3L in Figure \ref{fig:mds_allchr}
+For instance, the majority of the variation along 3L in \Figure \ref{fig:mds_allchr}
 is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
 there is also a relatively small dip on the right end of the arm (near the centromere).
-In contrast, Supplemental Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
+In contrast, Supplemental \Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
 remaining structure is dominated by the dip near the centromere.
@@
 Indeed, correlations between the recombination rate in each window and the position on the first MDS coordinate are highly significant
-(Spearman's $\rho=0.54$, $p<2 \times 10^{-16}$; Figures \ref{fig:drosophila_recomb_rate} and \ref{sfig:drosophila_gene_density_correlations}).
+(Spearman's $\rho=0.54$, $p<2 \times 10^{-16}$; \Figures \ref{fig:drosophila_recomb_rate} and \ref{sfig:drosophila_gene_density_correlations}).
 This is consistent with the hypothesis that variation
@@
         obtained after removing samples with inversions.
-        (A plot analogous to Figure \ref{fig:mds_allchr} is shown
-        in Supplemental Figure \ref{sfig:mds_allchr_noinversions}.)
+        (A plot analogous to \Figure \ref{fig:mds_allchr} is shown
+        in Supplemental \Figure \ref{sfig:mds_allchr_noinversions}.)
         The second plot (in blue) shows
@@
 three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
-notably a particularly large, polymorphic inversion on 8p23 (Figure \ref{fig:mds_human}). 
-% for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}) 
+notably a particularly large, polymorphic inversion on 8p23 (\Figure \ref{fig:mds_human}). 
+% for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (\Figure \ref{fig:mds_human}) 
 Similar plots for all chromosomes are shown
-in Supplementary Figures \ref{sfig:mds12_chr1_8_human}, \ref{sfig:mds12_chr9_16_human}, and \ref{sfig:mds12_chr17_22_human}.
+in Supplementary \Figures \ref{sfig:mds12_chr1_8_human}, \ref{sfig:mds12_chr9_16_human}, and \ref{sfig:mds12_chr17_22_human}.
 PCA plots of many outlying windows show a characteristic trimodal shape 
-(shown for chromosome 8 in Figure \ref{sfig:pca_chr8_outliers}),
+(shown for chromosome 8 in \Figure \ref{sfig:pca_chr8_outliers}),
 presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
@@
 and found that, remarkably, 
-the inversion on chromosome 8 is still the most striking outlying signal (Figure \ref{sfig:mds1_along_allchr_human}). 
+the inversion on chromosome 8 is still the most striking outlying signal (\Figure \ref{sfig:mds1_along_allchr_human}). 
 Further investigation with a denser set of SNPs,
@@
 we applied the method to all chromosomes together.
-Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
+Results for chromosome 3 are shown in \Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
 and other chromosomes are similar:
 % the two types of population structure with higher gene density 
-% (colored green and purple in Figure \ref{fig:mds12_medicago})
+% (colored green and purple in \Figure \ref{fig:mds12_medicago})
 % are nearly identical across all chromosomes,
 % but population structure associated with low gene density (colored orange)
-% shows more chromsome-specific patterns (see Supplemental Figure \ref{sfig:pca_drosophila_allchr}).
+% shows more chromsome-specific patterns (see Supplemental \Figure \ref{sfig:pca_drosophila_allchr}).
 across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
@@
 which are shown juxtaposed with
-the first MDS coordinate of each window in Figure \ref{fig:mds_medicago},
-and are significantly correlated, as shown in Supplemental Figure \ref{sfig:mds_gene_count}.
+the first MDS coordinate of each window in \Figure \ref{fig:mds_medicago},
+and are significantly correlated, as shown in Supplemental \Figure \ref{sfig:mds_gene_count}.
 (Values shown are the number of start and end positions of each predicted mRNA transcript,
@@
        and the remaining corner (green) intermediate.
-       Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:mds_medicago_allchr}.
+       Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:mds_medicago_allchr}.
        \textbf{(below)}
-        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
+        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in \Figure \ref{fig:mds12_medicago}. 
         Each point corresponds to a sample, colored by country of origin.
-        Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:pca_peaks_medicago_allchr}.
+        Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:pca_peaks_medicago_allchr}.
        \label{fig:mds12_medicago}
@@
 indicating that we can see variation on a sufficiently fine scale.
-Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in Figure \ref{fig:diagram}),
+Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in \Figure \ref{fig:diagram}),
 along with the method used to discover common structure,
@@
 Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs;
-PCA plots for each arm are shown in Figure \ref{sfig:pca_drosophila_allchr}.
+PCA plots for each arm are shown in \Figure \ref{sfig:pca_drosophila_allchr}.
 
@@
 We excluded the sex chromosomes and the mitochondria.
-PCA plots for each chromosome, separately, are shown in Figure \ref{sfig:pca_human_allchr}.
+PCA plots for each chromosome, separately, are shown in \Figure \ref{sfig:pca_human_allchr}.
 % We use the allele that has highest frequency in the samples as the reference allele for each position.
@@
 Each of the 8 chromosomes has 3--5 million SNPs;
-PCA plots for these are shown in Figure \ref{sfig:pca_medicago_allchr}.
+PCA plots for these are shown in \Figure \ref{sfig:pca_medicago_allchr}.
 We did not use the mitochondria or chloroplasts.
@@
 \bibliography{references}  
+\ifsubmission
+    \clearpage
+    \newcounter{bibend}
+    \setcounter{bibend}{\value{page}}
+\fi
+
+\ifsubmission\processdelayedfloats\fi
+
+\ifsubmission
+    \clearpage
+    \setcounter{page}{\thebibend}
+\fi
 
@@
 \renewcommand{\thefigure}{S\arabic{figure}}
+\ifsubmission
+    \setcounter{postfigure}{0}
+    \renewcommand{\thepostfigure}{S\arabic{postfigure}}
+\fi
 
@@
 
+% put in list of figures here 
+\ifsubmission
+    \listoffigures
+    \clearpage
+\fi
+
 \section{Supplementary Figures}
@@
     \caption{      
-        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_allchr},
+        PCA plots for the three sets of genomic windows colored in \Figure \ref{fig:mds_allchr},
         on each chromosome arm of \textit{Drosophila melanogaster}.
@@
         The first column shows the combined PCA plot for windows 
-        whose points are colored green in Figure \ref{fig:mds_allchr}; 
+        whose points are colored green in \Figure \ref{fig:mds_allchr}; 
         the second is for orange windows; 
@@
         across \textit{Drosophila melanogaster} chromosome arms: without inversions.
-         As in Figure \ref{fig:mds_allchr}, but after omitting for each chromosome arm individuals carrying the less frequent orientation
+         As in \Figure \ref{fig:mds_allchr}, but after omitting for each chromosome arm individuals carrying the less frequent orientation
          of any inversions on that chromosome arm.
@@
         Recombination rate, and the effects of population structure for \textit{Drosophila melanogaster}:
-        this shows the first MDS coordinate and recombination rate (in cM/Mbp), as in Supplementary Figure \ref{fig:drosophila_recomb_rate},
+        this shows the first MDS coordinate and recombination rate (in cM/Mbp), as in Supplementary \Figure \ref{fig:drosophila_recomb_rate},
         against each other.
-        Since the windows underlying estimates of Supplementary Figure \ref{fig:drosophila_recomb_rate} do not coincide,
+        Since the windows underlying estimates of Supplementary \Figure \ref{fig:drosophila_recomb_rate} do not coincide,
         to obtain correlations we divided the genome into 100Kbp bins, 
@@
     \caption{
-        PCA plots for regions colored in Figure \ref{sfig:mds_medicago_allchr}
+        PCA plots for regions colored in \Figure \ref{sfig:mds_medicago_allchr}
         on all 8 chromosomes of \textit{Medicago truncatula}:
@@
 
+\ifsubmission\processdelayedfloats\fi
+
 \end{document}

### 2016-10-05 writeup/local_pca_paper.tex
added author summary

@@
 can have on observed patterns of genetic variation.
+
+\section*{Author Summary}
+
+How individuals of a species in different parts of the world are related to
+each other is shaped by the history of migration and gene flow.  However,
+natural selection affects some parts of the genome in different ways than
+others: for instance, regions with more genes are more strongly affected by
+deleterious mutations. In this paper we develop a visualization method to see
+how patterns of mean relatedness vary along the genome, and aim to identify the
+underlying biological cause. In three species: Medicago truncatula, Drosophila
+melanogaster, and modern humans, we find significant heterogeneity in patterns
+of relatedness along the genome that is correlated with genome structure. For
+humans and Drosophila, this variation is closely connected to polymorphic
+chromosomal inversions, while for Medicago, and for Drosophila once inversions
+are removed, the variation is correlated with local gene density, and is
+perhaps due to linked selection.
 

### 2016-11-21 writeup/local_pca_paper.tex
Moved the reordering of the text to this branch while keeping correct figures.

--- 2016-09-30 writeup/local_pca_paper.tex
@@
 since the method is an visualization method, 
-we allow ourselves to be suprised by unexpected signals in the data.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
+we allow ourselves to be surprised by unexpected signals in the data.
+
+%%%%%%%%%%%%%%%%%
+\section{Materials and Methods}
 
@@
 and (5) combine similar windows to more accurately visualize local effects of population structure using PCA.
-Details of how we carried out each step are given in the Methods,
-and code to run the analyses is provided as an R package at \url{https://github.com/petrelharp/local_pca}.
+% Details of how we carried out each step are given in the Methods,
+% and code to run the analyses is provided as an R package at \url{https://github.com/petrelharp/local_pca}.
 
@@
 
-We applied the method to genomic datasets with good geographic sampling:
-380 African \textit{Drosophila melanogaster} from the Drosophila Genome Nexus \citep{lack2015drosophila},
-a worldwide dataset of humans,
-3,965 humans from several locations worldwide from the POPRES dataset \citep{nelson2008population},
-and 263 \textit{Medicago truncatula} from 24 countries around the Mediterranean basin 
-a range-wide dataset of the partially selfing weedy annual plant 
-from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
-
-In all three species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
-This implies that variation is due to meaningful heterogeneity in a biological process,
-since noise due to randomness in choice of local genealogical trees
-is not expected to show long distance correlations. 
-Below, we discuss the results and likely underlying causes.
-
-%%%%%%%%%%%
-\subsection{\textit{Drosophila melanogaster}}
-
-We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
-The first column of \Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
-of the matrix of dissimilarities between genomic windows:
-in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
-For each chromosome arm, the MDS visualization roughly resembles a triangle,
-sometimes with additional points.
-Since the relative position of each window in this plot shows the similarity between windows, 
-this suggests that there are at least three extreme manifestations of population structure 
-typified by windows found in the ``corners'' of the figure,
-and that other windows' patterns of relatedness may be a mixture of those extremes. 
-The next two columns of \Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
-plotted against the window's position along the genome,
-to show how the plot of the first column is laid out along the genome.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
-    \end{center}
-    \caption{
-        Variation in patterns of relatedness for windows across \textit{Drosophila melanogaster} chromosome arms.
-        In all plots, each point represents one window along the genome.
-         The first column shows the MDS visualization of relationships between windows, 
-         and the second and third columns show 
-        the two MDS coordinates
-         against the midpoint of each window;
-         rows correspond to chromosome arms.
-         Colors are consistent for plots in each row. 
-         Vertical lines show the breakpoints of known polymorphic inversions.   
-         Solid black lines are for the inversions we used in \Figure \ref{fig:pca_by_inversion},
-         while dotted grey lines are for other known inversions.     
-         \label{fig:mds_allchr}
-    }
-\end{figure}
-
-To help visualize how clustered windows with similar patterns of relatedness are along each chromosome arm, 
-we selected three ``extreme'' windows in the MDS plot
-and the 5\% of windows that are closest to it in the MDS coordinates,
-then highlighted these windows' positions along the genome,
-and created PCA plots for the windows, combined.
-Representative plots are shown for three groups of windows on each chromosome arm
-in \Figure \ref{fig:mds_allchr} (groups are shown in color),
-and in Supplemental \Figure \ref{fig:pca_by_pop} (PCA plots).
-The latter plots are quite different,
-showing that genomic windows in different regions of the MDS plot indeed show quite different patterns of relatedness.
-
-The most striking variation in patterns of relatedness turns out to be explained by
-several large inversions that are polymorphic in these samples, 
-discussed in \citet{corbett2012population} and \citet{langley2012genomic}.
-% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
-To depict this, \Figure \ref{fig:pca_by_inversion} shows
-the PCA plots in \Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
-Taking chromosome arm 2L as an example,
-the two regions of similar, extreme patterns of relatedness
-shown in green in the first row of \Figure \ref{fig:mds_allchr}
-lie directly around the breakpoints of the inversion In(2L)t,
-and the PCA plots in the first rows of \Figure \ref{fig:pca_by_inversion}
-shows that patterns of relatedness here are mostly determined by inversion orientation.
-The regions shown in purple on chromosome 2L lie near the centromere,
-and have patterns of relatedness reflective of two axes of variation,
-seen in \Figures \ref{fig:pca_by_pop} and \ref{fig:pca_by_inversion},
-which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
-(see \citet{lack2015drosophila} for more about admixture in this sample).
-The regions shown in orange on chromosome 2L mostly lie inside the inversion,
-and show patterns of relatedness that are a mixture between the other two,
-as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
-Similar results are found in other chromosome arms,
-albeit complicated by the coexistence of more than one polymorphic inversion;
-however, each breakpoint visibly affects patterns in the MDS coordinates
-(see vertical lines in \Figure \ref{fig:mds_allchr}).
-% \plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years.
-% Also cite \citet{corbettdetig2012sequencebased} for a computational method for detecting inversions.}
-% corbettdetig2012sequencebased suggests these inversions are recent-ish (but before out-of-Africa)
-% except that the situation of Chr3R is a little more complicated,
-% due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
-% However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
-% although we have not highlighted portions of the MDS plot that pick these out.
-% The inversions In(3L)P and In(3R)Mo, are at low frequency in these samples, 
-% and visibly affect the MDS plots although less than the more common inversion on each arm.
-% The very young inversion In(X)be shows little signal 
-% compared to the older inversion In(X)A \citep{corbett2012population}.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
-    \end{center}
-    \caption{
-        PCA plots for the three sets of genomic windows colored in \Figure \ref{fig:mds_allchr},
-        on each chromosome arm of \textit{Drosophila melanogaster}.
-        In all plots, each point represents a sample. 
-        The first column shows the combined PCA plot for windows 
-        whose points are colored green in \Figure \ref{fig:mds_allchr}; 
-        the second is for orange windows; 
-        and third is for purple windows.
-         In each, samples are colored by orientation of the polymorphic inversions 
-         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively
-         (data from \citep{lack2015drosophila}).
-         In each ``INV'' denotes an inverted genotype, ``ST'' denotes the standard orientation,
-         and ``N'' denotes unknown.
-        \label{fig:pca_by_inversion}
-    }
-\end{figure}
-
-To see how patterns of relatedness vary in the absence of polymorphic inversions,
-we performed the same analyses after removing, for each chromosome arm,
-any samples carrying inversions on that arm.
-In the result, shown in Supplemental \Figure \ref{sfig:mds_allchr_noinversions},
-the striking peaks associated with inversion breakpoints are gone,
-and previously smaller-scale variation now dominates the MDS visualization.
-For instance, the majority of the variation along 3L in \Figure \ref{fig:mds_allchr}
-is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
-there is also a relatively small dip on the right end of the arm (near the centromere).
-In contrast, Supplemental \Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
-remaining structure is dominated by the dip near the centromere.
-Without inversions, variation in patterns of relatedness shown in the MDS plots
-follows similar patterns to that previously seen in \textit{D.~melanogaster} recombination rate and diversity \citep{langley2012genomic,mackay2012drosophila}.
-Indeed, correlations between the recombination rate in each window and the position on the first MDS coordinate are highly significant
-(Spearman's $\rho=0.54$, $p<2 \times 10^{-16}$; \Figures \ref{fig:drosophila_recomb_rate} and \ref{sfig:drosophila_gene_density_correlations}).
-This is consistent with the hypothesis that variation
-is due to selection, since the strength of linked selection increases with local gene density, 
-measured in units of recombination distance.
-The number of genes -- measured as the number of transcription start and end sites within each window --
-was not significantly correlated with MDS coordinate ($p=0.22$).
-
-\begin{figure}
-    \begin{center}
-        \includegraphics{drosophila_recomb_mds}
-    \end{center}
-    \caption{
-        The effects of population structure without inversions is correlated to recombination rate in \textit{Drosophila melanogaster}.
-        The first plot (in red) shows the first MDS coordinate along the genome 
-        for windows of 10,000 SNPs,
-        obtained after removing samples with inversions.
-        (A plot analogous to \Figure \ref{fig:mds_allchr} is shown
-        in Supplemental \Figure \ref{sfig:mds_allchr_noinversions}.)
-        The second plot (in blue) shows
-        local average recombination rates in cM/Mbp, obtained as midpoint estimates for 100Kbp windows 
-        from the Drosophila recombination rate calculator \citep{fistonlavier2010drosophila} release 5,
-        using rates from \citet{comeron2012many}.
-        The third plot (in black) shows the number of genes' transcription start and end sites within each 100Kbp window, divided by two.
-        Transcription start and end sites were obtained from the RefGene table from the UCSC browser. %, assembly dm3, on August 11, 2016.
-        The histone gene cluster on chromosome arm 2L is excluded.
-        \label{fig:drosophila_recomb_rate}
-    }
-\end{figure}
-
-%%%%%%%%%%%
-\subsection{Human}
-
-As we did for the \textit{Drosophila} data, we applied our method separately to all 22 human autosomes.
-On each, variation in patterns of relatedness was dominated by a small number of windows
-having similar patterns of relatedness to each other that differed dramatically from the rest of the chromosome.
-These may be primarily inversions: outlying windows coincide with
-three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
-notably a particularly large, polymorphic inversion on 8p23 (\Figure \ref{fig:mds_human}). 
-% for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (\Figure \ref{fig:mds_human}) 
-Similar plots for all chromosomes are shown
-in Supplementary \Figures \ref{sfig:mds12_chr1_8_human}, \ref{sfig:mds12_chr9_16_human}, and \ref{sfig:mds12_chr17_22_human}.
-PCA plots of many outlying windows show a characteristic trimodal shape 
-(shown for chromosome 8 in \Figure \ref{sfig:pca_chr8_outliers}),
-presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
-(although we do not have data on orientation status).
-This trimodal shape has been proposed as a method to identify inversions \citep{ma2012investigation},
-but distinguishing this hypothesis from others,
-such as regions of low recombination rate,
-would require additional data.
-
-We also applied the method on all 22 autosomes together, 
-and found that, remarkably, 
-the inversion on chromosome 8 is still the most striking outlying signal (\Figure \ref{sfig:mds1_along_allchr_human}). 
-Further investigation with a denser set of SNPs,
-allowing a finer genomic resolution,
-may yield other patterns.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
-    \end{center}
-    \caption{
-         Variation in structure between windows on human chromosomes 8, 15, and 17. 
-         Each point in each plot represents a window.
-         The first column shows the MDS visualization of relationships between windows;
-         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
-         Rows, from top to bottom show chromosomes 8, 15, and 17. 
-         The vertical red lines show the breakpoints of known inversions from \citet{antonacci2009characterization}.
-        \label{fig:mds_human}
-    }
-\end{figure}
-
-%%%%%%%%%%%
-\subsection{\textit{Medicago truncatula}}
-
-Unlike the other two species,
-the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
-showed similar patterns of gradual change in patterns of relatedness across each chromosome,
-with no indications of chromosome-specific patterns.
-This consistency suggests that the factor affecting the population structure for each chromosome is the same,
-as might be caused by varying strengths of linked selection.
-To verify that variation in the effects of population structure is shared across chromosomes,
-we applied the method to all chromosomes together.
-Results for chromosome 3 are shown in \Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
-and other chromosomes are similar:
-% the two types of population structure with higher gene density 
-% (colored green and purple in \Figure \ref{fig:mds12_medicago})
-% are nearly identical across all chromosomes,
-% but population structure associated with low gene density (colored orange)
-% shows more chromsome-specific patterns (see Supplemental \Figure \ref{sfig:pca_drosophila_allchr}).
-across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
-which often have lower gene density and may therefore be less subject to linked selection.
-% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
-To verify that this is a possible explanation,
-we counted the number of genes found in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
-which are shown juxtaposed with
-the first MDS coordinate of each window in \Figure \ref{fig:mds_medicago},
-and are significantly correlated, as shown in Supplemental \Figure \ref{sfig:mds_gene_count}.
-(Values shown are the number of start and end positions of each predicted mRNA transcript,
-divided by two, assigned to the nearest window.)
-% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
-However, other genomic features, such as distance to centromere show roughly the same patterns,
-so we cannot rule out alternative hypotheses.
-In particular, the recombination rates estimated by \citet{paape2012finescale} appear visually to be similar,
-but were not available in a form that can be compared to our MDS values.
-
-We also found nearly identical results when choosing shorter windows of 1,000 SNPs;
-or choosing windows of equal length in base pairs rather than SNPs.
-Similarly, the results were not substantially changed
-when using weighted PCA to downweight the large group of Tunisian samples.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig6_Together_MDS_plot_chr3_final}
-       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
-    \end{center}
-    \caption{
-       MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3,
-       with corresponding PCA plots.
-       Each point in the plot represents a window;
-       the structure revealed by the MDS plot is strongly clustered along the chromosome,
-       with windows in the upper-right corner of the MDS plot (colored red) clustered around the centromere,
-       windows in the upper-left corner (purple) furthest from the centromere,
-       and the remaining corner (green) intermediate.
-       Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:mds_medicago_allchr}.
-       \textbf{(below)}
-        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in \Figure \ref{fig:mds12_medicago}. 
-        Each point corresponds to a sample, colored by country of origin.
-        Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:pca_peaks_medicago_allchr}.
-       \label{fig:mds12_medicago}
-       \label{fig:pca_medicago}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig5_MDS_and_gene_count_allchr_update_correction}
-    \end{center}
-    \caption{
-         MDS coordinate and gene density for each window in the \textit{Medicago} genome,
-         for chromosomes 1--8 (numbered above each pair of figures).
-         For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
-         The black plot below is gene count for each window against the position of each window.
-         \label{fig:mds_medicago}
-    }
-\end{figure}
-
-%%%%%%% %%%%%%%%%
-\section{Discussion}
-
-Our investigations have found substantial variation in the patterns of relatedness formed by population structure across the genomes
-of three diverse species,
-revealing distinct biological processes driving this variation in each species.
-More investigation, particularly on more species and datasets, will help to uncover what aspects of species history can explain these differences.
-With growing appreciation of the heterogeneous effects of selection across the genome,
-especially the importance of adaptive introgression and hybrid speciation \citep{pool2015natural,brandvain2014speciation,hufford2013genomic,fitzpatrick2010rapid,staubach2012genome},
-local adaptation \citep{lenormand2002limits,wang2014isolation},
-and inversion polymorphisms \citep{kirkpatrick2015chromosome,kirkpatrick2010chromosome},
-local PCA may prove to be a useful exploratory tool to discover important genomic features.
-% It is unclear whether the technique will be useful on reduced representation genotyping datasets 
-% due to marker density and issues with missing data --
-% our investigations with one such dataset were inconclusive --
-% but even low coverage, whole-genome sequence is very promising.
-
-We now discuss possible implications of this variation in the effects of population structure,
-the impact of various parameter choices in implementing the method,
-and possible additional applications.
-
-\paragraph{Chromosomal inversions}
-A major driver of variation in patterns of relatedness in two datasets we examined are inversions.
-This may be common,
-but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
-PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
-however, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
-or simply regions of reduced recombination
-without additional lines of evidence.
-Inversions show up in our method because across the inverted region,
-most gene trees share a common split that dates back to the origin of the inversion.
-However, in many applications, inversions are a nuisance.
-For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
-by regressing out the effect of linked SNPs on each other.
-Removing samples with the less common orientation of each inversion reduced,
-but did not eliminate, the signal of inversions
-seen in the \textit{Drosophila melanogaster} dataset,
-demonstrating that the genomic effects of transiently polymorphic inversions
-may outlast the inversions themselves.
-
-\paragraph{The effect of selection}
-It seems that
-the variation in patterns of relatedness we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
-must be explained somehow by linked selection.
-Furthermore, 
-the selection must be affecting many targets across the genome,
-since we see similar effects across long distances (even distinct chromosomes).
-For this reason, the most likely candidate may be selection against linked deleterious mutations, known as ``background selection'' \citep{charlesworth1993effect,charlesworth2013background}.
-Informally, background selection reduces the number of potential contributors to the gene pool 
-in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious};
-for this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently,
-Therefore, regions of the genome harboring many targets of local adaptation may show similar patterns,
-since migrant alleles in these regions will be selected against,
-and so locally gene trees will more closely reflect spatial proximity.
-
-A related possibility is that variation in patterns of relatedness is due to recent admixture between previously separated populations,
-the effects of which were not uniform across the genome due to selection.
-For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
-is due to selection against Neanderthal genes, leading to greater introgression in regions of lower gene density
-\citep{harris2016genetic,juric2016strength}.
-African \textit{Drosophila melanogaster} are known to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
-if selection regularly favors genes from one origin,
-this could lead to substantial variation in patterns of relatedness correlated with local gene density.
-
-There has been substantial debate over the relative impacts of different forms of selection.
-These have been difficult to disentangle in part because for the most part
-theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
-and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
-It may be possible to design more powerful statistics that make stronger use of spatial information.
-
-% How do we expect selection to affect population structure?
-% PCA summarizes patterns in kinship found in the genetic covariance matrix,
-% which is an average across locus-specific genealogies:
-% if individuals are closer in the genealogies of a given genomic region, 
-% they tend to be closer in the PCA maps for that region. 
-% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
-% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
-% Selective sweeps cause local recent ancestry or short trees.
-% Background selection causes shallow ones.
-% Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
-% characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
-% A genomic region with many targets for selection experiences
-% background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
-% which would tend to shorten genealogical trees in the region,
-% similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
-% % Balancing selection causes deep trees.
-% On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
-% and population structure locally describes which individuals have which alleles
-% rather than geographical proximity.
-% Finally, 
-% since recombination is suppressed between opposite orientations of a chromosomal inversion
-% near its breakpoints,
-% genealogies in these regions separate samples carrying the two orientations of the inversion.
-% In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
-% (indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
-% as a way to identify inversions).
-% if a chromosomal inversion is polymorphic in the sample, 
-% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints.
-Recombination suppression across inversions thus results in different genome structure and population structure.
-% Other effects, like noise, introgression might also influence population structure.
-% 
-% Many of the effects listed above (such as single selective sweeps or inversions) 
-% are not expected to have similar effects on population structure in different regions of the genome
-% because of randomness in which samples end up in which group;
-% but if these coincide with a region of reduced recombination (or an inversion),
-% these could drive major patterns of variation.
-% The more subtle effects of genome-wide linked selection could be shared across large regions
-% due to large-scale variation in gene density
-% (as by background selection or local adaptation at many genes).
-
-\paragraph{Parameter choices}
-There are several choices in the method that may in principle affect the results.
-As with whole-genome PCA,
-the choice of samples is important,
-as variation not strongly represented in the sample will not be discovered.
-The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
-but downweighting may be a better option that does not discard data
-(and here we present a method to do this).
-Next, the choice of window size may be important,
-although in our applications results were not sensitive to this,
-indicating that we can see variation on a sufficiently fine scale.
-Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in \Figure \ref{fig:diagram}),
-along with the method used to discover common structure,
-will affect results.
-We used MDS, applied to either each chromosome separately or to the entire genome;
-for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosome,
-but genome-wide, their signal is obscured by the numerous other signals of comparable strength.
-
-Besides window length, there is also the question of how to choose windows.
-In these applications we have used nonoverlapping windows with equal numbers of polymorphic sites.
-Alternatively, windows could be chosen to have equal length in genetic distance,
-so that each would have roughly the same number of independent trees.
-However, we found little change in results when using different window sizes
-or when measuring windows in physical distance (in bp).
-
-% More generally, there are many possible methods to discover common structure in different parts of the genome:
-% in this work we use PCA to visualize the effects of population structure,
-% but other methods, such as STRUCTURE \citep{falush2003inference},
-% SPA \citep{yang2012modelbased},
-% SpaceMix \citep{bradburd2015spatial},
-% or other matrix factorization methods \citep{engelhardt2010analysis}
-% may highlight different sources of variation in the data.
-% It is also possible that other methods for measuring dissimilarity between windows' covariance matrices
-% or for summarizing the matrix of pairwise distances between windows
-% would lead to different insights.
-
-Finally, our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
-and how many MDS coordinates to use when describing the distance matrix between windows,
-but in our exploration, changing these has not produced dramatically different results.
-These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
-we encourage the user to experiment.
-
-\paragraph{Applications}
-So-called cryptic relatedness between samples
-has been one of the major sources of confounding in genome-wide association studies (GWAS)
-and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
-Since the effects of population structure is not constant along the genome,
-this could in principle lead to an inflation of false positives in parts of the genome
-with stronger population structure than the genome-wide average.
-A method such as ours might be used to provide a more sensitive correction.
-Fortunately, in our human dataset this does not seem likely to have a strong effect:
-most variation is due to small, independent regions, possibly primarily inversions,
-and so may not have a major effect on GWAS.
-In the other species we examined, particularly \textit{Drosophila melanogaster},
-treating population structure as a single quantity would entail a substantial loss of power,
-and could potentially be misleading.
-
-%%%%%%%%%%%%%%%%%
-\section{Methods}
-
 \subsection{PCA in genomic windows}
 
-We first recoded sampled genotypes as numeric matrices in the usual manner,
+To begin,
+we first recoded sampled genotypes as numeric matrices in the usual manner,
 by recording the number of nonreference alleles seen at each locus for each sample.
@@
 Furthermore, to remove the effect of artifacts such as mutation rate variation,
-we also rescale each approximate covariance matrix 
-so that the underlying data matrix has trace norm equal to one.
+we also rescale each approximate covariance matrix to be of similar size
+(precisely, so that the underlying data matrix has trace norm equal to one).
 
@@
 
-The three publicly available datasets we used are summarized in Table \ref{tab:data_stats}.
-We converted each to a numeric matrix
-(with one row per polymorphic variant and one column per sample)
-by replacing each genotype with the number of nonreference alleles
-(or NA for missing data).
-% A normalization step (see below) ensures the result does not depend on the choice of reference allele.
+We applied the method to genomic datasets with good geographic sampling:
+380 African \textit{Drosophila melanogaster} from the Drosophila Genome Nexus \citep{lack2015drosophila},
+a worldwide dataset of humans,
+3,965 humans from several locations worldwide from the POPRES dataset \citep{nelson2008population},
+and 263 \textit{Medicago truncatula} from 24 countries around the Mediterranean basin 
+a range-wide dataset of the partially selfing weedy annual plant 
+from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
+as summarized in Table \ref{tab:data_stats}.
 
@@
 
-\section{Data access}
+\subsection{Data access}
 
@@
 and \textit{Drosophila} at the Drosophila Genome Nexus, \url{http://www.johnpool.net/genomes.html}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Results}
+
+In all three datasets:
+a worldwide sample of humans, African \textit{Drosophila melanogaster},
+and a rangewide sample of \textit{Medicago truncatula},
+PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
+This implies that variation is due to meaningful heterogeneity in a biological process,
+since noise due to randomness in choice of local genealogical trees
+is not expected to show long distance correlations. 
+Below, we discuss the results and likely underlying causes.
+
+%%%%%%%%%%%
+\subsection{\textit{Drosophila melanogaster}}
+
+We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
+The first column of \Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
+of the matrix of dissimilarities between genomic windows:
+in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
+For each chromosome arm, the MDS visualization roughly resembles a triangle,
+sometimes with additional points.
+Since the relative position of each window in this plot shows the similarity between windows, 
+this suggests that there are at least three extreme manifestations of population structure 
+typified by windows found in the ``corners'' of the figure,
+and that other windows' patterns of relatedness may be a mixture of those extremes. 
+The next two columns of \Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
+plotted against the window's position along the genome,
+to show how the plot of the first column is laid out along the genome.
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
+    \end{center}
+    \caption{
+        Variation in patterns of relatedness for windows across \textit{Drosophila melanogaster} chromosome arms.
+        In all plots, each point represents one window along the genome.
+         The first column shows the MDS visualization of relationships between windows, 
+         and the second and third columns show 
+        the two MDS coordinates
+         against the midpoint of each window;
+         rows correspond to chromosome arms.
+         Colors are consistent for plots in each row. 
+         Vertical lines show the breakpoints of known polymorphic inversions.   
+         Solid black lines are for the inversions we used in \Figure \ref{fig:pca_by_inversion},
+         while dotted grey lines are for other known inversions.     
+         \label{fig:mds_allchr}
+    }
+\end{figure}
+
+To help visualize how clustered windows with similar patterns of relatedness are along each chromosome arm, 
+we selected three ``extreme'' windows in the MDS plot
+and the 5\% of windows that are closest to it in the MDS coordinates,
+then highlighted these windows' positions along the genome,
+and created PCA plots for the windows, combined.
+Representative plots are shown for three groups of windows on each chromosome arm
+in \Figure \ref{fig:mds_allchr} (groups are shown in color),
+and in Supplemental \Figure \ref{fig:pca_by_pop} (PCA plots).
+The latter plots are quite different,
+showing that genomic windows in different regions of the MDS plot indeed show quite different patterns of relatedness.
+
+The most striking variation in patterns of relatedness turns out to be explained by
+several large inversions that are polymorphic in these samples, 
+discussed in \citet{corbett2012population} and \citet{langley2012genomic}.
+% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
+To depict this, \Figure \ref{fig:pca_by_inversion} shows
+the PCA plots in \Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
+Taking chromosome arm 2L as an example,
+the two regions of similar, extreme patterns of relatedness
+shown in green in the first row of \Figure \ref{fig:mds_allchr}
+lie directly around the breakpoints of the inversion In(2L)t,
+and the PCA plots in the first rows of \Figure \ref{fig:pca_by_inversion}
+shows that patterns of relatedness here are mostly determined by inversion orientation.
+The regions shown in purple on chromosome 2L lie near the centromere,
+and have patterns of relatedness reflective of two axes of variation,
+seen in \Figures \ref{fig:pca_by_pop} and \ref{fig:pca_by_inversion},
+which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
+(see \citet{lack2015drosophila} for more about admixture in this sample).
+The regions shown in orange on chromosome 2L mostly lie inside the inversion,
+and show patterns of relatedness that are a mixture between the other two,
+as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
+Similar results are found in other chromosome arms,
+albeit complicated by the coexistence of more than one polymorphic inversion;
+however, each breakpoint visibly affects patterns in the MDS coordinates
+(see vertical lines in \Figure \ref{fig:mds_allchr}).
+% \plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years.
+% Also cite \citet{corbettdetig2012sequencebased} for a computational method for detecting inversions.}
+% corbettdetig2012sequencebased suggests these inversions are recent-ish (but before out-of-Africa)
+% except that the situation of Chr3R is a little more complicated,
+% due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
+% However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
+% although we have not highlighted portions of the MDS plot that pick these out.
+% The inversions In(3L)P and In(3R)Mo, are at low frequency in these samples, 
+% and visibly affect the MDS plots although less than the more common inversion on each arm.
+% The very young inversion In(X)be shows little signal 
+% compared to the older inversion In(X)A \citep{corbett2012population}.
+
+\begin{figure}
+    \begin{center}
+       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
+    \end{center}
+    \caption{
+        PCA plots for the three sets of genomic windows colored in \Figure \ref{fig:mds_allchr},
+        on each chromosome arm of \textit{Drosophila melanogaster}.
+        In all plots, each point represents a sample. 
+        The first column shows the combined PCA plot for windows 
+        whose points are colored green in \Figure \ref{fig:mds_allchr}; 
+        the second is for orange windows; 
+        and third is for purple windows.
+         In each, samples are colored by orientation of the polymorphic inversions 
+         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively
+         (data from \citep{lack2015drosophila}).
+         In each ``INV'' denotes an inverted genotype, ``ST'' denotes the standard orientation,
+         and ``N'' denotes unknown.
+        \label{fig:pca_by_inversion}
+    }
+\end{figure}
+
+To see how patterns of relatedness vary in the absence of polymorphic inversions,
+we performed the same analyses after removing, for each chromosome arm,
+any samples carrying inversions on that arm.
+In the result, shown in Supplemental \Figure \ref{sfig:mds_allchr_noinversions},
+the striking peaks associated with inversion breakpoints are gone,
+and previously smaller-scale variation now dominates the MDS visualization.
+For instance, the majority of the variation along 3L in \Figure \ref{fig:mds_allchr}
+is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
+there is also a relatively small dip on the right end of the arm (near the centromere).
+In contrast, Supplemental \Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
+remaining structure is dominated by the dip near the centromere.
+Without inversions, variation in patterns of relatedness shown in the MDS plots
+follows similar patterns to that previously seen in \textit{D.~melanogaster} recombination rate and diversity \citep{langley2012genomic,mackay2012drosophila}.
+Indeed, correlations between the recombination rate in each window and the position on the first MDS coordinate are highly significant
+(Spearman's $\rho=0.54$, $p<2 \times 10^{-16}$; \Figures \ref{fig:drosophila_recomb_rate} and \ref{sfig:drosophila_gene_density_correlations}).
+This is consistent with the hypothesis that variation
+is due to selection, since the strength of linked selection increases with local gene density, 
+measured in units of recombination distance.
+The number of genes -- measured as the number of transcription start and end sites within each window --
+was not significantly correlated with MDS coordinate ($p=0.22$).
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{drosophila_recomb_mds}
+    \end{center}
+    \caption{
+        The effects of population structure without inversions is correlated to recombination rate in \textit{Drosophila melanogaster}.
+        The first plot (in red) shows the first MDS coordinate along the genome 
+        for windows of 10,000 SNPs,
+        obtained after removing samples with inversions.
+        (A plot analogous to \Figure \ref{fig:mds_allchr} is shown
+        in Supplemental \Figure \ref{sfig:mds_allchr_noinversions}.)
+        The second plot (in blue) shows
+        local average recombination rates in cM/Mbp, obtained as midpoint estimates for 100Kbp windows 
+        from the Drosophila recombination rate calculator \citep{fistonlavier2010drosophila} release 5,
+        using rates from \citet{comeron2012many}.
+        The third plot (in black) shows the number of genes' transcription start and end sites within each 100Kbp window, divided by two.
+        Transcription start and end sites were obtained from the RefGene table from the UCSC browser. %, assembly dm3, on August 11, 2016.
+        The histone gene cluster on chromosome arm 2L is excluded.
+        \label{fig:drosophila_recomb_rate}
+    }
+\end{figure}
+
+%%%%%%%%%%%
+\subsection{Human}
+
+As we did for the \textit{Drosophila} data, we applied our method separately to all 22 human autosomes.
+On each, variation in patterns of relatedness was dominated by a small number of windows
+having similar patterns of relatedness to each other that differed dramatically from the rest of the chromosome.
+These may be primarily inversions: outlying windows coincide with
+three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
+notably a particularly large, polymorphic inversion on 8p23 (\Figure \ref{fig:mds_human}). 
+% for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (\Figure \ref{fig:mds_human}) 
+Similar plots for all chromosomes are shown
+in Supplementary \Figures \ref{sfig:mds12_chr1_8_human}, \ref{sfig:mds12_chr9_16_human}, and \ref{sfig:mds12_chr17_22_human}.
+PCA plots of many outlying windows show a characteristic trimodal shape 
+(shown for chromosome 8 in \Figure \ref{sfig:pca_chr8_outliers}),
+presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
+(although we do not have data on orientation status).
+This trimodal shape has been proposed as a method to identify inversions \citep{ma2012investigation},
+but distinguishing this hypothesis from others,
+such as regions of low recombination rate,
+would require additional data.
+
+We also applied the method on all 22 autosomes together, 
+and found that, remarkably, 
+the inversion on chromosome 8 is still the most striking outlying signal (\Figure \ref{sfig:mds1_along_allchr_human}). 
+Further investigation with a denser set of SNPs,
+allowing a finer genomic resolution,
+may yield other patterns.
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
+    \end{center}
+    \caption{
+         Variation in structure between windows on human chromosomes 8, 15, and 17. 
+         Each point in each plot represents a window.
+         The first column shows the MDS visualization of relationships between windows;
+         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
+         Rows, from top to bottom show chromosomes 8, 15, and 17. 
+         The vertical red lines show the breakpoints of known inversions from \citet{antonacci2009characterization}.
+        \label{fig:mds_human}
+    }
+\end{figure}
+
+%%%%%%%%%%%
+\subsection{\textit{Medicago truncatula}}
+
+Unlike the other two species,
+the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
+showed similar patterns of gradual change in patterns of relatedness across each chromosome,
+with no indications of chromosome-specific patterns.
+This consistency suggests that the factor affecting the population structure for each chromosome is the same,
+as might be caused by varying strengths of linked selection.
+To verify that variation in the effects of population structure is shared across chromosomes,
+we applied the method to all chromosomes together.
+Results for chromosome 3 are shown in \Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
+and other chromosomes are similar:
+% the two types of population structure with higher gene density 
+% (colored green and purple in \Figure \ref{fig:mds12_medicago})
+% are nearly identical across all chromosomes,
+% but population structure associated with low gene density (colored orange)
+% shows more chromsome-specific patterns (see Supplemental \Figure \ref{sfig:pca_drosophila_allchr}).
+across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
+which often have lower gene density and may therefore be less subject to linked selection.
+% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
+To verify that this is a possible explanation,
+we counted the number of genes found in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
+which are shown juxtaposed with
+the first MDS coordinate of each window in \Figure \ref{fig:mds_medicago},
+and are significantly correlated, as shown in Supplemental \Figure \ref{sfig:mds_gene_count}.
+(Values shown are the number of start and end positions of each predicted mRNA transcript,
+divided by two, assigned to the nearest window.)
+% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
+However, other genomic features, such as distance to centromere show roughly the same patterns,
+so we cannot rule out alternative hypotheses.
+In particular, the recombination rates estimated by \citet{paape2012finescale} appear visually to be similar,
+but were not available in a form that can be compared to our MDS values.
+
+We also found nearly identical results when choosing shorter windows of 1,000 SNPs;
+or choosing windows of equal length in base pairs rather than SNPs.
+Similarly, the results were not substantially changed
+when using weighted PCA to downweight the large group of Tunisian samples.
+
+\begin{figure}
+    \begin{center}
+       \includegraphics{Fig6_Together_MDS_plot_chr3_final}
+       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
+    \end{center}
+    \caption{
+       MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3,
+       with corresponding PCA plots.
+       Each point in the plot represents a window;
+       the structure revealed by the MDS plot is strongly clustered along the chromosome,
+       with windows in the upper-right corner of the MDS plot (colored red) clustered around the centromere,
+       windows in the upper-left corner (purple) furthest from the centromere,
+       and the remaining corner (green) intermediate.
+       Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:mds_medicago_allchr}.
+       \textbf{(below)}
+        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in \Figure \ref{fig:mds12_medicago}. 
+        Each point corresponds to a sample, colored by country of origin.
+        Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:pca_peaks_medicago_allchr}.
+       \label{fig:mds12_medicago}
+       \label{fig:pca_medicago}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics{Fig5_MDS_and_gene_count_allchr_update_correction}
+    \end{center}
+    \caption{
+         MDS coordinate and gene density for each window in the \textit{Medicago} genome,
+         for chromosomes 1--8 (numbered above each pair of figures).
+         For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
+         The black plot below is gene count for each window against the position of each window.
+         \label{fig:mds_medicago}
+    }
+\end{figure}
+
+%%%%%%% %%%%%%%%%
+\section{Discussion}
+
+Our investigations have found substantial variation in the patterns of relatedness formed by population structure across the genomes
+of three diverse species,
+revealing distinct biological processes driving this variation in each species.
+More investigation, particularly on more species and datasets, will help to uncover what aspects of species history can explain these differences.
+With growing appreciation of the heterogeneous effects of selection across the genome,
+especially the importance of adaptive introgression and hybrid speciation \citep{pool2015natural,brandvain2014speciation,hufford2013genomic,fitzpatrick2010rapid,staubach2012genome},
+local adaptation \citep{lenormand2002limits,wang2014isolation},
+and inversion polymorphisms \citep{kirkpatrick2015chromosome,kirkpatrick2010chromosome},
+local PCA may prove to be a useful exploratory tool to discover important genomic features.
+% It is unclear whether the technique will be useful on reduced representation genotyping datasets 
+% due to marker density and issues with missing data --
+% our investigations with one such dataset were inconclusive --
+% but even low coverage, whole-genome sequence is very promising.
+
+We now discuss possible implications of this variation in the effects of population structure,
+the impact of various parameter choices in implementing the method,
+and possible additional applications.
+
+\paragraph{Chromosomal inversions}
+A major driver of variation in patterns of relatedness in two datasets we examined are inversions.
+This may be common,
+but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
+PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
+however, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
+or simply regions of reduced recombination
+without additional lines of evidence.
+Inversions show up in our method because across the inverted region,
+most gene trees share a common split that dates back to the origin of the inversion.
+However, in many applications, inversions are a nuisance.
+For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
+by regressing out the effect of linked SNPs on each other.
+Removing samples with the less common orientation of each inversion reduced,
+but did not eliminate, the signal of inversions
+seen in the \textit{Drosophila melanogaster} dataset,
+demonstrating that the genomic effects of transiently polymorphic inversions
+may outlast the inversions themselves.
+
+\paragraph{The effect of selection}
+It seems that
+the variation in patterns of relatedness we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
+must be explained somehow by linked selection.
+Furthermore, 
+the selection must be affecting many targets across the genome,
+since we see similar effects across long distances (even distinct chromosomes).
+For this reason, the most likely candidate may be selection against linked deleterious mutations, known as ``background selection'' \citep{charlesworth1993effect,charlesworth2013background}.
+Informally, background selection reduces the number of potential contributors to the gene pool 
+in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious};
+for this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently,
+Therefore, regions of the genome harboring many targets of local adaptation may show similar patterns,
+since migrant alleles in these regions will be selected against,
+and so locally gene trees will more closely reflect spatial proximity.
+
+A related possibility is that variation in patterns of relatedness is due to recent admixture between previously separated populations,
+the effects of which were not uniform across the genome due to selection.
+For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
+is due to selection against Neanderthal genes, leading to greater introgression in regions of lower gene density
+\citep{harris2016genetic,juric2016strength}.
+African \textit{Drosophila melanogaster} are known to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
+if selection regularly favors genes from one origin,
+this could lead to substantial variation in patterns of relatedness correlated with local gene density.
+
+There has been substantial debate over the relative impacts of different forms of selection.
+These have been difficult to disentangle in part because for the most part
+theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
+and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
+It may be possible to design more powerful statistics that make stronger use of spatial information.
+
+% How do we expect selection to affect population structure?
+% PCA summarizes patterns in kinship found in the genetic covariance matrix,
+% which is an average across locus-specific genealogies:
+% if individuals are closer in the genealogies of a given genomic region, 
+% they tend to be closer in the PCA maps for that region. 
+% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
+% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
+% Selective sweeps cause local recent ancestry or short trees.
+% Background selection causes shallow ones.
+% Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
+% characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
+% A genomic region with many targets for selection experiences
+% background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
+% which would tend to shorten genealogical trees in the region,
+% similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
+% % Balancing selection causes deep trees.
+% On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
+% and population structure locally describes which individuals have which alleles
+% rather than geographical proximity.
+% Finally, 
+% since recombination is suppressed between opposite orientations of a chromosomal inversion
+% near its breakpoints,
+% genealogies in these regions separate samples carrying the two orientations of the inversion.
+% In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
+% (indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
+% as a way to identify inversions).
+% if a chromosomal inversion is polymorphic in the sample, 
+% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints.
+Recombination suppression across inversions thus results in different genome structure and population structure.
+% Other effects, like noise, introgression might also influence population structure.
+% 
+% Many of the effects listed above (such as single selective sweeps or inversions) 
+% are not expected to have similar effects on population structure in different regions of the genome
+% because of randomness in which samples end up in which group;
+% but if these coincide with a region of reduced recombination (or an inversion),
+% these could drive major patterns of variation.
+% The more subtle effects of genome-wide linked selection could be shared across large regions
+% due to large-scale variation in gene density
+% (as by background selection or local adaptation at many genes).
+
+\paragraph{Parameter choices}
+There are several choices in the method that may in principle affect the results.
+As with whole-genome PCA,
+the choice of samples is important,
+as variation not strongly represented in the sample will not be discovered.
+The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
+but downweighting may be a better option that does not discard data
+(and here we present a method to do this).
+Next, the choice of window size may be important,
+although in our applications results were not sensitive to this,
+indicating that we can see variation on a sufficiently fine scale.
+Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in \Figure \ref{fig:diagram}),
+along with the method used to discover common structure,
+will affect results.
+We used MDS, applied to either each chromosome separately or to the entire genome;
+for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosome,
+but genome-wide, their signal is obscured by the numerous other signals of comparable strength.
+
+Besides window length, there is also the question of how to choose windows.
+In these applications we have used nonoverlapping windows with equal numbers of polymorphic sites.
+Alternatively, windows could be chosen to have equal length in genetic distance,
+so that each would have roughly the same number of independent trees.
+However, we found little change in results when using different window sizes
+or when measuring windows in physical distance (in bp).
+
+% More generally, there are many possible methods to discover common structure in different parts of the genome:
+% in this work we use PCA to visualize the effects of population structure,
+% but other methods, such as STRUCTURE \citep{falush2003inference},
+% SPA \citep{yang2012modelbased},
+% SpaceMix \citep{bradburd2015spatial},
+% or other matrix factorization methods \citep{engelhardt2010analysis}
+% may highlight different sources of variation in the data.
+% It is also possible that other methods for measuring dissimilarity between windows' covariance matrices
+% or for summarizing the matrix of pairwise distances between windows
+% would lead to different insights.
+
+Finally, our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
+and how many MDS coordinates to use when describing the distance matrix between windows,
+but in our exploration, changing these has not produced dramatically different results.
+These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
+we encourage the user to experiment.
+
+\paragraph{Applications}
+So-called cryptic relatedness between samples
+has been one of the major sources of confounding in genome-wide association studies (GWAS)
+and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
+Since the effects of population structure is not constant along the genome,
+this could in principle lead to an inflation of false positives in parts of the genome
+with stronger population structure than the genome-wide average.
+A method such as ours might be used to provide a more sensitive correction.
+Fortunately, in our human dataset this does not seem likely to have a strong effect:
+most variation is due to small, independent regions, possibly primarily inversions,
+and so may not have a major effect on GWAS.
+In the other species we examined, particularly \textit{Drosophila melanogaster},
+treating population structure as a single quantity would entail a substantial loss of power,
+and could potentially be misleading.
 
@@
     \caption{
-        PCA plots for all 22 huan autosomes from the POPRES data.
+        PCA plots for all 22 human autosomes from the POPRES data.
         \label{sfig:pca_human_allchr}
@@
         Recombination rate, and the effects of population structure for \textit{Drosophila melanogaster}:
-        this shows the first MDS coordinate and recombination rate (in cM/Mbp), as in Supplementary \Figure \ref{fig:drosophila_recomb_rate},
+        this shows the first MDS coordinate and recombination rate (in cM/Mbp), as in \Figure \ref{fig:drosophila_recomb_rate},
         against each other.
-        Since the windows underlying estimates of Supplementary \Figure \ref{fig:drosophila_recomb_rate} do not coincide,
+        Since the windows underlying estimates of \Figure \ref{fig:drosophila_recomb_rate} do not coincide,
         to obtain correlations we divided the genome into 100Kbp bins, 

### 2016-11-21 writeup/local_pca_paper.tex
added caption to supplementary figures

@@
     \caption{
-         MDS plots for human chromosome 1-8.
+         MDS plots for human chromosomes 1-8.
          The first column shows the MDS visualization of relationships between windows, 
@@
     \caption{
-        MDS plots for human chromosome 9-16.
+        MDS plots for human chromosomes 9-16, as in Supplemental Figure \ref{sfig:mds12_chr1_8_human}.
         \label{sfig:mds12_chr9_16_human}
@@
     \caption{
-        MDS plots for human chromosome 17-22.
+        MDS plots for human chromosomes 17-22, as in Supplemental Figure \ref{sfig:mds12_chr1_8_human}.
         \label{sfig:mds12_chr17_22_human}
@@
     \caption{
-        PCA plots comparison for outlier windows and non-outlier windows.
+        Comparison of PCA figures within outlying windows (center column)
+        and flanking non-outlying windows (left and right columns)
+        for the two windows having outlying MDS scores on chromosome 8.
         \label{sfig:pca_chr8_outliers}

### 2016-11-21 writeup/submission/cover-letter.tex
removed spurious comment in cover letter

--- 2016-09-07 writeup/submission/cover-letter.tex
@@
 \noheadline
-\signature{Peter Ralph and Han Li}
+\signature{Han Li and Peter Ralph}
 
 \namefrom{}
-\addrfrom{Peter Ralph \\ Fenton Hall \\ University of Oregon \\ Eugene, OR 97403-1222 USA \\ \textsc{phone:} (541) 346-5530 \\ \textsc{fax:} (541) 346-0987 \\ \href{mailto:plr@uoregon.edu}{plr@uoregon.edu}}
+\addrfrom{Han Li \\ Ray R Irani Hall \\ University of Southern California \\ Los Angeles, CA 90089 USA \\ \textsc{phone:} (541) 346-5530 \\ \textsc{fax:} (541) 346-0987 \\ \href{mailto:hli465@usc.edu}{hli465@usc.edu}}
 \phonefrom{(541) 346-5530}
 \faxfrom{(541) 346-0987}
-\emailfrom{plr@uoregon.edu}
+\emailfrom{hli465@usc.edu}
 
@@
 ``Local PCA Shows How the Effect of Population Structure Differs Along the Genome'',
-for consideration in \textit{JOURNAL XXX}.
+for consideration in \textit{Genetics}.
 In this paper, we set out to describe how background patterns of relatedness 
@@
 that are outliers for particular summaries of relatedness (e.g., $F_{ST}$-based scans for selection).
-We think the paper will be of interest to readers of \textit{JOURNAL XXX}
+We think the paper will be of interest to readers of \textit{Genetics}
 because the method fills an important gap in the genomicist's toolkit,
@@
 or Andy Clark (Cornell, \href{mailto:ac347@cornell.edu}{ac347@cornell.edu}).
+
+% If the reviewers would like to read a version of the paper that is single-spaced and has figures included, 
+% they can download it from \url{http://biorxiv.org/content/early/2016/08/21/070615}.
 

### 2016-12-15 medicago/medicago_data_setup.Rmd
minor notes

@@
 ```
+
+## Weights
+
+To check the effect of subsampling,
+here's a way to make a file of weights:
+```{r make_weights}
+samples$nsamples <- table(samples$Country.of.Origin)[samples$Country.of.Origin]
+samples$weight <- 1/pmax(10,samples$nsamples)
+
+bcf.samples <- vcf_samples("data/chr1-filtered-set-2014Apr15.bcf")
+bcf.samples[bcf.samples=="HM020-I"] <- "HM020"
+
+weights <- data.frame( ID=bcf.samples, weight=samples$weight[match(bcf.samples,samples$ID)] )
+
+write.table(weights, sep="\t", file="data/inverse-samplesize-weights.tsv", row.names=FALSE)
+```
+
+## Gene location information
+
+Non-TE related gene data downloaded from [jcvi.org's JBrowse](http://jcvi.org/medicago/jbrowse/?data=data%2Fjson%2Fmedicago&loc=chr1%3A9010377..51403300&tracks=gene_models&highlight=) on 4 September 2015.

### 2017-01-12 medicago/medicago_data_setup.Rmd
renamed package/ directory lostruct/ and changed load_all to library()

@@
 ```{r setup}
-devtools::load_all("../package")
+library(lostruct)
 library(Matrix)

### 2017-01-12 medicago/run_on_medicago.R

@@ -51,7 +51,7 @@ dir.create( opt$outdir, showWarnings=FALSE, recursive=TRUE )
 cat( jsonlite::toJSON( opt, pretty=TRUE ), file=file.path( opt$outdir, "config.json" ) )
 
 # setup
-devtools::load_all("../package")
+library(lostruct)
 
 # VCF files
 chroms <- paste0("chr",1:8)

### 2017-01-12 popres/popres_example.R

@@ -1,5 +1,4 @@
-library(devtools)
-load_all("../package")
+library(lostruct)
 
 # as in POPRES_SNPdata_recode12.R
 chr22 <- read_tped("POPRES_Genotypes_QC2_v2_TXT.tped.gz", chrom=22)

### 2017-01-14 sims/neutral/neutral-sim.py
neutral simulation script

@@ -1,0 +1,132 @@
+#!/usr/bin/python3.5
+description = '''
+Simulate.
+'''
+
+import gzip
+import sys
+from optparse import OptionParser
+import math
+import time
+
+import msprime
+
+def fileopt(fname,opts):
+    '''Return the file referred to by fname, open with options opts;
+    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
+    '''
+    if fname == "-":
+        if opts == "r":
+            fobj = sys.stdin
+        elif opts == "w":
+            fobj = sys.stdout
+        else:
+            print("Something not right here.")
+    elif fname[len(fname)-3:len(fname)]==".gz":
+        fobj = gzip.open(fname,opts)
+    else:
+        fobj = open(fname,opts)
+    return fobj
+
+parser = OptionParser(description=description)
+parser.add_option("-k","--nsamples",dest="nsamples",help="number of samples, total")
+parser.add_option("-N","--popsize",dest="popsize",help="size of each subpopulation",default=100)
+parser.add_option("-w","--width",dest="width",help="width of square grid, in populations",default=3)
+parser.add_option("-L","--length",dest="length",help="number of bp in the chromosome",default=1e4)
+parser.add_option("-m","--migr",dest="migr",help="migration proportion between adjacent populations",default=.01)
+parser.add_option("-r","--min_recomb",dest="min_recomb",help="minimum recombination rate",default=2.5e-8)
+parser.add_option("-R","--max_recomb",dest="max_recomb",help="maximum recombination rate",default=2.5e-8)
+parser.add_option("-p","--mapfile",dest="mapfile",help="name of map file")
+parser.add_option("-u","--mut_rate",dest="mut_rate",help="mutation rate",default=1e-7)
+parser.add_option("-o","--outfile",dest="outfile",help="name of output file (or '-' for stdout)",default="-")
+parser.add_option("-g","--logfile",dest="logfile",help="name of log file (or '-' for stdout)",default="-")
+(options,args) =  parser.parse_args()
+
+outfile = fileopt(options.outfile, "w")
+logfile = fileopt(options.logfile, "w")
+
+logfile.write("Options:\n")
+logfile.write(str(options)+"\n")
+
+popsize = float(options.popsize)
+width = int(options.width)
+nsamples = int(options.nsamples)
+length = float(options.length)
+migr = float(options.migr)
+mut_rate = float(options.mut_rate)
[…]

### 2017-01-19 sims/background/background-sim.py
background sims seem to work

@@ -11,6 +11,7 @@ import time
 import simuOpt
 simuOpt.setOptions(alleleType='mutant')
 import simuPOP as sim
+from simuPOP.demography import migr2DSteppingStoneRates
 import random
 from itertools import accumulate
 
@@ -44,10 +45,12 @@ parser.add_option("-a","--gamma_alpha",dest="gamma_alpha",help="alpha parameter 
 parser.add_option("-b","--gamma_beta",dest="gamma_beta",help="beta parameter in gamma distributed selection coefficients",default=.001)
 parser.add_option("-o","--outfile",dest="outfile",help="name of output file (or '-' for stdout)",default="-")
 parser.add_option("-g","--logfile",dest="logfile",help="name of log file (or '-' for stdout)",default="-")
+parser.add_option("-s","--selloci_file",dest="selloci_file",help="name of file to output selected locus information",default="sel_loci.txt")
 (options,args) =  parser.parse_args()
 
 outfile = fileopt(options.outfile, "w")
 logfile = fileopt(options.logfile, "w")
+selloci_file = options.selloci_file
 
 logfile.write("Options:\n")
 logfile.write(str(options)+"\n")
@@ -61,6 +64,7 @@ beta=float(options.gamma_beta)
 length=float(options.length)
 recomb_rate=float(options.recomb_rate)
 mut_rate=float(options.mut_rate)
+migr=float(options.migr)
 
 if length < nloci:
     raise ValueError("nloci cannot be larger than length")
@@ -115,7 +119,7 @@ pop = sim.Population(
         size=[popsize]*npops, 
         loci=[nloci], 
         lociPos=locus_position,
-        infoFields=['ind_id','fitness'])
+        infoFields=['ind_id','fitness','migrate_to'])
 
 pop.evolve(
     initOps=[
@@ -123,9 +127,13 @@ pop.evolve(
         sim.IdTagger(),
     ]+init_geno,
     preOps=[
+        sim.Migrator(
+            rate=migr2DSteppingStoneRates(
+                migr, m=width, n=width, diagonal=False, circular=False),
+            mode=sim.BY_PROBABILITY),
         sim.AcgtMutator(rate=[mut_rate], model='JC69'),
         sim.PyMlSelector(GammaDistributedFitness(alpha, beta),
-            output='>>sel_loci.txt'),
+            output='>>'+selloci_file),
     ],
     matingScheme=sim.RandomMating(
         ops=[

### 2017-01-19 sims/background/recombs-to-msprime.py

@@ -3,10 +3,6 @@ description = '''
 Convert simulations to msprime and VCF.
 '''
 
-# TO DO:
-#  - record locus locations and use this
-#  - put the simulation on a grid
-
 import gzip
 import sys
 from optparse import OptionParser
@@ -38,6 +34,7 @@ parser.add_option("-i","--infile",dest="infile",help="input file (output by back
 parser.add_option("-k","--nsamples",dest="nsamples",help="number of *diploid* samples, total")
 parser.add_option("-u","--mut_rate",dest="mut_rate",help="mutation rate",default=1e-7)
 parser.add_option("-o","--outfile",dest="outfile",help="name of output file (or '-' for stdout)",default="-")
+parser.add_option("-t","--treefile",dest="treefile",help="name of output file for trees (or '-' for stdout)")
 parser.add_option("-g","--logfile",dest="logfile",help="name of log file (or '-' for stdout)",default="-")
 (options,args) =  parser.parse_args()
 
@@ -116,16 +113,16 @@ while True:
             child,parent,ploid,*rec = [int(x) for x in line.split()]
             if child != prev_child:
                 raise ValueError("Recombs not in pairs:"+str(child)+"=="+str(prev_child))
-        print(child,child_p,parent,ploid)
+        # print(child,child_p,parent,ploid)
         if ind_to_time(child)>=ind_to_time(parent):
             raise ValueError(str(child)+" at "+str(ind_to_time(child))+" does not come after " + str(parent)+" at "+str(ind_to_time(parent)))
         start=0.0
         child_chrom=i2c(child,child_p)
-        print(".. Adding",child_chrom)
+        # print(".. Adding",child_chrom)
         args.add_individual(child_chrom,ind_to_time(child))
         for r in rec:
             breakpoint=random.uniform(locus_position[r],locus_position[r+1])
-            print("--- ",start,breakpoint)
+            # print("--- ",start,breakpoint)
             args.add_record(
                     left=start,
                     right=breakpoint,
@@ -133,7 +130,7 @@ while True:
                     children=(child_chrom,))
             start=breakpoint
             ploid=((ploid+1)%2)
-        print("--- ",start,length," |")
+        # print("--- ",start,length," |")
         args.add_record(
                 left=start,
                 right=length,
@@ -147,18 +144,16 @@ while True:
[…]

### 2017-01-25 lostruct/R/pc_dist.R
added sqrt to distance computation

@@ -8,7 +8,7 @@
 #' \code{values} is replaced by \code{values/sqrt(sum(values))}.
 #'
 #' The norm is also weighted by the weights \code{w}:
-#'   dist(A,B) = sum_{i,j}  w[i] * w[j] * ( A[i,j] - B[i,j] )^2
+#'   dist(A,B) = sqrt( sum_{i,j}  w[i] * w[j] * ( A[i,j] - B[i,j] )^2 )
 #'
 #' @param x Matrix as output by eigen_windows(): one row per window, with entries (sum squares), (eigenvalues), (eigenvectors).
 #' @param npc Number of pcas computed.
@@ -28,11 +28,11 @@ pc_dist <- function( x, npc=attr(x,"npc"), w=1, normalize="L1", mc.cores=1 ) {
     emat <- function (u) { matrix(u,ncol=npc) }
     out <- do.call( rbind, this.lapply( 1:n, function (i) {
                 sapply( 1:n, function (j) {
-                    dist_from_pcs( values[i,], emat(vectors[i,]), values[j,], emat(vectors[j,]) )
+                    dist_sq_from_pcs( values[i,], emat(vectors[i,]), values[j,], emat(vectors[j,]) )
                 } )
             } ) )
-    # symmetrize
-    return( (out + t(out))/2 )
+    # symmetrize and truncate negative nubmers (which are machine error)
+    return( sqrt(pmax(0,(out + t(out))/2)) )
 }
 
 
@@ -51,7 +51,7 @@ pc_dist <- function( x, npc=attr(x,"npc"), w=1, normalize="L1", mc.cores=1 ) {
 #' which means premultiplying U and V by sqrt(w); assume this has been done already.
 #'
 #' The code does not check that vectors1 and vectors2 are each orthonormal.
-dist_from_pcs <- function (values1,vectors1,values2,vectors2) {
+dist_sq_from_pcs <- function (values1,vectors1,values2,vectors2) {
     # this is X^T
     Xt <- crossprod(vectors2,vectors1)
     return( sum(values1^2) + sum(values2^2) - 2 * sum( values1 * colSums(Xt*(values2*Xt)) ) )

### 2017-01-25 lostruct/tests/testthat/test_pc_dist.R

@@ -1,6 +1,10 @@
 
 context("Testing pc_dist")
 
+expect_almost_equal <- function (object,expected,tol=10*sqrt(.Machine$double.eps),...) {
+    expect_true( all(abs(object-expected)<tol), ... )
+}
+
 set.seed(42)
 n <- 10
 
@@ -34,17 +38,17 @@ g2 <- function (k,eig) {
 kk <- c(1,5,10)
 for (k in kk) {
     # non-normalized
-    ff <- sum( (f(k,x.eig) - f(k,y.eig))^2 )
+    ff <- sqrt(sum( (f(k,x.eig) - f(k,y.eig))^2 ))
     fm <- matrix(c(0,ff,ff,0),nrow=2)
-    expect_equal( fm, pc_dist(elist(k),npc=k,normalize=FALSE) )
+    expect_almost_equal( fm, pc_dist(elist(k),npc=k,normalize=FALSE) )
     # normalized, L1
-    gg <- sum( (g1(k,x.eig) - g1(k,y.eig))^2 )
+    gg <- sqrt(sum( (g1(k,x.eig) - g1(k,y.eig))^2 ))
     gm <- matrix(c(0,gg,gg,0),nrow=2)
-    expect_equal( gm, pc_dist(elist(k),npc=k,normalize="L1") )
+    expect_almost_equal( gm, pc_dist(elist(k),npc=k,normalize="L1") )
     # normalized, L2
-    gg <- sum( (g2(k,x.eig) - g2(k,y.eig))^2 )
+    gg <- sqrt(sum( (g2(k,x.eig) - g2(k,y.eig))^2 ))
     gm <- matrix(c(0,gg,gg,0),nrow=2)
-    expect_equal( gm, pc_dist(elist(k),npc=k,normalize="L2") )
+    expect_almost_equal( gm, pc_dist(elist(k),npc=k,normalize="L2") )
 }
 
 ## weighted
@@ -68,16 +72,16 @@ expect_equal( sum( w * (y^2 %*% diag(w)) ), sum( y.eig$values^2 ) )
 kk <- c(1,5,10)
 for (k in kk) {
     # non-normalized
-    ff <- sum(  w * sweep( (f(k,x.eig) - f(k,y.eig))^2, 2, w, "*" ) )
+    ff <- sqrt( sum(  w * sweep( (f(k,x.eig) - f(k,y.eig))^2, 2, w, "*" ) ) )
[…]

### 2017-02-02 templated/summarize_run.Rmd
doc for templated report

--- 2016-05-28 medicago/summarize_run.Rmd
@@
 ```
-templater::render_template("summarize_run.Rmd",output="lostruct/results_type_snp_size_10000_jobid_324902/run_summary.html",change.rootdir=TRUE)
+templater::render_template("summarize_run.Rmd",output="lostruct_results/type_snp_size_10000_jobid_324902/run_summary.html",change.rootdir=TRUE)
 ```
@@
 # original data files
-chroms <- paste0("chr",1:8)
-bcf.files <- file.path( "..", "..", "data", paste0(chroms,"-filtered-set-2014Apr15.bcf") )
+chroms <- opt$bcf_file_names
+bcf.files <- opt$bcf_files
 names(bcf.files) <- chroms
+
+sample.ids <- vcf_samples(bcf.files[1])
+samp.file <- opt$sample_info
+samps <- read.table(samp.file,sep="\t",header=TRUE)
+samps <- droplevels( samps[match(sample.ids,samps$ID),] )
 
@@
 
-First, here are the first `r corner.npc` PCs for each corner, together:
+Here is the color scheme:
+```{r pop_colors, fig.height=2*fig.dim}
+pop.names <- levels(samps$population)
+pop.cols <- rainbow_hcl(nlevels(samps$population))
+pop.pch <- seq_len(nlevels(samps$population))
+plot( rep(1,length(pop.names)), seq_along(pop.names), pch=pop.pch, col=pop.cols, xlim=c(0,length(pop.names)),
+       xlab='', ylab='', xaxt='n', yaxt='n' )
+text( rep(1,length(pop.names)), seq_along(pop.names), labels=pop.names, pos=4 )
+```
+
+Here are all pairwise plots of the first `r corner.npc` PCs for each of the three corners:
 ```{r plot_corner_pca, fig.width=3*fig.dim}
-for (k in seq_along(mds.corners)) {
-    vectors <- matrix( corner.pca[[k]][-(1:(1+corner.npc))], ncol=corner.npc )
-    colnames(vectors) <- paste("PC", 1:corner.npc)
-    pairs(vectors)
+layout(t(1:3))
+for (i in 1:(corner.npc-1)) {
+    for (j in (i+1):corner.npc) {
+        for (k in 1:ncol(mds.corners)) {
+            vectors <- matrix( corner.pca[[k]][-(1:(1+corner.npc))], ncol=corner.npc )[,c(i,j)]
+            colnames(vectors) <- paste("PC", c(i,j))
+            plot(vectors, pch=pop.pch[samps$population], 
+                    col=pop.cols[samps$population] )
+            if (i==1 && j==2) {
+                mtext(paste("corner",k),side=3,cex=2)
+            }
+        }
+    }
 }

### 2017-02-02 templated/run_lostruct.R

@@ -1,0 +1,116 @@
+#!/usr/bin/Rscript
+library(optparse)
+
+invocation <- commandArgs()
+
+usage <- "\
+\
+Does a local PCA analysis.
+To run an analysis on windows of 1000 SNPs, for instance: \
+    Rscript run_lostruct.R -t snp -s 1000 \
+while to run it on windows of 10000 bps: \
+    Rscript run_lostruct.R -t bp -s 10000 \
+\
+Do
+    Rscript run_on_medicago.R -h\
+for options.\
+\
+"
+
+option_list <- list(
+    # input/output
+        make_option( c("-i","--input_dir"),   type="character",        help="Directory with input: indexed .bcf or .vcf.bgz files."),
+        make_option( c("-I","--sample_info"),   type="character",        help="File with ID and population of the samples."),
+        make_option( c("-t","--type"),   type="character",             help="Window by SNP or by bp?"),
+        make_option( c("-s","--size"),   type="integer",               help="Size of the window, in units of type."),
+        make_option( c("-k","--npc"),   type="integer",   default=2L,  help="Number of principal components to compute for each window. [default: %default]"),
+        make_option( c("-w","--weightfile"), type="character",            help="File name containing weights to apply to PCA."),
+        make_option( c("-m","--nmds"),   type="integer",   default=2L, help="Number of principal coordinates (MDS variables) to compute. [default: %default]"),
+        make_option( c("-o","--outdir"), type="character",             help="Directory to save results to.  [default: lostruct_results/type_%type_size_%size_jobid_%jobid/]"),
+        make_option( c("-j","--jobid"),  type="character", default=formatC(1e6*runif(1),width=6,format="d",flag="0"),   help="Unique job id. [default random]")
+    )
+opt <- parse_args(OptionParser(option_list=option_list,description=usage))
+if (is.null(opt$outdir)) { opt$outdir <- file.path("lostruct_results", 
+                                   sprintf( "type_%s_size_%d_weights_%s_jobid_%s", 
+                                           opt$type, 
+                                           opt$size, 
[…]

### 2017-02-03 medicago/summarize_run.Rmd
.

--- 2016-05-28 medicago/summarize_run.Rmd
@@
 names(bcf.files) <- chroms
+
+sample.ids <- vcf_samples(bcf.files[1])
+samp.file <- file.path( "..", "..", "data", "sample_info.tsv" )
+samps <- read.table(samp.file,sep="\t",header=TRUE)
+samps <- droplevels( samps[match(gsub("-I$","",sample.ids),samps$ID),] )
+samps$population <- samps$Country.of.Origin
 

### 2017-02-03 medicago/run_on_medicago.R

@@ -28,8 +28,8 @@ option_list <- list(
         make_option( c("-j","--jobid"),  type="character", default=formatC(1e6*runif(1),width=6,format="d",flag="0"),   help="Unique job id. [default random]")
     )
 opt <- parse_args(OptionParser(option_list=option_list,description=usage))
-if (is.null(opt$outdir)) { opt$outdir <- file.path("lostruct", 
-                                   sprintf( "results_type_%s_size_%d_weights_%s_jobid_%s", 
+if (is.null(opt$outdir)) { opt$outdir <- file.path("lostruct_results", 
+                                   sprintf( "type_%s_size_%d_weights_%s_jobid_%s", 
                                            opt$type, 
                                            opt$size, 
                                            if (is.null(opt$weightfile)) { "none" } else { gsub("[.].*","",basename(opt$weightfile)) }, 

### 2017-02-03 medicago/summarize_run.Rmd
added pdf output to report

@@
 knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
+```
+```{r plot_setup, include=FALSE}
+layout_heights <- function (k,dl=0,ncol=1) {
+    # to set up layout without 'dl' lines between plots
+    # use like layout(1:5,heights=layout_heights(5))
+    if (k==1) return(1)
+    layout(matrix(seq_len(k*ncol),ncol=ncol))  # this changes par("csi")
+    ds <- dl*par("lheight")*par("csi")
+    eps=par("mai")[c(1,3)]
+    dh=(par("din")[2]-sum(eps)-(k-1)*ds)/k
+    return(c(eps[2]+dh+ds/2,rep(dh+ds,k-2),eps[1]+dh+ds/2)/par("din")[2])
+}
+pdf_copy <- function (
+                width=6,
+                height=width*knitr::opts_current$get("fig.height")/knitr::opts_current$get("fig.width"),
+                plot.id=NULL,
+                filename
+                ) {
+    if (missing(filename)) {
+        file.id <- if (is.null(plot.id)) { "" } else { paste0("_",plot.id) }
+        filename <- knitr::fig_path(paste(file.id,".pdf",sep=""))
+    }
+    cat("pdf version at:",filename)
+    dev.print( file=filename, device=pdf, 
+            width=width, height=height,
+            pointsize=10,
+            family="sans")
+}
+do.pdfs <- TRUE
 ```
@@
 pairs( mds.coords[,mds.cols], pch=20, col=adjustcolor(chrom.cols,0.75) )
+if (do.pdfs) { pdf_copy() }
 ```
@@
 }
+if (do.pdfs) { pdf_copy() }
 ```
@@
 # plot corners and MDS along the chromosome
+spacing <- 1
+opar <- par(mar=c(4,4,2,1)+.1,mgp=c(2.5,0.8,0))
 layout(matrix(c(rep(1,length(mds.cols)),1+seq_along(mds.cols)),ncol=2),
-        widths=c(1,2), heights=c(rep(1,length(mds.cols)-1),1.25))
+        widths=c(1,2), heights=layout_heights(length(mds.cols),dl=spacing,ncol=2))
 plot( mds.coords[,mds.cols[1:2]], pch=cpch, 
@@
      xlab="MDS coordinate 1", ylab="MDS coordinate 2" )
-opar <- par(mar=c(par("mar"),1.1)[c(5,2,5,4)])
+opar2 <- par(mar=c(par("mar"),spacing/2)[c(5,2,3,4)])
 for (k in mds.cols) {
     lastone <- (k==mds.cols[length(mds.cols)])
-    if (lastone) {
-        par(mar=c(opar$mar,0.2)[c(1,2,5,4)])
-    }
+    if (lastone) { par(mar=c(par("mar"),opar2$mar[1])[c(5,2,3,4)]) }
     chrom.plot( mds.coords[,k], pch=20, 
@@
             col=adjustcolor(ccols,0.75) )
+    # do this for all but first
+    par(mar=c(par("mar"),spacing/2)[c(1,2,5,4)])
 }
 par(opar)
+if (do.pdfs) { pdf_copy() }
 ```
@@
         }
+        if (do.pdfs) { pdf_copy(plot.id=paste(i,j,sep="_")) }
     }

### 2017-02-03 templated/summarize_run.Rmd
added pdf output to report

@@
 knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
+do.pdfs <- FALSE
+```
+```{r plot_setup, include=FALSE}
+layout_heights <- function (k,dl=0,ncol=1) {
+    # to set up layout without 'dl' lines between plots
+    # use like layout(1:5,heights=layout_heights(5))
+    if (k==1) return(1)
+    layout(matrix(seq_len(k*ncol),ncol=ncol))  # this changes par("csi")
+    ds <- dl*par("lheight")*par("csi")
+    eps=par("mai")[c(1,3)]
+    dh=(par("din")[2]-sum(eps)-(k-1)*ds)/k
+    return(c(eps[2]+dh+ds/2,rep(dh+ds,k-2),eps[1]+dh+ds/2)/par("din")[2])
+}
+pdf_copy <- function (
+                width=6,
+                height=width*knitr::opts_current$get("fig.height")/knitr::opts_current$get("fig.width"),
+                plot.id=NULL,
+                filename
+                ) {
+    if (missing(filename)) {
+        file.id <- if (is.null(plot.id)) { "" } else { paste0("_",plot.id) }
+        filename <- knitr::fig_path(paste(file.id,".pdf",sep=""))
+    }
+    cat("pdf version at:",filename)
+    dev.print( file=filename, device=pdf, 
+            width=width, height=height,
+            pointsize=10,
+            family="sans")
+}
 ```
@@
     plot(0, type='n', xlim=range(chrom.offsets/1e6), ylim=range(y,finite=TRUE), 
-         xlab='', xaxt='n', ylab=ylab, main=main)
+         xlab='', xaxt='n', yaxt='n', ylab=ylab, main=main)
     for (k in 1:floor(length(chroms)/2)) {

### 2017-02-05 writeup/review-responses.tex
added response to reviewers to latex

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Genetics}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Han Li \\
    \emph{and} Peter Ralph \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are pleased to submit a revision of our manuscript,

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
Han Li and
Peter Ralph
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
    Please understand that incremental changes will not be sufficient.  Adding
    simulations to strengthen key claims will be necessary, particularly addressing
    the impacts of mutation rate and recombination rate variation with more depth,
    the concern regarding PC switching (Reviewer 1), and the concern regarding the
    impacts of variation in missingness by sub-population (Reviewer 2).
\end{quote}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    The paper is generally well written and clear; it addresses an important
    problem, and clearly makes some progress on it. However, it suffers from having
    no grounding in either theory or empirical demonstration that it really can find
    the structures that are claimed. I find the arguments that it finds inversions
    compelling, though not watertight, and I am not yet convinced that it is finding
    ubiquitous background selection.  To make this claim, significant extra work is
    required.
\end{quote}

\begin{quote}
    In short, the approach is interesting but not sufficiently explored to produce
    compelling evidence for the implications that are claimed.  Putting a large
    amount of effort into simulations may alleviate these concerns somewhat.

    Specific points: What does this method find? I'm concerned about:
    (a) variation in the recombination rate
    and (b) variation in the mutation rate, creating spurious structure.

    The first possibility is that massively varying information quantity
    within windows could lead to a small number of such windows having their
    orientation reversed: that is, PC1 becomes PC2 and vice versa. (Or PC2 and PC3
    could switch). This would lead to such windows having unusual properties and
    hence appearing as evidence of an inversion.

    I do agree with the authors that significant outliers would be found at
    inversions. However, even if the PC switching does not occur, or the model could
    handle it, the evidence for selection is weaker.  If the two types of variation
    described above exist, with no selection, I would still expect a ``continuous
    triangle'' of results (as seen left of Fig 2, top left of Fig 6) with extrema
    described by windows with the most information, and points placed at different
    extremum having low recombination rate (because by chance, these will get an
    approximately fixed local tree, corresponding on average to the genome-wide
    population structure).

    Addressing this is likely quite hard, though the authors may be able to think of
    something that separates these effects from selection.
\end{quote}

\begin{point}{}
\ldots variation in the recombination rate \ldots 
\end{point}

\begin{point}{}
\ldots and variation in the mutation rate, creating spurious structure.
\end{point}

\begin{point}{PC switching}
\ldots could lead to a small number of such windows having their
orientation reversed: that is, PC1 becomes PC2 and vice versa. (Or PC2 and PC3
could switch).
\end{point}

\begin{point}{p6}
``here, we use k=2...''  - you have to show that $k>2$ is the same.
\end{point}

\begin{point}{p15}
``We also found nearly identical results when choosing shorter windows of 1,000 SNPs'' - again, show this.
\end{point}

\begin{point}{p15}
 ``or choosing windows of equal length in base pairs rather than SNPs'' - once again.
\end{point}

\begin{point}{}
Using 2 PCs is common practice: only if this is the end of an analysis and the
PCA was done for visualisation. Here you are using it for something so should
keep all the relavant PCs.
\end{point}

\begin{point}{}
I'm surprised that PCAdmix isn't referenced. It is using a very similar
method, albeit with different goals. In particular, the approach of placing all
points into a single, genome-wide PC space solves many of the problems that this
approach has (though I agree there may be benefits to the approach described here)
\end{point}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    This is an interesting and well written paper. It was a pleasant read. I have three main general
    comments:
\end{quote}

\begin{point}{Related work:} 
The authors provide an introduction of the main concepts, as well as some
intuition of what the method is doing and how, but I found comparison to previous approaches
to be somewhat missing. To some extent, this is due to the fact that the main goal of their
analysis is somewhat vaguely ``finding heterogeneity'', which leads to the applications of
detecting chromosomal inversions and evidence for background selection. It would help to
have a well defined set of hypotheses, test the method’s accuracy  using simulation (see next
comment), and compare to previous efforts in similar domains.
\end{point}

\begin{point}{Validation:}
In several occasions, the authors seem to introduce a potential problem in their
approach, and provide a solution to it. This is generally rather intuitive, but it would really help
to have simulations of some sort to show that the issue arises and leads to a problem, and that
their approach does address the specific problem.
\end{point}

\begin{point}{}
The use of weighted PCA to cope with unbalanced sample size could be better demonstrated.
Although the current explanation makes intuitive sense, this approach does not seem to be
used in previous work. The authors could design a simulation that supports their approach.
\end{point}

\begin{point}{}
It is conceivable that some subpopulations will have more missingness in some windows. That
may skew the resulting PCs by selecting different sample sizes for the different windows (as
discussed in Appendix B) . This could distort the PCs, so that variation reflects underlying
variation in missingness. Would be good to discuss this potential issue and provide simulations.
\end{point}

\begin{point}{Appendix A:}
when using jackknife to estimate variance, each window is being divided in 10
``independent'' resampling units. Due to LD, these 10 blocks are likely correlated, which would
bias the estimates of variance. This is probably not a problem because both signal and noise
could be equally biased, but the authors may want to consider this potential issue. I wonder if
the correlation with recombination rate may be partially explained by this.
\end{point}

\begin{point}{}
Is it possible to explain the results of Figure 6 just considering neutral variation in local
ancestry due to recent admixture? This may explain why ancestry seems to explain a fair
amount of variance in the lower plots of Fig 6. Local PCA has been previously used by others to
detect local ancestry blocks, e.g. see the PCAdmix approach by Brisbin et al. The authors
discuss the possibility that admixture is driving the differentiation, but do not test whether their
observations agree with neutrality.
\end{point}

\begin{point}{}
``to remove the effect of artifacts such as mutation rate variation, we also rescale each
approximate covariance matrix to be of similar size (precisely, so that the underlying data
matrix has trace norm equal to one''. This potential issue is a bit unclear to me, since I would
expect that scaling the volume of local trees would not result in changed distances in PC
space. Perhaps the authors could show via simulations that this creates a problem, and that the
normalization addresses it.
\end{point}

\begin{point}{Figure 7:}
are MDS coordinates correlated with recombination rates in this case?
\end{point}

\begin{point}{Application:}
 is what the authors seem to be proposing not already accounted for by linear
mixed model association approaches? If not, this should be clarified. Either way, this paragraph
could be dropped.
\end{point}

\begin{point}{Introduction:}
 ``it is not necessarily clear what aspects of demography should be included in the
concept.'' I find it a bit weird to describe selection as an ``aspect of demography''. Although it
could be seen as such within a coalescent framework, that seems to be just a useful
representation. The authors may consider rewording`.
\end{point}

\begin{point}{}
Paragraph starting in ``Since the definition...''. The notation is a bit unclear. Please check that it
is clear which PC the text refers to.
\end{point}

\begin{point}{}
Would the authors be able to provide a sense for the directionality of effects in Figure 4? It
would be interesting if the authors tried to further characterize regions that are similar due to
higher recombination rates. E.g. is there more/less density of polymorphisms in these regions?
\end{point}

\begin{point}{Page 13:}
typo: ``figures 6 and 6''.
\end{point}

\begin{point}{}
Typo in abstract, line 6 ``, We show'' -> ``. We show''.
\end{point}

\begin{point}{}
Typo: end of introduction ``an visualization''. The whole sentence is a bit weird. The authors just
stated focus is on clustering, not on looking for outliers, but what does it mean that ``we allow
ourselves to be surprised by unexpected signals in the data''?
\end{point}

\begin{point}{}
``There has been substantial debate over the relative impacts of different forms of selection.''
Citation needed.
\end{point}

\begin{point}{}
``Results using larger numbers of PCs were nearly identical''. It would be interesting to have a
supplementary table.
\end{point}

\begin{point}{}
Table 1 legend seems a bit redundant. Columns are self-explanatory.
\end{point}

\begin{point}{}
It would help to have numbered lines and references.
\end{point}

### 2017-02-05 writeup/local_pca_paper.tex
background sims seem to work
added sqrt to distance computation
added response to reviewers to latex

@@
 %\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
+\usepackage{graphicx}				% Use pdf, png, jpg, or eps with pdflatex; use eps in DVI mode
 								% TeX will automatically convert eps --> pdf in pdflatex		
@@
 \DeclareMathOperator{\tr}{tr}
+
+%% for submitting with responses to reviews
+\usepackage{lineno}
+
+% from http://tex.stackexchange.com/questions/43648/why-doesnt-lineno-number-a-paragraph-when-it-is-followed-by-an-align-equation/55297#55297
+\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+  \renewenvironment{#1}%
+     {\linenomath\csname old#1\endcsname}%
+     {\csname oldend#1\endcsname\endlinenomath}}% 
+\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+  \patchAmsMathEnvironmentForLineno{#1}%
+  \patchAmsMathEnvironmentForLineno{#1*}}%
+\AtBeginDocument{%
+\patchBothAmsMathEnvironmentsForLineno{equation}%
+\patchBothAmsMathEnvironmentsForLineno{align}%
+\patchBothAmsMathEnvironmentsForLineno{flalign}%
+\patchBothAmsMathEnvironmentsForLineno{alignat}%
+\patchBothAmsMathEnvironmentsForLineno{gather}%
+\patchBothAmsMathEnvironmentsForLineno{multline}%
+}
+
+% Responses to reviews
+\input{review-response-commands}
+% set this to show line numbers and include responses to reviews or not
+\newif\ifreviewresponses
+\reviewresponsestrue  % include them
+% \reviewresponsesfalse  % don't include them
+\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file
 
@@
     = 
+    \left(
     \frac{ \sum_{\ell=1}^k \lambda_{\ell i}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell i})^2 }
     + \frac{ \sum_{\ell=1}^k \lambda_{\ell j}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell j})^2 }
-    - 2 \sum_{\ell, m=1}^k (V(i)^T V(j))^2_{\ell m} .
+    - 2 \sum_{\ell, m=1}^k (V(i)^T V(j))^2_{\ell m} 
+    \right)^{1/2}.
 \end{align}
@@
 (especially inversion status of \textit{D.~melanogaster} samples).
+Jaime Ashander and Jerome Kelleher provided assistance in performing the simulations.
 Thanks also go to Yaniv Brandvain, Barbara Engelhardt, Charles Langley, Graham Coop, and Jeremy Berg for helpful comments
@@
 
+% reviews here if toggled above
+\includereviews
+
 \end{document}

### 2017-02-09 writeup/local_pca_paper.tex
added correlations

@@
 
+\section{Supplementary Tables}
+
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Wed Feb  8 16:17:12 2017
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrrr}
+  \hline
+ & 10000snp, 2 PCs & 1000snp, 2 PCs & 10000snp, 5 PCs & 100000bp, 2 PCs & 10000bp, 2 PCs \\ 
+  \hline
+10000snp, 2 PCs & 1.00 & 0.87 & 0.96 & 0.90 & 0.88 \\ 
+  1000snp, 2 PCs & 0.68 & 1.00 & 0.73 & 0.68 & 0.94 \\ 
+  10000snp, 5 PCs & 0.96 & 0.92 & 1.00 & 0.88 & 0.93 \\ 
+  100000bp, 2 PCs & 0.90 & 0.87 & 0.88 & 1.00 & 0.87 \\ 
+  10000bp, 2 PCs & 0.68 & 0.93 & 0.72 & 0.67 & 1.00 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+[[2]]
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Wed Feb  8 16:17:12 2017
+\begin{table}[ht]
+\centering
+    \begin{tabular}{rrrrrr}
+      \hline
+     MDS1 & 10000 SNPs, 2 PCs & 1000 SNPs, 2 PCs & 10000 SNPs, 5 PCs & 100000bp, 2 PCs & 10000bp, 2 PCs \\ 
+      \hline
+      10000 SNPs, 2 PCs & 1.00 & 0.54 & 0.93 & 0.87 & 0.56 \\ 
+      1000 SNPs, 2 PCs & 0.82 & 1.00 & 0.76 & 0.83 & 0.92 \\ 
+      10000 SNPs, 5 PCs & 0.93 & 0.50 & 1.00 & 0.83 & 0.52 \\ 
+      100000bp, 2 PCs & 0.87 & 0.59 & 0.84 & 1.00 & 0.58 \\ 
+      10000bp, 2 PCs & 0.83 & 0.92 & 0.77 & 0.84 & 1.00 \\ 
+       \hline
+    \end{tabular}
+    \begin{tabular}{rrrrrr}
+      \hline
+     MDS2 & 10000snp, 2 PCs & 1000snp, 2 PCs & 10000snp, 5 PCs & 100000bp, 2 PCs & 10000bp, 2 PCs \\ 
+      \hline
+      10000snp, 2 PCs & 1.00 & 0.54 & 0.93 & 0.87 & 0.56 \\ 
+      1000snp, 2 PCs & 0.82 & 1.00 & 0.76 & 0.83 & 0.92 \\ 
+      10000snp, 5 PCs & 0.93 & 0.50 & 1.00 & 0.83 & 0.52 \\ 
+      100000bp, 2 PCs & 0.87 & 0.59 & 0.84 & 1.00 & 0.58 \\ 
+      10000bp, 2 PCs & 0.83 & 0.92 & 0.77 & 0.84 & 1.00 \\ 
+       \hline
+    \end{tabular}
+    \begin{caption}
+        Correlations between MDS coordinates of genomic regions
+        between runs with different parameter values.
+        To produce these, 
+        we first ran the algorithm with the specified window size and number of PCs 
+        ($k$ in equation \plr{XXX})
+        on the full \textit{Medicago truncatula} dataset.
+        Then to obtain the correlation between results obtained from parameters A in the row of the matrix above
+        and parameters B in the column of the matrix above,
+        we mapped the windows of B to those of A by averaging MDS coordinates of any windows of B whose midpoints lay in the corresponding window of A;
+        we then computed the correlation between the MDS coordinates of A and the averaged MDS coordinates of B.
+        This is not a symmetric operation, so these matrices are not symmetric.
+        As expected, parameter values with smaller windows produce noisier estimates.
+        Plots of MDS values along the genome are visually nearly identical
+        for parameter sets having similar window sizes --
+        full reports are available as supplementary material on Data Dryad \plr{XXX}.
+        \label{tab:param_cors}
+    \end{caption}
+\end{table}
+
 \section{Supplementary Figures}

### 2017-02-09 medicago/compare-parameter-runs.R

@@ -67,20 +67,54 @@ compare_mds <- function (d1,d2,k) {
 }
 
 # Produces a matrix with upper triangle correlations in MDS1, and lower triangle in MDS2
-mds.cors <- matrix(NA,nrow=nrow(params),ncol=ncol(params))
-for (i in 1:(nrow(params)-1)) {
-    for (j in (i+1):nrow(params)) {
-        mds.cors[i,j] <- compare_mds(params$outdir[i],params$outdir[j],1)
-        mds.cors[j,i] <- compare_mds(params$outdir[i],params$outdir[j],2)
+mds.cors <- list( matrix(NA,nrow=nrow(params),ncol=ncol(params)) )[c(1,1)]
+for (i in 1:nrow(params)) {
+    for (j in 1:nrow(params)) {
+        mds.cors[[1]][i,j] <- compare_mds(params$outdir[i],params$outdir[j],1)
+        mds.cors[[2]][j,i] <- compare_mds(params$outdir[i],params$outdir[j],2)
     }
 }
-rownames(mds.cors) <- sprintf("window=%d%s, %d PCs", params$size, params$type, params$npc)
-mds.cors
+for (k in 1:2) {
+    colnames(mds.cors[[k]]) <- rownames(mds.cors[[k]]) <- sprintf("%d%s, %d PCs", params$size, params$type, params$npc)
+}
 
-#                             [,1]      [,2]      [,3]      [,4]      [,5]
-# window=10000snp, 2 PCs        NA 0.8701600 0.9593167 0.9024207 0.8765862
-# window=1000snp, 2 PCs  0.8195672        NA 0.7253759 0.6820970 0.9355702
-# window=10000snp, 5 PCs 0.9319838 0.5027238        NA 0.8831645 0.9284340
-# window=100000bp, 2 PCs 0.8724878 0.5896812 0.8378923        NA 0.8724009
-# window=10000bp, 2 PCs  0.8279086 0.9172536 0.7738415 0.8373688        NA
+library(xtable)
 
+options(digits=2)
+lapply(mds.cors,xtable)
+
[…]

### 2017-02-12 medicago/summarize_run.Rmd
tweaks

@@
 
+# output file patterns
+corner.covmat.file <- "corner_covmat_%d.csv"
+mds.corners.file <- "mds_corners.csv"
+
 # read in mds
@@
 mds.corners <- corners( mds.coords[,mds.cols[1:2]], prop=.05 )
+write.csv(mds.corners,file=mds.corners.file,row.names=FALSE)
 # set up colors and pchs for corners
@@
         } )
+for (k in 1:length(corner.covmats)) {
+    write.csv(corner.covmats[[k]], file=sprintf(corner.covmat.file,k),row.names=FALSE)
+}
 corner.pca <- lapply( corner.covmats, function (covmat) {

### 2017-02-12 sims/background/background-sim.py

@@ -1,4 +1,4 @@
-#!/usr/bin/env python
+#!/usr/bin/env python3.5
 description = '''
 Simulate AND write to msprime/vcf.
 '''

### 2017-02-12 sims/background/install_notes.sh

@@ -1,0 +1,9 @@
+# to (re)-install
+
+source $HOME/cmb/bin/python3_setup.sh
+
+# for msprime
+source /usr/usc/hdf5/1.8.12/setup.sh
+export PKG_CONFIG_PATH=$HOME/cmb/msprime
+
+python3 setup.py install --prefix /home/rcf-40/pralph/cmb/python3.5

### 2017-03-02 templated/summarize_run.Rmd
slight updates to template

@@
 sample.ids <- vcf_samples(bcf.files[1])
+if (!is.null(opt$sample_info)) {
     samp.file <- opt$sample_info
@@
     samps <- droplevels( samps[match(sample.ids,samps$ID),] )
+} else {
+    samps <- data.frame( 
+            ID=sample.ids,
+            population=factor(rep("pop",length(sample.ids))) )
+}
 
@@
          xlab='', xaxt='n', yaxt='n', ylab=ylab, main=main)
-    for (k in 1:floor(length(chroms)/2)) {
+    if (length(chroms)>1) for (k in 1:floor(length(chroms)/2)) {
         rect( xleft=chrom.dividers[2*k-1]/1e6, xright=chrom.dividers[2*k]/1e6, 

### 2017-03-02 sims/background/background-sim.py

@@ -32,7 +32,8 @@ def fileopt(fname,opts):
 parser = OptionParser(description=description)
 parser.add_option("-T","--generations",dest="generations",help="number of generations to run for")
 parser.add_option("-N","--popsize",dest="popsize",help="size of each subpopulation",default=100)
-parser.add_option("-w","--width",dest="width",help="width of square grid, in populations",default=3)
+parser.add_option("-w","--gridwidth",dest="gridwidth",help="width of rectangular grid, in populations",default=3)
+parser.add_option("-y","--gridheight",dest="gridheight",help="height of rectangular grid, in populations (default: equal to gridwidth)")
 parser.add_option("-L","--length",dest="length",help="number of bp in the chromosome",default=1e4)
 parser.add_option("-l","--nloci",dest="nloci",help="number of selected loci",default=20)
 parser.add_option("-m","--migr",dest="migr",help="migration proportion between adjacent populations",default=.01)
@@ -53,7 +54,7 @@ parser.add_option("-s","--selloci_file",dest="selloci_file",help="name of file t
 import simuOpt
 simuOpt.setOptions(alleleType='mutant')
 import simuPOP as sim
-from simuPOP.demography import migr2DSteppingStoneRates
+from simuPOP.demography import migr2DSteppingStoneRates, migrSteppingStoneRates
 
 if options.outfile is not None:
     outfile = fileopt(options.outfile, "w")
@@ -69,7 +70,12 @@ logfile.flush()
 generations=int(options.generations)
 popsize=int(options.popsize)
 nloci=int(options.nloci)
-width=int(options.width)
+gridwidth=int(options.gridwidth)
+if options.gridheight is not None:
+    gridheight=int(options.gridheight)
+else:
+    gridheight=gridwidth
+
 alpha=float(options.gamma_alpha)
 beta=float(options.gamma_beta)
 length=float(options.length)
@@ -80,14 +86,13 @@ migr=float(options.migr)
 nsamples=int(options.nsamples)
 ancestor_age=float(options.ancestor_age)
 
-npops=width*width
+npops=gridwidth*gridheight
 
 # increase spacing between loci as we go along the chromosome
-spacing_fac=9
 rel_positions=[0.0 for k in range(nloci)]
 for k in range(nloci):
[…]

### 2017-03-02 sims/background/run_lostruct.R

@@ -1,0 +1,1 @@
+../../templated/run_lostruct.R

### 2017-03-02 templated/run_lostruct.R

@@ -20,7 +20,7 @@ for options.\
 option_list <- list(
     # input/output
         make_option( c("-i","--input_dir"),   type="character",        help="Directory with input: indexed .bcf or .vcf.gz files. (REQUIRED)"),
-        make_option( c("-I","--sample_info"),   type="character",        help="File with ID and population of the samples. (required to run summarize_run.Rmd from the results))"),
+        make_option( c("-I","--sample_info"),   type="character",        help="File with columns labeled 'ID' and 'population' describing populations of samples in bcf file.)"),
         make_option( c("-t","--type"),   type="character",             help="Window by SNP or by bp? (REQUIRED)"),
         make_option( c("-s","--size"),   type="integer",               help="Size of the window, in units of type. (REQUIRED)"),
         make_option( c("-k","--npc"),   type="integer",   default=2L,  help="Number of principal components to compute for each window. [default: %default]"),
@@ -41,8 +41,10 @@ if (is.null(opt$input_dir) || is.null(opt$type) || is.null(opt$size)) { stop(usa
 opt$start.time <- Sys.time()
 opt$run.dir <- normalizePath(".")
 
-if (is.null(opt$sample_info) || !file.exists(opt$sample_info)) {
-    stop(sprintf("Sample info file %s does not exist - won't be able to produce report afterwards.",opt$sample_info))
+if (is.null(opt$sample_info)) {
+    warning("No sample information file - report from summarize_run.Rmd won't have colors.")
+} else if  (!file.exists(opt$sample_info)) {
+    stop(sprintf("Sample info file %s does not exist.",opt$sample_info))
 } else {
     opt$sample_info <- normalizePath(opt$sample_info)
 }

### 2017-03-08 templated/summarize_run.Rmd
added corner labeling to template

@@
     samps <- droplevels( samps[match(sample.ids,samps$ID),] )
+    samps$population <- factor(samps$population)
 } else {
@@
 # this is where to plot windows at when plotting with all chromosomes
-regions$pos <- chrom.offsets[regions$chrom]+(regions$start+regions$end)/2
+regions$pos <- chrom.offsets[match(regions$chrom,chroms)]+(regions$start+regions$end)/2
 
@@
 }
+# centroids of the corners in MDS space
+corner.mds <- do.call(rbind, lapply(1:ncol(mds.corners), 
+                function (ii){
+                    colMeans(mds.coords[mds.corners[,ii],-(1:2)])
+                } ) )
 if (do.pdfs) { pdf_copy() }
@@
      xlab="MDS coordinate 1", ylab="MDS coordinate 2" )
+points( corner.mds, pch=20, cex=5,
+        col=adjustcolor(corner.cols,0.25))
+text( corner.mds, labels=seq_len(nrow(corner.mds)), 
+        col=corner.cols, cex=2, lwd=2 )
 opar2 <- par(mar=c(par("mar"),spacing/2)[c(5,2,3,4)])
@@
 pop.pch <- seq_len(nlevels(samps$population))
-plot( rep(1,length(pop.names)), seq_along(pop.names), pch=pop.pch, col=pop.cols, xlim=c(0,length(pop.names)),
+plot( rep(1,length(pop.names)), seq_along(pop.names), 
+       cex=4, pch=pop.pch, col=pop.cols, xlim=c(0,length(pop.names)),
        xlab='', ylab='', xaxt='n', yaxt='n' )

### 2017-03-08 sims/background/neutral-split-sim.py

@@ -19,6 +19,8 @@ parser.add_argument('--chrom_length', '-L', type=float, help="Length of chromoso
 parser.add_argument('--mut_rate', '-u', default=1e-7, type=float, help="Mutation rate per bp per generation.")
 parser.add_argument('--recomb_rate', '-r', default=1e-7, type=float, help="Recombination rate per bp per generation.")
 parser.add_argument('--relative_split_time', '-T', default=0.25, type=float, help="Time since rearrangement of populations in units of Ne.")
+parser.add_argument('--relative_fast_m', '-m', default=10, type=float, help="Migration rate for 'close' pops in units of Ne.")
+parser.add_argument('--relative_slow_m', '-M', default=0.1, type=float, help="Migration rate for 'distant' pops in units of Ne.")
 
 args = parser.parse_args()
 
@@ -27,9 +29,9 @@ if not os.path.isdir(args.outdir):
 
 base_options = {
         'Ne' : 1000,
-        'm_rel' : 10,
-        'M_rel' : 0.1,
-        'T_rel' : 0.25,
+        'm_rel' : args.relative_fast_m,
+        'M_rel' : args.relative_slow_m,
+        'T_rel' : args.relative_split_time,
         'nsamples' : args.nsamples,
         'chrom_len' : args.chrom_length,
         'treefile' : "chrom{}.trees",
@@ -52,9 +54,11 @@ with open(os.path.join(args.outdir,"config.json"),"w") as configfile:
 
 with open(os.path.join(args.outdir,"samples.tsv"),"w") as samplefile:
     samplefile.write("ID\tpopulation\n")
+    indiv=0
     for pop in range(1,5):
         for k in range(base_options['nsamples']):
-            samplefile.write("{}\t{}\n".format(k,pop))
+            samplefile.write("msp_{}\tpop_{}\n".format(indiv,pop))
+            indiv+=1
 
 for chrom,opts in options.items():
 

### 2017-04-22 sims/background/threeway-background-sim.py
.

@@ -3,6 +3,10 @@ description = '''
 Simulate with simuPOP AND write to msprime/vcf:
     a population arrangement switches from A<-B|C to A|B->C 
     at some point in the past.
+
+Output in selloci is of the form
+   loc a1 a2 fitness
+meaning locus number, allele1, allele2, fitness when it is first seen.
 '''
 
 import gzip

### 2017-04-22 templated/summarize_run.Rmd
small changes

@@
     samps <- read.table(samp.file,sep="\t",header=TRUE)
-    samps <- droplevels( samps[match(sample.ids,samps$ID),] )
+    names(samps) <- tolower(names(samps))
+    samps <- droplevels( samps[match(sample.ids,samps$id),] )
     samps$population <- factor(samps$population)

### 2017-04-22 sims/background/neutral-split-sim.py

@@ -97,16 +97,26 @@ for chrom,opts in options.items():
     logfile.write(time.strftime('     %X %x %Z\n'))
     logfile.flush()
 
-    mut_seed=random.randrange(1,1000)
     rng = msprime.RandomGenerator(mut_seed)
-    ts.generate_mutations(opts['mut_rate'],rng)
+    nodes = msprime.NodeTable()
+    edgesets = msprime.EdgesetTable()
+    sites = msprime.SiteTable()
+    mutations = msprime.MutationTable()
+    minimal_ts.dump_tables(nodes=nodes, edgesets=edgesets)
+    mutgen = msprime.MutationGenerator(rng, opts['mut_rate'])
+    mutgen.generate(nodes, edgesets, sites, mutations)
+
+    mutated_ts = msprime.load_tables(
+        nodes=nodes, edgesets=edgesets, sites=sites, mutations=mutations)
+
+    del ts
 
     logfile.write("  done generating mutations! Writing out data.\n")
     logfile.write(time.strftime('     %X %x %Z\n'))
     logfile.flush()
 
-    ts.dump(opts['treefile'])
-    ts.write_vcf(open(opts['vcffile'],'w'),ploidy=1)
+    mutated_ts.dump(opts['treefile'])
+    mutated_ts.write_vcf(open(opts['vcffile'],'w'),ploidy=1)
 
 
 logfile.write("Done!\n")

### 2017-04-22 sims/background/threeway-background-sim.py

@@ -38,9 +38,9 @@ def fileopt(fname,opts):
 parser = argparse.ArgumentParser(description=description)
 parser.add_argument('--relative_switch_time', '-w', default=0.25, type=float, 
         help="Time since rearrangement of populations in units of population size.")
-parser.add_argument('--relative_fast_M', '-m', default=10, type=float, 
+parser.add_argument('--relative_fast_M', '-M', default=10, type=float, 
         help="Migration rate for 'close' pops in units of population size.")
-parser.add_argument('--relative_slow_m', '-M', default=0.1, type=float, 
+parser.add_argument('--relative_slow_m', '-m', default=0.1, type=float, 
         help="Migration rate for 'distant' pops in units of population size.")
 parser.add_argument("--generations", "-T", type=int, help="number of generations to run for before the switch")
 parser.add_argument("--popsize", "-N", type=int, help="size of each subpopulation", default=100)

### 2017-04-22 sims/background/threeway-neutral-split-sim.py

@@ -97,16 +97,27 @@ for chrom,opts in options.items():
     logfile.write(time.strftime('     %X %x %Z\n'))
     logfile.flush()
 
-    mut_seed=random.randrange(1,1000)
     rng = msprime.RandomGenerator(mut_seed)
-    ts.generate_mutations(opts['mut_rate'],rng)
+    nodes = msprime.NodeTable()
+    edgesets = msprime.EdgesetTable()
+    sites = msprime.SiteTable()
+    mutations = msprime.MutationTable()
+    minimal_ts.dump_tables(nodes=nodes, edgesets=edgesets)
+    mutgen = msprime.MutationGenerator(rng, opts['mut_rate'])
+    mutgen.generate(nodes, edgesets, sites, mutations)
+
+    mutated_ts = msprime.load_tables(
+        nodes=nodes, edgesets=edgesets, sites=sites, mutations=mutations)
+
+    del ts
+
 
     logfile.write("  done generating mutations! Writing out data.\n")
     logfile.write(time.strftime('     %X %x %Z\n'))
     logfile.flush()
 
-    ts.dump(opts['treefile'])
-    ts.write_vcf(open(opts['vcffile'],'w'),ploidy=1)
+    mutated_ts.dump(opts['treefile'])
+    mutated_ts.write_vcf(open(opts['vcffile'],'w'),ploidy=1)
 
 
 logfile.write("Done!\n")

### 2017-04-24 templated/summarize_run.Rmd
warning for  unmatching IDs

--- 2017-03-08 templated/summarize_run.Rmd
@@
     samps <- read.table(samp.file,sep="\t",header=TRUE)
-    samps <- droplevels( samps[match(sample.ids,samps$ID),] )
+    names(samps) <- tolower(names(samps))
+    # hack for msprime output
+    if (all(grepl("^msp_",sample.ids)) & is.numeric(samps$id)) {
+        samps$id <- factor(paste0("msp_", samps$id))
+    }
+    drop_ids <- setdiff(sample.ids, levels(samps$id))
+    if (length(drop_ids)>0) {
+        warning(sprintf("These samples have no information in the samples file, %s:\n  %s.", samp.file, paste(drop_ids,collapse=" ")))
+    }
+    samps <- droplevels( samps[match(sample.ids,samps$id),] )
     samps$population <- factor(samps$population)
 } else {
+    warning("No populatio information in the sample file, %s.", samp.file)
     samps <- data.frame( 

### 2017-05-26 sims/local/summarize_run.Rmd
minor fixen

--- 2017-03-08 templated/summarize_run.Rmd
@@
     samps <- read.table(samp.file,sep="\t",header=TRUE)
-    samps <- droplevels( samps[match(sample.ids,samps$ID),] )
+    names(samps) <- tolower(names(samps))
+    # hack for msprime output
+    if (all(grepl("^msp_",sample.ids)) & is.numeric(samps$id)) {
+        # samps$id <- factor(paste0("msp_", samps$id))
+        samps$id <- paste0("msp_", 0:(nrow(samps)-1))
+    }
+    drop_ids <- setdiff(sample.ids, levels(samps$id))
+    if (length(drop_ids)>0) {
+        warning(sprintf("These samples have no information in the samples file, %s:\n  %s.", samp.file, paste(drop_ids,collapse=" ")))
+    }
+    samps <- droplevels( samps[match(sample.ids,samps$id),] )
     samps$population <- factor(samps$population)
 } else {
+    warning("No populatio information in the sample file, %s.", samp.file)
     samps <- data.frame( 

### 2017-05-26 lostruct/R/enclosing_circle.R

@@ -58,7 +58,7 @@ getCircleFrom3 <- function(xy) {
         v1  <- rowSums(xy^2)                    # first vector in the numerator
         v2x <- c( xDeltaB, -xDeltaC,  xDeltaA)  # 2nd vector numerator for Mx
         v2y <- c(-yDeltaB,  yDeltaC, -yDeltaA)  # 2nd vector numerator for My
-        ctr <- c(t(v1) %*% v2y, t(v1) %*% v2x) / (2 * (t(y) %*% v2x))  # center
+        ctr <- c(t(v1) %*% v2y, t(v1) %*% v2x) / as.vector(2 * (t(y) %*% v2x))  # center
     }
 
     return(list(ctr=ctr, rad=rad))

### 2017-05-26 sims/local/run_lostruct.R

@@ -1,0 +1,130 @@
+#!/usr/bin/Rscript
+library(optparse)
+
+invocation <- commandArgs()
+
+usage <- "\
+\
+Does a local PCA analysis.
+To run an analysis on windows of 1000 SNPs, for instance: \
+    Rscript run_lostruct.R -t snp -s 1000 \
+while to run it on windows of 10000 bps: \
+    Rscript run_lostruct.R -t bp -s 10000 \
+\
+Do
+    Rscript run_on_medicago.R -h\
+for options.\
+\
+"
+
+option_list <- list(
+    # input/output
+        make_option( c("-i","--input_dir"),   type="character",        help="Directory with input: indexed .bcf or .vcf.gz files. (REQUIRED)"),
+        make_option( c("-I","--sample_info"),   type="character",        help="File with columns labeled 'ID' and 'population' describing populations of samples in bcf file.)"),
+        make_option( c("-t","--type"),   type="character",             help="Window by SNP or by bp? (REQUIRED)"),
+        make_option( c("-s","--size"),   type="integer",               help="Size of the window, in units of type. (REQUIRED)"),
+        make_option( c("-k","--npc"),   type="integer",   default=2L,  help="Number of principal components to compute for each window. [default: %default]"),
+        make_option( c("-w","--weightfile"), type="character",            help="File name containing weights to apply to PCA."),
+        make_option( c("-m","--nmds"),   type="integer",   default=2L, help="Number of principal coordinates (MDS variables) to compute. [default: %default]"),
+        make_option( c("-o","--outdir"), type="character",             help="Directory to save results to.  [default: lostruct_results/type_%type_size_%size_jobid_%jobid/]"),
+        make_option( c("-j","--jobid"),  type="character", default=formatC(1e6*runif(1),width=6,format="d",flag="0"),   help="Unique job id. [default random]")
+    )
+opt <- parse_args(OptionParser(option_list=option_list,description=usage))
+if (is.null(opt$outdir)) { opt$outdir <- file.path("lostruct_results", 
+                                   sprintf( "type_%s_size_%d_weights_%s_jobid_%s", 
+                                           opt$type, 
[…]

### 2017-06-21 writeup/review-responses.tex
reviews; added some recomb stuff

@@
 
+Thanks for the positive feedback and the useful suggestions.
+We agree that more extensive exploration 
+using simulations would help bolster understanding of the method,
+and have now done so.
+This took a substantial amount of work, because genome-scale forwards-time simulations
+with a large number of loci under selection
+is at or beyond the current limits of computation,
+depending on the number of individuals simulated.
+
 %%%%%%%%%%%%%%
@@
 \begin{point}{}
-\ldots variation in the recombination rate \ldots 
+\ldots variation in the recombination rate \ldots creating spurious structure.
 \end{point}
+
+\reply{
+    We address this in two ways.
+    This point is addressed by comparing results with windows of different types --
+    windows of equal length in bp (or in SNPs) have different lengths in cM;
+    since these different choices show nearly identical patterns,
+    recombination rate variation cannot be driving the results.
+}
 
@@
 \end{point}
+
+\reply{
+
+}
 
@@
 \end{point}
+
+\reply{
+    This is a natural concern.
+    However, the only point at which we compare PCs in a way that could be sensitive to ordering
+    is in determining the window size -- in computing the distance between windows
+    we use a measure which is invariant under ordering.
+    We have made this more clear by moving the note about flipping signs of PCs
+    to the appendix on window choice \llname{ll:moved_pc_note}
+    and added more explicit notes about this to \llname{ll:another_pc_point} and \revref.
+}
 
@@
 \end{point}
+
+\reply{
+    This is a good point; the question is which the ``relevant'' PCs are.
+    \citet{novembre2008interpreting} showed that under isolation by distance,
+    the top two PCs should reflect the two-dimensional nature of the range,
+    and higher PCs are generally much less interpretable;
+    we used $k=2$ with this in mind.
+    We have changed this sentence \revref.
+}
 
@@
 \end{point}
+
+\reply{
+    First: we think that ``finding heterogeneity'' is in fact a well-defined goal,
+    although it was not that well-defined in the paper;
+    we have hopefully improved on this in the Introduction \revref.
+    Expanding a bit more:
+    We strongly agree that methods that seek to test well-defined hypotheses
+    are extremely useful and powerful.
+    We also feel that methods for visualization and exploration are also useful --
+    a primary example here being PCA.
+    If PCA is useful -- and we think that it is --
+    then it should be important to also know how much the thing that PCA is summarizing
+    varies along the genome,
+    in the same way that knowing the mean of some quantity in a population 
+    is only of limited usefulness
+    without also knowing the corresponding populaion variance.
+}
 

### 2017-06-21 writeup/local_pca_paper.tex
.
reviews; added some recomb stuff

@@
 
+% some figs with many points make the pdfs render slow
+\newif\ifusepngs
+\usepngstrue
+
 \usepackage{color}
@@
 
+To-do: cite \citet{duforetfrebourg2015detecting}.
+
 % NOTE on historical uses of 'population structure'
@@
 
-As reviewed above,
-it is common to describe variation along the genome of simple statistics such as $F_{ST}$,
-and also to use methods such as PCA to visualize large-scale patterns in genome-wide relatedness.
-In this paper we 
-apply PCA locally along the genome,
-thus describing in an unbiased, descriptive way
-how patterns of mean relatedness vary systematically along the genome, 
+It is common to describe variation along the genome of simple statistics such as $F_{ST}$
+and to interpret the results in terms of the action of selection 
+\citep{speciation_islands,differentiation_islands,flycatchers}.
+However, it has been pointed out that a given pattern (e.g., valleys of $F_{ST}$)
+can be caused by more than one biological process \citep{hahn_fst,flycatchers2},
+which in retrospect is unsuprising given that we are using a single statistic
+to describe a complex process.
+It is also common to use methods such as PCA to visualize large-scale patterns in 
+mean genome-wide relatedness.
+In this paper we show if and how patterns of mean relatedness vary systematically along the genome, 
 in a way particularly suited to large samples from geographically distributed populations.
-Geographic population structure sets the stage by establishing similar patterns of relatedness across much of the genome;
-we then aim to describe how this structure is affected by selection and other factors.
+Geographic population structure sets the stage by establishing ``background'' patterns of relatedness;
+our method then describes how this structure is affected by selection and other factors.
 Our aim is not to identify outlier loci,
-but rather to describe larger-scale variation shared by many parts of the genome.
-By doing this with three taxa with diverse population histories,
-we can compare relative contributions of different sorts of variation across taxa;
-since the method is an visualization method, 
-we allow ourselves to be surprised by unexpected signals in the data.
+but rather to describe larger-scale variation shared by many parts of the genome;
+correlation of this variation with known genomic features
+can then be used to uncover its source. \revpoint{2}{1}
 
@@
 
-The top few principal components are generally good summaries of population structure; 
-we follow common practice and usually only use the first two (referred to as $\pcone$ and $\pctwo$).
+The top 2--5 principal components are generally good summaries of population structure; 
+for ease of visualization we usually only use the first two (referred to as $\pcone$ and $\pctwo$),
+and check that results hold using more. \revpoint{1}{7}
 The above procedure can be performed on any subset of the data;
@@
 the result after applying to all SNPs in the $j^\text{th}$ window.
-
-Since the definition of eigenvectors does not specify their sign,
-when comparing between windows we choose signs to best match each other:
-after choosing $\pcone_1$, for instance, 
-if $u$ is the first eigenvector obtained from the covariance matrix
-for window $j$,
-then we next choose $\pcone_j = \pm u$,
-where the sign is chosen according to which of 
-$\| \pcone_{1} - u \|$ or
-$\| \pcone_{1} + u \|$ 
-is smaller.
+(Note, however, that our measure of dissimilarity between windows
+does not depend on PC ordering.) \linelabel{ll:another_pc_point}
 
@@
 obtained using the top $k$ PCs,
-which is invariant under reflections and rotations and yet contains all other information about the PCs.
+which is invariant under ordering of the PCs \revpoint{1}{3},
+reflections, and rotations and yet contains all other information about the PCs.
 (For results shown here, we use $k=2$;
@@
     \begin{center}
+        \ifusepngs
+           \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv.png}
+       \else
            \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
+       \fi
     \end{center}
@@
 In particular, the recombination rates estimated by \citet{paape2012finescale} appear visually to be similar,
-but were not available in a form that can be compared to our MDS values.
+but were not available in a form that can be reliably compared to our MDS values. \revpoint{2}{8}
 
@@
 averaged over samples.
-(Recall that the signs for PCs are chosen to match each other.)
-Then, the mean variance across windows is
+The definition of eigenvectors does not specify their sign,
+and so when comparing between windows we choose signs to best match each other:
+after choosing $\pcone_1$, for instance, 
+if $u$ is the first eigenvector obtained from the covariance matrix
+for window $j$,
+then we next choose $\pcone_j = \pm u$,
+where the sign is chosen according to which of 
+$\| \pcone_{1} - u \|$ or
+$\| \pcone_{1} + u \|$ 
+is smaller.
+\linelabel{ll:moved_pc_note}
+
+After doing this, the mean variance across windows is
 \begin{align*}
@@
     \begin{center}
+        \ifusepngs
+           \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human.png}
+       \else
            \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human}
+       \fi
     \end{center}
@@
     \begin{center}
+        \ifusepngs
+            \includegraphics{MDS_allchr_Together_plot_samples_noinv.png}
+        \else
             \includegraphics{MDS_allchr_Together_plot_samples_noinv}
+        \fi
     \end{center}
@@
     \begin{center}
+        \ifusepngs
+           \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr.png}
+       \else
            \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr}
+       \fi
     \end{center}

### 2017-06-21 medicago/recomb_rates/plot_gene_density.R

@@ -1,0 +1,109 @@
+# using data from Tim Paape, 8/17/2016:
+# Hi Peter R
+# 
+# Here are the complete LDhat output files for all 1kb sliding window and all
+# chromosomes, with positions of each window on each chromosome (sliding
+# windows).  In our GBE paper we published only chromosomes 2, 3 and 5. The
+# other 5 chromosomes are unpublished recombination rates  for Medicago, so
+# depending on what you want to do perhaps we can have that discussion later. 
+# 
+# This data comes from SNPs called using our version 3.0 gene annotations (also
+# used in the Branca et al. 2011 PNAS paper, 26 accessions). Since then there
+# have been v3.5 and 4.0 gene annotations and possible slightly modified
+# scaffolds in the genome assembly versions.  Also, if you need I can find the
+# centromere positions if you need them for v3.0.
+# 
+# cheers
+# Tim
+
+recomb <- read.table("from_paape/07_recom.txt",header=TRUE)
+recomb$mid <- (recomb$sStart + recomb$sEnd)/2
+levels(recomb$chr) <- gsub("MtChr","chr",levels(recomb$chr))
+
+# map <- read.csv("from_paape/Mtruncatula_mapbased_recomb_2may_b.csv",header=TRUE)
+# map <- map[,c("chr","start","end","cM","cm.bp...1.000.000","ave.cM...Mbp....3.window.moving.average")]
+
+mds <- do.call( rbind, lapply( 1:8, function (k) {
+                          read.table(file.path("..","results",sprintf("mds_chr%d_window_10000snp.tsv",k)), header=TRUE)
+        } ) )
+mds$mid <- (mds$start+mds$end)/2
+levels(mds$chrom) <- tolower(levels(mds$chrom))
+
+xlims <- lapply(levels(mds$chrom), function (this.chrom) { range(subset(recomb,chr==this.chrom)$mid, subset(mds,chrom==this.chrom)$mid) } )
+names(xlims) <- levels(mds$chrom)
+
+pdf(file="recomb_and_mds.pdf", width=6, height=6, pointsize=10)
+layout( matrix(1:16,nrow=4), heights=c(1,1.2,1,1.2), widths=c(1.2,1,1,1) )
+for (this.chrom in paste0('chr',1:8)) {
+    left.mar <- if (this.chrom %in% paste0('chr',1:2)) { 4} else {0}
+    par(mar=c(0.25,left.mar,2,2)+.1)
[…]

### 2017-06-25 writeup/review-responses.tex
.

@@
 
+\reply{
+    We made a stab at checking this,
+    and obtained the best current version of the Medicago recombination map
+    from Tim Paape and Peter Tiffin.
+    There are two versions: a very coarse physical map, and a fine-scale map estimated using LDhat.
+    Both are on a previous version of the assembly, 
+    and the LDhat estimates from only three chromosomes
+    have been published.
+    Given uncertainties in using and interpreting these data,
+    and the further uncertainty in what the relevant window size should be
+    (strength of background selection depends on ``nearby'' density of sites under selection,
+    but how close ``nearby'' is depends on many factors),
+    we decided not to carry though this analysis.
+}
+
 \begin{point}{Application:}

### 2017-06-25 medicago/recomb_rates/plot_gene_density.R

@@ -20,6 +20,15 @@ recomb <- read.table("from_paape/07_recom.txt",header=TRUE)
 recomb$mid <- (recomb$sStart + recomb$sEnd)/2
 levels(recomb$chr) <- gsub("MtChr","chr",levels(recomb$chr))
 
+# convert to BED
+recomb_bed <- data.frame(chrom=recomb$chr,
+                         chromStart=recomb$sStart,
+                         chromEnd=recomb$sEnd)
+write.table(recomb_bed, file="from_paape/07_recomb.bed", row.names=FALSE, quote=FALSE, col.names=FALSE, sep='\t')
+# then did liftOver - from source as at http://genome-source.cse.ucsc.edu/gitweb/?p=kent.git;a=blob;f=src/userApps/README
+# and 
+#  liftOver from_paape/07_recomb.bed from_paape/07_recomb.bed liftOver/Mt35.liftover.chain from_paape/07_recomb.bedfrom_paape/07_recomb_mt35.bed from_paape/07_recomb.bedfrom_paape/07_recomb_unmapped.bed 
+
 # map <- read.csv("from_paape/Mtruncatula_mapbased_recomb_2may_b.csv",header=TRUE)
 # map <- map[,c("chr","start","end","cM","cm.bp...1.000.000","ave.cM...Mbp....3.window.moving.average")]
 

### 2017-06-26 writeup/talk/files/export-layers-svg.sh
final version

@@ -1,0 +1,29 @@
+#!/bin/bash
+# CC-BY https://answers.launchpad.net/inkscape/+question/48164
+# $1 is the file to extract layers from
+# $2 is not needed but can be used eg with "-d 300" to send the png resolution or any other inkscape switch through
+#
+# needs xmlstarlet
+#
+# Notes:
+#  - doesn't deal with spaces in object names
+
+type xmlstarlet >/dev/null 2>&1 || { echo >&2 "can't find xmlstarlet"; exit 1; }
+
+TMPFILE=$(mktemp /tmp/output.XXXXXXXXX).svg
+cp $1 $TMPFILE
+
+YESLAYERS=${*:2}
+ALLLAYERS=$(xmlstarlet sel -t -m "//*[@inkscape:groupmode=\"layer\"]" -v "concat(@inkscape:label,' ')" $TMPFILE|sort -Vr)
+NOLAYERS=$(comm -1 -3 <(echo $YESLAYERS|tr ' ' '\n'|sort) <(echo $ALLLAYERS|tr ' ' '\n'|sort))
+
+for layer in $NOLAYERS
+do
+    echo $layer
+    id=$(xmlstarlet sel -t -m "//*[@inkscape:label=\"$layer\"]" -v "@id" $TMPFILE)
+    xmlstarlet ed -S -L -d "//*[@id=\"$id\"]" $TMPFILE
+done
+
+inkscape --without-gui --export-pdf=/dev/stdout $TMPFILE
+
+# [ -f $TMPFILE ] && rm $TMPFILE

### 2017-07-17 sims/background/background-introgression-fixed-s-sim.py
introgression with fixed, regular loci

@@ -1,12 +1,6 @@
 #!/usr/bin/env python3
 description = '''
-Simulate with simuPOP AND write to msprime/vcf:
-    a population arrangement switches from A<-B|C to A|B->C 
-    at some point in the past.
-
-Output in selloci is of the form
-   loc a1 a2 fitness
-meaning locus number, allele1, allele2, fitness when it is first seen.
+Simulates on a rectangular grid that was bisected (in 'width') for some period in the past.
 '''
 
 import gzip
@@ -36,36 +30,70 @@ def fileopt(fname,opts):
     return fobj
 
 parser = argparse.ArgumentParser(description=description)
-parser.add_argument('--relative_switch_time', '-w', default=0.25, type=float, 
-        help="Time since rearrangement of populations in units of population size.")
-parser.add_argument('--relative_fast_M', '-M', default=10, type=float, 
-        help="Migration rate for 'close' pops in units of population size.")
-parser.add_argument('--relative_slow_m', '-m', default=0.1, type=float, 
-        help="Migration rate for 'distant' pops in units of population size.")
-parser.add_argument("--generations", "-T", type=int, help="number of generations to run for before the switch")
-parser.add_argument("--popsize", "-N", type=int, help="size of each subpopulation", default=100)
-parser.add_argument("--length", "-L", type=float, help="number of bp in the chromosome", default=1e4)
-parser.add_argument("--nloci", "-l", type=int, help="number of selected loci", default=20)
-parser.add_argument("--sel_mut_rate", "-u", type=float, help="mutation rate of selected alleles", default=1e-7)
-parser.add_argument("--recomb_rate", "-r", type=float, help="recombination rate", default=1e-7)
-parser.add_argument("--selection_coef","-S", type=float, help="strength of selection",default=.1)
-parser.add_argument("--nsamples", "-k", type=int, help="number of *diploid* samples, total")
-parser.add_argument("--ancestor_age", "-A", type=float, help="time to ancestor above beginning of sim")
-parser.add_argument("--mut_rate", "-U", type=float, help="mutation rate", default=1e-7)
+parser.add_argument("--post_generations","-R", type=int, dest="post_generations",
+        help="number of generations since the division")
[…]

### 2017-07-17 sims/local/summarize_run.Rmd
.

@@
 knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
-do.pdfs <- FALSE
+if (!exists("do.pdfs")) { do.pdfs <- FALSE }
 ```
@@
 }
+knitr::knit_hooks$set(slide.mar = function(before, options, envir) {
+                if (before) {
+                    par(mai=c(.4,.4,.05,.05),  # margin for presentations
+                        mgp=c(2,0.7,0))
+                }
+            })
+knitr::opts_chunk$set(slide.mar=TRUE)
 pdf_copy <- function (
-                width=6,
+                width=5, 
                 height=width*knitr::opts_current$get("fig.height")/knitr::opts_current$get("fig.width"),
@@
     abline( v=chrom.dividers/1e6, lty=3, col=adjustcolor("grey",0.5), lwd=2 )
-    if (chrom.labels) axis( 1, at=chrom.mids/1e6, labels=chroms, las=0, tick=FALSE )
+    if (chrom.labels) axis( 1, at=chrom.mids/1e6, labels=paste("chromosome", chroms), las=0, tick=FALSE )
     points( regions$pos/1e6, y, ...)
@@
                 } ) )
-if (do.pdfs) { pdf_copy() }
 ```
@@
 spacing <- 1
-opar <- par(mar=c(4,4,2,1)+.1,mgp=c(2.5,0.8,0))
+opar <- par(mar=par("mar"),mgp=c(0.5,0.8,0))
 layout(matrix(c(rep(1,length(mds.cols)),1+seq_along(mds.cols)),ncol=2),
@@
      col=adjustcolor(ccols,0.75),  asp=1,
+     xaxt='n', yaxt='n',
      xlab="MDS coordinate 1", ylab="MDS coordinate 2" )
@@
 text( corner.mds, labels=seq_len(nrow(corner.mds)), 
-        col=corner.cols, cex=2, lwd=2 )
+        col="black", lwd=2 )
 opar2 <- par(mar=c(par("mar"),spacing/2)[c(5,2,3,4)])
@@
             colnames(vectors) <- paste("PC", c(i,j))
+            par(mgp=c(0.7,0.7,0), mar=c(2,2,2,0)+.1)
             plot(vectors, pch=pop.pch[samps$population], 
-                    col=pop.cols[samps$population] )
+                    col=pop.cols[samps$population],
+                    xaxt='n', yaxt='n' )
             if (i==1 && j==2) {
-                mtext(paste("corner",k),side=3,cex=2)
+                mtext(paste("corner",k),side=3)
             }

### 2017-07-18 writeup/local_pca_paper.tex
added some refs

@@
 
-To-do: cite \citet{duforetfrebourg2015detecting}.
-
 % NOTE on historical uses of 'population structure'
@@
 These observations motivate many methods to search for genetic loci under selection,
-as for example in \citet{huertasnchez2013genetic} and \citet{martin2016natural}.
+as for example in \citet{huertasnchez2013genetic}, \citet{martin2016natural}, 
+and \citet{duforetfrebourg2015detecting}.
 
@@
 adaptive introgression and/or local adaptation 
-(e.g., \citep{pease2013accurate,ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting});
+(e.g., \citet{pease2013accurate,ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting});
 and theoretical expectations predict that geographic patterns of relatedness should depend on selection
@@
 Patterns in genome-wide relatedness are often summarized
-by applying principal components analysis (PCA, \citep{patterson2006population}) 
+by applying principal components analysis (PCA, \citet{patterson2006population}) 
 to the genetic covariance matrix,
-as pioneered by \citetx[Menozzi et al]{menozzi1978synthetic}.
+as pioneered by \citet{menozzi1978synthetic}.
 % About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}.
@@
 Modeling such ``background'' kinship between samples
-is essential to genome-wide association studies (GWAS, \citep{price2006principal,astle2009population}),
+is essential to genome-wide association studies (GWAS, \citet{price2006principal,astle2009population}),
 % and PCA has often been used in this context \citep{price2006principal}.
@@
 
+PCA has been applied to genomic windows in methods to infer
+tracts of local ancestry in recently admixed populations
+\citep{bryc2010genomewide,brisbin2012pcadmix},
+and to identify putative chromosomal inversions \citep{ma2012investigation}.
+\revpoint{1}{8}
+
 A note on nomenclature: In this work we describe variation in patterns of relatedness using local PCA,
@@
 A number of general methods for dimensionality reduction also use a strategy of ``local PCA''
-(e.g., \citep{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear}),
+(e.g., \citet{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear}),
 performing PCA not on the entire dataset but instead on subsets of observations,
@@
 and to interpret the results in terms of the action of selection 
-\citep{speciation_islands,differentiation_islands,flycatchers}.
-However, it has been pointed out that a given pattern (e.g., valleys of $F_{ST}$)
-can be caused by more than one biological process \citep{hahn_fst,flycatchers2},
+(e.g., \citet{turner2005genomic,ellegren2012genomic}).
+However, a given pattern (e.g., valleys of $F_{ST}$)
+can be caused by more than one biological process \citep{cruickshank2014reanalysis,burri2015linked},
 which in retrospect is unsuprising given that we are using a single statistic

### 2017-07-18 writeup/review-responses.tex
added some refs
medicago recomb rate resolution

--- 2017-06-21 writeup/review-responses.tex
@@
 
+\reply{
+    Good point: we now reference this work \revref.
+}
+
 %%%%%%%%%%%%%%
@@
 
+\reply{
+    This is a good point, but XXX
+    We now cite PCAdmix \revreffull{1}{8}.
+}
+
 \begin{point}{}
@@
 \end{point}
+
+\reply{
+    We made a stab at checking this,
+    and obtained the best version to date of the Medicago recombination map
+    from Tim Paape and Peter Tiffin.
+    There are two versions: a very coarse physical map, and a fine-scale map estimated using LDhat.
+    However, both are on version 3.0 of the assembly,
+    while all other coordinates (sequencing data; gene annotations) are in version 4.0.
+    Furthermore, as Peter Tiffin told us,
+    ``apparently there are no files that translate Mt3.0 to Mt4.0 locations (yes, seems a bit silly)."
+    There is a liftOver chain file for translating 3.5 to 4.0, and
+    ``the differences in the Mt3.0 and Mt3.5 assemblies are, however, apparently relatively minor''.
+    On this basis, we produced the desired figure assuming that Mt3.0 coordinates
+    are the same as Mt3.5 coordinates, included to satisfy the reviewers' curiosity: \\
+    \includegraphics{../medicago/recomb_rates/recomb_and_mds} \\
+    However, given uncertainties in this mapping, the relatively poor match of window sizes,
+    the large number of unmappable windows, and the nature of the recombination data
+    (produced with LDhat, not with actual observations of recombiations),
+    we decided not to include this (but have provided a note, \revref).
+}
 

### 2017-07-18 writeup/local_pca_paper.tex
medicago recomb rate resolution

@@
 so we cannot rule out alternative hypotheses.
-In particular, the recombination rates estimated by \citet{paape2012finescale} appear visually to be similar,
-but were not available in a form that can be reliably compared to our MDS values. \revpoint{2}{8}
+In particular, fine-scale recombination rate estimates are not available in a form mappable
+to Mt4.0 coordinates (although those in \citet{paape2012finescale} appear visually similar).
+\revpoint{2}{8}
 

### 2017-07-18 medicago/recomb_rates/plot_gene_density.R

@@ -15,19 +15,47 @@
 # 
 # cheers
 # Tim
+####
+# 
+# In short, the LDhat estimates are from the 3.0 assembly, while annotations are from 4.0.
+# There exists a 3.5 -> 4.0 liftOver chain file, obtained from
+#   https://de.cyverse.org/dl/d/7D271AC5-463F-4840-B1E3-25FC2D5AE015/Mt35.liftover.chain
+# (linked to from http://www.medicagogenome.org/downloads )
+#
+####
+# ... BUT as Peter Tiffin says, 8/21/16:
+# 
+# I checked with Joseph and Peng.  Apparently there are no files that translate
+# Mt3.0 to Mt4.0 locations (yes, seems a bit silly).  
+# 
+# The differences in the Mt3.0 and Mt3.5 assemblies are, however, apparently
+# relatively minor and although there is not a liftOver file 3.5 --> 4.0 there
+# is a gene conversion table at JCVI Medicago downloads page.  I'm not sure it
+# is worth it, but I suppose one could use the Mt3.0 gene names (which should
+# be unchanged for Mt3.5 -- just more genes added) and then use for a rough
+# translation between 3.0 and 4.0 locations.
+#
+####
 
 recomb <- read.table("from_paape/07_recom.txt",header=TRUE)
-recomb$mid <- (recomb$sStart + recomb$sEnd)/2
 levels(recomb$chr) <- gsub("MtChr","chr",levels(recomb$chr))
 
 # convert to BED
 recomb_bed <- data.frame(chrom=recomb$chr,
                          chromStart=recomb$sStart,
-                         chromEnd=recomb$sEnd)
+                         chromEnd=recomb$sEnd,
+                         name=1:nrow(recomb))
 write.table(recomb_bed, file="from_paape/07_recomb.bed", row.names=FALSE, quote=FALSE, col.names=FALSE, sep='\t')
 # then did liftOver - from source as at http://genome-source.cse.ucsc.edu/gitweb/?p=kent.git;a=blob;f=src/userApps/README
 # and 
-#  liftOver from_paape/07_recomb.bed from_paape/07_recomb.bed liftOver/Mt35.liftover.chain from_paape/07_recomb.bedfrom_paape/07_recomb_mt35.bed from_paape/07_recomb.bedfrom_paape/07_recomb_unmapped.bed 
[…]

### 2017-07-30 writeup/local_pca_paper.tex
moved over to new argrecorder set-up

@@
     Descriptive statistics for each dataset used.
-    % with the chosen window sizes.
-    % See text for how window sizes were chosen;
-    Columns 2--4 give statistics describing the window sizes used for most analyses for each dataset
-    (windows had a fixed number of SNPs).
-    The final column gives the percent variance explained by the top two principal components,
-    averaged across independent PCA of each window.
+    \revpoint{2}{18}
+    % % with the chosen window sizes.
+    % % See text for how window sizes were chosen;
+    % Columns 2--4 give statistics describing the window sizes used for most analyses for each dataset
+    % (windows had a fixed number of SNPs).
+    % The final column gives the percent variance explained by the top two principal components,
+    % averaged across independent PCA of each window.
     \label{tab:data_stats}
@@
 Below, we discuss the results and likely underlying causes.
+
+%%%%%%%%%%%
+\subsection{Validation}
+
+Address
+mutation rate variation;
+recombination rate variation;
+choice of number of PCs;
+choice of window size;
+variation in missingness;
+maybe differing sample sizes and reweighting.
 

### 2017-07-30 sims/background/background-fixed-s-sim.py

@@ -9,7 +9,7 @@ from optparse import OptionParser
 import math
 import time
 import random
-from ftprime import RecombCollector, ind_to_chrom, mapa_labels
+from ftprime import RecombCollector
 import msprime
 
 def fileopt(fname,opts):
@@ -44,7 +44,7 @@ parser.add_option("-k","--nsamples",dest="nsamples",help="number of *diploid* sa
 parser.add_option("-A","--ancestor_age",dest="ancestor_age",help="time to ancestor above beginning of sim")
 parser.add_option("-U","--mut_rate",dest="mut_rate",help="mutation rate",default=1e-7)
 parser.add_option("-t","--treefile",dest="treefile",help="name of output file for trees (default: not output)",default=None)
-
+parser.add_option("-I","--simplify_interval",dest="simplify_interval",default=500)
 parser.add_option("-o","--outfile",dest="outfile",help="name of output VCF file (default: not output)",default=None)
 parser.add_option("-g","--logfile",dest="logfile",help="name of log file (or '-' for stdout)",default="-")
 parser.add_option("-s","--selloci_file",dest="selloci_file",help="name of file to output selected locus information",default="sel_loci.txt")
@@ -173,35 +173,38 @@ pop.evolve(
         ] ),
     postOps=[
         sim.Stat(numOfSegSites=sim.ALL_AVAIL, step=50),
-        sim.PyEval(r"'Gen: %2d #seg sites: %d\n' % (gen, numOfSegSites)", step=50)
+        sim.PyEval(r"'Gen: %2d #seg sites: %d\n' % (gen, numOfSegSites)", step=50),
+        sim.PyOperator(lambda pop: rc.simplify(pop.indInfo("ind_id")) or True, 
+                       step=options.simplify_interval),
     ],
     gen = generations
 )
-
 
 logfile.write("Done simulating!\n")
 logfile.write(time.strftime('%X %x %Z')+"\n")
 logfile.write("----------\n")
 logfile.flush()
 
-# writes out events in this form:
-# offspringID parentID startingPloidy rec1 rec2 ....
-
 locations = [pop.subPopIndPair(x)[0] for x in range(pop.popSize())]
-rc.add_diploid_samples(nsamples=args.nsamples, sample_ids=pop.indInfo("ind_id"),
[…]

### 2017-07-30 sims/background/background-introgression-fixed-s-sim.py

@@ -8,7 +8,7 @@ import sys, os
 import math
 import time
 import random
-from ftprime import RecombCollector, ind_to_chrom, mapa_labels
+from ftprime import RecombCollector
 import msprime
 import argparse
 
@@ -62,7 +62,10 @@ parser.add_argument("--mut_rate","-U", type=float, dest="mut_rate",
         help="mutation rate",default=1e-7)
 parser.add_argument("--treefile","-t", type=str, dest="treefile",
         help="name of output file for trees (default: not output)",default=None)
-parser.add_argument("--seed", "-d", type=int, help="random seed", default=random.randrange(1,1000))
+parser.add_argument("--simplify_interval", "-I", dest="simplify_interval", type=int,
+        help="Interval between simplify steps.", default=500)
+parser.add_argument("--seed", "-d", dest="seed", type=int,
+        help="random seed", default=random.randrange(1,1000))
 
 parser.add_argument("--outfile","-o", type=str, dest="outfile",
         help="name of output VCF file (default: not output)",default=None)
@@ -136,7 +139,6 @@ logfile.write(str(locus_position)+"\n")
 logfile.write("----------\n")
 logfile.flush()
 
-
 init_geno=[sim.InitGenotype(freq=init_freqs[k],loci=init_classes[k]) for k in range(len(init_freqs))]
 
 def fitness_fun(loc, alleles):
@@ -154,11 +156,15 @@ pop = sim.Population(
 id_tagger = sim.IdTagger()
 id_tagger.apply(pop)
 
-# record recombinations
-rc = RecombCollector(
-        first_gen=pop.indInfo("ind_id"), ancestor_age=args.ancestor_age, 
-                              length=2*args.length, 
-                              locus_position=locus_position+[args.length, 2*args.length])
+# set up recomb collector
+first_gen = pop.indInfo("ind_id")
+init_ts = msprime.simulate(2*len(first_gen),
+                           length=max(locus_position))
+haploid_labels = [(k,p) for k in first_gen 
+                        for p in (0,1)]
+node_ids = {x:j for x, j in zip(haploid_labels, init_ts.samples())}
+rc = RecombCollector(ts=init_ts, node_ids=node_ids,
[…]

### 2017-07-30 sims/background/background-introgression-sim.py

@@ -8,7 +8,7 @@ import sys, os
 import math
 import time
 import random
-from ftprime import RecombCollector, ind_to_chrom, mapa_labels
+from ftprime import RecombCollector
 import msprime
 import argparse
 
@@ -64,7 +64,10 @@ parser.add_argument("--mut_rate","-U", type=float, dest="mut_rate",
         help="mutation rate",default=1e-7)
 parser.add_argument("--treefile","-t", type=str, dest="treefile",
         help="name of output file for trees (default: not output)",default=None)
-parser.add_argument("--seed", "-d", type=int, help="random seed", default=random.randrange(1,1000))
+parser.add_argument("--simplify_interval", "-I", dest="simplify_interval", type=int,
+        help="Interval between simplify steps.", default=500)
+parser.add_argument("--seed", "-d", dest="seed", type=int,
+        help="random seed", default=random.randrange(1,1000))
 
 parser.add_argument("--outfile","-o", type=str, dest="outfile",
         help="name of output VCF file (default: not output)",default=None)
@@ -178,11 +181,15 @@ pop = sim.Population(
 id_tagger = sim.IdTagger()
 id_tagger.apply(pop)
 
-# record recombinations
-rc = RecombCollector(
-        first_gen=pop.indInfo("ind_id"), ancestor_age=args.ancestor_age, 
-                              length=args.length, 
-                              locus_position=locus_position)
+# set up recomb collector
+first_gen = pop.indInfo("ind_id")
+init_ts = msprime.simulate(2*len(first_gen),
+                           length=max(locus_position))
+haploid_labels = [(k,p) for k in first_gen 
+                        for p in (0,1)]
+node_ids = {x:j for x, j in zip(haploid_labels, init_ts.samples())}
+rc = RecombCollector(ts=init_ts, node_ids=node_ids,
+                     locus_position=locus_position)
 
 migr_rates = migrRates(args.migr, m=args.gridwidth, n=args.gridheight, barrier=False)
 barrier_rates = migrRates(args.migr, m=args.gridwidth, n=args.gridheight, barrier=True)
@@ -221,7 +228,9 @@ pop.evolve(
         ] ),
     postOps=[
[…]

### 2017-07-30 sims/background/background-sim.py

@@ -8,7 +8,7 @@ import sys, os
 import math
 import time
 import random
-from ftprime import RecombCollector, ind_to_chrom, mapa_labels
+from ftprime import RecombCollector
 import msprime
 import argparse
 
@@ -60,15 +60,18 @@ parser.add_argument("--mut_rate","-U", type=float, dest="mut_rate",
         help="mutation rate",default=1e-7)
 parser.add_argument("--treefile","-t", type=str, dest="treefile",
         help="name of output file for trees (default: not output)",default=None)
-parser.add_argument("--seed", "-d", type=int, help="random seed", default=random.randrange(1,1000))
+parser.add_argument("--simplify_interval", "-I", dest="simplify_interval", type=int,
+        help="Interval between simplify steps.", default=500)
+parser.add_argument("--seed", "-d", dest="seed", type=int,
+        help="random seed", default=random.randrange(1,1000))
 
 parser.add_argument("--outfile","-o", type=str, dest="outfile",
         help="name of output VCF file (default: not output)",default=None)
 parser.add_argument("--logfile","-g", type=str, dest="logfile",
         help="name of log file (or '-' for stdout)",default="-")
 parser.add_argument("--selloci_file","-s", type=str, dest="selloci_file",
-        help="name of file to output selected locus information",default="sel_loci.txt")
-parser.add_argument("--samples_file", "-e",
+        help="name of file to output selected locus information",default="(dir)/sel_loci.txt")
+parser.add_argument("--samples_file", "-e", type=str, dest="samples_file",
 	help="name of file to output information on samples (default=(dir)/samples.tsv)")
 
 args = parser.parse_args()
@@ -155,10 +158,15 @@ pop = sim.Population(
 id_tagger = sim.IdTagger()
 id_tagger.apply(pop)
 
-# record recombinations
-rc = RecombCollector(
-        first_gen=pop.indInfo("ind_id"), ancestor_age=args.ancestor_age, 
-                              length=args.length, locus_position=locus_position)
+# set up recomb collector
+first_gen = pop.indInfo("ind_id")
+init_ts = msprime.simulate(2*len(first_gen),
[…]

### 2017-07-30 sims/background/threeway-background-fixed-s-sim.py

@@ -14,7 +14,7 @@ import sys, os
 import math
 import time
 import random
-from ftprime import RecombCollector, ind_to_chrom, mapa_labels
+from ftprime import RecombCollector
 import msprime
 import argparse
 
@@ -52,6 +52,8 @@ parser.add_argument("--selection_coef","-S", type=float, help="strength of selec
 parser.add_argument("--nsamples", "-k", type=int, help="number of *diploid* samples, total")
 parser.add_argument("--ancestor_age", "-A", type=float, help="time to ancestor above beginning of sim")
 parser.add_argument("--mut_rate", "-U", type=float, help="mutation rate", default=1e-7)
+parser.add_argument("--simplify_interval", "-I", dest="simplify_interval", type=int,
+        help="Interval between simplify steps.", default=500)
 parser.add_argument("--seed", "-d", type=int, help="random seed", default=random.randrange(1,1000))
 
 parser.add_argument("--treefile", "-t", help="name of output file for trees (default: not output)", default=None)
@@ -127,10 +129,15 @@ pop = sim.Population(
 id_tagger = sim.IdTagger()
 id_tagger.apply(pop)
 
-# record recombinations
-rc = RecombCollector(
-        first_gen=pop.indInfo("ind_id"), ancestor_age=args.ancestor_age, 
-                              length=args.length, locus_position=locus_position)
+# set up recomb collector
+first_gen = pop.indInfo("ind_id")
+init_ts = msprime.simulate(2*len(first_gen),
+                           length=max(locus_position))
+haploid_labels = [(k,p) for k in first_gen 
+                        for p in (0,1)]
+node_ids = {x:j for x, j in zip(haploid_labels, init_ts.samples())}
+rc = RecombCollector(ts=init_ts, node_ids=node_ids,
+                     locus_position=locus_position)
 
 
 migr_init = [ [ 0, args.slow_m, args.slow_m ],
@@ -171,7 +178,9 @@ pop.evolve(
         ] ),
     postOps=[
         sim.Stat(numOfSegSites=sim.ALL_AVAIL, step=50),
-        sim.PyEval(r"'Gen: %2d #seg sites: %d\n' % (gen, numOfSegSites)", step=50)
[…]

### 2017-07-30 sims/background/threeway-background-sim.py

@@ -14,7 +14,7 @@ import sys, os
 import math
 import time
 import random
-from ftprime import RecombCollector, ind_to_chrom, mapa_labels
+from ftprime import RecombCollector
 import msprime
 import argparse
 
@@ -53,6 +53,8 @@ parser.add_argument("--gamma_beta", "-b", type=float, help="beta parameter in ga
 parser.add_argument("--nsamples", "-k", type=int, help="number of *diploid* samples, total")
 parser.add_argument("--ancestor_age", "-A", type=float, help="time to ancestor above beginning of sim")
 parser.add_argument("--mut_rate", "-U", type=float, help="mutation rate", default=1e-7)
+parser.add_argument("--simplify_interval", "-I", dest="simplify_interval", type=int,
+        help="Interval between simplify steps.", default=500)
 parser.add_argument("--seed", "-d", type=int, help="random seed", default=random.randrange(1,1000))
 
 parser.add_argument("--treefile", "-t", help="name of output file for trees (default: not output)", default=None)
@@ -158,10 +160,18 @@ args.total_generations=args.generations+args.switch_time
 id_tagger = sim.IdTagger()
 id_tagger.apply(pop)
 
-# record recombinations
-rc = RecombCollector(
-        first_gen=pop.indInfo("ind_id"), ancestor_age=args.ancestor_age, 
-                              length=args.length, locus_position=locus_position)
+# set up recomb collector
+first_gen = pop.indInfo("ind_id")
+init_ts = msprime.simulate(2*len(first_gen),
+                           length=max(locus_position))
+haploid_labels = [(k,p) for k in first_gen 
+                        for p in (0,1)]
+node_ids = {x:j for x, j in zip(haploid_labels, init_ts.samples())}
+rc = RecombCollector(ts=init_ts, node_ids=node_ids,
+                     locus_position=locus_position)
+
+migr_rates = migrRates(args.migr, m=args.gridwidth, n=args.gridheight, barrier=False)
+barrier_rates = migrRates(args.migr, m=args.gridwidth, n=args.gridheight, barrier=True)
 
 pop.evolve(
     initOps=[
@@ -190,7 +200,9 @@ pop.evolve(
         ] ),
     postOps=[
[…]

### 2017-08-08 templated/summarize_run.Rmd
allow pdf output in template

@@
 knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
-do.pdfs <- FALSE
+# set do.pdfs to TRUE to output PDFs of figures as well
+if (!exists("do.pdfs")) { do.pdfs <- FALSE }
 ```

### 2017-10-08 sims/msp/summarize_run.Rmd
.

--- 2017-08-08 templated/summarize_run.Rmd
@@
 ```{r get_corners}
-mds.corners <- corners( mds.coords[,mds.cols[1:2]], prop=.05 )
+mds.corners <- corners( mds.coords[,mds.cols[1:2]], prop=.01 )
 # set up colors and pchs for corners

### 2017-10-08 templated/bcf_and_index.sh

@@ -14,7 +14,7 @@ do
         echo "File $1 does not exist."
     else 
         OUT=${1%.vcf}.bcf
-        echo "Converting $! to $OUT (and indexing)."
+        echo "Converting $1 to $OUT (and indexing)."
         bcftools convert -o $OUT -O b $1
         bcftools index $OUT
     fi

### 2017-10-08 sims/msp/summarize_run.Rmd
.

@@
 pop.cols <- rainbow_hcl(nlevels(samps$population))
-pop.pch <- seq_len(nlevels(samps$population))
+# max pch value is 25
+pop.pch <- seq_len(nlevels(samps$population)) %% 25
 plot( rep(1,length(pop.names)), seq_along(pop.names), 

### 2017-10-08 sims/msp/summarize_run.Rmd
.

--- 2017-10-08 sims/msp/summarize_run.Rmd
@@
 pop.names <- levels(samps$population)
+dim(pop.names) <- c(10,10)
 pop.cols <- rainbow_hcl(nlevels(samps$population))
-pop.pch <- seq_len(nlevels(samps$population))
-plot( rep(1,length(pop.names)), seq_along(pop.names), 
-       cex=4, pch=pop.pch, col=pop.cols, xlim=c(0,length(pop.names)),
+# max pch value is 25
+# pop.pch <- seq_len(nlevels(samps$population)) %% 25
+pop.pch <- rep(c(0:4,14:18),10)
+plot( row(pop.names), col(pop.names),
+       cex=4, pch=pop.pch, col=pop.cols,
        xlab='', ylab='', xaxt='n', yaxt='n' )

### 2017-10-08 sims/msp/summarize_run.Rmd
updated corner pca

@@
 corner.npc <- 4
-corner.regions <- lapply( 1:ncol(mds.corners), function (k) {
-            regions[ mds.corners[,k],]
-    } )
-corner.covmats <- lapply( 1:ncol(mds.corners), function (k) {
+corner.regions <- vector(mode='list', length=ncol(mds.corners))
+for (k in seq_along(corner.regions)) {
+    corner.regions[[k]] <- regions[ mds.corners[,k],]
+}
+corner.covmats <- vector(mode='list', length=ncol(mds.corners))
+for (k in seq_along(corner.covmats)) {
     reg <- regions[ mds.corners[,k], ]
     qfun <- multi_vcf_query_fn( chrom.list=chroms, file=bcf.files, regions=reg )
-            running_cov(qfun,1:nrow(reg))
-        } )
-corner.pca <- lapply( corner.covmats, function (covmat) {
-            cov_pca(covmat=covmat,k=corner.npc,w=opt$weights)
-    } )
+    corner.covmats[[k]] <- running_cov(qfun,1:nrow(reg), normalize.rows=TRUE)
+}
+corner.pca <- vector(mode='list', length=ncol(mds.corners))
+for (k in seq_along(corner.pca)) {
+    corner.pca[[k]] <- cov_pca(covmat=covmat, k=corner.npc, w=opt$weights)
+}
 ```

### 2017-10-08 lostruct/R/cov_pca.R

@@ -53,6 +53,8 @@ cov_pca <- function (x,k,w=1,
 #'
 #' @param f A function such that f(k) returns the k-th chunk of data.
 #' @param n The number of chunks of data, or a vector of indices.
+#' @param normalize.rows Whether to subtract row means.
+#"
 #' @return A numeric matrix:
 #' If a and b are vectors, then
 #'    mean(a[1:(n+m)]) == (n/(n+m)) * mean(a[1:n]) + (1/(n+m)) * sum(a[(n+1):(n+m)])
@@ -60,9 +62,12 @@ cov_pca <- function (x,k,w=1,
 #'    cov(a,b) = sum(a*b)/(n-1) - sum(a)*sum(b)/(n*(n-1))
 #' where the sums are over the n shared nonmissings.
 #' @export
-running_cov <- function (f,n) {
+running_cov <- function (f, n, normalize.rows=TRUE) {
     if (is.numeric(n) && length(n)==1) { n <- seq_len(n) }
     x <- f(n[1])
+    if (normalize.rows){
+        x <- sweep(x, 1, rowMeans(x,na.rm=TRUE))
+    }
     z <- !is.na(x)
     colm <- colMeans(x,na.rm=TRUE)  # estimate of the column means
     x <- sweep(x,2,colm,"-")

### 2017-12-21 writeup/review-responses.tex
moved over to new argrecorder set-up
latex wrangling

@@
 
+\reply{
+    Good point; we've cut this down.
+}
+
 \begin{point}{}
@@
 \end{point}
+
+\reply{
+    We greatly prefer named references rather than numbers, but
+    the \LaTeX source code is available at \url{https://github.com/petrelharp/local_pca} 
+    (run `make local\_pca\_paper.pdf' in the `writeup/' directory) 
+    -- the reviewer is welcome to change the formatting and recompile.
+}

### 2018-05-14 templated/summarize_run.Rmd
fix to template

--- 2017-08-08 templated/summarize_run.Rmd
@@
 corner.npc <- 4
-corner.regions <- lapply( 1:ncol(mds.corners), function (k) {
-            regions[ mds.corners[,k],]
-    } )
-corner.covmats <- lapply( 1:ncol(mds.corners), function (k) {
+corner.regions <- vector(mode='list', length=ncol(mds.corners))
+for (k in seq_along(corner.regions)) {
+    corner.regions[[k]] <- regions[ mds.corners[,k],]
+}
+corner.covmats <- vector(mode='list', length=ncol(mds.corners))
+for (k in seq_along(corner.covmats)) {
     reg <- regions[ mds.corners[,k], ]
     qfun <- multi_vcf_query_fn( chrom.list=chroms, file=bcf.files, regions=reg )
-            running_cov(qfun,1:nrow(reg))
-        } )
-corner.pca <- lapply( corner.covmats, function (covmat) {
-            cov_pca(covmat=covmat,k=corner.npc,w=opt$weights)
-    } )
+    corner.covmats[[k]] <- running_cov(qfun,1:nrow(reg), normalize.rows=TRUE)
+}
+corner.pca <- vector(mode='list', length=ncol(mds.corners))
+for (k in seq_along(corner.pca)) {
+    corner.pca[[k]] <- cov_pca(covmat=corner.covmats[[k]], k=corner.npc, w=opt$weights)
+}
 ```

### 2018-05-14 templated/summarize_run.Rmd
config missing error msg

@@
 ```{r data_setup, include=FALSE}
+if (!file.exists("config.json")) {
+    stop(paste("File", file.path(getwd(), "config.json"), "does not exist. Cannot continue."))
+}
 opt <- fromJSON("config.json")

### 2018-08-08 writeup/local_pca_paper.tex
latex wrangling
add sup figures

@@
 
+%% for submitting with responses to reviews
+\usepackage{lineno}
+
 \newif\ifsubmission
-% \submissiontrue
-\submissionfalse
+\submissiontrue
+% \submissionfalse
 \ifsubmission
@@
     \usepackage{setspace}
-    \doublespacing
-    \usepackage{lineno}
-    \linenumbers
+    % \doublespacing
     \bibliographystyle{plos2015}
@@
     \newcommand{\Figures}{{Figures }}
+    \renewcommand{\linelabel}[1]{}
 \fi
@@
 \DeclareMathOperator{\tr}{tr}
-
-%% for submitting with responses to reviews
-\usepackage{lineno}
 
@@
 \end{flushleft}
+
+\ifsubmission\linenumbers\fi
 
@@
     Descriptive statistics for each dataset used.
-    \revpoint{2}{18}
+    % \revpoint{2}{18}  %% no line numbers here; won't work
     % % with the chosen window sizes.
@@
 
+%%%%%%%%%%simulations
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_012720}
+    \end{center}
+    \caption{
+        MDS visualizations of simulation run_012720 (RECOMBTYPE, ``step"; SELTYPE, ``neutral")
+        \label{sfig:mds_run_012720}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_027034}
+    \end{center}
+    \caption{
+        MDS visualizations of simulation run_027034 (RECOMBTYPE, ``base"; SELTYPE, ``neutral")
+        \label{sfig:mds_run_027034}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_015598}
+    \end{center}
+    \caption{
+        MDS visualizations of simulation run_015598 (RECOMBTYPE, ``hotspot"; SELTYPE, ``neutral")
+        \label{sfig:mds_run_015598}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corner_pca-1_run_015598}
+    \end{center}
+    \caption{
+        PCA plots for regions colored in \Figure \ref{sfig:mds_run_015598}
+        \label{sfig:pca_corner_run_015598}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_005464}
+    \end{center}
+    \caption{
+        MDS visualizations of simulation run_005464 (RECOMBTYPE, ``base"; SELTYPE, ``selected")
+        \label{sfig:mds_run_005464}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_031486}
+    \end{center}
+    \caption{
+        MDS visualizations of simulation run_031486 (RECOMBTYPE, ``base"; SELTYPE, ``selected")
+        \label{sfig:mds_run_031486}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_009673}
+    \end{center}
+    \caption{
+        MDS visualizations of simulation run_009673 (RECOMBTYPE, ``step"; SELTYPE, ``selected")
+        \label{sfig:mds_run_009673}
+    }
+\end{figure}
+
 \ifsubmission\processdelayedfloats\fi

### 2018-08-08 writeup/local_pca_paper.tex
fixed up latex

@@
 \newif\ifsubmission
-\submissiontrue
-% \submissionfalse
+% \submissiontrue
+\submissionfalse
 \ifsubmission
@@
         \ifusepngs
-            \includegraphics{MDS_allchr_Together_plot_samples_noinv.png}
+            \includegraphics[width=\textwidth]{MDS_allchr_Together_plot_samples_noinv.png}
         \else
-            \includegraphics{MDS_allchr_Together_plot_samples_noinv}
+            \includegraphics[width=\textwidth]{MDS_allchr_Together_plot_samples_noinv}
         \fi
@@
     \caption{
-        MDS visualizations of simulation run_012720 (RECOMBTYPE, ``step"; SELTYPE, ``neutral")
+        MDS visualizations of simulation run 012720 (RECOMBTYPE, ``step"; SELTYPE, ``neutral")
         \label{sfig:mds_run_012720}
@@
     \caption{
-        MDS visualizations of simulation run_027034 (RECOMBTYPE, ``base"; SELTYPE, ``neutral")
+        MDS visualizations of simulation run 027034 (RECOMBTYPE, ``base"; SELTYPE, ``neutral")
         \label{sfig:mds_run_027034}
@@
     \caption{
-        MDS visualizations of simulation run_015598 (RECOMBTYPE, ``hotspot"; SELTYPE, ``neutral")
+        MDS visualizations of simulation run 015598 (RECOMBTYPE, ``hotspot"; SELTYPE, ``neutral")
         \label{sfig:mds_run_015598}
@@
     \caption{
-        MDS visualizations of simulation run_005464 (RECOMBTYPE, ``base"; SELTYPE, ``selected")
+        MDS visualizations of simulation run 005464 (RECOMBTYPE, ``base"; SELTYPE, ``selected")
         \label{sfig:mds_run_005464}
@@
     \caption{
-        MDS visualizations of simulation run_031486 (RECOMBTYPE, ``base"; SELTYPE, ``selected")
+        MDS visualizations of simulation run 031486 (RECOMBTYPE, ``base"; SELTYPE, ``selected")
         \label{sfig:mds_run_031486}
@@
     \caption{
-        MDS visualizations of simulation run_009673 (RECOMBTYPE, ``step"; SELTYPE, ``selected")
+        MDS visualizations of simulation run 009673 (RECOMBTYPE, ``step"; SELTYPE, ``selected")
         \label{sfig:mds_run_009673}

### 2018-08-09 writeup/local_pca_paper.tex
add missingness plots

@@
 
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_Drosophila_missings_against_MDS1_win103}
+    \end{center}
+    \caption{
+        The missingness ratio along genome and the correlation of MDS against missingness in Drosophila.
+        \label{sfig:dpgp_missingness}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_Medicago_missing_against_MDS1_win104}
+    \end{center}
+    \caption{
+        The missingness ratio along genome and the correlation of MDS against missingness in Medicago.
+        \label{sfig:medicago_missingness}
+    }
+\end{figure}
+
 \ifsubmission\processdelayedfloats\fi

### 2018-08-10 writeup/local_pca_paper.tex
tidied table

@@
       \hline
- & 10000snp, 2 PCs & 1000snp, 2 PCs & 10000snp, 5 PCs & 100000bp, 2 PCs & 10000bp, 2 PCs \\ 
+          & 10000 SNPs & 1000 SNPs & 10000 SNPs & 100000bp & 10000bp \\ 
+     MDS1 & 2 PCs & 2 PCs & 5 PCs & 2 PCs & 2 PCs \\ 
       \hline
-10000snp, 2 PCs & 1.00 & 0.87 & 0.96 & 0.90 & 0.88 \\ 
-  1000snp, 2 PCs & 0.68 & 1.00 & 0.73 & 0.68 & 0.94 \\ 
-  10000snp, 5 PCs & 0.96 & 0.92 & 1.00 & 0.88 & 0.93 \\ 
+  10000 SNPs, 2 PCs & 1.00 & 0.87 & 0.96 & 0.90 & 0.88 \\ 
+  1000 SNPs, 2 PCs & 0.68 & 1.00 & 0.73 & 0.68 & 0.94 \\ 
+  10000 SNPs, 5 PCs & 0.96 & 0.92 & 1.00 & 0.88 & 0.93 \\ 
   100000bp, 2 PCs & 0.90 & 0.87 & 0.88 & 1.00 & 0.87 \\ 
@@
        \hline
-\end{tabular}
-\end{table}
-
-[[2]]
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Wed Feb  8 16:17:12 2017
-\begin{table}[ht]
-\centering
-    \begin{tabular}{rrrrrr}
        \hline
-     MDS1 & 10000 SNPs, 2 PCs & 1000 SNPs, 2 PCs & 10000 SNPs, 5 PCs & 100000bp, 2 PCs & 10000bp, 2 PCs \\ 
+        MDS2 & &&&&\\
       \hline
@@
       10000 SNPs, 5 PCs & 0.93 & 0.50 & 1.00 & 0.83 & 0.52 \\ 
-      100000bp, 2 PCs & 0.87 & 0.59 & 0.84 & 1.00 & 0.58 \\ 
-      10000bp, 2 PCs & 0.83 & 0.92 & 0.77 & 0.84 & 1.00 \\ 
-       \hline
-    \end{tabular}
-    \begin{tabular}{rrrrrr}
-      \hline
-     MDS2 & 10000snp, 2 PCs & 1000snp, 2 PCs & 10000snp, 5 PCs & 100000bp, 2 PCs & 10000bp, 2 PCs \\ 
-      \hline
-      10000snp, 2 PCs & 1.00 & 0.54 & 0.93 & 0.87 & 0.56 \\ 
-      1000snp, 2 PCs & 0.82 & 1.00 & 0.76 & 0.83 & 0.92 \\ 
-      10000snp, 5 PCs & 0.93 & 0.50 & 1.00 & 0.83 & 0.52 \\ 
       100000bp, 2 PCs & 0.87 & 0.59 & 0.84 & 1.00 & 0.58 \\ 

### 2018-08-16 sims/example/simple_example.Rmd
simple examples

---
title: "Simple simulation"
date: "`r date()`"
---

```{r setup, include=FALSE}
library(lostruct)
library(colorspace)
library(jsonlite)
library(RColorBrewer)
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
# set do.pdfs to TRUE to output PDFs of figures as well
if (!exists("do.pdfs")) { do.pdfs <- FALSE }
```
```{r plot_setup, include=FALSE}
layout_heights <- function (k,dl=0,ncol=1) {
    # to set up layout without 'dl' lines between plots
    # use like layout(1:5,heights=layout_heights(5))
    if (k==1) return(1)
    layout(matrix(seq_len(k*ncol),ncol=ncol))  # this changes par("csi")
    ds <- dl*par("lheight")*par("csi")
    eps=par("mai")[c(1,3)]
    dh=(par("din")[2]-sum(eps)-(k-1)*ds)/k
    return(c(eps[2]+dh+ds/2,rep(dh+ds,k-2),eps[1]+dh+ds/2)/par("din")[2])
}
pdf_copy <- function (
                width=6,
                height=width*knitr::opts_current$get("fig.height")/knitr::opts_current$get("fig.width"),
                plot.id=NULL,
                filename
                ) {
    if (missing(filename)) {
        file.id <- if (is.null(plot.id)) { "" } else { paste0("_",plot.id) }
        filename <- knitr::fig_path(paste(file.id,".pdf",sep=""))
    }
    cat("pdf version at:",filename)
    dev.print( file=filename, device=pdf, 
            width=width, height=height,
            pointsize=10,
            family="sans")
}
```

Let's suppose there are three populations,
and on the first quarter of the chromosome,
the first two are close to each other;
on the middle hald, the last two are close,
and on the last quarter the first and the last are.

```{r simple_data, cache=TRUE}
base_covmat <- matrix(c(1.0, 0.7, 0.0,
                        0.7, 1.0, 0.0,
                        0.0, 0.0, 1.0), nrow=3)
pop_covmats <- list(base_covmat, 
                    base_covmat[3:1, 3:1], 
                    base_covmat[c(1,3,2), c(1,3,2)])
struct_breakpoints <- c(0.0, 0.25, 0.5, 0.75, 1.0)
struct_indices <- c(1,2,2,3)
nsamples <- c(50, 50, 50)
ind_covmat <- diag(sum(nsamples))
for (k in seq_along(nsamples)) {
    ii <- sum(nsamples[seq_len(k-1)]) + 1:nsamples[k] 
    ind_covmat[ii, ii] <- rexp(nsamples[k]^2, 100)
    ind_covmat[ii, ii] <- ind_covmat[ii, ii] + t(ind_covmat[ii, ii])
}
pop_chols <- lapply(lapply(pop_covmats, chol),
                    function (x) { 
                        (x[rep(1:3, nsamples), rep(1:3, nsamples)] / sqrt(rep(nsamples, nsamples))
                         + ind_covmat)
                    })
nwindows <- 99
nloci <- 100
win.fn <- function (n, ...) {
    geno <- matrix(rnorm(sum(nsamples) * nloci), nrow=nloci)
    cholmat <- pop_chols[[struct_indices[findInterval(n/nwindows, struct_breakpoints, rightmost.closed=TRUE)]]]
    return( geno %*% cholmat )
}
attr(win.fn, "max.n") <- nwindows
attr(win.fn,"samples") <- paste0(rep(letters[seq_along(nsamples)], nsamples),
                                 unlist(lapply(nsamples, seq_len)))
attr(win.fn,"region") <- function (n) { 
    data.frame( chrom="matrix", start=((n-1)*nloci+1), end=n*nloci ) }
all.pcas <- eigen_windows(win.fn, k=2, do.windows=1:nwindows)
regions <- region(win.fn)()
all.lengths <- c("chrom1"=nwindows)
pc.distmat <- pc_dist( all.pcas, npc=2 )
na.inds <- is.na( all.pcas[,1] ) # there may be windows with missing data
mds.coords <- cbind( data.frame( 
                        chrom=rep(names(all.lengths),all.lengths),
                        window=unlist(lapply(all.lengths,seq_len)) ),
                        cmdscale( pc.distmat[!na.inds,!na.inds], k=2 )[ ifelse( na.inds, NA, cumsum(!na.inds) ), ] )
mds.cols <- (1:ncol(mds.coords))[-(1:2)]
```

OK, now we plot them:
```{r plotting_setup, echo=FALSE}
chroms <- unique(regions$chrom)
chrom.starts <- tapply( regions$start, regions$chrom, min, na.rm=TRUE )
chrom.ends <- tapply( regions$end, regions$chrom, max, na.rm=TRUE )
chrom.spacing <- floor(.05*mean(chrom.ends))
chrom.offsets <- c(0,cumsum(chrom.spacing+chrom.ends))
names(chrom.offsets) <- c(chroms,"end")
chrom.dividers <- c(0,chrom.offsets[-1])-chrom.spacing/2
chrom.mids <- chrom.dividers[-1] - diff(chrom.dividers)/2

# this is where to plot windows at when plotting with all chromosomes
regions$pos <- chrom.offsets[match(regions$chrom,chroms)]+(regions$start+regions$end)/2

chrom.cols <- rainbow_hcl(length(chroms), c=90, end=.9*360)[as.numeric(regions$chrom)]

#' Set up to plot all chromosomes together
#' will plot the vector of values 'y' along all the chromosomes.
chrom.plot <- function (y,ylab='',main='',chrom.labels=TRUE,...) {
    plot(0, type='n', xlim=range(chrom.offsets/1e6), ylim=range(y,finite=TRUE), 
         xlab='', xaxt='n', yaxt='n', ylab=ylab, main=main)
    if (length(chroms)>1) for (k in 1:floor(length(chroms)/2)) {
        rect( xleft=chrom.dividers[2*k-1]/1e6, xright=chrom.dividers[2*k]/1e6, 
             ybottom=par("usr")[3], ytop=par("usr")[4], 
             border=NA, col=adjustcolor("grey",0.25) )
    }
    abline( v=chrom.dividers/1e6, lty=3, col=adjustcolor("grey",0.5), lwd=2 )
    if (chrom.labels) axis( 1, at=chrom.mids/1e6, labels=paste("chromosome", chroms), las=0, tick=FALSE )
    points( regions$pos/1e6, y, ...)
}

pairs( mds.coords[,mds.cols], pch=20, col=adjustcolor(chrom.cols,0.75) )

plot_everything <- function () {
    mds.corners <- corners( mds.coords[,mds.cols[1:2]], prop=.05 )
    corner.cols <- brewer.pal(3,"Dark2")
    corner.pch <- c(15,17,19)
    ccols <- rep("black",nrow(mds.coords))
    cpch <- rep(20,nrow(mds.coords))
    for (k in 1:ncol(mds.corners)) {
        ccols[ mds.corners[,k] ] <- corner.cols[k]
        cpch[ mds.corners[,k] ] <- corner.pch[k]
    }
    # centroids of the corners in MDS space
    corner.mds <- do.call(rbind, lapply(1:ncol(mds.corners), 
                    function (ii){
                        colMeans(mds.coords[mds.corners[,ii],-(1:2)])
                    } ) )

    spacing <- 1
    opar <- par(mar=c(4,4,2,1)+.1,mgp=c(2.5,0.8,0))
    layout(matrix(c(rep(1,length(mds.cols)),1+seq_along(mds.cols)),ncol=2),
            widths=c(1,2), heights=layout_heights(length(mds.cols),dl=spacing,ncol=2))
    plot( mds.coords[,mds.cols[1:2]], pch=cpch, 
         col=adjustcolor(ccols,0.75),  asp=1,
         xaxt='n', yaxt='n',
         xlab="MDS coordinate 1", ylab="MDS coordinate 2" )
    points( corner.mds, pch=20, cex=5,
            col=adjustcolor(corner.cols,0.25))
    text( corner.mds, labels=seq_len(nrow(corner.mds)), 
            col=corner.cols, cex=2, lwd=2 )
    opar2 <- par(mar=c(par("mar"),spacing/2)[c(5,2,3,4)])
    for (k in mds.cols) {
        lastone <- (k==mds.cols[length(mds.cols)])
        if (lastone) { par(mar=c(par("mar"),opar2$mar[1])[c(5,2,3,4)]) }
        chrom.plot( mds.coords[,k], pch=20, 
                xlab=if (lastone) { "Position (Mb)"} else { "" }, # main=paste("MDS coordinate",match(k,mds.cols)),
                chrom.labels=lastone,
                ylab=colnames(mds.coords)[k],
                col=adjustcolor(ccols,0.75) )
        # do this for all but first
        par(mar=c(par("mar"),spacing/2)[c(1,2,5,4)])
    }
    par(opar)
    pdf_copy()
}
```
```{r base_case}
plot_everything()
```

## With missing data

Now we'll do the same thing, but with 50% missing data
on the first half of the chromosome.
```{r missing_data, cache=TRUE, depends="simple_data"}
win.fn <- function (n, ...) {
    geno <- matrix(rnorm(sum(nsamples) * nloci), nrow=nloci)
    cholmat <- pop_chols[[struct_indices[findInterval(n/nwindows, struct_breakpoints, rightmost.closed=TRUE)]]]
    out <- geno %*% cholmat
    if (n < nwindows/2) {
        out[rbinom(length(out), 1, 0.5)==1] <- NA
    }
    return(out)
}
attr(win.fn, "max.n") <- nwindows
attr(win.fn,"samples") <- paste0(rep(letters[seq_along(nsamples)], nsamples),
                                 unlist(lapply(nsamples, seq_len)))
attr(win.fn,"region") <- function (n) { 
    data.frame( chrom="matrix", start=((n-1)*nloci+1), end=n*nloci ) }
all.pcas <- eigen_windows(win.fn, k=2, do.windows=1:nwindows)
pc.distmat <- pc_dist( all.pcas, npc=2 )
na.inds <- is.na( all.pcas[,1] ) # there may be windows with missing data
mds.coords <- cbind( data.frame( 
                        chrom=rep(names(all.lengths),all.lengths),
                        window=unlist(lapply(all.lengths,seq_len)) ),
                        cmdscale( pc.distmat[!na.inds,!na.inds], k=2 )[ ifelse( na.inds, NA, cumsum(!na.inds) ), ] )
mds.cols <- (1:ncol(mds.coords))[-(1:2)]

plot_everything()
```

## With very different sample sizes

Now we'll do the same thing, but with very different sample sizes.
```{r sample_diffs, cache=TRUE, depends="simple_data"}
nsamples <- c(900, 10, 10)
ind_covmat <- diag(sum(nsamples))
for (k in seq_along(nsamples)) {
    ii <- sum(nsamples[seq_len(k-1)]) + 1:nsamples[k] 
    ind_covmat[ii, ii] <- rexp(nsamples[k]^2, 15)
    ind_covmat[ii, ii] <- ind_covmat[ii, ii] + t(ind_covmat[ii, ii])
}
pop_chols <- lapply(lapply(pop_covmats, chol),
                    function (x) { 
                        (x[rep(1:3, nsamples), rep(1:3, nsamples)] / sqrt(rep(nsamples, nsamples))
                         + ind_covmat)
                    })
nwindows <- 99
nloci <- 50
win.fn <- function (n, ...) {
    geno <- matrix(rnorm(sum(nsamples) * nloci), nrow=nloci)
    cholmat <- pop_chols[[struct_indices[findInterval(n/nwindows, struct_breakpoints, rightmost.closed=TRUE)]]]
    return( geno %*% cholmat )
}
attr(win.fn, "max.n") <- nwindows
attr(win.fn,"samples") <- paste0(rep(letters[seq_along(nsamples)], nsamples),
                                 unlist(lapply(nsamples, seq_len)))
attr(win.fn,"region") <- function (n) { 
    data.frame( chrom="matrix", start=((n-1)*nloci+1), end=n*nloci ) }
all.pcas <- eigen_windows(win.fn, k=2, do.windows=1:nwindows)
all.lengths <- c("chrom1"=nwindows)
pc.distmat <- pc_dist( all.pcas, npc=2 )
na.inds <- is.na( all.pcas[,1] ) # there may be windows with missing data
mds.coords <- cbind( data.frame( 
                        chrom=rep(names(all.lengths),all.lengths),
                        window=unlist(lapply(all.lengths,seq_len)) ),
                        cmdscale( pc.distmat[!na.inds,!na.inds], k=2 )[ ifelse( na.inds, NA, cumsum(!na.inds) ), ] )
mds.cols <- (1:ncol(mds.coords))[-(1:2)]

plot_everything()
```

And, *reweighted*:
```{r reweighted, cache=TRUE, depends="sample_diffs"}
all.pcas <- eigen_windows(win.fn, k=2, do.windows=1:nwindows, w=1/rep(sqrt(nsamples),nsamples))
all.lengths <- c("chrom1"=nwindows)
pc.distmat <- pc_dist( all.pcas, npc=2, w=1/rep(sqrt(nsamples),nsamples))
na.inds <- is.na( all.pcas[,1] ) # there may be windows with missing data
mds.coords <- cbind( data.frame( 
                        chrom=rep(names(all.lengths),all.lengths),
                        window=unlist(lapply(all.lengths,seq_len)) ),
                        cmdscale( pc.distmat[!na.inds,!na.inds], k=2 )[ ifelse( na.inds, NA, cumsum(!na.inds) ), ] )
mds.cols <- (1:ncol(mds.coords))[-(1:2)]

plot_everything()
```

### 2018-08-16 templated/summarize_run.Rmd
simple examples

@@
 } else {
-    warning("No populatio information in the sample file, %s.", samp.file)
+    warning("No population information in the sample file, %s.", samp.file)
     samps <- data.frame( 

### 2018-08-16 sims/example/simple_example.Rmd
adjustments to simple tests

@@
 
+**Note:** to copy the resulting figures to the `writeup/` directory:
+```
+for x in figure/simple_example/*pdf; do mv $x "../../writeup/simple_example_$(basename ${x%-1.pdf}.pdf)"; done
+```
+
 Let's suppose there are three populations,
@@
 ```{r simple_data, cache=TRUE}
-base_covmat <- matrix(c(1.0, 0.7, 0.0,
-                        0.7, 1.0, 0.0,
+base_covmat <- matrix(c(0.75, 0.25, 0.0,
+                        0.25, 0.75, 0.0,
                         0.0, 0.0, 1.0), nrow=3)
@@
 nsamples <- c(50, 50, 50)
-ind_covmat <- diag(sum(nsamples))
-for (k in seq_along(nsamples)) {
-    ii <- sum(nsamples[seq_len(k-1)]) + 1:nsamples[k] 
-    ind_covmat[ii, ii] <- rexp(nsamples[k]^2, 100)
-    ind_covmat[ii, ii] <- ind_covmat[ii, ii] + t(ind_covmat[ii, ii])
-}
+# ind_covmat <- diag(sum(nsamples))
+# for (k in seq_along(nsamples)) {
+#     ii <- sum(nsamples[seq_len(k-1)]) + 1:nsamples[k] 
+#     ind_covmat[ii, ii] <- rexp(nsamples[k]^2, 100)
+#     ind_covmat[ii, ii] <- ind_covmat[ii, ii] + t(ind_covmat[ii, ii])
+# }
+ind_covmat <- 0
 pop_chols <- lapply(lapply(pop_covmats, chol),
@@
 nwindows <- 99
-nloci <- 100
+nloci <- 400
 win.fn <- function (n, ...) {
@@
 mds.cols <- (1:ncol(mds.coords))[-(1:2)]
-
 plot_everything()
@@
 nwindows <- 99
-nloci <- 50
+nloci <- 400
 win.fn <- function (n, ...) {

### 2018-08-16 writeup/local_pca_paper.tex
adjustments to simple tests

@@
 
+\paragraph{Testing}
+Figure \ref{fig:test_missing} shows tests with lots of missing data.
+Also the covariance matrix was chosen to have the top two eigenvalues equal,
+so that eigenvector switching is expected.
+
 \subsection{Visualization of results}
@@
 to find only the top $k$ eigenvectors.
+
+\paragraph{Testing}
+To demonstrate the utilitiy of this method XXX
+see Figure \ref{fig:test_weighted_pca}.
 
@@
 
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=0.82\textwidth]{simple_example_base_case}
+        \includegraphics[width=0.82\textwidth]{simple_example_missing_data}
+    \end{center}
+    \caption{
+        From simple example -- base case; and with 50\% missing data.
+    } \label{fig:test_missing}.
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=0.82\textwidth]{simple_example_sample_diffs}
+        \includegraphics[width=0.82\textwidth]{simple_example_reweighted}
+    \end{center}
+    \caption{
+        From simple example -- different sample sizes; unweighted (top); weighted by $1/\sqrt{n_i}$ (bottom).
+    } \label{fig:test_weighted_pca}.
+\end{figure}
+
 \ifsubmission\processdelayedfloats\fi

### 2018-08-21 writeup/local_pca_paper.tex
run_016913 simulation added
discussion of tests

@@
 
-\paragraph{Testing}
-Figure \ref{fig:test_missing} shows tests with lots of missing data.
-Also the covariance matrix was chosen to have the top two eigenvalues equal,
-so that eigenvector switching is expected.
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Visualization of results}
@@
 and performing PCA on all, aggregated, regions.
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{Testing}
+
+We tested the method using two types of simulation.
+First, to verify expected behavior, we simulated ``genomes'' as an independent sequence
+of correlated Gaussian ``genotypes'', using a different covariance matrix
+in the first quarter, middle half, and last quarter of the chromosome.
+To verify robustness to missing data,
+we ran the method after randomly dropping 50\% of the genotypes in the first half of the genome;
+if the method is misled by missing data, then it will distinguish the two halves of the chromosome
+rather than the segments having different covariance matrices.
+Details are given in Appendix \ref{apx:gaussian_sims}.
+
+To provide a realistic test,
+we next used forwards-time, individual-based simulations, implemented using SLiM v3 \citep{slim_v3},
+which are described in detail in Appendix \ref{apx:slim_sims}.
+To provide realistic population structure for PCA to identify,
+each simulation ahd at least 5,000 diploid individuals,
+living across a continuous square range,
+with Gaussian dispersal and local density-dependent competition.
+Each genome was modeled on human chromosome 7, which is $1.54 \times 10^8$bp long,
+with an overall recombination rate of 1.6785 crossovers per chromosome per generation.
+To improve speed, we used tskit \citep{tskit_paper} to record tree sequences
+and add neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
+Most simulations were neutral, but we also included linked selection, of two types.
+First, % run_031486
+we introduced selected mutations into two regions, 
+which extended from 1/3 to 1/2 and from 5/6 to the end of the genome respectively.
+These had selection coefficients from a Gamma distribution with shape 2 and mean 0.005
+at a rate of $10^{-10}$ per bp, that were either beneficial (with probability $1/30$)
+or deleterious (otherwise).
+Second, % run_016913
+to roughly model a recent expansion followed by local adaptation,
+we introduced mutations in the same manner as above,
+except that mutations were no longer unconditionally deleterious or beneficial:
+each selection coefficient was multiplied by a factor depending on the spatial location
+of the individual being evaluated, varying linearly from -1 at the left side of the range
+to +1 at the right edge.
+In all simulations, genome-wide PCA displayed a map of the population range, as expected.
 
@@
 
-Address
+\plr{Address:
     mutation rate variation;
@@
     variation in missingness;
-maybe differing sample sizes and reweighting.
+    maybe differing sample sizes and reweighting.  }
+
+Simple non-population-based simulations with Gaussian ``genotypes''
+showed that the method performs as expected, clearly separating regions of the genome
+with different underlying covariance matrices
+without being affected by extreme differences in amount of missing data
+(Supplemental Figure \ref{fig:test_missing}).
+This simulation also verifies insensitivity to ordering of top PCs,
+since it was performed using a covariance matrix with the top two eigenvalues equal,
+so that the order of empirical eigenvectors (PCs) switches randomly.
+The same type of simulations with dramatically different sample sizes
+showed that using weighted PCA substantially improved identification of different genomic regions
+(Supplemental Figure \ref{fig:test_weighted_pca}).
+
+Individual-based simulations using SLiM \citep{slim_v3} allowed us to test the effects
+of recombination and mutation rate variation, as well as linked selection.
+Results are shown in Supplemental Figures \ref{sfig:slim_sims1} and \ref{sfig:slim_sims2}.
+As expected, varying recombination rate stepwise by a factor of 64
+did not induce patterns in the MDS visualizations correlated with recombination rate.
+Since varying mutation rate with a fixed recombination map
+is equivalent to varying the recombination map and remapping windows,
+we used the same simulations to show that variation in mutation rate similarly has no effect.
+On the other hand,
+a recombination map with hotspots
+(the HapMap human female map for chromosome 7 \citep{hapmap})
+induced outliers at long regions of low recombination rate
+(also as expected).
+
+Simulations with linked selection produced mixed results
+(Figure \ref{sfig:linked_selection}).
+The method strongly identified the regions under spatially varying linked selection.
+It also identified the regions (although less unambiguously)
+with constant selection and stepwise varying recombination rate,
+but did not clearly identify them with constant recombination rate.
+This difference is likely because recombination rates are overall lower in the first case,
+leading to a stronger effect of linked seleciton.
+These tests are not meant to be comprehensive survey of linked selection,
+but only to demonstrate that linked selection can produce signals simlar to what we see in real data.
 
@@
     \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_027034}
        \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_012720}
-    \end{center}
-    \caption{
-        MDS visualizations of simulation run 012720 (RECOMBTYPE, ``step"; SELTYPE, ``neutral")
-        \label{sfig:mds_run_012720}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_027034}
-    \end{center}
-    \caption{
-        MDS visualizations of simulation run 027034 (RECOMBTYPE, ``base"; SELTYPE, ``neutral")
-        \label{sfig:mds_run_027034}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
        \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_015598}
-    \end{center}
-    \caption{
-        MDS visualizations of simulation run 015598 (RECOMBTYPE, ``hotspot"; SELTYPE, ``neutral")
-        \label{sfig:mds_run_015598}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
        \includegraphics[width=0.82\textwidth]{FigS_plot_corner_pca-1_run_015598}
+    \end{center}
+    \caption{
+        MDS visualizations of the results of individual-based simulations using SLiM
+        (see Appendix \ref{apx:slim_sims} for details).
+        All simulations are neutral, and recombination is:
+        \textbf{(top)} constant;
+        \textbf{(top middle)} varies stepwise by factors of two in seven equal-length segments,
+        with highest rates on the ends, so the middle segment has a recombination rate 64 times lower
+        than the ends;
+        \textbf{(bottom middle)} according to the HapMap human female chromosome 7 map.
+        The \textbf{bottom} figure shows PCA maps corresponding to the three colored windows
+        of the last (HapMap) situation;
+        the outlying regions are long regions of low recombination rate,
+        so that region can be dominated by a few correlated trees,
+        similar to an inversion.
+    } \label{sfig:slim_sims1}
+\end{figure}
+
+\begin{figure}
+    \begin{center}
     \end{center}
@@
     \caption{
+        \plr{KEEP}
         MDS visualizations of simulation run 031486 (RECOMBTYPE, ``base"; SELTYPE, ``selected")
@@
     \caption{
+        \plr{KEEP}
         MDS visualizations of simulation run 009673 (RECOMBTYPE, ``step"; SELTYPE, ``selected")
@@
     \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_016913}
+    \end{center}
+    \caption{
+        MDS visualizations of simulation run 009673 (RECOMBTYPE, ``base"; SELTYPE, ``balancing")
+        \label{sfig:mds_run_016913}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corner_pca-1_run_016913}
+    \end{center}
+    \caption{
+        PCA plots for regions colored in \Figure \ref{sfig:mds_run_016913}
+        \label{sfig:pca_corner_run_016913}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
        \includegraphics[width=0.82\textwidth]{FigS_Drosophila_missings_against_MDS1_win103}

### 2018-08-22 writeup/review-responses.tex
first pass at responding to all comments

--- 2017-07-18 writeup/review-responses.tex
@@
 \begin{point}{}
-\ldots variation in the recombination rate \ldots creating spurious structure.
+    \ldots (a) variation in the recombination rate and 
+    (b) variation in the mutation rate \ldots creating spurious structure.
 \end{point}
@@
 \reply{
-    We address this in two ways.
-    This point is addressed by comparing results with windows of different types --
+    Since we only look at SNPs,
+    a 1Kb window with mutation rate $\mu$ and recombination rate $\rho$
+    is equivalent to a 10Kb window with mutation rate $\mu/10$ and recombination rate $\rho/10$,
+    and the ratio $\mu/\rho$ determines the ``signal-to-noise'' ratio.
+    For this reason, varying recombination rate in simulations,
+    and varying window size and type in analysis
+    suffices to check for the effects of both recombination and mutation rate.
+    We have now added simulation tests showing that large-scale variation in recombination rate
+    does not create spurious structure \llname{ll:test_recomb_rate};
+    except in the presence of long regions of low recombination,
+    as expected \llname{ll:hotspot_recomb}
+    The point is addressed by comparing results with windows of different types --
     windows of equal length in bp (or in SNPs) have different lengths in cM;
-    since these different choices show nearly identical patterns,
-    recombination rate variation cannot be driving the results.
-}
-
-\begin{point}{}
-\ldots and variation in the mutation rate, creating spurious structure.
-\end{point}
-
-\reply{
-
+    since these different choices show nearly identical patterns in all cases we have examined,
+    recombination rate variation cannot be driving the results
+    (e.g., compare Figures \ref{fig:mds_allchr} and Supplemental Figure \plr{XXX}).
 }
@@
     and added more explicit notes about this to \llname{ll:another_pc_point} and \revref.
+    Furthermore, one of our simulation tests
+    is explicitly designed to test for this effect. \llname{ll:pc_switching_test}
 }
@@
 
+\reply{
+    We have added a representative figure (for the Drosophila data, $k=5$),
+    see Figure \ref{fig:dmel_k5}. \revref
+}
+
 \begin{point}{p15}
@@
 
+\reply{
+    We have added this: see supplemental figure \ref{sfig:mds_medicago_allchr_1e3}. \revref
+}
+
 \begin{point}{p15}
@@
 \end{point}
+
+\reply{
+    Again, we have added this: see supplemental figure \plr{XXX}. \revref
+}
 
@@
 \reply{
-    Good point: we now reference this work \revref.
+    Good point: we now reference this work and discuss it's similarity. \revref
 }
@@
 \end{quote}
+
+Thanks! We're glad you enjoyed it.
 
@@
     is only of limited usefulness
-    without also knowing the corresponding populaion variance.
+    without also knowing the corresponding population variance.
 }
@@
 
+\reply{
+    We have added a number of simulations testing various aspects of the method,
+    and have tried to incorporate the results in the paper without cluttering it up
+    with the results of sanity checks.
+    See sections \ref{ss:testing_methods} and \ref{ss:validation};
+    in particular the Gaussian simulations described in appendix \ref{apx:gaussian_sims}.
+}
+
 \begin{point}{}
@@
 
+\reply{
+    We felt that weighted PCA was a nice but unnecessary complication to this work,
+    so have removed it entirely.
+}
+
 \begin{point}{}
@@
 may skew the resulting PCs by selecting different sample sizes for the different windows (as
-discussed in Appendix B) . This could distort the PCs, so that variation reflects underlying
+discussed in Appendix B). This could distort the PCs, so that variation reflects underlying
 variation in missingness. Would be good to discuss this potential issue and provide simulations.
 \end{point}
+
+\reply{
+    We have tested the method against a large quantity of missing data
+    whereby the amount of missingness does not vary by population \llname{ll:test_missing}.
+    However, missingness that is structured by population could absolutely cause variation in PCs.
+    This could happen, for instance, with RAD data wherein allelic dropout is driven by
+    population-specific polymorphism in the restriction site.
+    The same problem to a lesser degree would be caused by mapping bias,
+    and in fact we see windows around the breakpoints of the polymorphic inversions
+    of the DPGP data having higher rates of missingness.
+    We definately would like our method to be insensitive to missing data,
+    but if missingness is in fact correlated with changes in patterns of relatedness (as in these cases),
+    correlation of MDS scores with amount of missingness is not in fact a problem.
+    As we don't think this point is particularly helpful to the reader, 
+    we have provided a plot of missingness against MDS score (Figure \ref{sfig:dpgp_missingness})
+    and mentioned in the text \revref,
+    but have not added a lengthy discussion.
+}
 
@@
 \end{point}
+
+\reply{
+    That's true, but we are unaware of a better option,
+    and since the jackknife is only used to pick an appropriate window size,
+    it does not seem crucial.  \revref
+    Since results are the same across different window sizes,
+    we don't think this can explain a correlation wtih recombination rate.
+}
 
@@
 \reply{
-    This is a good point, but XXX
-    We now cite PCAdmix \revreffull{1}{8}.
+    This is a good point, but any \emph{neutral} variation in local ancestry
+    is not expected to vary systematically along the chromosome,
+    as we now discuss more explicitly \llname{ll:not_neutral}.
+    It well could be that the patterns we see are due to selection acting differentially
+    on introgressed segments,
+    a scenario that would fall under ``linked selection''.
+    We discuss this in the Discussion \revref.
+    We also now cite and discuss the differences to PCAdmix \revreffull{1}{7}.
 }
@@
 
+\reply{
+    We agree that demonstrating this issue would be nice in principle,
+    but for this audience would be getting too far down in the weeds,
+    and including the requested simulations would inflate an already lengthy paper.
+    Additionally, we aim to provide a good method, not to prove that the method
+    is better than all other possible methods.
+    For the reviewer's sake, here is some more about our rationale.
+    Suppose that the covariance matrix is the same along the whole genome,
+    except that it is multiplied by an overall constant that varies.
+    Since PCs are scaled to have norm 1, this overall scaling will indeed not affect the PCs,
+    but we don't compare the PCs directly -- we compare the low-dimensional approximations
+    to the covariance matrix, and the scaling we propose would remove the effect
+    of this overall constant.
+    It is not clear that variation in mutation rate would act like an overall multiple
+    of the covariance matrix,
+    it since we're interested in patterns rather than magnitudes of relatedness,
+    it seems a wise thing to remove.
+}
+
 \begin{point}{Figure 7:}
-are MDS coordinates correlated with recombination rates in this case?
+    Are MDS coordinates correlated with recombination rates in this case?
 \end{point}
@@
 \begin{point}{Application:}
- is what the authors seem to be proposing not already accounted for by linear
+Is what the authors seem to be proposing not already accounted for by linear
 mixed model association approaches? If not, this should be clarified. Either way, this paragraph
@@
 \end{point}
+
+\reply{
+    It is not accounted for; we've added some more explanation to this section
+    (which we think should stay in). \revref
+}
 
@@
 could be seen as such within a coalescent framework, that seems to be just a useful
-representation. The authors may consider rewording`.
+representation. The authors may consider rewording.
 \end{point}
+
+\reply{
+    Good point, but we think that the sense of unease is perhaps desireable here:
+    certainly differential survival is an aspect of demography;
+    but as the rest of the paragraph goes on to discuss,
+    this quickly spills over into selection, which probably should not be included.
+    We've opted to keep this. \revref
+}
 
@@
 \end{point}
+
+\reply{
+    We've rewritten this section;
+    hopefully it is clearer. \revref
+}
 
@@
 
+\reply{
+    We agree, that further investigation into these patterns and their correlates
+    would be very interesting and fruitful;
+    however, we think it's outside of the scope of this paper.
+}
+
 \begin{point}{Page 13:}
@@
 
+\reply{Fixed. \revref}
+
 \begin{point}{}
-Typo in abstract, line 6 ``, We show'' -> ``. We show''.
+Typo in abstract, line 6 ``, We show'' $\rightarrow$ ``. We show''.
 \end{point}
+
+\reply{Fixed.}
 
@@
 
+\reply{
+    We have removed this sentence.
+}
+
 \begin{point}{}
@@
 \end{point}
+
+\reply{
+    We have added some relevant recent citations,
+    but apologize for not being aware of the appropriate recent review to refer to.
+    \revref
+}
 
@@
 
+\reply{
+    We have included the plots for Drosphila data using $k=5$:
+    see Figure \ref{fig:dmel_k5},
+    and hope that the readers will take our word on this point more generally.
+}
+
 \begin{point}{}
@@
 \end{point}
+
+\reply{
+    We've made these changes.
+}

### 2018-08-22 sims/example/simple_example.Rmd
first pass at responding to all comments

@@
 if (!exists("do.pdfs")) { do.pdfs <- FALSE }
+set.seed(23)
 ```
@@
                         0.0, 0.0, 1.0), nrow=3)
+nugget <- 0.1
 pop_covmats <- list(base_covmat, 
@@
 nsamples <- c(50, 50, 50)
-# ind_covmat <- diag(sum(nsamples))
-# for (k in seq_along(nsamples)) {
-#     ii <- sum(nsamples[seq_len(k-1)]) + 1:nsamples[k] 
-#     ind_covmat[ii, ii] <- rexp(nsamples[k]^2, 100)
-#     ind_covmat[ii, ii] <- ind_covmat[ii, ii] + t(ind_covmat[ii, ii])
-# }
-ind_covmat <- 0
-pop_chols <- lapply(lapply(pop_covmats, chol),
-                    function (x) { 
-                        (x[rep(1:3, nsamples), rep(1:3, nsamples)] / sqrt(rep(nsamples, nsamples))
-                         + ind_covmat)
-                    })
+pop_chols <- lapply(pop_covmats, function (X) {
+                        Y <- X[rep(1:3, nsamples), rep(1:3, nsamples)] 
+                        Y <- Y + nugget * diag(nrow(Y))
+                        return(chol(Y))
+                    } )
 nwindows <- 99
@@
 Now we'll do the same thing, but with very different sample sizes.
-```{r sample_diffs, cache=TRUE, depends="simple_data"}
+```{r sample_diffs, depends="simple_data"}
 nsamples <- c(900, 10, 10)
@@
     ii <- sum(nsamples[seq_len(k-1)]) + 1:nsamples[k] 
-    ind_covmat[ii, ii] <- rexp(nsamples[k]^2, 15)
-    ind_covmat[ii, ii] <- ind_covmat[ii, ii] + t(ind_covmat[ii, ii])
+    this_covmat <- nugget * diag(nsamples[k])
+    this_covmat[abs(row(this_covmat)-col(this_covmat)) <= 1] <- nugget
+    ind_covmat[ii, ii] <- this_covmat
 }
-pop_chols <- lapply(lapply(pop_covmats, chol),
-                    function (x) { 
-                        (x[rep(1:3, nsamples), rep(1:3, nsamples)] / sqrt(rep(nsamples, nsamples))
-                         + ind_covmat)
-                    })
+pop_chols <- lapply(pop_covmats, function (X) {
+                        Y <- X[rep(1:3, nsamples), rep(1:3, nsamples)] 
+                        Y <- Y + nugget * diag(nrow(Y)) + ind_covmat
+                        return(chol(Y))
+                    } )
 nwindows <- 99
@@
 And, *reweighted*:
-```{r reweighted, cache=TRUE, depends="sample_diffs"}
+```{r reweighted, depends="sample_diffs"}
 all.pcas <- eigen_windows(win.fn, k=2, do.windows=1:nwindows, w=1/rep(sqrt(nsamples),nsamples))

### 2018-08-22 writeup/local_pca_paper.tex
first pass at responding to all comments

@@
 \ifsubmission
-    % % stuff for submission
         \usepackage[nofiglist,nomarkers,figuresonly]{endfloat}
-    \usepackage{setspace}
-    % \doublespacing
+        \renewcommand\includegraphics{\relax}
+        \renewcommand{\includegraphics}[2][]{\relax}
+        \renewcommand{\figurename}{Fig.}
     \bibliographystyle{plos2015}
@@
     \newcommand\citep{\cite}
-    \renewcommand\includegraphics{\relax}
-    \renewcommand{\includegraphics}[2][]{\relax}
-    \renewcommand{\figurename}{Fig.}
     \newcommand{\Figure}{Fig }
@@
     % for the nice version
-    \usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
+    \usepackage[square,sort,comma,numbers]{natbib} 
+    % \bibpunct{(}{)}{;}{author-year}{}{,} 
     \bibliographystyle{plainnat}
@@
     \newcommand{\Figures}{{Figures }}
-    \renewcommand{\linelabel}[1]{}
+    % \renewcommand{\linelabel}[1]{}
 \fi
@@
 is an average of the relationships across locus-specific genealogical trees,
-which can be strongly affected on intermediate genomic scales by linked selection and other factors,
+which can be strongly affected on intermediate genomic scales by linked selection and other factors.
 We show how to use local principal components analysis to describe this 
@@
 as for instance reduced mean relatedness between individuals from distinct populations.
-However, it is not necessarily clear what aspects of demography should be included in the concept.
+However, it is not necessarily clear what aspects of demography should be included in the concept. \revpoint{2}{10}
 For instance, 
@@
 many species show chromosome-scale variation in diversity and divergence
-(e.g., \citep{langley2012genomic});
+(e.g., \citet{langley2012genomic});
 species phylogenies can differ along the genome 
@@
 
-PCA has been applied to genomic windows in methods to infer
-tracts of local ancestry in recently admixed populations
-\citep{bryc2010genomewide,brisbin2012pcadmix},
-and to identify putative chromosomal inversions \citep{ma2012investigation}.
-\revpoint{1}{8}
+Others have applied PCA to windows of the genome:
+\citet{ma2012investigation} used local PCA much as we do 
+to identify putative chromosomal inversions.
+\citet{bryc2010genomewide} and \citet{brisbin2012pcadmix}
+use PCA to infer tracts of local ancestry in recently admixed populations,
+but by projecting each genomic window onto the axes of a single, globally-defined PCA
+rather than doing PCA separately on each window.  \revpoint{1}{7}
 
@@
 for ease of visualization we usually only use the first two (referred to as $\pcone$ and $\pctwo$),
-and check that results hold using more. \revpoint{1}{7}
+and check that results hold using more. \revpoint{1}{6}
 The above procedure can be performed on any subset of the data;
@@
 
-Several of the datasets we use have unbalanced representations of diverged populations,
-which can have a strong impact on the results of PCA.
-(The principal axes may describe variation \emph{within} an overrepresented group
-rather than more significant variation between groups.)
-Therefore, to check that sampling patterns do not affect our results,
-we compared to a variant of PCA that gives roughly equal weight to each group of samples,
-rather than to each sample.
-The rationale and implementation of this method are described in Appendix \ref{apx:weighted_pca}.
+% Several of the datasets we use have unbalanced representations of diverged populations,
+% which can have a strong impact on the results of PCA.
+% (The principal axes may describe variation \emph{within} an overrepresented group
+% rather than more significant variation between groups.)
+% Therefore, to check that sampling patterns do not affect our results,
+% we compared to a variant of PCA that gives roughly equal weight to each group of samples,
+% rather than to each sample.
+% The rationale and implementation of this method are described in Appendix \ref{apx:weighted_pca}.
 
@@
 obtained using the top $k$ PCs,
-which is invariant under ordering of the PCs \revpoint{1}{3},
+which is invariant under ordering of the PCs \revpoint{1}{2},
 reflections, and rotations and yet contains all other information about the PCs.
@@
 % \end{align}
-
 To measure the similarity of patterns of relatedness for the $i^\text{th}$ window and $j^\text{th}$ window,
@@
 \subsection{Testing}
+\label{ss:testing_methods}
 
@@
 To provide a realistic test,
-we next used forwards-time, individual-based simulations, implemented using SLiM v3 \citep{slim_v3},
+we next used forwards-time, individual-based simulations, implemented using SLiM v3 \citep{haller2017flexible},
 which are described in detail in Appendix \ref{apx:slim_sims}.
@@
 with an overall recombination rate of 1.6785 crossovers per chromosome per generation.
-To improve speed, we used tskit \citep{tskit_paper} to record tree sequences
+To improve speed, we used tskit \citep{kelleher2018efficient} to record tree sequences
 and add neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
@@
 \subsection{Validation}
-
-\plr{Address:
-    mutation rate variation;
-    recombination rate variation;
-    choice of number of PCs;
-    choice of window size;
-    variation in missingness;
-    maybe differing sample sizes and reweighting.  }
+\label{ss:validation}
 
@@
 so that the order of empirical eigenvectors (PCs) switches randomly.
-The same type of simulations with dramatically different sample sizes
-showed that using weighted PCA substantially improved identification of different genomic regions
-(Supplemental Figure \ref{fig:test_weighted_pca}).
-
-Individual-based simulations using SLiM \citep{slim_v3} allowed us to test the effects
+
+Individual-based simulations using SLiM \citep{haller2017flexible} allowed us to test the effects
 of recombination and mutation rate variation, as well as linked selection.
@@
 did not induce patterns in the MDS visualizations correlated with recombination rate.
+\linelabel{ll:test_recomb_rate}
 Since varying mutation rate with a fixed recombination map
@@
 a recombination map with hotspots
-(the HapMap human female map for chromosome 7 \citep{hapmap})
+(the HapMap human female map for chromosome 7 \citep{hapmap2007second})
 induced outliers at long regions of low recombination rate
-(also as expected).
+(also as expected). \linelabel{ll:hotspot_recomb}
 
 Simulations with linked selection produced mixed results
-(Figure \ref{sfig:linked_selection}).
+(Figure \ref{sfig:slim_sims2}).
 The method strongly identified the regions under spatially varying linked selection.
@@
 to show how the plot of the first column is laid out along the genome.
+The patterns did not depend on the number of PCs used
+(see \Figure \ref{fig:dmel_k5} for the same plot with $k=5$ PCs), \revpoint{1}{3}
+and are not driven by variation in missingness (see \Figure \ref{sfig:dpgp_missingness}). \revpoint{2}{4}
 
@@
 The latter plots are quite different,
-showing that genomic windows in different regions of the MDS plot indeed show quite different patterns of relatedness.
+showing that genomic windows in different regions of the MDS plot 
+indeed show quite different patterns of relatedness.
 
@@
 (see vertical lines in \Figure \ref{fig:mds_allchr}).
+
 % \plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years.
@@
 we applied the method to all chromosomes together.
-Results for chromosome 3 are shown in \Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
+Results for chromosome 3 are shown in \Figure \ref{fig:mds12_medicago}, \revpoint{2}{13}
 and other chromosomes are similar:
-% the two types of population structure with higher gene density 
-% (colored green and purple in \Figure \ref{fig:mds12_medicago})
-% are nearly identical across all chromosomes,
-% but population structure associated with low gene density (colored orange)
-% shows more chromsome-specific patterns (see Supplemental \Figure \ref{sfig:pca_drosophila_allchr}).
 across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
@@
 
-We also found nearly identical results when choosing shorter windows of 1,000 SNPs;
-or choosing windows of equal length in base pairs rather than SNPs.
-Similarly, the results were not substantially changed
-when using weighted PCA to downweight the large group of Tunisian samples.
+We also found nearly identical results when choosing shorter windows of 1,000 SNPs
+(Supplemental Figure \ref{sfig:mds_medicago_allchr_1e3}; \revpoint{1}{4}
+or choosing windows of equal length in base pairs rather than SNPs. \revpoint{1}{5}
+% Similarly, the results were not substantially changed
+% when using weighted PCA to downweight the large group of Tunisian samples.
 
@@
     \caption{
+        \textbf{(top)}
        MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3,
@@
        Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:mds_medicago_allchr}.
-       \textbf{(below)}
-        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in \Figure \ref{fig:mds12_medicago}. 
+       \textbf{(bottom)}
+        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in the top figure.
         Each point corresponds to a sample, colored by country of origin.
@@
 the variation in patterns of relatedness we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
-must be explained somehow by linked selection.
+must be explained somehow by linked selection --
+the effects of neutral processes are not expected to show such large, chromosome-scale correlations.  \linelabel{ll:not_neutral}
 Furthermore, 
@@
 A related possibility is that variation in patterns of relatedness is due to recent admixture between previously separated populations,
-the effects of which were not uniform across the genome due to selection.
+the effects of which were not uniform across the genome due to selection. \revpoint{2}{6}
 For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
@@
 
-There has been substantial debate over the relative impacts of different forms of selection.
+There has been substantial debate over the relative impacts of different forms of selection
+\citep{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining}.
+\revpoint{2}{16}
 These have been difficult to disentangle in part because for the most part
@@
 and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
+Modern ``mixed model'' methods \citep[e.g.][]{loh2015efficient} 
+account for this with either a single, genome-wide kinship matrix
+or one constructed using only sites unliked to the focal SNP. \revpoint{2}{9}
 Since the effects of population structure is not constant along the genome,
@@
 with stronger population structure than the genome-wide average.
-A method such as ours might be used to provide a more sensitive correction.
+A method such as ours might be used to estimate local kinship matrices,
+thus providing a more sensitive correction,
+although doing so without removing the signal itself could be challenging.
 Fortunately, in our human dataset this does not seem likely to have a strong effect:
@@
 and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
-averaged over samples.
+averaged over samples. \revpoint{2}{11}
 The definition of eigenvectors does not specify their sign,
@@
 and $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
-We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}:
+We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}: \revpoint{2}{5}
 we divide the $j^\text{th}$ window into 10 equal-sized pieces,
@@
 
-\section{Weighted PCA}
-\label{apx:weighted_pca}
-
-Principal components analysis can be thought of as finding a good low-dimensional matrix factorization \citep{engelhardt2010analysis}
-that well-approximates the original data in the least-squares sense:
-if $C$ is the $N \times N$ genetic covariance matrix,
-then to find the top $k$ principal components, 
-we find an orthogonal $N \times k$ matrix $U$,
-and a $k \times k$ diagonal matrix $\Lambda$ with diagonal entries $\Lambda_{ii}=\lambda_i$ to minimize
-\begin{align} \label{eqn:objective}
-    \| C - U \Lambda U^T \|^2 = \sum_{ij} \left( C_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 .
-\end{align}
-The columns of $U$, known as the principal components, are the eigenvectors of $C$,
-the entries of $\lambda$ are the eigenvalues of $C$, 
-and the proportion of variance explained by the $m^\text{th}$ component is
+%%%%%%%%%%%%%%%%%%%%%
+\section{Simulations}
+
+We implemented two types of simulation:
+first, simple simulations of Gaussian ``genotypes'' 
+where the expectation of variation in ``population structure'' was clear;
+and next, individual-based simulations with explicit genomes, using SLiM.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Gaussian simulations}
+\label{apx:gaussian_sims}
+
+We simulated genotypes at each locus independently,
+drawing each vector of genotypes from a multivariate Gaussian distribution
+with zero mean and covariance matrix $\Sigma$.
+Sampled individuals came from three populations,
+and each $\Sigma_{ij}$ depends on which populations the individuals $i$ and $j$ are in,
+as well as the location along the chromosome.
+There are three population-level mean relatedness matrices along the genome,
+which apply to the first quarter ($S^{(1)}$),
+the middle half ($S^{(2)}$),
+and the last quarter ($S^{(3)}$), respectively:
 \begin{align*}
-    \frac{ \lambda_m^2 }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m U_{im} U_{jm} )^2 }{ \sum_{ij} C_{ij}^2 } .
+    S^{(1)} &= 
+    \begin{bmatrix}
+        0.75 &  0.25 &  0.0 \\
+        0.25 &  0.75 &  0.0 \\
+        0.0  &  0.0  &  1.0 \\
+    \end{bmatrix} \\
+    S^{(2)} &= 
+    \begin{bmatrix}
+        1.0  &  0.0  &  0.0 \\
+        0.0  &  0.75 &  0.25\\
+        0.0  &  0.25 &  0.75\\
+    \end{bmatrix} \\
+    S^{(3)} &= 
+    \begin{bmatrix}
+        0.75 &  0.0  &  0.25\\
+        0.0  &  1.0  &  0.0 \\
+        0.25 &  0.0  &  0.75\\
+    \end{bmatrix} \\
 \end{align*}
-
-Thinking about the problem as a least-squares approximation problem
-makes it clear why unbalanced sample sizes can result in undesireable outcomes.
-If we want to describe variation \emph{between} populations,
-but 80\% of the samples are from a single population,
-then unless populations are highly differentiated, 
-a better approximation to $C$ may be obtained by using the columns of $U$ to describe variation \emph{within} the overrepresented population
-rather than between the populations.
-A common workaround is to remove samples,
-but a more elegant solution can be found by reweighting the objective function in \eqref{eqn:objective}.
-Let $w_{i}$ be a weight associated with sample $i$,
-$W$ the diagonal matrix with $w$ along the diagonal,
-and instead seek to minimize
-\begin{align} \label{eqn:weighted_objective}
-    \| W^{1/2} (C - U \Lambda U^T) W^{1/2} \|^2 = \sum_{ij} w_i w_j \left( G_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 ,
-\end{align}
-and now for convenince we require $U$ to be orthogonal in $\ell_2(w)$, i.e., that $U^T W U =I$.
-We then would choose $w$ to give roughly equal weight to each \emph{population},
-instead of each individual.
-We have used with good results the weightings
-$w_i = 1/\max(10,n_i)$,
-where $n_i$ is, if there are discrete populations,
-the number of samples in the same population as sample $i$;
-or, for continuously sampled individuals,
-the number of samples within a certain distance of sample $i$.
-
-To solve \eqref{eqn:weighted_objective},
-let $\lambda$ and $V$ denote the top $k$ eigenvalues and eigenvectors of $W^{1/2} C W^{1/2}$,
-so that $V \Lambda V^T$ is the rank $k$ matrix closest in least squares to $W^{1/2} C W^{1/2}$;
-so if we define $U = W^{-1/2} V$
-then $U^T W U = V^T V = I$,
-and 
-\begin{align*}
-    W^{-1/2} V \Lambda V^T W^{-1/2} 
-    =
-    U \Lambda U^T
-\end{align*}
-is the low-dimensional approximation to $C$.
-The proportion of variance explained is calculated from eigenvalues as before,
-but has the interpretation
-\begin{align*}
-    \frac{ \lambda_m^2 }{ \sum_\ell \lambda_\ell^2 } 
-    = 
-    \frac{ \sum_{ij} w_i w_j ( \lambda_m U_{im} U_{jm} )^2 }{ \sum_{ij} w_i w_j C_{ij}^2 } .
-\end{align*}
-In our R implementation we use the Spectra library \citep{qiu2016rspectra}
-to find only the top $k$ eigenvectors.
-
-\paragraph{Testing}
-To demonstrate the utilitiy of this method XXX
-see Figure \ref{fig:test_weighted_pca}.
+If indiviudals $i$ and $j$ are in populations $p(i)$ and $p(j)$ respectively,
+then the covariance between their genotypes is $\Sigma_{ij} = S_{p(i),p(j)}$, 
+using the appropriate $S$ for that segment of the genome.
+The variance of individual $i$'s genotype is $\Sigma_{ii} = S_{p(i),p(i)} + 0.1$.
+
+We first created ``genotypes'' in this way with fifty individuals from each of the three populations;
+running our method on a genome with 99 windows of 400 loci each
+produced the first plot in Figure \ref{fig:test_missing}.
+These matrices are chosen so that the top two eigenvalues $\Sigma$ are the same (both 50.1),
+and so the ordering of the top two PCs is arbitrary.
+If our method was sensitive to PC ordering,
+then half the windows in each region that have one ordering would cluster with each other,
+separate from the other half. \linelabel{ll:pc_switching_test}
+
+We then marked each genotype in the first half of the chromosome
+as missing, independently, with probability $1/2$ and ran our method again,
+producing the second plot of Figure \ref{fig:test_missing}. \linelabel{ll:test_missing}
+If our method was influenced by missing data,
+we would expect the first half of the chromosome to separate from the second
+in the MDS plot.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{SLiM simulations}
+\label{apx:slim_sims}
+
+Our SLiM simulations were constructed as follows.
+Individuals are diploid, and genomes have a length of 153,520,244 bp.
+Recombination was either 
+(a) flat, with a constant rate of $10^{-9}$;
+(b) according to the human female HapMap map for chromosome 7;
+or (c) constant in each of seven equal-sized regions, beginning at $2.04\times 10^{-8}$,
+descending by a factor of four for three steps, and then ascending by a factor of four for three steps,
+so that the middle seventh has the lowest recombination rate,
+and the outer two sevenths has a rage 64 times higher.
+Selected mutations are introduced at a rate of $10^{-10}$ per bp per individual per generation,
+and have selection coefficients drawn from a Gamma distribution with mean 0.005 and shape 2;
+each coefficient are either positive or negative with probabilities $1/30$ and $29/30$ respectively.
+Each simulation was run for 50,000 generations.
+
+Each individual has a spatial position in the two-dimensional square of width $W=8$.
+Each time step, each individual chooses the nearest other to mate with,
+producing a random, Poisson distributed number of offspring with mean $1/3$.
+Offspring are assigned random spatial locations displaced from their parent's
+by a bivariate Gaussian with mean zero and standard deviation $\sigma = 0.2$,
+reflected to stay within the habitat range.
+
+Each individual survives to the next time step
+with probability equal to their fitness.
+Fitness values are determined multiplicatively by the effects of each mutation,
+but are multiplied by an additional factor determined by the local density of individuals.
+This factor is equal to $\rho / (1 + C)$,
+where $\rho = 2 \pi K \sigma^2$ is the carrying capacity per circle of radius $\sigma$;
+$K=100$ is the mean equilibrium population density;
+and $C$ is the sum of a Gaussian kernel with standard deviation $\sigma = 0.1$
+between the focal individual and all other individuals within distance $3 \sigma$.
+To avoid edge effects, fitnesses are further multiplied by $\min(1, z)$,
+where $z$ is the distance to the nearest boundary.
+This produces populations that fluctuate at equilibrium around 6,000 individuals in total,
+fairly evenly spread across the square.
+
+In one additional simulation, we modified fitnesses by multiplying the selective effect
+of each allele in each individual by multiplying it by $2x/W - 1$, 
+where $x$ is the $x$ coordinate of the individual.
+This makes the effect of each allele opposite on the left and on the right,
+and neutral in the middle,
+and leads to a moderate number of balanced polymorphisms.
 
@@
     \begin{center}
+        \includegraphics[width=1\textwidth]{Fig1_allchr_Together_MDS_plot_PCAk5_win103}
+    \end{center}
+    \caption{      
+        MSD visualizations for each chromosome arm of \textit{Drosophila melanogaster},
+        as in \Figure \ref{fig:mds_allchr},
+        except that the method was run using five PCs ($k=5$) instead of two.
+         \label{fig:dmel_k5}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
        \includegraphics[width=1\textwidth]{FigS_pca_plots_allchr_drosophila}
@@
     \caption{
-        MDS visualizations of the effects of population structure for all 8 chromosomes of the \textit{Medicago truncatula} data.
+        MDS visualizations of the effects of population structure for all 8 chromosomes 
+        of the \textit{Medicago truncatula} data, using windows of $10^4$ SNPs.
         \label{sfig:mds_medicago_allchr}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+        \ifusepngs
+           \includegraphics[width=0.85\textwidth]{Medicago_MDS_plot_allchr_win103_PCAk2.png}
+       \else
+           \includegraphics[width=0.85\textwidth]{Medicago_MDS_plot_allchr_win103_PCAk2}
+       \fi
+    \end{center}
+    \caption{
+        MDS visualizations of the effects of population structure for all 8 chromosomes 
+        of the \textit{Medicago truncatula} data as in \Figure \ref{sfig:mds_medicago_allchr},
+        but using windows of $10^3$ SNPs.
+        \label{sfig:mds_medicago_allchr_1e3}
     }
@@
     \begin{center}
-    \end{center}
-    \caption{
-        PCA plots for regions colored in \Figure \ref{sfig:mds_run_015598}
-        \label{sfig:pca_corner_run_015598}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_005464}
-    \end{center}
-    \caption{
-        MDS visualizations of simulation run 005464 (RECOMBTYPE, ``base"; SELTYPE, ``selected")
-        \label{sfig:mds_run_005464}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_031486}
-    \end{center}
-    \caption{
-        \plr{KEEP}
-        MDS visualizations of simulation run 031486 (RECOMBTYPE, ``base"; SELTYPE, ``selected")
-        \label{sfig:mds_run_031486}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_009673}
-    \end{center}
-    \caption{
-        \plr{KEEP}
-        MDS visualizations of simulation run 009673 (RECOMBTYPE, ``step"; SELTYPE, ``selected")
-        \label{sfig:mds_run_009673}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_016913}
-    \end{center}
-    \caption{
-        MDS visualizations of simulation run 009673 (RECOMBTYPE, ``base"; SELTYPE, ``balancing")
-        \label{sfig:mds_run_016913}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_031486} % base, selected
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_009673} % step, selected
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_016913} % base, balancing
        \includegraphics[width=0.82\textwidth]{FigS_plot_corner_pca-1_run_016913}
@@
     \caption{
-        PCA plots for regions colored in \Figure \ref{sfig:mds_run_016913}
-        \label{sfig:pca_corner_run_016913}
-    }
+        MDS visualizations of the results of individual-based simulations using SLiM
+        (see Appendix \ref{apx:slim_sims} for details).
+        All simulations incorporate linked selection by allowing selected mutations
+        to appear in the same two regions of the genome: 
+        the one-sixth of the genome immediately before the halfway point,
+        and the last one-sixth of the genome.
+        \textbf{(top)} Constant recombination rate.
+        \textbf{(top middle)} Stepwise varying recombination rate 
+        (as described in \Figure \ref{sfig:slim_sims1}).
+        \textbf{(bottom middle)} Constant recombination rate
+        with spatially varying effects of selection.
+        \textbf{(bottom)} PCA plots corresponding to the highlighted corners 
+        of the last MDS visualization, showing how spatially varying linked selection 
+        has affected patterns of relatedness.
+    } \label{sfig:slim_sims2}
 \end{figure}
@@
     \caption{
-        The missingness ratio along genome and the correlation of MDS against missingness in Drosophila.
+        The proportion of data in each window that are missing,
+        compared to the value of the first MDS coordinate
+        for the \textit{Drosophila melanogaster} data from \Figure \ref{fig:mds_allchr}.
         \label{sfig:dpgp_missingness}
@@
 
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.82\textwidth]{FigS_Medicago_missing_against_MDS1_win104}
-    \end{center}
-    \caption{
-        The missingness ratio along genome and the correlation of MDS against missingness in Medicago.
-        \label{sfig:medicago_missingness}
-    }
-\end{figure}
+% \begin{figure}
+%     \begin{center}
+%        \includegraphics[width=0.82\textwidth]{FigS_Medicago_missing_against_MDS1_win104}
+%     \end{center}
+%     \caption{
+%         The missingness ratio along genome and the correlation of MDS against missingness in Medicago.
+%         \label{sfig:medicago_missingness}
+%     }
+% \end{figure}
 
@@
     \caption{
-        From simple example -- base case; and with 50\% missing data.
+        MDS visualizations of the Gaussian genotypes described in Appendix \ref{apx:gaussian_sims},
+        for 50 individuals from each of three populations.
+        \textbf{(top)} The first quarter, middle half, and final quarter of the chromosome
+        each have different population structure, as expected,
+        despite the possibility for PC switching within each.
+        \textbf{(bottom)} The same picture results even after marking a random 50\% of the genotypes
+        in the first half of the chromosome as missing.
     } \label{fig:test_missing}.
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-        \includegraphics[width=0.82\textwidth]{simple_example_sample_diffs}
-        \includegraphics[width=0.82\textwidth]{simple_example_reweighted}
-    \end{center}
-    \caption{
-        From simple example -- different sample sizes; unweighted (top); weighted by $1/\sqrt{n_i}$ (bottom).
-    } \label{fig:test_weighted_pca}.
 \end{figure}

### 2018-08-22 writeup/weighted_pca.tex
first pass at responding to all comments

--- 2016-09-25 writeup/local_pca_paper.tex
@@
-\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
-\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
-%\geometry{landscape}                		% Activate for for rotated page geometry
-%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
-								% TeX will automatically convert eps --> pdf in pdflatex		
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{setspace}
-\usepackage[hidelinks]{hyperref}
-\usepackage[utf8]{inputenc}
-
-% % stuff for submission
-% \usepackage[nomarkers,figuresonly]{endfloat}
-% \usepackage{setspace}
-% \doublespacing
-% \usepackage{lineno}
-% \linenumbers
-% \bibliographystyle{plos2015}
-% \newcommand\citet{\cite}
-% \newcommand\citep{\cite}
-
-% % for the nice version
-\usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
-\bibliographystyle{plainnat}
-
-\usepackage{color}
-\newcommand{\plr}[1]{{\em \color{blue} #1}}
-
-\newcommand{\pcone}{PC1}
-\newcommand{\pctwo}{PC2}
-\newcommand{\pc}[1]{PC#1}
-\newcommand{\given}{\,\vert\,}
-\newcommand{\st}{\,\colon\,}
-\renewcommand{\and}{\,\&\,}
-\newcommand{\E}{\mathbb{E}}
-\renewcommand{\P}{\mathbb{P}}
-\DeclareMathOperator{\sgn}{sgn}
-\DeclareMathOperator{\var}{Var}
-\DeclareMathOperator{\cov}{Cov}
-\DeclareMathOperator{\tr}{tr}
-
-\title{Local PCA Shows How the Effect of Population Structure Differs Along the Genome}
-\author{Han Li, Peter Ralph}
-%\date{}							% Activate to display a given date or no date
-
-\begin{document}
-\maketitle
-%\section{}
-%\subsection{}
-% \doublespacing
-
-\section*{Abstract}
-
-Population structure leads to systematic patterns in measures of mean relatedness
-between individuals in large genomic datasets,
-which are often discovered and visualized using dimension reduction techniques such as principal component analysis (PCA).
-Mean relatedness
-is an average of the relationships across locus-specific genealogical trees,
-which can be strongly affected on intermediate genomic scales by linked selection and other factors,
-We show how to use local principal components analysis to describe this 
-meso-scale heterogeneity in patterns of relatedness,
-and apply the method to genomic data from three species,
-finding in each that the effect of population structure 
-can vary substantially across only a few megabases.
-In a global human dataset, localized heterogeneity is likely explained by polymorphic chromosomal inversions.
-In a range-wide dataset of \textit{Medicago truncatula},
-factors that produce heterogeneity are shared between chromosomes,
-correlate with local gene density,
-and may be caused by background selection or local adaptation.
-In a dataset of primarily African \textit{Drosophila melanogaster}, large-scale heterogeneity across each chromosome arm
-is explained by known chromosomal inversions thought to be under recent selection,
-and after removing samples carrying inversions, remaining heterogeneity is correlated with recombination rate and gene density,
-again suggesting a role for linked selection.
-% The method is implemented as an R package, called lostruct.
-The visualization method provides a flexible new way to discover biological drivers of genetic variation,
-and its application to data highlights the strong effects that linked selection and chromosomal inversions
-can have on observed patterns of genetic variation.
-
-\section{Introduction}
-
-% NOTE on historical uses of 'population structure'
-% Population Structure in Toads, Albert P. Blair, The American Naturalist, Vol. 77, No. 773 (Nov. - Dec., 1943), pp. 563-568
-% "In the last decade it has become increasingly apparent that the development of diversity within species populations is intimately concerned with such factors as size of breeding populations, periodic fluctuation of population size, sex ratio, activity range and differential survival of progeny.
-The sum total of these factors, population structure, determines to a great extent the relative importance of the roles which mutation, natural selection and random fluctuation of gene frequency play in the genetic configuration of the species."
-% Wright 1949, The Genetical Structure of Populations, says that:
-% "I propose to discuss certain genetic aspects of population structure. The term is used to include such matters as numbers, composition by age and sex, and state of subdivision."
-
-\citet{wright1949genetical} defined \emph{population structure} to encompass ``such matters as numbers, composition by age and sex, and state of subdivision'',
-where ``subdivision'' refers to restricted migration between subpopulations.
-The phrase is also commonly used to refer to the genetic patterns that result from this process,
-as for instance reduced mean relatedness between individuals from distinct populations.
-However, it is not necessarily clear what aspects of demography should be included in the concept.
-For instance, 
-\citet{blair1943population} defines \emph{population structure} to be the sum total of
-``such factors as size of breeding populations, periodic fluctuation of population size, sex ratio, activity range and \emph{differential survival of progeny}'' (emphasis added).
-The definition is similar to Wright's, but differs in including the effects of natural selection.
-On closer examination, incorporating differential survival or fecundity makes the concept less clear:
-should a randomly mating population consisting of two types that are partially reproductively isolated from each other
-be said to show population structure or not?
-Whatever the definition,
-it is clear that due to natural selection, the effects of population structure --
-the \emph{realized} patterns of genetic relatedness --
-differ depending on which portion of the genome is being considered.
-For instance, strongly locally adapted alleles of a gene will be selected against in migrants to different habitats,
-increasing genetic differentiation between populations near to this gene.
-Similarly, newly adaptive alleles spread first in local populations.
-These observations motivate many methods to search for genetic loci under selection,
-as for example in \citet{huertasnchez2013genetic} and \citet{martin2016natural}.
-
-% The phrase ``population structure'',
-% used to describe patterns in genome-wide mean kinship,
-% calls to mind to reduced gene flow between subpopulations. %, often because of geographical isolation.
-% However, it is widely recognized that because of selection the effects of gene flow are not equal everywhere on the genome,
-% and patterns of polymorphism within species and divergence between species can vary significantly depending on factors including local gene density.
-% This implies that, counterintuitively, 
-% the population structure of a species depends on which part of the genome is being examined.
-
-% The kinship matrix contains the kinship coefficient for pairwise individuals. 
-% Kinship coefficient defines the genetic relatedness between individuals.
-% It is the probability that two alleles randomly selected from two individuals are inherited from the most recent common ancester.
-% It could be estimated from given pedigree or from genome-wide covariances of genotype markers. 
-% It is well-known that for kinship matrix, actual relatednesses have a lot of noise about the expected value, and depend on where on the genome you look; this is why scans for selective sweeps work.
-% Populations are often structured in some way while there are systematic genetic variation between populations.
-
-% Population structure leads to systematic patterns in genome-wide mean kinship,
-% and so it is said that visualizations of kinship depict population structure,
-% rather than the \emph{effects} of population structure (which would be more accurate).
-% The \emph{kinship coefficient} for a pair of individuals
-% gives the expected proportion of their genome
-% that the two have inherited identically by descent;
-% for a single individual it is the inbreeding coefficient.
-% However, as \citet{wright1949genetical} wrote,
-% ``It has probably occurred to the reader that the coefficient of inbreeding
-% may mean very different things in different cases.''
-% The kinship coefficient originally referred to the expected probability
-% of coinheritance within a given pedigree;
-% while modern applications to ``unrelated'' individuals
-% use a genetic covariance matrix to estimate the \emph{realized} proportion of the genome
-% coinherited from sufficiently recent ancestors.
-
-These realized patterns of genetic relatedness
-summarize the shapes of the genealogical trees % that relate the samples
-at each location along the genome.
-Since these trees vary along the genome, so does relatedness,
-but averaging over sufficiently many trees we hope to get a stable estimate
-that doesn't depend much on the genetic markers chosen.
-This is not guaranteed:
-for instance,
-relatedness on sex chromosomes is expected to differ from the autosomes;
-and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
-\citep{kim2002hitchhiking,charlesworth1993effect,barton2000genetic}.
-Indeed,
-many species show chromosome-scale variation in diversity and divergence
-(e.g., \citep{langley2012genomic});
-species phylogenies can differ along the genome 
-due to incomplete lineage sorting,
-adaptive introgression and/or local adaptation 
-(e.g., \citep{pease2013accurate,ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting});
-and theoretical expectations predict that geographic patterns of relatedness should depend on selection
-\citep{charlesworth2003review}.
-
-Patterns in genome-wide relatedness are often summarized
-by applying principal components analysis (PCA, \citep{patterson2006population}) 
-to the genetic covariance matrix,
-as pioneered by \citet{menozzi1978synthetic}.
-% About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}.
-% Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.
-The results of PCA can be related to the genealogical history of the samples, 
-such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}, 
-and sometimes produce ``maps'' of population structure
-that reflect the samples' geographic origin distorted by rates of gene flow
-\citep{novembre2008genes}.
-% although other patterns can be caused by recent migration or nongeographic kinship patterns.  % \citep{astle2009population}.
-% Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
-% Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
-% Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}.
-% However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}
-
-Modeling such ``background'' kinship between samples
-is essential to genome-wide association studies (GWAS, \citep{price2006principal,astle2009population}),
-% and PCA has often been used in this context \citep{price2006principal}.
-and so understanding variation in kinship along the genome 
-% can help us to have a better understanding of the relation between genome structure and population structure, 
-could lead to more generally powerful methods,
-and may be essential for doing GWAS in species with substantial heterogeneity in realized patterns of mean relatedness along the genome.
-
-%% summary of the method: this happens in the method
-% To investigate how population structure varies along the genome in several datasets, 
-% we cut each genome into windows (with hundreds to thousands of SNPs in each), 
-% applied PCA to each window, 
-% and visualized the major ways that population structure, as summarized by PCA, varies among windows.
-% % In this project, we used SNP data for human, \textit{Medicago truncatula}, and
-% % whole genome sequencing data for \textit{Drosophila}.  
-% % Based on the principal components, we can estimate the similarity of population structure contained in each genome window.  
-% To quantify similarity of population structure between
-% windows, we constructed for each window an approximate, scaled covariance
-% matrix based on the first few principal components, and measured the pairwise Euclidean distance
-% between those matrices.  
-% We then use multidimensional scaling to visualize the
-% relationships between windows, which reduces the pairwise distance matrix to
-% lower dimension while preserving the distance information between windows as
-% well as possible \citep{borg2005modern}.
-
-A note on nomenclature: In this work we describe variation in patterns of relatedness using local PCA,
-where ``local'' refers to proximity along the genome.
-A number of general methods for dimensionality reduction also use a strategy of ``local PCA''
-(e.g., \citep{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear}),
-performing PCA not on the entire dataset but instead on subsets of observations,
-providing local pictures which are then stitched back together to give a global picture.
-At first sight,
-this differs from our method in that we restrict to subsets of \emph{variables} instead of subsets of observations.
-However, if we flip perspectives and think of each genetic variant as an observation,
-our method shares common threads, although 
-our method does not subsequently use adjacency along the genome,
-as we aim to identify similar regions that may be distant.
-% Future methods for visualization of genomic data
-% may benefit from other advances in this substantial literature 
-% (reviewed in \citep{vandermaaten2009dimensionality}).
-
-As reviewed above,
-it is common to describe variation along the genome of simple statistics such as $F_{ST}$,
-and also to use methods such as PCA to visualize large-scale patterns in genome-wide relatedness.
-In this paper we 
-apply PCA locally along the genome,
-thus describing in an unbiased, descriptive way
-how patterns of mean relatedness vary systematically along the genome, 
-in a way particularly suited to large samples from geographically distributed populations.
-Geographic population structure sets the stage by establishing similar patterns of relatedness across much of the genome;
-we then aim to describe how this structure is affected by selection and other factors.
-Our aim is not to identify outlier loci,
-but rather to describe larger-scale variation shared by many parts of the genome.
-By doing this with three taxa with diverse population histories,
-we can compare relative contributions of different sorts of variation across taxa;
-since the method is an visualization method, 
-we allow ourselves to be suprised by unexpected signals in the data.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
-
-As depicted in Figure \ref{fig:diagram}, the general steps to the method are:
-(1) divide the genome into windows,
-(2) summarize the patterns of relatedness in each window,
-(3) measure dissimilarity in relatedness between each pair of windows,
-(4) visualize the resulting dissimilarity matrix using multidimensional scaling (MDS),
-and (5) combine similar windows to more accurately visualize local effects of population structure using PCA.
-Details of how we carried out each step are given in the Methods,
-and code to run the analyses is provided as an R package at \url{https://github.com/petrelharp/local_pca}.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{the-method-diagram}
-    \end{center}
-    \caption{
-         An illustration of the method; see Methods for details.
-         % (1) the genome is divided into windows;
-         % (2) population structure is summarized in each with PCA, and 
-         % (3) dissimilarities for each are computed;
-         % (4) the patterns of dissimilarity are visualized with MDS,
-         % and (5) genomic regions clustering together are aggregated and visualized, again with PCA.
-         \label{fig:diagram}
-    }
-\end{figure}
-
-We applied the method to genomic datasets with good geographic sampling:
-380 African \textit{Drosophila melanogaster} from the Drosophila Genome Nexus \citep{lack2015drosophila},
-a worldwide dataset of humans,
-3,965 humans from several locations worldwide from the POPRES dataset \citep{nelson2008population},
-and 263 \textit{Medicago truncatula} from 24 countries around the Mediterranean basin 
-a range-wide dataset of the partially selfing weedy annual plant 
-from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
-
-In all three species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
-This implies that variation is due to meaningful heterogeneity in a biological process,
-since noise due to randomness in choice of local genealogical trees
-is not expected to show long distance correlations. 
-Below, we discuss the results and likely underlying causes.
-
-%%%%%%%%%%%
-\subsection{\textit{Drosophila melanogaster}}
-
-We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
-The first column of Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
-of the matrix of dissimilarities between genomic windows:
-in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
-For each chromosome arm, the MDS visualization roughly resembles a triangle,
-sometimes with additional points.
-Since the relative position of each window in this plot shows the similarity between windows, 
-this suggests that there are at least three extreme manifestations of population structure 
-typified by windows found in the ``corners'' of the figure,
-and that other windows' patterns of relatedness may be a mixture of those extremes. 
-The next two columns of Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
-plotted against the window's position along the genome,
-to show how the plot of the first column is laid out along the genome.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
-    \end{center}
-    \caption{
-        Variation in patterns of relatedness for windows across \textit{Drosophila melanogaster} chromosome arms.
-        In all plots, each point represents one window along the genome.
-         The first column shows the MDS visualization of relationships between windows, 
-         and the second and third columns show 
-        the two MDS coordinates
-         against the midpoint of each window;
-         rows correspond to chromosome arms.
-         Colors are consistent for plots in each row. 
-         Vertical lines show the breakpoints of known polymorphic inversions.   
-         Solid black lines are for the inversions we used in Figure \ref{fig:pca_by_inversion},
-         while dotted grey lines are for other known inversions.     
-         \label{fig:mds_allchr}
-    }
-\end{figure}
-
-To help visualize how clustered windows with similar patterns of relatedness are along each chromosome arm, 
-we selected three ``extreme'' windows in the MDS plot
-and the 5\% of windows that are closest to it in the MDS coordinates,
-then highlighted these windows' positions along the genome,
-and created PCA plots for the windows, combined.
-Representative plots are shown for three groups of windows on each chromosome arm
-in Figure \ref{fig:mds_allchr} (groups are shown in color),
-and in Supplemental Figure \ref{fig:pca_by_pop} (PCA plots).
-The latter plots are quite different,
-showing that genomic windows in different regions of the MDS plot indeed show quite different patterns of relatedness.
-
-% Fig fig:pca_by_pop WAS HERe
-
-The most striking variation in patterns of relatedness turns out to be explained by
-several large inversions that are polymorphic in these samples, 
-discussed in \citet{corbett2012population} and \citet{langley2012genomic}.
-% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
-To depict this, Figure \ref{fig:pca_by_inversion} shows
-the PCA plots in Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
-Taking chromosome arm 2L as an example,
-the two regions of similar, extreme patterns of relatedness
-shown in green in the first row of Figure \ref{fig:mds_allchr}
-lie directly around the breakpoints of the inversion In(2L)t,
-and the PCA plots in the first rows of Figure \ref{fig:pca_by_inversion}
-shows that patterns of relatedness here are mostly determined by inversion orientation.
-The regions shown in purple on chromosome 2L lie near the centromere,
-and have patterns of relatedness reflective of two axes of variation,
-seen in Figure \ref{fig:pca_by_pop},
-which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
-(see \citet{lack2015drosophila} for more about admixture in this sample).
-The regions shown in orange on chromosome 2L mostly lie inside the inversion,
-and show patterns of relatedness that are a mixture between the other two,
-as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
-Similar results are found in other chromosome arms,
-albeit complicated by the coexistence of more than one polymorphic inversion;
-however, each breakpoint visibly affects patterns in the MDS coordinates
-(see vertical lines in Figure \ref{fig:mds_allchr}).
-% \plr{\citet{corbett2012population} says In(1)Be is very young, $\sim$ 60 years.
-% Also cite \citet{corbettdetig2012sequencebased} for a computational method for detecting inversions.}
-% corbettdetig2012sequencebased suggests these inversions are recent-ish (but before out-of-Africa)
-% except that the situation of Chr3R is a little more complicated,
-% due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
-% However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
-% although we have not highlighted portions of the MDS plot that pick these out.
-% The inversions In(3L)P and In(3R)Mo, are at low frequency in these samples, 
-% and visibly affect the MDS plots although less than the more common inversion on each arm.
-% The very young inversion In(X)be shows little signal 
-% compared to the older inversion In(X)A \citep{corbett2012population}.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
-    \end{center}
-    \caption{
-         As in Figure 2, except that samples are colored by orientation of the polymorphic inversions 
-         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
-         (data from \citep{lack2015drosophila})
-         In each ``INV'' denotes an inverted genotype, ``ST'' denotes the standard orientation,
-         and ``N'' denotes unknown.
-        \label{fig:pca_by_inversion}
-    }
-\end{figure}
-
-To see how patterns of relatedness vary in the absence of polymorphic inversions,
-we performed the same analyses after removing, for each chromosome arm,
-any samples carrying inversions on that arm.
-In the result, shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions},
-the striking peaks associated with inversion breakpoints are gone,
-and previously smaller-scale variation now dominates the MDS visualization.
-For instance, the majority of the variation along 3L in Figure \ref{fig:mds_allchr}
-is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
-there is also a relatively small dip on the right end of the arm (near the centromere).
-In contrast, Supplemental Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
-remaining structure is dominated by the dip near the centromere.
-Without inversions, variation in patterns of relatedness shown in the MDS plots
-follows similar patterns to that previously seen in \textit{D.~melanogaster} recombination rate and diversity \citep{langley2012genomic,mackay2012drosophila}.
-Indeed, correlations between the recombination rate in each window and the position on the first MDS coordinate are highly significant
-(Spearman's $\rho=0.54$, $p<2 \times 10^{-16}$; Figures \ref{fig:drosophila_recomb_rate} and \ref{sfig:drosophila_gene_density_correlations}).
-This is consistent with the hypothesis that variation
-is due to selection, since the strength of linked selection increases with local gene density, 
-measured in units of recombination distance.
-The number of genes -- measured as the number of transcription start and end sites within each window --
-was not significantly correlated with MDS coordinate ($p=0.22$).
-
-\begin{figure}
-    \begin{center}
-        \includegraphics{drosophila_recomb_mds}
-    \end{center}
-    \caption{
-        The effects of population structure without inversions is correlated to recombination rate in \textit{Drosophila melanogaster}.
-        The first plot (in red) shows the first MDS coordinate along the genome 
-        for windows of 10,000 SNPs,
-        obtained after removing samples with inversions.
-        (A plot analogous to Figure \ref{fig:mds_allchr} is shown
-        in Supplemental Figure \ref{sfig:mds_allchr_noinversions}.)
-        The second plot (in blue) shows
-        local average recombination rates in cM/Mbp, obtained as midpoint estimates for 100Kbp windows 
-        from the Drosophila recombination rate calculator \citep{fistonlavier2010drosophila} release 5,
-        using rates from \citet{comeron2012many}.
-        The third plot (in black) shows the number of genes' transcription start and end sites within each 100Kbp window, divided by two.
-        Transcription start and end sites were obtained from the RefGene table from the UCSC browser. %, assembly dm3, on August 11, 2016.
-        The histone gene cluster on chromosome arm 2L is excluded.
-        \label{fig:drosophila_recomb_rate}
-    }
-\end{figure}
-
-%%%%%%%%%%%
-\subsection{Human}
-
-As we did for the \textit{Drosophila} data, we applied our method separately to all 22 human autosomes.
-On each, variation in patterns of relatedness was dominated by a small number of windows
-having similar patterns of relatedness to each other that differed dramatically from the rest of the chromosome.
-These may be primarily inversions: outlying windows coincide with
-three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
-notably a particularly large, polymorphic inversion on 8p23 (Figure \ref{fig:mds_human}). 
-% for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}) 
-Similar plots for all chromosomes are shown
-in Supplementary Figures \ref{sfig:mds12_chr1_8_human}, \ref{sfig:mds12_chr9_16_human}, and \ref{sfig:mds12_chr17_22_human}.
-PCA plots of many outlying windows show a characteristic trimodal shape 
-(shown for chromosome 8 in Figure \ref{sfig:pca_chr8_outliers}),
-presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
-(although we do not have data on orientation status).
-This trimodal shape has been proposed as a method to identify inversions \citep{ma2012investigation},
-but distinguishing this hypothesis from others,
-such as regions of low recombination rate,
-would require additional data.
-
-We also applied the method on all 22 autosomes together, 
-and found that, remarkably, 
-the inversion on chromosome 8 is still the most striking outlying signal (Figure \ref{sfig:mds1_along_allchr_human}). 
-Further investigation with a denser set of SNPs,
-allowing a finer genomic resolution,
-may yield other patterns.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
-    \end{center}
-    \caption{
-         Variation in structure between windows on human chromosomes 8, 15, and 17. 
-         Each point in each plot represents a window.
-         The first column shows the MDS visualization of relationships between windows;
-         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
-         Rows, from top to bottom show chromosomes 8, 15, and 17. 
-         The vertical red lines show the breakpoints of known inversions from \citet{antonacci2009characterization}.
-        \label{fig:mds_human}
-    }
-\end{figure}
-
-%%%%%%%%%%%
-\subsection{\textit{Medicago truncatula}}
-
-Unlike the other two species,
-the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
-showed similar patterns of gradual change in patterns of relatedness across each chromosome,
-with no indications of chromosome-specific patterns.
-This consistency suggests that the factor affecting the population structure for each chromosome is the same,
-as might be caused by varying strengths of linked selection.
-To verify that variation in the effects of population structure is shared across chromosomes,
-we applied the method to all chromosomes together.
-Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
-and other chromosomes are similar:
-% the two types of population structure with higher gene density 
-% (colored green and purple in Figure \ref{fig:mds12_medicago})
-% are nearly identical across all chromosomes,
-% but population structure associated with low gene density (colored orange)
-% shows more chromsome-specific patterns (see Supplemental Figure \ref{sfig:pca_drosophila_allchr}).
-across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
-which often have lower gene density and may therefore be less subject to linked selection.
-% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
-To verify that this is a possible explanation,
-we counted the number of genes found in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
-which are shown juxtaposed with
-the first MDS coordinate of each window in Figure \ref{fig:mds_medicago},
-and are significantly correlated, as shown in Supplemental Figure \ref{sfig:mds_gene_count}.
-(Values shown are the number of start and end positions of each predicted mRNA transcript,
-divided by two, assigned to the nearest window.)
-% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
-However, other genomic features, such as distance to centromere show roughly the same patterns,
-so we cannot rule out alternative hypotheses.
-In particular, the recombination rates estimated by \citet{paape2012finescale} appear visually to be similar,
-but were not available in a form that can be compared to our MDS values.
-
-We also found nearly identical results when choosing shorter windows of 1,000 SNPs;
-or choosing windows of equal length in base pairs rather than SNPs.
-Similarly, the results were not substantially changed
-when using weighted PCA to downweight the large group of Tunisian samples.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig6_Together_MDS_plot_chr3_final}
-       \includegraphics{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
-    \end{center}
-    \caption{
-       MDS visualization of patterns of relatedness on \textit{M.~truncatula} chromosome 3,
-       with corresponding PCA plots.
-       Each point in the plot represents a window;
-       the structure revealed by the MDS plot is strongly clustered along the chromosome,
-       with windows in the upper-right corner of the MDS plot (colored red) clustered around the centromere,
-       windows in the upper-left corner (purple) furthest from the centromere,
-       and the remaining corner (green) intermediate.
-       Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:mds_medicago_allchr}.
-       \textbf{(below)}
-        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
-        Each point corresponds to a sample, colored by country of origin.
-        Plots for remaining chromosomes are shown in Supplemental Figure \ref{sfig:pca_peaks_medicago_allchr}.
-       \label{fig:mds12_medicago}
-       \label{fig:pca_medicago}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{Fig5_MDS_and_gene_count_allchr_update_correction}
-    \end{center}
-    \caption{
-         MDS coordinate and gene density for each window in the \textit{Medicago} genome,
-         for chromosomes 1--8 (numbered above each pair of figures).
-         For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
-         The black plot below is gene count for each window against the position of each window.
-         \label{fig:mds_medicago}
-    }
-\end{figure}
-
-%%%%%%% %%%%%%%%%
-\section{Discussion}
-
-Our investigations have found substantial variation in the patterns of relatedness formed by population structure across the genomes
-of three diverse species,
-revealing distinct biological processes driving this variation in each species.
-More investigation, particularly on more species and datasets, will help to uncover what aspects of species history can explain these differences.
-With growing appreciation of the heterogeneous effects of selection across the genome,
-especially the importance of adaptive introgression and hybrid speciation \citep{pool2015natural,brandvain2014speciation,hufford2013genomic,fitzpatrick2010rapid,staubach2012genome},
-local adaptation \citep{lenormand2002limits,wang2014isolation},
-and inversion polymorphisms \citep{kirkpatrick2015chromosome,kirkpatrick2010chromosome},
-local PCA may prove to be a useful exploratory tool to discover important genomic features.
-% It is unclear whether the technique will be useful on reduced representation genotyping datasets 
-% due to marker density and issues with missing data --
-% our investigations with one such dataset were inconclusive --
-% but even low coverage, whole-genome sequence is very promising.
-
-We now discuss possible implications of this variation in the effects of population structure,
-the impact of various parameter choices in implementing the method,
-and possible additional applications.
-
-\paragraph{Chromosomal inversions}
-A major driver of variation in patterns of relatedness in two datasets we examined are inversions.
-This may be common,
-but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
-PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
-however, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
-or simply regions of reduced recombination
-without additional lines of evidence.
-Inversions show up in our method because across the inverted region,
-most gene trees share a common split that dates back to the origin of the inversion.
-However, in many applications, inversions are a nuisance.
-For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
-by regressing out the effect of linked SNPs on each other.
-Removing samples with the less common orientation of each inversion reduced,
-but did not eliminate, the signal of inversions
-seen in the \textit{Drosophila melanogaster} dataset,
-demonstrating that the genomic effects of transiently polymorphic inversions
-may outlast the inversions themselves.
-
-\paragraph{The effect of selection}
-It seems that
-the variation in patterns of relatedness we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
-must be explained somehow by linked selection.
-Furthermore, 
-the selection must be affecting many targets across the genome,
-since we see similar effects across long distances (even distinct chromosomes).
-For this reason, the most likely candidate may be selection against linked deleterious mutations, known as ``background selection'' \citep{charlesworth1993effect,charlesworth2013background}.
-Informally, background selection reduces the number of potential contributors to the gene pool 
-in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious};
-for this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently,
-Therefore, regions of the genome harboring many targets of local adaptation may show similar patterns,
-since migrant alleles in these regions will be selected against,
-and so locally gene trees will more closely reflect spatial proximity.
-
-A related possibility is that variation in patterns of relatedness is due to recent admixture between previously separated populations,
-the effects of which were not uniform across the genome due to selection.
-For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
-is due to selection against Neanderthal genes, leading to greater introgression in regions of lower gene density
-\citep{harris2016genetic,juric2016strength}.
-African \textit{Drosophila melanogaster} are known to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
-if selection regularly favors genes from one origin,
-this could lead to substantial variation in patterns of relatedness correlated with local gene density.
-
-There has been substantial debate over the relative impacts of different forms of selection.
-These have been difficult to disentangle in part because for the most part
-theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
-and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
-It may be possible to design more powerful statistics that make stronger use of spatial information.
-
-% How do we expect selection to affect population structure?
-% PCA summarizes patterns in kinship found in the genetic covariance matrix,
-% which is an average across locus-specific genealogies:
-% if individuals are closer in the genealogies of a given genomic region, 
-% they tend to be closer in the PCA maps for that region. 
-% Different DNA segments may have different gene tree and therefore different population structure for those segments. 
-% Second, the strength of linked selection differs for different DNA segments, and produces different population structure in region under linked selection compared to other region. 
-% Selective sweeps cause local recent ancestry or short trees.
-% Background selection causes shallow ones.
-% Strong selective sweeps of beneficial alleles can lead to relatively long genomic regions
-% characterized by short genealogical trees \citep{przeworski2005signature,garud2013selective}.
-% A genomic region with many targets for selection experiences
-% background selection and/or recurrent selective sweeps \citep{stephan1992effect,coop2012patterns},
-% which would tend to shorten genealogical trees in the region,
-% similar to a reduction in effective population size \citep{hudson1995deleterious,sattath2011pervasive}.
-% % Balancing selection causes deep trees.
-% On the other hand, balancing selection leads to very deep trees \citep{gao2014footprints},
-% and population structure locally describes which individuals have which alleles
-% rather than geographical proximity.
-% Finally, 
-% since recombination is suppressed between opposite orientations of a chromosomal inversion
-% near its breakpoints,
-% genealogies in these regions separate samples carrying the two orientations of the inversion.
-% In the resulting extended block, population structure shows two (for haploids) or three (for diploids) clusters
-% (indeed, \citep{ma2012investigation} has proposed using trimodality of local PCA plots
-% as a way to identify inversions).
-% if a chromosomal inversion is polymorphic in the sample, 
-% the regions around the breakpoints of inversions usually have high linkage disequilibrium and the two directions of a inversion will have different linked alleles around the breakpoints.
-Recombination suppression across inversions thus results in different genome structure and population structure.
-% Other effects, like noise, introgression might also influence population structure.
-% 
-% Many of the effects listed above (such as single selective sweeps or inversions) 
-% are not expected to have similar effects on population structure in different regions of the genome
-% because of randomness in which samples end up in which group;
-% but if these coincide with a region of reduced recombination (or an inversion),
-% these could drive major patterns of variation.
-% The more subtle effects of genome-wide linked selection could be shared across large regions
-% due to large-scale variation in gene density
-% (as by background selection or local adaptation at many genes).
-
-\paragraph{Parameter choices}
-There are several choices in the method that may in principle affect the results.
-As with whole-genome PCA,
-the choice of samples is important,
-as variation not strongly represented in the sample will not be discovered.
-The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
-but downweighting may be a better option that does not discard data
-(and here we present a method to do this).
-Next, the choice of window size may be important,
-although in our applications results were not sensitive to this,
-indicating that we can see variation on a sufficiently fine scale.
-Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in Figure \ref{fig:diagram}),
-along with the method used to discover common structure,
-will affect results.
-We used MDS, applied to either each chromosome separately or to the entire genome;
-for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosome,
-but genome-wide, their signal is obscured by the numerous other signals of comparable strength.
-
-Besides window length, there is also the question of how to choose windows.
-In these applications we have used nonoverlapping windows with equal numbers of polymorphic sites.
-Alternatively, windows could be chosen to have equal length in genetic distance,
-so that each would have roughly the same number of independent trees.
-However, we found little change in results when using different window sizes
-or when measuring windows in physical distance (in bp).
-
-% More generally, there are many possible methods to discover common structure in different parts of the genome:
-% in this work we use PCA to visualize the effects of population structure,
-% but other methods, such as STRUCTURE \citep{falush2003inference},
-% SPA \citep{yang2012modelbased},
-% SpaceMix \citep{bradburd2015spatial},
-% or other matrix factorization methods \citep{engelhardt2010analysis}
-% may highlight different sources of variation in the data.
-% It is also possible that other methods for measuring dissimilarity between windows' covariance matrices
-% or for summarizing the matrix of pairwise distances between windows
-% would lead to different insights.
-
-Finally, our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
-and how many MDS coordinates to use when describing the distance matrix between windows,
-but in our exploration, changing these has not produced dramatically different results.
-These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
-we encourage the user to experiment.
-
-\paragraph{Applications}
-So-called cryptic relatedness between samples
-has been one of the major sources of confounding in genome-wide association studies (GWAS)
-and so methods must account for it by modeling population structure or kinship \citep{astle2009population,yang2014advantages}.
-Since the effects of population structure is not constant along the genome,
-this could in principle lead to an inflation of false positives in parts of the genome
-with stronger population structure than the genome-wide average.
-A method such as ours might be used to provide a more sensitive correction.
-Fortunately, in our human dataset this does not seem likely to have a strong effect:
-most variation is due to small, independent regions, possibly primarily inversions,
-and so may not have a major effect on GWAS.
-In the other species we examined, particularly \textit{Drosophila melanogaster},
-treating population structure as a single quantity would entail a substantial loss of power,
-and could potentially be misleading.
-
-%%%%%%%%%%%%%%%%%
-\section{Methods}
-
-\subsection{PCA in genomic windows}
-
-We first recoded sampled genotypes as numeric matrices in the usual manner,
-by recording the number of nonreference alleles seen at each locus for each sample.
-We then divided the genome into contiguous segments 
-(``windows'')
-% recoded matrix into contiguous matrices that have the same columns but fewer rows than the original matrix, 
-and applied principal component analysis (PCA) as described in \citet{mcvean2009genealogical}
-separately to the submatrices that corresponded to each window.
-The choice of window length entails a tradeoff between signal and noise,
-since shorter windows allow better resolution along the genome but provide less precise estimates of relatedness.
-A method for choosing a window length to balance these considerations is given in Appendix \ref{apx:window_length}.
-% Specifically, we did PCA as follows:
-Precisely,
-denote by $Z$ the $L\times N$ recoded genotype matrix for a given window ($L$ is the number of SNPs and $N$ is the sample size), 
-and by $\overline{Z_{s}}$ the mean of non-missing entries for allele $s$, 
-so that $\overline{Z_{s}}=\frac{1}{n_s}\sum_j Z_{sj}$, 
-where the sum is over the $n_s$ nonmissing genotypes.
-We first compute the mean-centered matrix $X$, as $X_{si}=Z_{si}-\overline{Z_{s}}$,
-and preserving missingness.
-(This mean-centering makes the result not depend on the choice of reference allele,
-exactly if there is no missing data, and approximately otherwise.)
-Next, we find the covariance matrix of $X$, denoted $C$,
-as $C_{ij} = \frac{1}{m_{ij}-1} \sum_s X_{si} X_{sj} - \frac{1}{m_{ij}(m_{ij}-1)} (\sum_s X_{si})(\sum_s X_{sj})$,
-where all sums are over the $m_{ij}$ sites where both sample $i$ and sample $j$ have nonmissing genotypes.
-% We compute the covariance matrix using the R function cov() , with use=``pairwise",
-% which computes the covariance between each pair of individuals using all complete pairs of SNPs on those individuals.
-The principal components are the eigenvectors of $C$, 
-normalized to have Euclidean length equal to one,
-and ordered by magnitude of the eigenvalues.
-
-The top few principal components are generally good summaries of population structure; 
-we follow common practice and usually only use the first two (referred to as $\pcone$ and $\pctwo$).
-The above procedure can be performed on any subset of the data;
-for future reference, denote by $\pcone_j$ and $\pctwo_j$
-the result after applying to all SNPs in the $j^\text{th}$ window.
-
-Since the definition of eigenvectors does not specify their sign,
-when comparing between windows we choose signs to best match each other:
-after choosing $\pcone_1$, for instance, 
-if $u$ is the first eigenvector obtained from the covariance matrix
-for window $j$,
-then we next choose $\pcone_j = \pm u$,
-where the sign is chosen according to which of 
-$\| \pcone_{1} - u \|$ or
-$\| \pcone_{1} + u \|$ 
-is smaller.
-
-Several of the datasets we use have unbalanced representations of diverged populations,
-which can have a strong impact on the results of PCA.
-(The principal axes may describe variation \emph{within} an overrepresented group
-rather than more significant variation between groups.)
-Therefore, to check that sampling patterns do not affect our results,
-we compared to a variant of PCA that gives roughly equal weight to each group of samples,
-rather than to each sample.
-The rationale and implementation of this method are described in Appendix \ref{apx:weighted_pca}.
-
-\subsection{Similarity of patterns of relatedness between windows}
-
-We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
-in the space defined by the top principal components.
-However, 
-we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
-since rotations or reflections of these imply identical patterns of relatedness.
-Instead, we compare the low-dimensional approximations of the local covariance matrices
-obtained using the top $k$ PCs,
-which is invariant under reflections and rotations and yet contains all other information about the PCs.
-(For results shown here, we use $k=2$;
-results using larger numbers of PCs were nearly identical.)
-Furthermore, to remove the effect of artifacts such as mutation rate variation,
-we also rescale each approximate covariance matrix 
-so that the underlying data matrix has trace norm equal to one.
-
-To do this, define the $N \times k$ matrix $V(i)$ so that $V(i)_{\cdot \ell}$, 
-the $\ell^\text{th}$ column of $V(i)$,
-is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
-multiplied by $( \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} )^{1/2}$,
-where $\lambda_{\ell i}$ is the $\ell^\text{th}$ eigenvalue of the genetic covariance matrix.
-% \begin{equation}
-%     \begin{aligned}
-%         V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} ,
-%         \qquad
-%         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} 
-%     \end{aligned}
-% \end{equation}
-Then, the rescaled, rank $k$ approximate covariance matrix for the $i^\text{th}$ window is
-\begin{align} \label{eqn:est_cov}
-    M(i) &= \sum_{\ell=1}^k V(i)_{\cdot \ell} V(i)_{\cdot \ell}^T .
-\end{align}
-% \begin{align}
-%     M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
-% \end{align}
-
-To measure the similarity of patterns of relatedness for the $i^\text{th}$ window and $j^\text{th}$ window,
-we then use
-Euclidean distance $D_{ij}$ between the matrices $M(i)$ and $M(j)$:
-$D_{ij}^2 = \sum_{k\ell} ( M(i)_{k,\ell} - M(j)_{k,\ell} )^2$.
-
-The goal of comparing PC plots up to rotation and reflection 
-turned out to be equivalent to comparing rank-$k$ approximations to local covariance matrices.
-This suggests instead directly comparing entire local covariance matrices. 
-However, with thousands of samples and tens of thousands of windows,
-computing the distance matrix would take months of CPU time,
-while as defined above, $D$ can be computed in minutes using the following method.
-Since for square matrices $A$ and $B$,
-$\sum_{ij} (A_{ij}-B_{ij})^2 = \sum_{ij} (A^2_{ij} + B^2_{ij}) - 2 \tr(A^T B)$,
-then due to the orthogonality of eigenvectors and the cyclic invariance of trace,
-$D_{ij}$ can be computed efficiently as
-\begin{align}
-    D_{ij} 
-    = 
-    \frac{ \sum_{\ell=1}^k \lambda_{\ell i}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell i})^2 }
-    + \frac{ \sum_{\ell=1}^k \lambda_{\ell j}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell j})^2 }
-    - 2 \sum_{\ell, m=1}^k (V(i)^T V(j))^2_{\ell m} .
-\end{align}
-% from the code for pc_dist():
-% #' In the unweighted case, suppose that (u) and (v) are sets of vectors, and that 
-% #'    A = a_1 * u_1 u_1^T + ... + a_j * u_j u_j^T = U diag(a) U^T
-% #'    B = b_1 * v_1 v_1^T + ... + b_j * v_k v_k^T = V diag(b) V^T,
-% #' where U is the matrix whose columns are (u), and likewise for V.
-% #' Then 
-% #'    (A-B)^T (A-B) = A^T A + B^T B - A^T B - B^T A
-% #' and so
-% #'    ||A-B||^2 = ||A||^2 + ||B||^2 - 2 tr( A^T B )
-% #' By the cyclic invariance of trace, if X = U^T V then
-% #'    tr( A^T B ) = tr( U diag(a) U^T V diag(b) V^T ) = tr( diag(a) X diag(b) X^T ) .
-% \begin{align}
-%     \begin{split}
-%     D_{ij} &= 
-%         \left \{
-%                 \left ( 
-%                     V_{i1}\cdot V_{i1}^{} 
-%                 \right )^{2}
-%                 + \left ( V_{i2}\cdot V_{i2}^{} \right )^{2}
-%                 + \left ( V_{j1}\cdot V_{j1}^{} \right )^{2} 
-%                 +\left ( V_{j2}\cdot V_{j2}^{} \right )^{2} 
-%         \right. \\ 
-%         & \left. \qquad {}  
-%             -2\left [ 
-%                 \left ( V_{i1}\cdot V_{j1}^{} \right )^{2}
-%                 + \left ( V_{i1}\cdot V_{j2}^{} \right )^{2}
-%                 + \left ( V_{i2}\cdot V_{j1}^{} \right )^{2} 
-%                 + \left ( V_{i2}\cdot V_{j2}^{} \right )^{2}
-%             \right ]
-%         \right \}^{1/2}
-%     \end{split}
-% \end{align}
-
-\subsection{Visualization of results}
-
-We use multidimensional scaling (MDS) to visualize relationships between windows
-as summarized by the dissimilarity matrix $D$.
-MDS produces a set of $m$ coordinates for each window that
-give the arrangement in $m$-dimensional space that best recapitulates the original distance matrix.
-For results here, we use $m=2$ to produce one- or two-dimensional visualizations of relationships between windows' patterns of relatedness.
-
-We then locate variation in patterns of relatedness along the genome
-by choosing collections of windows that are nearby in MDS coordinates,
-and map their positions along the genome.
-A visualization of the effects of population structure across the entire collection is formed by extracting the corresponding genomic regions
-and performing PCA on all, aggregated, regions.
-
-\subsection{Datasets}
-% Recode the DNA sequence to a matrix consisting of 0,1,2 (and NA).
-
-The three publicly available datasets we used are summarized in Table \ref{tab:data_stats}.
-We converted each to a numeric matrix
-(with one row per polymorphic variant and one column per sample)
-by replacing each genotype with the number of nonreference alleles
-(or NA for missing data).
-% A normalization step (see below) ensures the result does not depend on the choice of reference allele.
-
-\paragraph{\textit{Drosophila melanogaster}:}
-We used whole-genome sequencing data 
-from the Drosophila Genome Nexus (\url{http://www.johnpool.net/genomes.html}, \citep{lack2015drosophila}),
-consisting of the Drosophila Population Genomics Project phases 1--3 \citep{langley2012genomic,pool2012population},
-and additional African genomes \citep{lack2015drosophila}.
-After removing 20 genomes with more than 8\% missing data,
-we were left with 380 samples from 16 countries across Africa and Europe.
-Since the \textit{Drosophila} samples are from inbred lines or haploid embryos, 
-we treat the samples as haploid when recoding;
-regions with residual heterozygosity were marked as missing in the original dataset;
-we also removed positions with more than 20\% missing data. 
-% (We chose these cutoffs as the tails of the relevant empirical distributions.)
-Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs;
-PCA plots for each arm are shown in Figure \ref{sfig:pca_drosophila_allchr}.
-
-\paragraph{Human:}
-We also used genomic data from the entire POPRES dataset \citep{nelson2008population},
-which has array-derived genotype information for 447,267 SNPs across the 22 autosomes
-of 3,965 samples in total: 346 African-Americans, 73 Asians, 3,187 Europeans and 359 Indian Asians.
-Since these data derive from genotyping arrays, the SNP density is much lower than the other datasets,
-which are each derived from whole genome sequencing.
-We excluded the sex chromosomes and the mitochondria.
-PCA plots for each chromosome, separately, are shown in Figure \ref{sfig:pca_human_allchr}.
-% We use the allele that has highest frequency in the samples as the reference allele for each position.
-
-% lack2015drosophila: describes the DGN, including the AGES samples.  reports diversity and mentions inversions but mostly a 'data' paper
-% pool2012population: describes 129 subsaharan samples (DPGP3)
-
-\paragraph{\textit{Medicago truncatula}:}
-Finally, we used whole-genome sequencing data from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
-which has 263 samples from 24 countries,
-primarily distributed around the Mediterranean basin.
-Each of the 8 chromosomes has 3--5 million SNPs;
-PCA plots for these are shown in Figure \ref{sfig:pca_medicago_allchr}.
-We did not use the mitochondria or chloroplasts.
-
-\begin{table}[ht]
-\centering
-    \begin{tabular}{p{0.8in}rrrr}
-  \hline
-    species 
-    & \parbox[t]{.8in}{\# SNPs per \\ window} 
-    & \parbox[t]{1in}{mean window\\ length (bp)}
-    & \parbox[t]{1.2in}{mean \# windows \\ per chromosome} 
-    & \parbox[t]{1.4in}{mean \% variance ex-\\plained by top 2 PCs} \\ 
-  \hline
-  \textit{Drosophila melanogaster} & 1,000 & 9,019 & 2,674 & 0.53 \\ 
-  Human & 100 & 636,494 & 203 & 0.55 \\ 
-  \textit{Medicago truncatula} & 10,000 & 102,580 & 467 & 0.50 \\ 
-   \hline
-\end{tabular}
-\caption{
-    Descriptive statistics for each dataset used.
-    % with the chosen window sizes.
-    % See text for how window sizes were chosen;
-    Columns 2--4 give statistics describing the window sizes used for most analyses for each dataset
-    (windows had a fixed number of SNPs).
-    The final column gives the percent variance explained by the top two principal components,
-    averaged across independent PCA of each window.
-    \label{tab:data_stats}
-}
-\end{table}
-
-\section{Data access}
-
-The methods described here
-are implemented in an open-source R package
-available at \url{https://github.com/petrelharp/local_pca},
-as well as scripts to perform all analyses from VCF files
-at various parameter settings.
-
-Datasets are available as follows:
-human (POPRES) at dbGaP with accession number phs000145.v4.p2,
-\textit{Medicago} at the Medicago Hapmap \url{http://www.medicagohapmap.org/},
-and \textit{Drosophila} at the Drosophila Genome Nexus, \url{http://www.johnpool.net/genomes.html}.
-
-%%%%%%%%
-\section*{Acknowledgements}
-
-We are indebted to John Pool, Russ Corbett-Detig, Matilde Cordeiro, and Peter Chang 
-for assistance with obtaining data and interpreting results
-(especially inversion status of \textit{D.~melanogaster} samples).
-Thanks also go to Yaniv Brandvain, Barbara Engelhardt, Charles Langley, Graham Coop, and Jeremy Berg for helpful comments
-and for encouraging the project.
-
-%%%%%%%%
-\section*{Disclosure declaration}
-
-The authors declare no conflicts of interest.
-
-\bibliography{references}
-
-\appendix
-\setcounter{table}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-
-\section{Choosing window length}
-\label{apx:window_length}
-
-The choice of window length entails a balance between signal and noise.
-In very short windows, genealogies of the samples will only be represented by a few trees,
-so variation between windows represents demographic noise rather than meaningful variation in patterns of relatedness.
-Longer windows generally have more distinct trees (and SNPs), 
-allowing for less noisy estimation of local patterns of relatedness.
-However, to better resolve meaningful signal, i.e., differences in patterns of relatedness along the genome,
-we would like reasonably short windows.
-% Window length choice therefore entails a signal versus noise tradeoff
-% in the estimates of population structure.
-% If we use the first principal component as a measure of population structure, 
-% then to choose the best window length,
-% we need to find a balance between the standard error of the first principal components for each window 
-% and the standard deviation between windows.
-
-Since we summarize patterns of relatedness using relative positions in the principal component maps,
-we quantify ``noise'' as the standard error of a sample's position on PC1 in a particular window,
-averaged across windows and samples,
-and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
-averaged over samples.
-(Recall that the signs for PCs are chosen to match each other.)
-Then, the mean variance across windows is
-\begin{align*}
-    \sigma_\text{signal}^2
-    = 
-    \frac{1}{N} \sum_{j=1}^{N}
-        \frac{1}{L}\sum_{i=1}^{L}\left ( \pcone_{ij} -\overline\pcone_{j} \right )^{2} ,
-\end{align*}
-where $\pcone_{ij}$ is the position of the $i^\text{th}$ individual on $\pcone$ in window $j$,
-and $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
-We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}:
-we divide the $j^\text{th}$ window into 10 equal-sized pieces,
-and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
-then the estimate of the squared standard error is
-$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$.
-Averaging over samples and windows,
-\begin{align*}
-    \sigma^2_\text{noise}
-    &=
-    \frac{1}{N} \sum_{j=1}^{N} \frac{1}{L}\sum_{i=1}^{L} \sigma^2_{ij} .
-\end{align*}
-
-For the main analysis, we defined windows to each consist of the same number of neighboring SNPs,
-and calculated $\sigma^2_\text{signal}$ and $\sigma^2_\text{noise}$
-for a range of window sizes (i.e., numbers of SNPs).
-For our main results we
-chose the smallest window for which $\sigma^2_\text{signal}$ was consistently larger than $\sigma^2_\text{noise}$ (but checked other sizes);
-the values for various window sizes across \textit{Drosophila} chromosomes are shown in Table \ref{tab:window_sizes}.
-% and the choices for all taxa are in Table \ref{tab:data_stats}.
-% In practice, even substantially different window sizes result in the same genome-wide patterns.
-In the cases we examined, we found nearly identical results after varying window size,
-and choosing windows to be of the same physical length (in bp) rather than in numbers of SNPs.
-
-% # Variances within and between windows for Drosophila, multiplied by 1000:
-% w <- structure(list(chrom = structure(c(1L, 1L, 2L, 2L, 3L, 3L, 4L, 
-% 4L, 5L, 5L), .Label = c("2L", "2R", "3L", "3R", "X"), class = "factor"), 
-%     type = structure(c(1L, 2L, 1L, 2L, 1L, 2L, 1L, 2L, 1L, 2L
-%     ), .Label = c("noise", "signal"), class = "factor"), `100` = c(2.05209, 
-%     2.75625, 2.18089, 2.77729, 2.07936, 2.601, 1.95364, 2.58064, 
-%     2.48004, 2.61121), `500` = c(1.64025, 2.69361, 1.91844, 2.704, 
-%     1.99809, 2.52004, 1.764, 2.51001, 2.04304, 2.43049), `1000` = c(1.18336, 
-%     2.22784, 1.63216, 2.65225, 1.64025, 2.401, 1.43641, 2.44036, 
-%     1.53664, 2.304), `10000` = c(0.169, 0.676, 0.576, 2.31361, 
-%     0.73441, 1.681, 0.58564, 1.96249, 1.61604, 0.324), `1e+05` = c(0.03969, 
-%     0.31329, 0.13456, 1.82329, 0.24649, 1.89225, 0.20164, 1.39876, 
-%     0.169, 1.14244)), .Names = c("chrom", "type", "100", "500", 
-% "1000", "10000", "1e+05"), row.names = c(NA, -10L), class = "data.frame")
-% print(xtable(w),include.rownames=FALSE)
-
-\begin{table}[ht]
-\centering
-    \begin{tabular}{cccrrrrr}
-  \hline
-        & & & \multicolumn{5}{c}{window length (SNPs)} \\
- & chrom.\ arm  & & 100 & 500 & 1,000 & 10,000 & 100,000 \\ 
-  \hline
-    & 2L & $\sigma^2_\text{noise}$  & 2.05  &  1.64  &  1.18  &  0.17  &  0.04 \\
-    & 	 & $\sigma^2_\text{signal}$ & 2.76  &  2.69  &  2.23  &  0.68  &  0.31 \\
-    & 2R & $\sigma^2_\text{noise}$  & 2.18  &  1.92  &  1.63  &  0.58  &  0.13 \\
-    & 	 & $\sigma^2_\text{signal}$ & 2.78  &  2.70  &  2.65  &  2.31  &  1.82 \\
-    & 3L & $\sigma^2_\text{noise}$  & 2.08  &  2.00  &  1.64  &  0.73  &  0.25 \\
-    & 	 & $\sigma^2_\text{signal}$ & 2.60  &  2.52  &  2.40  &  1.68  &  1.89 \\
-    & 3R & $\sigma^2_\text{noise}$  & 1.95  &  1.76  &  1.44  &  0.59  &  0.20 \\
-    & 	 & $\sigma^2_\text{signal}$ & 2.58  &  2.51  &  2.44  &  1.96  &  1.40 \\
-    & X  & $\sigma^2_\text{noise}$  & 2.48  &  2.04  &  1.54  &  1.62  &  0.17 \\
-    & 	 & $\sigma^2_\text{signal}$ & 2.61  &  2.43  &  2.30  &  0.32  &  1.14 \\
-   \hline
-\end{tabular}
-\caption{
-    Measures of signal and noise,
-    computed separately for each chromosome arm in the \textit{Drosophila} dataset,
-    at different window sizes.
-    All values are multiplied by $1,000$
-    (so typical variation is of order of 50\% of the actual values).
-    Starting at windows of 1,000 SNPs, the signal (variation of PC1 between windows)
-    starts to be substantially larger than the noise (standard error of PC1 for each window).
-} \label{tab:window_sizes}
-\end{table}
-
+%%%%%%%%%%%%%%%%%%%%%%
 \section{Weighted PCA}
@@
 to find only the top $k$ eigenvectors.
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
-    \end{center}
-    \caption{      
-        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_allchr},
-        on each chromosome arm of \textit{Drosophila melanogaster}.
-        In all plots, each point represents a sample. 
-        The first column shows the combined PCA plot for windows 
-        whose points are colored green in Figure \ref{fig:mds_allchr}; 
-        the second is for orange windows; 
-        and third is for purple windows.
-         \label{fig:pca_by_pop}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=1\textwidth]{FigS_pca_plots_allchr_drosophila}
-    \end{center}
-    \caption{
-        PCA plots for chromosome arms 2L, 2R, 3L, 3R and X of the \textit{Drosophila melanogaster} dataset.
-        \label{sfig:pca_drosophila_allchr}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human}
-    \end{center}
-    \caption{
-        PCA plots for all 22 huan autosomes from the POPRES data.
-        \label{sfig:pca_human_allchr}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=1\textwidth]{FigS_pca_plots_medicago_allchr}
-    \end{center}
-    \caption{
-        PCA plots for all 8 chromosomes in the \textit{Medicago truncatula} dataset.
-        \label{sfig:pca_medicago_allchr}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-        \includegraphics{MDS_allchr_Together_plot_samples_noinv}
-    \end{center}
-    \caption{
-        Variation in structure for windows of 1,000 SNPs 
-        across \textit{Drosophila melanogaster} chromosome arms: without inversions.
-         As in Figure \ref{fig:mds_allchr}, but after omitting for each chromosome arm individuals carrying the less frequent orientation
-         of any inversions on that chromosome arm.
-         The values differ from those in \ref{fig:drosophila_recomb_rate}
-         in the window size used and that some MDS values were inverted
-         (but relative orientation is meaningless as chromosome arms were run separately,
-         unlike for \textit{Medicago}).
-        In all plots, each point represents one window along the genome.
-         The first column shows the MDS visualization of relationships between windows, 
-         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
-         rows correspond to chromosome arms.
-         Colors are consistent for plots in each row. 
-         Vertical lines show the breakpoints of known polymorphic inversions.   
-         \label{sfig:mds_allchr_noinversions}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-        \includegraphics{drosophila_recomb_mds_correlation}
-    \end{center}
-    \caption{
-        Recombination rate, and the effects of population structure for \textit{Drosophila melanogaster}:
-        this shows the first MDS coordinate and recombination rate (in cM/Mbp), as in Supplementary Figure \ref{fig:drosophila_recomb_rate},
-        against each other.
-        Since the windows underlying estimates of Supplementary Figure \ref{fig:drosophila_recomb_rate} do not coincide,
-        to obtain correlations we divided the genome into 100Kbp bins, 
-        and for each variable (recombination rate and MDS coordinate 1)
-        averaged the values of each overlapping bin with weight proportional to the proportion of overlap.
-        The correlation coefficient and $p$-values for each linear regression are as follows:
-        2L: correlation $=0.52$, $r^2=0.27$;
-        2R: correlation $=0.43$, $r^2=0.18$;
-        3L: correlation $=0.47$, $r^2=0.21$;
-        3R: correlation $=0.46$, $r^2=0.21$;
-        X:  correlation $=0.50$, $r^2=0.24$.
-        \label{sfig:drosophila_gene_density_correlations}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr1_8_all_valid_and_pre}
-    \end{center}
-    \caption{
-         MDS plots for human chromosome 1-8.
-         The first column shows the MDS visualization of relationships between windows, 
-         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
-         rows correspond to chromosomes.        
-         Colorful vertical lines show the breakpoints of known valid inversions, 
-         while grey vertical lines show the breakpoints of predicted inversions.
-        \label{sfig:mds12_chr1_8_human}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr9_16_all_valid_and_pre}
-    \end{center}
-    \caption{
-        MDS plots for human chromosome 9-16.
-        \label{sfig:mds12_chr9_16_human}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr17_22_all_valid_and_pre}
-    \end{center}
-    \caption{
-        MDS plots for human chromosome 17-22.
-        \label{sfig:mds12_chr17_22_human}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.9\textwidth]{FigS_PCA_chr8_outliers_combine_update}
-    \end{center}
-    \caption{
-        PCA plots comparison for outlier windows and non-outlier windows.
-        \label{sfig:pca_chr8_outliers}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=1\textwidth]{FigS_MDS1D_plot_all_chr_human}
-    \end{center}
-    \caption{
-        MDS visualization of variation in the effects of population structure amongst windows across \emph{all} human autosomes simultaneously.
-        The small group of windows with positive outlying MDS values lie around the inversion at 8p23.
-        \label{sfig:mds1_along_allchr_human}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics{MDS_1D_against_gene_count_all_chr_win104_with_lm_update}
-    \end{center}
-    \caption{
-        First MDS coordinate against gene density for all 8 chromosomes of \textit{M.~truncatula}.
-        The first MDS coordinate is significantly correlated with gene count ($r=0.149$, $p=2.2\times 10^{-16}$). 
-        \label{sfig:mds_gene_count}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr}
-    \end{center}
-    \caption{
-        MDS visualizations of the effects of population structure for all 8 chromosomes of the \textit{Medicago truncatula} data.
-        \label{sfig:mds_medicago_allchr}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.82\textwidth]{FigS_pca_plots_for_Medicago_allchr_3peaks_byMDS}
-    \end{center}
-    \caption{
-        PCA plots for regions colored in Figure \ref{sfig:mds_medicago_allchr}
-        on all 8 chromosomes of \textit{Medicago truncatula}:
-        (A) green, (B) orange, and (C) purple.
-        \label{sfig:pca_peaks_medicago_allchr}
-    }
-\end{figure}
-
-\end{document}

### 2018-08-22 sims/background/background-fixed-s-sim.py

@@ -1,258 +1,0 @@
-#!/usr/bin/env python3
-description = '''
-Simulate AND write to msprime/vcf.
-'''
-
-import gzip
-import sys
-from optparse import OptionParser
-import math
-import time
-import random
-from ftprime import RecombCollector
-import msprime
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = OptionParser(description=description)
-parser.add_option("-T","--generations",dest="generations",help="number of generations to run for")
-parser.add_option("-N","--popsize",dest="popsize",help="size of each subpopulation",default=100)
-parser.add_option("-w","--gridwidth",dest="gridwidth",help="width of rectangular grid, in populations",default=3)
-parser.add_option("-y","--gridheight",dest="gridheight",help="height of rectangular grid, in populations (default: equal to gridwidth)")
-parser.add_option("-L","--length",dest="length",help="number of bp in the chromosome",default=1e4)
-parser.add_option("-l","--nloci",dest="nloci",help="number of selected loci",default=20)
-parser.add_option("-m","--migr",dest="migr",help="migration proportion between adjacent populations",default=.01)
-parser.add_option("-u","--sel_mut_rate",dest="sel_mut_rate",help="mutation rate of selected alleles",default=1e-7)
-parser.add_option("-r","--recomb_rate",dest="recomb_rate",help="recombination rate",default=2.5e-8)
-parser.add_option("-S","--selection_coef",help="strength of selection",default=.1)
-parser.add_option("-k","--nsamples",dest="nsamples",help="number of *diploid* samples, total")
-parser.add_option("-A","--ancestor_age",dest="ancestor_age",help="time to ancestor above beginning of sim")
[…]

### 2018-08-22 sims/background/background-introgression-fixed-s-sim.py

@@ -1,286 +1,0 @@
-#!/usr/bin/env python3
-description = '''
-Simulates on a rectangular grid that was bisected (in 'width') for some period in the past.
-'''
-
-import gzip
-import sys, os
-import math
-import time
-import random
-from ftprime import RecombCollector
-import msprime
-import argparse
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = argparse.ArgumentParser(description=description)
-parser.add_argument("--post_generations","-R", type=int, dest="post_generations",
-        help="number of generations since the division")
-parser.add_argument("--split_generations","-S", type=int, dest="split_generations",
-        help="number of generations of the split")
-parser.add_argument("--pre_generations","-T", type=int, dest="pre_generations",
-        help="number of generations to run before the split")
-parser.add_argument("--popsize","-N", type=int, dest="popsize",
-        help="size of each subpopulation",default=100)
-parser.add_argument("--gridwidth","-w", type=int, dest="gridwidth",
-        help="width of rectangular grid, in populations",default=3)
-parser.add_argument("--gridheight","-y", type=int, dest="gridheight",
-        help="height of rectangular grid, in populations (default: equal to gridwidth)")
-parser.add_argument("--length","-L", type=float, dest="length",
-        help="number of bp in the chromosome",default=1e4)
-parser.add_argument("--nloci","-l", type=int, dest="nloci",
-        help="number of selected loci",default=20)
-parser.add_argument("--migr","-m", type=float, dest="migr",
-        help="migration proportion between adjacent populations",default=.01)
[…]

### 2018-08-22 sims/background/background-introgression-sim.py

@@ -1,311 +1,0 @@
-#!/usr/bin/env python3
-description = '''
-Simulates on a rectangular grid that was bisected (in 'width') for some period in the past.
-'''
-
-import gzip
-import sys, os
-import math
-import time
-import random
-from ftprime import RecombCollector
-import msprime
-import argparse
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = argparse.ArgumentParser(description=description)
-parser.add_argument("--post_generations","-R", type=int, dest="post_generations",
-        help="number of generations since the division")
-parser.add_argument("--split_generations","-S", type=int, dest="split_generations",
-        help="number of generations of the split")
-parser.add_argument("--pre_generations","-T", type=int, dest="pre_generations",
-        help="number of generations to run before the split")
-parser.add_argument("--popsize","-N", type=int, dest="popsize",
-        help="size of each subpopulation",default=100)
-parser.add_argument("--gridwidth","-w", type=int, dest="gridwidth",
-        help="width of rectangular grid, in populations",default=3)
-parser.add_argument("--gridheight","-y", type=int, dest="gridheight",
-        help="height of rectangular grid, in populations (default: equal to gridwidth)")
-parser.add_argument("--length","-L", type=float, dest="length",
-        help="number of bp in the chromosome",default=1e4)
-parser.add_argument("--nloci","-l", type=int, dest="nloci",
-        help="number of selected loci",default=20)
-parser.add_argument("--migr","-m", type=float, dest="migr",
-        help="migration proportion between adjacent populations",default=.01)
[…]

### 2018-08-22 sims/background/background-sim.py

@@ -1,280 +1,0 @@
-#!/usr/bin/env python3
-description = '''
-Simulate AND write to msprime/vcf.
-'''
-
-import gzip
-import sys, os
-import math
-import time
-import random
-from ftprime import RecombCollector
-import msprime
-import argparse
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = argparse.ArgumentParser(description=description)
-parser.add_argument("--generations","-T", type=int, dest="generations",
-        help="number of generations to run for")
-parser.add_argument("--popsize","-N", type=int, dest="popsize",
-        help="size of each subpopulation",default=100)
-parser.add_argument("--gridwidth","-w", type=int, dest="gridwidth",
-        help="width of rectangular grid, in populations",default=3)
-parser.add_argument("--gridheight","-y", type=int, dest="gridheight",
-        help="height of rectangular grid, in populations (default: equal to gridwidth)")
-parser.add_argument("--length","-L", type=float, dest="length",
-        help="number of bp in the chromosome",default=1e4)
-parser.add_argument("--nloci","-l", type=int, dest="nloci",
-        help="number of selected loci",default=20)
-parser.add_argument("--migr","-m", type=float, dest="migr",
-        help="migration proportion between adjacent populations",default=.01)
-parser.add_argument("--sel_mut_rate","-u", type=float, dest="sel_mut_rate",
-        help="mutation rate of selected alleles",default=1e-7)
-parser.add_argument("--recomb_rate","-r", type=float, dest="recomb_rate",
-        help="recombination rate",default=2.5e-8)
-parser.add_argument("--gamma_alpha","-a", type=float, dest="gamma_alpha",
[…]

### 2018-08-22 sims/background/install_notes.sh

@@ -1,9 +1,0 @@
-# to (re)-install
-
-source $HOME/cmb/bin/python3_setup.sh
-
-# for msprime
-source /usr/usc/hdf5/1.8.12/setup.sh
-export PKG_CONFIG_PATH=$HOME/cmb/msprime
-
-python3 setup.py install --prefix /home/rcf-40/pralph/cmb/python3.5

### 2018-08-22 sims/background/neutral-split-sim.py

@@ -1,131 +1,0 @@
-#!/usr/bin/python3
-usage = '''
-Simulate a bisplit neutral simulation with different Ne on different chromosomes.
-'''
-
-import sys, os
-import time
-import json
-import random
-import msprime
-
-import argparse
-
-parser = argparse.ArgumentParser(description=usage)
-parser.add_argument('--outdir', '-o', help="Output directory.")
-parser.add_argument('--Ne', '-N', nargs="+", type=int, help="List of effective population sizes used for each *chromosome*.")
-parser.add_argument('--nsamples', '-n', type=int, help="Number of samples per population.")
-parser.add_argument('--chrom_length', '-L', type=float, help="Length of chromosome in bp.")
-parser.add_argument('--mut_rate', '-u', default=1e-7, type=float, help="Mutation rate per bp per generation.")
-parser.add_argument('--recomb_rate', '-r', default=1e-7, type=float, help="Recombination rate per bp per generation.")
-parser.add_argument('--relative_split_time', '-T', default=0.25, type=float, help="Time since rearrangement of populations in units of Ne.")
-parser.add_argument('--relative_fast_m', '-m', default=10, type=float, help="Migration rate for 'close' pops in units of Ne.")
-parser.add_argument('--relative_slow_m', '-M', default=0.1, type=float, help="Migration rate for 'distant' pops in units of Ne.")
-
-args = parser.parse_args()
-
-if not os.path.isdir(args.outdir):
-    os.mkdir(args.outdir)
-
-base_options = {
-        'Ne' : 1000,
-        'm_rel' : args.relative_fast_m,
-        'M_rel' : args.relative_slow_m,
-        'T_rel' : args.relative_split_time,
-        'nsamples' : args.nsamples,
-        'chrom_len' : args.chrom_length,
-        'treefile' : "chrom{}.trees",
-        'vcffile' : "chrom{}.vcf",
-        'mut_rate' : args.mut_rate,
-        'recomb_rate' : args.recomb_rate,
-     }
-
-logfile=sys.stdout
-
-options = { k:base_options.copy() for k in range(len(args.Ne)) }
-for k in range(len(args.Ne)):
-    options[k]['Ne'] = args.Ne[k]
-
[…]

### 2018-08-22 sims/background/other_attempts/background-sim-only.py

@@ -1,158 +1,0 @@
-#!/usr/bin/env python3.5
-description = '''
-Simulate.
-'''
-
-import gzip
-import sys
-from optparse import OptionParser
-import math
-import time
-import random
-from itertools import accumulate
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = OptionParser(description=description)
-parser.add_option("-T","--generations",dest="generations",help="number of generations to run for")
-parser.add_option("-N","--popsize",dest="popsize",help="size of each subpopulation",default=100)
-parser.add_option("-w","--width",dest="width",help="width of square grid, in populations",default=3)
-parser.add_option("-L","--length",dest="length",help="number of bp in the chromosome",default=1e4)
-parser.add_option("-l","--nloci",dest="nloci",help="number of selected loci",default=20)
-parser.add_option("-m","--migr",dest="migr",help="migration proportion between adjacent populations",default=.01)
-parser.add_option("-u","--mut_rate",dest="mut_rate",help="mutation rate of selected alleles",default=1e-7)
-parser.add_option("-r","--recomb_rate",dest="recomb_rate",help="recombination rate",default=2.5e-8)
-parser.add_option("-a","--gamma_alpha",dest="gamma_alpha",help="alpha parameter in gamma distributed selection coefficients",default=.23)
-parser.add_option("-b","--gamma_beta",dest="gamma_beta",help="beta parameter in gamma distributed selection coefficients",default=5.34)
-parser.add_option("-o","--outfile",dest="outfile",help="name of output file (or '-' for stdout)",default="-")
-parser.add_option("-g","--logfile",dest="logfile",help="name of log file (or '-' for stdout)",default="-")
[…]

### 2018-08-22 sims/background/other_attempts/ftprime-background-sim.py

@@ -1,179 +1,0 @@
-#!/usr/bin/env python3.5
-description = '''
-Simulate.
-'''
-
-import gzip
-import sys
-from optparse import OptionParser
-import math
-import time
-import msprime
-import simuOpt
-simuOpt.setOptions(alleleType='mutant')
-import simuPOP as sim
-import random
-import ftprime
-from itertools import accumulate
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = OptionParser(description=description)
-parser.add_option("-T","--generations",dest="generations",help="number of generations to run for")
-parser.add_option("-k","--nsamples",dest="nsamples",help="number of *diploid* samples, total")
-parser.add_option("-N","--popsize",dest="popsize",help="size of each subpopulation",default=100)
-parser.add_option("-w","--width",dest="width",help="width of square grid, in populations",default=3)
-parser.add_option("-L","--length",dest="length",help="number of bp in the chromosome",default=1e4)
-parser.add_option("-l","--nloci",dest="nloci",help="number of selected loci",default=20)
-parser.add_option("-m","--migr",dest="migr",help="migration proportion between adjacent populations",default=.01)
-parser.add_option("-u","--mut_rate",dest="mut_rate",help="mutation rate",default=1e-7)
-parser.add_option("-r","--recomb_rate",dest="recomb_rate",help="recombination rate",default=2.5e-8)
-parser.add_option("-a","--gamma_alpha",dest="gamma_alpha",help="alpha parameter in gamma distributed selection coefficients",default=1)
-parser.add_option("-b","--gamma_beta",dest="gamma_beta",help="beta parameter in gamma distributed selection coefficients",default=.001)
[…]

### 2018-08-22 sims/background/other_attempts/recombs-to-msprime.py

@@ -1,226 +1,0 @@
-#!/usr/bin/env python
-description = '''
-Convert simulations to msprime and VCF.
-'''
-
-import gzip
-import sys
-from optparse import OptionParser
-import math
-import time
-import random
-import msprime
-import ftprime
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = OptionParser(description=description)
-parser.add_option("-i","--infile",dest="infile",help="input file (output by background-sim.py)")
-parser.add_option("-k","--nsamples",dest="nsamples",help="number of *diploid* samples, total")
-parser.add_option("-A","--ancestor_age",dest="ancestor_age",help="time to ancestor above beginning of sim")
-parser.add_option("-u","--mut_rate",dest="mut_rate",help="mutation rate",default=1e-7)
-parser.add_option("-o","--outfile",dest="outfile",help="name of output file (or '-' for stdout)",default="-")
-parser.add_option("-t","--treefile",dest="treefile",help="name of output file for trees (or '-' for stdout)")
-parser.add_option("-g","--logfile",dest="logfile",help="name of log file (or '-' for stdout)",default="-")
-(options,args) =  parser.parse_args()
-
-infile = fileopt(options.infile, "r")
-outfile = fileopt(options.outfile, "w")
-logfile = fileopt(options.logfile, "w")
-
-logfile.write("Options:\n")
-logfile.write(str(options)+"\n")
-
-nsamples=int(options.nsamples)
-mut_rate=float(options.mut_rate)
-ancestor_age=float(options.ancestor_age)
-
-# Read in from the file::
-#   generations
-#   length
-#   N
-
-name,val = infile.readline().split(":")
-if name != "# generations":
-    raise ValueError("Bad file format.")
-else:
-    generations = int(val.strip())
-print("generations: ",generations)
-
-name,val = infile.readline().split(":")
[…]

### 2018-08-22 sims/background/other_attempts/simupop-notes.py

@@ -1,55 +1,0 @@
-
-# from http://simupop.sourceforge.net/manual_svn/build/userGuide_ch6_sec3.html
-# to do our own recombination, create an OffspringGenerator,
-# which gets called as a during-reproduction operator to create offspring,
-# for instance:
-def RandomMating(numOffspring=1., sexMode=RANDOM_SEX,
-        ops=MendelianGenoTransmitter(), subPopSize=[],
-        subPops=ALL_AVAIL, weight=0, selectionField='fitness'):
-    'A basic diploid sexual random mating scheme.'
-    return HomoMating(
-        chooser=RandomParentsChooser(True, selectionField),
-        generator=OffspringGenerator(ops, numOffspring, sexMode),
-        subPopSize=subPopSize,
-        subPops=subPops,
-        weight=weight)
-
-# In defining an OffspringGenerator...
-# "Note that you need to specify all needed operators if you use parameter ops
-# to change the operators used in a mating scheme (see Example
-# HeteroMatingWeight). That is to say, you can use ops=Recombinator() to
-# replace a default MendelianGenoTransmitter(), but you have to use
-# ops=[IdTagger(), MendelianGenoTransmitter()] if you would like to add a
-# during-mating operator to the default one."
-
-# "A customized genotype transmitter is only a Python during-mating operator.
-# Although it is possible to define a function and use a PyOperator directly
-# (Example PyOperator), it is much better to derive an operator from
-# PyOperator, as the case in Example newOperator:"
-
-class sexSpecificRecombinator(PyOperator):
-    def __init__(self, intensity=0, rates=0, loci=[], convMode=NO_CONVERSION,
-            maleIntensity=0, maleRates=0, maleLoci=[], maleConvMode=NO_CONVERSION,
-            *args, **kwargs):
-        # This operator is used to recombine maternal chromosomes
-        self.Recombinator = Recombinator(rates, intensity, loci, convMode)
-        # This operator is used to recombine paternal chromosomes
-        self.maleRecombinator = Recombinator(maleRates, maleIntensity,
-            maleLoci, maleConvMode)
-        #
-        PyOperator.__init__(self, func=self.transmitGenotype, *args, **kwargs)
-    #
-    def transmitGenotype(self, pop, off, dad, mom):
-        # Form the first homologous copy of offspring.
-        self.Recombinator.transmitGenotype(mom, off, 0)
-        # Form the second homologous copy of offspring.
-        self.maleRecombinator.transmitGenotype(dad, off, 1)
[…]

### 2018-08-22 sims/background/run_lostruct.R

@@ -1,1 +1,0 @@
-../../templated/run_lostruct.R

### 2018-08-22 sims/background/test_simuPOP/simple-background-sim.py

@@ -1,203 +1,0 @@
-#!/usr/bin/env python3
-description = '''
-Simulate a single population with background selection.
-'''
-
-import gzip
-import sys, os
-import math
-import time
-import random
-from ftprime import RecombCollector, ind_to_chrom, mapa_labels
-import msprime
-
-import argparse
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = argparse.ArgumentParser(description=description)
-parser.add_argument("--generations","-T", type=int, help="number of generations to run for")
-parser.add_argument("--popsize","-N", type=int, help="size of each subpopulation",default=100)
-parser.add_argument("--length","-L", type=float, help="number of bp in the chromosome",default=1e4)
-parser.add_argument("--nloci","-l", type=int, help="number of selected loci",default=20)
-parser.add_argument("--sel_mut_rate","-u", type=float, help="mutation rate of selected alleles",default=1e-7)
-parser.add_argument("--recomb_rate","-r", type=float, help="recombination rate",default=2.5e-8)
-parser.add_argument("--selection_coef","-S", type=float, help="strength of selection",default=.1)
-parser.add_argument("--nsamples","-k", type=int, help="number of *diploid* samples, total")
-parser.add_argument("--ancestor_age","-A", type=int, help="time to ancestor above beginning of sim")
-parser.add_argument("--mut_rate","-U", type=float, help="mutation rate",default=1e-7)
-parser.add_argument("--treefile","-t", type=str, help="name of output file for trees (default: not output)")
-
-parser.add_argument("--outfile","-o", type=str, help="name of output VCF file (default: not output)")
[…]

### 2018-08-22 sims/background/test_simuPOP/simple-background-sim_no-msprime.py

@@ -1,142 +1,0 @@
-#!/usr/bin/env python3
-description = '''
-Simulate a single population with varying levels of background selection.
-'''
-
-import gzip
-import sys, os
-import math
-import time
-import random
-
-import argparse
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = argparse.ArgumentParser(description=description)
-parser.add_argument("--generations","-T", type=int, help="number of generations to run for")
-parser.add_argument("--popsize","-N", type=int, help="size of each subpopulation",default=100)
-parser.add_argument("--length","-L", type=float, help="number of bp in the chromosome",default=1e4)
-parser.add_argument("--nloci","-l", type=int, help="number of selected loci",default=20)
-parser.add_argument("--sel_mut_rate","-u", type=float, help="mutation rate of selected alleles",default=1e-7)
-parser.add_argument("--recomb_rate","-r", type=float, help="recombination rate",default=2.5e-8)
-parser.add_argument("--selection_coef","-S", type=float, help="strength of selection",default=.1)
-parser.add_argument("--nsamples","-k", type=int, help="number of *diploid* samples, total")
-parser.add_argument("--ancestor_age","-A", type=int, help="time to ancestor above beginning of sim")
-parser.add_argument("--mut_rate","-U", type=float, help="mutation rate",default=1e-7)
-parser.add_argument("--treefile","-t", type=str, help="name of output file for trees (default: not output)")
-
-parser.add_argument("--outfile","-o", type=str, help="name of output VCF file (default: not output)")
-parser.add_argument("--logfile","-g", type=str, help="name of log file (or '-' for stdout)",default="-")
[…]

### 2018-08-22 sims/background/test_simuPOP/simple-background-sim_test_output.py

@@ -1,143 +1,0 @@
-#!/usr/bin/env python3
-description = '''
-Simulate a single population with varying levels of background selection.
-'''
-
-import gzip
-import sys, os
-import math
-import time
-import random
-
-import argparse
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = argparse.ArgumentParser(description=description)
-parser.add_argument("--generations","-T", type=int, help="number of generations to run for")
-parser.add_argument("--popsize","-N", type=int, help="size of each subpopulation",default=100)
-parser.add_argument("--length","-L", type=float, help="number of bp in the chromosome",default=1e4)
-parser.add_argument("--nloci","-l", type=int, help="number of selected loci",default=20)
-parser.add_argument("--sel_mut_rate","-u", type=float, help="mutation rate of selected alleles",default=1e-7)
-parser.add_argument("--recomb_rate","-r", type=float, help="recombination rate",default=2.5e-8)
-parser.add_argument("--selection_coef","-S", type=float, help="strength of selection",default=.1)
-parser.add_argument("--nsamples","-k", type=int, help="number of *diploid* samples, total")
-parser.add_argument("--ancestor_age","-A", type=int, help="time to ancestor above beginning of sim")
-parser.add_argument("--mut_rate","-U", type=float, help="mutation rate",default=1e-7)
-parser.add_argument("--treefile","-t", type=str, help="name of output file for trees (default: not output)")
-
-parser.add_argument("--outfile","-o", type=str, help="name of output VCF file (default: not output)")
-parser.add_argument("--logfile","-g", type=str, help="name of log file (or '-' for stdout)",default="-")
[…]

### 2018-08-22 sims/background/test_simuPOP/simple-uneven-loci-background-sim.py

@@ -1,216 +1,0 @@
-#!/usr/bin/env python3
-description = '''
-Simulate a single population with varying levels of background selection along a chromosome.
-'''
-
-import gzip
-import sys, os
-import math
-import time
-import random
-from ftprime import RecombCollector, ind_to_chrom, mapa_labels
-import msprime
-
-import argparse
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = argparse.ArgumentParser(description=description)
-parser.add_argument("--generations","-T", type=int, help="number of generations to run for")
-parser.add_argument("--popsize","-N", type=int, help="size of each subpopulation",default=100)
-parser.add_argument("--length","-L", type=float, help="number of bp in the chromosome",default=1e4)
-parser.add_argument("--nloci","-l", type=int, help="number of selected loci",default=20)
-parser.add_argument("--sel_mut_rate","-u", type=float, help="mutation rate of selected alleles",default=1e-7)
-parser.add_argument("--recomb_rate","-r", type=float, help="recombination rate",default=2.5e-8)
-parser.add_argument("--selection_coef","-S", type=float, help="strength of selection",default=.1)
-parser.add_argument("--nsamples","-k", type=int, help="number of *diploid* samples, total")
-parser.add_argument("--ancestor_age","-A", type=int, help="time to ancestor above beginning of sim")
-parser.add_argument("--mut_rate","-U", type=float, help="mutation rate",default=1e-7)
-parser.add_argument("--treefile","-t", type=str, help="name of output file for trees (default: not output)")
-parser.add_argument("--outfile","-o", type=str, help="name of output VCF file (default: not output)")
[…]

### 2018-08-22 sims/background/test_simuPOP/speed_test.sh

@@ -1,19 +1,0 @@
-#!/bin/bash
-
-# note %M is:
-#   M      Maximum resident set size of the process during its lifetime, in Kilobytes.
-
-
-/usr/bin/time --format='elapsed: %E / kernel: %S / user: %U / mem: %M' \
-    python3 simple-background-sim.py --generations 1000 --popsize 100 --nsamples 10 \
-        --nloci 2 --recomb_rate 1e-6 --length 1e6 --ancestor_age 10 --mut_rate 0 --sel_mut_rate 0 -s /dev/stdout -e /dev/stdout &> speed_test.log
-
-
-/usr/bin/time --format='elapsed: %E / kernel: %S / user: %U / mem: %M' \
-    python3 simple-background-sim_no-msprime.py --generations 1000 --popsize 100 --nsamples 10 \
-        --nloci 2 --recomb_rate 1e-6 --length 1e6 --ancestor_age 10 --mut_rate 0 --sel_mut_rate 0 -s /dev/stdout -e /dev/stdout &> speed_test_no_msprime.log
-
-
-/usr/bin/time --format='elapsed: %E / kernel: %S / user: %U / mem: %M' \
-    python3 simple-background-sim_test_output.py --generations 1000 --popsize 100 --nsamples 10 \
-        --nloci 2 --recomb_rate 1e-6 --length 1e6 --ancestor_age 10 --mut_rate 0 --sel_mut_rate 0 -s /dev/stdout -e /dev/stdout -t test.recomb &> speed_test_file_output.log

### 2018-08-22 sims/background/threeway-background-fixed-s-sim.py

@@ -1,261 +1,0 @@
-#!/usr/bin/env python3
-description = '''
-Simulate with simuPOP AND write to msprime/vcf:
-    a population arrangement switches from A<-B|C to A|B->C 
-    at some point in the past.
-
-Output in selloci is of the form
-   loc a1 a2 fitness
-meaning locus number, allele1, allele2, fitness when it is first seen.
-'''
-
-import gzip
-import sys, os
-import math
-import time
-import random
-from ftprime import RecombCollector
-import msprime
-import argparse
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = argparse.ArgumentParser(description=description)
-parser.add_argument('--relative_switch_time', '-w', default=0.25, type=float, 
-        help="Time since rearrangement of populations in units of population size.")
-parser.add_argument('--relative_fast_M', '-M', default=10, type=float, 
-        help="Migration rate for 'close' pops in units of population size.")
-parser.add_argument('--relative_slow_m', '-m', default=0.1, type=float, 
-        help="Migration rate for 'distant' pops in units of population size.")
-parser.add_argument("--generations", "-T", type=int, help="number of generations to run for before the switch")
-parser.add_argument("--popsize", "-N", type=int, help="size of each subpopulation", default=100)
-parser.add_argument("--length", "-L", type=float, help="number of bp in the chromosome", default=1e4)
-parser.add_argument("--nloci", "-l", type=int, help="number of selected loci", default=20)
-parser.add_argument("--sel_mut_rate", "-u", type=float, help="mutation rate of selected alleles", default=1e-7)
[…]

### 2018-08-22 sims/background/threeway-background-sim.py

@@ -1,278 +1,0 @@
-#!/usr/bin/env python3
-description = '''
-Simulate with simuPOP AND write to msprime/vcf:
-    a population arrangement switches from A<-B|C to A|B->C 
-    at some point in the past.
-
-Output in selloci is of the form
-   loc a1 a2 fitness
-meaning locus number, allele1, allele2, fitness when it is first seen.
-'''
-
-import gzip
-import sys, os
-import math
-import time
-import random
-from ftprime import RecombCollector
-import msprime
-import argparse
-
-def fileopt(fname,opts):
-    '''Return the file referred to by fname, open with options opts;
-    if fname is "-" return stdin/stdout; if fname ends with .gz run it through gzip.
-    '''
-    if fname == "-":
-        if opts == "r":
-            fobj = sys.stdin
-        elif opts == "w":
-            fobj = sys.stdout
-        else:
-            print("Something not right here.")
-    elif fname[len(fname)-3:len(fname)]==".gz":
-        fobj = gzip.open(fname,opts)
-    else:
-        fobj = open(fname,opts)
-    return fobj
-
-parser = argparse.ArgumentParser(description=description)
-parser.add_argument('--relative_switch_time', '-w', default=0.25, type=float, 
-        help="Time since rearrangement of populations in units of population size.")
-parser.add_argument('--relative_fast_M', '-M', default=10, type=float, 
-        help="Migration rate for 'close' pops in units of population size.")
-parser.add_argument('--relative_slow_m', '-m', default=0.1, type=float, 
-        help="Migration rate for 'distant' pops in units of population size.")
-parser.add_argument("--generations", "-T", type=int, help="number of generations to run for before the switch")
-parser.add_argument("--popsize", "-N", type=int, help="size of each subpopulation", default=100)
-parser.add_argument("--length", "-L", type=float, help="number of bp in the chromosome", default=1e4)
-parser.add_argument("--nloci", "-l", type=int, help="number of selected loci", default=20)
-parser.add_argument("--sel_mut_rate", "-u", type=float, help="mutation rate of selected alleles", default=1e-7)
[…]

### 2018-08-22 sims/background/threeway-neutral-split-sim.py

@@ -1,132 +1,0 @@
-#!/usr/bin/python3
-usage = '''
-Simulate a three-population neutral simulation with different Ne on different chromosomes.
-'''
-
-import sys, os
-import time
-import json
-import random
-import msprime
-
-import argparse
-
-parser = argparse.ArgumentParser(description=usage)
-parser.add_argument('--outdir', '-o', help="Output directory.")
-parser.add_argument('--Ne', '-N', nargs="+", type=int, help="List of effective population sizes used for each *chromosome*.")
-parser.add_argument('--nsamples', '-n', type=int, help="Number of samples per population.")
-parser.add_argument('--chrom_length', '-L', type=float, help="Length of chromosome in bp.")
-parser.add_argument('--mut_rate', '-u', default=1e-7, type=float, help="Mutation rate per bp per generation.")
-parser.add_argument('--recomb_rate', '-r', default=1e-7, type=float, help="Recombination rate per bp per generation.")
-parser.add_argument('--relative_split_time', '-T', default=0.25, type=float, help="Time since rearrangement of populations in units of Ne.")
-parser.add_argument('--relative_fast_m', '-m', default=10, type=float, help="Migration rate for 'close' pops in units of Ne.")
-parser.add_argument('--relative_slow_m', '-M', default=0.1, type=float, help="Migration rate for 'distant' pops in units of Ne.")
-
-args = parser.parse_args()
-
-if not os.path.isdir(args.outdir):
-    os.mkdir(args.outdir)
-
-base_options = {
-        'Ne' : 1000,
-        'm_rel' : args.relative_fast_m,
-        'M_rel' : args.relative_slow_m,
-        'T_rel' : args.relative_split_time,
-        'nsamples' : args.nsamples,
-        'chrom_len' : args.chrom_length,
-        'treefile' : "chrom{}.trees",
-        'vcffile' : "chrom{}.vcf",
-        'mut_rate' : args.mut_rate,
-        'recomb_rate' : args.recomb_rate,
-     }
-
-logfile=sys.stdout
-
-options = { k:base_options.copy() for k in range(len(args.Ne)) }
-for k in range(len(args.Ne)):
-    options[k]['Ne'] = args.Ne[k]
-
-logfile.write("Begun: writing to {}\n".format(args.outdir))
[…]

### 2018-08-22 writeup/review-responses.tex
better

@@
 \reply{
-    We have added a representative figure (for the Drosophila data, $k=5$),
+    We have added a table including correlations between MDS values between $k=2$ and $k=5$
+    for the Medicago data (Table \ref{tab:param_cors}),
+    and a representative figure (for the Drosophila data, $k=5$),
     see Figure \ref{fig:dmel_k5}. \revref
@@
 \reply{
-    We have added this: see supplemental figure \ref{sfig:mds_medicago_allchr_1e3}. \revref
+    The same table includes correlations between MDS values for different window lengths 
+    (Table \ref{tab:param_cors}). \revref
 }
@@
 \reply{
-    Again, we have added this: see supplemental figure \plr{XXX}. \revref
+    Again, we address this in Table \ref{tab:param_cors}. \revref
 }
@@
 \reply{
-    We have included the plots for Drosphila data using $k=5$:
-    see Figure \ref{fig:dmel_k5},
-    and hope that the readers will take our word on this point more generally.
+    We have included a table of correlations (Table \ref{tab:param_cors}),
+    as well as the plots for Drosphila data using $k=5$
+    (Figure \ref{fig:dmel_k5}),
+    and hope that readers will take our word on this point more generally.
 }

### 2018-08-22 writeup/local_pca_paper.tex
better

@@
 
-We also found nearly identical results when choosing shorter windows of 1,000 SNPs
-(Supplemental Figure \ref{sfig:mds_medicago_allchr_1e3}; \revpoint{1}{4}
-or choosing windows of equal length in base pairs rather than SNPs. \revpoint{1}{5}
+The results were highly consistent across window sizes, window types (SNPs or bp),
+and number of PCs (Table \ref{tab:param_cors}).
+\revpoint{1}{4} \revpoint{1}{5} \revpoint{1}{6}
+% We also found nearly identical results when choosing shorter windows of 1,000 SNPs
+% (Supplemental Figure \ref{sfig:mds_medicago_allchr_1e3}; \revpoint{1}{4}
+% or choosing windows of equal length in base pairs rather than SNPs. \revpoint{1}{5}
 % Similarly, the results were not substantially changed
@@
     \end{tabular}
-    \begin{caption}
+    \caption{
         Correlations between MDS coordinates of genomic regions
@@
         we first ran the algorithm with the specified window size and number of PCs 
-        ($k$ in equation \plr{XXX})
+        ($k$ in equation \eqref{eqn:est_cov})
         on the full \textit{Medicago truncatula} dataset.
@@
         This is not a symmetric operation, so these matrices are not symmetric.
-        As expected, parameter values with smaller windows produce noisier estimates.
-        Plots of MDS values along the genome are visually nearly identical
-        for parameter sets having similar window sizes --
-        full reports are available as supplementary material on Data Dryad \plr{XXX}.
-        \label{tab:param_cors}
-    \end{caption}
+        As expected, parameter values with smaller windows produce noisier estimates,
+        but plots of MDS values along the genome are visually very similar.
+    } \label{tab:param_cors}
 \end{table}
@@
     \caption{      
-        MSD visualizations for each chromosome arm of \textit{Drosophila melanogaster},
+        MDS visualizations for each chromosome arm of \textit{Drosophila melanogaster},
         as in \Figure \ref{fig:mds_allchr},
@@
 
-\begin{figure}
-    \begin{center}
-        \ifusepngs
-           \includegraphics[width=0.85\textwidth]{Medicago_MDS_plot_allchr_win103_PCAk2.png}
-       \else
-           \includegraphics[width=0.85\textwidth]{Medicago_MDS_plot_allchr_win103_PCAk2}
-       \fi
-    \end{center}
-    \caption{
-        MDS visualizations of the effects of population structure for all 8 chromosomes 
-        of the \textit{Medicago truncatula} data as in \Figure \ref{sfig:mds_medicago_allchr},
-        but using windows of $10^3$ SNPs.
-        \label{sfig:mds_medicago_allchr_1e3}
-    }
-\end{figure}
+% \begin{figure}
+%     \begin{center}
+%         \ifusepngs
+%            \includegraphics[width=0.85\textwidth]{Medicago_MDS_plot_allchr_win103_PCAk2.png}
+%        \else
+%            \includegraphics[width=0.85\textwidth]{Medicago_MDS_plot_allchr_win103_PCAk2}
+%        \fi
+%     \end{center}
+%     \caption{
+%         MDS visualizations of the effects of population structure for all 8 chromosomes 
+%         of the \textit{Medicago truncatula} data as in \Figure \ref{sfig:mds_medicago_allchr},
+%         but using windows of $10^3$ SNPs.
+%         \label{sfig:mds_medicago_allchr_1e3}
+%     }
+% \end{figure}
 

### 2018-08-23 writeup/review-responses.tex
edit of reviews

@@
 
-We are pleased to submit a revision of our manuscript,
+We are pleased to finally submit a revised version of our manuscript, 
+``Local PCA shows how the effect of population structure differs along the genome''.
+As you will see, at the reviewers' requests 
+we have put substantial effort into additional validation of the method,
+including simple simulations to back up some of our mathematical claims about the method
+(e.g., about PC switching),
+and large individual-based simulations with large numbers of loci under weak selection
+on realistically-sized chromosomes.
+The latter simulations, 
+which verify that our method can detect the action of linked selection,
+were made possible with tools only developed (some by us) over the last year.
+
+In adding simulation results and the additional verifications requested by the reviewers,
+we have endeavored not to expand the manuscript (or the supplementary figures) too much.
+
+In addition to the addition of many more tests,
+we have also tried to shift the focus away from background selection specifically
+to more general linked selection.
+We think it likely that background selection is a contributing factor,
+but that the relative contributions of different sorts of selection
+on genomic landscapes of diversity, divergence, and relatedness
+are almost completely unknown, and constitute a major open question in the field.
+We hope this addresses Reviewer 1's skepticism regarding background selection.
+
+In addition to the main document, with line numbers and hyperlinked references
+from the responses to reviews,
+we have also uploaded a version highlighting the changes between this version
+and the submitted one.
+We hope you find the revised manuscript an improvement as well.
 
@@
 \begin{quote}
-    Please understand that incremental changes will not be sufficient.  Adding
-    simulations to strengthen key claims will be necessary, particularly addressing
-    the impacts of mutation rate and recombination rate variation with more depth,
-    the concern regarding PC switching (Reviewer 1), and the concern regarding the
-    impacts of variation in missingness by sub-population (Reviewer 2).
+    Adding simulations to strengthen key claims will be necessary, particularly
+    addressing the impacts of mutation rate and recombination rate variation
+    with more depth, the concern regarding PC switching (Reviewer 1), and the
+    concern regarding the impacts of variation in missingness by sub-population
+    (Reviewer 2).
 \end{quote}
@@
 and have now done so.
-This took a substantial amount of work, because genome-scale forwards-time simulations
-with a large number of loci under selection
-is at or beyond the current limits of computation,
-depending on the number of individuals simulated.
+See below for more detailed responses.
 
@@
     required.
-\end{quote}
 
-\begin{quote}
     In short, the approach is interesting but not sufficiently explored to produce
@@
     amount of effort into simulations may alleviate these concerns somewhat.
+\end{quote}
 
+We have backed off somewhat on our previous speculation that the observed patterns
+are due to background selection,
+and have also put a large amount of effort into simulations (discussed below).
+We hope these alleviate your concerns.
+
+\begin{quote}
     Specific points: What does this method find? I'm concerned about:
@@
 
+Specific points are addressed below.
+Briefly, switching of PC 1 and 2 cannot mislead the method
+because we measure distances in a way that does not depend on PC ordering.
+It is possible that switching of PC 2 and 3 could mislead if we are using only $k=2$ of the PCs;
+however, we see the same patterns when using more PCs, so we know this is not happening.
+
+Nearly all of the inversion-like signals we find are well-characterized in the literature.
+The evidence for remaining inversion is \emph{not} watertight,
+because (as we point out), we cannot distinguish them from long regions of low recombination.
+The fact that nearly all of the strong, localized signals coincide with known polymorphic inversions
+induces us to emphasize inversions.
+
 \begin{point}{}
@@
 \reply{
-    Since we only look at SNPs,
+    We have now added simulation tests showing that large-scale variation in recombination rate
+    does not create spurious structure \llname{ll:test_recomb_rate};
+    except in the presence of long regions of low recombination,
+    as expected \llname{ll:hotspot_recomb}.
+    We also check for the effect of this in the real data:
+    since we only look at SNPs,
     a 1Kb window with mutation rate $\mu$ and recombination rate $\rho$
@@
     and the ratio $\mu/\rho$ determines the ``signal-to-noise'' ratio.
-    For this reason, varying recombination rate in simulations,
-    and varying window size and type in analysis
-    suffices to check for the effects of both recombination and mutation rate.
-    We have now added simulation tests showing that large-scale variation in recombination rate
-    does not create spurious structure \llname{ll:test_recomb_rate};
-    except in the presence of long regions of low recombination,
-    as expected \llname{ll:hotspot_recomb}
-    The point is addressed by comparing results with windows of different types --
-    windows of equal length in bp (or in SNPs) have different lengths in cM;
-    since these different choices show nearly identical patterns in all cases we have examined,
-    recombination rate variation cannot be driving the results
-    (e.g., compare Figures \ref{fig:mds_allchr} and Supplemental Figure \plr{XXX}).
+    Therefore, we can check for effects due to recombination
+    and mutation rate variation 
+    by comparing results with windows of different types --
+    windows of equal length in bp, or in SNPs, have different lengths in cM.
+    These different choices show nearly identical patterns in all cases we have examined,
+    as is summarized in Table \ref{tab:window_sizes},
+    making it unlikely that recombination or mutation rate variation could be driving the results.
 }
@@
     However, the only point at which we compare PCs in a way that could be sensitive to ordering
-    is in determining the window size -- in computing the distance between windows
-    we use a measure which is invariant under ordering.
+    is in determining the window size -- we compute the distance between windows
+    in a way that is invariant under ordering.
     We have made this more clear by moving the note about flipping signs of PCs
@@
     Expanding a bit more:
-    We strongly agree that methods that seek to test well-defined hypotheses
+    We agree that methods that seek to test well-defined hypotheses
     are extremely useful and powerful.
-    We also feel that methods for visualization and exploration are also useful --
-    a primary example here being PCA.
-    If PCA is useful -- and we think that it is --
+    We feel that methods for visualization and exploration are also useful --
+    a primary example here being PCA (as used for visualization).
+    If PCA is useful -- or more generally, dimension reduction --
     then it should be important to also know how much the thing that PCA is summarizing
@@
     We have tested the method against a large quantity of missing data
-    whereby the amount of missingness does not vary by population \llname{ll:test_missing}.
+    in which the amount of missingness does not vary by population \llname{ll:test_missing}.
     However, missingness that is structured by population could absolutely cause variation in PCs.
@@
     We definately would like our method to be insensitive to missing data,
-    but if missingness is in fact correlated with changes in patterns of relatedness (as in these cases),
+    but if missingness is in fact correlated with the signal we are looking for 
+    (changes in patterns of relatedness, as it is in these cases),
     correlation of MDS scores with amount of missingness is not in fact a problem.
@@
     we have provided a plot of missingness against MDS score (Figure \ref{sfig:dpgp_missingness})
-    and mentioned in the text \revref,
+    and mentioned it in the text \revref,
     but have not added a lengthy discussion.
@@
     and since the jackknife is only used to pick an appropriate window size,
-    it does not seem crucial.  \revref
+    it does not seem crucial.  {\revref}
     Since results are the same across different window sizes,
@@
     We agree that demonstrating this issue would be nice in principle,
-    but for this audience would be getting too far down in the weeds,
-    and including the requested simulations would inflate an already lengthy paper.
-    Additionally, we aim to provide a good method, not to prove that the method
-    is better than all other possible methods.
+    but think that for this audience would be getting too far down in the weeds.
+    (Besides, we aim to provide a good method, not to prove that the method
+    is better than all other possible methods.)
     For the reviewer's sake, here is some more about our rationale.
@@
 would be interesting if the authors tried to further characterize regions that are similar due to
-higher recombination rates. E.g. is there more/less density of polymorphisms in these regions?
+higher recombination rates. E.g.\ is there more/less density of polymorphisms in these regions?
 \end{point}
@@
 \reply{
-    We've made these changes.
+    Now provided.
 }

### 2018-08-23 writeup/local_pca_paper.tex
edit of reviews

@@
     \newcommand{\Figures}{{Figures }}
-    % \renewcommand{\linelabel}[1]{}
+    % \renewcommand{\llabel}[1]{}
 \fi
@@
 \citet{bryc2010genomewide} and \citet{brisbin2012pcadmix}
-use PCA to infer tracts of local ancestry in recently admixed populations,
+used PCA to infer tracts of local ancestry in recently admixed populations,
 but by projecting each genomic window onto the axes of a single, globally-defined PCA
@@
 can be caused by more than one biological process \citep{cruickshank2014reanalysis,burri2015linked},
-which in retrospect is unsuprising given that we are using a single statistic
+which in retrospect is unsurprising given that we are using a single statistic
 to describe a complex process.
@@
 (Note, however, that our measure of dissimilarity between windows
-does not depend on PC ordering.) \linelabel{ll:another_pc_point}
+does not depend on PC ordering.) \llabel{ll:another_pc_point}
 
@@
 the $\ell^\text{th}$ column of $V(i)$,
-is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
+is equal to the $\ell^\text{th}$ principal component of the $i^\text{th}$ window,
 multiplied by $( \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} )^{1/2}$,
@@
 To provide realistic population structure for PCA to identify,
-each simulation ahd at least 5,000 diploid individuals,
+each simulation had at least 5,000 diploid individuals,   
 living across a continuous square range,
@@
 To improve speed, we used tskit \citep{kelleher2018efficient} to record tree sequences
-and add neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
+and added neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
 Most simulations were neutral, but we also included linked selection, of two types.
@@
 did not induce patterns in the MDS visualizations correlated with recombination rate.
-\linelabel{ll:test_recomb_rate}
+\llabel{ll:test_recomb_rate}
 Since varying mutation rate with a fixed recombination map
 is equivalent to varying the recombination map and remapping windows,
-we used the same simulations to show that variation in mutation rate similarly has no effect.
+this also indicates that the method is not misled by variation in mutation rate.
 On the other hand,
@@
 induced outliers at long regions of low recombination rate
-(also as expected). \linelabel{ll:hotspot_recomb}
+(also as expected). \llabel{ll:hotspot_recomb}
 
@@
 This difference is likely because recombination rates are overall lower in the first case,
-leading to a stronger effect of linked seleciton.
+leading to a stronger effect of linked selection.
 These tests are not meant to be comprehensive survey of linked selection,
-but only to demonstrate that linked selection can produce signals simlar to what we see in real data.
+but only to demonstrate that linked selection can produce signals similar to what we see in real data.
 
@@
 (see \Figure \ref{fig:dmel_k5} for the same plot with $k=5$ PCs), \revpoint{1}{3}
-and are not driven by variation in missingness (see \Figure \ref{sfig:dpgp_missingness}). \revpoint{2}{4}
+and are only weakly correlated with variation in missingness 
+(see \Figure \ref{sfig:dpgp_missingness}). \revpoint{2}{4}
 
@@
 The results were highly consistent across window sizes, window types (SNPs or bp),
-and number of PCs (Table \ref{tab:param_cors}).
-\revpoint{1}{4} \revpoint{1}{5} \revpoint{1}{6}
+and number of PCs, as shown in Table \ref{tab:param_cors}.
+\revpoint{1}{3} \revpoint{1}{4} \revpoint{1}{5}
 % We also found nearly identical results when choosing shorter windows of 1,000 SNPs
@@
          For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
-         The black plot below is gene count for each window against the position of each window.
+         The black plot below is gene count for each window against the middle position of each window.
          \label{fig:mds_medicago}
@@
 \paragraph{The effect of selection}
-It seems that
-the variation in patterns of relatedness we see in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets
-must be explained somehow by linked selection --
-the effects of neutral processes are not expected to show such large, chromosome-scale correlations.  \linelabel{ll:not_neutral}
-Furthermore, 
-the selection must be affecting many targets across the genome,
-since we see similar effects across long distances (even distinct chromosomes).
-For this reason, the most likely candidate may be selection against linked deleterious mutations, known as ``background selection'' \citep{charlesworth1993effect,charlesworth2013background}.
-Informally, background selection reduces the number of potential contributors to the gene pool 
-in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious};
-for this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently,
+Neutral processes are not expected to produce the chromosome-scale correlations we see in
+patterns of relatedness 
+in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets, \llabel{ll:not_neutral}
+because correlations induced by neutral processes 
+should extend no further than does linkage disequilibrium
+(i.e., much less than a chromosome's length).
+This suggests that they are produced by linked selection,
+a hypothesis backed up by correlations with gene density and recombination rate.
+We have also shown with simulations that linked selection can, in at least some circumstances,
+produce the sorts of patterns we observe.
+How might selection cause variation in patterns of relatedness?
+For instance, background selection
+(the effect on linked sites of selection against deleterious mutations
+\citet{charlesworth1993effect,charlesworth2013background}),
+can informally be thought of as reducing the number of potential contributors to the gene pool 
+in regions of the genome with many possible deleterious mutations \citep{hudson1995deleterious}.
+For this reason, if it acts in a spatial context, it is expected to induce samples from nearby locations to cluster together more frequently.
 Therefore, regions of the genome harboring many targets of local adaptation may show similar patterns,
@@
 and so locally gene trees will more closely reflect spatial proximity.
-
-A related possibility is that variation in patterns of relatedness is due to recent admixture between previously separated populations,
+Other forms of selection, such as hard sweeps on new mutations,
+repeated selection on standing variation, local adaptation,
+or temporally fluctuating selection,
+could clearly lead to variation 
+in geographic patterns of relatedness in a similar way.
+
+Another possible contributor is recent admixture between previously separated populations,
 the effects of which were not uniform across the genome due to selection. \revpoint{2}{6}
@@
 There has been substantial debate over the relative impacts of different forms of selection
-\citep{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining}.
+(e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining}).
 \revpoint{2}{16}
@@
 We used MDS, applied to either each chromosome separately or to the entire genome;
-for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosome,
+for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosomes,
 but genome-wide, their signal is obscured by the numerous other signals of comparable strength.
@@
 In these applications we have used nonoverlapping windows with equal numbers of polymorphic sites.
-Alternatively, windows could be chosen to have equal length in genetic distance,
+Alternatively, windows could be chosen to have equal length in genetic distance, % or have equal length in physical distance
 so that each would have roughly the same number of independent trees.
@@
 account for this with either a single, genome-wide kinship matrix
-or one constructed using only sites unliked to the focal SNP. \revpoint{2}{9}
+or one constructed using only sites unliked to the focal SNP. \revpoint{2}{9}   %unlike ---> unlink?
 Since the effects of population structure is not constant along the genome,
@@
 is smaller.
-\linelabel{ll:moved_pc_note}
+\llabel{ll:moved_pc_note}
 
@@
     All values are multiplied by $1,000$
-    (so typical variation is of order of 50\% of the actual values).
+    (so typical variation is of order of 50\% of the actual values).  %A little confused about this line. What does this 50% mean?
     Starting at windows of 1,000 SNPs, the signal (variation of PC1 between windows)
@@
 then half the windows in each region that have one ordering would cluster with each other,
-separate from the other half. \linelabel{ll:pc_switching_test}
+separate from the other half. \llabel{ll:pc_switching_test}
 
@@
 as missing, independently, with probability $1/2$ and ran our method again,
-producing the second plot of Figure \ref{fig:test_missing}. \linelabel{ll:test_missing}
+producing the second plot of Figure \ref{fig:test_missing}. \llabel{ll:test_missing}
 If our method was influenced by missing data,
@@
         the second is for orange windows; 
-        and third is for purple windows.
+        and the third is for purple windows.
          \label{fig:pca_by_pop}
@@
          rows correspond to chromosome arms.
-         Colors are consistent for plots in each row. 
+%         Colors are consistent for plots in each row.   %this plot is all black points. 
          Vertical lines show the breakpoints of known polymorphic inversions.   

### 2018-08-26 writeup/local_pca_paper.tex
minor fixes

--- 2018-08-23 writeup/local_pca_paper.tex
@@
 To improve speed, we used tskit \citep{kelleher2018efficient} to record tree sequences
-and added neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
+and to add neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
 Most simulations were neutral, but we also included linked selection, of two types.
@@
 In these applications we have used nonoverlapping windows with equal numbers of polymorphic sites.
-Alternatively, windows could be chosen to have equal length in genetic distance, % or have equal length in physical distance
-so that each would have roughly the same number of independent trees.
 However, we found little change in results when using different window sizes
@@
 account for this with either a single, genome-wide kinship matrix
-or one constructed using only sites unliked to the focal SNP. \revpoint{2}{9}   %unlike ---> unlink?
+or one constructed using only sites unlinked to the focal SNP. \revpoint{2}{9}   %unlike ---> unlink?
 Since the effects of population structure is not constant along the genome,
@@
     at different window sizes.
-    All values are multiplied by $1,000$
-    (so typical variation is of order of 50\% of the actual values).  %A little confused about this line. What does this 50% mean?
+    All values are multiplied by $1,000$.
     Starting at windows of 1,000 SNPs, the signal (variation of PC1 between windows)
@@
          rows correspond to chromosome arms.
-%         Colors are consistent for plots in each row.   %this plot is all black points. 
          Vertical lines show the breakpoints of known polymorphic inversions.   

### 2018-08-31 writeup/review-responses.tex
response update

@@
 Using 2 PCs is common practice: only if this is the end of an analysis and the
-PCA was done for visualisation. Here you are using it for something so should
-keep all the relavant PCs.
+PCA was done for visualization. Here you are using it for something so should
+keep all the relevant PCs.
 \end{point}
@@
 detecting chromosomal inversions and evidence for background selection. It would help to
-have a well defined set of hypotheses, test the method’s accuracy  using simulation (see next
+have a well defined set of hypotheses, test the method'۪s accuracy  using simulation (see next
 comment), and compare to previous efforts in similar domains.
@@
     of the DPGP data having higher rates of missingness.
-    We definately would like our method to be insensitive to missing data,
+    We definitely would like our method to be insensitive to missing data,
     but if missingness is in fact correlated with the signal we are looking for 
@@
     Since results are the same across different window sizes,
-    we don't think this can explain a correlation wtih recombination rate.
+    we don't think this can explain a correlation with recombination rate.
 }
@@
     It is not clear that variation in mutation rate would act like an overall multiple
-    of the covariance matrix,
-    it since we're interested in patterns rather than magnitudes of relatedness,
+    of the covariance matrix.
+    Since we're interested in patterns rather than magnitudes of relatedness,
     it seems a wise thing to remove.
@@
 \reply{
-    Good point, but we think that the sense of unease is perhaps desireable here:
+    Good point, but we think that the sense of unease is perhaps desirable here:
     certainly differential survival is an aspect of demography;
@@
     We have included a table of correlations (Table \ref{tab:param_cors}),
-    as well as the plots for Drosphila data using $k=5$
+    as well as the plots for Drosophila data using $k=5$
     (Figure \ref{fig:dmel_k5}),

### 2018-09-15 medicago/medicago_data_setup.Rmd
wrong comma, closes #7

@@
 # there may be windows with missing data
-na.inds <- is.na( snp.pca[1,] )
+na.inds <- is.na( snp.pca[,1] )
 chr1.mds <- cmdscale( chr1.pcdist[!na.inds,!na.inds], eig=TRUE, k=4 )

### 2018-09-18 writeup/local_pca_paper.tex
resubmitting now

@@
 correlate with local gene density,
-and may be caused by background selection or local adaptation.
+and may be caused by linked selection, such as background selection or local adaptation.
 In a dataset of primarily African \textit{Drosophila melanogaster}, large-scale heterogeneity across each chromosome arm
@@
 can have on observed patterns of genetic variation.
+
+% Article summary
+% Principal component analysis is often used to describe overall "population structure", or patterns of relatedness arising from past demographic history, among a set of genomes.
+In this paper, we describe how the patterns uncovered by PCA differ at a local scale along the genome.
+We find that these patterns can be driven strongly by chromosomal inversions (and perhaps also regions of low recombination), and suggest that the strong remaining heterogeneity may be caused by the past action of linked selection.
 
@@
 obtained using the top $k$ PCs,
-which is invariant under ordering of the PCs \revpoint{1}{2},
+which is invariant under reordering of the PCs, \revpoint{1}{2}
 reflections, and rotations and yet contains all other information about the PCs.
-(For results shown here, we use $k=2$;
-results using larger numbers of PCs were nearly identical.)
+(For results shown here, we use $k=2$.)
 Furthermore, to remove the effect of artifacts such as mutation rate variation,
@@
 we then use
-Euclidean distance $D_{ij}$ between the matrices $M(i)$ and $M(j)$:
+Euclidean distance $D_{ij}$ between the matrices $M(i)$ and $M(j)$,
+defined by
 $D_{ij}^2 = \sum_{k\ell} ( M(i)_{k,\ell} - M(j)_{k,\ell} )^2$.
@@
 in the first quarter, middle half, and last quarter of the chromosome.
+The details of the simulation,
+also designed to detect sensitivity to PC switching,
+are given in Appendix \ref{apx:gaussian_sims}.
 To verify robustness to missing data,
@@
 rather than the segments having different covariance matrices.
-Details are given in Appendix \ref{apx:gaussian_sims}.
 
@@
 To improve speed, we used tskit \citep{kelleher2018efficient} to record tree sequences 
+in SLiM \citep{haller2018treesequence}
 and to add neutral mutations afterwards, at a rate of $10^{-9}$ per bp per generation.
@@
 without being affected by extreme differences in amount of missing data
-(Supplemental Figure \ref{fig:test_missing}).
+(Supplemental Figure \ref{sfig:test_missing}).
 This simulation also verifies insensitivity to ordering of top PCs,
@@
 of recombination and mutation rate variation, as well as linked selection.
-Results are shown in Supplemental Figures \ref{sfig:slim_sims1} and \ref{sfig:slim_sims2}.
 As expected, varying recombination rate stepwise by a factor of 64
-did not induce patterns in the MDS visualizations correlated with recombination rate.
+did not induce patterns in the MDS visualizations correlated with recombination rate
+(Supplemental Figure \ref{sfig:slim_sims1}).
 \llabel{ll:test_recomb_rate}
@@
 The patterns did not depend on the number of PCs used
-(see \Figure \ref{fig:dmel_k5} for the same plot with $k=5$ PCs), \revpoint{1}{3}
+(see \Figure \ref{sfig:dmel_k5} for the same plot with $k=5$ PCs), \revpoint{1}{3}
 and are only weakly correlated with variation in missingness 
@@
 in \Figure \ref{fig:mds_allchr} (groups are shown in color),
-and in Supplemental \Figure \ref{fig:pca_by_pop} (PCA plots).
+and in Supplemental \Figure \ref{sfig:pca_by_pop} (PCA plots).
 The latter plots are quite different,
@@
 To depict this, \Figure \ref{fig:pca_by_inversion} shows
-the PCA plots in \Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
+the PCA plots in Supplemental \Figure \ref{sfig:pca_by_pop} recolored by the orientation of the inversion for each sample.
 Taking chromosome arm 2L as an example,
@@
 and have patterns of relatedness reflective of two axes of variation,
-seen in \Figures \ref{fig:pca_by_pop} and \ref{fig:pca_by_inversion},
+seen in \Figure \ref{fig:pca_by_inversion} and Supplemental \Figure \ref{sfig:pca_by_pop},
 which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
@@
 any samples carrying inversions on that arm.
-In the result, shown in Supplemental \Figure \ref{sfig:mds_allchr_noinversions},
+In the result, shown in \Figure \ref{fig:drosophila_recomb_rate} and
+Supplemental \Figure \ref{sfig:mds_allchr_noinversions},
 the striking peaks associated with inversion breakpoints are gone,
@@
 and number of PCs, as shown in Table \ref{tab:param_cors}.
-\revpoint{1}{3} \revpoint{1}{4} \revpoint{1}{5}
+\revpoint{1}{4} \revpoint{1}{5}
 % We also found nearly identical results when choosing shorter windows of 1,000 SNPs
@@
        \label{fig:mds12_medicago}
-       \label{fig:pca_medicago}
     }
@@
 \paragraph{Chromosomal inversions}
-A major driver of variation in patterns of relatedness in two datasets we examined are inversions.
+A major driver of variation in patterns of relatedness in two datasets we examined seems to be inversions.
 This may be common,
@@
 or simply regions of reduced recombination
-without additional lines of evidence.
+without additional lines of evidence.  \llabel{ll:inversion_caveat}
 Inversions show up in our method because across the inverted region,
@@
 \citep{harris2016genetic,juric2016strength}.
-African \textit{Drosophila melanogaster} are known to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
+African \textit{Drosophila melanogaster} are thought to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
 if selection regularly favors genes from one origin,
@@
 There has been substantial debate over the relative impacts of different forms of selection
-(e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining}).
+(e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining,stankowski2018tempo}).
 \revpoint{2}{16}
@@
 The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
-but downweighting may be a better option that does not discard data
-(and here we present a method to do this).
+but downweighting may be a better option that does not discard data.
 Next, the choice of window size may be important,
-although in our applications results were not sensitive to this,
-indicating that we can see variation on a sufficiently fine scale.
+although in our applications results were not sensitive to this.
 Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in \Figure \ref{fig:diagram}),
@@
 running our method on a genome with 99 windows of 400 loci each
-produced the first plot in Figure \ref{fig:test_missing}.
+produced the first plot in Figure \ref{sfig:test_missing}.
 These matrices are chosen so that the top two eigenvalues $\Sigma$ are the same (both 50.1),
@@
 as missing, independently, with probability $1/2$ and ran our method again,
-producing the second plot of Figure \ref{fig:test_missing}. \llabel{ll:test_missing}
+producing the second plot of Figure \ref{sfig:test_missing}. \llabel{ll:test_missing}
 If our method was influenced by missing data,
@@
 so that the middle seventh has the lowest recombination rate,
-and the outer two sevenths has a rage 64 times higher.
+and the outer two sevenths has a rate 64 times higher.
 Selected mutations are introduced at a rate of $10^{-10}$ per bp per individual per generation,
@@
     \begin{center}
+       \includegraphics[width=1\textwidth]{FigS_pca_plots_allchr_drosophila}
+    \end{center}
+    \caption{
+        PCA plots for chromosome arms 2L, 2R, 3L, 3R and X of the \textit{Drosophila melanogaster} dataset.
+        \label{sfig:pca_drosophila_allchr}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+        \ifusepngs
+           \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human.png}
+       \else
+           \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human}
+       \fi
+    \end{center}
+    \caption{
+        PCA plots for all 22 human autosomes from the POPRES data.
+        \label{sfig:pca_human_allchr}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=1\textwidth]{FigS_pca_plots_medicago_allchr}
+    \end{center}
+    \caption{
+        PCA plots for all 8 chromosomes in the \textit{Medicago truncatula} dataset.
+        \label{sfig:pca_medicago_allchr}
+    }
+\end{figure}
+
+%%%%%%%%%%simulations
+
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=0.82\textwidth]{simple_example_base_case}
+        \includegraphics[width=0.82\textwidth]{simple_example_missing_data}
+    \end{center}
+    \caption{
+        MDS visualizations of the Gaussian genotypes described in Appendix \ref{apx:gaussian_sims},
+        for 50 individuals from each of three populations.
+        \textbf{(top)} The first quarter, middle half, and final quarter of the chromosome
+        each have different population structure, as expected,
+        despite the possibility for PC switching within each.
+        \textbf{(bottom)} The same picture results even after marking a random 50\% of the genotypes
+        in the first half of the chromosome as missing.
+    } \label{sfig:test_missing}.
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_027034}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_012720}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_015598}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corner_pca-1_run_015598}
+    \end{center}
+    \caption{
+        MDS visualizations of the results of individual-based simulations using SLiM
+        (see Appendix \ref{apx:slim_sims} for details).
+        All simulations are neutral, and recombination is:
+        \textbf{(top)} constant;
+        \textbf{(top middle)} varies stepwise by factors of two in seven equal-length segments,
+        with highest rates on the ends, so the middle segment has a recombination rate 64 times lower
+        than the ends;
+        \textbf{(bottom middle)} according to the HapMap human female chromosome 7 map.
+        The \textbf{bottom} figure shows PCA maps corresponding to the three colored windows
+        of the last (HapMap) situation;
+        the outlying regions are long regions of low recombination rate,
+        so that region can be dominated by a few correlated trees,
+        similar to an inversion.
+    } \label{sfig:slim_sims1}
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_031486} % base, selected
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_009673} % step, selected
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_016913} % base, balancing
+       \includegraphics[width=0.82\textwidth]{FigS_plot_corner_pca-1_run_016913}
+    \end{center}
+    \caption{
+        MDS visualizations of the results of individual-based simulations using SLiM
+        (see Appendix \ref{apx:slim_sims} for details).
+        All simulations incorporate linked selection by allowing selected mutations
+        to appear in the same two regions of the genome: 
+        the one-sixth of the genome immediately before the halfway point,
+        and the last one-sixth of the genome.
+        \textbf{(top)} Constant recombination rate.
+        \textbf{(top middle)} Stepwise varying recombination rate 
+        (as described in \Figure \ref{sfig:slim_sims1}).
+        \textbf{(bottom middle)} Constant recombination rate
+        with spatially varying effects of selection.
+        \textbf{(bottom)} PCA plots corresponding to the highlighted corners 
+        of the last MDS visualization, showing how spatially varying linked selection 
+        has affected patterns of relatedness.
+    } \label{sfig:slim_sims2}
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=1\textwidth]{Fig1_allchr_Together_MDS_plot_PCAk5_win103}
+    \end{center}
+    \caption{      
+        MDS visualizations for each chromosome arm of \textit{Drosophila melanogaster},
+        as in \Figure \ref{fig:mds_allchr},
+        except that the method was run using five PCs ($k=5$) instead of two.
+         \label{sfig:dmel_k5}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
+       \includegraphics[width=0.82\textwidth]{FigS_Drosophila_missings_against_MDS1_win103}
+    \end{center}
+    \caption{
+        The proportion of data in each window that are missing,
+        compared to the value of the first MDS coordinate
+        for the \textit{Drosophila melanogaster} data from \Figure \ref{fig:mds_allchr}.
+        \label{sfig:dpgp_missingness}
+    }
+\end{figure}
+
+% \begin{figure}
+%     \begin{center}
+%        \includegraphics[width=0.82\textwidth]{FigS_Medicago_missing_against_MDS1_win104}
+%     \end{center}
+%     \caption{
+%         The missingness ratio along genome and the correlation of MDS against missingness in Medicago.
+%         \label{sfig:medicago_missingness}
+%     }
+% \end{figure}
+
+\begin{figure}
+    \begin{center}
        \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
@@
         and the third is for purple windows.
-         \label{fig:pca_by_pop}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-        \includegraphics[width=1\textwidth]{Fig1_allchr_Together_MDS_plot_PCAk5_win103}
-    \end{center}
-    \caption{      
-        MDS visualizations for each chromosome arm of \textit{Drosophila melanogaster},
-        as in \Figure \ref{fig:mds_allchr},
-        except that the method was run using five PCs ($k=5$) instead of two.
-         \label{fig:dmel_k5}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=1\textwidth]{FigS_pca_plots_allchr_drosophila}
-    \end{center}
-    \caption{
-        PCA plots for chromosome arms 2L, 2R, 3L, 3R and X of the \textit{Drosophila melanogaster} dataset.
-        \label{sfig:pca_drosophila_allchr}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-        \ifusepngs
-           \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human.png}
-       \else
-           \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human}
-       \fi
-    \end{center}
-    \caption{
-        PCA plots for all 22 human autosomes from the POPRES data.
-        \label{sfig:pca_human_allchr}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=1\textwidth]{FigS_pca_plots_medicago_allchr}
-    \end{center}
-    \caption{
-        PCA plots for all 8 chromosomes in the \textit{Medicago truncatula} dataset.
-        \label{sfig:pca_medicago_allchr}
+         \label{sfig:pca_by_pop}
     }
@@
         \label{sfig:mds_gene_count}
-    }
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-        \ifusepngs
-           \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr.png}
-       \else
-           \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr}
-       \fi
-    \end{center}
-    \caption{
-        MDS visualizations of the effects of population structure for all 8 chromosomes 
-        of the \textit{Medicago truncatula} data, using windows of $10^4$ SNPs.
-        \label{sfig:mds_medicago_allchr}
     }
@@
     \begin{center}
+        \ifusepngs
+           \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr.png}
+       \else
+           \includegraphics[width=0.85\textwidth]{FigS_Together_MDS_plot_allchr}
+       \fi
+    \end{center}
+    \caption{
+        MDS visualizations of the effects of population structure for all 8 chromosomes 
+        of the \textit{Medicago truncatula} data, using windows of $10^4$ SNPs.
+        \label{sfig:mds_medicago_allchr}
+    }
+\end{figure}
+
+\begin{figure}
+    \begin{center}
        \includegraphics[width=0.82\textwidth]{FigS_pca_plots_for_Medicago_allchr_3peaks_byMDS}
@@
 
-%%%%%%%%%%simulations
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_027034}
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_012720}
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_015598}
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corner_pca-1_run_015598}
-    \end{center}
-    \caption{
-        MDS visualizations of the results of individual-based simulations using SLiM
-        (see Appendix \ref{apx:slim_sims} for details).
-        All simulations are neutral, and recombination is:
-        \textbf{(top)} constant;
-        \textbf{(top middle)} varies stepwise by factors of two in seven equal-length segments,
-        with highest rates on the ends, so the middle segment has a recombination rate 64 times lower
-        than the ends;
-        \textbf{(bottom middle)} according to the HapMap human female chromosome 7 map.
-        The \textbf{bottom} figure shows PCA maps corresponding to the three colored windows
-        of the last (HapMap) situation;
-        the outlying regions are long regions of low recombination rate,
-        so that region can be dominated by a few correlated trees,
-        similar to an inversion.
-    } \label{sfig:slim_sims1}
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_031486} % base, selected
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_009673} % step, selected
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corners-1_run_016913} % base, balancing
-       \includegraphics[width=0.82\textwidth]{FigS_plot_corner_pca-1_run_016913}
-    \end{center}
-    \caption{
-        MDS visualizations of the results of individual-based simulations using SLiM
-        (see Appendix \ref{apx:slim_sims} for details).
-        All simulations incorporate linked selection by allowing selected mutations
-        to appear in the same two regions of the genome: 
-        the one-sixth of the genome immediately before the halfway point,
-        and the last one-sixth of the genome.
-        \textbf{(top)} Constant recombination rate.
-        \textbf{(top middle)} Stepwise varying recombination rate 
-        (as described in \Figure \ref{sfig:slim_sims1}).
-        \textbf{(bottom middle)} Constant recombination rate
-        with spatially varying effects of selection.
-        \textbf{(bottom)} PCA plots corresponding to the highlighted corners 
-        of the last MDS visualization, showing how spatially varying linked selection 
-        has affected patterns of relatedness.
-    } \label{sfig:slim_sims2}
-\end{figure}
-
-\begin{figure}
-    \begin{center}
-       \includegraphics[width=0.82\textwidth]{FigS_Drosophila_missings_against_MDS1_win103}
-    \end{center}
-    \caption{
-        The proportion of data in each window that are missing,
-        compared to the value of the first MDS coordinate
-        for the \textit{Drosophila melanogaster} data from \Figure \ref{fig:mds_allchr}.
-        \label{sfig:dpgp_missingness}
-    }
-\end{figure}
-
-% \begin{figure}
-%     \begin{center}
-%        \includegraphics[width=0.82\textwidth]{FigS_Medicago_missing_against_MDS1_win104}
-%     \end{center}
-%     \caption{
-%         The missingness ratio along genome and the correlation of MDS against missingness in Medicago.
-%         \label{sfig:medicago_missingness}
-%     }
-% \end{figure}
-
-\begin{figure}
-    \begin{center}
-        \includegraphics[width=0.82\textwidth]{simple_example_base_case}
-        \includegraphics[width=0.82\textwidth]{simple_example_missing_data}
-    \end{center}
-    \caption{
-        MDS visualizations of the Gaussian genotypes described in Appendix \ref{apx:gaussian_sims},
-        for 50 individuals from each of three populations.
-        \textbf{(top)} The first quarter, middle half, and final quarter of the chromosome
-        each have different population structure, as expected,
-        despite the possibility for PC switching within each.
-        \textbf{(bottom)} The same picture results even after marking a random 50\% of the genotypes
-        in the first half of the chromosome as missing.
-    } \label{fig:test_missing}.
-\end{figure}
-
 \ifsubmission\processdelayedfloats\fi

### 2018-10-23 writeup/review-responses.tex
start at responses

--- 2018-08-22 writeup/review-responses.tex
@@
 
-We are pleased to submit a revision of our manuscript,
+We are pleased to submit a revised version of our manuscript, 
+``Local PCA shows how the effect of population structure differs along the genome''.
 
@@
 \pagebreak
+\setcounter{page}{1}
 
@@
 \begin{quote}
-    Please understand that incremental changes will not be sufficient.  Adding
-    simulations to strengthen key claims will be necessary, particularly addressing
-    the impacts of mutation rate and recombination rate variation with more depth,
-    the concern regarding PC switching (Reviewer 1), and the concern regarding the
-    impacts of variation in missingness by sub-population (Reviewer 2).
+     Please include a response to each of the reviewers' comments. It is most
+    important that you address the following in a revised manuscript: Reviewer 1's
+    second comment; and more strongly addressing reviewer 2's concern on the issue
+    of the number of PCs included would be helpful as the concern remains. Finally,
+    I add some minor comments, and the senior editor has suggested other
+    discussions in the literature to link to (extensive discussion is not needed).
 \end{quote}
 
-Thanks for the positive feedback and the useful suggestions.
-We agree that more extensive exploration 
-using simulations would help bolster understanding of the method,
-and have now done so.
-This took a substantial amount of work, because genome-scale forwards-time simulations
-with a large number of loci under selection
-is at or beyond the current limits of computation,
-depending on the number of individuals simulated.
+\begin{point}{Line 17.}
+    ``meso-scale'' is vague (and more typically restricted to the physical domain?) - perhaps omit or replace
+\end{point}
+
+\reply{
+    Good point; we've replaced it with ``intermediate-scale'' (maybe better?). \revref
+}
+
+\begin{point}{Line 44.}
+    After ``partially reproductively isolated'' (e.g.
+due to partial post-zygotic incompatibilities) to make it more accessible/clear to a reader following the argument?
+I stumbled at first reading ``random mating'' and ``partially reproductively isolated'' but that apparent contradiction is exactly the point!
+\end{point}
+
+\reply{
+    Another good point; we've got the word ``post-zygotic'' in there now. \revref
+}
+
+\begin{point}{Line 66.}
+    Minor wording : PCA on the genetic data matrix (getting the genetic covariance matrix is an intermediate step in some algorithms); And Menozzi et al used frequency data, so maybe it's best to say inspired by pioneering work by Menozzi et al?
+\end{point}
+
+\reply{
+    We agree, that is more accurate. Thanks; fixed. \revref
+}
+
+\begin{point}{Line 395--396} ``correlations [...] should not extend no further than does linkage disequilibrium'' - I get the point, but it's a tautology as written
+\end{point}
+
+\reply{
+    We've clarified this to ``correlations in patterns of relatedness'',
+    which we think isn't tautological any more,
+    if LD is correlations in allelic identity, anyhow? \revref
+}
+
+\begin{point}{Figures S5 and S6:} 
+    Include a panel which gives a legend/image to explain the colors being used for the bottom PCA such that the reader can know what it will look like when one of the PCAs in that row is mirroring geography.
+\end{point}
+
+\reply{
+    Done.
+}
+
+\begin{point}{Senior editor comments:}
+    It would be helpful to expand the discussion of variation in relationships
+    along the genome - as written, the discussion is strongly focused on the
+    "classical" Drosophila-centred linked selection literature.  Specifically,
+    there has been a lot of discussion in the phylogenetics world about how to
+    estimate species trees from the set of locus-specific genealogies (where the
+    "consensus" tree can be quite wrong); and there are methods that use the
+    distribution of genealogical relationships to estimate population structure
+    (Konrad Lohse, Asger Hobolth etc).  It would be helpful to make these broader
+    links.  
+\end{point}
+
+\reply{
+}
 
@@
 \begin{quote}
-    The paper is generally well written and clear; it addresses an important
-    problem, and clearly makes some progress on it. However, it suffers from having
-    no grounding in either theory or empirical demonstration that it really can find
-    the structures that are claimed. I find the arguments that it finds inversions
-    compelling, though not watertight, and I am not yet convinced that it is finding
-    ubiquitous background selection.  To make this claim, significant extra work is
-    required.
-\end{quote}
+    Li \& Ralph in their manuscript show how local PCA can be used to visualise genomic regions that differ from each other in the mean relatedness among samples.
+While a similar approach has be previously used to detect inversions in human genome (Ma & Amos 2012), they have applied their method to also other than human datasets and invoke other interpretations to the observed patterns than inversions.
 
-\begin{quote}
-    In short, the approach is interesting but not sufficiently explored to produce
-    compelling evidence for the implications that are claimed.  Putting a large
-    amount of effort into simulations may alleviate these concerns somewhat.
+    I appreciate the quality of writing (readability, clarity, organisation) and also the time spent amending the manuscript after the first review.
+The figures were clear and demonstrated the points they should.
+The method was very well explained and the code is understandable to use and easily accessible.
+The manuscript was fun to read.
 
-    Specific points: What does this method find? I'm concerned about:
-    (a) variation in the recombination rate
-    and (b) variation in the mutation rate, creating spurious structure.
-
-    The first possibility is that massively varying information quantity
-    within windows could lead to a small number of such windows having their
-    orientation reversed: that is, PC1 becomes PC2 and vice versa. (Or PC2 and PC3
-    could switch). This would lead to such windows having unusual properties and
-    hence appearing as evidence of an inversion.
-
-    I do agree with the authors that significant outliers would be found at
-    inversions. However, even if the PC switching does not occur, or the model could
-    handle it, the evidence for selection is weaker.  If the two types of variation
-    described above exist, with no selection, I would still expect a ``continuous
-    triangle'' of results (as seen left of Fig 2, top left of Fig 6) with extrema
-    described by windows with the most information, and points placed at different
-    extremum having low recombination rate (because by chance, these will get an
-    approximately fixed local tree, corresponding on average to the genome-wide
-    population structure).
-
-    Addressing this is likely quite hard, though the authors may be able to think of
-    something that separates these effects from selection.
+    I have not previously reviewed the manuscript and but I think that most of the comments of previous reviewers were well reacted upon.
+However, the resulting manuscript is to a new reader a bit unclear in terms of the application and usability of the method and ability to distinguish between different scenarios leading to observed patterns and I think further simulations of mechanisms of interest should be performed with the aim to either demonstrate what the limitations of the method are for potential users or to further explore biological processes hinted at.
 \end{quote}
@@
 \begin{point}{}
-    \ldots (a) variation in the recombination rate and 
-    (b) variation in the mutation rate \ldots creating spurious structure.
+    Is the method meant to facilitate discovery of new inversions?
+If so, what number of samples are necessary?
+Of what coverage (what level of missingness is allowed)?
+What size of inversions is possible to detect?
+What are ranges of the distance values between windows that could point to inversions (any significance test would be recommended)?
+How does the method compare to other methods that detect inversions (esp.
+Ma & Amos 2012)?
+If it is not meant to discover new inversions and it should only confirm known existing inversions, what is the success rate (with varying sample size, missingness)?
 \end{point}
@@
 \reply{
-    Since we only look at SNPs,
-    a 1Kb window with mutation rate $\mu$ and recombination rate $\rho$
-    is equivalent to a 10Kb window with mutation rate $\mu/10$ and recombination rate $\rho/10$,
-    and the ratio $\mu/\rho$ determines the ``signal-to-noise'' ratio.
-    For this reason, varying recombination rate in simulations,
-    and varying window size and type in analysis
-    suffices to check for the effects of both recombination and mutation rate.
-    We have now added simulation tests showing that large-scale variation in recombination rate
-    does not create spurious structure \llname{ll:test_recomb_rate};
-    except in the presence of long regions of low recombination,
-    as expected \llname{ll:hotspot_recomb}
-    The point is addressed by comparing results with windows of different types --
-    windows of equal length in bp (or in SNPs) have different lengths in cM;
-    since these different choices show nearly identical patterns in all cases we have examined,
-    recombination rate variation cannot be driving the results
-    (e.g., compare Figures \ref{fig:mds_allchr} and Supplemental Figure \plr{XXX}).
 }
 
-\begin{point}{PC switching}
-\ldots could lead to a small number of such windows having their
-orientation reversed: that is, PC1 becomes PC2 and vice versa. (Or PC2 and PC3
-could switch).
+\begin{point}{}
+    The authors argue that linked selection is the likely cause for the patterns they observed in Drosophila dataset since the recombination rate correlates with the position of the first MDS coordinate (from line number 321).
+At the same time, the authors mention that the ability to identify the windows could be influenced by recombination rate in the case of linked selection and also in the presence of recombination hotspots (section 3.1 Validation).
+As such, I do not agree with the authors, that it is enough to demonstrate that linked selection could cause these patterns but to use this in interpretation, they should be able to distinguish this explanation from the other possible causes in simulations first and then on real data (e.g.
+using recombination maps for Drosophila to simulate exactly the same regions could be of advantage).
 \end{point}
@@
 \reply{
-    This is a natural concern.
-    However, the only point at which we compare PCs in a way that could be sensitive to ordering
-    is in determining the window size -- in computing the distance between windows
-    we use a measure which is invariant under ordering.
-    We have made this more clear by moving the note about flipping signs of PCs
-    to the appendix on window choice \llname{ll:moved_pc_note}
-    and added more explicit notes about this to \llname{ll:another_pc_point} and \revref.
-    Furthermore, one of our simulation tests
-    is explicitly designed to test for this effect. \llname{ll:pc_switching_test}
 }
 
-\begin{point}{p6}
-``here, we use k=2...''  - you have to show that $k>2$ is the same.
+\begin{point}{}
+    The authors refer to supplementary Figures sometimes as "Figure S6" and sometimes as "Supplementary Figure S6".
+In certain journals, these two could denote different figures so I would select one way only.
 \end{point}
@@
 \reply{
-    We have added a table including correlations between MDS values between $k=2$ and $k=5$
-    for the Medicago data (Table \ref{tab:param_cors}),
-    and a representative figure (for the Drosophila data, $k=5$),
-    see Figure \ref{fig:dmel_k5}. \revref
 }
 
-\begin{point}{p15}
-``We also found nearly identical results when choosing shorter windows of 1,000 SNPs'' - again, show this.
+\begin{point}{}
+    I do not understand while when plotting the positions, the authors opted to show 2 graphs (for each MDS coordinate) when identifying outliers along the genome.
+While it is connected to the MDS plot they also show, I would rather try to express this in 1 value because I think that it would increase their ability to detect the outliers (e.g.
+Figure S6 - the 2 top left figures showing results of simulations with constant recombination rate when compared to MDS coordinate 1 and 2 - if the information from the two graphs was combined in 1 measure, I would expect the pattern to be more visible).
 \end{point}
@@
 \reply{
-    The same table includes correlations between MDS values for different window lengths 
-    (Table \ref{tab:param_cors}). \revref
-}
-
-\begin{point}{p15}
- ``or choosing windows of equal length in base pairs rather than SNPs'' - once again.
-\end{point}
-
-\reply{
-    Again, we address this in Table \ref{tab:param_cors}. \revref
-}
-
-\begin{point}{}
-Using 2 PCs is common practice: only if this is the end of an analysis and the
-PCA was done for visualisation. Here you are using it for something so should
-keep all the relavant PCs.
-\end{point}
-
-\reply{
-    This is a good point; the question is which the ``relevant'' PCs are.
-    \citet{novembre2008interpreting} showed that under isolation by distance,
-    the top two PCs should reflect the two-dimensional nature of the range,
-    and higher PCs are generally much less interpretable;
-    we used $k=2$ with this in mind.
-    We have changed this sentence \revref.
-}
-
-\begin{point}{}
-I'm surprised that PCAdmix isn't referenced. It is using a very similar
-method, albeit with different goals. In particular, the approach of placing all
-points into a single, genome-wide PC space solves many of the problems that this
-approach has (though I agree there may be benefits to the approach described here)
-\end{point}
-
-\reply{
-    Good point: we now reference this work and discuss it's similarity. \revref
 }
@@
 \begin{quote}
-    This is an interesting and well written paper. It was a pleasant read. I have three main general
-    comments:
+    The authors have put in a considerable effort to address the concerns of the reviewers.
+The simulation studies are now much better and the conclusions were weakened somewhat in light of them.
+The method definitely appears to be useful.
+I think that it could be an important part of the literature that moves us forward in the empirical understanding of how the genome evolves.
 \end{quote}
 
-Thanks! We're glad you enjoyed it.
+\begin{point}{}
+I only have one major comment, which is unfortunate because it could have been avoided if the simulation were implemented slightly differently.
+The authors didn't address the intention of my concern about the switching of PCs.
+To be very explicit: when you perform a PC analysis, there is some variability about the order that the "true" directions of the data comes out in terms of the magnitude of that PC (i.e.
+the order).
+This model uses K PCs for each small segment of genome, and each small segment is compared using a method called MDS.
+It is convincing that order doesn't affect comparisons of the segments *provided that all of the components are present*.
+However, if the magnitude of the noise is high, then it is quite plausible that the top K components in a segment of genome are not the top K for the whole genome.
+The simulation is useless for reassuring on this because it sets up a situation with 2 important PCs which are always included, and the claim that "the simulation also verifies insensitivity to ordering of top PCs" is not really true.
+The real data analyses are more reassuring, given the relatively similar results between K=2 and K=5, but I think the concern remains valid.
 
-\begin{point}{Related work:} 
-The authors provide an introduction of the main concepts, as well as some
-intuition of what the method is doing and how, but I found comparison to previous approaches
-to be somewhat missing. To some extent, this is due to the fact that the main goal of their
-analysis is somewhat vaguely ``finding heterogeneity'', which leads to the applications of
-detecting chromosomal inversions and evidence for background selection. It would help to
-have a well defined set of hypotheses, test the method’s accuracy  using simulation (see next
-comment), and compare to previous efforts in similar domains.
+I would like to see evidence that the windows that are found by this method are not associated with "which" PCs happen to be included in the top K.
+To do that, you could for example calculate the top M>K pcs, check the correlation with the overall PCs, and then plot "presence" of a PC (measured in correlation in the top K) against the MDS values.
+It would alternatively be possible to make an argument that it will always be fine if you use a large enough number of PCs, but this is argued to have severe computational problems (an argument I don't know that I buy).
+It would be enough to do this for retaining *one* PC in the simulation that has 2 *equally strong* PCs, but I have a suspicion that this case goes quite wrong and is anyway unrealistic.
+Better to keep more, e.g.
+5 when there is non-trivial structure in the truth, and see that the 4th or 5th swapping with the 6th or 7th doesn't affect much.
+To be clear: the authors shouldn't necessarily follow this route, but they should address the underlying concern.
 \end{point}
@@
 \reply{
-    First: we think that ``finding heterogeneity'' is in fact a well-defined goal,
-    although it was not that well-defined in the paper;
-    we have hopefully improved on this in the Introduction \revref.
-    Expanding a bit more:
-    We strongly agree that methods that seek to test well-defined hypotheses
-    are extremely useful and powerful.
-    We also feel that methods for visualization and exploration are also useful --
-    a primary example here being PCA.
-    If PCA is useful -- and we think that it is --
-    then it should be important to also know how much the thing that PCA is summarizing
-    varies along the genome,
-    in the same way that knowing the mean of some quantity in a population 
-    is only of limited usefulness
-    without also knowing the corresponding population variance.
 }
 
-\begin{point}{Validation:}
-In several occasions, the authors seem to introduce a potential problem in their
-approach, and provide a solution to it. This is generally rather intuitive, but it would really help
-to have simulations of some sort to show that the issue arises and leads to a problem, and that
-their approach does address the specific problem.
+\begin{point}{}
+The simulation with geographically linked selection is not comparable to the simulation with 1/30 beneficial selection and 29/30 negative selection, because the probability of any mutation being positive of negative is now 15/30 (which depends on where in the geography the allele is).
+This makes this scenario a factor 15 higher in positive selection strength.
+The scenarios are ok, but this imbalance likely explains the difference, as positive selection is more likely to stand out than negative selection?
 \end{point}
@@
 \reply{
-    We have added a number of simulations testing various aspects of the method,
-    and have tried to incorporate the results in the paper without cluttering it up
-    with the results of sanity checks.
-    See sections \ref{ss:testing_methods} and \ref{ss:validation};
-    in particular the Gaussian simulations described in appendix \ref{apx:gaussian_sims}.
 }
-
-\begin{point}{}
-The use of weighted PCA to cope with unbalanced sample size could be better demonstrated.
-Although the current explanation makes intuitive sense, this approach does not seem to be
-used in previous work. The authors could design a simulation that supports their approach.
-\end{point}
-
-\reply{
-    We felt that weighted PCA was a nice but unnecessary complication to this work,
-    so have removed it entirely.
-}
-
-\begin{point}{}
-It is conceivable that some subpopulations will have more missingness in some windows. That
-may skew the resulting PCs by selecting different sample sizes for the different windows (as
-discussed in Appendix B). This could distort the PCs, so that variation reflects underlying
-variation in missingness. Would be good to discuss this potential issue and provide simulations.
-\end{point}
-
-\reply{
-    We have tested the method against a large quantity of missing data
-    whereby the amount of missingness does not vary by population \llname{ll:test_missing}.
-    However, missingness that is structured by population could absolutely cause variation in PCs.
-    This could happen, for instance, with RAD data wherein allelic dropout is driven by
-    population-specific polymorphism in the restriction site.
-    The same problem to a lesser degree would be caused by mapping bias,
-    and in fact we see windows around the breakpoints of the polymorphic inversions
-    of the DPGP data having higher rates of missingness.
-    We definately would like our method to be insensitive to missing data,
-    but if missingness is in fact correlated with changes in patterns of relatedness (as in these cases),
-    correlation of MDS scores with amount of missingness is not in fact a problem.
-    As we don't think this point is particularly helpful to the reader, 
-    we have provided a plot of missingness against MDS score (Figure \ref{sfig:dpgp_missingness})
-    and mentioned in the text \revref,
-    but have not added a lengthy discussion.
-}
-
-\begin{point}{Appendix A:}
-when using jackknife to estimate variance, each window is being divided in 10
-``independent'' resampling units. Due to LD, these 10 blocks are likely correlated, which would
-bias the estimates of variance. This is probably not a problem because both signal and noise
-could be equally biased, but the authors may want to consider this potential issue. I wonder if
-the correlation with recombination rate may be partially explained by this.
-\end{point}
-
-\reply{
-    That's true, but we are unaware of a better option,
-    and since the jackknife is only used to pick an appropriate window size,
-    it does not seem crucial.  \revref
-    Since results are the same across different window sizes,
-    we don't think this can explain a correlation wtih recombination rate.
-}
-
-\begin{point}{}
-Is it possible to explain the results of Figure 6 just considering neutral variation in local
-ancestry due to recent admixture? This may explain why ancestry seems to explain a fair
-amount of variance in the lower plots of Fig 6. Local PCA has been previously used by others to
-detect local ancestry blocks, e.g. see the PCAdmix approach by Brisbin et al. The authors
-discuss the possibility that admixture is driving the differentiation, but do not test whether their
-observations agree with neutrality.
-\end{point}
-
-\reply{
-    This is a good point, but any \emph{neutral} variation in local ancestry
-    is not expected to vary systematically along the chromosome,
-    as we now discuss more explicitly \llname{ll:not_neutral}.
-    It well could be that the patterns we see are due to selection acting differentially
-    on introgressed segments,
-    a scenario that would fall under ``linked selection''.
-    We discuss this in the Discussion \revref.
-    We also now cite and discuss the differences to PCAdmix \revreffull{1}{7}.
-}
-
-\begin{point}{}
-``to remove the effect of artifacts such as mutation rate variation, we also rescale each
-approximate covariance matrix to be of similar size (precisely, so that the underlying data
-matrix has trace norm equal to one''. This potential issue is a bit unclear to me, since I would
-expect that scaling the volume of local trees would not result in changed distances in PC
-space. Perhaps the authors could show via simulations that this creates a problem, and that the
-normalization addresses it.
-\end{point}
-
-\reply{
-    We agree that demonstrating this issue would be nice in principle,
-    but for this audience would be getting too far down in the weeds,
-    and including the requested simulations would inflate an already lengthy paper.
-    Additionally, we aim to provide a good method, not to prove that the method
-    is better than all other possible methods.
-    For the reviewer's sake, here is some more about our rationale.
-    Suppose that the covariance matrix is the same along the whole genome,
-    except that it is multiplied by an overall constant that varies.
-    Since PCs are scaled to have norm 1, this overall scaling will indeed not affect the PCs,
-    but we don't compare the PCs directly -- we compare the low-dimensional approximations
-    to the covariance matrix, and the scaling we propose would remove the effect
-    of this overall constant.
-    It is not clear that variation in mutation rate would act like an overall multiple
-    of the covariance matrix,
-    it since we're interested in patterns rather than magnitudes of relatedness,
-    it seems a wise thing to remove.
-}
-
-\begin{point}{Figure 7:}
-    Are MDS coordinates correlated with recombination rates in this case?
-\end{point}
-
-\reply{
-    We made a stab at checking this,
-    and obtained the best version to date of the Medicago recombination map
-    from Tim Paape and Peter Tiffin.
-    There are two versions: a very coarse physical map, and a fine-scale map estimated using LDhat.
-    However, both are on version 3.0 of the assembly,
-    while all other coordinates (sequencing data; gene annotations) are in version 4.0.
-    Furthermore, as Peter Tiffin told us,
-    ``apparently there are no files that translate Mt3.0 to Mt4.0 locations (yes, seems a bit silly)."
-    There is a liftOver chain file for translating 3.5 to 4.0, and
-    ``the differences in the Mt3.0 and Mt3.5 assemblies are, however, apparently relatively minor''.
-    On this basis, we produced the desired figure assuming that Mt3.0 coordinates
-    are the same as Mt3.5 coordinates, included to satisfy the reviewers' curiosity: \\
-    \includegraphics{../medicago/recomb_rates/recomb_and_mds} \\
-    However, given uncertainties in this mapping, the relatively poor match of window sizes,
-    the large number of unmappable windows, and the nature of the recombination data
-    (produced with LDhat, not with actual observations of recombiations),
-    we decided not to include this (but have provided a note, \revref).
-}
-
-\begin{point}{Application:}
-Is what the authors seem to be proposing not already accounted for by linear
-mixed model association approaches? If not, this should be clarified. Either way, this paragraph
-could be dropped.
-\end{point}
-
-\reply{
-    It is not accounted for; we've added some more explanation to this section
-    (which we think should stay in). \revref
-}
-
-\begin{point}{Introduction:}
- ``it is not necessarily clear what aspects of demography should be included in the
-concept.'' I find it a bit weird to describe selection as an ``aspect of demography''. Although it
-could be seen as such within a coalescent framework, that seems to be just a useful
-representation. The authors may consider rewording.
-\end{point}
-
-\reply{
-    Good point, but we think that the sense of unease is perhaps desireable here:
-    certainly differential survival is an aspect of demography;
-    but as the rest of the paragraph goes on to discuss,
-    this quickly spills over into selection, which probably should not be included.
-    We've opted to keep this. \revref
-}
-
-\begin{point}{}
-Paragraph starting in ``Since the definition...''. The notation is a bit unclear. Please check that it
-is clear which PC the text refers to.
-\end{point}
-
-\reply{
-    We've rewritten this section;
-    hopefully it is clearer. \revref
-}
-
-\begin{point}{}
-Would the authors be able to provide a sense for the directionality of effects in Figure 4? It
-would be interesting if the authors tried to further characterize regions that are similar due to
-higher recombination rates. E.g. is there more/less density of polymorphisms in these regions?
-\end{point}
-
-\reply{
-    We agree, that further investigation into these patterns and their correlates
-    would be very interesting and fruitful;
-    however, we think it's outside of the scope of this paper.
-}
-
-\begin{point}{Page 13:}
-typo: ``figures 6 and 6''.
-\end{point}
-
-\reply{Fixed. \revref}
-
-\begin{point}{}
-Typo in abstract, line 6 ``, We show'' $\rightarrow$ ``. We show''.
-\end{point}
-
-\reply{Fixed.}
-
-\begin{point}{}
-Typo: end of introduction ``an visualization''. The whole sentence is a bit weird. The authors just
-stated focus is on clustering, not on looking for outliers, but what does it mean that ``we allow
-ourselves to be surprised by unexpected signals in the data''?
-\end{point}
-
-\reply{
-    We have removed this sentence.
-}
-
-\begin{point}{}
-``There has been substantial debate over the relative impacts of different forms of selection.''
-Citation needed.
-\end{point}
-
-\reply{
-    We have added some relevant recent citations,
-    but apologize for not being aware of the appropriate recent review to refer to.
-    \revref
-}
-
-\begin{point}{}
-``Results using larger numbers of PCs were nearly identical''. It would be interesting to have a
-supplementary table.
-\end{point}
-
-\reply{
-    We have included a table of correlations (Table \ref{tab:param_cors}),
-    as well as the plots for Drosphila data using $k=5$
-    (Figure \ref{fig:dmel_k5}),
-    and hope that readers will take our word on this point more generally.
-}
-
-\begin{point}{}
-Table 1 legend seems a bit redundant. Columns are self-explanatory.
-\end{point}
-
-\reply{
-    Good point; we've cut this down.
-}
-
-\begin{point}{}
-It would help to have numbered lines and references.
-\end{point}
-
-\reply{
-    We've made these changes.
-}

### 2018-10-23 writeup/review-responses_v1.tex
start at responses

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Genetics}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Han Li \\
    \emph{and} Peter Ralph \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are pleased to finally submit a revised version of our manuscript, 
``Local PCA shows how the effect of population structure differs along the genome'',
previously submitted with ID GENETICS/2016/197517.
As you will see, at the reviewers' requests 
we have put substantial effort into additional validation of the method,
including simple simulations to back up some of our mathematical claims about the method
(e.g., about PC switching),
and individual-based simulations with large numbers of loci under weak selection
on realistically-sized chromosomes.
The latter simulations, 
which verify that our method can detect the action of linked selection,
were made possible with tools only recently developed (in part by us).
Our additional tests are described in Section \ref{ss:testing_methods}.

In adding simulation results and the additional verifications requested by the reviewers,
we have endeavored not to expand the manuscript (or the supplementary figures) too much.

Besides theese additional verifications,
we have tried to shift the focus away from background selection specifically
to more general linked selection.
We think it likely that background selection is a contributing factor,
but that the relative contributions of different sorts of selection
on genomic landscapes of diversity, divergence, and relatedness
are almost completely unknown, and constitute a major open question in the field.
We hope this addresses Reviewer 1's skepticism regarding background selection.

In addition to the main document, with line numbers and hyperlinked references
from the responses to reviews,
we have also uploaded a version highlighting the changes between this version
and the submitted one (note that the submission system doesn't provide an appropriate category
for this file upload, so it is also marked ``Response to Reviewers'').

We hope you find the manuscript to be improved as much as we do.

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
Han Li and
Peter Ralph
}\\
\end{minipage}

\vskip 4em

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
    Adding simulations to strengthen key claims will be necessary, particularly
    addressing the impacts of mutation rate and recombination rate variation
    with more depth, the concern regarding PC switching (Reviewer 1), and the
    concern regarding the impacts of variation in missingness by sub-population
    (Reviewer 2).
\end{quote}

Thanks for the positive feedback and the useful suggestions.
We agree that more extensive exploration 
using simulations would help bolster understanding of the method,
and have now done so.
See below for more detailed responses.

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    The paper is generally well written and clear; it addresses an important
    problem, and clearly makes some progress on it. However, it suffers from having
    no grounding in either theory or empirical demonstration that it really can find
    the structures that are claimed. I find the arguments that it finds inversions
    compelling, though not watertight, and I am not yet convinced that it is finding
    ubiquitous background selection.  To make this claim, significant extra work is
    required.

    In short, the approach is interesting but not sufficiently explored to produce
    compelling evidence for the implications that are claimed.  Putting a large
    amount of effort into simulations may alleviate these concerns somewhat.
\end{quote}

We have backed off somewhat on our previous speculation that the observed patterns
are due to background selection,
and have also put a large amount of effort into simulations (discussed below).
We hope these alleviate your concerns.

\begin{quote}
    Specific points: What does this method find? I'm concerned about:
    (a) variation in the recombination rate
    and (b) variation in the mutation rate, creating spurious structure.

    The first possibility is that massively varying information quantity
    within windows could lead to a small number of such windows having their
    orientation reversed: that is, PC1 becomes PC2 and vice versa. (Or PC2 and PC3
    could switch). This would lead to such windows having unusual properties and
    hence appearing as evidence of an inversion.

    I do agree with the authors that significant outliers would be found at
    inversions. However, even if the PC switching does not occur, or the model could
    handle it, the evidence for selection is weaker.  If the two types of variation
    described above exist, with no selection, I would still expect a ``continuous
    triangle'' of results (as seen left of Fig 2, top left of Fig 6) with extrema
    described by windows with the most information, and points placed at different
    extremum having low recombination rate (because by chance, these will get an
    approximately fixed local tree, corresponding on average to the genome-wide
    population structure).

    Addressing this is likely quite hard, though the authors may be able to think of
    something that separates these effects from selection.
\end{quote}

Specific points are addressed below.
Briefly, switching of PC 1 and 2 cannot mislead the method
because we measure distances in a way that does not depend on PC ordering.
It is possible that switching of PC 2 and 3 could mislead if we are using only $k=2$ of the PCs;
however, we see the same patterns when using more PCs, so we know this is not happening.

As for inversions: we aren't claiming that every strong localized signal discovered by this method
is an inversion (and we've tried to make sure the paper doesn't imply that!),
but think we have good evidence for the taxa we study.
We hope the reviewer agrees that we find the large inversions of \emph{Drosophila},
where (a) PCA structure coincides with known inversion breakpoints, 
and (b) the local PCA plots separate samples carrying the two inversion orientations.
Nearly all of the inversion-like signals we find in human are likewise located at the positions 
of known polymorphic inversions; 
however, we agree that the evidence for remaining inversion is \emph{not} watertight,
because, as the reviwer points out (and we do as well),
we cannot distinguish them from long regions of low recombination.
The fact that nearly all of the strong, localized signals coincide with known polymorphic inversions
induces us to emphasize inversions.
We now more clearly emphasize this caveat in the Discussion \llname{ll:inversion_caveat}.

\begin{point}{}
    \ldots (a) variation in the recombination rate and 
    (b) variation in the mutation rate \ldots creating spurious structure.
\end{point}

\reply{
    We have now added simulation tests showing that large-scale variation in recombination rate
    does not create spurious structure \llname{ll:test_recomb_rate};
    except in the presence of long regions of low recombination,
    which we point is is as expected \llname{ll:hotspot_recomb}.
    We also check for the effect of this in the real data:
    since we only look at SNPs,
    a 1Kb window with mutation rate $\mu$ and recombination rate $\rho$
    is equivalent to a 10Kb window with mutation rate $\mu/10$ and recombination rate $\rho/10$,
    and the ratio $\mu/\rho$ determines the ``signal-to-noise'' ratio.
    Therefore, we can check for effects due to recombination
    and mutation rate variation 
    by comparing results with windows of different types --
    windows of equal length in bp, or in SNPs, have different lengths in cM.
    These different choices show nearly identical patterns in all cases we have examined,
    as is summarized in Table \ref{tab:window_sizes},
    making it unlikely that recombination or mutation rate variation could be driving the results.
}

\begin{point}{PC switching}
\ldots could lead to a small number of such windows having their
orientation reversed: that is, PC1 becomes PC2 and vice versa. (Or PC2 and PC3
could switch).
\end{point}

\reply{
    This is a natural concern.
    However, the only point at which we compare PCs in a way that could be sensitive to ordering
    is in determining the window size -- we compute the distance between windows
    in a way that is invariant under ordering.
    We have made this more clear by moving the note about flipping signs of PCs
    to the appendix on window choice \llname{ll:moved_pc_note}
    and added more explicit notes about this to \llname{ll:another_pc_point} and \revref.
    Furthermore, one of our simulation tests
    is explicitly designed to test for this effect. \llname{ll:pc_switching_test}
}

\begin{point}{p6}
``here, we use k=2...''  - you have to show that $k>2$ is the same.
\end{point}

\reply{
    We have added a table including correlations between MDS values between $k=2$ and $k=5$
    for the Medicago data (Table \ref{tab:param_cors}),
    and a representative figure (for the Drosophila data, $k=5$),
    see Figure \ref{sfig:dmel_k5}, \revref and \llname{rr:rev1:4}.
}

\begin{point}{p15}
``We also found nearly identical results when choosing shorter windows of 1,000 SNPs'' - again, show this.
\end{point}

\reply{
    The same table includes correlations between MDS values for different window lengths 
    (Table \ref{tab:param_cors}). \revref
}

\begin{point}{p15}
 ``or choosing windows of equal length in base pairs rather than SNPs'' - once again.
\end{point}

\reply{
    Again, we address this in Table \ref{tab:param_cors}. \revref
}

\begin{point}{}
Using 2 PCs is common practice: only if this is the end of an analysis and the
PCA was done for visualization. Here you are using it for something so should
keep all the relevant PCs.
\end{point}

\reply{
    This is a good point; the question is which the ``relevant'' PCs are.
    \citet{novembre2008interpreting} showed that under isolation by distance,
    the top two PCs should reflect the two-dimensional nature of the range,
    and higher PCs are generally much less interpretable;
    we used $k=2$ with this in mind.
    We have changed this sentence \revref.
}

\begin{point}{}
I'm surprised that PCAdmix isn't referenced. It is using a very similar
method, albeit with different goals. In particular, the approach of placing all
points into a single, genome-wide PC space solves many of the problems that this
approach has (though I agree there may be benefits to the approach described here)
\end{point}

\reply{
    Good point: we now reference this work and discuss it's similarity. \revref
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    This is an interesting and well written paper. It was a pleasant read. I have three main general
    comments:
\end{quote}

Thanks! We're glad you enjoyed it.

\begin{point}{Related work:} 
The authors provide an introduction of the main concepts, as well as some
intuition of what the method is doing and how, but I found comparison to previous approaches
to be somewhat missing. To some extent, this is due to the fact that the main goal of their
analysis is somewhat vaguely ``finding heterogeneity'', which leads to the applications of
detecting chromosomal inversions and evidence for background selection. It would help to
have a well defined set of hypotheses, test the method's accuracy using simulation (see next
comment), and compare to previous efforts in similar domains.
\end{point}

\reply{
    First: we think that ``finding heterogeneity'' is in fact a well-defined goal,
    although it was not that well-defined in the paper;
    we have hopefully improved on this in the Introduction \revref.
    Expanding a bit more:
    We agree that methods that seek to test well-defined hypotheses
    are extremely useful and powerful.
    We feel that methods for visualization and exploration are also useful --
    a primary example here being PCA (as used for visualization).
    If PCA is useful -- or more generally, dimension reduction --
    then it should be important to also know how much the thing that PCA is summarizing
    varies along the genome,
    in the same way that knowing the mean of some quantity in a population 
    is only of limited usefulness
    without also knowing the corresponding population variance.
}

\begin{point}{Validation:}
In several occasions, the authors seem to introduce a potential problem in their
approach, and provide a solution to it. This is generally rather intuitive, but it would really help
to have simulations of some sort to show that the issue arises and leads to a problem, and that
their approach does address the specific problem.
\end{point}

\reply{
    We have added a number of simulations testing various aspects of the method,
    and have tried to incorporate the results in the paper without cluttering it up
    with the results of sanity checks.
    See sections \ref{ss:testing_methods} and \ref{ss:validation};
    in particular the Gaussian simulations described in appendix \ref{apx:gaussian_sims}.
}

\begin{point}{}
The use of weighted PCA to cope with unbalanced sample size could be better demonstrated.
Although the current explanation makes intuitive sense, this approach does not seem to be
used in previous work. The authors could design a simulation that supports their approach.
\end{point}

\reply{
    We felt that weighted PCA was a nice but unnecessary complication to this work,
    so have removed it entirely.
}

\begin{point}{}
It is conceivable that some subpopulations will have more missingness in some windows. That
may skew the resulting PCs by selecting different sample sizes for the different windows (as
discussed in Appendix B). This could distort the PCs, so that variation reflects underlying
variation in missingness. Would be good to discuss this potential issue and provide simulations.
\end{point}

\reply{
    We have tested the method against a large quantity of missing data
    in which the amount of missingness does not vary by population \llname{ll:test_missing}.
    However, missingness that is structured by population could absolutely cause variation in PCs.
    This could happen, for instance, with RAD data wherein allelic dropout is driven by
    population-specific polymorphism in the restriction site.
    The same problem to a lesser degree would be caused by mapping bias,
    and in fact we see windows around the breakpoints of the polymorphic inversions
    of the DPGP data having higher rates of missingness.
    We definitely would like our method to be insensitive to missing data,
    but if missingness is in fact correlated with the signal we are looking for 
    (changes in patterns of relatedness, as it is in these cases),
    correlation of MDS scores with amount of missingness is not in fact a problem.
    As we don't think this point is particularly helpful to the reader, 
    we have provided a plot of missingness against MDS score (Figure \ref{sfig:dpgp_missingness})
    and mentioned it in the text \revref,
    but have not added a lengthy discussion.
}

\begin{point}{Appendix A:}
when using jackknife to estimate variance, each window is being divided in 10
``independent'' resampling units. Due to LD, these 10 blocks are likely correlated, which would
bias the estimates of variance. This is probably not a problem because both signal and noise
could be equally biased, but the authors may want to consider this potential issue. I wonder if
the correlation with recombination rate may be partially explained by this.
\end{point}

\reply{
    That's true, but we are unaware of a better option,
    and since the jackknife is only used to pick an appropriate window size,
    it does not seem crucial.  {\revref}
    Since results are the same across different window sizes,
    we don't think this can explain a correlation with recombination rate.
}

\begin{point}{}
Is it possible to explain the results of Figure 6 just considering neutral variation in local
ancestry due to recent admixture? This may explain why ancestry seems to explain a fair
amount of variance in the lower plots of Fig 6. Local PCA has been previously used by others to
detect local ancestry blocks, e.g. see the PCAdmix approach by Brisbin et al. The authors
discuss the possibility that admixture is driving the differentiation, but do not test whether their
observations agree with neutrality.
\end{point}

\reply{
    This is a good point, but any \emph{neutral} variation in local ancestry
    is not expected to vary systematically along the chromosome,
    as we now discuss more explicitly \llname{ll:not_neutral}.
    It well could be that the patterns we see are due to selection acting differentially
    on introgressed segments,
    a scenario that would fall under ``linked selection''.
    We discuss this in the Discussion \revref.
    We also now cite and discuss the differences to PCAdmix \revreffull{1}{7}.
}

\begin{point}{}
``to remove the effect of artifacts such as mutation rate variation, we also rescale each
approximate covariance matrix to be of similar size (precisely, so that the underlying data
matrix has trace norm equal to one''. This potential issue is a bit unclear to me, since I would
expect that scaling the volume of local trees would not result in changed distances in PC
space. Perhaps the authors could show via simulations that this creates a problem, and that the
normalization addresses it.
\end{point}

\reply{
    We agree that demonstrating this issue would be nice in principle,
    but think that for this audience would be getting too far down in the weeds.
    (Besides, we aim to provide a good method, not to prove that the method
    is better than all other possible methods.)
    For the reviewer's sake, here is some more about our rationale.
    Suppose that the covariance matrix is the same along the whole genome,
    except that it is multiplied by an overall constant that varies.
    Since PCs are scaled to have norm 1, this overall scaling will indeed not affect the PCs,
    but we don't compare the PCs directly -- we compare the low-dimensional approximations
    to the covariance matrix, and the scaling we propose would remove the effect
    of this overall constant.
    It is not clear that variation in mutation rate would act like an overall multiple
    of the covariance matrix, but
    since we're interested in patterns rather than magnitudes of relatedness,
    it seems a wise thing to remove.
}

\begin{point}{Figure 7:}
    Are MDS coordinates correlated with recombination rates in this case?
\end{point}

\reply{
    We made a stab at checking this,
    and obtained the best version to date of the Medicago recombination map
    from Tim Paape and Peter Tiffin.
    There are two versions: a very coarse physical map, and a fine-scale map estimated using LDhat.
    However, both are on version 3.0 of the assembly,
    while all other coordinates (sequencing data; gene annotations) are in version 4.0.
    Furthermore, as Peter Tiffin told us,
    ``apparently there are no files that translate Mt3.0 to Mt4.0 locations (yes, seems a bit silly)."
    There is a liftOver chain file for translating 3.5 to 4.0, and
    ``the differences in the Mt3.0 and Mt3.5 assemblies are, however, apparently relatively minor''.
    On this basis, we produced the desired figure assuming that Mt3.0 coordinates
    are the same as Mt3.5 coordinates, included to satisfy the reviewers' curiosity: \\
    \includegraphics{../medicago/recomb_rates/recomb_and_mds} \\
    However, given uncertainties in this mapping, the relatively poor match of window sizes,
    the large number of unmappable windows, and the nature of the recombination data
    (produced with LDhat, not with actual observations of recombinations),
    we decided not to include this (but have provided a note, \revref).
}

\begin{point}{Application:}
Is what the authors seem to be proposing not already accounted for by linear
mixed model association approaches? If not, this should be clarified. Either way, this paragraph
could be dropped.
\end{point}

\reply{
    It is not accounted for; we've added some more explanation to this section
    (which we think should stay in). \revref
}

\begin{point}{Introduction:}
 ``it is not necessarily clear what aspects of demography should be included in the
concept.'' I find it a bit weird to describe selection as an ``aspect of demography''. Although it
could be seen as such within a coalescent framework, that seems to be just a useful
representation. The authors may consider rewording.
\end{point}

\reply{
    Good point, but we think that the sense of unease is perhaps desirable here:
    certainly differential survival is an aspect of demography;
    but as the rest of the paragraph goes on to discuss,
    this quickly spills over into selection, which probably should not be included.
    We've opted to keep this. \revref
}

\begin{point}{}
Paragraph starting in ``Since the definition...''. The notation is a bit unclear. Please check that it
is clear which PC the text refers to.
\end{point}

\reply{
    We've rewritten this section;
    hopefully it is clearer. \revref
}

\begin{point}{}
Would the authors be able to provide a sense for the directionality of effects in Figure 4? It
would be interesting if the authors tried to further characterize regions that are similar due to
higher recombination rates. E.g.\ is there more/less density of polymorphisms in these regions?
\end{point}

\reply{
    We agree, that further investigation into these patterns and their correlates
    would be very interesting and fruitful;
    however, we think it's outside of the scope of this paper.
}

\begin{point}{Page 13:}
typo: ``figures 6 and 6''.
\end{point}

\reply{Fixed. \revref}

\begin{point}{}
Typo in abstract, line 6 ``, We show'' $\rightarrow$ ``. We show''.
\end{point}

\reply{Fixed.}

\begin{point}{}
Typo: end of introduction ``an visualization''. The whole sentence is a bit weird. The authors just
stated focus is on clustering, not on looking for outliers, but what does it mean that ``we allow
ourselves to be surprised by unexpected signals in the data''?
\end{point}

\reply{
    We have removed this sentence.
}

\begin{point}{}
``There has been substantial debate over the relative impacts of different forms of selection.''
Citation needed.
\end{point}

\reply{
    We have added some relevant recent citations,
    but apologize for not being aware of the appropriate recent review to refer to.
    \revref
}

\begin{point}{}
``Results using larger numbers of PCs were nearly identical''. It would be interesting to have a
supplementary table.
\end{point}

\reply{
    We have included a table of correlations (Table \ref{tab:param_cors}),
    as well as the plots for Drosophila data using $k=5$
    (Figure \ref{sfig:dmel_k5}),
    and hope that readers will take our word on this point more generally.
}

\begin{point}{}
Table 1 legend seems a bit redundant. Columns are self-explanatory.
\end{point}

\reply{
    Good point; we've cut this down.
}

\begin{point}{}
It would help to have numbered lines and references.
\end{point}

\reply{
    Now provided.
}

### 2018-10-23 writeup/local_pca_paper.tex
biorxiv latex
start at responses

@@
 \input{review-response-commands}
+\newif\ifbiorxiv
+\biorxivfalse
+% \biorxivtrue
+\ifbiorxiv
+    \renewcommand{\revpoint}[2]{\relax}
+    \renewcommand{\llabel}[1]{\relax}
+\fi
 % set this to show line numbers and include responses to reviews or not
@@
 % \reviewresponsesfalse  % don't include them
+\ifbiorxiv\reviewresponsesfalse\fi
 \newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file
@@
 \vspace*{0.35in}
+
+\ifbiorxiv\nolinenumbers\fi
 
@@
 
-\ifsubmission\linenumbers\fi
-
 \section*{Abstract}
@@
 We show how to use local principal components analysis to describe this 
-meso-scale heterogeneity in patterns of relatedness,
+intermediate-scale heterogeneity in patterns of relatedness, \revpoint{AE}{1}
 and apply the method to genomic data from three species,
@@
 as for instance reduced mean relatedness between individuals from distinct populations.
-However, it is not necessarily clear what aspects of demography should be included in the concept. \revpoint{2}{10}
+However, it is not necessarily clear what aspects of demography should be included in the concept. % \revpoint{2}{10}
 For instance, 
@@
 On closer examination, incorporating differential survival or fecundity makes the concept less clear:
-should a randomly mating population consisting of two types that are partially reproductively isolated from each other
+should a randomly mating population consisting of two types 
+that exhibit partial post-zygotic reproductive isolation from each other \revpoint{AE}{2}
 be said to show population structure or not?
@@
 by applying principal components analysis (PCA, \citet{patterson2006population}) 
-to the genetic covariance matrix,
-as pioneered by \citet{menozzi1978synthetic}.
+to the genetic data matrix,
+as inspired by the pioneering work of \citet{menozzi1978synthetic}. \revpoint{AE}{3}
 % About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}.
@@
 but by projecting each genomic window onto the axes of a single, globally-defined PCA
-rather than doing PCA separately on each window.  \revpoint{1}{7}
+rather than doing PCA separately on each window.  % \revpoint{1}{7}
 
@@
 correlation of this variation with known genomic features
-can then be used to uncover its source. \revpoint{2}{1}
+can then be used to uncover its source. % \revpoint{2}{1}
 
@@
 for ease of visualization we usually only use the first two (referred to as $\pcone$ and $\pctwo$),
-and check that results hold using more. \revpoint{1}{6}
+and check that results hold using more. % \revpoint{1}{6}
 The above procedure can be performed on any subset of the data;
@@
 obtained using the top $k$ PCs,
-which is invariant under reordering of the PCs, \revpoint{1}{2}
+which is invariant under reordering of the PCs, % \revpoint{1}{2}
 reflections, and rotations and yet contains all other information about the PCs.
@@
 The patterns did not depend on the number of PCs used
-(see \Figure \ref{sfig:dmel_k5} for the same plot with $k=5$ PCs), \revpoint{1}{3}
+(see \Figure \ref{sfig:dmel_k5} for the same plot with $k=5$ PCs), % \revpoint{1}{3}
 and are only weakly correlated with variation in missingness 
-(see \Figure \ref{sfig:dpgp_missingness}). \revpoint{2}{4}
+(see \Figure \ref{sfig:dpgp_missingness}). % \revpoint{2}{4}
 
@@
 we applied the method to all chromosomes together.
-Results for chromosome 3 are shown in \Figure \ref{fig:mds12_medicago}, \revpoint{2}{13}
+Results for chromosome 3 are shown in \Figure \ref{fig:mds12_medicago}, % \revpoint{2}{13}
 and other chromosomes are similar:
@@
 to Mt4.0 coordinates (although those in \citet{paape2012finescale} appear visually similar).
-\revpoint{2}{8}
+% \revpoint{2}{8}
 
@@
 and number of PCs, as shown in Table \ref{tab:param_cors}.
-\revpoint{1}{4} \revpoint{1}{5}
+% \revpoint{1}{4} \revpoint{1}{5}
 % We also found nearly identical results when choosing shorter windows of 1,000 SNPs
@@
 in the \textit{Medicago truncatula} and \textit{Drosophila melanogaster} datasets, \llabel{ll:not_neutral}
-because correlations induced by neutral processes 
+because correlations in patterns of relatedness induced by neutral processes \revpoint{AE}{4}
 should extend no further than does linkage disequilibrium
@@
 (e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining,stankowski2018tempo}).
-\revpoint{2}{16}
+% \revpoint{2}{16}
 These have been difficult to disentangle in part because for the most part
@@
 account for this with either a single, genome-wide kinship matrix
-or one constructed using only sites unlinked to the focal SNP. \revpoint{2}{9}   %unlike ---> unlink?
+or one constructed using only sites unlinked to the focal SNP. % \revpoint{2}{9}   %unlike ---> unlink?
 Since the effects of population structure is not constant along the genome,
@@
 and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
-averaged over samples. \revpoint{2}{11}
+averaged over samples. % \revpoint{2}{11}
 The definition of eigenvectors does not specify their sign,
@@
 and $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
-We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}: \revpoint{2}{5}
+We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}: % \revpoint{2}{5}
 we divide the $j^\text{th}$ window into 10 equal-sized pieces,
@@
         similar to an inversion.
+        The \textbf{(inset)} provides a key to the locations of the individuals
+        on the spatial lanscape.
     } \label{sfig:slim_sims1}
@@
         has affected patterns of relatedness.
+        The \textbf{(inset)} provides a key to the locations of the individuals
+        on the spatial lanscape.
     } \label{sfig:slim_sims2}

### 2018-10-23 writeup/make_pop_legend.R

@@ -1,0 +1,16 @@
+# Make an inset legend for the "corner pca" plots with space.
+library(colorspace)
+
+grid.width <- 4
+grid.mat <- matrix(1:(grid.width^2), nrow=grid.width)
+pop.cols <- rainbow_hcl(grid.width^2)
+pop.pch <- seq_len(grid.width^2)
+
+pdf(file="inset_pop_key.pdf", width=1, height=1, pointsize=10)
+par(mar=c(0,0,0,0))
+    plot(row(grid.mat), col(grid.mat),
+         cex=3, pch=pop.pch, col=pop.cols, lwd=2,
+         xlab='', ylab='', xaxt='n', yaxt='n',
+         xlim=c(0.5, grid.width+0.5),
+         ylim=c(0.5, grid.width+0.5) )
+dev.off()

### 2018-10-26 writeup/review-responses.tex
responses

@@
 \reply{
+    Good suggestion; we've added (hopefully helpful) additional discussion.
+    \revref and \revreffull{1}{2}
 }
@@
     Li \& Ralph in their manuscript show how local PCA can be used to visualise genomic regions that differ from each other in the mean relatedness among samples.
-While a similar approach has be previously used to detect inversions in human genome (Ma & Amos 2012), they have applied their method to also other than human datasets and invoke other interpretations to the observed patterns than inversions.
+While a similar approach has be previously used to detect inversions in human genome (Ma \& Amos 2012), they have applied their method to also other than human datasets and invoke other interpretations to the observed patterns than inversions.
 
@@
 The manuscript was fun to read.
+\end{quote}
 
+Thanks!  We put a lot of time into the work and the writing,
+and that's really nice to hear.
+
+\begin{quote}
     I have not previously reviewed the manuscript and but I think that most of the comments of previous reviewers were well reacted upon.
@@
 \end{quote}
+
+A similar point was raised by another reviewer in the previous round.
+It might be reasonable, as \citet{ma2012investigation} did,
+to use these results to develop an inference method for a particular signal.
+We have rather chosen to follow the lead of genomic PCA,
+which, despite having connections to demographic quantities,
+is most widely-used as a purely descriptive tool,
+for communicating population genetics resuls.
+We've tried to make this a bit more clear in the introduction. \llname{ll:descriptive}
 
@@
 How does the method compare to other methods that detect inversions (esp.
-Ma & Amos 2012)?
+Ma \& Amos 2012)?
 If it is not meant to discover new inversions and it should only confirm known existing inversions, what is the success rate (with varying sample size, missingness)?
@@
 \reply{
+    These are excellent questions for a method designed to detect inversions.
+    However, we do not present such a method,
+    we rather present a visualization method in which, empirically, inversions often leave a large (but not quantified) signal.
+    We are not sure if it would be productive to turn this into a quantitative method, either:
+    as we have now clarified in the paper \llname{ll:inversion_caveat},
+    distinguishing inversions from regions of low recombination
+    likely requires other sources of information.
 }
@@
 At the same time, the authors mention that the ability to identify the windows could be influenced by recombination rate in the case of linked selection and also in the presence of recombination hotspots (section 3.1 Validation).
-As such, I do not agree with the authors, that it is enough to demonstrate that linked selection could cause these patterns but to use this in interpretation, they should be able to distinguish this explanation from the other possible causes in simulations first and then on real data (e.g.
-using recombination maps for Drosophila to simulate exactly the same regions could be of advantage).
+As such, I do not agree with the authors, that it is enough to demonstrate that linked selection could cause these patterns but to use this in interpretation, they should be able to distinguish this explanation from the other possible causes in simulations first and then on real data (e.g., using recombination maps for Drosophila to simulate exactly the same regions could be of advantage).
 \end{point}
@@
 \reply{
+    We don't share the reviewer's concern about variation in recombination rate:
+    a region of low recombination could be dominated by a single tree,
+    and therefore provide a strongly divergent PCA signal,
+    there is no reason this would provide a \emph{consistent} signal
+    across a large fraction of a chromosome,
+    unless that chromosomal region had much lower recombination rate than we see here.
+    More generally, we heartily agree that we would love to have available a test
+    that could prove to what degree these (or any) varying patterns of relatedness along genomes
+    are due to linked selection, demographic noise, or some other force.
+    However, this is one of the major unsolved question in the field,
+    and is subject to substantial debate 
+    (see for instance \citet{jensen2014unfounded,harris2018unfounded}
+    and subsequent rejoinders).
+    We've added more to the discussion on this. \revref
 }
@@
 \reply{
+    Good point; that's fixed.
 }
@@
 \reply{
+    That's a good suggestion -- perhaps, to test for (or, visualize) outlierness,
+    one could plot the deviation from the MDS centroid.
+    We chose to plot each MDS coordinate separately
+    so as to better see how the structure of the MDS plot lays out along the genome.
+    For instance, in Figure \ref{fig:mds_allchr} on chromosome arm 3L,
+    it is interesting that the green arm of the MDS plot localizes to the endpoints of the inversion,
+    while the purple arm localizes to near the centromere.
+    If only a measure of outlierness were plotted,
+    it would be more difficult to discern patterns like this,
+    and therefore develop hypotheses explaining the observed patterns.
 }
@@
 I would like to see evidence that the windows that are found by this method are not associated with "which" PCs happen to be included in the top K.
-To do that, you could for example calculate the top M>K pcs, check the correlation with the overall PCs, and then plot "presence" of a PC (measured in correlation in the top K) against the MDS values.
+To do that, you could for example calculate the top $M>K$ pcs, check the correlation with the overall PCs, and then plot "presence" of a PC (measured in correlation in the top K) against the MDS values.
 It would alternatively be possible to make an argument that it will always be fine if you use a large enough number of PCs, but this is argued to have severe computational problems (an argument I don't know that I buy).
@@
 \reply{
+    Apologies for not realizing the precise nature of your original concern.
+    Yes, we agree that if strongly different PCs switched order 
+    right at the boundary of which PCs were included,
+    then that could easily show up as (spurious) structure in our MDS plots.
+    We have added discussion of this point. \revref
+    However, for our empirical data we think we have strong evidence that the differences between windows 
+    are not being driven by this effect:
+    as the reviewer also mentions, the effect is expected to dissappear when using a larger number of PCs,
+    which does not happen in our applications.
+    Furthermore, and speculatively,
+    our method of choosing a window size looks for a big enough window that the PCs within a window 
+    are relatively stable,
+    and so might help avoid sensitive PC switching.
 }
@@
 \reply{
+    Good point;
+    we've added a note about this. \revref
 }

### 2018-10-26 writeup/local_pca_paper.tex
responses

@@
     \newcommand{\Figures}{Fig }
+    \newcommand{\SFigure}{Sup Fig }
 \else
@@
     \newcommand{\Figures}{{Figures }}
+    \newcommand{\SFigure}{{Supplementary Figure }}
     % \renewcommand{\llabel}[1]{}
@@
 our method then describes how this structure is affected by selection and other factors.
-Our aim is not to identify outlier loci,
-but rather to describe larger-scale variation shared by many parts of the genome;
-correlation of this variation with known genomic features
-can then be used to uncover its source. % \revpoint{2}{1}
+The method is descriptive:
+it does not aim to identify outlier loci,
+but rather to describe larger-scale variation shared by many parts of the genome,
+and give clues about its source. \llabel{ll:descriptive}
+% If such variation is found, correlation of this variation with known genomic features
+% can then be used to uncover its source. % \revpoint{2}{1}
 
@@
 We excluded the sex chromosomes and the mitochondria.
-PCA plots for each chromosome, separately, are shown in \Figure \ref{sfig:pca_human_allchr}.
+PCA plots for each chromosome, separately, are shown in \SFigure \ref{sfig:pca_human_allchr}.
 % We use the allele that has highest frequency in the samples as the reference allele for each position.
@@
 Each of the 8 chromosomes has 3--5 million SNPs;
-PCA plots for these are shown in \Figure \ref{sfig:pca_medicago_allchr}.
+PCA plots for these are shown in \SFigure \ref{sfig:pca_medicago_allchr}.
 We did not use the mitochondria or chloroplasts.
@@
 Simulations with linked selection produced mixed results
-(Figure \ref{sfig:slim_sims2}).
-The method strongly identified the regions under spatially varying linked selection.
+(\SFigure \ref{sfig:slim_sims2}).
+The method strongly identified the regions under spatially varying linked selection
+(also our simulation with strongest positive selection). \revpoint{2}{2}
 It also identified the regions (although less unambiguously)
@@
 The patterns did not depend on the number of PCs used
-(see \Figure \ref{sfig:dmel_k5} for the same plot with $k=5$ PCs), % \revpoint{1}{3}
+(see \SFigure \ref{sfig:dmel_k5} for the same plot with $k=5$ PCs), % \revpoint{1}{3}
 and are only weakly correlated with variation in missingness 
-(see \Figure \ref{sfig:dpgp_missingness}). % \revpoint{2}{4}
+(see \SFigure \ref{sfig:dpgp_missingness}). % \revpoint{2}{4}
 
@@
 and found that, remarkably, 
-the inversion on chromosome 8 is still the most striking outlying signal (\Figure \ref{sfig:mds1_along_allchr_human}). 
+the inversion on chromosome 8 is still the most striking outlying signal (\SFigure \ref{sfig:mds1_along_allchr_human}). 
 Further investigation with a denser set of SNPs,
@@
 
+\paragraph{Genealogical noise?} \revpoint{AE}{6}
+The field of phylogenetics has long had to deal with the fact that 
+there can be a great number of different local phylogenies along the genome,
+even between species
+\citep{avise1983mitochondrial,pamilo1988relationships,hobolth2007genomic}.
+Loosely, we are studying the within-species patterns that might lead to such
+incomplete lineage sorting in the future.
+The neutral distribution of variation in these patterns 
+has been used to infer demographic history,
+both between species \citep{slatkin2006concordance}
+and within species \citep{beeravolu2018blockwise}.
+If these distinct phylogenies are merely a result of neutral stochasticity,
+there is not expected to be a correlation between local phylogeny
+and other genomic features.
+However, in some cases the local assortment of ancestral diversity 
+and subsequent introgression between sister taxa
+shows clear signs of selection, 
+as for instance in the wild tomato clade \citep{pease2016phylogenomics}.
+
 \paragraph{The effect of selection}
@@
 and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
-It may be possible to design more powerful statistics that make stronger use of spatial information.
+It may be possible to design more powerful statistics that make stronger use of spatial information,
+or even using the variation in relatedness that we observe here,
+similar to the method of \citet{beeravolu2018blockwise}. \revpoint{1}{2}
 
@@
 but in our exploration, changing these has not produced dramatically different results.
-These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
+However, this choice could easily be important:
+for instance, if the $k^\text{th}$ and $(k+1)^\text{st}$ PCs 
+are sufficiently different but have similar eigenvalues,
+then small amounts of noise could cause these to switch,
+leading to spuriously inferred differences between windows in which one or the other
+was included in the top $k$ PCs.
+This does not seem to be a problem in our applications,
+as changing the number of PCs did not affect the qualitative results. \revpoint{2}{1}
+These choices are all part of more general techniques in dimension reduction and high-dimensional data visualization;
 we encourage the user to experiment.
@@
 running our method on a genome with 99 windows of 400 loci each
-produced the first plot in Figure \ref{sfig:test_missing}.
+produced the first plot in \SFigure \ref{sfig:test_missing}.
 These matrices are chosen so that the top two eigenvalues $\Sigma$ are the same (both 50.1),
@@
 as missing, independently, with probability $1/2$ and ran our method again,
-producing the second plot of Figure \ref{sfig:test_missing}. \llabel{ll:test_missing}
+producing the second plot of \SFigure \ref{sfig:test_missing}. \llabel{ll:test_missing}
 If our method was influenced by missing data,

### 2018-10-29 writeup/review-responses.tex
revise update

@@
 is most widely-used as a purely descriptive tool,
-for communicating population genetics resuls.
+for communicating population genetics results.
 We've tried to make this a bit more clear in the introduction. \llname{ll:descriptive}
@@
     are not being driven by this effect:
-    as the reviewer also mentions, the effect is expected to dissappear when using a larger number of PCs,
+    as the reviewer also mentions, the effect is expected to disappear when using a larger number of PCs,
     which does not happen in our applications.

### 2018-10-29 writeup/review-responses_v1.tex
revise update

@@
 
-Besides theese additional verifications,
+Besides these additional verifications,
 we have tried to shift the focus away from background selection specifically

### 2018-10-29 writeup/local_pca_paper.tex
revise update

@@
 relatedness on sex chromosomes is expected to differ from the autosomes;
-and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
+and positive or negative selection on particular loci can dramatically distort shapes of nearby genealogies
 \citep{kim2002hitchhiking,charlesworth1993effect,barton2000genetic}.
@@
 without being affected by extreme differences in amount of missing data
-(Supplemental Figure \ref{sfig:test_missing}).
+(Supplementary Figure \ref{sfig:test_missing}).
 This simulation also verifies insensitivity to ordering of top PCs,
@@
 did not induce patterns in the MDS visualizations correlated with recombination rate
-(Supplemental Figure \ref{sfig:slim_sims1}).
+(Supplementary Figure \ref{sfig:slim_sims1}).
 \llabel{ll:test_recomb_rate}
@@
 in \Figure \ref{fig:mds_allchr} (groups are shown in color),
-and in Supplemental \Figure \ref{sfig:pca_by_pop} (PCA plots).
+and in Supplementary \Figure \ref{sfig:pca_by_pop} (PCA plots).
 The latter plots are quite different,
@@
 To depict this, \Figure \ref{fig:pca_by_inversion} shows
-the PCA plots in Supplemental \Figure \ref{sfig:pca_by_pop} recolored by the orientation of the inversion for each sample.
+the PCA plots in Supplementary \Figure \ref{sfig:pca_by_pop} recolored by the orientation of the inversion for each sample.
 Taking chromosome arm 2L as an example,
@@
 and have patterns of relatedness reflective of two axes of variation,
-seen in \Figure \ref{fig:pca_by_inversion} and Supplemental \Figure \ref{sfig:pca_by_pop},
+seen in \Figure \ref{fig:pca_by_inversion} and Supplementary \Figure \ref{sfig:pca_by_pop},
 which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
@@
 In the result, shown in \Figure \ref{fig:drosophila_recomb_rate} and
-Supplemental \Figure \ref{sfig:mds_allchr_noinversions},
+Supplementary \Figure \ref{sfig:mds_allchr_noinversions},
 the striking peaks associated with inversion breakpoints are gone,
@@
 there is also a relatively small dip on the right end of the arm (near the centromere).
-In contrast, Supplemental \Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
+In contrast, Supplementary \Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
 remaining structure is dominated by the dip near the centromere.
@@
         (A plot analogous to \Figure \ref{fig:mds_allchr} is shown
-        in Supplemental \Figure \ref{sfig:mds_allchr_noinversions}.)
+        in Supplementary \Figure \ref{sfig:mds_allchr_noinversions}.)
         The second plot (in blue) shows
@@
 the first MDS coordinate of each window in \Figure \ref{fig:mds_medicago},
-and are significantly correlated, as shown in Supplemental \Figure \ref{sfig:mds_gene_count}.
+and are significantly correlated, as shown in Supplementary \Figure \ref{sfig:mds_gene_count}.
 (Values shown are the number of start and end positions of each predicted mRNA transcript,
@@
        and the remaining corner (green) intermediate.
-       Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:mds_medicago_allchr}.
+       Plots for remaining chromosomes are shown in Supplementary \Figure \ref{sfig:mds_medicago_allchr}.
        \textbf{(bottom)}
@@
         Each point corresponds to a sample, colored by country of origin.
-        Plots for remaining chromosomes are shown in Supplemental \Figure \ref{sfig:pca_peaks_medicago_allchr}.
+        Plots for remaining chromosomes are shown in Supplementary \Figure \ref{sfig:pca_peaks_medicago_allchr}.
        \label{fig:mds12_medicago}
@@
     \caption{
-        MDS plots for human chromosomes 9-16, as in Supplemental Figure \ref{sfig:mds12_chr1_8_human}.
+        MDS plots for human chromosomes 9-16, as in Supplementary Figure \ref{sfig:mds12_chr1_8_human}.
         \label{sfig:mds12_chr9_16_human}
@@
     \caption{
-        MDS plots for human chromosomes 17-22, as in Supplemental Figure \ref{sfig:mds12_chr1_8_human}.
+        MDS plots for human chromosomes 17-22, as in Supplementary Figure \ref{sfig:mds12_chr1_8_human}.
         \label{sfig:mds12_chr17_22_human}

### 2018-10-29 writeup/review-responses.tex
tweaks to the responses

@@
 
-We are pleased to submit a revised version of our manuscript, 
+We are writing to submit a revised version of our manuscript, 
 ``Local PCA shows how the effect of population structure differs along the genome''.
+We appreciate the thoughtful comments by the reviewers and the editors,
+and have made changes to address all of the points raised.
+The changes are mostly to what is discussed;
+we sympathize with the additional testing suggested by Reviewer 1,
+but think that it falls outside the scope of this paper,
+and would be appropriate for follow-up work that builds on our descriptive method
+to create more specific, hypothesis-driven tools.
+
+We hope you find the revised manuscript to be as improved as we do.
 
@@
     along the genome - as written, the discussion is strongly focused on the
-    "classical" Drosophila-centred linked selection literature.  Specifically,
+    ``classical'' Drosophila-centred linked selection literature.  Specifically,
     there has been a lot of discussion in the phylogenetics world about how to
     estimate species trees from the set of locus-specific genealogies (where the
-    "consensus" tree can be quite wrong); and there are methods that use the
+    ``consensus'' tree can be quite wrong); and there are methods that use the
     distribution of genealogical relationships to estimate population structure
@@
     Good suggestion; we've added (hopefully helpful) additional discussion.
-    \revref and \revreffull{1}{2}
+    {\revref} and \revreffull{1}{2}
 }
@@
     We don't share the reviewer's concern about variation in recombination rate:
-    a region of low recombination could be dominated by a single tree,
+    although a region of low recombination could be dominated by a single tree,
     and therefore provide a strongly divergent PCA signal,
@@
     across a large fraction of a chromosome,
-    unless that chromosomal region had much lower recombination rate than we see here.
+    as we see here.
+    We've tried to make this argument more clear. {\llname{ll:not_neutral}}
     More generally, we heartily agree that we would love to have available a test
@@
 \begin{point}{}
-    The authors refer to supplementary Figures sometimes as "Figure S6" and sometimes as "Supplementary Figure S6".
+    The authors refer to supplementary Figures sometimes as ``Figure S6'' and sometimes as ``Supplementary Figure S6''. 
     In certain journals, these two could denote different figures so I would select one way only.
@@
 The authors didn't address the intention of my concern about the switching of PCs.
-To be very explicit: when you perform a PC analysis, there is some variability about the order that the "true" directions of the data comes out in terms of the magnitude of that PC (i.e.
+To be very explicit: when you perform a PC analysis, there is some variability about the order that the ``true'' directions of the data comes out in terms of the magnitude of that PC (i.e.
 the order).
@@
 However, if the magnitude of the noise is high, then it is quite plausible that the top K components in a segment of genome are not the top K for the whole genome.
-The simulation is useless for reassuring on this because it sets up a situation with 2 important PCs which are always included, and the claim that "the simulation also verifies insensitivity to ordering of top PCs" is not really true.
+The simulation is useless for reassuring on this because it sets up a situation with 2 important PCs which are always included, and the claim that ``the simulation also verifies insensitivity to ordering of top PCs'' is not really true.
 The real data analyses are more reassuring, given the relatively similar results between K=2 and K=5, but I think the concern remains valid.
 
-I would like to see evidence that the windows that are found by this method are not associated with "which" PCs happen to be included in the top K.
-To do that, you could for example calculate the top $M>K$ pcs, check the correlation with the overall PCs, and then plot "presence" of a PC (measured in correlation in the top K) against the MDS values.
+    I would like to see evidence that the windows that are found by this method are not associated with ``which'' PCs happen to be included in the top K.
+To do that, you could for example calculate the top $M>K$ pcs, check the correlation with the overall PCs, and then plot ``presence'' of a PC (measured in correlation in the top K) against the MDS values.
 It would alternatively be possible to make an argument that it will always be fine if you use a large enough number of PCs, but this is argued to have severe computational problems (an argument I don't know that I buy).
-It would be enough to do this for retaining *one* PC in the simulation that has 2 *equally strong* PCs, but I have a suspicion that this case goes quite wrong and is anyway unrealistic.
+It would be enough to do this for retaining *one* PC in the simulation that has 2 \emph{equally strong} PCs, but I have a suspicion that this case goes quite wrong and is anyway unrealistic.
 Better to keep more, e.g.
@@
 \reply{
-    Apologies for not realizing the precise nature of your original concern.
+    Apologies for not realizing the precise nature of the original concern.
     Yes, we agree that if strongly different PCs switched order 
@@
     then that could easily show up as (spurious) structure in our MDS plots.
-    We have added discussion of this point. \revref
-    However, for our empirical data we think we have strong evidence that the differences between windows 
-    are not being driven by this effect:
-    as the reviewer also mentions, the effect is expected to disappear when using a larger number of PCs,
-    which does not happen in our applications.
+    However, as the reviewer says,
+    spurious structure created by this problem would dissappear when using a larger number of PCs
+    -- this does not happen with our empirical results.
+    We have added discussion of these points. {\revref}
     Furthermore, and speculatively,

### 2018-10-29 writeup/local_pca_paper.tex
tweaks to the responses

@@
 but rather to describe larger-scale variation shared by many parts of the genome,
-and give clues about its source. \llabel{ll:descriptive}
+and give clues about the source of this variation. \llabel{ll:descriptive}
 % If such variation is found, correlation of this variation with known genomic features
@@
 there is also a relatively small dip on the right end of the arm (near the centromere).
-In contrast, Supplementary \Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
+In contrast, Figure \ref{fig:drosophila_recomb_rate} 
+and Supplementary \Figure \ref{sfig:mds_allchr_noinversions} 
+show that after removing polymorphic inversions,
 remaining structure is dominated by the dip near the centromere.
@@
 \citep{avise1983mitochondrial,pamilo1988relationships,hobolth2007genomic}.
-Loosely, we are studying the within-species patterns that might lead to such
-incomplete lineage sorting in the future.
+The within-species patterns we observe might contribute to such
+incomplete lineage sorting among future descendant species of a given population.
 The neutral distribution of variation in these patterns 
@@
 Another possible contributor is recent admixture between previously separated populations,
-the effects of which were not uniform across the genome due to selection. \revpoint{2}{6}
+the effects of which were not uniform across the genome due to selection. % \revpoint{2}{6}
 For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
@@
 These have been difficult to disentangle in part because for the most part
-theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
-and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
-It may be possible to design more powerful statistics that make stronger use of spatial information,
-or even using the variation in relatedness that we observe here,
+theory makes predictions which are only strictly valid 
+in randomly mating (i.e., unstructured) populations,
+and it is unclear to what extent the spatial structure observed in most real populations 
+will affect these predictions.
+Developing a method to distinguish these forms of selection from each other
+and from the effects of demography is a major challenge to the field.
+It may be possible to make progress using statistics that make stronger use of spatial information,
+such as the variation in relatedness that we observe here,
 similar to the method of \citet{beeravolu2018blockwise}. \revpoint{1}{2}
@@
 but in our exploration, changing these has not produced dramatically different results.
-However, this choice could easily be important:
+However, this choice could in some situations be important:
 for instance, if the $k^\text{th}$ and $(k+1)^\text{st}$ PCs 

### 2018-10-30 writeup/review-responses.tex
final tweaks

@@
 
-We hope you find the revised manuscript to be as improved as we do.
+We have also included a color-coded manuscript highlighting differences 
+between this version and the last one (but the line numbers referred to in the 
+``response to reviewers'' refer to the standard manuscript).
+The file is named ``\url{local_pca_paper-diff164c6.pdf}'',
+and is labeled another ``response to reviewers'' for lack of a better category in the submission system.
+
+\vspace{5em}
 
@@
     These are excellent questions for a method designed to detect inversions.
-    However, we do not present such a method,
-    we rather present a visualization method in which, empirically, inversions often leave a large (but not quantified) signal.
+    However, we do not set out to make such a method,
+    but rather a visualization method in which, empirically, inversions often leave a large signal.
     We are not sure if it would be productive to turn this into a quantitative method, either:
-    as we have now clarified in the paper \llname{ll:inversion_caveat},
+    as we have further clarified in the paper \llname{ll:inversion_caveat},
     distinguishing inversions from regions of low recombination
@@
     However, this is one of the major unsolved question in the field,
-    and is subject to substantial debate 
-    (see for instance \citet{jensen2014unfounded,harris2018unfounded}
-    and subsequent rejoinders).
+    and is subject to substantial debate.
+    % (see for instance \citet{jensen2014unfounded,harris2018unfounded} and subsequent rejoinders).
     We've added more to the discussion on this. \revref
@@
     are relatively stable,
-    and so might help avoid sensitive PC switching.
+    and so might help prevent sensitivity to PC switching.
 }

### 2018-10-30 writeup/local_pca_paper.tex
final tweaks

@@
 by applying principal components analysis (PCA, \citet{patterson2006population}) 
-to the genetic data matrix,
+to the genotype matrix,
 as inspired by the pioneering work of \citet{menozzi1978synthetic}. \revpoint{AE}{3}
@@
 (\SFigure \ref{sfig:slim_sims2}).
-The method strongly identified the regions under spatially varying linked selection
-(also our simulation with strongest positive selection). \revpoint{2}{2}
+The method strongly identified the distinct regions 
+in the simulation with spatially varying linked selection.
 It also identified the regions (although less unambiguously)
@@
 but did not clearly identify them with constant recombination rate.
-This difference is likely because recombination rates are overall lower in the first case,
-leading to a stronger effect of linked selection.
+The differences in power between these three scenarios are likely
+explained by varying strength of linked selection -- 
+the simulation with spatially varying selection was also the case with strongest positive selection,
+and recombination rates were overall lower in the simulation with stepwise varying recombination rate
+than in the simulation with constant rate. \revpoint{2}{2}
 These tests are not meant to be comprehensive survey of linked selection,

### 2019-06-05 lostruct/R/pc_dist.R
avoid pmin

@@ -26,14 +26,15 @@ pc_dist <- function( x, npc=attr(x,"npc"), w=1, normalize="L1", mc.cores=1 ) {
     n <- nrow(values)
     # could speed this up by a factor of two taking advantage of symmetry
     emat <- function (u) { matrix(u,ncol=npc) }
-    out <- do.call( rbind, this.lapply( 1:n, function (i) {
+    out <- do.call( cbind, this.lapply( 1:n, function (i) {
                 sapply( 1:n, function (j) {
                     dist_sq_from_pcs( values[i,], emat(vectors[i,]), values[j,], emat(vectors[j,]) )
                 } )
             } ) )
     # symmetrize and truncate negative numbers (which are machine error)
-    out <- sqrt(pmax(0,(out+t(out))/2))
-    dim(out) <- c(n,n)
+    out <- (out + t(out))/2
+    out[out < 0] <- 0
+    out <- sqrt(out)
     return(out)
 }
 

### 2020-09-13 templated/summarize_run.Rmd

@@
 # original data files
-chroms <- opt$bcf_file_names
+chroms <- opt$chrom_names
 bcf.files <- opt$bcf_files
@@
 (this is done without a ton of memory by accumulating the covariance matrix in `running_cov`):
-```{r corner_pca, cache=TRUE}
+```{r corner_pca, cache=TRUE, echo=-1}
+options(datatable.fread.input.cmd.message=FALSE)
 corner.npc <- 4

### 2021-02-24 templated/summarize_run.Rmd

@@
     samp.file <- opt$sample_info
-    samps <- read.table(samp.file,sep="\t",header=TRUE)
+    samps <- read.table(samp.file,sep="\t",header=TRUE, stringsAsFactors=TRUE)
     names(samps) <- tolower(names(samps))

### 2021-05-20 medicago/summarize_run.Rmd
more stringsAsFactors

@@
 # read in mds
-mds.coords <- read.csv(mds.file,header=TRUE)
+mds.coords <- read.csv(mds.file,header=TRUE,stringsAsFactors=TRUE)
 mds.cols <- (1:ncol(mds.coords))[-(1:2)]
@@
 # position information
-regions <- do.call( rbind, lapply( regions.files, read.csv, header=TRUE ) )
+regions <- do.call( rbind, lapply( regions.files, read.csv, header=TRUE,stringsAsFactors=TRUE ) )
 # figure out where to plot things at

### 2021-05-20 templated/summarize_run.Rmd
more stringsAsFactors

@@
 # read in mds
-mds.coords <- read.csv(mds.file,header=TRUE)
+mds.coords <- read.csv(mds.file,header=TRUE, stringsAsFactors=TRUE)
 mds.cols <- (1:ncol(mds.coords))[-(1:2)]
@@
 # position information
-regions <- do.call( rbind, lapply( regions.files, read.csv, header=TRUE ) )
+regions <- do.call( rbind, lapply( regions.files, read.csv, header=TRUE, stringsAsFactors=TRUE ) )
 # figure out where to plot things at
